### 2021-04-13 scripts/ZhangetalReanalysis.Rmd
first commit of files

---
title: "Forest cover runoff relationship"
author: "Willem Vervoort, Eliana Nervi, Jimena Alonso"
date: "01/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
require(tidyverse)
require(lubridate)
require(mgcv) # to do the generalise modelling
```

## Introduction

Zhang et al. (2017) presented a review of changes in streamflow due to changes in forest cover based on a global data set. However, in the actual analysis, the authors used very basuic linear regression to look at individual variables and factors. However, these factors are fo course all interrelated, and we really should be analysing the marginal trends, taking into account the other factors and variables in the data set. The idea of this paper is to use the fantastic data set collected by Zhang et al. (2017) and to reanalyse the relationship between change in forest, but to use more advanced statistical techniques that allow discovery of the marginal effects of the different drivers of streamflow change.

In particular, we want to use generalised linear (GLM) and generalised additive models (GAM) to be able to include both factors and continuous variables, and to be able to analyses these drivers as non-linear effects (if there are clear hypotheses to support non-linear relationships).

## The data

The data consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small catchments (<1000 km2) and one table for large catchments (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the catchment originates from the small or large table.

```{r}
Zhang_small <- read_csv("TablesZhangetal2017_small.csv")
Zhang_large <- read_csv("TablesZhangetal2017_large.csv")
```

### change names of columns

the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple

```{r}
names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
```

## Simple generalised linear model

This will use the pacakge `mgcv`, which can do both generalised linear modelling and generalised additive modelling

We will start with a simple linear model which essentially predicts the change in streamflow $\Delta Qf\%$ as a function of the other variables:
$\Delta Qf\% ~ Area + Pa + Forest type + Hydrological regime + \Delta F\% + Precipitation data type + Assessment technique$

### Small catchments

```{r}
linear_model_small <- gam(DeltaQf_perc ~ DeltaF_perc + Area_km2 + Pa_mm + Forest_type + 
                            Hydrological_regime + Precip_data_type +  Assessment_technique, data = Zhang_small)
summary(linear_model_small)
gam.check(linear_model_small)

```
This shows that the change in forestry is a highly signficant predictor of the change in flow, and this indicates a decrease in the change in flow withe an increase in the change in forestry. In other words it shows the same relationship as figure 2 in Zhang et al.
However, it is dependent on rainfall (with higher rainfall showing steeper declines) and the type of forest (with Mixed forest types causing a greater decline in flow with an increase in forestry). None of the other variables are significant.

The residuals of the model are slightly problematic, they are skewed and show some increase in variance. This is probably due to the distribution of the y-variable (the change in flow).

Check distribution of the y-variable

```{r}
hist(Zhang_small$DeltaQf_perc)
min(Zhang_small$DeltaQf_perc)
```
Doing a log transformation on the y-variable is problematic as the change in flow ranges from -200% to +200%. However, it is possible when you divide the percentage by 100, and then use a log10(y + 1) transformation.

```{r}
linear_model_small_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + Area_km2 + Pa_mm + Forest_type + 
                            Hydrological_regime + Precip_data_type +  Assessment_technique, data = Zhang_small)
summary(linear_model_small_log)
gam.check(linear_model_small_log)
```
This improves the residuals, but the results of the analysis is the same.

### Large catchments

```{r}
linear_model_large <- gam(DeltaQf_perc ~ DeltaF_perc + Area_km2 + Pa_mm + Forest_type + Hydrological_regime + Precip_data_type +
                    Assessment_technique, data = Zhang_large)
summary(linear_model_large)
gam.check(linear_model_large)
```
The results for the last catchments shows that in this case the change in forestry is not significant at p < 0.05 as the p-value is 0.11. Increasing Rainfall is associated with a lower change in runoff, while having a snow dominated hydrological regime is associated with a higher change in runoff (at p <0.05). The "SH" assessment technique (combined use of hydrographs and statistical techniques) is also associated with a lower change in runoff.

The residuals are again a little skewed, again because the change in runoff percentage data is skewed.

```{r}
hist(Zhang_large$DeltaQf_perc)
```

Using a similar transformation as for the small catchments results on the change in streamflow, and a log transformation of the rainfall however results in quite a different model outcome

```{r}
linear_model_large_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + log10(Pa_mm) + Forest_type + Hydrological_regime + Precip_data_type +
                    Assessment_technique, data = Zhang_large)
summary(linear_model_large_log)
gam.check(linear_model_large_log)

```
The residuals are much better distributed. And the results now show that the change in forestry is a significant predictor, resulting in a decrease of the change in stream flow with an increase in the change in forestry.. In addition, increasing rainfall also decreases the change in streamflow. while snow dominated  regimes, hydrological modelling techniques and statistical and hydrograph approaches predict higher changes in streamflow.

## is the separation in large and small arbitrary

Stack the two data frames and run the GAM and check if the size of the catchment is significant.

```{r}
Zhang_all <- bind_rows(Zhang_small, Zhang_large)

linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + log10(Pa_mm) + Forest_type + Hydrological_regime + Precip_data_type +
                    Assessment_technique, data = Zhang_all)
summary(linear_model_all_log)
gam.check(linear_model_all_log)

```

## Work to be done

1. Extend the analysis to include a smooth non-linear rainfall response?  
2. In the paper, they also use PET which is not in the data, se we need to find some value of PET for each catchment. I wonder how you define PET for the whole of the Amazon?
3. We also probably want to extract the "continent" and I think also Latitude  
4. We also need to check if we know of more published data that is not in the current dataset

Key arguments we are making are:  
1. regression approach is too simple, ignores possible interactions and possible non-linear behaviour (for example rainfall).  
2. Is the 1000 km2 is an arbitrary cut-off?  Or is there an interaction with the assessment technique?  
3. Can you actually define "annual rainfall" or "annual PET" for a large catchment? Would there not be a major effect of spatial variability?

### 2021-05-27 paper/Forest_and_Water/Forest_and_Water.Rmd
updating paper

---
title: Do larger catchments respond different to forest cover change? Re-analysing a global data set.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: The University of Sydney, INIA
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
    footnote: 2
address:
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: INIA, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Departamento de Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

Introduction
==========================

## Introduction

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013]. The historic work highlights provides a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (or Australian watersheds) a 100% decrease in stream flow for watersheds with 100% forest cover. However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.

There have been several review papers aiming to summarize different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984] and more generally [@jackson2005; @zhang2017]. These studies are aiming to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent review [@zhang2017] developed an impressive database of watershed studies in relation to changes in streamflow due to changes in forest cover based on a global data set. This dataset, which covers over 250 studies are described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.

The conclusions of the paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^. While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover. Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed [@zhang2017].

Encouraged by the work presented by @zhang2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion by presenting further analysis of the data and by adding further watersheds and enhancements to the data base.

In particular, the main method in the work by @zhang2017 is using simple linear regression. And the main assumption is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds. Given the fantastic data set collected, the analysis can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.

In particular, the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds and spatial coordinates and 2) to analyse the possibility of non-linear and partial effects of the different factors and variables in the data base using generalised linear (GLM) and generalised additive models (GAM @wood2006). Finally we hope to point to further research that can expand our work and that outlined @zhang2017 to better understand the impact of forest cover change on streamflow.

Methods
==========

The staerting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this dataset

References {#references .unnumbered}
==========

### 2021-05-27 paper/Forest_and_Water/Forest_and_Water.tex
updating paper

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Journal of Hydrology} % Sets Journal name

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short Paper},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing

% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Short Paper}
    \author[The University of Sydney, INIA]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso\corref{2}}
   \ead{jalonso@fing.edu.uy} 
      \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney,
Sydney, NSW 2006, Australia}
    \address[INIA]{INIA, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
Engineering, Universidad de la República, 11200 Montevideo, Departamento
de Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}

 \end{frontmatter}

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

There has been an long and on-going discussion in the hydrologcal
literature around the impact of forests on streamflow (Andréassian,
2004; Brown et al., 2013, 2005; Jackson et al., 2005; Zhang et al.,
2017). The historic work highlights provides a general consensus that if
forest areas increase, streamflow decreases and vice-versa. The most
dramatic result in relation to this, is Figure 5 in Zhang et al. (2011)
indicating (or Australian watersheds) a 100\% decrease in stream flow
for watersheds with 100\% forest cover. However, on the other end of the
spectrum, in a series of French watersheds (Cosandey et al., 2005),
there was no change in streamflow characteristics in 2 of the three
watersheds studied in relation to deforestation.

There have been several review papers aiming to summarize different
studies across the globe, in relation to paired watershed studies (Bosch
and Hewlett, 1982; Brown et al., 2005) and more generally (Jackson et
al., 2005; Zhang et al., 2017). These studies are aiming to generalize
the individual findings and to identify if there are global trends or
relationships that can be developed. The most recent review (Zhang et
al., 2017) developed an impressive database of watershed studies in
relation to changes in streamflow due to changes in forest cover based
on a global data set. This dataset, that covers over 250 studies are
described in terms of the change in streamflow as a result of the change
in forest cover, where studies related to both forestation (increase in
forest cover) and deforestation (decrease in forest cover) were
included.

The conclusions of the paper (Zhang et al., 2017) suggest that there is
a distinct difference in the change in flow as a result of forestation
or deforestation between small watersheds, defined as \textless{} 1000
km\textsuperscript{2} and large watersheds \textgreater{} 1000
km\textsuperscript{2}. While for small watersheds there was no real
change in runoff with changes in cover, for large watersheds there was a
clear trend showing a decrease in runoff with and increase in forest
cover. Their main conclusion was that the response in annual runoff to
forest cover was scale dependent and appeared to be more sensitive to
forest cover change in water limited watersheds relative to energy
limited watershed (Zhang et al., 2017).

Encouraged by the work presented by Zhang et al. (2017) and the
fantastic database of studies presented by these authors, we believe we
can add to the discussion by presenting further analysis of the data and
by adding further watersheds and enhancements to the data base.

In particular, the main method in the work by Zhang et al. (2017) is
using simple linear regression. And the main assumption is that the
threshold at 1000 km\textsuperscript{2} is a distinct separation between
``small'' and ``large'' watersheds. Given the fantastic data set
collected, the analysis can be easily expanded to look at interactions
between the terms and to test the assumption of a distinct threshold at
1000 km\textsuperscript{2}.

In particular, the objective of this paper is to 1) enhance the data set
from ({\textbf{???}}) with further watersheds and spatial coordinates
and 2) to analyse the possibility of non-linear and partial effects of
the different factors and variables in the data base using generalised
linear (GLM) and generalised additive models (GAM Wood (2006)). Finally
we hope to point to further research that can expand our work and that
outlined Zhang et al. (2017) to better understand the impact of forest
cover change on streamflow.

\hypertarget{front-matter}{%
\section{Front matter}\label{front-matter}}

The author names and affiliations could be formatted in two ways:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Group the authors per affiliation.
\item
  Use footnotes to indicate the affiliations.
\end{enumerate}

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-andreassian2004}{}%
Andréassian, V., 2004. Waters and forests: From historical controversy
to scientific debate. Journal of Hydrology 291, 1--27.
doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2003.12.015}{https://doi.org/10.1016/j.jhydrol.2003.12.015}

\leavevmode\hypertarget{ref-hewlett1984}{}%
Bosch, J.M., Hewlett, J.D., 1982. A review of catchment experiments to
determine the effect of vegetation changes on water yield and
evapotranspiration. Journal of Hydrology 55, 3--23.

\leavevmode\hypertarget{ref-brown2013}{}%
Brown, A.E., Western, A.W., McMahon, T.A., Zhang, L., 2013. Impact of
forest cover changes on annual streamflow and flow duration curves.
Journal of Hydrology 483, 39--50.
doi:\href{https://doi.org/http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}

\leavevmode\hypertarget{ref-brown2005}{}%
Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A.,
2005. A review of paired catchment studies for determining changes in
water yield resulting from alterations in vegetation. Journal of
Hydrology 310, 28--61.

\leavevmode\hypertarget{ref-cosandey2005}{}%
Cosandey, C., Andréassian, V., Martin, C., Didon-Lescot, J.F., Lavabre,
J., Folton, N., Mathys, N., Richard, D., 2005. The hydrological impact
of the mediterranean forest: A review of french research. Journal of
Hydrology 301, 235--249.
doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2004.06.040}{https://doi.org/10.1016/j.jhydrol.2004.06.040}

\leavevmode\hypertarget{ref-jackson2005}{}%
Jackson, R.B., Jobbagy, E.G., Avissar, R., Roy, S.B., Barrett, D.J.,
Cook, C.W., Farley, K.A., Maitre, D.C. le, McCarl, B.A., Murray, B.C.,
2005. Trading water for carbon with biological carbon sequestration.
Science 310, 1944--1947.
doi:\href{https://doi.org/10.1126/science.1119282}{10.1126/science.1119282}

\leavevmode\hypertarget{ref-wood2006}{}%
Wood, S., 2006. Generalized additive models: An introduction with r. CRC
Press, Boca Raton, FL.

\leavevmode\hypertarget{ref-zhang2011}{}%
Zhang, L., Zhao, F., Chen, Y., Dixon, R.N.M., 2011. Estimating effects
of plantation expansion and climate variability on streamflow for
catchments in australia. Water Resources Research 47, W12539.
doi:\href{https://doi.org/10.1029/2011wr010711}{10.1029/2011wr010711}

\leavevmode\hypertarget{ref-zhang2017}{}%
Zhang, M., Liu, N., Harper, R., Li, Q., Liu, K., Wei, X., Ning, D., Hou,
Y., Liu, S., 2017. A global review on hydrological responses to forest
change across multiple spatial scales: Importance of scale, climate,
forest type and hydrological regime. Journal of Hydrology 546, 44--59.
doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2016.12.040}{https://doi.org/10.1016/j.jhydrol.2016.12.040}

\end{document}

### 2021-05-27 scripts/ZhangetalReanalysis.Rmd
updating paper

@@
 require(lubridate)
-require(mgcv) # to do the generalise modelling
+require(mgcv) # to do the generalised modelling
 ```
@@
 
-Zhang et al.
-(2017) presented a review of changes in streamflow due to changes in forest cover based on a global data set.
-However, in the actual analysis, the authors used very basuic linear regression to look at individual variables and factors.
-However, these factors are fo course all interrelated, and we really should be analysing the marginal trends, taking into account the other factors and variables in the data set.
-The idea of this paper is to use the fantastic data set collected by Zhang et al.
-(2017) and to reanalyse the relationship between change in forest, but to use more advanced statistical techniques that allow discovery of the marginal effects of the different drivers of streamflow change.
+There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013].
+The historic work highlights provides a general consensus that if forest areas increase, streamflow decreases and vice-versa.
+The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (or Australian watersheds) a 100% decrease in stream flow for watersheds with 100% forest cover.
+However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
 
-In particular, we want to use generalised linear (GLM) and generalised additive models (GAM) to be able to include both factors and continuous variables, and to be able to analyses these drivers as non-linear effects (if there are clear hypotheses to support non-linear relationships).
+There have been several review papers aiming to summarize different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1982] and more generally [@jackson2005; @zhang2017].
+These studies are aiming to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
+The most recent review [@zhang2017] developed an impressive database of watershed studies in relation to changes in streamflow due to changes in forest cover based on a global data set.
+This dataset, that covers over 250 studies are described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+
+The conclusions of the paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^.
+While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover.
+Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed [@zhang2017].
+
+Encouraged by the work presented by @zhang2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion by presenting further analysis of the data and by adding further watersheds and enhancements to the data base.
+
+In particular, the main method in the work by @zhang2017 is using simple linear regression.
+And the main assumption is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds.
+Given the fantastic data set collected, the analysis can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.
+
+In particular, the objective of this paper is to 1) enhance the data set from @Zhang2017 with further watersheds and spatial coordinates and 2) to analyse the possibility of non-linear and partial effects of the different factors and variables in the data base using generalised linear (GLM) and generalised additive models (GAM @wood2006).
+Finally we hope to point to further research that can expand our work and that outlined @zhang2017 to better understand the impact of forest cover change on streamflow.
 
@@
 
-The data consists of the two tables in the supplementary data from Zhang et al.
-(2017), which means there is one table for small catchments (<1000 km2) and one table for large catchments (> 1000 km2).
-We might combine these tables as we can easily add a variable indicating whether the catchment originates from the small or large table.
+The starting point consists of the two tables in the supplementary data from Zhang et al.
+(2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2).
+The columns in the data represent We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
 
@@
 
-### Small catchments
+### Small watersheds
 
@@
 
-### Large catchments
+### Large watersheds
 
@@
 ```
-The results for the last catchments shows that in this case the change in forestry is not significant at p < 0.05 as the p-value is 0.11.
+The results for the last watersheds shows that in this case the change in forestry is not significant at p < 0.05 as the p-value is 0.11.
 Increasing Rainfall is associated with a lower change in runoff, while having a snow dominated hydrological regime is associated with a higher change in runoff (at p <0.05).
@@
 
-Using a similar transformation as for the small catchments results on the change in streamflow, and a log transformation of the rainfall however results in quite a different model outcome
+Using a similar transformation as for the small watersheds results on the change in streamflow, and a log transformation of the rainfall however results in quite a different model outcome
 
@@
 
-Stack the two data frames and run the GAM and check if the size of the catchment is significant.
+Stack the two data frames and run the GAM and check if the size of the watershed is significant.
 
@@
 1. Extend the analysis to include a smooth non-linear rainfall response?  
-2. In the paper, they also use PET which is not in the data, se we need to find some value of PET for each catchment. I wonder how you define PET for the whole of the Amazon?
+2. In the paper, they also use PET which is not in the data, se we need to find some value of PET for each watershed. I wonder how you define PET for the whole of the Amazon?
 3. We also probably want to extract the "continent" and I think also Latitude  
@@
 2. Is the 1000 km2 is an arbitrary cut-off?  Or is there an interaction with the assessment technique?  
-3. Can you actually define "annual rainfall" or "annual PET" for a large catchment? Would there not be a major effect of spatial variability?
+3. Can you actually define "annual rainfall" or "annual PET" for a large watershed? Would there not be a major effect of spatial variability?
+4. time period after which measurement of decrease in flow is reached: harvesting, growth cycles etc. There is a need for standardisation or incorporation of this variability.

### 2021-05-30 paper/Forest_and_Water/Forest_and_Water.Rmd
paper intro & methods and new watersheds

@@
 
-The staerting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this dataset
+## The original data set
+
+The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
+The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in % ($\Delta$F%) and the change in streamflow in % ($\Delta$Qf(%), based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
+
+Table 1  Summary of abbreviations of factors used in the @zhang2017 data set
+
+|Factor | Abbreviation | Definition |
+|:------------|:-------|:----------|
+| forest type | CF |coniferous forest | 
+|  | BF | broadleaf forest |
+|  | MF |mixed forest  |
+|hydrological regime|RD|rain dominated|
+| |SD|snow dominated|
+| climate type| EL| energy limited|
+| |WL| water limited|
+| |EQ| equitant|
+|precipitation data type|OB|observed|
+| |SG|spatial gridded|
+| |MD|modelled|
+|assessment technique|PWE|paired watershed experiment|
+| |QPW|quasi-paired watershed experiment|
+| |HM|hydrological modelling|
+| |EA|elastictity analysis|
+| |SH|combined use of statistical methods|
+| | |and hydrographs|
+
+While the @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration is not part of the data set.
+We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watreshed data sets in our analysis.
+Some small naming errors and citations for some of the data sets for some of the small watersheds were fixed as we were familiar with the studies.
+But overall the original data set was not changed.
+
+## Additional data collection
+
+To enhance the existing data set, this study added additional variables.
+The first variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
+Using this information  annual average reference evapotranspiration (E0) was extracted from **XXXXX** if value of E0 were not available from the original papers.
+For large water watersheds, this value, similar to annual average rainfall, is an approximation of the climate at the location.
+
+The length of the study can be a variable influencing the change in flow [e.g.
+@jackson2005] and therefore, the length of the different studies was extracted from the references provided by @zhang2017.
+
+Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
+In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
+Table 6 in @zhang2011)
+
+The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research.
+
+## Statistical modelling
+
+To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
+In particular we fitted the following initial model to replicate the variables in the analysis from @zhang2017:
+
+$$\tag{1}
+\begin{aligned}
+\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}$$
+
+However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals.
+As a result we transformed $\Delta \%Q$ and $\Delta \%forest$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is either $\Delta Q$ or $\Delta forest$.
+This means that the model residuals are $\sim N(0,\sigma^2)$ results in the following equation:
+
+$$\tag{2}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}$$
+
+In this model, the assumption is that all continuous variables (such as Pa) can have a linear or non-linear relationship with $log10(\Delta Q)$.
+This means that a smooth function $s()$ is applied to the variable.
+To restrict the smoothness of the fit, the smoothness factor $k$ is restricted to a value of 3 [@wood2006].
+This restriction was applied to smooth variables throughout this paper and we have dropped this from the subsequent equations.
+
+For the model in equation 2, we only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the analysis.
+Given that in @zhang2017, dryness ($\frac{E0}{Pa}$) is used to look at variations in the change in flow, we also fitted the following model:
+
+$$\tag{3}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}$$
+
+Subsequently, using the full data set, including the additional watersheds and the additional variables the following two models were fitted:
+
+$$\tag{4}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
+& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+\end{aligned}$$
+
+$$\tag{5}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
+& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+\end{aligned}$$
+
+The results were analysed to identify:  
+1. the significance of the different variables  
+2. the direction of the categorical or shape of the smooth variables
 

### 2021-05-30 paper/Forest_and_Water/Forest_and_Water.tex
paper intro & methods and new watersheds

@@
 \bibliographystyle{elsarticle-harv}
+\usepackage{longtable}
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Short Paper},
+            pdftitle={Do larger catchments respond different to forest cover change? Re-analysing a global data set.},
             colorlinks=false,
@@
 
-  \title{Short Paper}
+  \title{Do larger catchments respond different to forest cover change?
+Re-analysing a global data set.}
     \author[The University of Sydney, INIA]{R. Willem Vervoort\corref{1}}
@@
  \end{frontmatter}
-
-\emph{Text based on elsarticle sample manuscript, see
-\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}
 
@@
 relation to changes in streamflow due to changes in forest cover based
-on a global data set. This dataset, that covers over 250 studies are
+on a global data set. This dataset, which covers over 250 studies are
 described in terms of the change in streamflow as a result of the change
@@
 In particular, the objective of this paper is to 1) enhance the data set
-from ({\textbf{???}}) with further watersheds and spatial coordinates
+from Zhang et al. (2017) with further watersheds and spatial coordinates
 and 2) to analyse the possibility of non-linear and partial effects of
@@
 
-\hypertarget{front-matter}{%
-\section{Front matter}\label{front-matter}}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-The author names and affiliations could be formatted in two ways:
+\hypertarget{the-original-data-set}{%
+\subsection{The original data set}\label{the-original-data-set}}
 
-\begin{enumerate}
-\def\labelenumi{(\arabic{enumi})}
-\item
-  Group the authors per affiliation.
-\item
-  Use footnotes to indicate the affiliations.
-\end{enumerate}
+The starting point of this paper is the data base of studies which were
+included in Zhang et al. (2017) as supplementary material. The columns
+in this data set are the watershed number, the watershed name, the Area
+in km\textsuperscript{2}, the annual average precipitation (Pa) in mm,
+the forest type, hydrological regime, and climate type, the change in
+forest cover in \% (\(\Delta\)F\%) and the change in streamflow in \%
+(\(\Delta\)Qf(\%), based on equation 1 in Zhang et al. (2017)), the
+precipitation data type, the assessment technique, and the source of the
+info, which is a citation. Several of these columns contain
+abbreviations to describe the different variables, which are summarised
+in Table 1.
 
-See the front matter of this document for examples. You are recommended
-to conform your choice to the journal you are submitting to.
+Table 1 Summary of abbreviations of factors used in the Zhang et al.
+(2017) data set
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+Factor & Abbreviation & Definition\tabularnewline
+\midrule
+\endhead
+forest type & CF & coniferous forest\tabularnewline
+& BF & broadleaf forest\tabularnewline
+& MF & mixed forest\tabularnewline
+hydrological regime & RD & rain dominated\tabularnewline
+& SD & snow dominated\tabularnewline
+climate type & EL & energy limited\tabularnewline
+& WL & water limited\tabularnewline
+& EQ & equitant\tabularnewline
+precipitation data type & OB & observed\tabularnewline
+& SG & spatial gridded\tabularnewline
+& MD & modelled\tabularnewline
+assessment technique & PWE & paired watershed experiment\tabularnewline
+& QPW & quasi-paired watershed experiment\tabularnewline
+& HM & hydrological modelling\tabularnewline
+& EA & elastictity analysis\tabularnewline
+& SH & combined use of statistical methods\tabularnewline
+& & and hydrographs\tabularnewline
+\bottomrule
+\end{longtable}
+
+While the Zhang et al. (2017) use the dryness index in their analysis,
+potential or reference evapotranspiration is not part of the data set.
+We combined the tables for small (\textless{} 1000
+km\textsuperscript{2}) and large (\textgreater= 1000
+km\textsuperscript{2}) watreshed data sets in our analysis. Some small
+naming errors and citations for some of the data sets for some of the
+small watersheds were fixed as we were familiar with the studies. But
+overall the original data set was not changed.
+
+\hypertarget{additional-data-collection}{%
+\subsection{Additional data
+collection}\label{additional-data-collection}}
+
+To enhance the existing data set, this study added additional variables.
+The first variables added were the latitude and longitude for the center
+of the watershed as an approximation of its spatial location. Using this
+information annual average reference evapotranspiration (E0) was
+extracted from \textbf{XXXXX} if value of E0 were not available from the
+original papers. For large water watersheds, this value, similar to
+annual average rainfall, is an approximation of the climate at the
+location.
+
+The length of the study can be a variable influencing the change in flow
+(e.g.~Jackson et al., 2005) and therefore, the length of the different
+studies was extracted from the references provided by Zhang et al.
+(2017).
+
+Several additional data points from watershed studies were extracted
+from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988),
+Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010),
+Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were
+checked against the existing studies to prevent overlap. In the citation
+column in the data set, in general the main reference for the calculated
+change in streamflow was used, because sometimes the original study did
+not provide the quantification of the change in streamflow (i.e.~Table 6
+in Zhang et al. (2011))
+
+The final column in the improved data set is a ``notes'' column, which
+is not further used in the analysis, but gives context to some of the
+data for future research.
+
+\hypertarget{statistical-modelling}{%
+\subsection{Statistical modelling}\label{statistical-modelling}}
+
+To estimate how the change in streamflow is affected by the change in
+forest cover while considering the effects of the other variables, we
+applied generalised additive modelling (GAM) (Wood, 2006). In particular
+we fitted the following initial model to replicate the variables in the
+analysis from Zhang et al. (2017):
+
+\[\tag{1}
+\begin{aligned}
+\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}\]
+
+However, the overall skewed distribution of the predictant
+(\(\Delta \%Q\)) is problematic, and this results in a skewed
+distribution of the GAM model residuals. As a result we transformed
+\(\Delta \%Q\) and \(\Delta \%forest\) back to fractions (0 - 1) and log
+transformed using \(log10(x + 1)\), where \(x\) is either \(\Delta Q\)
+or \(\Delta forest\). This means that the model residuals are
+\(\sim N(0,\sigma^2)\) results in the following equation:
+
+\[\tag{2}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}\]
+
+In this model, the assumption is that all continuous variables (such as
+Pa) can have a linear or non-linear relationship with
+\(log10(\Delta Q)\). This means that a smooth function \(s()\) is
+applied to the variable. To restrict the smoothness of the fit, the
+smoothness factor \(k\) is restricted to a value of 3 (Wood, 2006). This
+restriction was applied to smooth variables throughout this paper and we
+have dropped this from the subsequent equations.
+
+For the model in equation 2, we only used the data from Zhang et al.
+(2017) to make sure that the additional watersheds added to the data set
+did not influence the analysis. Given that in Zhang et al. (2017),
+dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
+in flow, we also fitted the following model:
+
+\[\tag{3}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\end{aligned}\]
+
+Subsequently, using the full data set, including the additional
+watersheds and the additional variables the following two models were
+fitted:
+
+\[\tag{4}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
+& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+\end{aligned}\]
+
+\[\tag{5}
+\begin{aligned}
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
+& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+\end{aligned}\]
+
+The results were analysed to identify:\\
+1. the significance of the different variables\\
+2. the direction of the categorical or shape of the smooth variables
 
@@
 doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2003.12.015}{https://doi.org/10.1016/j.jhydrol.2003.12.015}
+
+\leavevmode\hypertarget{ref-borg1988}{}%
+Borg, H., Bell, R.W., Loh, I.C., 1988. Streamflow and stream salinity in
+a small water supply catchment in southwest western australia after
+reforestation. Journal of Hydrology 103, 323--333.
+doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(88)90141-2}{https://doi.org/10.1016/0022-1694(88)90141-2}
 
@@
 
+\leavevmode\hypertarget{ref-pena-arancibia2012}{}%
+Peña-Arancibia, J.L., Dijk, A.I.J.M. van, Guerschman, J.P., Mulligan,
+M., Bruijnzeel, L.A., McVicar, T.R., 2012. Detecting changes in
+streamflow after partial woodland clearing in two large catchments in
+the seasonal tropics. Journal of Hydrology 416-417, 60--71.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2011.11.036}{https://doi.org/10.1016/j.jhydrol.2011.11.036}
+
+\leavevmode\hypertarget{ref-rodriguez2010}{}%
+Rodriguez, D.A., Tomasella, J., Linhares, C., 2010. Is the forest
+conversion to pasture affecting the hydrological response of amazonian
+catchments? Signals in the ji-paraná basin. Hydrological Processes 24,
+1254--1269.
+doi:\href{https://doi.org/https://doi.org/10.1002/hyp.7586}{https://doi.org/10.1002/hyp.7586}
+
+\leavevmode\hypertarget{ref-ruprechtetal1991}{}%
+Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A.,
+Stoneman, G.L., 1991. Early hydrological response to intense forest
+thinning in southwestern australia. Journal of Hydrology 127, 261--277.
+doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(91)90118-2}{https://doi.org/10.1016/0022-1694(91)90118-2}
+
+\leavevmode\hypertarget{ref-thornton2007}{}%
+Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007. The
+brigalow catchment study: II*. Clearing brigalow (acacia harpophylla)
+for cropping or pasture increases runoff. Australian Journal of Soil
+Research 45, 496--511.
+doi:\href{https://doi.org/doi:10.1071/SR07064}{doi:10.1071/SR07064}
+
 \leavevmode\hypertarget{ref-wood2006}{}%
@@
 
+\leavevmode\hypertarget{ref-zhao2010}{}%
+Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010. Evaluation of methods
+for estimating the effects of vegetation change and climate variability
+on streamflow. Water Resources Research 46, W03505.
+doi:\href{https://doi.org/10.1029/2009wr007702}{10.1029/2009wr007702}
+
+\leavevmode\hypertarget{ref-zhou2010}{}%
+Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang,
+C., 2010. Forest recovery and river discharge at the regional scale of
+guangdong province, china. Water Resources Research 46.
+doi:\href{https://doi.org/https://doi.org/10.1029/2009WR008829}{https://doi.org/10.1029/2009WR008829}
+
 \end{document}

### 2021-09-23 paper/Forest_and_Water/Forest_and_Water.Rmd
rewrite of introduction and methods

@@
 
-There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013].
-The historic work highlights provides a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (or Australian watersheds) a 100% decrease in stream flow for watersheds with 100% forest cover.
+There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017].
+The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
+The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian watersheds) a 100% decrease in stream flow for watersheds with 100% forest cover.
 However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
 
-There have been several review papers aiming to summarize different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984] and more generally [@jackson2005; @zhang2017].
-These studies are aiming to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent review [@zhang2017] developed an impressive database of watershed studies in relation to changes in streamflow due to changes in forest cover based on a global data set.
-This dataset, which covers over 250 studies are described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+<!-- Include Filoso paper -->
 
-The conclusions of the paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^.
+Several review papers have summarized different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
+These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
+The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover.
+The @zhang2017 dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
+In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+
+The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^.
 While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover.
@@
 
-Encouraged by the work presented by @zhang2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion by presenting further analysis of the data and by adding further watersheds and enhancements to the data base.
+The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
+Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
+For a subset of quantitative data (37) they showed a relationship between catchment size and decline in streamflow.
 
-In particular, the main method in the work by @zhang2017 is using simple linear regression.
-And the main assumption is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds.
-Given the fantastic data set collected, the analysis can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.
+A final summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
+However, this paper aims to explain the variation in the data using the Fuh model, and in particular aims to link the variation in the observed data to variations in the exponent *m* in the model.
+A key observation is that in drier environments, the effects of deforestation are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
 
-In particular, the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds and spatial coordinates and 2) to analyse the possibility of non-linear and partial effects of the different factors and variables in the data base using generalised linear (GLM) and generalised additive models (GAM @wood2006).
-Finally we hope to point to further research that can expand our work and that outlined @zhang2017 to better understand the impact of forest cover change on streamflow.
+Encouraged by the work presented by @zhang2017 and @filoso2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion.
+In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.
+
+In particular, the main method in the work by @zhang2017 is using simple linear regression, and in @filoso2017 the focus is mainly on classification.
+As @zhang2017 points out, the main assumption in their work is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds, but the subset of data in @filoso2017 does not appear to support this.
+And while te work @filoso2017 provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact.
+This is because the work had a strict criterion to select quantitative studies.
+However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.
+
+As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
+Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
 
@@
 
-## The original data set
+## The original data sets
 
@@
 
-While the @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration is not part of the data set.
-We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watreshed data sets in our analysis.
-Some small naming errors and citations for some of the data sets for some of the small watersheds were fixed as we were familiar with the studies.
-But overall the original data set was not changed.
+While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
+We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watershed data sets in our analysis.
 
@@
 
-To enhance the existing data set, this study added additional variables.
-The first variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
-Using this information  annual average reference evapotranspiration (E0) was extracted from **XXXXX** if value of E0 were not available from the original papers.
-For large water watersheds, this value, similar to annual average rainfall, is an approximation of the climate at the location.
+To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from @filoso2017.
+In particular, we focussed on the 37 data points included in the quantitative analysis in @filoso2017.
+
+In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
+Using this information  annual average reference evapotranspiration (E0) was extracted from **XXXXX** if a value of E0 was not available from the original papers.
+For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
 
 The length of the study can be a variable influencing the change in flow [e.g.
-@jackson2005] and therefore, the length of the different studies was extracted from the references provided by @zhang2017.
+@jackson2005] and therefore, the length, starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
 
@@
 
-The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research.
+The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
 
@@
 
+Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
+However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
+@jackson2005) and possibly the climate (as indicated by either Pa and E0 or latitude and longitude @filoso2017; @zhou2015).
+
+However, there could be further confounding factors, which are eluded to by @filoso2017:  
+- the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.  
+- the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
+
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
-In particular we fitted the following initial model to replicate the variables in the analysis from @zhang2017:
+
+The first model applied in this analysis is based on the main hypothesis outlined above, can the change in streamflow be predicted from the change in forest cover, modulated by area, the length of the study and the climate.
 
@@
 \begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
 \end{aligned}$$
 
-However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals.
+However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals, which violates the linear model assumptions.
 As a result we transformed $\Delta \%Q$ and $\Delta \%forest$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is either $\Delta Q$ or $\Delta forest$.
-This means that the model residuals are $\sim N(0,\sigma^2)$ results in the following equation:
+This means that the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:
 
 $$\tag{2}
+\begin{aligned}
+log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
+\end{aligned}$$
+
+A second model included all the variables in the analysis from @zhang2017 in one model:
+
+$$\tag{3}
 \begin{aligned}
@@
 
-In this model, the assumption is that all continuous variables (such as Pa) can have a linear or non-linear relationship with $log10(\Delta Q)$.
+In this model, no direct interactions are assumed, and the assumption is that all continuous variables (such as Pa) can have a linear or non-linear relationship with $log10(\Delta Q)$.
 This means that a smooth function $s()$ is applied to the variable.
 To restrict the smoothness of the fit, the smoothness factor $k$ is restricted to a value of 3 [@wood2006].
-This restriction was applied to smooth variables throughout this paper and we have dropped this from the subsequent equations.
+This restriction was applied to smooth variables throughout this paper and we have dropped this from the notation in subsequent equations.
 
-For the model in equation 2, we only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the analysis.
+For the model in equation 3, we only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the analysis.
 Given that in @zhang2017, dryness ($\frac{E0}{Pa}$) is used to look at variations in the change in flow, we also fitted the following model:
 
-$$\tag{3}
+$$\tag{4}
 \begin{aligned}
@@
 
-$$\tag{4}
+$$\tag{5}
 \begin{aligned}
 log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
 \end{aligned}$$
 
-$$\tag{5}
+$$\tag{6}
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
+& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
+& {forest~type} + {climate~type} + {assessment~type} + \\  
+& {hydrologic~regime} + \varepsilon
 \end{aligned}$$

### 2021-09-23 paper/Forest_and_Water/Forest_and_Water.Rmd
update paper with first analysis

@@
 
+<!----
+The starting point consists of the two tables in the supplementary data from Zhang et al.
+(2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2).
+We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
+--->
+```{r, echo = F, message = F}
+require(tidyverse)
+require(lubridate)
+
+Zhang_small <- read_csv("data/TablesZhangetal2017_small.csv")
+Zhang_large <- read_csv("data/TablesZhangetal2017_large.csv")
+```
+
+<!--- ### change names of columns
+
+the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple --->
+
+```{r, echo = F}
+names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
+names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
+names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
+names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
+names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
+names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
+```
+
+<!-- combine both tables -->
+
+```{r, echo = F}
+Zhang_small$From <- as.numeric(Zhang_small$From)
+Zhang_small$To <- as.numeric(Zhang_small$To)
+Zhang_all <- bind_rows(Zhang_large,Zhang_small)
+```
+
 Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
 However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
-@jackson2005) and possibly the climate (as indicated by either Pa and E0 or latitude and longitude @filoso2017; @zhou2015).
+@jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
 
 However, there could be further confounding factors, which are eluded to by @filoso2017:
+
 - the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.
+
 - the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
@@
 \begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
+\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + Latitude + Longitude + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
+log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + Latitude + Longitude + \varepsilon
 \end{aligned}$$
@@
 
+Results
+===========
+
+<!-- analysis -->
+```{r model2, echo = F}
+require(mgcv)
+linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+
 References {#references .unnumbered}

### 2021-09-23 paper/Forest_and_Water/Forest_and_Water.tex
update paper with first analysis

@@
 literature around the impact of forests on streamflow (Andréassian,
-2004; Brown et al., 2013, 2005; Jackson et al., 2005; Zhang et al.,
-2017). The historic work highlights provides a general consensus that if
-forest areas increase, streamflow decreases and vice-versa. The most
-dramatic result in relation to this, is Figure 5 in Zhang et al. (2011)
-indicating (or Australian watersheds) a 100\% decrease in stream flow
-for watersheds with 100\% forest cover. However, on the other end of the
-spectrum, in a series of French watersheds (Cosandey et al., 2005),
-there was no change in streamflow characteristics in 2 of the three
-watersheds studied in relation to deforestation.
+2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al.,
+2005; Zhang et al., 2017). The historic work highlights a general
+consensus that if forest areas increase, streamflow decreases and
+vice-versa. The most dramatic result in relation to this, is Figure 5 in
+Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
+decrease in stream flow for watersheds with 100\% forest cover. However,
+on the other end of the spectrum, in a series of French watersheds
+(Cosandey et al., 2005), there was no change in streamflow
+characteristics in 2 of the three watersheds studied in relation to
+deforestation.
 
-There have been several review papers aiming to summarize different
-studies across the globe, in relation to paired watershed studies (Bosch
-and Hewlett, 1982; Brown et al., 2005) and more generally (Jackson et
-al., 2005; Zhang et al., 2017). These studies are aiming to generalize
-the individual findings and to identify if there are global trends or
-relationships that can be developed. The most recent review (Zhang et
-al., 2017) developed an impressive database of watershed studies in
-relation to changes in streamflow due to changes in forest cover based
-on a global data set. This dataset, which covers over 250 studies are
-described in terms of the change in streamflow as a result of the change
-in forest cover, where studies related to both forestation (increase in
-forest cover) and deforestation (decrease in forest cover) were
-included.
+Several review papers have summarized different studies across the
+globe, in relation to paired watershed studies (Bosch and Hewlett, 1982;
+Brown et al., 2005), related to reforestation in particular (Filoso et
+al., 2017), and more generally (Jackson et al., 2005; Zhang et al.,
+2017). These studies aim to generalize the individual findings and to
+identify if there are global trends or relationships that can be
+developed. The most recent reviews (Filoso et al., 2017; Zhang et al.,
+2017) developed an impressive global database of watershed studies in
+relation to changes in streamflow due to changes in forest cover. The
+Zhang et al. (2017) dataset, which covers over 250 studies, is described
+in terms of the change in streamflow as a result of the change in forest
+cover, where studies related to both forestation (increase in forest
+cover) and deforestation (decrease in forest cover) were included. In
+contrast, the paper by Filoso et al. (2017) focused primarily on
+reforestation, and covered an equally impressive database of 167 studies
+using a systematic review. In this case the collected data is mostly
+coded as count data and only a subset of 37 studies was analysed for
+actual water yield change.
 
-The conclusions of the paper (Zhang et al., 2017) suggest that there is
-a distinct difference in the change in flow as a result of forestation
-or deforestation between small watersheds, defined as \textless{} 1000
-km\textsuperscript{2} and large watersheds \textgreater{} 1000
-km\textsuperscript{2}. While for small watersheds there was no real
-change in runoff with changes in cover, for large watersheds there was a
-clear trend showing a decrease in runoff with and increase in forest
-cover. Their main conclusion was that the response in annual runoff to
-forest cover was scale dependent and appeared to be more sensitive to
-forest cover change in water limited watersheds relative to energy
-limited watershed (Zhang et al., 2017).
+The conclusions of the first paper (Zhang et al., 2017) suggest that
+there is a distinct difference in the change in flow as a result of
+forestation or deforestation between small watersheds, defined as
+\textless{} 1000 km\textsuperscript{2} and large watersheds
+\textgreater{} 1000 km\textsuperscript{2}. While for small watersheds
+there was no real change in runoff with changes in cover, for large
+watersheds there was a clear trend showing a decrease in runoff with and
+increase in forest cover. Their main conclusion was that the response in
+annual runoff to forest cover was scale dependent and appeared to be
+more sensitive to forest cover change in water limited watersheds
+relative to energy limited watershed (Zhang et al., 2017).
 
-Encouraged by the work presented by Zhang et al. (2017) and the
-fantastic database of studies presented by these authors, we believe we
-can add to the discussion by presenting further analysis of the data and
-by adding further watersheds and enhancements to the data base.
+The second study (Filoso et al., 2017) was a systematic review which
+classified the historical research and highlighted gaps in the spatial
+distribution, the types of studies and the types of analysis. Their main
+conclusion was also that reforestation decreases streamflow, but that
+there were many interacting factors. For a subset of quantitative data
+(37) they showed a relationship between catchment size and decline in
+streamflow.
+
+A final summary paper that includes much of the same data as Zhang et
+al. (2017) and Filoso et al. (2017) is Zhou et al. (2015), which has one
+author in common with Zhang et al. (2017). However, this paper aims to
+explain the variation in the data using the Fuh model, and in particular
+aims to link the variation in the observed data to variations in the
+exponent \emph{m} in the model. A key observation is that in drier
+environments, the effects of deforestation are much greater than in
+wetter environments, which is also suggested by Figure 4 in Zhang et al.
+(2017).
+
+Encouraged by the work presented by Zhang et al. (2017) and Filoso et
+al. (2017) and the fantastic database of studies presented by these
+authors, we believe we can add to the discussion. In this paper, the aim
+is to develop further analysis of the collected data and expanding and
+combining the two data sets to provide further depth.
 
 In particular, the main method in the work by Zhang et al. (2017) is
-using simple linear regression. And the main assumption is that the
-threshold at 1000 km\textsuperscript{2} is a distinct separation between
-``small'' and ``large'' watersheds. Given the fantastic data set
-collected, the analysis can be easily expanded to look at interactions
-between the terms and to test the assumption of a distinct threshold at
-1000 km\textsuperscript{2}.
+using simple linear regression, and in Filoso et al. (2017) the focus is
+mainly on classification. As Zhang et al. (2017) points out, the main
+assumption in their work is that the threshold at 1000
+km\textsuperscript{2} is a distinct separation between ``small'' and
+``large'' watersheds, but the subset of data in Filoso et al. (2017)
+does not appear to support this. And while te work Filoso et al. (2017)
+provides important insights in study types, analysis types and broad
+classification, there is limited quantification of actual impact. This
+is because the work had a strict criterion to select quantitative
+studies. However, given the fantastic data sets collected, the analyses
+can be easily expanded to look at interactions between the terms and to
+test the assumption of a distinct threshold at 1000
+km\textsuperscript{2}.
 
-In particular, the objective of this paper is to 1) enhance the data set
-from Zhang et al. (2017) with further watersheds and spatial coordinates
-and 2) to analyse the possibility of non-linear and partial effects of
-the different factors and variables in the data base using generalised
-linear (GLM) and generalised additive models (GAM Wood (2006)). Finally
-we hope to point to further research that can expand our work and that
-outlined Zhang et al. (2017) to better understand the impact of forest
-cover change on streamflow.
+As a result the objective of this paper is to 1) enhance the data set
+from Zhang et al. (2017) with further watersheds (such as from Filoso et
+al. (2017)) and spatial coordinates and 2) to analyse the possibility of
+non-linear, interactions and partial effects of the different factors
+and variables in the data using generalised linear (GLM) and generalised
+additive models (GAM Wood (2006)). Finally, this work aims to point to
+further research that can expand this area of work, based on the
+collected data, to better understand the impact of forest cover change
+on streamflow.
 
@@
 
-\hypertarget{the-original-data-set}{%
-\subsection{The original data set}\label{the-original-data-set}}
+\hypertarget{the-original-data-sets}{%
+\subsection{The original data sets}\label{the-original-data-sets}}
 
@@
 
-While the Zhang et al. (2017) use the dryness index in their analysis,
-potential or reference evapotranspiration is not part of the data set.
-We combined the tables for small (\textless{} 1000
-km\textsuperscript{2}) and large (\textgreater= 1000
-km\textsuperscript{2}) watreshed data sets in our analysis. Some small
-naming errors and citations for some of the data sets for some of the
-small watersheds were fixed as we were familiar with the studies. But
-overall the original data set was not changed.
+While Zhang et al. (2017) use the dryness index in their analysis,
+potential or reference evapotranspiration was not originally included as
+part of the published data set. We combined the tables for small
+(\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000
+km\textsuperscript{2}) watershed data sets in our analysis.
 
@@
 
-To enhance the existing data set, this study added additional variables.
-The first variables added were the latitude and longitude for the center
-of the watershed as an approximation of its spatial location. Using this
-information annual average reference evapotranspiration (E0) was
-extracted from \textbf{XXXXX} if value of E0 were not available from the
-original papers. For large water watersheds, this value, similar to
-annual average rainfall, is an approximation of the climate at the
-location.
+To enhance the existing data set, this study added additional variables
+and cross-checked the studies with the data set from Filoso et al.
+(2017). In particular, we focussed on the 37 data points included in the
+quantitative analysis in Filoso et al. (2017).
+
+In addition, additional variables added were the latitude and longitude
+for the center of the watershed as an approximation of its spatial
+location. Using this information annual average reference
+evapotranspiration (E0) was extracted from \textbf{XXXXX} if a value of
+E0 was not available from the original papers. For large watersheds,
+this value, similar to annual average rainfall, is only an approximation
+of the climate at the location.
 
 The length of the study can be a variable influencing the change in flow
-(e.g.~Jackson et al., 2005) and therefore, the length of the different
-studies was extracted from the references provided by Zhang et al.
-(2017).
+(e.g.~Jackson et al., 2005) and therefore, the length, starting data and
+completion date of the different studies was extracted from the
+references provided by Zhang et al. (2017).
 
@@
 is not further used in the analysis, but gives context to some of the
-data for future research.
+data for future research and highlights some of the discrepancies that
+we found between the original papers and the data in the tables from
+Zhang et al. (2017).
 
@@
 
+Building on the analyses by Zhang et al. (2017) and Filoso et al.
+(2017), and combining their conclusions, the main hypothesis to test is
+that the change in streamflow is impacted by the change in forest cover.
+However, this change is clearly modulated by the area under
+consideration (affecting the length of the flowpaths Zhou et al.
+(2015)), the length of the study (c.f. Jackson et al. (2005)) and
+possibly the climate (as indicated by either Pa and E0 or latitude and
+longitude Filoso et al. (2017); Zhou et al. (2015)).
+
+However, there could be further confounding factors, which are eluded to
+by Filoso et al. (2017):\\
+- the type of analysis, i.e.~paired catchment studies, modelling, time
+series analysis etc.\\
+- the age of the study, assuming that historical studies might not have
+had the ability to measure at the accuracy that currently is available
+to researchers, or that more careful historical attention to detail in
+field studies might have been lost more recently due to reductions in
+research investment.
+
 To estimate how the change in streamflow is affected by the change in
 forest cover while considering the effects of the other variables, we
-applied generalised additive modelling (GAM) (Wood, 2006). In particular
-we fitted the following initial model to replicate the variables in the
-analysis from Zhang et al. (2017):
+applied generalised additive modelling (GAM) (Wood, 2006).
+
+The first model applied in this analysis is based on the main hypothesis
+outlined above, can the change in streamflow be predicted from the
+change in forest cover, modulated by area, the length of the study and
+the climate.
 
@@
 \begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
 \end{aligned}\]
@@
 (\(\Delta \%Q\)) is problematic, and this results in a skewed
-distribution of the GAM model residuals. As a result we transformed
-\(\Delta \%Q\) and \(\Delta \%forest\) back to fractions (0 - 1) and log
-transformed using \(log10(x + 1)\), where \(x\) is either \(\Delta Q\)
-or \(\Delta forest\). This means that the model residuals are
-\(\sim N(0,\sigma^2)\) results in the following equation:
+distribution of the GAM model residuals, which violates the linear model
+assumptions. As a result we transformed \(\Delta \%Q\) and
+\(\Delta \%forest\) back to fractions (0 - 1) and log transformed using
+\(log10(x + 1)\), where \(x\) is either \(\Delta Q\) or
+\(\Delta forest\). This means that the model residuals approximate
+\(\sim N(0,\sigma^2)\) and this results in the following equation:
 
 \[\tag{2}
+\begin{aligned}
+log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
+\end{aligned}\]
+
+A second model included all the variables in the analysis from Zhang et
+al. (2017) in one model:
+
+\[\tag{3}
 \begin{aligned}
@@
 
-In this model, the assumption is that all continuous variables (such as
-Pa) can have a linear or non-linear relationship with
-\(log10(\Delta Q)\). This means that a smooth function \(s()\) is
-applied to the variable. To restrict the smoothness of the fit, the
-smoothness factor \(k\) is restricted to a value of 3 (Wood, 2006). This
-restriction was applied to smooth variables throughout this paper and we
-have dropped this from the subsequent equations.
+In this model, no direct interactions are assumed, and the assumption is
+that all continuous variables (such as Pa) can have a linear or
+non-linear relationship with \(log10(\Delta Q)\). This means that a
+smooth function \(s()\) is applied to the variable. To restrict the
+smoothness of the fit, the smoothness factor \(k\) is restricted to a
+value of 3 (Wood, 2006). This restriction was applied to smooth
+variables throughout this paper and we have dropped this from the
+notation in subsequent equations.
 
-For the model in equation 2, we only used the data from Zhang et al.
+For the model in equation 3, we only used the data from Zhang et al.
 (2017) to make sure that the additional watersheds added to the data set
@@
 
-\[\tag{3}
+\[\tag{4}
 \begin{aligned}
@@
 
-\[\tag{4}
+\[\tag{5}
 \begin{aligned}
 log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
 \end{aligned}\]
 
-\[\tag{5}
+\[\tag{6}
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
+log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
+& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
+& {forest~type} + {climate~type} + {assessment~type} + \\  
+& {hydrologic~regime} + \varepsilon
 \end{aligned}\]
@@
 doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2004.06.040}{https://doi.org/10.1016/j.jhydrol.2004.06.040}
+
+\leavevmode\hypertarget{ref-filoso2017}{}%
+Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017. Impacts of
+forest restoration on water yield: A systematic review. PLOS ONE 12,
+e0183210.
+doi:\href{https://doi.org/10.1371/journal.pone.0183210}{10.1371/journal.pone.0183210}
 
@@
 
+\leavevmode\hypertarget{ref-zhou2015}{}%
+Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G.,
+Scott, D.F., Zhou, S., Han, L., Su, Y., 2015. Global pattern for the
+effect of climate and land cover on water yield. Nature Communications
+6, 5918.
+doi:\href{https://doi.org/10.1038/ncomms6918}{10.1038/ncomms6918}
+
 \leavevmode\hypertarget{ref-zhou2010}{}%

### 2021-09-27 paper/Forest_and_Water/Forest_and_Water.Rmd
changes WV 27092021

@@
 As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
+
+Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
+However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
+@jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
+
+However, there could be further confounding factors, which are eluded to by @filoso2017:
+
+* the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.
+
+* the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
+
 Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
@@
 
-Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
-@jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
-
-However, there could be further confounding factors, which are eluded to by @filoso2017:
-
-- the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.
-
-- the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
-
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
@@
 ```
+<!-- This shows that Area is not significant -->
+
+<!-- include Lat and Long (later E0) -->
+
+```{r model2a, echo = F}
+linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+
+<!-- include all the other numerical variables -->
+
+```{r model2a, echo = F}
+Zhang_all <- Zhang_all %>%
+  mutate(length = To - From)
+linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+
+<!-- model 2b also include categorical variables -->
+```{r model2b, echo = F}
+
+linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k=3) + s(Pa_mm, k=3) + s(From,k = 3) + s(length, k=3) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+<!-- model 2c include smooth variables -->
+
+```{r model2c, echo = F}
+
+linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k=3) + s(Pa_mm, k=3) + s(From,k = 3) + s(length, k=3) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+
+No evidence of effect of area
+
+<!-- Should area be log transformed? model 2d -->
+
+```{r model2d, echo = F}
+
+linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + s(Pa_mm, k=3) + From + length + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime + Latitude + Longitude, data = Zhang_all)
+summary(linear_model_all_log)
+gam.check(linear_model_all_log)
+```
+```{r}
+Zhang_all %>%
+  ggplot(aes(Assessment_technique,log10(Area_km2))) + geom_boxplot()
+```
+
+```{r}
+Zhang_all %>%
+  ggplot(aes(DeltaF_perc, log10(DeltaQf_perc/100 + 1), colour = Assessment_technique,size = log10(Area_km2))) + geom_point()
+```
+```{r}
+tiff("flow_forest_byArea.tiff", width = 2000, height = 1400, res = 300)
+Zhang_all %>%
+  ggplot(aes(DeltaF_perc/100, (DeltaQf_perc/100 + 1), colour = Assessment_technique, size = log10(Area_km2))) + geom_point() +
+  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
+  xlab("fractional change in forestry") + scale_y_log10() + scale_size_continuous(name = "log10 Area in km2") +
+  scale_colour_discrete(name = "Assessment Technique")
+dev.off()
+```
+
+```{r}
+Zhang_all %>%
+  ggplot(aes(Latitude, log10(DeltaQf_perc/100 + 1), colour = DeltaF_perc,size = log10(Area_km2))) + geom_point()
+```
+
+Check the size distribition of the catchments
+
+```{r}
+Zhang_all %>%
+  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
+  scale_x_log10()
+total <- nrow(Zhang_all)
+length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
+```
 

### 2021-10-07 paper/Forest_and_Water/Forest_and_Water.Rmd
update paper 08102021

@@
 require(lubridate)
+require(mgcv)
 
-Zhang_small <- read_csv("data/TablesZhangetal2017_small.csv")
-Zhang_large <- read_csv("data/TablesZhangetal2017_large.csv")
+Zhang_small <- read_csv("../../data/TablesZhangetal2017_small.csv")
+Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv")
 ```
@@
 However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals, which violates the linear model assumptions.
-As a result we transformed $\Delta \%Q$ and $\Delta \%forest$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is either $\Delta Q$ or $\Delta forest$.
-This means that the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:
+As a result we transformed $\Delta \%Q$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is $\Delta Q$.
+After transformation the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:
 
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + Latitude + Longitude + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \%{forest~cover}
++ Pa + Area + Latitude + 
+\\ &Longitude + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \% forest~cover + s(Pa, k = 3) + s(Area, k = 3) +  \\
+& {forest~type} + {climate~type} + {assessment~type} + \\  
+& {hydrologic~regime} + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + 
+\\  &{climate~type} + {assessment~type} + \\  &{hydrologic~regime} + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(Pa) + s(Area) + s(Latitude) + \\
+& s(Longitude) + s(begin_{year}) + s(length_{study}) +\\
+&{forest~type} + {climate~type} + {assessment~type} +\\
+& {hydrologic~regime} + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
-& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
-& {forest~type} + {climate~type} + {assessment~type} + \\  
-& {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover +
+s(\frac{E0}{Pa}) + s(Area) + \\ 
+& s(Latitude) + s(Longitude) + s(begin_{year}) + \\ 
+& s(length_{study}) + {forest~type} + {climate~type} + \\ 
+& {assessment~type} + {hydrologic~regime} + \varepsilon
 \end{aligned}$$
-
+<!--
 The results were analysed to identify:  
@@
 2. the direction of the categorical or shape of the smooth variables
+-->
 
@@
 <!-- analysis -->
+Create a new data frame in which all the change in forestry is positive, but an additional column is added that indicates whether it was a forest cover increase or decrease.
+
+```{r}
+Zhang_all2 <- Zhang_all %>%
+  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
+                              "decrease", "increase"),
+         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
+                                  -1*DeltaF_perc,
+                                  DeltaF_perc))
+
+```
+
+linear regression just looking at the change in forestry and the sign of the change (deforestation or forestation)
+
+```{r model1, echo = F}
+linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
+                             Forest_Sign, 
+                           data = Zhang_all2)
+summary(linear_model_simple)
+gam.check(linear_model_simple)
+
+```
+Change in forest cover has a significant impact on the change in flow, and the change in flow is different under deforestation compared to reforestation.
+In fact, reforestation has a 32% lower impact than deforestation.
+Overall model explanatory power is just `r summary(linear_model_simple)$r.squared`
+
 ```{r model2, echo = F}
-require(mgcv)
-linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm, data = Zhang_all)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
+
+linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
+summary(linear_model2)
+gam.check(linear_model2)
 ```
+Area is not a significant predictor of the change in flow.
+But decreases in forestry definitely have a different effect from increases in flow.
+
 <!-- This shows that Area is not significant -->
@@
 
-```{r model2a, echo = F}
+```{r model2a1, echo = F}
 Zhang_all <- Zhang_all %>%
@@
 ```{r}
-Zhang_all %>%
-  ggplot(aes(DeltaF_perc, log10(DeltaQf_perc/100 + 1), colour = Assessment_technique,size = log10(Area_km2))) + geom_point()
+Zhang_all2 %>%
+  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), shape = Forest_Sign)) + geom_point()
 ```
 ```{r}
-tiff("flow_forest_byArea.tiff", width = 2000, height = 1400, res = 300)
-Zhang_all %>%
-  ggplot(aes(DeltaF_perc/100, (DeltaQf_perc/100 + 1), colour = Assessment_technique, size = log10(Area_km2))) + geom_point() +
-  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
-  xlab("fractional change in forestry") + scale_y_log10() + scale_size_continuous(name = "log10 Area in km2") +
-  scale_colour_discrete(name = "Assessment Technique")
+tiff("flow_forest_byArea.tiff", width = 2500, height = 1800, res = 300)
+Zhang_all2 %>%
+  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
+             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
+  theme_bw() + ylab("% change in flow") +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5))) +
+  xlab("% change in forestry") + #scale_y_log10() +
+  scale_size_continuous(name = "log10(Area in km2)") +
+  scale_colour_discrete(name = "Assessment Technique") +
+  scale_shape_discrete(name = "Forest cover direction")
 dev.off()
@@
 ```
+```{r}
+Zhang_all %>%
+  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
+```
+```{r}
+#windows()
+Zhang_all %>%
+  ggplot(aes(log10(length), (DeltaQf_perc/100 + 1), colour = DeltaF_perc/100, size = log10(Area_km2))) + geom_point() +
+  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
+  xlab("Length of study") + scale_size_continuous(name = "log10 Area in km2") +
+  scale_colour_continuous(name = "change in forest cover")
+```
 

### 2021-10-07 paper/Forest_and_Water/Forest_and_Water.tex
update paper 08102021

@@
 \bibliographystyle{elsarticle-harv}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable}
+\usepackage{graphicx}
 \ifxetex
@@
 and variables in the data using generalised linear (GLM) and generalised
-additive models (GAM Wood (2006)). Finally, this work aims to point to
-further research that can expand this area of work, based on the
-collected data, to better understand the impact of forest cover change
-on streamflow.
+additive models (GAM Wood (2006)).
+
+Building on the analyses by Zhang et al. (2017) and Filoso et al.
+(2017), and combining their conclusions, the main hypothesis to test is
+that the change in streamflow is impacted by the change in forest cover.
+However, this change is clearly modulated by the area under
+consideration (affecting the length of the flowpaths Zhou et al.
+(2015)), the length of the study (c.f. Jackson et al. (2005)) and
+possibly the climate (as indicated by either E0/Pa or latitude and
+longitude Filoso et al. (2017); Zhou et al. (2015)).
+
+However, there could be further confounding factors, which are eluded to
+by Filoso et al. (2017):
+
+\begin{itemize}
+\item
+  the type of analysis, i.e.~paired catchment studies, modelling, time
+  series analysis etc.
+\item
+  the age of the study, assuming that historical studies might not have
+  had the ability to measure at the accuracy that currently is available
+  to researchers, or that more careful historical attention to detail in
+  field studies might have been lost more recently due to reductions in
+  research investment.
+\end{itemize}
+
+Finally, this work aims to point to further research that can expand
+this area of work, based on the collected data, to better understand the
+impact of forest cover change on streamflow.
 
@@
 
-Building on the analyses by Zhang et al. (2017) and Filoso et al.
-(2017), and combining their conclusions, the main hypothesis to test is
-that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under
-consideration (affecting the length of the flowpaths Zhou et al.
-(2015)), the length of the study (c.f. Jackson et al. (2005)) and
-possibly the climate (as indicated by either Pa and E0 or latitude and
-longitude Filoso et al. (2017); Zhou et al. (2015)).
+\begin{verbatim}
+## Warning: NAs introduced by coercion
 
-However, there could be further confounding factors, which are eluded to
-by Filoso et al. (2017):\\
-- the type of analysis, i.e.~paired catchment studies, modelling, time
-series analysis etc.\\
-- the age of the study, assuming that historical studies might not have
-had the ability to measure at the accuracy that currently is available
-to researchers, or that more careful historical attention to detail in
-field studies might have been lost more recently due to reductions in
-research investment.
+## Warning: NAs introduced by coercion
+\end{verbatim}
 
@@
 \begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
+\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + Latitude + Longitude + \varepsilon
 \end{aligned}\]
@@
 distribution of the GAM model residuals, which violates the linear model
-assumptions. As a result we transformed \(\Delta \%Q\) and
-\(\Delta \%forest\) back to fractions (0 - 1) and log transformed using
-\(log10(x + 1)\), where \(x\) is either \(\Delta Q\) or
-\(\Delta forest\). This means that the model residuals approximate
+assumptions. As a result we transformed \(\Delta \%Q\) back to fractions
+(0 - 1) and log transformed using \(log10(x + 1)\), where \(x\) is
+\(\Delta Q\). After transformation the model residuals approximate
 \(\sim N(0,\sigma^2)\) and this results in the following equation:
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \%{forest~cover}
++ Pa + Area + Latitude + 
+\\ &Longitude + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \% forest~cover + s(Pa, k = 3) + s(Area, k = 3) +  \\
+& {forest~type} + {climate~type} + {assessment~type} + \\  
+& {hydrologic~regime} + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + 
+\\  &{climate~type} + {assessment~type} + \\  &{hydrologic~regime} + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(Pa) + s(Area) + s(Latitude) + \\
+& s(Longitude) + s(begin_{year}) + s(length_{study}) +\\
+&{forest~type} + {climate~type} + {assessment~type} +\\
+& {hydrologic~regime} + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
-& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
-& {forest~type} + {climate~type} + {assessment~type} + \\  
-& {hydrologic~regime} + \varepsilon
+log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover +
+s(\frac{E0}{Pa}) + s(Area) + \\ 
+& s(Latitude) + s(Longitude) + s(begin_{year}) + \\ 
+& s(length_{study}) + {forest~type} + {climate~type} + \\ 
+& {assessment~type} + {hydrologic~regime} + \varepsilon
 \end{aligned}\]
 
-The results were analysed to identify:\\
-1. the significance of the different variables\\
-2. the direction of the categorical or shape of the smooth variables
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\begin{verbatim}
+## Loading required package: mgcv
+\end{verbatim}
+
+\begin{verbatim}
+## Loading required package: nlme
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'nlme'
+\end{verbatim}
+
+\begin{verbatim}
+## The following object is masked from 'package:dplyr':
+## 
+##     collapse
+\end{verbatim}
+
+\begin{verbatim}
+## This is mgcv 1.8-35. For overview type 'help("mgcv-package")'.
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + Area_km2 + Pa_mm
+## 
+## Parametric coefficients:
+##               Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)  4.430e-02  1.759e-02   2.519   0.0123 *  
+## DeltaF_perc -1.345e-03  1.468e-04  -9.163   <2e-16 ***
+## Area_km2    -1.657e-08  3.057e-08  -0.542   0.5883    
+## Pa_mm       -7.989e-06  1.164e-05  -0.686   0.4930    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =   0.21   Deviance explained = 21.8%
+## GCV = 0.018116  Scale est. = 0.017882  n = 310
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  4 / 4
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + Area_km2 + Pa_mm + 
+##     Latitude + Longitude
+## 
+## Parametric coefficients:
+##               Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)  7.296e-02  1.964e-02   3.715 0.000243 ***
+## DeltaF_perc -1.552e-03  1.526e-04 -10.173  < 2e-16 ***
+## Area_km2    -2.963e-08  2.996e-08  -0.989 0.323525    
+## Pa_mm       -1.382e-05  1.205e-05  -1.147 0.252393    
+## Latitude    -1.276e-03  2.826e-04  -4.516 9.07e-06 ***
+## Longitude   -2.121e-05  8.975e-05  -0.236 0.813323    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.271   Deviance explained = 28.3%
+## GCV = 0.017125  Scale est. = 0.016787  n = 304
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2a-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  6 / 6
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + 
+##     Pa_mm + Latitude + Longitude + From + length
+## 
+## Parametric coefficients:
+##                   Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)     -1.027e+00  9.040e-01  -1.136 0.256902    
+## DeltaF_perc     -1.267e-03  1.594e-04  -7.947 5.83e-14 ***
+## log10(Area_km2) -2.003e-02  4.386e-03  -4.567 7.63e-06 ***
+## Pa_mm           -7.871e-06  1.255e-05  -0.627 0.531281    
+## Latitude        -1.084e-03  2.834e-04  -3.827 0.000163 ***
+## Longitude        6.268e-05  9.560e-05   0.656 0.512595    
+## From             5.612e-04  4.583e-04   1.225 0.221809    
+## length          -2.931e-06  7.036e-06  -0.417 0.677294    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.303   Deviance explained = 32.1%
+## GCV = 0.016008  Scale est. = 0.01553   n = 268
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2a1-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  8 / 8
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k = 3) + 
+##     s(Pa_mm, k = 3) + s(From, k = 3) + s(length, k = 3) + Precip_data_type + 
+##     Assessment_technique + Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                               Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -0.0962315  0.0558563  -1.723   0.0862 .  
+## DeltaF_perc                 -0.0008839  0.0001711  -5.165 4.98e-07 ***
+## Precip_data_typeOB          -0.0329926  0.0418383  -0.789   0.4311    
+## Precip_data_typeSG           0.0595846  0.0474089   1.257   0.2100    
+## Assessment_techniqueEA, HM   0.0143199  0.1329271   0.108   0.9143    
+## Assessment_techniqueHM       0.0910165  0.0445991   2.041   0.0423 *  
+## Assessment_techniquePWE      0.2041286  0.0457201   4.465 1.22e-05 ***
+## Assessment_techniquePWE, HM  0.0977846  0.1370941   0.713   0.4764    
+## Assessment_techniqueQPW      0.0850446  0.0731186   1.163   0.2459    
+## Assessment_techniqueSH       0.1066145  0.0473168   2.253   0.0251 *  
+## Forest_typeCF               -0.0029430  0.0268072  -0.110   0.9127    
+## Forest_typeMF               -0.0463517  0.0265852  -1.744   0.0825 .  
+## Hydrological_regimeSD        0.0111796  0.0319040   0.350   0.7263    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##              edf Ref.df     F p-value  
+## s(Area_km2) 1.00  1.000 0.301  0.5839  
+## s(Pa_mm)    1.25  1.437 0.640  0.3574  
+## s(From)     1.00  1.000 3.194  0.0751 .
+## s(length)   1.00  1.000 0.540  0.4631  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.274   Deviance explained = 31.9%
+## GCV = 0.017635  Scale est. = 0.016474  n = 262
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2b-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 8 iterations.
+## The RMS GCV score gradient at convergence was 2.042593e-08 .
+## The Hessian was positive definite.
+## Model rank =  21 / 21 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##               k'  edf k-index p-value    
+## s(Area_km2) 2.00 1.00    0.97    0.29    
+## s(Pa_mm)    2.00 1.25    0.78  <2e-16 ***
+## s(From)     2.00 1.00    0.85    0.01 ** 
+## s(length)   2.00 1.00    0.90    0.02 *  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k = 3) + 
+##     s(Pa_mm, k = 3) + s(From, k = 3) + s(length, k = 3) + Precip_data_type + 
+##     Assessment_technique + Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                               Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -0.0962315  0.0558563  -1.723   0.0862 .  
+## DeltaF_perc                 -0.0008839  0.0001711  -5.165 4.98e-07 ***
+## Precip_data_typeOB          -0.0329926  0.0418383  -0.789   0.4311    
+## Precip_data_typeSG           0.0595846  0.0474089   1.257   0.2100    
+## Assessment_techniqueEA, HM   0.0143199  0.1329271   0.108   0.9143    
+## Assessment_techniqueHM       0.0910165  0.0445991   2.041   0.0423 *  
+## Assessment_techniquePWE      0.2041286  0.0457201   4.465 1.22e-05 ***
+## Assessment_techniquePWE, HM  0.0977846  0.1370941   0.713   0.4764    
+## Assessment_techniqueQPW      0.0850446  0.0731186   1.163   0.2459    
+## Assessment_techniqueSH       0.1066145  0.0473168   2.253   0.0251 *  
+## Forest_typeCF               -0.0029430  0.0268072  -0.110   0.9127    
+## Forest_typeMF               -0.0463517  0.0265852  -1.744   0.0825 .  
+## Hydrological_regimeSD        0.0111796  0.0319040   0.350   0.7263    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##              edf Ref.df     F p-value  
+## s(Area_km2) 1.00  1.000 0.301  0.5839  
+## s(Pa_mm)    1.25  1.437 0.640  0.3574  
+## s(From)     1.00  1.000 3.194  0.0751 .
+## s(length)   1.00  1.000 0.540  0.4631  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.274   Deviance explained = 31.9%
+## GCV = 0.017635  Scale est. = 0.016474  n = 262
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2c-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 8 iterations.
+## The RMS GCV score gradient at convergence was 2.042593e-08 .
+## The Hessian was positive definite.
+## Model rank =  21 / 21 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##               k'  edf k-index p-value    
+## s(Area_km2) 2.00 1.00    0.97   0.330    
+## s(Pa_mm)    2.00 1.25    0.78  <2e-16 ***
+## s(From)     2.00 1.00    0.85   0.015 *  
+## s(length)   2.00 1.00    0.90   0.095 .  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+No evidence of effect of area
+
+\begin{verbatim}
+## Warning in eval(predvars, data, env): NaNs produced
+
+## Warning in eval(predvars, data, env): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + 
+##     s(Pa_mm, k = 3) + From + length + Precip_data_type + Assessment_technique + 
+##     Forest_type + Hydrological_regime + Latitude + Longitude
+## 
+## Parametric coefficients:
+##                               Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -1.2633792  1.1731739  -1.077  0.28260    
+## DeltaF_perc                 -0.0010709  0.0001693  -6.324 1.21e-09 ***
+## log10(Area_km2)             -0.0106973  0.0079411  -1.347  0.17921    
+## From                         0.0006320  0.0005902   1.071  0.28528    
+## length                       0.0002020  0.0007979   0.253  0.80032    
+## Precip_data_typeOB          -0.0496200  0.0405500  -1.224  0.22226    
+## Precip_data_typeSG           0.0379293  0.0474175   0.800  0.42455    
+## Assessment_techniqueEA, HM   0.0190552  0.1279391   0.149  0.88173    
+## Assessment_techniqueHM       0.0665563  0.0447830   1.486  0.13852    
+## Assessment_techniquePWE      0.1335480  0.0529257   2.523  0.01226 *  
+## Assessment_techniquePWE, HM  0.0724841  0.1341116   0.540  0.58936    
+## Assessment_techniqueQPW      0.0585720  0.0713099   0.821  0.41224    
+## Assessment_techniqueSH       0.0649925  0.0482180   1.348  0.17895    
+## Forest_typeCF                0.0129954  0.0267350   0.486  0.62735    
+## Forest_typeMF               -0.0171200  0.0262238  -0.653  0.51448    
+## Hydrological_regimeSD        0.0425574  0.0316727   1.344  0.18031    
+## Latitude                    -0.0010038  0.0003307  -3.035  0.00267 ** 
+## Longitude                    0.0001756  0.0001025   1.713  0.08803 .  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##            edf Ref.df     F p-value
+## s(Pa_mm) 1.004  1.008 1.285   0.256
+## 
+## R-sq.(adj) =  0.328   Deviance explained = 37.5%
+## GCV = 0.016441  Scale est. = 0.015249  n = 262
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2d-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 5 iterations.
+## The RMS GCV score gradient at convergence was 2.648312e-07 .
+## The Hessian was positive definite.
+## Model rank =  20 / 20 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##          k' edf k-index p-value    
+## s(Pa_mm)  2   1     0.8  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Assessment_technique,}\KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_boxplot}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc, }\KeywordTok{log10}\NormalTok{(DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in FUN(X[[i]], ...): NaNs produced
+
+## Warning in FUN(X[[i]], ...): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 2 rows containing missing values (geom_point).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-5-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{tiff}\NormalTok{(}\StringTok{"flow_forest_byArea.tiff"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{2000}\NormalTok{, }\DataTypeTok{height =} \DecValTok{1400}\NormalTok{, }\DataTypeTok{res =} \DecValTok{300}\NormalTok{)}
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc}\OperatorTok{/}\DecValTok{100}\NormalTok{, (DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ Assessment_technique, }\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"log10 (fractional change in flow + 1)"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"fractional change in forestry"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{scale_y_log10}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10 Area in km2"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_colour_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Assessment Technique"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in self$trans$transform(x): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Transformation introduced infinite values in continuous y-axis
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 2 rows containing missing values (geom_point).
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{dev.off}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Latitude, }\KeywordTok{log10}\NormalTok{(DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ DeltaF_perc,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in FUN(X[[i]], ...): NaNs produced
+
+## Warning in FUN(X[[i]], ...): NaNs produced
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 8 rows containing missing values (geom_point).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-7-1.pdf}
+
+Check the size distribition of the catchments
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Area_km2)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_log10}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-8-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{total <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(Zhang_all)}
+\KeywordTok{length}\NormalTok{(Zhang_all}\OperatorTok{$}\NormalTok{Area_km2[Zhang_all}\OperatorTok{$}\NormalTok{Area_km2}\OperatorTok{<}\DecValTok{10}\NormalTok{])}\OperatorTok{/}\NormalTok{total}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 0.6570513
+\end{verbatim}
 

### 2021-10-12 paper/Forest_and_Water/Forest_and_Water.Rmd
adding MODIS ET data large catchments

@@
 
-linear regression just looking at the change in forestry and the sign of the change (deforestation or forestation)
+We start of with a simple linear regression (i.e.
+following @zhang2017) just looking at the change in flow as a result in the percent change forestry and including the direction of the change, increase in forest cover, or decrease in forest cover.
 

### 2021-10-13 paper/Forest_and_Water/Forest_and_Water.Rmd
updates paper 14102021

@@
 In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
-Using this information  annual average reference evapotranspiration (E0) was extracted from **XXXXX** if a value of E0 was not available from the original papers.
+Using this information  annual average potential evapotranspiration (E_{pot}) was extracted from the MODIS16 ET product, if a value of E0 was not available from the original papers.
+This involved downloading the MODIS product for PET at 500 m scale for the approximate catchment centroid using the package MODISTools via the [MODIS/VIIRS subsets](https://modis.ornl.gov/data/modis_webservice.html) for the period 1 January 2000 - 31 December 2020.
+The average annual PET calculated from this series was used as the an approximation of average annual E0.
 For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 The length of the study can be a variable influencing the change in flow [e.g.
-@jackson2005] and therefore, the length, starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
+@jackson2005], as for example, more mature plantations are thought to have smaller impacts on flow.
+Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
 
@@
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
+
+**This methods section is no longer correct and needs rewriting**
 
@@
 <!-- analysis -->
-Create a new data frame in which all the change in forestry is positive, but an additional column is added that indicates whether it was a forest cover increase or decrease.
+#### description of the data
+
+The overall dataset contains `r nrow(Zhang_all)` observations of changes in flow.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
+The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
+Note also the large number of studies with 100\% forest cover reduction.
+
+```{r}
+Zhang_all %>%
+  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm), 
+               names_to = "variable", values_to = "value") %>%
+  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() + facet_wrap(~variable, scales = "free")
+```
+
+The changes in forest cover contain both positive (forestation) and negative values (deforestation).
+In @zhang 2017, these changes were analysed jointly, which assumes that the effect on the change in flow is linear and non-hysteretic.
+However, it is a reasonable hypothesis that this is not the case, and that the impact of an increase in forest cover is different from the same fractional decrease in forest cover.
+To be able to analyse this difference, all the change in forest cover is converted to positive values, but an additional column is added that indicates whether it was a forest cover increase or decrease.
 
@@
 ```
+
+#### The initial relationship between change in forest cover and streamflow
 
@@
 ```
-Change in forest cover has a significant impact on the change in flow, and the change in flow is different under deforestation compared to reforestation.
-In fact, reforestation has a 32% lower impact than deforestation.
-Overall model explanatory power is just `r summary(linear_model_simple)$r.squared`
+While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly indicates the hypothesised relationship between the change in forest cover and the change in flow.
+The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%.
+However the change in flow is different for forest cover decreases compared to forest cover increases.
+In fact, forest cover increases decrease flow by 32\% less than a similar decrease in forest cover causes flow to increase.
+So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.32)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.
+
+It is however clear from lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017).
+
+<!-- At the moment using catchment rainfall, we probably want to straight away use P/E0, and explain that rainfall itself might not be a good predictor.
+Alternatively we can first show that Latitude is a predictor -->
 
@@
 ```
-Area is not a significant predictor of the change in flow.
-But decreases in forestry definitely have a different effect from increases in flow.
 
-<!-- This shows that Area is not significant -->
+Including area and annual precipitation does not really improve the overall explaining power of the model, in fact, annual precipitation appears to be only a very small confounding factor, representing only a -0.01/% partial effect in the change in streamflow, holding all other factors constant.
+The catchment area does not appear to have an effect at all in contrast to earlier reported studies [@zhang2017; @filoso2017].
+The main effects remain the change in forest cover and whether this is an increase or decrease
 
+#### The effect of location on the globe
 <!-- include Lat and Long (later E0) -->
@@
 ```{r model2a, echo = F}
-linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all)
+linear_model_all_log <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
 summary(linear_model_all_log)

### 2021-10-13 paper/Forest_and_Water/Forest_and_Water.tex
updates paper 14102021

@@
 for the center of the watershed as an approximation of its spatial
-location. Using this information annual average reference
-evapotranspiration (E0) was extracted from \textbf{XXXXX} if a value of
-E0 was not available from the original papers. For large watersheds,
-this value, similar to annual average rainfall, is only an approximation
-of the climate at the location.
+location. Using this information annual average potential
+evapotranspiration (E\_\{pot\}) was extracted from the MODIS16 ET
+product, if a value of E0 was not available from the original papers.
+This involved downloading the MODIS product for PET at 500 m scale for
+the approximate catchment centroid using the package MODISTools via the
+\href{https://modis.ornl.gov/data/modis_webservice.html}{MODIS/VIIRS
+subsets} for the period 1 January 2000 - 31 December 2020. The average
+annual PET calculated from this series was used as the an approximation
+of average annual E0. For large watersheds, this value, similar to
+annual average rainfall, is only an approximation of the climate at the
+location.
 
 The length of the study can be a variable influencing the change in flow
-(e.g.~Jackson et al., 2005) and therefore, the length, starting data and
+(e.g.~Jackson et al., 2005), as for example, more mature plantations are
+thought to have smaller impacts on flow. Therefore, the length of the
+study calculate as the difference between the starting data and
 completion date of the different studies was extracted from the
@@
 
+\textbf{This methods section is no longer correct and needs rewriting}
+
 The first model applied in this analysis is based on the main hypothesis
@@
 
-\begin{verbatim}
-## Loading required package: mgcv
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: nlme
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'nlme'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:dplyr':
-## 
-##     collapse
-\end{verbatim}
-
-\begin{verbatim}
-## This is mgcv 1.8-35. For overview type 'help("mgcv-package")'.
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in eval(predvars, data, env): NaNs produced
-
-## Warning in eval(predvars, data, env): NaNs produced
-\end{verbatim}
+\hypertarget{description-of-the-data}{%
+\paragraph{description of the data}\label{description-of-the-data}}
+
+The overall dataset contains 312 observations of changes in flow. The
+overall distribution of changes in flow is highly skewed as is the
+distribution of changes in forest cover and Area. The values of changes
+in flow greater than 100\% and smaller than -100\% clearly create long
+tails on the change in flow distribution. Note also the large number of
+studies with 100\% forest cover reduction.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{pivot_longer}\NormalTok{(}\KeywordTok{c}\NormalTok{(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm), }
+               \DataTypeTok{names_to =} \StringTok{"variable"}\NormalTok{, }\DataTypeTok{values_to =} \StringTok{"value"}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(value)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill =} \StringTok{"steelblue"}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{facet_wrap}\NormalTok{(}\OperatorTok{~}\NormalTok{variable, }\DataTypeTok{scales =} \StringTok{"free"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}
+
+The changes in forest cover contain both positive (forestation) and
+negative values (deforestation). In ({\textbf{???}}) 2017, these changes
+were analysed jointly, which assumes that the effect on the change in
+flow is linear and non-hysteretic. However, it is a reasonable
+hypothesis that this is not the case, and that the impact of an increase
+in forest cover is different from the same fractional decrease in forest
+cover. To be able to analyse this difference, all the change in forest
+cover is converted to positive values, but an additional column is added
+that indicates whether it was a forest cover increase or decrease.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 <-}\StringTok{ }\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{Forest_Sign =} \KeywordTok{ifelse}\NormalTok{(DeltaF_perc }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{,}
+                              \StringTok{"decrease"}\NormalTok{, }\StringTok{"increase"}\NormalTok{),}
+         \DataTypeTok{DeltaF_perc_pos =} \KeywordTok{ifelse}\NormalTok{(DeltaF_perc }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{,}
+                                  \DecValTok{-1}\OperatorTok{*}\NormalTok{DeltaF_perc,}
+\NormalTok{                                  DeltaF_perc))}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{the-initial-relationship-between-change-in-forest-cover-and-streamflow}{%
+\paragraph{The initial relationship between change in forest cover and
+streamflow}\label{the-initial-relationship-between-change-in-forest-cover-and-streamflow}}
+
+We start of with a simple linear regression (i.e.~following Zhang et al.
+(2017)) just looking at the change in flow as a result in the percent
+change forestry and including the direction of the change, increase in
+forest cover, or decrease in forest cover.
 
@@
 ## Formula:
-## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + Area_km2 + Pa_mm
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign
 ## 
@@
 ##                     Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)  4.430e-02  1.759e-02   2.519   0.0123 *  
-## DeltaF_perc -1.345e-03  1.468e-04  -9.163   <2e-16 ***
-## Area_km2    -1.657e-08  3.057e-08  -0.542   0.5883    
-## Pa_mm       -7.989e-06  1.164e-05  -0.686   0.4930    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =   0.21   Deviance explained = 21.8%
-## GCV = 0.018116  Scale est. = 0.017882  n = 310
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}
+## (Intercept)           8.8466     6.5008   1.361    0.175    
+## DeltaF_perc_pos       0.5364     0.1017   5.274 2.52e-07 ***
+## Forest_Signincrease -32.4760     6.9367  -4.682 4.26e-06 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =   0.16   Deviance explained = 16.6%
+## GCV = 3278.3  Scale est. = 3246.8    n = 312
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model1-1.pdf}
 
@@
 ## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  4 / 4
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in eval(predvars, data, env): NaNs produced
-
-## Warning in eval(predvars, data, env): NaNs produced
-\end{verbatim}
+## Model required no smoothing parameter selectionModel rank =  3 / 3
+\end{verbatim}
+
+While the overall variance explained in this model is not high at 0.16,
+it clearly indicates the hypothesised relationship between the change in
+forest cover and the change in flow. The model suggests that for every
+1\% change in forest cover, on the average, the flow changes 0.5\%.
+However the change in flow is different for forest cover decreases
+compared to forest cover increases. In fact, forest cover increases
+decrease flow by 32\% less than a similar decrease in forest cover
+causes flow to increase. So roughly speaking, a 1\% forest cover
+increase on the average decreases flow by \((1 - 0.32)*0.5\%\), while a
+the percentage forest cover decrease will increase flow by 0.5\%.
+
+It is however clear from lack of explaining power, that there could be
+confounding factors, as alluded to in the methods. The obvious ones
+being catchment dryness and area (following Zhang et al. (2017)).
 
@@
 ## Formula:
-## log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + Area_km2 + Pa_mm + 
-##     Latitude + Longitude
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm
 ## 
@@
 ##                       Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)  7.296e-02  1.964e-02   3.715 0.000243 ***
-## DeltaF_perc -1.552e-03  1.526e-04 -10.173  < 2e-16 ***
-## Area_km2    -2.963e-08  2.996e-08  -0.989 0.323525    
-## Pa_mm       -1.382e-05  1.205e-05  -1.147 0.252393    
-## Latitude    -1.276e-03  2.826e-04  -4.516 9.07e-06 ***
-## Longitude   -2.121e-05  8.975e-05  -0.236 0.813323    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =  0.271   Deviance explained = 28.3%
-## GCV = 0.017125  Scale est. = 0.016787  n = 304
+## (Intercept)          2.323e+01  9.189e+00   2.528   0.0120 *  
+## DeltaF_perc_pos      5.424e-01  1.030e-01   5.265 2.64e-07 ***
+## Forest_Signincrease -3.324e+01  6.955e+00  -4.779 2.73e-06 ***
+## Area_km2            -1.867e-06  1.322e-05  -0.141   0.8878    
+## Pa_mm               -1.067e-02  4.944e-03  -2.158   0.0317 *  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.168   Deviance explained = 17.9%
+## GCV = 3269.3  Scale est. = 3217      n = 312
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  5 / 5
+\end{verbatim}
+
+Including area and annual precipitation does not really improve the
+overall explaining power of the model, in fact, annual precipitation
+appears to be only a very small confounding factor, representing only a
+-0.01/\% partial effect in the change in streamflow, holding all other
+factors constant. The catchment area does not appear to have an effect
+at all in contrast to earlier reported studies (Filoso et al., 2017;
+Zhang et al., 2017). The main effects remain the change in forest cover
+and whether this is an increase or decrease
+
+\hypertarget{the-effect-of-location-on-the-globe}{%
+\paragraph{The effect of location on the
+globe}\label{the-effect-of-location-on-the-globe}}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc + Forest_Sign + Area_km2 + Pa_mm + 
+##     Latitude + Longitude
+## 
+## Parametric coefficients:
+##                       Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)          3.474e+01  1.055e+01   3.295  0.00110 ** 
+## DeltaF_perc         -4.951e-01  1.064e-01  -4.652 4.95e-06 ***
+## Forest_Signincrease -3.222e+00  1.160e+01  -0.278  0.78140    
+## Area_km2            -9.488e-06  1.339e-05  -0.709  0.47903    
+## Pa_mm               -1.007e-02  5.309e-03  -1.897  0.05883 .  
+## Latitude            -3.806e-01  1.239e-01  -3.071  0.00233 ** 
+## Longitude            1.074e-03  3.955e-02   0.027  0.97834    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.172   Deviance explained = 18.8%
+## GCV = 3336.1  Scale est. = 3259.8    n = 306
 \end{verbatim}
@@
 ## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  6 / 6
+## Model required no smoothing parameter selectionModel rank =  7 / 7
 \end{verbatim}
@@
 ##                   Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)     -1.027e+00  9.040e-01  -1.136 0.256902    
-## DeltaF_perc     -1.267e-03  1.594e-04  -7.947 5.83e-14 ***
-## log10(Area_km2) -2.003e-02  4.386e-03  -4.567 7.63e-06 ***
-## Pa_mm           -7.871e-06  1.255e-05  -0.627 0.531281    
-## Latitude        -1.084e-03  2.834e-04  -3.827 0.000163 ***
-## Longitude        6.268e-05  9.560e-05   0.656 0.512595    
-## From             5.612e-04  4.583e-04   1.225 0.221809    
-## length          -2.931e-06  7.036e-06  -0.417 0.677294    
+## (Intercept)     -6.962e-01  1.110e+00  -0.627 0.531002    
+## DeltaF_perc     -1.253e-03  1.612e-04  -7.772  1.8e-13 ***
+## log10(Area_km2) -1.886e-02  5.186e-03  -3.637 0.000332 ***
+## Pa_mm           -7.362e-06  1.257e-05  -0.586 0.558457    
+## Latitude        -1.070e-03  2.823e-04  -3.791 0.000186 ***
+## Longitude        6.832e-05  9.635e-05   0.709 0.478879    
+## From             3.951e-04  5.604e-04   0.705 0.481473    
+## length          -3.424e-04  7.026e-04  -0.487 0.626447    
 ## ---
@@
 ## R-sq.(adj) =  0.303   Deviance explained = 32.1%
-## GCV = 0.016008  Scale est. = 0.01553   n = 268
+## GCV = 0.016004  Scale est. = 0.015526  n = 268
 \end{verbatim}
@@
 ##               k'  edf k-index p-value    
-## s(Area_km2) 2.00 1.00    0.97    0.29    
+## s(Area_km2) 2.00 1.00    0.97   0.265    
 ## s(Pa_mm)    2.00 1.25    0.78  <2e-16 ***
-## s(From)     2.00 1.00    0.85    0.01 ** 
-## s(length)   2.00 1.00    0.90    0.02 *  
+## s(From)     2.00 1.00    0.85   0.005 ** 
+## s(length)   2.00 1.00    0.90   0.060 .  
 ## ---
@@
 ##               k'  edf k-index p-value    
-## s(Area_km2) 2.00 1.00    0.97   0.330    
+## s(Area_km2) 2.00 1.00    0.97   0.280    
 ## s(Pa_mm)    2.00 1.25    0.78  <2e-16 ***
 ## s(From)     2.00 1.00    0.85   0.015 *  
-## s(length)   2.00 1.00    0.90   0.095 .  
+## s(length)   2.00 1.00    0.90   0.040 *  
 ## ---
@@
 ##          k' edf k-index p-value   
-## s(Pa_mm)  2   1     0.8  <2e-16 ***
+## s(Pa_mm)  2   1     0.8   0.005 **
 ## ---
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-6-1.pdf}
 
@@
 \begin{Highlighting}[]
-\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc, }\KeywordTok{log10}\NormalTok{(DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }\DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-## Warning in FUN(X[[i]], ...): NaNs produced
-
-## Warning in FUN(X[[i]], ...): NaNs produced
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 2 rows containing missing values (geom_point).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-7-1.pdf}
 
@@
 \begin{Highlighting}[]
-\KeywordTok{tiff}\NormalTok{(}\StringTok{"flow_forest_byArea.tiff"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{2000}\NormalTok{, }\DataTypeTok{height =} \DecValTok{1400}\NormalTok{, }\DataTypeTok{res =} \DecValTok{300}\NormalTok{)}
-\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc}\OperatorTok{/}\DecValTok{100}\NormalTok{, (DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ Assessment_technique, }\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"log10 (fractional change in flow + 1)"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"fractional change in forestry"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{scale_y_log10}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10 Area in km2"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_colour_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Assessment Technique"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in self$trans$transform(x): NaNs produced
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Transformation introduced infinite values in continuous y-axis
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 2 rows containing missing values (geom_point).
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+\KeywordTok{tiff}\NormalTok{(}\StringTok{"flow_forest_byArea.tiff"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{2500}\NormalTok{, }\DataTypeTok{height =} \DecValTok{1800}\NormalTok{, }\DataTypeTok{res =} \DecValTok{300}\NormalTok{)}
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }
+             \DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.5}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"% change in flow"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{axis.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\DecValTok{2}\NormalTok{)),}
+        \DataTypeTok{axis.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\FloatTok{1.5}\NormalTok{))) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"% change in forestry"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\CommentTok{#scale_y_log10() +}
+\StringTok{  }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10(Area in km2)"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_colour_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Assessment Technique"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_shape_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Forest cover direction"}\NormalTok{)}
 \KeywordTok{dev.off}\NormalTok{()}
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-9-1.pdf}
 
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-8-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-10-1.pdf}
 
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(length)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 42 rows containing non-finite values (stat_bin).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-11-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#windows()}
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\KeywordTok{log10}\NormalTok{(length), (DeltaQf_perc}\OperatorTok{/}\DecValTok{100} \OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{), }\DataTypeTok{colour =}\NormalTok{ DeltaF_perc}\OperatorTok{/}\DecValTok{100}\NormalTok{, }\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"log10 (fractional change in flow + 1)"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Length of study"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10 Area in km2"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_colour_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"change in forest cover"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 42 rows containing missing values (geom_point).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-12-1.pdf}
+
 \hypertarget{references}{%

### 2021-10-21 paper/Forest_and_Water/Forest_and_Water.Rmd
small update manuscript

@@
 require(mgcv)
+suppressWarnings(require(knitr))
+suppressWarnings(require(bookdown))
+suppressWarnings(require(pander))
 
@@
 
-```{r, echo = F}
+```{r, echo = F, warning = F}
 Zhang_small$From <- as.numeric(Zhang_small$From)
@@
 <!-- analysis -->
-#### description of the data
+## description of the data
 
@@
 
-```{r}
+```{r, echo = F}
 Zhang_all %>%
@@
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang 2017, these changes were analysed jointly, which assumes that the effect on the change in flow is linear and non-hysteretic.
+In @zhang2017, these changes were analysed jointly, which assumes that the effect on the change in flow is linear and non-hysteretic.
 However, it is a reasonable hypothesis that this is not the case, and that the impact of an increase in forest cover is different from the same fractional decrease in forest cover.
@@
 
-```{r}
+```{r, echo = F}
 Zhang_all2 <- Zhang_all %>%
@@
 
-#### The initial relationship between change in forest cover and streamflow
+## The initial relationship between change in forest cover and streamflow
 
 We start of with a simple linear regression (i.e.
-following @zhang2017) just looking at the change in flow as a result in the percent change forestry and including the direction of the change, increase in forest cover, or decrease in forest cover.
+following @zhang2017) to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover.
 
@@
 ```
-While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly indicates the hypothesised relationship between the change in forest cover and the change in flow.
+While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesised relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%.
@@
 The catchment area does not appear to have an effect at all in contrast to earlier reported studies [@zhang2017; @filoso2017].
-The main effects remain the change in forest cover and whether this is an increase or decrease
+The main effects remain the change in forest cover and whether this is an increase or decrease.
 
-#### The effect of location on the globe
+## The effect of location on the globe
 <!-- include Lat and Long (later E0) -->
 
-```{r model2a, echo = F}
-linear_model_all_log <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
+```{r global_map, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
+
+include_graphics("draft_map.png")
+```
+
+As indicated, a second hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
+As the global map (\@ref(fig:global_map)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
+Of interest is whether the spatial clustering also indicates a difference in reponse to forest cover change.
+
+```{r model3, echo = F}
+model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
+summary(model3_includeLatLong)
+gam.check(model3_includeLatLong)
 ```
@@
 
-```{r model2a1, echo = F}
-Zhang_all <- Zhang_all %>%
+```{r model4, echo = F}
+Zhang_all2 <- Zhang_all2 %>%
   mutate(length = To - From)
-linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
+model4_all_num <- gam(DeltaQf_perc ~ DeltaF_perc+ Forest_Sign + Area_km2 + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all2)
+summary(model4_all_num)
+gam.check(model4_all_num)
 ```
@@
 <!-- model 2b also include categorical variables -->
-```{r model2b, echo = F}
+```{r model5, echo = F}
 
-linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k=3) + s(Pa_mm, k=3) + s(From,k = 3) + s(length, k=3) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
+model5_all <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + s(Area_km2) + s(Pa_mm) + s(From) + s(length) +  Assessment_technique + s(Latitude) + Longitude,  data = Zhang_all2)
+summary(model5_all)
+gam.check(model5_all)
+plot(model5_all)
 ```
@@
 
-linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + s(Area_km2, k=3) + s(Pa_mm, k=3) + s(From,k = 3) + s(length, k=3) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
+model6_all <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + te(Longitude,Latitude, k =5) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all2)
+summary(model6_all)
+gam.check(model6_all)
+plot(model6_all)
 ```
@@
 
-<!-- Should area be log transformed? model 2d -->
-
-```{r model2d, echo = F}
-
-linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + s(Pa_mm, k=3) + From + length + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime + Latitude + Longitude, data = Zhang_all)
-summary(linear_model_all_log)
-gam.check(linear_model_all_log)
-```
 ```{r}
@@
 ```{r}
-Zhang_all %>%
-  ggplot(aes(Latitude, log10(DeltaQf_perc/100 + 1), colour = DeltaF_perc,size = log10(Area_km2))) + geom_point()
+Zhang_all2 %>%
+  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc )) + geom_point(alpha = 0.5)
 ```

### 2021-11-14 paper/Forest_and_Water/Forest_and_Water.Rmd
updates on paper 14112021

@@
   This is the abstract.
-
   It consists of two paragraphs.
-
 journal: "Journal of Hydrology"
@@
 output: rticles::elsevier_article
+editor_options: 
+  chunk_output_type: console
 ---
@@
 In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
-Using this information  annual average potential evapotranspiration (E_{pot}) was extracted from the MODIS16 ET product, if a value of E0 was not available from the original papers.
-This involved downloading the MODIS product for PET at 500 m scale for the approximate catchment centroid using the package MODISTools via the [MODIS/VIIRS subsets](https://modis.ornl.gov/data/modis_webservice.html) for the period 1 January 2000 - 31 December 2020.
-The average annual PET calculated from this series was used as the an approximation of average annual E0.
+Using this information  reference evapotranspiration (E_{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET0) Climate Databasev2 [@trabucco2018], if a value of E0 was not available from the original papers.
 For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
+
+Similar to @zhang2017, the "dryness index" was calculated as:
+$$\tag{1}
+\begin{aligned}
+D = \frac{E0}{Pa}
+\end{aligned}$$
 
@@
 Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv")
+
+```
+
+<!-- read in the ET data -->
+
+```{r, echo = F, warning = F, message=F}
+PET_large <- read_csv("../../data/largeET.csv")
+PET_small <- read_csv("../../data/smallET.csv")
+
+# drop columns and make sure the titles are the same
+PET_large <- PET_large %>%
+  rename(Latitude = Latitiude,
+         E0 = rvalue_1,
+         `Watershed #` = Watershed)
+
+PET_small <- PET_small %>%
+  select(site_name, lat, long, rvalue_1) %>%
+  rename(Latitude = lat,
+         Longitude = long,
+         `Watershed #` = site_name,
+         E0 = rvalue_1)
+
+# 
+PET_all <- bind_rows(PET_large, PET_small)
+# There are 7 missing E0 values
+# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
+# [1]  38 278 269 243 236 299 250
+```
+
+```{r, echo =F, eval = F}
+PET_all %>%
+  ggplot(aes(`Watershed #`,E0)) + geom_bar(stat="identity", fill= "steelblue")
+```
+
+```{r, echo =F, eval = F}
+PET_all %>%
+  ggplot(aes(Longitude,Latitude, colour = E0)) + geom_point() + coord_equal()
+
 ```
@@
 Zhang_all <- bind_rows(Zhang_large,Zhang_small)
+
+```
+
+<!--now combine with E0 values -->
+```{r, echo = F}
+Zhang_all <- left_join(Zhang_all %>% select(-c(E0, Longitude, Latitude)),PET_all, by = "Watershed #")
+# calculate dryness index
+Zhang_all <- Zhang_all %>%
+  mutate(Dryness = E0/Pa_mm)
 ```
@@
 
-**This methods section is no longer correct and needs rewriting**
-
-The first model applied in this analysis is based on the main hypothesis outlined above, can the change in streamflow be predicted from the change in forest cover, modulated by area, the length of the study and the climate.
-
-$$\tag{1}
-\begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + Latitude + Longitude + \varepsilon
-\end{aligned}$$
-
-However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals, which violates the linear model assumptions.
-As a result we transformed $\Delta \%Q$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is $\Delta Q$.
-After transformation the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:
+The general model tested is
 
@@
 \begin{aligned}
-log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \%{forest~cover}
-+ Pa + Area + Latitude + 
-\\ &Longitude + \varepsilon
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}$$
 
-A second model included all the variables in the analysis from @zhang2017 in one model:
+In this model, no direct interactions are assumed.
+The changes in forest cover contain both positive (forestation) and negative values (deforestation).
+In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic.
+However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
+Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease.
+A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$.
+This means that a smooth function $s()$ is applied to the $Z_i$ variables.
 
-$$\tag{3}
-\begin{aligned}
-log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \% forest~cover + s(Pa, k = 3) + s(Area, k = 3) +  \\
-& {forest~type} + {climate~type} + {assessment~type} + \\  
-& {hydrologic~regime} + \varepsilon
-\end{aligned}$$
+For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the results.
+Subsequently the analysis was repeated and the additionally identified watersheds were added.
 
-In this model, no direct interactions are assumed, and the assumption is that all continuous variables (such as Pa) can have a linear or non-linear relationship with $log10(\Delta Q)$.
-This means that a smooth function $s()$ is applied to the variable.
-To restrict the smoothness of the fit, the smoothness factor $k$ is restricted to a value of 3 [@wood2006].
-This restriction was applied to smooth variables throughout this paper and we have dropped this from the notation in subsequent equations.
-
-For the model in equation 3, we only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the analysis.
-Given that in @zhang2017, dryness ($\frac{E0}{Pa}$) is used to look at variations in the change in flow, we also fitted the following model:
-
-$$\tag{4}
-\begin{aligned}
-log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + 
-\\  &{climate~type} + {assessment~type} + \\  &{hydrologic~regime} + \varepsilon
-\end{aligned}$$
-
-Subsequently, using the full data set, including the additional watersheds and the additional variables the following two models were fitted:
-
-$$\tag{5}
-\begin{aligned}
-log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(Pa) + s(Area) + s(Latitude) + \\
-& s(Longitude) + s(begin_{year}) + s(length_{study}) +\\
-&{forest~type} + {climate~type} + {assessment~type} +\\
-& {hydrologic~regime} + \varepsilon
-\end{aligned}$$
-
-$$\tag{6}
-\begin{aligned}
-log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover +
-s(\frac{E0}{Pa}) + s(Area) + \\ 
-& s(Latitude) + s(Longitude) + s(begin_{year}) + \\ 
-& s(length_{study}) + {forest~type} + {climate~type} + \\ 
-& {assessment~type} + {hydrologic~regime} + \varepsilon
-\end{aligned}$$
-<!--
-The results were analysed to identify:  
+More generally the results were analysed to identify:  
 1. the significance of the different variables  
 2. the direction of the categorical or shape of the smooth variables
--->
 
@@
 Zhang_all %>%
-  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm), 
+  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
                names_to = "variable", values_to = "value") %>%
@@
 ```
-
-The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were analysed jointly, which assumes that the effect on the change in flow is linear and non-hysteretic.
-However, it is a reasonable hypothesis that this is not the case, and that the impact of an increase in forest cover is different from the same fractional decrease in forest cover.
-To be able to analyse this difference, all the change in forest cover is converted to positive values, but an additional column is added that indicates whether it was a forest cover increase or decrease.
 
@@
 
+# quick plot to show differences
+Zhang_all2 %>%
+  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
+  geom_point() +  
+  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
+  theme_bw() + scale_colour_manual(name = "direction of forest change", values = c("increase" = "blue", "decrease" = "red")) +
+  xlab("Change in forest cover %") + ylab("Change in flow %")
+
 ```
+
+This immediately shows that for the data related to forest decreases, there is almost always a positive flow change.
+In other words, flow almost always increased.
+However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
+However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
 
@@
 
-We start of with a simple linear regression (i.e.
-following @zhang2017) to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover.
+Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover.
 
@@
 ```
-While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesised relationship between the change in forest cover and the change in flow.
+
+```{r, echo =F}
+pander(summary(linear_model_simple$p.table))
+```
+
+While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%.
@@
 
-It is however clear from lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017).
+It is however clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017).
 
@@
 
+```{r, echo = F}
+pander(summary(linear_model2)$p.table)
+```
+
 Including area and annual precipitation does not really improve the overall explaining power of the model, in fact, annual precipitation appears to be only a very small confounding factor, representing only a -0.01/% partial effect in the change in streamflow, holding all other factors constant.
-The catchment area does not appear to have an effect at all in contrast to earlier reported studies [@zhang2017; @filoso2017].
-The main effects remain the change in forest cover and whether this is an increase or decrease.
+In contrast to earlier reported studies [@zhang2017; @filoso2017], watershed area has no effect on the change in stream flow.
+This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds The main effects remain the change in forest cover and whether this is an increase or decrease.
 
@@
 
-include_graphics("draft_map.png")
+include_graphics("../../data/FAOET0data.png")
 ```
@@
 As the global map (\@ref(fig:global_map)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
-Of interest is whether the spatial clustering also indicates a difference in reponse to forest cover change.
+Of interest is whether the spatial clustering also indicates a difference in response to forest cover change.
 
 ```{r model3, echo = F}
-model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
+model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
 summary(model3_includeLatLong)
@@
 
-<!-- include all the other numerical variables -->
+```{r, echo =F}
+pander(summary(model3_includeLatLong)$p.table)
+```
+
+This linear model shows that there is a significant gradient in the Latitude and with annual average rainfall, with watersheds closer to the equator having lower changes in the runoff compare to watersheds further away from the equator.
+This suggests an influence of radiation, which will be tested next.
+In addition, the model suggests an influence of the annual average rainfall, with wetter watersheds having slightly lower changes in runoff.
+The total explaining power of the model is still low at `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further confounding factors currently not included in the model.
+
+There is no relationship with Longitude, suggesting that the different continents do not show a trend in the East-West direction.
+
+## Impact of the dryness index
+
+The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017.
+Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variables.
+
+<!-- include Dryness index -->
 
 ```{r model4, echo = F}
-Zhang_all2 <- Zhang_all2 %>%
-  mutate(length = To - From)
-model4_all_num <- gam(DeltaQf_perc ~ DeltaF_perc+ Forest_Sign + Area_km2 + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all2)
-summary(model4_all_num)
-gam.check(model4_all_num)
+#Zhang_all2 <- Zhang_all2 %>%
+#  mutate(length = To - From)
+model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 +  
+                        Latitude + Longitude + Dryness, data = Zhang_all2)
+summary(model4_dryness)
+gam.check(model4_dryness)
 ```
+
+```{r, echo =F}
+pander(summary(model4_dryness)$p.table)
+```
+
+The results from this model confirm that dryness is a significant confounding factor of the change in streamflow as function of the change in forest cover change.
+In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is ~14\% greater.
+However, more interesting, Latitude remains a significant predictor with each degree in latitude causing an -0.31\% change in runoff.
+This indicates that Dryness (i.e.
+an increase in radiation) alone does not explain the trend in the Latitude and some other unknown confounding factor is captured by Latitude.
 
@@
 ```{r model5, echo = F}
+Zhang_all2 <- Zhang_all2 %>%
+  mutate(length = To - From)
 
-model5_all <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + s(Area_km2) + s(Pa_mm) + s(From) + s(length) +  Assessment_technique + s(Latitude) + Longitude,  data = Zhang_all2)
+model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + s(Dryness) + length +  Assessment_technique + Latitude ,  data = Zhang_all2 %>% filter(Dryness < 4))
 summary(model5_all)
@@
 
-model6_all <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + te(Longitude,Latitude, k =5) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all2)
+model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + te(Dryness,Latitude, k =5) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all2)
 summary(model6_all)
@@
 plot(model6_all)
+```
+
+```{r}
+Zhang_all2 %>%
+  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Dryness)) + geom_point()
+```
+
+```{r}
+Zhang_all2 %>%
+  ggplot(aes(Assessment_technique,Dryness)) + geom_boxplot()
+```
+
+```{r}
+Zhang_all2 %>%
+  filter(Dryness > 4)
+
 ```
@@
 Zhang_all2 %>%
-  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc )) + geom_point(alpha = 0.5)
+  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
 ```
 
-Check the size distribition of the catchments
+<!--Check the size distribution of the catchments -->
 
@@
 ```{r}
-Zhang_all %>%
+Zhang_all2 %>%
   ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
 ```
-```{r}
-#windows()
-Zhang_all %>%
-  ggplot(aes(log10(length), (DeltaQf_perc/100 + 1), colour = DeltaF_perc/100, size = log10(Area_km2))) + geom_point() +
-  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
-  xlab("Length of study") + scale_size_continuous(name = "log10 Area in km2") +
-  scale_colour_continuous(name = "change in forest cover")
-```
+Essentially, the analysis shows at the moment that their is no evidence that the size of the catchment influences the
 

### 2021-11-14 paper/Forest_and_Water/Forest_and_Water.tex
updates on paper 14112021

--- 2021-09-23 paper/Forest_and_Water/Forest_and_Water.tex
@@
   \begin{abstract}
-  This is the abstract.
-
-  It consists of two paragraphs.
+  This is the abstract. It consists of two paragraphs.
   \end{abstract}
@@
 for the center of the watershed as an approximation of its spatial
-location. Using this information annual average reference
-evapotranspiration (E0) was extracted from \textbf{XXXXX} if a value of
-E0 was not available from the original papers. For large watersheds,
-this value, similar to annual average rainfall, is only an approximation
-of the climate at the location.
+location. Using this information reference evapotranspiration (E\_\{0\})
+was extracted from the Global Aridity Index and Potential
+Evapo-Transpiration (ET0) Climate Databasev2 (Trabucco and Zomer, 2018),
+if a value of E0 was not available from the original papers. For large
+watersheds, this value, similar to annual average rainfall, is only an
+approximation of the climate at the location.
 
 The length of the study can be a variable influencing the change in flow
-(e.g.~Jackson et al., 2005) and therefore, the length, starting data and
+(e.g.~Jackson et al., 2005), as for example, more mature plantations are
+thought to have smaller impacts on flow. Therefore, the length of the
+study calculate as the difference between the starting data and
 completion date of the different studies was extracted from the
@@
 
+Similar to Zhang et al. (2017), the ``dryness index'' was calculated as:
+\[\tag{1}
+\begin{aligned}
+D = \frac{E0}{Pa}
+\end{aligned}\]
+
 \hypertarget{statistical-modelling}{%
@@
 
-Building on the analyses by Zhang et al. (2017) and Filoso et al.
-(2017), and combining their conclusions, the main hypothesis to test is
-that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under
-consideration (affecting the length of the flowpaths Zhou et al.
-(2015)), the length of the study (c.f. Jackson et al. (2005)) and
-possibly the climate (as indicated by either Pa and E0 or latitude and
-longitude Filoso et al. (2017); Zhou et al. (2015)).
+\begin{verbatim}
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   Watershed = col_double(),
+##   Latitiude = col_double(),
+##   Longitude = col_double(),
+##   rvalue_1 = col_double()
+## )
+\end{verbatim}
 
-However, there could be further confounding factors, which are eluded to
-by Filoso et al. (2017):\\
-- the type of analysis, i.e.~paired catchment studies, modelling, time
-series analysis etc.\\
-- the age of the study, assuming that historical studies might not have
-had the ability to measure at the accuracy that currently is available
-to researchers, or that more careful historical attention to detail in
-field studies might have been lost more recently due to reductions in
-research investment.
+\begin{verbatim}
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   site_name = col_double(),
+##   lat = col_double(),
+##   long = col_double(),
+##   `Area(km2)` = col_double(),
+##   `Forest typ` = col_character(),
+##   C = col_character(),
+##   rvalue_1 = col_double()
+## )
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 7 rows containing missing values (position_stack).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-3-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}
 
@@
 
-The first model applied in this analysis is based on the main hypothesis
-outlined above, can the change in streamflow be predicted from the
-change in forest cover, modulated by area, the length of the study and
-the climate.
-
-\[\tag{1}
-\begin{aligned}
-\Delta \%Q \sim ~&\Delta \%forest + Pa + Area + {climate~type} + \varepsilon
-\end{aligned}\]
-
-However, the overall skewed distribution of the predictant
-(\(\Delta \%Q\)) is problematic, and this results in a skewed
-distribution of the GAM model residuals, which violates the linear model
-assumptions. As a result we transformed \(\Delta \%Q\) and
-\(\Delta \%forest\) back to fractions (0 - 1) and log transformed using
-\(log10(x + 1)\), where \(x\) is either \(\Delta Q\) or
-\(\Delta forest\). This means that the model residuals approximate
-\(\sim N(0,\sigma^2)\) and this results in the following equation:
+The general model tested is
 
@@
 \begin{aligned}
-log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + {climate~type} + \varepsilon
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}\]
 
-A second model included all the variables in the analysis from Zhang et
-al. (2017) in one model:
+In this model, no direct interactions are assumed. The changes in forest
+cover contain both positive (forestation) and negative values
+(deforestation). In Zhang et al. (2017), these changes were jointly
+analysed, assuming the effect on the change in flow was linear and
+non-hysteretic. However, the impact of an increase in forest cover can
+be different from the same fractional decrease in forest cover.
+Therefore all the change in forest cover data is converted to positive
+values, and an additional column (\emph{sign\_\{forest cover\}}) is
+added that indicates whether it was a forest cover increase or decrease.
+A further assumption in the model is that all continuous variables
+\(Z_i\) (such as annual precipitation (Pa)) can have a linear or
+non-linear relationship with \(\Delta Q \%\). This means that a smooth
+function \(s()\) is applied to the \(Z_i\) variables.
 
-\[\tag{3}
-\begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
-\end{aligned}\]
+For the model in equation 2, we initially only used the data from Zhang
+et al. (2017) to make sure that the additional watersheds added to the
+data set did not influence the results. Subsequently the analysis was
+repeated and the additionally identified watersheds were added.
 
-In this model, no direct interactions are assumed, and the assumption is
-that all continuous variables (such as Pa) can have a linear or
-non-linear relationship with \(log10(\Delta Q)\). This means that a
-smooth function \(s()\) is applied to the variable. To restrict the
-smoothness of the fit, the smoothness factor \(k\) is restricted to a
-value of 3 (Wood, 2006). This restriction was applied to smooth
-variables throughout this paper and we have dropped this from the
-notation in subsequent equations.
-
-For the model in equation 3, we only used the data from Zhang et al.
-(2017) to make sure that the additional watersheds added to the data set
-did not influence the analysis. Given that in Zhang et al. (2017),
-dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
-in flow, we also fitted the following model:
-
-\[\tag{4}
-\begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
-\end{aligned}\]
-
-Subsequently, using the full data set, including the additional
-watersheds and the additional variables the following two models were
-fitted:
-
-\[\tag{5}
-\begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
-& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
-\end{aligned}\]
-
-\[\tag{6}
-\begin{aligned}
-log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
-& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
-& {forest~type} + {climate~type} + {assessment~type} + \\  
-& {hydrologic~regime} + \varepsilon
-\end{aligned}\]
-
-The results were analysed to identify:\\
+More generally the results were analysed to identify:\\
 1. the significance of the different variables\\
 2. the direction of the categorical or shape of the smooth variables
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{description-of-the-data}{%
+\subsection{description of the data}\label{description-of-the-data}}
+
+The overall dataset contains 312 observations of changes in flow. The
+overall distribution of changes in flow is highly skewed as is the
+distribution of changes in forest cover and Area. The values of changes
+in flow greater than 100\% and smaller than -100\% clearly create long
+tails on the change in flow distribution. Note also the large number of
+studies with 100\% forest cover reduction.
+
+\begin{verbatim}
+## Warning: Removed 7 rows containing non-finite values (stat_bin).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-8-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-9-1.pdf}
+
+This immediately shows that for the data related to forest decreases,
+there is almost always a positive flow change. In other words, flow
+almost always increased. However, for increases in forest cover, this is
+not the case, and flow can both increase and decrease. However in both
+cases the variability in the reported change in flow increases with the
+increase in forest cover change.
+
+\hypertarget{the-initial-relationship-between-change-in-forest-cover-and-streamflow}{%
+\subsection{The initial relationship between change in forest cover and
+streamflow}\label{the-initial-relationship-between-change-in-forest-cover-and-streamflow}}
+
+Following Zhang et al. (2017), the first step is to use a linear
+regression to investigate the percent change in flow as a result in the
+percent change forestry and modulated by the direction of the change,
+either an increase in forest cover, or decrease in forest cover.
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign
+## 
+## Parametric coefficients:
+##                     Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)           8.8466     6.5008   1.361    0.175    
+## DeltaF_perc_pos       0.5364     0.1017   5.274 2.52e-07 ***
+## Forest_Signincrease -32.4760     6.9367  -4.682 4.26e-06 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =   0.16   Deviance explained = 16.6%
+## GCV = 3278.3  Scale est. = 3246.8    n = 312
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model1-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  3 / 3
+\end{verbatim}
+
+\begin{longtable}[]{@{}ccc@{}}
+\toprule
+\begin{minipage}[b]{0.11\columnwidth}\centering
+Length\strut
+\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
+Class\strut
+\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
+Mode\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.11\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+NULL\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+NULL\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+While the overall variance explained in this model is not high at 0.16,
+it clearly supports the hypothesized relationship between the change in
+forest cover and the change in flow. The model suggests that for every
+1\% change in forest cover, on the average, the flow changes 0.5\%.
+However the change in flow is different for forest cover decreases
+compared to forest cover increases. In fact, forest cover increases
+decrease flow by 32\% less than a similar decrease in forest cover
+causes flow to increase. So roughly speaking, a 1\% forest cover
+increase on the average decreases flow by \((1 - 0.32)*0.5\%\), while a
+the percentage forest cover decrease will increase flow by 0.5\%.
+
+It is however clear from the lack of explaining power, that there could
+be confounding factors, as alluded to in the methods. The obvious ones
+being catchment dryness and area (following Zhang et al. (2017)).
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm
+## 
+## Parametric coefficients:
+##                       Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)          2.323e+01  9.189e+00   2.528   0.0120 *  
+## DeltaF_perc_pos      5.424e-01  1.030e-01   5.265 2.64e-07 ***
+## Forest_Signincrease -3.324e+01  6.955e+00  -4.779 2.73e-06 ***
+## Area_km2            -1.867e-06  1.322e-05  -0.141   0.8878    
+## Pa_mm               -1.067e-02  4.944e-03  -2.158   0.0317 *  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.168   Deviance explained = 17.9%
+## GCV = 3269.3  Scale est. = 3217      n = 312
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  5 / 5
+\end{verbatim}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+23.23\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+9.189\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.528\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.01197\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.5424\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.103\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.265\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2.64e-07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-33.24\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+6.955\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-4.779\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2.733e-06\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Area\_km2}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-1.867e-06\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+1.322e-05\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.1412\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.8878\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Pa\_mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.01067\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.004944\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.158\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.03169\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+Including area and annual precipitation does not really improve the
+overall explaining power of the model, in fact, annual precipitation
+appears to be only a very small confounding factor, representing only a
+-0.01/\% partial effect in the change in streamflow, holding all other
+factors constant. In contrast to earlier reported studies (Filoso et
+al., 2017; Zhang et al., 2017), watershed area has no effect on the
+change in stream flow. This supports our approach (in contrast to Zhang
+et al. (2017)) to consider watershed area as a continuous variable and
+making no separation between larger and smaller watersheds The main
+effects remain the change in forest cover and whether this is an
+increase or decrease.
+
+\hypertarget{the-effect-of-location-on-the-globe}{%
+\subsection{The effect of location on the
+globe}\label{the-effect-of-location-on-the-globe}}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\end{figure}
+
+As indicated, a second hypothesis relates to whether there is a strong
+spatial global gradient as captured by latitude and longitude. As the
+global map (@ref(fig:global\_map)) shows, the distribution of case study
+catchments covers multiple continents and shows some distinct clustering
+in parts of the world. Of interest is whether the spatial clustering
+also indicates a difference in response to forest cover change.
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + 
+##     Latitude + Longitude
+## 
+## Parametric coefficients:
+##                       Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)          3.504e+01  1.026e+01   3.415 0.000724 ***
+## DeltaF_perc_pos      5.147e-01  1.027e-01   5.014 9.07e-07 ***
+## Forest_Signincrease -3.793e+01  7.230e+00  -5.247 2.90e-07 ***
+## Area_km2            -6.282e-06  1.327e-05  -0.474 0.636137    
+## Pa_mm               -1.304e-02  5.055e-03  -2.580 0.010339 *  
+## Latitude            -3.119e-01  1.255e-01  -2.484 0.013520 *  
+## Longitude           -1.058e-02  3.965e-02  -0.267 0.789747    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.183   Deviance explained = 19.9%
+## GCV = 3230.5  Scale est. = 3158      n = 312
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model3-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  7 / 7
+\end{verbatim}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+35.04\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+10.26\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.415\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.0007243\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.5147\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1027\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.014\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+9.071e-07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-37.93\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+7.23\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.247\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2.898e-07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Area\_km2}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-6.282e-06\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+1.327e-05\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4736\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.6361\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Pa\_mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.01304\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.005055\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.58\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.01034\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.3119\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1255\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.484\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.01352\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Longitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.01058\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.03965\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.2669\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.7897\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+This linear model shows that there is a significant gradient in the
+Latitude and with annual average rainfall, with watersheds closer to the
+equator having lower changes in the runoff compare to watersheds further
+away from the equator. This suggests an influence of radiation, which
+will be tested next. In addition, the model suggests an influence of the
+annual average rainfall, with wetter watersheds having slightly lower
+changes in runoff. The total explaining power of the model is still low
+at 0.18 suggesting further confounding factors currently not included in
+the model.
+
+There is no relationship with Longitude, suggesting that the different
+continents do not show a trend in the East-West direction.
+
+\hypertarget{impact-of-the-dryness-index}{%
+\subsection{Impact of the dryness
+index}\label{impact-of-the-dryness-index}}
+
+The previous analysis suggests an influence of radiation on
+evapotranspiration, and most likely related to the dryness index, as
+also indicated in Zhang et al. (2017). Increased evapotranspiration
+could lead to drier catchments, unless balanced by rainfall (such as
+possibly in the tropics). This model introduces the dryness index as a
+linear variable and drops the annual average precipitation as a
+variables.
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Latitude + 
+##     Longitude + Dryness
+## 
+## Parametric coefficients:
+##                       Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)          6.031e-01  8.582e+00   0.070 0.944022    
+## DeltaF_perc_pos      5.006e-01  1.036e-01   4.831 2.18e-06 ***
+## Forest_Signincrease -3.898e+01  7.253e+00  -5.375 1.55e-07 ***
+## Area_km2            -7.227e-06  1.317e-05  -0.549 0.583604    
+## Latitude            -3.121e-01  1.249e-01  -2.500 0.012974 *  
+## Longitude           -6.775e-03  3.958e-02  -0.171 0.864199    
+## Dryness              1.370e+01  3.628e+00   3.776 0.000193 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.204   Deviance explained = 21.9%
+## GCV = 3199.2  Scale est. = 3125.7    n = 305
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model4-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  7 / 7
+\end{verbatim}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.6031\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+8.582\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.07027\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.944\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.5006\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1036\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+4.831\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2.178e-06\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-38.98\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+7.253\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.375\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+1.549e-07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Area\_km2}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-7.227e-06\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+1.317e-05\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.5487\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.5836\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.3121\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1249\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.5\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.01297\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Longitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-0.006775\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.03958\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.1712\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.8642\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{Dryness}\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+13.7\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+3.628\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.776\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.0001927\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+The results from this model confirm that dryness is a significant
+confounding factor of the change in streamflow as function of the change
+in forest cover change. In fact if the dryness index doubles
+(remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa,
+which is very dry), the change in runoff is \textasciitilde14\% greater.
+However, more interesting, Latitude remains a significant predictor with
+each degree in latitude causing an -0.31\% change in runoff. This
+indicates that Dryness (i.e.~an increase in radiation) alone does not
+explain the trend in the Latitude and some other unknown confounding
+factor is captured by Latitude.
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + 
+##     s(Dryness) + length + Assessment_technique + Latitude
+## 
+## Parametric coefficients:
+##                              Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                  19.80962   20.90807   0.947  0.34433    
+## DeltaF_perc_pos               0.34576    0.10727   3.223  0.00144 ** 
+## Forest_Signincrease         -33.34803    7.26563  -4.590 7.06e-06 ***
+## length                       -0.16503    0.26850  -0.615  0.53935    
+## Assessment_techniqueEA, HM    6.97930   47.05195   0.148  0.88220    
+## Assessment_techniqueHM        1.39799   16.63165   0.084  0.93308    
+## Assessment_techniquePWE       4.12984   19.29598   0.214  0.83070    
+## Assessment_techniquePWE, HM  -5.44499   49.38888  -0.110  0.91230    
+## Assessment_techniqueQPW      -9.50504   26.72374  -0.356  0.72239    
+## Assessment_techniqueSH        2.52746   18.97865   0.133  0.89416    
+## Latitude                     -0.30515    0.09629  -3.169  0.00172 ** 
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##             edf Ref.df     F p-value
+## s(Area_km2)   1      1 0.513   0.474
+## s(Dryness)    1      1 0.002   0.964
+## 
+## R-sq.(adj) =  0.185   Deviance explained = 22.2%
+## GCV = 2181.8  Scale est. = 2072.8    n = 260
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model5-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 15 iterations.
+## The RMS GCV score gradient at convergence was 0.0001863276 .
+## The Hessian was positive definite.
+## Model rank =  29 / 29 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##             k' edf k-index p-value  
+## s(Area_km2)  9   1    1.06   0.825  
+## s(Dryness)   9   1    0.86   0.015 *
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model5-2.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model5-3.pdf}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + te(Dryness, Latitude, 
+##     k = 5) + Precip_data_type + Assessment_technique + Forest_type + 
+##     Hydrological_regime
+## 
+## Parametric coefficients:
+##                             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -15.0968    26.7857  -0.564 0.573480    
+## DeltaF_perc_pos               0.4516     0.1187   3.805 0.000175 ***
+## Forest_Signincrease         -27.5255     9.0389  -3.045 0.002552 ** 
+## Precip_data_typeOB           -5.2815    18.9655  -0.278 0.780856    
+## Precip_data_typeSG           11.3166    21.3449   0.530 0.596417    
+## Assessment_techniqueEA, HM   45.5460    57.7615   0.789 0.431077    
+## Assessment_techniqueHM       21.7074    19.8327   1.095 0.274689    
+## Assessment_techniquePWE      29.7823    20.4186   1.459 0.145827    
+## Assessment_techniquePWE, HM  51.4437    62.8574   0.818 0.413831    
+## Assessment_techniqueQPW      19.1277    31.5696   0.606 0.545090    
+## Assessment_techniqueQPW, EA  29.9396    35.2621   0.849 0.396591    
+## Assessment_techniqueSH       28.3985    22.2674   1.275 0.203273    
+## Forest_typeCF                 5.0733    11.4701   0.442 0.658619    
+## Forest_typeMF               -13.2933    10.8133  -1.229 0.219997    
+## Hydrological_regimeSD        13.3327    13.2135   1.009 0.313855    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                       edf Ref.df     F  p-value    
+## te(Dryness,Latitude) 10.7  12.97 3.432 6.26e-05 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.249   Deviance explained = 31.1%
+## GCV = 3281.3  Scale est. = 2999.2    n = 299
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2c-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 7 iterations.
+## The RMS GCV score gradient at convergence was 0.002397792 .
+## The Hessian was positive definite.
+## Model rank =  39 / 39 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##                        k'  edf k-index p-value
+## te(Dryness,Latitude) 24.0 10.7       1    0.52
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model2c-2.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Dryness)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-14-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Assessment_technique,Dryness)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_boxplot}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 7 rows containing non-finite values (stat_boxplot).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-15-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(Dryness }\OperatorTok{>}\StringTok{ }\DecValTok{4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 4 x 25
+##   `Watershed #` `Watershed name`     Area_km2 Pa_mm Forest_type Hydrological_re~
+##           <dbl> <chr>                   <dbl> <dbl> <chr>       <chr>           
+## 1            76 Beaver  Creek,  AZ ~     1.47   426 CF          SD              
+## 2            90 Cantua                 119      442 MF          RD              
+## 3           295 White Spar, Ariz., ~     1      549 BF          RD              
+## 4           225 Natural DRDages, Ar~     0.05   452 BF          RD              
+## # ... with 19 more variables: DeltaF_perc <dbl>, DeltaQf_perc <dbl>,
+## #   Precip_data_type <chr>, Assessment_technique <chr>, Source of info <chr>,
+## #   Length of study (years) <dbl>, Notes <chr>, From <dbl>, To <dbl>,
+## #   Location Details AL=aprox location given by author, PL=point location given by author, GE= coordinates guessed from google maps <chr>,
+## #   notes <chr>,
+## #   Location Details AL=aprox location given by author, PL=point location given by author, GM= coordinates guessed from google maps <chr>,
+## #   Latitude <dbl>, Longitude <dbl>, E0 <dbl>, Dryness <dbl>,
+## #   Forest_Sign <chr>, DeltaF_perc_pos <dbl>, length <dbl>
+\end{verbatim}
+
+No evidence of effect of area
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Assessment_technique,}\KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_boxplot}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-17-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }\DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-18-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{tiff}\NormalTok{(}\StringTok{"flow_forest_byArea.tiff"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{2500}\NormalTok{, }\DataTypeTok{height =} \DecValTok{1800}\NormalTok{, }\DataTypeTok{res =} \DecValTok{300}\NormalTok{)}
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }
+             \DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.5}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"% change in flow"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{axis.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\DecValTok{2}\NormalTok{)),}
+        \DataTypeTok{axis.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\FloatTok{1.5}\NormalTok{))) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"% change in forestry"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\CommentTok{#scale_y_log10() +}
+\StringTok{  }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10(Area in km2)"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_colour_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Assessment Technique"}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_shape_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Forest cover direction"}\NormalTok{)}
+\KeywordTok{dev.off}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Longitude, Latitude, }\DataTypeTok{colour =}\NormalTok{ DeltaF_perc, }\DataTypeTok{size =}\NormalTok{ DeltaQf_perc}\OperatorTok{/}\DecValTok{100}\NormalTok{ )) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.5}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
+
+Check the size distribition of the catchments
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Area_km2)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_log10}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-21-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{total <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(Zhang_all)}
+\KeywordTok{length}\NormalTok{(Zhang_all}\OperatorTok{$}\NormalTok{Area_km2[Zhang_all}\OperatorTok{$}\NormalTok{Area_km2}\OperatorTok{<}\DecValTok{10}\NormalTok{])}\OperatorTok{/}\NormalTok{total}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 0.6570513
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all2 }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(length)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 42 rows containing non-finite values (stat_bin).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-22-1.pdf}
+Essentially, the analysis shows at the moment that their is no evidence
+that the size of the catchment influences the
 
@@
 
+\leavevmode\hypertarget{ref-trabucco2018}{}%
+Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential
+evapo-transpiration (et0) climate database v2. CGIAR consortium for
+spatial information(CGIAR-csi).
+
 \leavevmode\hypertarget{ref-wood2006}{}%

### 2021-11-23 paper/Forest_and_Water/Forest_and_Water.Rmd
further paper updates and data correction

@@
     affiliation: INIA
+    footnote: 2
   - name: Jimena Alonso
@@
 In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
-Table 6 in @zhang2011)
+Table 6 in @zhang2011).
+We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017.
+This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
 
@@
 
+<!--remove the Amazon from the analysis -->
+```{r}
+Zhang_all <- Zhang_all %>%
+  filter(`Watershed #` > 1)
+```
+
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
@@
 
-In this model, no direct interactions are assumed.
+Here $X_i$ are factorial variables, while $Z_i$ are continous variables.
+The model assumes no direct interactions and all variables are additive.
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
@@
 Note also the large number of studies with 100\% forest cover reduction.
+Smaller watersheds dominate the database with 42% of the data from watersheds < 1 km^2^ and 65% of the data for watersheds < 10 km^2^.
 
-```{r, echo = F}
+```{r, echo = F, warning = F, message=F, eval =F}
+png("DataExploration.png", width = 2500, height = 1800, res = 300)
 Zhang_all %>%
@@
                names_to = "variable", values_to = "value") %>%
-  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() + facet_wrap(~variable, scales = "free")
+  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
+  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
+    c("Area_km2" = "Area (km2)",
+      "DeltaF_perc" = "Delta Forest (%)",
+      "DeltaQf_perc" = "Delta Flow (%)",
+      "E0" = "Annual Potential E (mm)",
+      "Pa_mm" = "Annual Rainfall (mm)"))) +   
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)),
+        axis.text.x = element_text(angle = 45, hjust = 1))
+dev.off()
+```
+
+```{r data_graphs, echo =F, fig.cap = "Overview of the distributions of some of the variables in the data set", out.width= "90%"}
+include_graphics("DataExploration.png")
 ```
@@
                                   DeltaF_perc))
+```
 
-# quick plot to show differences
+```{r, echo = F, warning = F, message=F, eval =F}
+# plot to show the difference
+png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
 Zhang_all2 %>%
@@
   xlab("Change in forest cover %") + ylab("Change in flow %")
+dev.off()
 
@@
 
-This immediately shows that for the data related to forest decreases, there is almost always a positive flow change.
+```{r increase_decrease, echo =F, fig.cap = "Changes in flow as a function of increases and decreases in forest cover", out.width= "90%"}
+include_graphics("Increase_decrease.png")
+```
+
+This shows that for the data related to forest decreases, there is almost always a positive flow change.
 In other words, flow almost always increased.
@@
                            data = Zhang_all2)
-summary(linear_model_simple)
-gam.check(linear_model_simple)
+#summary(linear_model_simple)
+#gam.check(linear_model_simple)
 
@@
 ```{r, echo =F}
-pander(summary(linear_model_simple$p.table))
+pander(round(summary(linear_model_simple)$p.table,2))
 ```
@@
 linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
-summary(linear_model2)
-gam.check(linear_model2)
+#summary(linear_model2)
+#gam.check(linear_model2)
 ```
@@
 ```{r, echo = F}
-pander(summary(linear_model2)$p.table)
+pander(round(summary(linear_model2)$p.table,2))
 ```
@@
 model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
-summary(model3_includeLatLong)
-gam.check(model3_includeLatLong)
+#summary(model3_includeLatLong)
+#gam.check(model3_includeLatLong)
 ```
@@
 ```{r, echo =F}
-pander(summary(model3_includeLatLong)$p.table)
+pander(round(summary(model3_includeLatLong)$p.table,2))
 ```
@@
 Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
-This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variables.
+This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable.
 
@@
                         Latitude + Longitude + Dryness, data = Zhang_all2)
-summary(model4_dryness)
-gam.check(model4_dryness)
+#summary(model4_dryness)
+#gam.check(model4_dryness)
 ```
@@
 ```{r, echo =F}
-pander(summary(model4_dryness)$p.table)
+pander(round(summary(model4_dryness)$p.table,2))
 ```
@@
 
-<!-- model 2b also include categorical variables -->
+However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data.
+These catchment are listed in Table XX:
+
+```{r, echo = F}
+pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`))
+```
+
 ```{r model5, echo = F}
@@
 model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + s(Dryness) + length +  Assessment_technique + Latitude ,  data = Zhang_all2 %>% filter(Dryness < 4))
-summary(model5_all)
-gam.check(model5_all)
-plot(model5_all)
+#summary(model5_all)
+#gam.check(model5_all)
+#plot(model5_all)
 ```
@@
 
-```{r model2c, echo = F}
+```{r model6_all, echo = F}
 
-model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + te(Dryness,Latitude, k =5) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all2)
+model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
+                    Precip_data_type + Assessment_technique + Forest_type +
+                    Hydrological_regime, data = Zhang_all2)
 summary(model6_all)
-gam.check(model6_all)
-plot(model6_all)
+#gam.check(model6_all)
+#plot(model6_all)
 ```
 
-```{r}
+```{r model6_reduc}
+model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
+                    Assessment_technique +
+                    Hydrological_regime, data = Zhang_all2)
+summary(model6_reduc)
+
+```
+
+```{r, echo = F}
 Zhang_all2 %>%
@@
 
-```{r}
+```{r, echo = F}
 Zhang_all2 %>%
-  ggplot(aes(Assessment_technique,Dryness)) + geom_boxplot()
+  ggplot(aes(Assessment_technique,Area_km2)) + geom_boxplot()
 ```
 
-```{r}
+```{r, eval =F}
 Zhang_all2 %>%
-  filter(Dryness > 4)
+  filter(Area_km2 > 3e+06)
 
@@
 
-```{r}
+```{r, eval =F}
 Zhang_all %>%
@@
 
-```{r}
+```{r, echo = F}
 Zhang_all2 %>%
@@
 ```
-```{r}
-tiff("flow_forest_byArea.tiff", width = 2500, height = 1800, res = 300)
+```{r, echo = F}
+png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
 Zhang_all2 %>%
@@
 
-```{r}
+```{r, eval =F}
 Zhang_all2 %>%
@@
 ```
-```{r}
+```{r, eval =F}
 Zhang_all2 %>%

### 2021-11-23 paper/Forest_and_Water/Forest_and_Water.tex
further paper updates and data correction

@@
    \ead{willem.vervoort@sydney.edu.au} 
-    \author[INIA]{Eliana Nervi}
+    \author[INIA]{Eliana Nervi\corref{2}}
    \ead{eliananervi@gmail.com} 
@@
 not provide the quantification of the change in streamflow (i.e.~Table 6
-in Zhang et al. (2011))
+in Zhang et al. (2011)). We also removed one data point from the
+analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al.
+(2017). This is because the cited reference (Roche, 1981) only relates
+to 1 and 1.5 ha paired watershed studies in French Guyana, and in which
+the actual change in forest cover is not recorded.
 
@@
 
-\begin{verbatim}
-## 
-## -- Column specification --------------------------------------------------------
-## cols(
-##   Watershed = col_double(),
-##   Latitiude = col_double(),
-##   Longitude = col_double(),
-##   rvalue_1 = col_double()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## -- Column specification --------------------------------------------------------
-## cols(
-##   site_name = col_double(),
-##   lat = col_double(),
-##   long = col_double(),
-##   `Area(km2)` = col_double(),
-##   `Forest typ` = col_character(),
-##   C = col_character(),
-##   rvalue_1 = col_double()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 7 rows containing missing values (position_stack).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-3-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-4-1.pdf}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all <-}\StringTok{ }\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(}\StringTok{`}\DataTypeTok{Watershed #}\StringTok{`} \OperatorTok{>}\StringTok{ }\DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-In this model, no direct interactions are assumed. The changes in forest
-cover contain both positive (forestation) and negative values
-(deforestation). In Zhang et al. (2017), these changes were jointly
-analysed, assuming the effect on the change in flow was linear and
-non-hysteretic. However, the impact of an increase in forest cover can
-be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data is converted to positive
-values, and an additional column (\emph{sign\_\{forest cover\}}) is
-added that indicates whether it was a forest cover increase or decrease.
-A further assumption in the model is that all continuous variables
-\(Z_i\) (such as annual precipitation (Pa)) can have a linear or
-non-linear relationship with \(\Delta Q \%\). This means that a smooth
-function \(s()\) is applied to the \(Z_i\) variables.
+Here \(X_i\) are factorial variables, while \(Z_i\) are continous
+variables. The model assumes no direct interactions and all variables
+are additive. The changes in forest cover contain both positive
+(forestation) and negative values (deforestation). In Zhang et al.
+(2017), these changes were jointly analysed, assuming the effect on the
+change in flow was linear and non-hysteretic. However, the impact of an
+increase in forest cover can be different from the same fractional
+decrease in forest cover. Therefore all the change in forest cover data
+is converted to positive values, and an additional column
+(\emph{sign\_\{forest cover\}}) is added that indicates whether it was a
+forest cover increase or decrease. A further assumption in the model is
+that all continuous variables \(Z_i\) (such as annual precipitation
+(Pa)) can have a linear or non-linear relationship with \(\Delta Q \%\).
+This means that a smooth function \(s()\) is applied to the \(Z_i\)
+variables.
 
@@
 
-The overall dataset contains 312 observations of changes in flow. The
+The overall dataset contains 311 observations of changes in flow. The
 overall distribution of changes in flow is highly skewed as is the
@@
 tails on the change in flow distribution. Note also the large number of
-studies with 100\% forest cover reduction.
-
-\begin{verbatim}
-## Warning: Removed 7 rows containing non-finite values (stat_bin).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-8-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-9-1.pdf}
-
-This immediately shows that for the data related to forest decreases,
-there is almost always a positive flow change. In other words, flow
-almost always increased. However, for increases in forest cover, this is
-not the case, and flow can both increase and decrease. However in both
-cases the variability in the reported change in flow increases with the
-increase in forest cover change.
+studies with 100\% forest cover reduction. Smaller watersheds dominate
+the database with 42\% of the data from watersheds \textless{} 1
+km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10
+km\textsuperscript{2}.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{DataExploration} \caption{Overview of the distributions of some of the variables in the data set}\label{fig:data_graphs}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of increases and decreases in forest cover}\label{fig:increase_decrease}
+\end{figure}
+
+This shows that for the data related to forest decreases, there is
+almost always a positive flow change. In other words, flow almost always
+increased. However, for increases in forest cover, this is not the case,
+and flow can both increase and decrease. However in both cases the
+variability in the reported change in flow increases with the increase
+in forest cover change.
 
@@
 
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign
-## 
-## Parametric coefficients:
-##                     Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)           8.8466     6.5008   1.361    0.175    
-## DeltaF_perc_pos       0.5364     0.1017   5.274 2.52e-07 ***
-## Forest_Signincrease -32.4760     6.9367  -4.682 4.26e-06 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =   0.16   Deviance explained = 16.6%
-## GCV = 3278.3  Scale est. = 3246.8    n = 312
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model1-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  3 / 3
-\end{verbatim}
-
-\begin{longtable}[]{@{}ccc@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \toprule
-\begin{minipage}[b]{0.11\columnwidth}\centering
-Length\strut
-\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
-Class\strut
-\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
-Mode\strut
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
 \end{minipage}\tabularnewline
@@
 \endhead
-\begin{minipage}[t]{0.11\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-NULL\strut
-\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
-NULL\strut
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+8.97\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.55\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.37\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.17\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.53\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.23\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-32.55\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.96\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-4.68\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
 \end{minipage}\tabularnewline
@@
 
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm
-## 
-## Parametric coefficients:
-##                       Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)          2.323e+01  9.189e+00   2.528   0.0120 *  
-## DeltaF_perc_pos      5.424e-01  1.030e-01   5.265 2.64e-07 ***
-## Forest_Signincrease -3.324e+01  6.955e+00  -4.779 2.73e-06 ***
-## Area_km2            -1.867e-06  1.322e-05  -0.141   0.8878    
-## Pa_mm               -1.067e-02  4.944e-03  -2.158   0.0317 *  
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =  0.168   Deviance explained = 17.9%
-## GCV = 3269.3  Scale est. = 3217      n = 312
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model2-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  5 / 5
-\end{verbatim}
-
 \begin{longtable}[]{@{}ccccc@{}}
 \toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
+\begin{minipage}[b]{0.31\columnwidth}\centering
 ~\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
 Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
 Std. Error\strut
@@
 t value\strut
-\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
 Pr(\textgreater\textbar t\textbar)\strut
@@
 \endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-23.23\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-9.189\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.528\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.01197\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+23.27\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+9.21\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.53\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.5424\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.103\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.265\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-2.64e-07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.2\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--33.24\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-6.955\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--4.779\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-2.733e-06\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-33.29\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.99\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-4.77\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--1.867e-06\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-1.322e-05\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.1412\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.8878\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.16\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.87\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.01067\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.004944\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.158\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.03169\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.13\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.03\strut
 \end{minipage}\tabularnewline
@@
 
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + 
-##     Latitude + Longitude
-## 
-## Parametric coefficients:
-##                       Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)          3.504e+01  1.026e+01   3.415 0.000724 ***
-## DeltaF_perc_pos      5.147e-01  1.027e-01   5.014 9.07e-07 ***
-## Forest_Signincrease -3.793e+01  7.230e+00  -5.247 2.90e-07 ***
-## Area_km2            -6.282e-06  1.327e-05  -0.474 0.636137    
-## Pa_mm               -1.304e-02  5.055e-03  -2.580 0.010339 *  
-## Latitude            -3.119e-01  1.255e-01  -2.484 0.013520 *  
-## Longitude           -1.058e-02  3.965e-02  -0.267 0.789747    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =  0.183   Deviance explained = 19.9%
-## GCV = 3230.5  Scale est. = 3158      n = 312
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model3-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  7 / 7
-\end{verbatim}
-
 \begin{longtable}[]{@{}ccccc@{}}
 \toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
+\begin{minipage}[b]{0.31\columnwidth}\centering
 ~\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
 Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
 Std. Error\strut
@@
 \endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-35.04\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-10.26\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.415\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.0007243\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+35.23\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+10.31\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.42\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.5147\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.1027\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.014\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-9.071e-07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.51\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+4.92\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--37.93\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-7.23\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.247\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-2.898e-07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-38.06\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.26\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.24\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--6.282e-06\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-1.327e-05\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.4736\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.6361\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.46\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.64\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.01304\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.005055\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.58\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.01034\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.54\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.3119\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.1255\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.484\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.01352\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.13\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.49\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.01058\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.03965\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.2669\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.7897\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.04\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.28\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.78\strut
 \end{minipage}\tabularnewline
@@
 linear variable and drops the annual average precipitation as a
-variables.
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Latitude + 
-##     Longitude + Dryness
-## 
-## Parametric coefficients:
-##                       Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)          6.031e-01  8.582e+00   0.070 0.944022    
-## DeltaF_perc_pos      5.006e-01  1.036e-01   4.831 2.18e-06 ***
-## Forest_Signincrease -3.898e+01  7.253e+00  -5.375 1.55e-07 ***
-## Area_km2            -7.227e-06  1.317e-05  -0.549 0.583604    
-## Latitude            -3.121e-01  1.249e-01  -2.500 0.012974 *  
-## Longitude           -6.775e-03  3.958e-02  -0.171 0.864199    
-## Dryness              1.370e+01  3.628e+00   3.776 0.000193 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## 
-## R-sq.(adj) =  0.204   Deviance explained = 21.9%
-## GCV = 3199.2  Scale est. = 3125.7    n = 305
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model4-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Model required no smoothing parameter selectionModel rank =  7 / 7
-\end{verbatim}
+variable.
 
@@
 \toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
+\begin{minipage}[b]{0.31\columnwidth}\centering
 ~\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
 Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
 Std. Error\strut
@@
 \endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.6031\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-8.582\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.07027\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.944\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+1.08\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+8.76\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.9\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.5006\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.1036\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-4.831\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-2.178e-06\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+4.74\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--38.98\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-7.253\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.375\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-1.549e-07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-39.12\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.28\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.37\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--7.227e-06\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-1.317e-05\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.5487\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.5836\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.54\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.59\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.3121\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.1249\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.5\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.01297\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.32\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.13\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.51\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--0.006775\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.03958\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.1712\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.8642\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.04\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.19\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.85\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
 \textbf{Dryness}\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-13.7\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-3.628\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.776\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.0001927\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+13.64\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+3.64\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.75\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
 \end{minipage}\tabularnewline
@@
 
+However, the result also indicates possible issues with the data, some
+of the Dryness values are very large (\textgreater{} 4) and these values
+have high leverage in the data. These catchment are listed in Table XX:
+
+\begin{longtable}[]{@{}ccc@{}}
+\toprule
+\begin{minipage}[b]{0.14\columnwidth}\centering
+Latitude\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Longitude\strut
+\end{minipage} & \begin{minipage}[b]{0.42\columnwidth}\centering
+Watershed name\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.14\columnwidth}\centering
+34.67\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-111.7\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Beaver Creek, AZ \#3-2\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+36.4\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-120.4\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Cantua\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+34.43\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-112.3\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+White Spar, Ariz., U.S.A, B\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+32.74\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-111.5\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Natural DRDages, Ariz., U.S.A, A\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
 \begin{verbatim}
@@
 ## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + 
-##     s(Dryness) + length + Assessment_technique + Latitude
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
+##     s(Latitude, bs = "ts") + Precip_data_type + Assessment_technique + 
+##     Forest_type + Hydrological_regime
 ## 
@@
 ##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                  19.80962   20.90807   0.947  0.34433    
-## DeltaF_perc_pos               0.34576    0.10727   3.223  0.00144 ** 
-## Forest_Signincrease         -33.34803    7.26563  -4.590 7.06e-06 ***
-## length                       -0.16503    0.26850  -0.615  0.53935    
-## Assessment_techniqueEA, HM    6.97930   47.05195   0.148  0.88220    
-## Assessment_techniqueHM        1.39799   16.63165   0.084  0.93308    
-## Assessment_techniquePWE       4.12984   19.29598   0.214  0.83070    
-## Assessment_techniquePWE, HM  -5.44499   49.38888  -0.110  0.91230    
-## Assessment_techniqueQPW      -9.50504   26.72374  -0.356  0.72239    
-## Assessment_techniqueSH        2.52746   18.97865   0.133  0.89416    
-## Latitude                     -0.30515    0.09629  -3.169  0.00172 ** 
+## (Intercept)                 -14.1541    26.6625  -0.531  0.59595    
+## DeltaF_perc_pos               0.4961     0.1165   4.257 2.86e-05 ***
+## Forest_Signincrease         -28.4355     8.6675  -3.281  0.00117 ** 
+## Precip_data_typeOB           -1.6722    18.5282  -0.090  0.92816    
+## Precip_data_typeSG           23.1345    21.4482   1.079  0.28172    
+## Assessment_techniqueEA, HM   43.8506    56.2639   0.779  0.43644    
+## Assessment_techniqueHM       14.5238    20.0687   0.724  0.46987    
+## Assessment_techniquePWE      24.0991    20.0296   1.203  0.22996    
+## Assessment_techniquePWE, HM  55.2315    62.4275   0.885  0.37709    
+## Assessment_techniqueQPW      16.5020    30.8514   0.535  0.59317    
+## Assessment_techniqueQPW, EA  36.7969    33.9341   1.084  0.27917    
+## Assessment_techniqueSH       22.5635    22.0306   1.024  0.30666    
+## Forest_typeCF                 2.3144    11.1946   0.207  0.83637    
+## Forest_typeMF               -16.6438    10.5710  -1.574  0.11654    
+## Hydrological_regimeSD        12.8365    12.4739   1.029  0.30437    
 ## ---
@@
 ##               edf Ref.df     F p-value    
-## s(Area_km2)   1      1 0.513   0.474
-## s(Dryness)    1      1 0.002   0.964
-## 
-## R-sq.(adj) =  0.185   Deviance explained = 22.2%
-## GCV = 2181.8  Scale est. = 2072.8    n = 260
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model5-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 15 iterations.
-## The RMS GCV score gradient at convergence was 0.0001863276 .
-## The Hessian was positive definite.
-## Model rank =  29 / 29 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##             k' edf k-index p-value  
-## s(Area_km2)  9   1    1.06   0.825  
-## s(Dryness)   9   1    0.86   0.015 *
+## s(Dryness)  8.493      9 5.989 < 2e-16 ***
+## s(Latitude) 3.907      9 1.695 0.00238 ** 
 ## ---
 ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.305   Deviance explained = 36.7%
+## GCV = 3065.1  Scale est. = 2783.3    n = 298
 \end{verbatim}
 
-\includegraphics{Forest_and_Water_files/figure-latex/model5-2.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model5-3.pdf}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{model6_reduc <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(Latitude, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\NormalTok{Assessment_technique }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2)}
+\KeywordTok{summary}\NormalTok{(model6_reduc)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 ## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + te(Dryness, Latitude, 
-##     k = 5) + Precip_data_type + Assessment_technique + Forest_type + 
-##     Hydrological_regime
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
+##     s(Latitude, bs = "ts") + Assessment_technique + Hydrological_regime
 ## 
@@
 ##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -15.0968    26.7857  -0.564 0.573480    
-## DeltaF_perc_pos               0.4516     0.1187   3.805 0.000175 ***
-## Forest_Signincrease         -27.5255     9.0389  -3.045 0.002552 ** 
-## Precip_data_typeOB           -5.2815    18.9655  -0.278 0.780856    
-## Precip_data_typeSG           11.3166    21.3449   0.530 0.596417    
-## Assessment_techniqueEA, HM   45.5460    57.7615   0.789 0.431077    
-## Assessment_techniqueHM       21.7074    19.8327   1.095 0.274689    
-## Assessment_techniquePWE      29.7823    20.4186   1.459 0.145827    
-## Assessment_techniquePWE, HM  51.4437    62.8574   0.818 0.413831    
-## Assessment_techniqueQPW      19.1277    31.5696   0.606 0.545090    
-## Assessment_techniqueQPW, EA  29.9396    35.2621   0.849 0.396591    
-## Assessment_techniqueSH       28.3985    22.2674   1.275 0.203273    
-## Forest_typeCF                 5.0733    11.4701   0.442 0.658619    
-## Forest_typeMF               -13.2933    10.8133  -1.229 0.219997    
-## Hydrological_regimeSD        13.3327    13.2135   1.009 0.313855    
+## (Intercept)                 -10.1399    18.3918  -0.551 0.581848    
+## DeltaF_perc_pos               0.4702     0.1145   4.107 5.26e-05 ***
+## Forest_Signincrease         -30.3156     8.2051  -3.695 0.000265 ***
+## Assessment_techniqueEA, HM   41.6430    55.5000   0.750 0.453687    
+## Assessment_techniqueHM       17.2590    18.9194   0.912 0.362425    
+## Assessment_techniquePWE      18.4065    18.6123   0.989 0.323542    
+## Assessment_techniquePWE, HM  53.7505    60.2840   0.892 0.373360    
+## Assessment_techniqueQPW       9.2813    29.4994   0.315 0.753278    
+## Assessment_techniqueQPW, EA  30.4467    32.8946   0.926 0.355457    
+## Assessment_techniqueSH       18.6532    21.0133   0.888 0.375472    
+## Hydrological_regimeSD        18.3329     9.5617   1.917 0.056210 .  
 ## ---
@@
 ##               edf Ref.df     F  p-value    
-## te(Dryness,Latitude) 10.7  12.97 3.432 6.26e-05 ***
+## s(Dryness)  8.497      9 5.950  < 2e-16 ***
+## s(Latitude) 3.725      9 1.884 0.000881 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.249   Deviance explained = 31.1%
-## GCV = 3281.3  Scale est. = 2999.2    n = 299
+## R-sq.(adj) =  0.302   Deviance explained = 35.4%
+## GCV = 2973.4  Scale est. = 2746.3    n = 304
 \end{verbatim}
 
-\includegraphics{Forest_and_Water_files/figure-latex/model2c-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 7 iterations.
-## The RMS GCV score gradient at convergence was 0.002397792 .
-## The Hessian was positive definite.
-## Model rank =  39 / 39 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                        k'  edf k-index p-value
-## te(Dryness,Latitude) 24.0 10.7       1    0.52
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model2c-2.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-17-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-18-1.pdf}
 
@@
 \NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Dryness)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
+\StringTok{  }\KeywordTok{filter}\NormalTok{(Area_km2 }\OperatorTok{>}\StringTok{ }\FloatTok{3e+06}\NormalTok{)}
 \end{Highlighting}
 \end{Shaded}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-14-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Assessment_technique,Dryness)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_boxplot}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: Removed 7 rows containing non-finite values (stat_boxplot).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-15-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{filter}\NormalTok{(Dryness }\OperatorTok{>}\StringTok{ }\DecValTok{4}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## # A tibble: 4 x 25
-##   `Watershed #` `Watershed name`     Area_km2 Pa_mm Forest_type Hydrological_re~
-##           <dbl> <chr>                   <dbl> <dbl> <chr>       <chr>           
-## 1            76 Beaver  Creek,  AZ ~     1.47   426 CF          SD              
-## 2            90 Cantua                 119      442 MF          RD              
-## 3           295 White Spar, Ariz., ~     1      549 BF          RD              
-## 4           225 Natural DRDages, Ar~     0.05   452 BF          RD              
-## # ... with 19 more variables: DeltaF_perc <dbl>, DeltaQf_perc <dbl>,
-## #   Precip_data_type <chr>, Assessment_technique <chr>, Source of info <chr>,
-## #   Length of study (years) <dbl>, Notes <chr>, From <dbl>, To <dbl>,
-## #   Location Details AL=aprox location given by author, PL=point location given by author, GE= coordinates guessed from google maps <chr>,
-## #   notes <chr>,
-## #   Location Details AL=aprox location given by author, PL=point location given by author, GM= coordinates guessed from google maps <chr>,
-## #   Latitude <dbl>, Longitude <dbl>, E0 <dbl>, Dryness <dbl>,
-## #   Forest_Sign <chr>, DeltaF_perc_pos <dbl>, length <dbl>
-\end{verbatim}
 
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-17-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }\DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-18-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{tiff}\NormalTok{(}\StringTok{"flow_forest_byArea.tiff"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{2500}\NormalTok{, }\DataTypeTok{height =} \DecValTok{1800}\NormalTok{, }\DataTypeTok{res =} \DecValTok{300}\NormalTok{)}
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(DeltaF_perc_pos, DeltaQf_perc, }\DataTypeTok{colour =}\NormalTok{ Assessment_technique,}\DataTypeTok{size =} \KeywordTok{log10}\NormalTok{(Area_km2), }
-             \DataTypeTok{shape =}\NormalTok{ Forest_Sign)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.5}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"% change in flow"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{axis.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\DecValTok{2}\NormalTok{)),}
-        \DataTypeTok{axis.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \KeywordTok{rel}\NormalTok{(}\FloatTok{1.5}\NormalTok{))) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"% change in forestry"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\CommentTok{#scale_y_log10() +}
-\StringTok{  }\KeywordTok{scale_size_continuous}\NormalTok{(}\DataTypeTok{name =} \StringTok{"log10(Area in km2)"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_colour_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Assessment Technique"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_shape_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Forest cover direction"}\NormalTok{)}
-\KeywordTok{dev.off}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-21-1.pdf}
 
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
-
-Check the size distribition of the catchments
-
 \begin{Shaded}
@@
 \StringTok{  }\KeywordTok{scale_x_log10}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-21-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \NormalTok{total <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(Zhang_all)}
@@
 \end{Shaded}
-
-\begin{verbatim}
-## [1] 0.6570513
-\end{verbatim}
 
@@
 
-\begin{verbatim}
-## Warning: Removed 42 rows containing non-finite values (stat_bin).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-22-1.pdf}
 Essentially, the analysis shows at the moment that their is no evidence
@@
 
+\leavevmode\hypertarget{ref-roche1981}{}%
+Roche, M., 1981. Watershed investigations for development of forest
+resources of the amazon region in french guyana. Tropical Agricultural
+Hydrology. J 75--82.
+
 \leavevmode\hypertarget{ref-rodriguez2010}{}%

### 2021-11-25 paper/Forest_and_Water/Forest_and_Water.Rmd
Update Forest_and_Water.Rmd

@@
   - code: INIA
-    address: INIA, Uruguay
+    address: Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
   - code: IMFIA
-    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Departamento de Montevideo, Uruguay
+    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
 footnote:
@@
 The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian watersheds) a 100% decrease in stream flow for watersheds with 100% forest cover.
+The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian watersheds) a 100% decrease in streamflow for watersheds with 100% forest cover.
 However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
@@
 
-include_graphics("../../data/FAOET0data.png")
+include_graphics("../../data/FAOET0data2.png")
 ```

### 2021-11-25 paper/Forest_and_Water/Forest_and_Water.tex
Update Forest_and_Water.tex

@@
 Sydney, NSW 2006, Australia}
-    \address[INIA]{INIA, Uruguay}
+    \address[INIA]{Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48
+km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
     \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
-Engineering, Universidad de la República, 11200 Montevideo, Departamento
-de Montevideo, Uruguay}
+Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
       \cortext[1]{Corresponding Author}
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
 \end{figure}

### 2021-11-26 paper/Forest_and_Water/Forest_and_Water.Rmd
further updates paper 26112021

--- 2021-11-23 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
 ---
-title: Do larger catchments respond different to forest cover change? Re-analysing a global data set.
+title: Do larger watersheds respond different to forest cover change? Re-analysing a global data set.
 author:
@@
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of quantitative data (37) they showed a relationship between catchment size and decline in streamflow.
+For a subset of quantitative data (37) they showed a relationship between watershed size and decline in streamflow.
 
@@
 
-* the type of analysis, i.e. paired catchment studies, modelling, time series analysis etc.
+* the type of analysis, i.e. paired watershed studies, modelling, time series analysis etc.
 
@@
 
+<!-- correct watershed #76 (Baker1984) to be 157 for DeltaQf_perc -->
+
+```{r}
+Zhang_all <- Zhang_all %>%
+  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
+```
+
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
@@
 However the change in flow is different for forest cover decreases compared to forest cover increases.
-In fact, forest cover increases decrease flow by 32\% less than a similar decrease in forest cover causes flow to increase.
-So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.32)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.
+In fact, forest cover increases decrease flow by 31\% less than a similar decrease in forest cover causes flow to increase.
+So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.31)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.
 
-It is however clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017).
+It is however clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being watershed dryness and area (following @zhang2017).
 
-<!-- At the moment using catchment rainfall, we probably want to straight away use P/E0, and explain that rainfall itself might not be a good predictor.
+<!-- At the moment using watershed rainfall, we probably want to straight away use P/E0, and explain that rainfall itself might not be a good predictor.
 Alternatively we can first show that Latitude is a predictor -->
@@
 
-```{r global_map, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
+```{r global_map, fig.cap="Distribution of included watersheds across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
 
@@
 As indicated, a second hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-As the global map (\@ref(fig:global_map)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
+As the global map (\@ref(fig:global_map)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
 Of interest is whether the spatial clustering also indicates a difference in response to forest cover change.
@@
 The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017.
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
 This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable.
@@
 However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data.
-These catchment are listed in Table XX:
+These watershed are listed in Table XX:
 
@@
 
-model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + s(Dryness) + length +  Assessment_technique + Latitude ,  data = Zhang_all2 %>% filter(Dryness < 4))
-#summary(model5_all)
+model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + s(length, bs="ts") + 
+                    Latitude ,  data = Zhang_all2)
+summary(model5_all)
 #gam.check(model5_all)
 #plot(model5_all)
+#pander(round(summary(model5_all)$p.table,2))
 ```
@@
 model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
+                    s(Area_km2, bs="ts") + s(Dryness, bs="ts" ) + 
+                    s(Latitude, bs="ts" ) +  s(length, bs="ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = Zhang_all2)
+                    Hydrological_regime
+                    , data = Zhang_all2)
 summary(model6_all)
@@
 model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
+                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts") + 
+                      s(Area_km2, bs="ts") +  s(length, bs="ts") +
                     Assessment_technique +
@@
 
-```{r, echo = F}
+Clearly Latitude is masking other factors including the assessment technique and the forest type
+
+<!-- Show boxplots of relationships with latitude-->
+```{r, eval =T, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(Assessment_technique, Latitude)) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
+```{r, eval =T, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(Forest_type, Latitude)) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
+```{r, eval =T, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(Hydrological_regime, Latitude)) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
+Clearly all have at least some relationship with Latitude, therefore are being masked if Latitude is included in the model.
+
+<!-- Take out Latitude -->
+```{r model7_noLat}
+model7_noLat <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + 
+                      s(Area_km2, bs="ts")  +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = Zhang_all2)
+summary(model7_noLat)
+plot(model7_noLat)
+```
+
+```{r, eval =T, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(DeltaQf_perc, length)) + 
+  geom_point(colour = "blue") + theme_bw()
+```
+
+In drier watersheds, changes in forest cover have greater impact on flow, which is similar to @zhang2017.
+This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
+
+In contrast to @filoso2017, we also did not identify an effect of the
+
+Given how skewed Dryness is due tot he few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
+
+<!--reduce drynes < 4 -->
+```{r model7_noLatb}
+model7_noLatb <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(Area_km2, bs="ts") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = Zhang_all2 %>% filter(Dryness < 4))
+summary(model7_noLatb)
+plot(model7_noLatb)
+```
+
+```{r, echo = F, eval =F}
 Zhang_all2 %>%
@@
 Zhang_all2 %>%
-  ggplot(aes(Assessment_technique,Area_km2)) + geom_boxplot()
+  ggplot(aes(Hydrological_regime,Forest_type)) + geom_count() + theme_bw()
 ```
 
-```{r, eval =F}
+```{r, eval =T, echo =F}
 Zhang_all2 %>%
-  filter(Area_km2 > 3e+06)
-
-```
-
-No evidence of effect of area
-
-```{r, eval =F}
-Zhang_all %>%
-  ggplot(aes(Assessment_technique,log10(Area_km2))) + geom_boxplot()
+  ggplot(aes(Assessment_technique,log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") +
+  theme_bw() + xlab("assessment technique") + ylab("log10(Area in km2)")
 ```
@@
 
-<!--Check the size distribution of the catchments -->
+<!--Check the size distribution of the watersheds -->
 
@@
 ```
-Essentially, the analysis shows at the moment that their is no evidence that the size of the catchment influences the
+
+```{r, eval =F, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(log10(Area_km2), DeltaQf_perc, colour = DeltaF_perc)) + 
+  geom_point(alpha = 0.5) + theme_bw()
+```
+
+```{r, eval =F, echo =F}
+Zhang_all2 %>%
+  mutate(size_class = round(log10(Area_km2),0)) %>%
+  ggplot(aes(log10(Area_km2), DeltaQf_perc, group = size_class)) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
+<!--- Analysis of Methods by dryness -->
+
+```{r, eval =F, echo =F}
+Zhang_all2 %>%
+  ggplot(aes(Assessment_technique, Dryness)) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
+Discussion
+======================
+
+Essentially, the analysis shows at the moment that in contrast to @zhang2017 there is no evidence that the size of a watershed influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds.
+In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
+
+As shown earlier, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analsysi (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+This is a confounding factor that is not easily addressed in the regression modelling attempted here.
+Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
+
+There are further confouding factors in the data, which were also classified by @filoso2017 and thesse create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow.
+
+Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
+While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
+Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
 

### 2021-11-26 paper/Forest_and_Water/Forest_and_Water.tex
further updates paper 26112021

--- 2021-11-23 paper/Forest_and_Water/Forest_and_Water.tex
@@
             pdfauthor={},
-            pdftitle={Do larger catchments respond different to forest cover change? Re-analysing a global data set.},
+            pdftitle={Do larger watersheds respond different to forest cover change? Re-analysing a global data set.},
             colorlinks=false,
@@
 
-  \title{Do larger catchments respond different to forest cover change?
+  \title{Do larger watersheds respond different to forest cover change?
 Re-analysing a global data set.}
@@
 there were many interacting factors. For a subset of quantitative data
-(37) they showed a relationship between catchment size and decline in
+(37) they showed a relationship between watershed size and decline in
 streamflow.
@@
 \item
-  the type of analysis, i.e.~paired catchment studies, modelling, time
+  the type of analysis, i.e.~paired watershed studies, modelling, time
   series analysis etc.
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang_all <-}\StringTok{ }\NormalTok{Zhang_all }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{DeltaQf_perc =} \KeywordTok{ifelse}\NormalTok{(}\StringTok{`}\DataTypeTok{Watershed #}\StringTok{`} \OperatorTok{==}\StringTok{ }\DecValTok{76}\NormalTok{,}\DecValTok{157}\NormalTok{,DeltaQf_perc))}
+\end{Highlighting}
+\end{Shaded}
+
 To estimate how the change in streamflow is affected by the change in
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-8.97\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.55\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.37\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.17\strut
+8.77\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.52\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.59\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.11\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.53\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.23\strut
+0.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.77\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--32.55\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.96\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--4.68\strut
+-30.9\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.86\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.27\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 
-While the overall variance explained in this model is not high at 0.16,
+While the overall variance explained in this model is not high at 0.19,
 it clearly supports the hypothesized relationship between the change in
@@
 compared to forest cover increases. In fact, forest cover increases
-decrease flow by 32\% less than a similar decrease in forest cover
+decrease flow by 31\% less than a similar decrease in forest cover
 causes flow to increase. So roughly speaking, a 1\% forest cover
-increase on the average decreases flow by \((1 - 0.32)*0.5\%\), while a
+increase on the average decreases flow by \((1 - 0.31)*0.5\%\), while a
 the percentage forest cover decrease will increase flow by 0.5\%.
@@
 be confounding factors, as alluded to in the methods. The obvious ones
-being catchment dryness and area (following Zhang et al. (2017)).
+being watershed dryness and area (following Zhang et al. (2017)).
 
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-23.27\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-9.21\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.53\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
+18.94\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.78\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.43\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.54\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.2\strut
+0.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.66\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--33.29\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.99\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--4.77\strut
+-31.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.9\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.35\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.16\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.87\strut
+-0.3\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.77\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.13\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.03\strut
+-1.75\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.08\strut
 \end{minipage}\tabularnewline
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
 \end{figure}
@@
 global map (@ref(fig:global\_map)) shows, the distribution of case study
-catchments covers multiple continents and shows some distinct clustering
+watersheds covers multiple continents and shows some distinct clustering
 in parts of the world. Of interest is whether the spatial clustering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-35.23\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-10.31\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.42\strut
+29.99\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+8.65\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.47\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.51\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-4.92\strut
+0.47\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.4\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--38.06\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.26\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.24\strut
+-37.11\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.09\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.46\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.64\strut
+-0.59\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.55\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.17\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.03\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.29\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.74\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
 0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.54\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Longitude}\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
 0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.31\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.13\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.49\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.04\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.28\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.03\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.28\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 changes in runoff. The total explaining power of the model is still low
-at 0.18 suggesting further confounding factors currently not included in
+at 0.22 suggesting further confounding factors currently not included in
 the model.
@@
 also indicated in Zhang et al. (2017). Increased evapotranspiration
-could lead to drier catchments, unless balanced by rainfall (such as
+could lead to drier watersheds, unless balanced by rainfall (such as
 possibly in the tropics). This model introduces the dryness index as a
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-1.08\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-8.76\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.12\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.9\strut
+10.27\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.44\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.38\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.17\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.5\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-4.74\strut
+0.46\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.17\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--39.12\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.28\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.37\strut
+-37.56\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.19\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.07\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.54\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.59\strut
+-0.76\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.45\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.32\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.13\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.51\strut
+-0.28\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.66\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.04\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.19\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.85\strut
+0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.03\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.4\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.69\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-13.64\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-3.64\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.75\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
+6.1\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+3.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.97\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.05\strut
 \end{minipage}\tabularnewline
@@
 of the Dryness values are very large (\textgreater{} 4) and these values
-have high leverage in the data. These catchment are listed in Table XX:
+have high leverage in the data. These watershed are listed in Table XX:
 
@@
 ## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(Latitude, bs = "ts") + Precip_data_type + Assessment_technique + 
-##     Forest_type + Hydrological_regime
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2, bs = "ts") + 
+##     s(Dryness, bs = "ts") + s(length, bs = "ts") + Latitude
 ## 
@@
 ##                      Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -14.1541    26.6625  -0.531  0.59595    
-## DeltaF_perc_pos               0.4961     0.1165   4.257 2.86e-05 ***
-## Forest_Signincrease         -28.4355     8.6675  -3.281  0.00117 ** 
-## Precip_data_typeOB           -1.6722    18.5282  -0.090  0.92816    
-## Precip_data_typeSG           23.1345    21.4482   1.079  0.28172    
-## Assessment_techniqueEA, HM   43.8506    56.2639   0.779  0.43644    
-## Assessment_techniqueHM       14.5238    20.0687   0.724  0.46987    
-## Assessment_techniquePWE      24.0991    20.0296   1.203  0.22996    
-## Assessment_techniquePWE, HM  55.2315    62.4275   0.885  0.37709    
-## Assessment_techniqueQPW      16.5020    30.8514   0.535  0.59317    
-## Assessment_techniqueQPW, EA  36.7969    33.9341   1.084  0.27917    
-## Assessment_techniqueSH       22.5635    22.0306   1.024  0.30666    
-## Forest_typeCF                 2.3144    11.1946   0.207  0.83637    
-## Forest_typeMF               -16.6438    10.5710  -1.574  0.11654    
-## Hydrological_regimeSD        12.8365    12.4739   1.029  0.30437    
+## (Intercept)          18.85584    5.84673   3.225 0.001419 ** 
+## DeltaF_perc_pos       0.37334    0.09042   4.129 4.90e-05 ***
+## Forest_Signincrease -33.25371    6.40370  -5.193 4.15e-07 ***
+## Latitude             -0.29544    0.08775  -3.367 0.000873 ***
 ## ---
@@
 ##                   edf Ref.df     F p-value  
-## s(Dryness)  8.493      9 5.989 < 2e-16 ***
-## s(Latitude) 3.907      9 1.695 0.00238 ** 
+## s(Area_km2) 3.174e-06      9 0.000  0.3914  
+## s(Dryness)  1.293e-05      9 0.000  0.5926  
+## s(length)   5.991e+00      9 1.221  0.0779 .
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.305   Deviance explained = 36.7%
-## GCV = 3065.1  Scale est. = 2783.3    n = 298
+## R-sq.(adj) =  0.227   Deviance explained = 25.2%
+## GCV =   2018  Scale est. = 1944.1    n = 273
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2, bs = "ts") + 
+##     s(Dryness, bs = "ts") + s(Latitude, bs = "ts") + s(length, 
+##     bs = "ts") + Precip_data_type + Assessment_technique + Forest_type + 
+##     Hydrological_regime
+## 
+## Parametric coefficients:
+##                             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                   2.5565    25.5148   0.100 0.920272    
+## DeltaF_perc_pos               0.3234     0.1058   3.057 0.002482 ** 
+## Forest_Signincrease         -27.2070     7.5951  -3.582 0.000411 ***
+## Precip_data_typeOB            2.9296    15.9005   0.184 0.853974    
+## Precip_data_typeSG           23.9270    18.5848   1.287 0.199159    
+## Assessment_techniqueEA, HM   14.8215    46.1143   0.321 0.748176    
+## Assessment_techniqueHM       -7.5091    18.4252  -0.408 0.683967    
+## Assessment_techniquePWE       8.9562    19.7884   0.453 0.651240    
+## Assessment_techniquePWE, HM  19.6741    52.6286   0.374 0.708856    
+## Assessment_techniqueQPW      -7.3110    28.3315  -0.258 0.796583    
+## Assessment_techniqueSH        2.8647    18.4698   0.155 0.876871    
+## Forest_typeCF                 4.9185     9.2133   0.534 0.593935    
+## Forest_typeMF               -10.6875     9.6354  -1.109 0.268444    
+## Hydrological_regimeSD         6.9324    10.5522   0.657 0.511825    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                   edf Ref.df     F p-value   
+## s(Area_km2) 1.754e-07      9 0.000 0.50299   
+## s(Dryness)  1.784e-06      9 0.000 0.87130   
+## s(Latitude) 3.607e+00      9 1.648 0.00213 **
+## s(length)   5.815e+00      9 1.067 0.11589   
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.231   Deviance explained = 29.6%
+## GCV = 2158.9  Scale est. = 1969.5    n = 267
 \end{verbatim}
@@
 \NormalTok{model6_reduc <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
-\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(Latitude, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(Latitude, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
+\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{  }\KeywordTok{s}\NormalTok{(length, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
 \StringTok{                    }\NormalTok{Assessment_technique }\OperatorTok{+}
@@
 ## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(Latitude, bs = "ts") + Assessment_technique + Hydrological_regime
+##     s(Latitude, bs = "ts") + s(Area_km2, bs = "ts") + s(length, 
+##     bs = "ts") + Assessment_technique + Hydrological_regime
 ## 
@@
 ##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -10.1399    18.3918  -0.551 0.581848    
-## DeltaF_perc_pos               0.4702     0.1145   4.107 5.26e-05 ***
-## Forest_Signincrease         -30.3156     8.2051  -3.695 0.000265 ***
-## Assessment_techniqueEA, HM   41.6430    55.5000   0.750 0.453687    
-## Assessment_techniqueHM       17.2590    18.9194   0.912 0.362425    
-## Assessment_techniquePWE      18.4065    18.6123   0.989 0.323542    
-## Assessment_techniquePWE, HM  53.7505    60.2840   0.892 0.373360    
-## Assessment_techniqueQPW       9.2813    29.4994   0.315 0.753278    
-## Assessment_techniqueQPW, EA  30.4467    32.8946   0.926 0.355457    
-## Assessment_techniqueSH       18.6532    21.0133   0.888 0.375472    
-## Hydrological_regimeSD        18.3329     9.5617   1.917 0.056210 .  
+## (Intercept)                  11.5706    17.3998   0.665  0.50666    
+## DeltaF_perc_pos               0.3094     0.1038   2.979  0.00317 ** 
+## Forest_Signincrease         -29.9159     7.1426  -4.188 3.88e-05 ***
+## Assessment_techniqueEA, HM   11.4108    45.6037   0.250  0.80262    
+## Assessment_techniqueHM       -4.6254    17.2047  -0.269  0.78827    
+## Assessment_techniquePWE       3.8045    18.5150   0.205  0.83736    
+## Assessment_techniquePWE, HM  17.3096    50.9143   0.340  0.73416    
+## Assessment_techniqueQPW     -14.1255    26.9676  -0.524  0.60088    
+## Assessment_techniqueSH       -0.6693    17.6262  -0.038  0.96974    
+## Hydrological_regimeSD        12.3022     8.4529   1.455  0.14680    
 ## ---
@@
 ##                   edf Ref.df     F  p-value    
-## s(Dryness)  8.497      9 5.950  < 2e-16 ***
-## s(Latitude) 3.725      9 1.884 0.000881 ***
+## s(Dryness)  1.026e-05      9 0.000 0.885487    
+## s(Latitude) 3.300e+00      9 1.829 0.000683 ***
+## s(Area_km2) 2.280e-06      9 0.000 0.848293    
+## s(length)   5.575e+00      9 0.901 0.176727    
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.302   Deviance explained = 35.4%
-## GCV = 2973.4  Scale est. = 2746.3    n = 304
+## R-sq.(adj) =  0.228   Deviance explained = 27.9%
+## GCV = 2085.2  Scale est. = 1941.1    n = 273
 \end{verbatim}
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-17-1.pdf}
+Clearly Latitude is masking other factors including the assessment
+technique and the forest type
 
 \includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-18-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-19-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
+
+Clearly all have at least some relationship with Latitude, therefore are
+being masked if Latitude is included in the model.
 
@@
 \begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{filter}\NormalTok{(Area_km2 }\OperatorTok{>}\StringTok{ }\FloatTok{3e+06}\NormalTok{)}
+\NormalTok{model7_noLat <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }
+\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{)  }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2)}
+\KeywordTok{summary}\NormalTok{(model7_noLat)}
 \end{Highlighting}
@@
 
-No evidence of effect of area
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
+##     s(Area_km2, bs = "ts") + Precip_data_type + Assessment_technique + 
+##     Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -12.5291    22.7513  -0.551 0.582282    
+## DeltaF_perc_pos               0.4348     0.1033   4.209 3.46e-05 ***
+## Forest_Signincrease         -27.2295     7.6821  -3.545 0.000461 ***
+## Precip_data_typeOB          -12.4249    15.4195  -0.806 0.421049    
+## Precip_data_typeSG           11.5676    17.1018   0.676 0.499348    
+## Assessment_techniqueEA, HM   19.1901    50.7190   0.378 0.705450    
+## Assessment_techniqueHM       21.6621    17.2625   1.255 0.210578    
+## Assessment_techniquePWE      42.2758    17.2104   2.456 0.014642 *  
+## Assessment_techniquePWE, HM  23.5495    52.6315   0.447 0.654903    
+## Assessment_techniqueQPW      22.1990    27.1887   0.816 0.414924    
+## Assessment_techniqueQPW, EA  41.0844    30.1936   1.361 0.174707    
+## Assessment_techniqueSH       38.3711    18.7848   2.043 0.042025 *  
+## Forest_typeCF                -3.7900     9.7088  -0.390 0.696562    
+## Forest_typeMF               -16.3363     9.3408  -1.749 0.081404 .  
+## Hydrological_regimeSD        -1.1868    10.9105  -0.109 0.913460    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                   edf Ref.df     F p-value  
+## s(Dryness)  3.995e+00      9 1.162   0.024 *
+## s(Area_km2) 2.228e-07      9 0.000   0.439  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.202   Deviance explained =   25%
+## GCV =   2510  Scale est. = 2350      n = 298
+\end{verbatim}
 
@@
 \begin{Highlighting}[]
-\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Assessment_technique,}\KeywordTok{log10}\NormalTok{(Area_km2))) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_boxplot}\NormalTok{()}
+\KeywordTok{plot}\NormalTok{(model7_noLat)}
 \end{Highlighting}
@@
 
+\includegraphics{Forest_and_Water_files/figure-latex/model7_noLat-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model7_noLat-2.pdf}
+
+\begin{verbatim}
+## Warning: Removed 32 rows containing missing values (geom_point).
+\end{verbatim}
+
 \includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-21-1.pdf}
+
+In drier watersheds, changes in forest cover have greater impact on
+flow, which is similar to Zhang et al. (2017). This is most likely
+because in these watersheds the overall flow is surface flow dominated
+and therefore the buffering that is afforded by the groundwater inputs
+is not as great. As we don't have a separate variable for groundwater
+inputs (although this effect is estimated in many studies), we cannot
+analyse this effect separately.
+
+In contrast to Filoso et al. (2017), we also did not identify an effect
+of the
+
+Given how skewed Dryness is due tot he few watersheds that have very
+high dryness values, it is worth investigating what excluding these 4
+watersheds from the data means for the relationships.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{model7_noLatb <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\CommentTok{#s(Latitude, bs="ts") + }
+\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2 }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(Dryness }\OperatorTok{<}\StringTok{ }\DecValTok{4}\NormalTok{))}
+\KeywordTok{summary}\NormalTok{(model7_noLatb)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
+##     s(Area_km2, bs = "ts") + Precip_data_type + Assessment_technique + 
+##     Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -16.8716    22.8324  -0.739 0.460577    
+## DeltaF_perc_pos               0.4205     0.1036   4.060 6.41e-05 ***
+## Forest_Signincrease         -27.4323     7.6284  -3.596 0.000383 ***
+## Precip_data_typeOB          -12.8820    15.2846  -0.843 0.400065    
+## Precip_data_typeSG           11.9340    16.9606   0.704 0.482257    
+## Assessment_techniqueEA, HM   25.8163    50.4996   0.511 0.609607    
+## Assessment_techniqueHM       26.4681    17.4660   1.515 0.130815    
+## Assessment_techniquePWE      48.3282    17.6030   2.745 0.006440 ** 
+## Assessment_techniquePWE, HM  29.5466    52.3323   0.565 0.572808    
+## Assessment_techniqueQPW      29.1804    27.2233   1.072 0.284707    
+## Assessment_techniqueQPW, EA  48.6163    30.1921   1.610 0.108492    
+## Assessment_techniqueSH       43.7353    18.9899   2.303 0.022019 *  
+## Forest_typeCF                -4.2723     9.6304  -0.444 0.657661    
+## Forest_typeMF               -14.2554     9.4382  -1.510 0.132089    
+## Hydrological_regimeSD        -2.6825    10.8541  -0.247 0.804980    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                   edf Ref.df     F p-value   
+## s(Dryness)  3.374e+00      9 1.576 0.00703 **
+## s(Area_km2) 3.036e-09      9 0.000 0.41938   
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.209   Deviance explained = 25.6%
+## GCV = 2462.7  Scale est. = 2308.7    n = 294
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(model7_noLatb)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-2.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-23-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-24-1.pdf}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-25-1.pdf}
 
@@
 
-Essentially, the analysis shows at the moment that their is no evidence
-that the size of the catchment influences the
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Essentially, the analysis shows at the moment that in contrast to Zhang
+et al. (2017) there is no evidence that the size of a watershed
+influences the change in the streamflow as a result of changes in
+forestry. If anything the scatter in the data (in the change in flow) is
+greater for the smaller watersheds then for the larger watersheds. In
+other words, the response to changes in forest cover is more consistent
+for larger watersheds than it is for smaller watersheds.
+
+As shown earlier, most of the smaller watersheds are ``real observed
+data'' using paired watershed studies, while for larger watersheds, the
+analysis are mostly based on modelling approximations using either
+elasticity analysis (EA), Hydrological modelling (HM) or a combined use
+of statistical methods (SH) or quasi paired watershed analsysi (QPW),
+thus all providing an approximation of the effect of forestry on
+streamflow rather than a direct comparison of watersheds. This is a
+confounding factor that is not easily addressed in the regression
+modelling attempted here. Furthermore, the catchments analysed using EA,
+are concentrated in the drier end of the Dryness index scale compared to
+the other methods, with only the paired watershed experiment (PWE)
+assessment technique covering the full range of dryness indices.
+
+There are further confouding factors in the data, which were also
+classified by Filoso et al. (2017) and thesse create biases in the data
+set that can impact the overall assessment. For example, snow dominated
+hydrological regimes (SD), which are weakly significant, are dominated
+by Coniferous Forests (CF), while the majority of the rain dominated
+regimes are all broadleaf forests (BF). However, the forest type
+classification is very coarse and does not fully capture possible
+physiological differences that could affect evapotranspiration and
+therefore changes in streamflow.
+
+Apart from a difficulty of analysing complex confounding factors in the
+data, a general limitation of the type of analysis presented is that
+this work does not consider the spatial arrangement of the forest
+clearing in the catchments. While for fully or almost fully cleared
+smaller catchments this might not be an issue, it is perceivable that
+for larger catchments being partially cleared, a interaction between
+spatial location and clearing could be a factor in determining the
+change in streamflow. Clearing head water catchments on shallower soils
+might have a larger impact than clearing in downstream areas on deeper
+soils.
 

### 2021-11-26 paper/Forest_and_Water/Forest_and_Water.tex

@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
 \end{figure}

### 2021-11-28 paper/Forest_and_Water/Forest_and_Water.Rmd
further changes to the paper

@@
 In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
-Using this information  reference evapotranspiration (E_{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET0) Climate Databasev2 [@trabucco2018], if a value of E0 was not available from the original papers.
+Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
 For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 
-Similar to @zhang2017, the "dryness index" was calculated as:
+Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
 $$\tag{1}
 \begin{aligned}
-D = \frac{E0}{Pa}
+D = \frac{E_{0}}{Pa}
 \end{aligned}$$
@@
 <!--remove the Amazon from the analysis -->
-```{r}
+```{r, echo = F}
 Zhang_all <- Zhang_all %>%
@@
 
-```{r}
+```{r,echo = F}
 Zhang_all <- Zhang_all %>%
@@
 
-The general model tested is
+The general model tested is:
 
@@
 This means that a smooth function $s()$ is applied to the $Z_i$ variables.
+For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant [@wood2006].
 
@@
 
-Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover.
+Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
+
+$$\tag{3}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\end{aligned}$$
 
@@
 ```{r, echo =F}
-pander(round(summary(linear_model_simple)$p.table,2))
+pander(round(summary(linear_model_simple)$p.table,2), caption="Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
 ```
 
-While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow.
+While the overall variance explained in this model is not high with an adjusted *r^2^* of `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%.
@@
 
-It is however clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being watershed dryness and area (following @zhang2017).
+It is however it is clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods.
+The obvious ones being watershed dryness and area (following @zhang2017):
+
+$$\tag{4}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
+\end{aligned}$$
+Where $Pa_mm$ is the annual average rainfall in mm.
 
@@
 ```{r, echo = F}
-pander(round(summary(linear_model2)$p.table,2))
+pander(round(summary(linear_model2)$p.table,2), caption="Summary of the second model, taking into account the annual rainfall and the area of the watershed")
 ```
@@
 
-As indicated, a second hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
+As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
 As the global map (\@ref(fig:global_map)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
-Of interest is whether the spatial clustering also indicates a difference in response to forest cover change.
+Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
+
+$$\tag{5}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
+\end{aligned}$$
 
@@
 ```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2))
+pander(round(summary(model3_includeLatLong)$p.table,2), caption= "Results of the model including Latitude and Longitude")
 ```
@@
 In addition, the model suggests an influence of the annual average rainfall, with wetter watersheds having slightly lower changes in runoff.
-The total explaining power of the model is still low at `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further confounding factors currently not included in the model.
+The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors that are currently not included in the model.
 
-There is no relationship with Longitude, suggesting that the different continents do not show a trend in the East-West direction.
+There is no relationship with Longitude, suggesting that the watersheds across different continents do not show an East-West direction trend.
 
@@
 
-The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017.
+Latitude might be indicating an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017.
 Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
-This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable.
+The North-South trend in dryness would be related to Latitude.
+This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from.
 
@@
 ```{r, echo =F}
-pander(round(summary(model4_dryness)$p.table,2))
+pander(round(summary(model4_dryness)$p.table,2), caption = "Results of the model replacing the annual precipitation with the dryness index")
 ```
@@
 However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data.
-These watershed are listed in Table XX:
+These watersheds are listed in Table 5:
 
 ```{r, echo = F}
-pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`))
+pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "Watersheds for which the dryness inddex > 4")
 ```
+
+### Are some of the variables possibly non-linear?
+
+The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response.
+This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackon2005).
+In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.
 
@@
 model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + s(length, bs="ts") + 
-                    Latitude ,  data = Zhang_all2)
+                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + 
+                    s(length, bs="ts") + s(Latitude, bs = "ts") ,
+                  data = Zhang_all2)
 summary(model5_all)
@@
 #plot(model5_all)
-#pander(round(summary(model5_all)$p.table,2))
+pander(round(summary(model5_all)$p.table,2))
+pander(round(summary(model5_all)$s.table,2))
 ```
 <!-- model 2c include smooth variables -->
+
+The overall explaining power of the model increases with an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)`, but little changes in the significance of the variables.
+
+A final full model
 
@@
 
-```{r model6_reduc}
+```{r model6_reduc, eval = F}
 model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
                     s(Dryness, bs="ts" ) + 
-                      s(Area_km2, bs="ts")  +
+                      s(Area_km2, bs="ts")  + s(length, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
@@
 ```
+
+```{r model7_noLat_intLength}
+model7_noLat_intlength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="tp" ) + 
+                      s(Area_km2, bs="tp")  + s(length, bs = "tp") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, 
+                    data = Zhang_all2)
+summary(model7_noLat_intlength)
+plot(model7_noLat_intlength)
+```
+
+The flexible nature of the splines and the shrinkage applied to the splines means that the length variable can capture variation in the data, but it is unclear what exactly is captured.
+The shape of the resulting conditional response does not reflect the reponses highlighted in @filoso2017 and @jackson2005.
+Reducing the flexibility of the splines, or fitting a linear term results in "length" not being significant, which suggests that this variable is not helpful in explaining the variation in the data.
 

### 2021-11-28 paper/Forest_and_Water/Forest_and_Water.tex
further changes to the paper

@@
 Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
-decrease in stream flow for watersheds with 100\% forest cover. However,
+decrease in streamflow for watersheds with 100\% forest cover. However,
 on the other end of the spectrum, in a series of French watersheds
@@
 percent change forestry and modulated by the direction of the change,
-either an increase in forest cover, or decrease in forest cover.
+either an increase in forest cover, or decrease in forest cover:
+
+\[\tag{3}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\end{aligned}\]
 
@@
 
-While the overall variance explained in this model is not high at 0.19,
-it clearly supports the hypothesized relationship between the change in
-forest cover and the change in flow. The model suggests that for every
-1\% change in forest cover, on the average, the flow changes 0.5\%.
-However the change in flow is different for forest cover decreases
-compared to forest cover increases. In fact, forest cover increases
-decrease flow by 31\% less than a similar decrease in forest cover
-causes flow to increase. So roughly speaking, a 1\% forest cover
-increase on the average decreases flow by \((1 - 0.31)*0.5\%\), while a
-the percentage forest cover decrease will increase flow by 0.5\%.
-
-It is however clear from the lack of explaining power, that there could
-be confounding factors, as alluded to in the methods. The obvious ones
-being watershed dryness and area (following Zhang et al. (2017)).
+While the overall variance explained in this model is not high with an
+adjusted \emph{r\textsuperscript{2}} of 0.19, it clearly supports the
+hypothesized relationship between the change in forest cover and the
+change in flow. The model suggests that for every 1\% change in forest
+cover, on the average, the flow changes 0.5\%. However the change in
+flow is different for forest cover decreases compared to forest cover
+increases. In fact, forest cover increases decrease flow by 31\% less
+than a similar decrease in forest cover causes flow to increase. So
+roughly speaking, a 1\% forest cover increase on the average decreases
+flow by \((1 - 0.31)*0.5\%\), while a the percentage forest cover
+decrease will increase flow by 0.5\%.
+
+It is however it is clear from the lack of explaining power, that there
+could be confounding factors, as alluded to in the methods. The obvious
+ones being watershed dryness and area (following Zhang et al. (2017)):
+
+\[\tag{4}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
+\end{aligned}\] Where \(Pa_mm\) is the annual average rainfall in mm.
+
+\begin{longtable}[]{@{}ccccc@{}}
+\caption{Summary of the second model, taking into account the annual
+rainfall and the area of the watershed}\tabularnewline
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+18.94\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.78\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.43\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.66\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-31.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.9\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.35\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Area\_km2}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.77\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Pa\_mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.75\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.08\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+Including area and annual precipitation does not really improve the
+overall explaining power of the model, in fact, annual precipitation
+appears to be only a very small confounding factor, representing only a
+-0.01/\% partial effect in the change in streamflow, holding all other
+factors constant. In contrast to earlier reported studies (Filoso et
+al., 2017; Zhang et al., 2017), watershed area has no effect on the
+change in stream flow. This supports our approach (in contrast to Zhang
+et al. (2017)) to consider watershed area as a continuous variable and
+making no separation between larger and smaller watersheds The main
+effects remain the change in forest cover and whether this is an
+increase or decrease.
+
+\hypertarget{the-effect-of-location-on-the-globe}{%
+\subsection{The effect of location on the
+globe}\label{the-effect-of-location-on-the-globe}}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\end{figure}
+
+As indicated, a further hypothesis relates to whether there is a strong
+spatial global gradient as captured by latitude and longitude. As the
+global map (@ref(fig:global\_map)) shows, the distribution of case study
+watersheds covers multiple continents and shows some distinct clustering
+in parts of the world. Of interest is whether the spatial clustering
+also indicates a difference in response to forest cover change:
+
+\[\tag{5}
+\begin{aligned}
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
+\end{aligned}\]
+
+\begin{longtable}[]{@{}ccccc@{}}
+\caption{Results of the model including Latitude and
+Longitude}\tabularnewline
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+29.99\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+8.65\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.47\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.47\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.4\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-37.11\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.09\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Area\_km2}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.59\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.55\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Pa\_mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.17\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.03\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.29\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.74\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Longitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.03\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.28\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.78\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+This linear model shows that there is a significant gradient in the
+Latitude and with annual average rainfall, with watersheds closer to the
+equator having lower changes in the runoff compare to watersheds further
+away from the equator. This suggests an influence of radiation, which
+will be tested next. In addition, the model suggests an influence of the
+annual average rainfall, with wetter watersheds having slightly lower
+changes in runoff. The total explaining power of the model is still low
+with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further
+factors that are currently not included in the model.
+
+There is no relationship with Longitude, suggesting that the watersheds
+across different continents do not show an East-West direction trend.
+
+\hypertarget{impact-of-the-dryness-index}{%
+\subsection{Impact of the dryness
+index}\label{impact-of-the-dryness-index}}
+
+Latitude might be indicating an influence of radiation on
+evapotranspiration, and most likely related to the dryness index, as
+also indicated in Zhang et al. (2017). Increased evapotranspiration
+could lead to drier watersheds, unless balanced by rainfall (such as
+possibly in the tropics). The North-South trend in dryness would be
+related to Latitude. This model introduces the dryness index as a linear
+variable and drops the annual average precipitation as a variable, as
+dryness is calculated from.
 
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-18.94\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.78\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.43\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.5\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.66\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--31.54\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.9\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.35\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.3\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.77\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1.75\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.08\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-Including area and annual precipitation does not really improve the
-overall explaining power of the model, in fact, annual precipitation
-appears to be only a very small confounding factor, representing only a
--0.01/\% partial effect in the change in streamflow, holding all other
-factors constant. In contrast to earlier reported studies (Filoso et
-al., 2017; Zhang et al., 2017), watershed area has no effect on the
-change in stream flow. This supports our approach (in contrast to Zhang
-et al. (2017)) to consider watershed area as a continuous variable and
-making no separation between larger and smaller watersheds The main
-effects remain the change in forest cover and whether this is an
-increase or decrease.
-
-\hypertarget{the-effect-of-location-on-the-globe}{%
-\subsection{The effect of location on the
-globe}\label{the-effect-of-location-on-the-globe}}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
-\end{figure}
-
-As indicated, a second hypothesis relates to whether there is a strong
-spatial global gradient as captured by latitude and longitude. As the
-global map (@ref(fig:global\_map)) shows, the distribution of case study
-watersheds covers multiple continents and shows some distinct clustering
-in parts of the world. Of interest is whether the spatial clustering
-also indicates a difference in response to forest cover change.
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-29.99\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-8.65\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.47\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.47\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.4\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--37.11\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.09\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.59\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.55\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.17\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.03\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.29\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.74\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.03\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.28\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.78\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-This linear model shows that there is a significant gradient in the
-Latitude and with annual average rainfall, with watersheds closer to the
-equator having lower changes in the runoff compare to watersheds further
-away from the equator. This suggests an influence of radiation, which
-will be tested next. In addition, the model suggests an influence of the
-annual average rainfall, with wetter watersheds having slightly lower
-changes in runoff. The total explaining power of the model is still low
-at 0.22 suggesting further confounding factors currently not included in
-the model.
-
-There is no relationship with Longitude, suggesting that the different
-continents do not show a trend in the East-West direction.
-
-\hypertarget{impact-of-the-dryness-index}{%
-\subsection{Impact of the dryness
-index}\label{impact-of-the-dryness-index}}
-
-The previous analysis suggests an influence of radiation on
-evapotranspiration, and most likely related to the dryness index, as
-also indicated in Zhang et al. (2017). Increased evapotranspiration
-could lead to drier watersheds, unless balanced by rainfall (such as
-possibly in the tropics). This model introduces the dryness index as a
-linear variable and drops the annual average precipitation as a
-variable.
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
 10.27\strut
@@
 of the Dryness values are very large (\textgreater{} 4) and these values
-have high leverage in the data. These watershed are listed in Table XX:
+have high leverage in the data. These watersheds are listed in Table XX:
 

### 2021-11-30 paper/Forest_and_Water/Forest_and_Water.Rmd
further changes to paper 01122021

@@
 Zhang_small <- read_csv("../../data/TablesZhangetal2017_small.csv")
-Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv")
-
+Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv") %>%
+  mutate(Complete = 1,
+         `Possible duplicate` = 0)
+new_data <- read_csv("../../data/TableNewData.csv")
 ```
@@
 names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
+names(new_data)[3:4] <- c("Area_km2", "Pa_mm")
 names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
 names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
+names(new_data)[5:6] <- c("Forest_type", "Hydrological_regime")
 names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
 names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
+names(new_data)[9:10] <- c("Precip_data_type", "Assessment_technique")
+
 ```
@@
 Zhang_small$To <- as.numeric(Zhang_small$To)
-Zhang_all <- bind_rows(Zhang_large,Zhang_small)
+Zhang_all <- bind_rows(Zhang_large,Zhang_small) %>%
+  mutate(dataset = "original Zhang et al data")
+new_data <- new_data %>%
+  mutate(dataset = "new data")
+All_data <- bind_rows(Zhang_all, new_data)
+```
 
+<!--filter duplicates -->
+```{r, echo = F}
+All_data <- All_data %>%
+  mutate(`Possible duplicate` = 
+           ifelse(is.na(`Possible duplicate`)==T,0,`Possible duplicate`),
+         `Possible duplicate` = as.numeric(`Possible duplicate`)) %>%
+  filter(`Possible duplicate` != 1)
 ```
+Removing possible duplicates removes 29 watersheds from the overall data set
 
@@
 ```{r, echo = F}
-Zhang_all <- left_join(Zhang_all %>% select(-c(E0, Longitude, Latitude)),PET_all, by = "Watershed #")
+All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
+All_data <- All_data %>%
+  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
+  select(-c(E0.x,E0.y)) #%>%
+  #rename(Latitude = Latitude.x,
+#         Longitude = Longitude.x)
+
 # calculate dryness index
-Zhang_all <- Zhang_all %>%
+All_data <- All_data %>%
   mutate(Dryness = E0/Pa_mm)
@@
 ```{r, echo = F}
-Zhang_all <- Zhang_all %>%
-  filter(`Watershed #` > 1)
+All_data <- All_data %>%
+  filter(`Watershed #` != 1)
 ```
@@
 ```{r,echo = F}
-Zhang_all <- Zhang_all %>%
+All_data <- All_data %>%
   mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
@@
 png("DataExploration.png", width = 2500, height = 1800, res = 300)
-Zhang_all %>%
+All_data %>%
   pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
@@
 ```{r, echo = F}
-Zhang_all2 <- Zhang_all %>%
+All_data2 <- All_data %>%
   mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
@@
 png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
-Zhang_all2 %>%
+All_data2 %>%
   ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
@@
                              Forest_Sign, 
-                           data = Zhang_all2)
+                           data = All_data2 %>% filter(dataset == "original Zhang et al data"))
+#summary(linear_model_simple)
+#gam.check(linear_model_simple)
+
+```
+
+```{r model1b, echo = F}
+linear_model_simple_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
+                             Forest_Sign, 
+                           data = All_data2)
 #summary(linear_model_simple)
@@
 ```
+```{r, echo =F}
+pander(round(summary(linear_model_simple_b)$p.table,2), caption="Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets")
+```
 
 While the overall variance explained in this model is not high with an adjusted *r^2^* of `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow.
-The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%.
+The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
 However the change in flow is different for forest cover decreases compared to forest cover increases.
-In fact, forest cover increases decrease flow by 31\% less than a similar decrease in forest cover causes flow to increase.
-So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.31)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.
+In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase.
+So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.
 
-It is however it is clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods.
+Checking this result against including the new data indicates that this mainly strengthens the difference between the forest cover increases and decreases, and decreases the mean decrease in flow as a result of forest cover change
+
+It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
 The obvious ones being watershed dryness and area (following @zhang2017):
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
 \end{aligned}$$
-Where $Pa_mm$ is the annual average rainfall in mm.
+
+Where $Pa_mm$ is the annual average rainfall in mm, and a log base 10 transformation is applied to the area variable given that the distribution of Area is higly skewed (Figure 1)
 
@@
 
-linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
+linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm, data = All_data2)
 #summary(linear_model2)
@@
 
-Including area and annual precipitation does not really improve the overall explaining power of the model, in fact, annual precipitation appears to be only a very small confounding factor, representing only a -0.01/% partial effect in the change in streamflow, holding all other factors constant.
-In contrast to earlier reported studies [@zhang2017; @filoso2017], watershed area has no effect on the change in stream flow.
+Including area and annual precipitation slightly improves the overall explaining power of the model.
+Annual precipitation is in fact not significant.
+Relative to earlier reported studies [@zhang2017; @filoso2017], the log base 10 transformed watershed area indicates a p-value of only `r round(summary(linear_model2)$p.table[4,4],2)`, suggesting a marginal impact on the change in stream flow.
 This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds The main effects remain the change in forest cover and whether this is an increase or decrease.
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}$$
@@
 ```{r model3, echo = F}
-model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
+model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm + Latitude + Longitude, data = All_data2)
 #summary(model3_includeLatLong)
@@
 ```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2), caption= "Results of the model including Latitude and Longitude")
+pander(round(summary(model3_includeLatLong)$p.table,2), caption= "Results of the model including Latitude and Longitude including new data")
 ```
 
-This linear model shows that there is a significant gradient in the Latitude and with annual average rainfall, with watersheds closer to the equator having lower changes in the runoff compare to watersheds further away from the equator.
-This suggests an influence of radiation, which will be tested next.
-In addition, the model suggests an influence of the annual average rainfall, with wetter watersheds having slightly lower changes in runoff.
+There appears to be no significant latitudinal gradient.
 The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors that are currently not included in the model.
 
-There is no relationship with Longitude, suggesting that the watersheds across different continents do not show an East-West direction trend.
+Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow as also indicated in @zhang2017.
 
@@
 
-Latitude might be indicating an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017.
+Climate, and in particular evapotranspiration would be expected to have a significant effect on the streamflow change as represented by the dryness index, which is also indicated in @zhang2017.
 Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
-The North-South trend in dryness would be related to Latitude.
-This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from.
+This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from the precipitation.
+It also drops the Latitude and Longitude as these are indicated not to be significant.
 
@@
 #  mutate(length = To - From)
-model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 +  
-                        Latitude + Longitude + Dryness, data = Zhang_all2)
+model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) +  
+                        Dryness, data = All_data2)
 #summary(model4_dryness)
@@
 ```{r model5, echo = F}
-Zhang_all2 <- Zhang_all2 %>%
+All_data2 <- All_data2 %>%
   mutate(length = To - From)
@@
 model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + 
-                    s(length, bs="ts") + s(Latitude, bs = "ts") ,
-                  data = Zhang_all2)
+                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
+                    s(length, bs="ts") ,
+                  data = All_data2)
 summary(model5_all)
@@
 
-The overall explaining power of the model increases with an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)`, but little changes in the significance of the variables.
+Including non-linearity increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)`, but creates few changes in the significance of the variables.
+However, it also can create a chance of overfitting, as the smoothing splines allow significant flexibility, which will be investigated below.
 
-A final full model
+However, first a final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included in the model
 
@@
 model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Area_km2, bs="ts") + s(Dryness, bs="ts" ) + 
-                    s(Latitude, bs="ts" ) +  s(length, bs="ts") +
+                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts" ) + 
+                     s(length, bs="ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
                     Hydrological_regime
-                    , data = Zhang_all2)
-summary(model6_all)
+                    , data = All_data2)
+#summary(model6_all)
 #gam.check(model6_all)
 #plot(model6_all)
+pander(round(summary(model6_all)$p.table,2))
+pander(round(summary(model6_all)$s.table,2))
+plot(model6_all, select = 3)
 ```
 
-```{r model6_reduc, eval = F}
+```{r model6_reduc, eval = F, echo = F}
 model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts") + 
-                      s(Area_km2, bs="ts") +  s(length, bs="ts") +
+                    s(Dryness, bs="ts" )  + 
+                      s(log10(Area_km2), bs="ts") +  length +
                     Assessment_technique +
@@
 ```{r, eval =T, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Assessment_technique, Latitude)) + 
+All_data2 %>%
+  ggplot(aes(Assessment_technique, length)) + 
   geom_boxplot(colour = "blue") + theme_bw()
@@
 
-```{r, eval =T, echo =F}
+```{r, eval =F, echo =F}
 Zhang_all2 %>%
@@
 Clearly all have at least some relationship with Latitude, therefore are being masked if Latitude is included in the model.
-
-<!-- Take out Latitude -->
-```{r model7_noLat}
-model7_noLat <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + 
-                      s(Area_km2, bs="ts")  + s(length, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = Zhang_all2)
-summary(model7_noLat)
-plot(model7_noLat)
-```
-
-```{r model7_noLat_intLength}
-model7_noLat_intlength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="tp" ) + 
-                      s(Area_km2, bs="tp")  + s(length, bs = "tp") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, 
-                    data = Zhang_all2)
-summary(model7_noLat_intlength)
-plot(model7_noLat_intlength)
-```
 
@@
                     s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(Area_km2, bs="ts") +
+                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = Zhang_all2 %>% filter(Dryness < 4))
+                    Hydrological_regime, data = All_data2 %>% filter(length< 60))
 summary(model7_noLatb)

### 2021-12-02 paper/Forest_and_Water/Forest_and_Water.Rmd
updates paper 03122021

@@
 The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response.
-This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackon2005).
+This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005).
 In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.
@@
 
+```{r, eval =T, echo =F}
+All_data2 %>%
+  group_by(Assessment_technique) %>%
+  summarise(n = n())
+```
+
+Maybe remove data for low numbers of assessment techniques and rerun the analysis (<10)
+
+```{r, eval =T, echo =F}
+All_data2 %>%
+  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") + theme_bw()
+```
+
 ```{r, eval =F, echo =F}
@@
 ```{r, eval =T, echo =F}
-Zhang_all2 %>%
+All_data2 %>%
   ggplot(aes(DeltaQf_perc, length)) + 
@@
 
-```{r, echo = F, eval =F}
-Zhang_all2 %>%
-  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Dryness)) + geom_point()
-```
-
-```{r, echo = F}
-Zhang_all2 %>%
-  ggplot(aes(Hydrological_regime,Forest_type)) + geom_count() + theme_bw()
-```
-
-```{r, eval =T, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Assessment_technique,log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") +
-  theme_bw() + xlab("assessment technique") + ylab("log10(Area in km2)")
-```
-
-```{r, echo = F}
-Zhang_all2 %>%
-  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), shape = Forest_Sign)) + geom_point()
-```
 ```{r, echo = F}
 png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
-Zhang_all2 %>%
+All_data2 %>%
   ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
@@
 
-As shown earlier, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analsysi (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+As shown earlier, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 
-There are further confouding factors in the data, which were also classified by @filoso2017 and thesse create biases in the data set that can impact the overall assessment.
+There are further confouding factors in the data, which were also classified by @filoso2017 and these create biases in the data set that can impact the overall assessment.
 For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).

### 2021-12-02 paper/Forest_and_Water/Forest_and_Water.tex
updates paper 03122021

@@
 for the center of the watershed as an approximation of its spatial
-location. Using this information reference evapotranspiration (E\_\{0\})
-was extracted from the Global Aridity Index and Potential
-Evapo-Transpiration (ET0) Climate Databasev2 (Trabucco and Zomer, 2018),
-if a value of E0 was not available from the original papers. For large
-watersheds, this value, similar to annual average rainfall, is only an
-approximation of the climate at the location.
+location. Using this information reference evapotranspiration
+(E\textsubscript{0}) was extracted from the Global Aridity Index and
+Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2
+(Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not
+available from the original papers. For large watersheds, this value,
+similar to annual average rainfall, is only an approximation of the
+climate at the location.
 
@@
 
-Similar to Zhang et al. (2017), the ``dryness index'' was calculated as:
-\[\tag{1}
+Similar to Zhang et al. (2017), the ``dryness index'' was calculated
+from the reference evapotranspiration and the annual average rainfall
+(Pa) as: \[\tag{1}
 \begin{aligned}
-D = \frac{E0}{Pa}
+D = \frac{E_{0}}{Pa}
 \end{aligned}\]
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all <-}\StringTok{ }\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{filter}\NormalTok{(}\StringTok{`}\DataTypeTok{Watershed #}\StringTok{`} \OperatorTok{>}\StringTok{ }\DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all <-}\StringTok{ }\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{DeltaQf_perc =} \KeywordTok{ifelse}\NormalTok{(}\StringTok{`}\DataTypeTok{Watershed #}\StringTok{`} \OperatorTok{==}\StringTok{ }\DecValTok{76}\NormalTok{,}\DecValTok{157}\NormalTok{,DeltaQf_perc))}
-\end{Highlighting}
-\end{Shaded}
+Removing possible duplicates removes 29 watersheds from the overall data
+set
 
@@
 This means that a smooth function \(s()\) is applied to the \(Z_i\)
-variables.
+variables. For the smoothing function we applied thin plate regression
+splines with an additional shrinkage penalty which means the terms can
+be shrunk to 0 if not significant (Wood, 2006).
 
@@
 
-The overall dataset contains 311 observations of changes in flow. The
+The overall dataset contains 312 observations of changes in flow. The
 overall distribution of changes in flow is highly skewed as is the
@@
 \begin{longtable}[]{@{}ccccc@{}}
+\caption{Summary results of the first regression model predicting change
+in streamflow from change in forest cover and accounting for the
+direction of the change}\tabularnewline
 \toprule
@@
 \midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
 \endhead
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-8.77\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.52\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.59\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.11\strut
+8.65\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.56\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.56\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.12\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.5\strut
+0.45\strut
 \end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.77\strut
+5.26\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--30.9\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.86\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.27\strut
+-29.17\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.79\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.04\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 
+\begin{longtable}[]{@{}ccccc@{}}
+\caption{Summary results of the first regression model predicting change
+in streamflow from change in forest cover and accounting for the
+direction of the change including the new data sets}\tabularnewline
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+9.43\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.54\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.7\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.09\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.44\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-36.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.59\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.53\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
 While the overall variance explained in this model is not high with an
-adjusted \emph{r\textsuperscript{2}} of 0.19, it clearly supports the
+adjusted \emph{r\textsuperscript{2}} of 0.18, it clearly supports the
 hypothesized relationship between the change in forest cover and the
 change in flow. The model suggests that for every 1\% change in forest
-cover, on the average, the flow changes 0.5\%. However the change in
+cover, on the average, the flow changes 0.45\%. However the change in
 flow is different for forest cover decreases compared to forest cover
-increases. In fact, forest cover increases decrease flow by 31\% less
+increases. In fact, forest cover increases decrease flow by 29\% less
 than a similar decrease in forest cover causes flow to increase. So
 roughly speaking, a 1\% forest cover increase on the average decreases
-flow by \((1 - 0.31)*0.5\%\), while a the percentage forest cover
-decrease will increase flow by 0.5\%.
-
-It is however it is clear from the lack of explaining power, that there
-could be confounding factors, as alluded to in the methods. The obvious
-ones being watershed dryness and area (following Zhang et al. (2017)):
+flow by \((1 - 0.29)*0.45\%\), while a the percentage forest cover
+decrease will increase flow by 0.45\%.
+
+Checking this result against including the new data indicates that this
+mainly strengthens the difference between the forest cover increases and
+decreases, and decreases the mean decrease in flow as a result of forest
+cover change
+
+It is however it is clear from the lack of explaining power for the
+model, that there could be confounding factors, as alluded to in the
+methods. The obvious ones being watershed dryness and area (following
+Zhang et al. (2017)):
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
-\end{aligned}\] Where \(Pa_mm\) is the annual average rainfall in mm.
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
+\end{aligned}\]
+
+Where \(Pa_mm\) is the annual average rainfall in mm, and a log base 10
+transformation is applied to the area variable given that the
+distribution of Area is higly skewed (Figure 1)
 
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-18.94\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.78\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.43\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.02\strut
+22.55\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+9.16\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.46\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.5\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.66\strut
+0.34\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.26\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--31.54\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.9\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.35\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.3\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.77\strut
+-35.52\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.67\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.27\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{log10(Area\_km2)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.12\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+1.72\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.81\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.07\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1.75\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.08\strut
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.32\strut
 \end{minipage}\tabularnewline
@@
 
-Including area and annual precipitation does not really improve the
-overall explaining power of the model, in fact, annual precipitation
-appears to be only a very small confounding factor, representing only a
--0.01/\% partial effect in the change in streamflow, holding all other
-factors constant. In contrast to earlier reported studies (Filoso et
-al., 2017; Zhang et al., 2017), watershed area has no effect on the
+Including area and annual precipitation slightly improves the overall
+explaining power of the model. Annual precipitation is in fact not
+significant. Relative to earlier reported studies (Filoso et al., 2017;
+Zhang et al., 2017), the log base 10 transformed watershed area
+indicates a p-value of only 0.07, suggesting a marginal impact on the
 change in stream flow. This supports our approach (in contrast to Zhang
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
+\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}\]
@@
 \begin{longtable}[]{@{}ccccc@{}}
-\caption{Results of the model including Latitude and
-Longitude}\tabularnewline
+\caption{Results of the model including Latitude and Longitude including
+new data}\tabularnewline
 \toprule
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-29.99\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-8.65\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.47\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
+23.29\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+9.96\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.34\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.47\strut
+0.35\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+3.32\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-37.21\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.2\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{log10(Area\_km2)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.28\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+1.73\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.89\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.06\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Pa\_mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.83\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.41\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.05\strut
 \end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.4\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--37.11\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.09\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.59\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.55\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.17\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.55\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.58\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Longitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
 0.03\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.29\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.74\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.03\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.28\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.78\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.23\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.82\strut
 \end{minipage}\tabularnewline
@@
 
-This linear model shows that there is a significant gradient in the
-Latitude and with annual average rainfall, with watersheds closer to the
-equator having lower changes in the runoff compare to watersheds further
-away from the equator. This suggests an influence of radiation, which
-will be tested next. In addition, the model suggests an influence of the
-annual average rainfall, with wetter watersheds having slightly lower
-changes in runoff. The total explaining power of the model is still low
-with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further
-factors that are currently not included in the model.
-
-There is no relationship with Longitude, suggesting that the watersheds
-across different continents do not show an East-West direction trend.
+There appears to be no significant latitudinal gradient. The total
+explaining power of the model is still low with an adjusted
+\emph{r\textsuperscript{2}} of 0.22 suggesting further factors that are
+currently not included in the model.
+
+Note that in this case the significance of the Area variable increases,
+and generally indicates that larger watersheds would be expected to have
+a lower change in streamflow as also indicated in Zhang et al. (2017).
 
@@
 
-Latitude might be indicating an influence of radiation on
-evapotranspiration, and most likely related to the dryness index, as
-also indicated in Zhang et al. (2017). Increased evapotranspiration
-could lead to drier watersheds, unless balanced by rainfall (such as
-possibly in the tropics). The North-South trend in dryness would be
-related to Latitude. This model introduces the dryness index as a linear
-variable and drops the annual average precipitation as a variable, as
-dryness is calculated from.
+Climate, and in particular evapotranspiration would be expected to have
+a significant effect on the streamflow change as represented by the
+dryness index, which is also indicated in Zhang et al. (2017). Increased
+evapotranspiration could lead to drier watersheds, unless balanced by
+rainfall (such as possibly in the tropics). This model introduces the
+dryness index as a linear variable and drops the annual average
+precipitation as a variable, as dryness is calculated from the
+precipitation. It also drops the Latitude and Longitude as these are
+indicated not to be significant.
 
 \begin{longtable}[]{@{}ccccc@{}}
+\caption{Results of the model replacing the annual precipitation with
+the dryness index}\tabularnewline
 \toprule
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-10.27\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.44\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.38\strut
+10.46\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.66\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.36\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.46\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.17\strut
+0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.93\strut
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--37.56\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.19\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.07\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Area\_km2}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.76\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.45\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.28\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.66\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+-35.67\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.73\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.23\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{log10(Area\_km2)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+1.76\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.99\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.05\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Dryness}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+6.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+3.08\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
 0.03\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.4\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.69\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Dryness}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-6.1\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-3.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.97\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.05\strut
 \end{minipage}\tabularnewline
@@
 of the Dryness values are very large (\textgreater{} 4) and these values
-have high leverage in the data. These watersheds are listed in Table XX:
-
-\begin{longtable}[]{@{}ccc@{}}
-\toprule
-\begin{minipage}[b]{0.14\columnwidth}\centering
-Latitude\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
-Longitude\strut
-\end{minipage} & \begin{minipage}[b]{0.42\columnwidth}\centering
-Watershed name\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.14\columnwidth}\centering
-34.67\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--111.7\strut
-\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
-Beaver Creek, AZ \#3-2\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.14\columnwidth}\centering
-36.4\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--120.4\strut
-\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
-Cantua\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.14\columnwidth}\centering
-34.43\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--112.3\strut
-\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
-White Spar, Ariz., U.S.A, B\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.14\columnwidth}\centering
-32.74\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
--111.5\strut
-\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
-Natural DRDages, Ariz., U.S.A, A\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
+have high leverage in the data. These watersheds are listed in Table 5:
+
+Quitting from lines 461-462 (Forest\_and\_Water.Rmd) Error in filter(.,
+Dryness \textgreater{} 4) : object `Zhang\_all2' not found Calls:
+\ldots{} withVisible -\textgreater{} eval -\textgreater{} eval
+-\textgreater{} pander -\textgreater{} \%\textgreater\% -\textgreater{}
+select -\textgreater{} filter In addition: Warning messages: 1: In
+eval(expr, envir, enclos) : NAs introduced by coercion 2: In eval(expr,
+envir, enclos) : NAs introduced by coercion
+
+\hypertarget{are-some-of-the-variables-possibly-non-linear}{%
+\subsubsection{Are some of the variables possibly
+non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
+
+The work by Filoso et al. (2017) and earlier by Jackson et al. (2005)
+has indicated that the length of the study might influence the response.
+This links to the idea from Kuczera (1987) that the effect of logging or
+deforestation or reforestation reduces with the length of time post
+intervention (see also Jackson et al. (2005)). In addition to adding the
+length as a variable, all continuous variables (Dryness, Area, length
+and Latitude) were considered non-linear in this model and as indicated
+a shrinkage smoothing spline (Wood, 2006) was applied to these
+variables.
 
@@
 ## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2, bs = "ts") + 
-##     s(Dryness, bs = "ts") + s(length, bs = "ts") + Latitude
+## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(log10(Area_km2), 
+##     bs = "ts") + s(Dryness, bs = "ts") + s(length, bs = "ts")
 ## 
@@
 ##                     Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)          18.85584    5.84673   3.225 0.001419 ** 
-## DeltaF_perc_pos       0.37334    0.09042   4.129 4.90e-05 ***
-## Forest_Signincrease -33.25371    6.40370  -5.193 4.15e-07 ***
-## Latitude             -0.29544    0.08775  -3.367 0.000873 ***
+## (Intercept)          17.0942     6.0776   2.813  0.00526 ** 
+## DeltaF_perc_pos       0.2606     0.1067   2.442  0.01521 *  
+## Forest_Signincrease -34.5961     6.1052  -5.667 3.59e-08 ***
 ## ---
@@
 ##                       edf Ref.df     F p-value   
-## s(Area_km2) 3.174e-06      9 0.000  0.3914  
-## s(Dryness)  1.293e-05      9 0.000  0.5926  
-## s(length)   5.991e+00      9 1.221  0.0779 .
+## s(log10(Area_km2)) 1.7399      9 0.782 0.01465 * 
+## s(Dryness)         0.8731      9 0.570 0.01514 * 
+## s(length)          8.8276      9 3.097 0.00114 **
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.227   Deviance explained = 25.2%
-## GCV =   2018  Scale est. = 1944.1    n = 273
+## R-sq.(adj) =  0.287   Deviance explained =   32%
+## GCV = 2171.2  Scale est. = 2065.7    n = 297
 \end{verbatim}
 
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+17.09\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.08\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.81\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.26\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.44\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-34.6\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.67\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.74\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.78\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.87\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.57\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+8.83\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+Including non-linearity increases the overall explaining power of the
+model to an adjusted \emph{r\textsuperscript{2}} of 0.29, but creates
+few changes in the significance of the variables. However, it also can
+create a chance of overfitting, as the smoothing splines allow
+significant flexibility, which will be investigated below.
+
+However, first a final full model which includes the remaining
+categorical variables (Precipitation data type, Assessment technique,
+Forest type and Hydrological regime) are included in the model
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.36\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-27.3\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+19.84\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-1.38\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.17\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.01\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-21.54\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+7.29\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-2.96\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeOB}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-8.13\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+15.17\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.54\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.59\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeSG}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+17.33\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+17.29\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.32\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueEA, HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+13.93\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+46.74\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.3\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.77\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueHM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+32.96\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+13.92\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.37\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniquePWE}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+53.24\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.81\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+4.16\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniquePWE, HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+33.54\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+47.07\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.71\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.48\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+39.8\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+22.06\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.8\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW, EA}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+45.97\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+26.25\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.75\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.08\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueSH}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+44.89\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+13.49\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.33\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeCF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-2.2\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.52\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.26\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.8\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeMF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.56\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.96\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.73\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.46\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Hydrological\_regimeSD}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+4.59\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+10.25\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.45\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.65\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.07\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.16\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.45\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.39\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+8.73\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.04\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model6_all-1.pdf}
+
+Clearly Latitude is masking other factors including the assessment
+technique and the forest type
+
 \begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2, bs = "ts") + 
-##     s(Dryness, bs = "ts") + s(Latitude, bs = "ts") + s(length, 
-##     bs = "ts") + Precip_data_type + Assessment_technique + Forest_type + 
-##     Hydrological_regime
-## 
-## Parametric coefficients:
-##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                   2.5565    25.5148   0.100 0.920272    
-## DeltaF_perc_pos               0.3234     0.1058   3.057 0.002482 ** 
-## Forest_Signincrease         -27.2070     7.5951  -3.582 0.000411 ***
-## Precip_data_typeOB            2.9296    15.9005   0.184 0.853974    
-## Precip_data_typeSG           23.9270    18.5848   1.287 0.199159    
-## Assessment_techniqueEA, HM   14.8215    46.1143   0.321 0.748176    
-## Assessment_techniqueHM       -7.5091    18.4252  -0.408 0.683967    
-## Assessment_techniquePWE       8.9562    19.7884   0.453 0.651240    
-## Assessment_techniquePWE, HM  19.6741    52.6286   0.374 0.708856    
-## Assessment_techniqueQPW      -7.3110    28.3315  -0.258 0.796583    
-## Assessment_techniqueSH        2.8647    18.4698   0.155 0.876871    
-## Forest_typeCF                 4.9185     9.2133   0.534 0.593935    
-## Forest_typeMF               -10.6875     9.6354  -1.109 0.268444    
-## Hydrological_regimeSD         6.9324    10.5522   0.657 0.511825    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                   edf Ref.df     F p-value   
-## s(Area_km2) 1.754e-07      9 0.000 0.50299   
-## s(Dryness)  1.784e-06      9 0.000 0.87130   
-## s(Latitude) 3.607e+00      9 1.648 0.00213 **
-## s(length)   5.815e+00      9 1.067 0.11589   
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.231   Deviance explained = 29.6%
-## GCV = 2158.9  Scale est. = 1969.5    n = 267
+## Warning: Removed 5 rows containing non-finite values (stat_boxplot).
 \end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
+
+\begin{verbatim}
+## # A tibble: 8 x 2
+##   Assessment_technique     n
+##   <chr>                <int>
+## 1 EA                      32
+## 2 EA, HM                   1
+## 3 HM                      53
+## 4 PWE                    185
+## 5 PWE, HM                  1
+## 6 QPW                      7
+## 7 QPW, EA                  4
+## 8 SH                      26
+\end{verbatim}
+
+Maybe remove data for low numbers of assessment techniques and rerun the
+analysis (\textless10)
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-22-1.pdf}
+
+Clearly all have at least some relationship with Latitude, therefore are
+being masked if Latitude is included in the model.
+
+The flexible nature of the splines and the shrinkage applied to the
+splines means that the length variable can capture variation in the
+data, but it is unclear what exactly is captured. The shape of the
+resulting conditional response does not reflect the reponses highlighted
+in Filoso et al. (2017) and Jackson et al. (2005). Reducing the
+flexibility of the splines, or fitting a linear term results in
+``length'' not being significant, which suggests that this variable is
+not helpful in explaining the variation in the data.
+
+\begin{verbatim}
+## Warning: Removed 5 rows containing missing values (geom_point).
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-25-1.pdf}
+
+In drier watersheds, changes in forest cover have greater impact on
+flow, which is similar to Zhang et al. (2017). This is most likely
+because in these watersheds the overall flow is surface flow dominated
+and therefore the buffering that is afforded by the groundwater inputs
+is not as great. As we don't have a separate variable for groundwater
+inputs (although this effect is estimated in many studies), we cannot
+analyse this effect separately.
+
+In contrast to Filoso et al. (2017), we also did not identify an effect
+of the
+
+Given how skewed Dryness is due tot he few watersheds that have very
+high dryness values, it is worth investigating what excluding these 4
+watersheds from the data means for the relationships.
 
@@
 \begin{Highlighting}[]
-\NormalTok{model6_reduc <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
-\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(Latitude, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
-\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{  }\KeywordTok{s}\NormalTok{(length, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Assessment_technique }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2)}
-\KeywordTok{summary}\NormalTok{(model6_reduc)}
+\NormalTok{model7_noLatb <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
+\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\CommentTok{#s(Latitude, bs="ts") + }
+\StringTok{                      }\KeywordTok{s}\NormalTok{(}\KeywordTok{log10}\NormalTok{(Area_km2), }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(length,}\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
+\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ All_data2 }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(length}\OperatorTok{<}\StringTok{ }\DecValTok{60}\NormalTok{))}
+\KeywordTok{summary}\NormalTok{(model7_noLatb)}
 \end{Highlighting}
@@
 ## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(Latitude, bs = "ts") + s(Area_km2, bs = "ts") + s(length, 
-##     bs = "ts") + Assessment_technique + Hydrological_regime
+##     s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + Precip_data_type + 
+##     Assessment_technique + Forest_type + Hydrological_regime
 ## 
@@
 ##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                  11.5706    17.3998   0.665  0.50666    
-## DeltaF_perc_pos               0.3094     0.1038   2.979  0.00317 ** 
-## Forest_Signincrease         -29.9159     7.1426  -4.188 3.88e-05 ***
-## Assessment_techniqueEA, HM   11.4108    45.6037   0.250  0.80262    
-## Assessment_techniqueHM       -4.6254    17.2047  -0.269  0.78827    
-## Assessment_techniquePWE       3.8045    18.5150   0.205  0.83736    
-## Assessment_techniquePWE, HM  17.3096    50.9143   0.340  0.73416    
-## Assessment_techniqueQPW     -14.1255    26.9676  -0.524  0.60088    
-## Assessment_techniqueSH       -0.6693    17.6262  -0.038  0.96974    
-## Hydrological_regimeSD        12.3022     8.4529   1.455  0.14680    
+## (Intercept)                 -12.5901    18.2493  -0.690 0.490850    
+## DeltaF_perc_pos               0.2764     0.1021   2.708 0.007197 ** 
+## Forest_Signincrease         -23.5034     6.9108  -3.401 0.000772 ***
+## Precip_data_typeOB          -11.8739    14.1588  -0.839 0.402419    
+## Precip_data_typeSG           16.9396    16.3313   1.037 0.300543    
+## Assessment_techniqueEA, HM   -8.2703     3.5091  -2.357 0.019143 *  
+## Assessment_techniqueHM       28.0181    12.3042   2.277 0.023557 *  
+## Assessment_techniquePWE      39.2429    12.0840   3.247 0.001310 ** 
+## Assessment_techniquePWE, HM -14.4779     5.5623  -2.603 0.009753 ** 
+## Assessment_techniqueQPW      33.8032    20.7827   1.627 0.105002    
+## Assessment_techniqueQPW, EA  35.5789    25.2225   1.411 0.159506    
+## Assessment_techniqueSH       42.8100    12.7078   3.369 0.000864 ***
+## Forest_typeCF                -3.8554     8.1687  -0.472 0.637324    
+## Forest_typeMF                -4.0151     8.6538  -0.464 0.643047    
+## Hydrological_regimeSD         6.4841    10.0569   0.645 0.519641    
 ## ---
@@
 ##                          edf Ref.df     F p-value  
-## s(Dryness)  1.026e-05      9 0.000 0.885487    
-## s(Latitude) 3.300e+00      9 1.829 0.000683 ***
-## s(Area_km2) 2.280e-06      9 0.000 0.848293    
-## s(length)   5.575e+00      9 0.901 0.176727    
+## s(Dryness)         3.586e+00      9 1.665  0.0132 *
+## s(log10(Area_km2)) 7.053e-01      9 0.293  0.0510 .
+## s(length)          2.302e-10      9 0.000  0.9953  
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.228   Deviance explained = 27.9%
-## GCV = 2085.2  Scale est. = 1941.1    n = 273
-\end{verbatim}
-
-Clearly Latitude is masking other factors including the assessment
-technique and the forest type
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-18-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-19-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
-
-Clearly all have at least some relationship with Latitude, therefore are
-being masked if Latitude is included in the model.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{model7_noLat <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
-\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }
-\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{)  }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2)}
-\KeywordTok{summary}\NormalTok{(model7_noLat)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(Area_km2, bs = "ts") + Precip_data_type + Assessment_technique + 
-##     Forest_type + Hydrological_regime
-## 
-## Parametric coefficients:
-##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -12.5291    22.7513  -0.551 0.582282    
-## DeltaF_perc_pos               0.4348     0.1033   4.209 3.46e-05 ***
-## Forest_Signincrease         -27.2295     7.6821  -3.545 0.000461 ***
-## Precip_data_typeOB          -12.4249    15.4195  -0.806 0.421049    
-## Precip_data_typeSG           11.5676    17.1018   0.676 0.499348    
-## Assessment_techniqueEA, HM   19.1901    50.7190   0.378 0.705450    
-## Assessment_techniqueHM       21.6621    17.2625   1.255 0.210578    
-## Assessment_techniquePWE      42.2758    17.2104   2.456 0.014642 *  
-## Assessment_techniquePWE, HM  23.5495    52.6315   0.447 0.654903    
-## Assessment_techniqueQPW      22.1990    27.1887   0.816 0.414924    
-## Assessment_techniqueQPW, EA  41.0844    30.1936   1.361 0.174707    
-## Assessment_techniqueSH       38.3711    18.7848   2.043 0.042025 *  
-## Forest_typeCF                -3.7900     9.7088  -0.390 0.696562    
-## Forest_typeMF               -16.3363     9.3408  -1.749 0.081404 .  
-## Hydrological_regimeSD        -1.1868    10.9105  -0.109 0.913460    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                   edf Ref.df     F p-value  
-## s(Dryness)  3.995e+00      9 1.162   0.024 *
-## s(Area_km2) 2.228e-07      9 0.000   0.439  
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.202   Deviance explained =   25%
-## GCV =   2510  Scale est. = 2350      n = 298
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{plot}\NormalTok{(model7_noLat)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model7_noLat-1.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model7_noLat-2.pdf}
-
-\begin{verbatim}
-## Warning: Removed 32 rows containing missing values (geom_point).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-21-1.pdf}
-
-In drier watersheds, changes in forest cover have greater impact on
-flow, which is similar to Zhang et al. (2017). This is most likely
-because in these watersheds the overall flow is surface flow dominated
-and therefore the buffering that is afforded by the groundwater inputs
-is not as great. As we don't have a separate variable for groundwater
-inputs (although this effect is estimated in many studies), we cannot
-analyse this effect separately.
-
-In contrast to Filoso et al. (2017), we also did not identify an effect
-of the
-
-Given how skewed Dryness is due tot he few watersheds that have very
-high dryness values, it is worth investigating what excluding these 4
-watersheds from the data means for the relationships.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{model7_noLatb <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
-\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\CommentTok{#s(Latitude, bs="ts") + }
-\StringTok{                      }\KeywordTok{s}\NormalTok{(Area_km2, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ Zhang_all2 }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(Dryness }\OperatorTok{<}\StringTok{ }\DecValTok{4}\NormalTok{))}
-\KeywordTok{summary}\NormalTok{(model7_noLatb)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(Area_km2, bs = "ts") + Precip_data_type + Assessment_technique + 
-##     Forest_type + Hydrological_regime
-## 
-## Parametric coefficients:
-##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -16.8716    22.8324  -0.739 0.460577    
-## DeltaF_perc_pos               0.4205     0.1036   4.060 6.41e-05 ***
-## Forest_Signincrease         -27.4323     7.6284  -3.596 0.000383 ***
-## Precip_data_typeOB          -12.8820    15.2846  -0.843 0.400065    
-## Precip_data_typeSG           11.9340    16.9606   0.704 0.482257    
-## Assessment_techniqueEA, HM   25.8163    50.4996   0.511 0.609607    
-## Assessment_techniqueHM       26.4681    17.4660   1.515 0.130815    
-## Assessment_techniquePWE      48.3282    17.6030   2.745 0.006440 ** 
-## Assessment_techniquePWE, HM  29.5466    52.3323   0.565 0.572808    
-## Assessment_techniqueQPW      29.1804    27.2233   1.072 0.284707    
-## Assessment_techniqueQPW, EA  48.6163    30.1921   1.610 0.108492    
-## Assessment_techniqueSH       43.7353    18.9899   2.303 0.022019 *  
-## Forest_typeCF                -4.2723     9.6304  -0.444 0.657661    
-## Forest_typeMF               -14.2554     9.4382  -1.510 0.132089    
-## Hydrological_regimeSD        -2.6825    10.8541  -0.247 0.804980    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                   edf Ref.df     F p-value   
-## s(Dryness)  3.374e+00      9 1.576 0.00703 **
-## s(Area_km2) 3.036e-09      9 0.000 0.41938   
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.209   Deviance explained = 25.6%
-## GCV = 2462.7  Scale est. = 2308.7    n = 294
+## Rank: 39/42
+## R-sq.(adj) =  0.273   Deviance explained = 31.4%
+## GCV = 2119.6  Scale est. = 1992.7    n = 289
 \end{verbatim}
@@
 \includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-2.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-23-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-24-1.pdf}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-25-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-3.pdf}
 
@@
 elasticity analysis (EA), Hydrological modelling (HM) or a combined use
-of statistical methods (SH) or quasi paired watershed analsysi (QPW),
+of statistical methods (SH) or quasi paired watershed analysis (QPW),
 thus all providing an approximation of the effect of forestry on
@@
 There are further confouding factors in the data, which were also
-classified by Filoso et al. (2017) and thesse create biases in the data
+classified by Filoso et al. (2017) and these create biases in the data
 set that can impact the overall assessment. For example, snow dominated
@@
 
+\leavevmode\hypertarget{ref-kuczera1987}{}%
+Kuczera, G., 1987. Prediction of water yield reductions following a
+bushfire in ash-mixed species eucalypt forest. Journal of Hydrology 94,
+215--236.
+doi:\href{https://doi.org/Doi:\%2010.1016/0022-1694(87)90054-0}{Doi: 10.1016/0022-1694(87)90054-0}
+
 \leavevmode\hypertarget{ref-pena-arancibia2012}{}%

### 2022-01-01 paper/Forest_and_Water/Forest_and_Water.Rmd
final changes old laptop

@@
     email: willem.vervoort@sydney.edu.au
-    affiliation: The University of Sydney, INIA
+    affiliation: DARE, The University of Sydney
     footnote: 1
@@
 address:
+  - code: DARE
+    address: ARC Training Centre Data Analytics for Resources and the Environment
   - code: The University of Sydney
@@
   - code: INIA
-    address: Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
+    address: Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
   - code: IMFIA
@@
       "Pa_mm" = "Annual Rainfall (mm)"))) +   
-  theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)),
+  theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5)),
         axis.text.x = element_text(angle = 45, hjust = 1))
@@
   facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
-  theme_bw() + scale_colour_manual(name = "direction of forest change", values = c("increase" = "blue", "decrease" = "red")) +
+  theme_bw() + 
+    theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        legend.text = element_text(size = rel(1.2)),
+        legend.title = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5))) + 
+  scale_colour_manual(name = "Forest change direction", 
+                      values = c("increase" = "blue", "decrease" = "red")) +
   xlab("Change in forest cover %") + ylab("Change in flow %")
@@
 ```{r, echo = F}
-pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "Watersheds for which the dryness inddex > 4")
+pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "Watersheds for which the dryness index > 4")
 ```
@@
 pander(round(summary(model6_all)$s.table,2))
-plot(model6_all, select = 3)
+plot(model6_all, select = 2)
 ```
@@
 
-```{r, eval =T, echo =F}
+```{r, eval =F, echo =F}
+png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
 All_data2 %>%
   ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw()
+  geom_boxplot(colour = "blue") + theme_bw() +
+    theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)))
+dev.off()
+```
+
+```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F}
+
+include_graphics("AssessmentTechnique_byArea.png")
 ```
@@
 
+\newpage
+
 <!--reduce drynes < 4 -->
-```{r model7_noLatb}
+```{r model7_noLatb, echo = F}
 model7_noLatb <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
                     Hydrological_regime, data = All_data2 %>% filter(length< 60))
+#summary(model7_noLatb)
+#plot(model7_noLatb)
+pander(round(summary(model7_noLatb)$p.table,2))
+pander(round(summary(model7_noLatb)$s.table,2))
+plot(model7_noLatb)
 summary(model7_noLatb)
-plot(model7_noLatb)
+```
+
+### remove the assessment techniques with very small numbers
+
+```{r model8_reduc_assess, echo = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
+
+model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
+#summary(model7_noLatb)
+#plot(model7_noLatb)
+pander(round(summary(model8_assess)$p.table,2))
+pander(round(summary(model8_assess)$s.table,2))
+plot(model8_assess)
+summary(model8_assess)
+
 ```
@@
         axis.text = element_text(size = rel(1.5))) +
-  xlab("% change in forestry") + #scale_y_log10() +
+  xlab("% change in forest cover") + #scale_y_log10() +
   scale_size_continuous(name = "log10(Area in km2)") +
@@
 
-There are further confouding factors in the data, which were also classified by @filoso2017 and these create biases in the data set that can impact the overall assessment.
+There are further confounding factors in the data, which were also classified by @filoso2017 and these create biases in the data set that can impact the overall assessment.
 For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).

### 2022-01-01 paper/Forest_and_Water/Forest_and_Water.tex
final changes old laptop

@@
 Re-analysing a global data set.}
-    \author[The University of Sydney, INIA]{R. Willem Vervoort\corref{1}}
+    \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
    \ead{willem.vervoort@sydney.edu.au} 
@@
    \ead{jalonso@fing.edu.uy} 
+      \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
     \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney,
 Sydney, NSW 2006, Australia}
-    \address[INIA]{Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48
-km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \address[INIA]{Project Manager, FPTA 357 Instituto Nacional de Investigación
+Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100
+Canelones, Uruguay}
     \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
@@
 
-Quitting from lines 461-462 (Forest\_and\_Water.Rmd) Error in filter(.,
+Quitting from lines 471-472 (Forest\_and\_Water.Rmd) Error in filter(.,
 Dryness \textgreater{} 4) : object `Zhang\_all2' not found Calls:
@@
 
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-22-1.pdf}
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments}\label{fig:assessment}
+\end{figure}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{model7_noLatb <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(DeltaQf_perc }\OperatorTok{~}\StringTok{ }\NormalTok{DeltaF_perc_pos }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_Sign }\OperatorTok{+}\StringTok{ }
-\StringTok{                    }\KeywordTok{s}\NormalTok{(Dryness, }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\OperatorTok{+}\StringTok{ }\CommentTok{#s(Latitude, bs="ts") + }
-\StringTok{                      }\KeywordTok{s}\NormalTok{(}\KeywordTok{log10}\NormalTok{(Area_km2), }\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}\StringTok{ }\KeywordTok{s}\NormalTok{(length,}\DataTypeTok{bs=}\StringTok{"ts"}\NormalTok{) }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Precip_data_type }\OperatorTok{+}\StringTok{  }\NormalTok{Assessment_technique }\OperatorTok{+}\StringTok{ }\NormalTok{Forest_type }\OperatorTok{+}
-\StringTok{                    }\NormalTok{Hydrological_regime, }\DataTypeTok{data =}\NormalTok{ All_data2 }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(length}\OperatorTok{<}\StringTok{ }\DecValTok{60}\NormalTok{))}
-\KeywordTok{summary}\NormalTok{(model7_noLatb)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Dryness, bs = "ts") + 
-##     s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + Precip_data_type + 
-##     Assessment_technique + Forest_type + Hydrological_regime
-## 
-## Parametric coefficients:
-##                             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -12.5901    18.2493  -0.690 0.490850    
-## DeltaF_perc_pos               0.2764     0.1021   2.708 0.007197 ** 
-## Forest_Signincrease         -23.5034     6.9108  -3.401 0.000772 ***
-## Precip_data_typeOB          -11.8739    14.1588  -0.839 0.402419    
-## Precip_data_typeSG           16.9396    16.3313   1.037 0.300543    
-## Assessment_techniqueEA, HM   -8.2703     3.5091  -2.357 0.019143 *  
-## Assessment_techniqueHM       28.0181    12.3042   2.277 0.023557 *  
-## Assessment_techniquePWE      39.2429    12.0840   3.247 0.001310 ** 
-## Assessment_techniquePWE, HM -14.4779     5.5623  -2.603 0.009753 ** 
-## Assessment_techniqueQPW      33.8032    20.7827   1.627 0.105002    
-## Assessment_techniqueQPW, EA  35.5789    25.2225   1.411 0.159506    
-## Assessment_techniqueSH       42.8100    12.7078   3.369 0.000864 ***
-## Forest_typeCF                -3.8554     8.1687  -0.472 0.637324    
-## Forest_typeMF                -4.0151     8.6538  -0.464 0.643047    
-## Hydrological_regimeSD         6.4841    10.0569   0.645 0.519641    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                          edf Ref.df     F p-value  
-## s(Dryness)         3.586e+00      9 1.665  0.0132 *
-## s(log10(Area_km2)) 7.053e-01      9 0.293  0.0510 .
-## s(length)          2.302e-10      9 0.000  0.9953  
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Rank: 39/42
-## R-sq.(adj) =  0.273   Deviance explained = 31.4%
-## GCV = 2119.6  Scale est. = 1992.7    n = 289
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{plot}\NormalTok{(model7_noLatb)}
-\end{Highlighting}
-\end{Shaded}
+\newpage
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.36\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-12.59\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+18.25\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.69\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.49\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.28\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.71\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-23.5\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+6.91\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-3.4\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeOB}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-11.87\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+14.16\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.84\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeSG}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+16.94\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+16.33\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.04\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueEA, HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-8.27\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+3.51\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-2.36\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueHM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+28.02\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.3\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.28\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniquePWE}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+39.24\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.08\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.25\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniquePWE, HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-14.48\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+5.56\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-2.6\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+33.8\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+20.78\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.63\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.11\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW, EA}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+35.58\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+25.22\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.41\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.16\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueSH}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+42.81\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.71\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.37\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeCF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-3.86\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.17\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.47\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.64\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeMF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-4.02\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.65\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.46\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.64\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Hydrological\_regimeSD}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+6.48\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+10.06\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.64\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.52\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.59\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.67\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.71\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.29\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.05\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 \includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-3.pdf}
+
+\hypertarget{remove-the-assessment-techniques-with-very-small-numbers}{%
+\subsubsection{remove the assessment techniques with very small
+numbers}\label{remove-the-assessment-techniques-with-very-small-numbers}}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.34\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-18.71\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+19.02\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.98\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.33\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.29\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.83\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-23.26\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+7.15\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-3.25\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Precip\_data\_typeOB}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-11.25\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+14.19\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.79\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.43\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Precip\_data\_typeSG}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+18.43\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+16.43\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.26\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueHM}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+32.02\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+12.76\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.51\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniquePWE}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+44.48\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+12.97\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.43\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+38.12\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+21.21\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.8\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueSH}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+46.79\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+13.17\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.55\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_typeCF}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.3\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+8.23\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.4\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.69\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_typeMF}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.88\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+8.69\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.45\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.66\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Hydrological\_regimeSD}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+5.73\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+10.11\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.57\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.57\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.51\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.96\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.7\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.24\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.08\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.99\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-1.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-2.pdf}
+\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-3.pdf}
 
@@
 
-There are further confouding factors in the data, which were also
+There are further confounding factors in the data, which were also
 classified by Filoso et al. (2017) and these create biases in the data

### 2022-01-01 paper/Forest_and_Water/Forest_and_Water.Rmd
test new laptop

@@
 
-Here $X_i$ are factorial variables, while $Z_i$ are continous variables.
+Here $X_i$ are factorial variables, while $Z_i$ are continuous variables.
 The model assumes no direct interactions and all variables are additive.
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic.
+In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation.
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.

### 2022-01-10 paper/Forest_and_Water/Forest_and_Water.Rmd
adding catchments from Filoso et al 201

@@
     affiliation: INIA
-    footnote: 2
   - name: Jimena Alonso
@@
     affiliation: IMFIA
-    footnote: 2
 address:
@@
     text: "Corresponding Author"
-  - code: 2
-    text: "Equal contribution"
 abstract: |
@@
 csl: elsevier-harvard.csl
-output: rticles::elsevier_article
+output:   
+  bookdown::pdf_book:
+    base_format: rticles::elsevier_article
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
@@
 ==========================
-
-## Introduction
 
@@
 The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in % ($\Delta$F%) and the change in streamflow in % ($\Delta$Qf(%), based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
 Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
@@
 This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
+Furthermore, the change in flow for watershed #76 was corrected to 157% based on the original publication [@baker1984].
+Finally, on review of all the data in @zhang2017, 29 potential duplicates were identified and flagged in the data.
 
@@
 require(mgcv)
+require(visreg)
+require(patchwork)
 suppressWarnings(require(knitr))
@@
 
-```{r, echo =F, eval = F}
-PET_all %>%
-  ggplot(aes(`Watershed #`,E0)) + geom_bar(stat="identity", fill= "steelblue")
-```
-
-```{r, echo =F, eval = F}
-PET_all %>%
-  ggplot(aes(Longitude,Latitude, colour = E0)) + geom_point() + coord_equal()
-
-```
-
 <!--- ### change names of columns
-
 the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple --->
@@
 ```
-Removing possible duplicates removes 29 watersheds from the overall data set
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}$$
@@
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease.
-A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$.
+Therefore all the change in forest cover data is converted to positive values, and an additional column ($sign_{forest cover}$) is added that indicates whether it was a forest cover increase or decrease.
+A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Qf \%$.
 This means that a smooth function $s()$ is applied to the $Z_i$ variables.
@@
 
-The overall dataset contains `r nrow(Zhang_all)` observations of changes in flow.
+The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` observations.
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
@@
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller watersheds dominate the database with 42% of the data from watersheds < 1 km^2^ and 65% of the data for watersheds < 10 km^2^.
+Smaller watersheds dominate the database with 42\% of the data from watersheds < 1 km^2^ and 65\% of the data for watersheds < 10 km^2^.
 
@@
 
-```{r data_graphs, echo =F, fig.cap = "Overview of the distributions of some of the variables in the data set", out.width= "90%"}
-include_graphics("DataExploration.png")
+(ref:datagraphscaption) foo
+
+```{r datagraphs, echo=F, out.width="90%", fig.cap="(ref:datagraphscaption)"}
+
+include_graphics("./DataExploration.png")
 ```
@@
 
-```{r increase_decrease, echo =F, fig.cap = "Changes in flow as a function of increases and decreases in forest cover", out.width= "90%"}
+```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of increases and decreases in forest cover", out.width="90%"}
 include_graphics("Increase_decrease.png")
@@
 
-## The initial relationship between change in forest cover and streamflow
+## The general relationship between change in forest cover and streamflow
 
-Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
+Following @zhang2017, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
 \end{aligned}$$
@@
 #summary(linear_model_simple)
-#gam.check(linear_model_simple)
-
+r_sq <- round(summary(linear_model_simple_b)$r.sq,2)
 ```
@@
 ```{r, echo =F}
-pander(round(summary(linear_model_simple)$p.table,2), caption="Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
-```
-```{r, echo =F}
-pander(round(summary(linear_model_simple_b)$p.table,2), caption="Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets")
+pander(round(summary(linear_model_simple)$p.table,2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
 ```
 
-While the overall variance explained in this model is not high with an adjusted *r^2^* of `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow.
+```{r, echo =F}
+pander(round(summary(linear_model_simple_b)$p.table,2), caption="(\\#tab:tabmodel1b) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets")
+```
+
+The overall variance explained in this model is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 
-Checking this result against including the new data indicates that this mainly strengthens the difference between the forest cover increases and decreases, and decreases the mean decrease in flow as a result of forest cover change
+```{r, eval = F, echo = F}
+png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
+par(mfrow=c(2,2))
+gam.check(linear_model_simple)
+par(mfrow=c(1,1))
+dev.off()
+```
+```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
+include_graphics("residual_plot_model1.png")
+```
+
+Of importance here is to highlight the residuals of this regression.
+These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)).
+In other words, the distribution of the residuals is somewhat fat-tailed.
+We will discuss this later.
+
+Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1) and Table \@ref(tab:tabmodel1b)), and the result indication a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
 \end{aligned}$$
@@
 ```{r, echo = F}
-pander(round(summary(linear_model2)$p.table,2), caption="Summary of the second model, taking into account the annual rainfall and the area of the watershed")
+pander(round(summary(linear_model2)$p.table,2), caption="(\\#tab:out-model2) Summary of the second model, taking into account the annual rainfall and the area of the watershed")
 ```
 
-Including area and annual precipitation slightly improves the overall explaining power of the model.
+Including area and annual precipitation slightly improves the overall explaining power of the model (Table \@ref(tab:out-model2)).
 Annual precipitation is in fact not significant.
 Relative to earlier reported studies [@zhang2017; @filoso2017], the log base 10 transformed watershed area indicates a p-value of only `r round(summary(linear_model2)$p.table[4,4],2)`, suggesting a marginal impact on the change in stream flow.
-This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds The main effects remain the change in forest cover and whether this is an increase or decrease.
+This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds.
+The main effects in the model remain the change in forest cover and whether this is an increase or decrease.
 
@@
 
-```{r global_map, fig.cap="Distribution of included watersheds across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
+```{r globalmap, fig.cap="Distribution of included watersheds across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
 
@@
 As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-As the global map (\@ref(fig:global_map)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
+As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
 Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}$$
@@
 ```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2), caption= "Results of the model including Latitude and Longitude including new data")
+pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model including Latitude and Longitude including new data")
 ```
 
-There appears to be no significant latitudinal gradient.
-The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors that are currently not included in the model.
+There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the watersheds across the globe has little influence on the overall result.
+The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
-Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow as also indicated in @zhang2017.
+Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow, as also indicated in @zhang2017.
 
@@
 
-Climate, and in particular evapotranspiration would be expected to have a significant effect on the streamflow change as represented by the dryness index, which is also indicated in @zhang2017.
+Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both @zhang2017 and @jackson2005.
 Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
@@
 ```{r, echo =F}
-pander(round(summary(model4_dryness)$p.table,2), caption = "Results of the model replacing the annual precipitation with the dryness index")
+pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model replacing the annual precipitation with the dryness index")
 ```
 
-The results from this model confirm that dryness is a significant confounding factor of the change in streamflow as function of the change in forest cover change.
+The results from this model (Table \@ref(tab:out-model4)) confirm that dryness is a significant confounding factor related to the change in streamflow as a function of the change in forest cover change.
 In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is ~14\% greater.
@@
 
+```{r, echo = F}
+pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) Watersheds for which the dryness index > 4")
+```
+
 However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data.
-These watersheds are listed in Table 5:
-
-```{r, echo = F}
-pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "Watersheds for which the dryness index > 4")
-```
+These watersheds are listed in Table \@ref(tab:drytable).
 
@@
                   data = All_data2)
-summary(model5_all)
+#summary(model5_all)
 #gam.check(model5_all)
 #plot(model5_all)
-pander(round(summary(model5_all)$p.table,2))
-pander(round(summary(model5_all)$s.table,2))
 ```
+
+```{r mfive-linear, echo = F}
+pander(round(summary(model5_all)$p.table,2), caption = "Statistical summary for the linear terms in the model with non-linear terms")
+```
+
+```{r mfive-smooth, echo = F}
+pander(round(summary(model5_all)$s.table,2), caption = "Statistical summary for the smooth terms in the model with non-linear terms")
+
+```
+
 <!-- model 2c include smooth variables -->
 
-Including non-linearity increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)`, but creates few changes in the significance of the variables.
-However, it also can create a chance of overfitting, as the smoothing splines allow significant flexibility, which will be investigated below.
+Including non-linearity increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
+However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
 
-However, first a final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included in the model
+A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.
 
@@
 #plot(model6_all)
-pander(round(summary(model6_all)$p.table,2))
-pander(round(summary(model6_all)$s.table,2))
-plot(model6_all, select = 2)
 ```
 
+```{r msix-linear, echo = F}
+pander(round(summary(model6_all)$p.table,2), caption = "Statistical summary for the linear terms the full model")
+#plot(model6_all)
+```
+
+```{r msix-smooth}
+pander(round(summary(model6_all)$s.table,2), caption = "Statistical summary for the smooth terms for the full model")
+#plot(model6_all)
+```
+
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of `r round(summary(model6_all)$dev.expl,2)`.
+This indicates that the categorical variables explain a limited amount of the variance.
+However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant.
+In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
+
+```{r visualisesmooths,echo =F, eval = F}
+#require(visreg)
+png("model6_smooths.png", width = 2500, height = 1800, res = 300)
+#par(mfrow=c(2,2))
+D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
+L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
+A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
+D / (L + A) 
+# visreg(model6_all,"Dryness",gg=T) + theme_bw()
+# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
+# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
+#par(mfrow=c(1,1))
+dev.off()
+```
+
+```{r smoothsmodel6, fig.cap="Visualisation of the smooth variables in the model", out.width="90%", echo=F}
+
+include_graphics("model6_smooths.png")
+```
+
+Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area) and the change in flow is essentially linear and does not need to be smoothed (this is the value of using penalized smooths following @wood2006).
+It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
+In addition, both the length and Dryness variables show non-linearity, but this does not show a clear trend.
+
+<!--
 ```{r model6_reduc, eval = F, echo = F}
@@
 
-Clearly Latitude is masking other factors including the assessment technique and the forest type
-
-<!-- Show boxplots of relationships with latitude-->
-```{r, eval =T, echo =F}
+```{r, eval =F, echo =F}
 All_data2 %>%
@@
 
-```{r, eval =T, echo =F}
-All_data2 %>%
-  group_by(Assessment_technique) %>%
-  summarise(n = n())
-```
-
 Maybe remove data for low numbers of assessment techniques and rerun the analysis (<10)
-
-```{r, eval =F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw() +
-    theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)))
-dev.off()
-```
-
-```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
 
@@
 ```
+ -->
 
-Clearly all have at least some relationship with Latitude, therefore are being masked if Latitude is included in the model.
-
-The flexible nature of the splines and the shrinkage applied to the splines means that the length variable can capture variation in the data, but it is unclear what exactly is captured.
-The shape of the resulting conditional response does not reflect the reponses highlighted in @filoso2017 and @jackson2005.
-Reducing the flexibility of the splines, or fitting a linear term results in "length" not being significant, which suggests that this variable is not helpful in explaining the variation in the data.
-
-```{r, eval =T, echo =F}
+<!--
+```{r, eval =F, echo =F}
 All_data2 %>%
@@
 ```
+-->
 
-In drier watersheds, changes in forest cover have greater impact on flow, which is similar to @zhang2017.
-This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
-As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
-
-In contrast to @filoso2017, we also did not identify an effect of the
-
-Given how skewed Dryness is due tot he few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
-
-\newpage
-
-<!--reduce drynes < 4 -->
-```{r model7_noLatb, echo = F}
-model7_noLatb <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+<!--filter lenght < 60 -->
+```{r model7_noLatb, echo = F, eval = F}
+model7_nolongLength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                     s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
@@
 #plot(model7_noLatb)
-pander(round(summary(model7_noLatb)$p.table,2))
-pander(round(summary(model7_noLatb)$s.table,2))
-plot(model7_noLatb)
-summary(model7_noLatb)
+#pander(round(summary(model7_nolongLength)$p.table,2))
+#plot(model7_noLatb)
+#summary(model7_nolongLength)
+```
+```{r restrictlength, echo = F}
+pander(round(summary(model7_nolongLength)$s.table,2), caption = "Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
 ```
 
-### remove the assessment techniques with very small numbers
+The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
+The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005.
+One reason could be the few data points with very long data series, but highly variable responses (Figure \@ref(fig:smoothsmodel6)).
+Reducing the flexibility of the splines, removing any studies longer than 60 years (Table \@ref(tab:restrictlength)), or fitting a linear term results in "length" not being significant, suggesting that this variable is not helpful in explaining the variation in the data.
 
-```{r model8_reduc_assess, echo = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
-#summary(model7_noLatb)
-#plot(model7_noLatb)
-pander(round(summary(model8_assess)$p.table,2))
-pander(round(summary(model8_assess)$s.table,2))
-plot(model8_assess)
-summary(model8_assess)
-
-```
-
-```{r, echo = F}
+```{r, echo = F, eval =F}
 png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
@@
 
-```{r, eval =F}
+```{r, eval =F, echo =F}
 Zhang_all2 %>%
@@
 
-```{r, eval =F}
+```{r, eval =F, echo =F}
 Zhang_all %>%
@@
 
+### remove the assessment techniques with very small numbers
+
+One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).
+
+```{r tableassess, eval =T, echo =F}
+Table_assessment <- All_data2 %>%
+  group_by(Assessment_technique) %>%
+  summarise(n = n()) %>%
+  arrange(desc(n))
+
+pander(Table_assessment)
+```
+
+```{r, eval = T, echo =F}
+png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
+All_data2 %>%
+  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") + theme_bw() +
+    theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)))
+dev.off()
+```
+
+```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F}
+
+include_graphics("AssessmentTechnique_byArea.png")
+```
+
+```{r model8_reduc_assess, echo = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
+
+model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
+#summary(model7_noLatb)
+#plot(model7_noLatb)
+pander(round(summary(model8_assess)$p.table,2))
+pander(round(summary(model8_assess)$s.table,2))
+#plot(model8_assess)
+#summary(model8_assess)
+
+```
+
+In drier watersheds, changes in forest cover have greater impact on flow, which is similar to @zhang2017.
+This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
+
+In contrast to @filoso2017, we also did not identify an effect of the
+
+Given how skewed Dryness is due to the few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
+
+### Other possible bias in the data
+
 There are further confounding factors in the data, which were also classified by @filoso2017 and these create biases in the data set that can impact the overall assessment.

### 2022-01-10 paper/Forest_and_Water/Forest_and_Water.tex
adding catchments from Filoso et al 201

@@
   \linenumbers % turns line numbering on
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
 \usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 \bibliographystyle{elsarticle-harv}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={Do larger watersheds respond different to forest cover change? Re-analysing a global data set.},
+            colorlinks=false,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+
+\setcounter{secnumdepth}{5}
+% Pandoc toggle for numbering sections (defaults to be off)
+
+% Pandoc syntax highlighting
 \usepackage{color}
@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable}
-\usepackage{graphicx}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Do larger watersheds respond different to forest cover change? Re-analysing a global data set.},
-            colorlinks=false,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-
-\setcounter{secnumdepth}{0}
-% Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
-
-% Pandoc citation processing
-
-% Pandoc header
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
 \begin{document}
+
 \begin{frontmatter}
 
-  \title{Do larger watersheds respond different to forest cover change?
-Re-analysing a global data set.}
+  \title{Do larger watersheds respond different to forest cover change? Re-analysing a global data set.}
     \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
@@
       \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
-    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney,
-Sydney, NSW 2006, Australia}
-    \address[INIA]{Project Manager, FPTA 357 Instituto Nacional de Investigación
-Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100
-Canelones, Uruguay}
-    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
-Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
+    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
+    \address[INIA]{Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
       \cortext[1]{Corresponding Author}
@@
   \begin{abstract}
-  This is the abstract. It consists of two paragraphs.
+  This is the abstract.
+  It consists of two paragraphs.
   \end{abstract}
@@
 
-There has been an long and on-going discussion in the hydrologcal
-literature around the impact of forests on streamflow (Andréassian,
-2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al.,
-2005; Zhang et al., 2017). The historic work highlights a general
-consensus that if forest areas increase, streamflow decreases and
-vice-versa. The most dramatic result in relation to this, is Figure 5 in
-Zhang et al. (2011) indicating (for Australian watersheds) a 100\%
-decrease in streamflow for watersheds with 100\% forest cover. However,
-on the other end of the spectrum, in a series of French watersheds
-(Cosandey et al., 2005), there was no change in streamflow
-characteristics in 2 of the three watersheds studied in relation to
-deforestation.
-
-Several review papers have summarized different studies across the
-globe, in relation to paired watershed studies (Bosch and Hewlett, 1982;
-Brown et al., 2005), related to reforestation in particular (Filoso et
-al., 2017), and more generally (Jackson et al., 2005; Zhang et al.,
-2017). These studies aim to generalize the individual findings and to
-identify if there are global trends or relationships that can be
-developed. The most recent reviews (Filoso et al., 2017; Zhang et al.,
-2017) developed an impressive global database of watershed studies in
-relation to changes in streamflow due to changes in forest cover. The
-Zhang et al. (2017) dataset, which covers over 250 studies, is described
-in terms of the change in streamflow as a result of the change in forest
-cover, where studies related to both forestation (increase in forest
-cover) and deforestation (decrease in forest cover) were included. In
-contrast, the paper by Filoso et al. (2017) focused primarily on
-reforestation, and covered an equally impressive database of 167 studies
-using a systematic review. In this case the collected data is mostly
-coded as count data and only a subset of 37 studies was analysed for
-actual water yield change.
-
-The conclusions of the first paper (Zhang et al., 2017) suggest that
-there is a distinct difference in the change in flow as a result of
-forestation or deforestation between small watersheds, defined as
-\textless{} 1000 km\textsuperscript{2} and large watersheds
-\textgreater{} 1000 km\textsuperscript{2}. While for small watersheds
-there was no real change in runoff with changes in cover, for large
-watersheds there was a clear trend showing a decrease in runoff with and
-increase in forest cover. Their main conclusion was that the response in
-annual runoff to forest cover was scale dependent and appeared to be
-more sensitive to forest cover change in water limited watersheds
-relative to energy limited watershed (Zhang et al., 2017).
-
-The second study (Filoso et al., 2017) was a systematic review which
-classified the historical research and highlighted gaps in the spatial
-distribution, the types of studies and the types of analysis. Their main
-conclusion was also that reforestation decreases streamflow, but that
-there were many interacting factors. For a subset of quantitative data
-(37) they showed a relationship between watershed size and decline in
-streamflow.
-
-A final summary paper that includes much of the same data as Zhang et
-al. (2017) and Filoso et al. (2017) is Zhou et al. (2015), which has one
-author in common with Zhang et al. (2017). However, this paper aims to
-explain the variation in the data using the Fuh model, and in particular
-aims to link the variation in the observed data to variations in the
-exponent \emph{m} in the model. A key observation is that in drier
-environments, the effects of deforestation are much greater than in
-wetter environments, which is also suggested by Figure 4 in Zhang et al.
+There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017).
+The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
+The most dramatic result in relation to this, is Figure 5 in Zhang et al.
+(2011) indicating (for Australian watersheds) a 100\% decrease in streamflow for watersheds with 100\% forest cover.
+However, on the other end of the spectrum, in a series of French watersheds (Cosandey et al., 2005), there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
+
+Several review papers have summarized different studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017).
+These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
+The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover.
+The Zhang et al.
+(2017) dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+In contrast, the paper by Filoso et al.
+(2017) focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
+In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+
+The conclusions of the first paper (Zhang et al., 2017) suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as \textless{} 1000 km\textsuperscript{2} and large watersheds \textgreater{} 1000 km\textsuperscript{2}.
+While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover.
+Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed (Zhang et al., 2017).
+
+The second study (Filoso et al., 2017) was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
+Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
+For a subset of quantitative data (37) they showed a relationship between watershed size and decline in streamflow.
+
+A final summary paper that includes much of the same data as Zhang et al.
+(2017) and Filoso et al.
+(2017) is Zhou et al.
+(2015), which has one author in common with Zhang et al.
 (2017).
-
-Encouraged by the work presented by Zhang et al. (2017) and Filoso et
-al. (2017) and the fantastic database of studies presented by these
-authors, we believe we can add to the discussion. In this paper, the aim
-is to develop further analysis of the collected data and expanding and
-combining the two data sets to provide further depth.
-
-In particular, the main method in the work by Zhang et al. (2017) is
-using simple linear regression, and in Filoso et al. (2017) the focus is
-mainly on classification. As Zhang et al. (2017) points out, the main
-assumption in their work is that the threshold at 1000
-km\textsuperscript{2} is a distinct separation between ``small'' and
-``large'' watersheds, but the subset of data in Filoso et al. (2017)
-does not appear to support this. And while te work Filoso et al. (2017)
-provides important insights in study types, analysis types and broad
-classification, there is limited quantification of actual impact. This
-is because the work had a strict criterion to select quantitative
-studies. However, given the fantastic data sets collected, the analyses
-can be easily expanded to look at interactions between the terms and to
-test the assumption of a distinct threshold at 1000
-km\textsuperscript{2}.
-
-As a result the objective of this paper is to 1) enhance the data set
-from Zhang et al. (2017) with further watersheds (such as from Filoso et
-al. (2017)) and spatial coordinates and 2) to analyse the possibility of
-non-linear, interactions and partial effects of the different factors
-and variables in the data using generalised linear (GLM) and generalised
-additive models (GAM Wood (2006)).
-
-Building on the analyses by Zhang et al. (2017) and Filoso et al.
-(2017), and combining their conclusions, the main hypothesis to test is
-that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under
-consideration (affecting the length of the flowpaths Zhou et al.
-(2015)), the length of the study (c.f. Jackson et al. (2005)) and
-possibly the climate (as indicated by either E0/Pa or latitude and
-longitude Filoso et al. (2017); Zhou et al. (2015)).
-
-However, there could be further confounding factors, which are eluded to
-by Filoso et al. (2017):
+However, this paper aims to explain the variation in the data using the Fuh model, and in particular aims to link the variation in the observed data to variations in the exponent \emph{m} in the model.
+A key observation is that in drier environments, the effects of deforestation are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al.
+(2017).
+
+Encouraged by the work presented by Zhang et al.
+(2017) and Filoso et al.
+(2017) and the fantastic database of studies presented by these authors, we believe we can add to the discussion.
+In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.
+
+In particular, the main method in the work by Zhang et al.
+(2017) is using simple linear regression, and in Filoso et al.
+(2017) the focus is mainly on classification.
+As Zhang et al.
+(2017) points out, the main assumption in their work is that the threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' watersheds, but the subset of data in Filoso et al.
+(2017) does not appear to support this.
+And while te work Filoso et al.
+(2017) provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact.
+This is because the work had a strict criterion to select quantitative studies.
+However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km\textsuperscript{2}.
+
+As a result the objective of this paper is to 1) enhance the data set from Zhang et al.
+(2017) with further watersheds (such as from Filoso et al.
+(2017)) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).
+
+Building on the analyses by Zhang et al.
+(2017) and Filoso et al.
+(2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
+However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths Zhou et al.
+(2015)), the length of the study (c.f.
+Jackson et al.
+(2005)) and possibly the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al.
+(2017); Zhou et al.
+(2015)).
+
+However, there could be further confounding factors, which are eluded to by Filoso et al. (2017):
 
@@
 \item
-  the type of analysis, i.e.~paired watershed studies, modelling, time
-  series analysis etc.
+  the type of analysis, i.e.~paired watershed studies, modelling, time series analysis etc.
 \item
-  the age of the study, assuming that historical studies might not have
-  had the ability to measure at the accuracy that currently is available
-  to researchers, or that more careful historical attention to detail in
-  field studies might have been lost more recently due to reductions in
-  research investment.
+  the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
 \end{itemize}
 
-Finally, this work aims to point to further research that can expand
-this area of work, based on the collected data, to better understand the
-impact of forest cover change on streamflow.
+Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
 
@@
 
-The starting point of this paper is the data base of studies which were
-included in Zhang et al. (2017) as supplementary material. The columns
-in this data set are the watershed number, the watershed name, the Area
-in km\textsuperscript{2}, the annual average precipitation (Pa) in mm,
-the forest type, hydrological regime, and climate type, the change in
-forest cover in \% (\(\Delta\)F\%) and the change in streamflow in \%
-(\(\Delta\)Qf(\%), based on equation 1 in Zhang et al. (2017)), the
-precipitation data type, the assessment technique, and the source of the
-info, which is a citation. Several of these columns contain
-abbreviations to describe the different variables, which are summarised
-in Table 1.
-
-Table 1 Summary of abbreviations of factors used in the Zhang et al.
-(2017) data set
+The starting point of this paper is the data base of studies which were included in Zhang et al.
+(2017) as supplementary material.
+The columns in this data set are the watershed number, the watershed name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in Zhang et al.
+(2017)), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
+
+Table 1 Summary of abbreviations of factors used in the Zhang et al. (2017) data set
 
@@
 \toprule
-Factor & Abbreviation & Definition\tabularnewline
-\midrule
-\endhead
-forest type & CF & coniferous forest\tabularnewline
-& BF & broadleaf forest\tabularnewline
-& MF & mixed forest\tabularnewline
-hydrological regime & RD & rain dominated\tabularnewline
-& SD & snow dominated\tabularnewline
-climate type & EL & energy limited\tabularnewline
-& WL & water limited\tabularnewline
-& EQ & equitant\tabularnewline
-precipitation data type & OB & observed\tabularnewline
-& SG & spatial gridded\tabularnewline
-& MD & modelled\tabularnewline
-assessment technique & PWE & paired watershed experiment\tabularnewline
-& QPW & quasi-paired watershed experiment\tabularnewline
-& HM & hydrological modelling\tabularnewline
-& EA & elastictity analysis\tabularnewline
-& SH & combined use of statistical methods\tabularnewline
-& & and hydrographs\tabularnewline
-\bottomrule
-\end{longtable}
-
-While Zhang et al. (2017) use the dryness index in their analysis,
-potential or reference evapotranspiration was not originally included as
-part of the published data set. We combined the tables for small
-(\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000
-km\textsuperscript{2}) watershed data sets in our analysis.
+Factor & Abbreviation & Definition \\
+\midrule
+\endhead
+forest type & CF & coniferous forest \\
+& BF & broadleaf forest \\
+& MF & mixed forest \\
+hydrological regime & RD & rain dominated \\
+& SD & snow dominated \\
+climate type & EL & energy limited \\
+& WL & water limited \\
+& EQ & equitant \\
+precipitation data type & OB & observed \\
+& SG & spatial gridded \\
+& MD & modelled \\
+assessment technique & PWE & paired watershed experiment \\
+& QPW & quasi-paired watershed experiment \\
+& HM & hydrological modelling \\
+& EA & elastictity analysis \\
+& SH & combined use of statistical methods \\
+& & and hydrographs \\
+\bottomrule
+\end{longtable}
+
+While Zhang et al. (2017) use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
+We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) watershed data sets in our analysis.
 
 \hypertarget{additional-data-collection}{%
-\subsection{Additional data
-collection}\label{additional-data-collection}}
-
-To enhance the existing data set, this study added additional variables
-and cross-checked the studies with the data set from Filoso et al.
-(2017). In particular, we focussed on the 37 data points included in the
-quantitative analysis in Filoso et al. (2017).
-
-In addition, additional variables added were the latitude and longitude
-for the center of the watershed as an approximation of its spatial
-location. Using this information reference evapotranspiration
-(E\textsubscript{0}) was extracted from the Global Aridity Index and
-Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2
-(Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not
-available from the original papers. For large watersheds, this value,
-similar to annual average rainfall, is only an approximation of the
-climate at the location.
-
-The length of the study can be a variable influencing the change in flow
-(e.g.~Jackson et al., 2005), as for example, more mature plantations are
-thought to have smaller impacts on flow. Therefore, the length of the
-study calculate as the difference between the starting data and
-completion date of the different studies was extracted from the
-references provided by Zhang et al. (2017).
-
-Several additional data points from watershed studies were extracted
-from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988),
-Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010),
-Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were
-checked against the existing studies to prevent overlap. In the citation
-column in the data set, in general the main reference for the calculated
-change in streamflow was used, because sometimes the original study did
-not provide the quantification of the change in streamflow (i.e.~Table 6
-in Zhang et al. (2011)). We also removed one data point from the
-analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al.
-(2017). This is because the cited reference (Roche, 1981) only relates
-to 1 and 1.5 ha paired watershed studies in French Guyana, and in which
-the actual change in forest cover is not recorded.
-
-The final column in the improved data set is a ``notes'' column, which
-is not further used in the analysis, but gives context to some of the
-data for future research and highlights some of the discrepancies that
-we found between the original papers and the data in the tables from
-Zhang et al. (2017).
-
-Similar to Zhang et al. (2017), the ``dryness index'' was calculated
-from the reference evapotranspiration and the annual average rainfall
-(Pa) as: \[\tag{1}
+\subsection{Additional data collection}\label{additional-data-collection}}
+
+To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from Filoso et al.
+(2017).
+In particular, we focussed on the 37 data points included in the quantitative analysis in Filoso et al.
+(2017).
+
+In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
+Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
+For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+
+The length of the study can be a variable influencing the change in flow (e.g.
+Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow.
+Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al.
+(2017).
+
+Several additional data points from watershed studies were extracted from Zhang et al.
+(2011), Zhao et al.
+(2010), Borg et al.
+(1988), Thornton et al.
+(2007), Zhou et al.
+(2010), Rodriguez et al.
+(2010), Ruprecht et al.
+(1991) and Peña-Arancibia et al.
+(2012), and these were checked against the existing studies to prevent overlap.
+In the citation column in the data set, in general the main reference for the calculated change in streamflow was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al.
+(2011)).
+We also removed one data point from the analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al.
+(2017).
+This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
+
+The final column in the improved data set is a ``notes'' column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al.
+(2017).
+
+Similar to Zhang et al. (2017), the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+\[\tag{1}
 \begin{aligned}
@@
 
-Removing possible duplicates removes 29 watersheds from the overall data
-set
-
-To estimate how the change in streamflow is affected by the change in
-forest cover while considering the effects of the other variables, we
-applied generalised additive modelling (GAM) (Wood, 2006).
+Removing possible duplicates removes 29 watersheds from the overall data set
+
+To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) (Wood, 2006).
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}\]
 
-Here \(X_i\) are factorial variables, while \(Z_i\) are continous
-variables. The model assumes no direct interactions and all variables
-are additive. The changes in forest cover contain both positive
-(forestation) and negative values (deforestation). In Zhang et al.
-(2017), these changes were jointly analysed, assuming the effect on the
-change in flow was linear and non-hysteretic. However, the impact of an
-increase in forest cover can be different from the same fractional
-decrease in forest cover. Therefore all the change in forest cover data
-is converted to positive values, and an additional column
-(\emph{sign\_\{forest cover\}}) is added that indicates whether it was a
-forest cover increase or decrease. A further assumption in the model is
-that all continuous variables \(Z_i\) (such as annual precipitation
-(Pa)) can have a linear or non-linear relationship with \(\Delta Q \%\).
-This means that a smooth function \(s()\) is applied to the \(Z_i\)
-variables. For the smoothing function we applied thin plate regression
-splines with an additional shrinkage penalty which means the terms can
-be shrunk to 0 if not significant (Wood, 2006).
-
-For the model in equation 2, we initially only used the data from Zhang
-et al. (2017) to make sure that the additional watersheds added to the
-data set did not influence the results. Subsequently the analysis was
-repeated and the additionally identified watersheds were added.
+Here \(X_i\) are factorial variables, while \(Z_i\) are continuous variables.
+The model assumes no direct interactions and all variables are additive.
+The changes in forest cover contain both positive (forestation) and negative values (deforestation).
+In Zhang et al.
+(2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation.
+However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
+Therefore all the change in forest cover data is converted to positive values, and an additional column (\(sign_{forest cover}\)) is added that indicates whether it was a forest cover increase or decrease.
+A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have a linear or non-linear relationship with \(\Delta Qf \%\).
+This means that a smooth function \(s()\) is applied to the \(Z_i\) variables.
+For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant (Wood, 2006).
+
+For the model in equation 2, we initially only used the data from Zhang et al.
+(2017) to make sure that the additional watersheds added to the data set did not influence the results.
+Subsequently the analysis was repeated and the additionally identified watersheds were added.
 
@@
 
-The overall dataset contains 312 observations of changes in flow. The
-overall distribution of changes in flow is highly skewed as is the
-distribution of changes in forest cover and Area. The values of changes
-in flow greater than 100\% and smaller than -100\% clearly create long
-tails on the change in flow distribution. Note also the large number of
-studies with 100\% forest cover reduction. Smaller watersheds dominate
-the database with 42\% of the data from watersheds \textless{} 1
-km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10
-km\textsuperscript{2}.
+The overall dataset contains 309 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+In contrast, the original dataset from Zhang et al.
+(2017) contained 312 observations.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
+The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
+Note also the large number of studies with 100\% forest cover reduction.
+Smaller watersheds dominate the database with 42\% of the data from watersheds \textless{} 1 km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10 km\textsuperscript{2}.
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{DataExploration} \caption{Overview of the distributions of some of the variables in the data set}\label{fig:data_graphs}
+\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{foo}\label{fig:datagraphs}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of increases and decreases in forest cover}\label{fig:increase_decrease}
+\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of increases and decreases in forest cover}\label{fig:increasedecrease}
 \end{figure}
 
-This shows that for the data related to forest decreases, there is
-almost always a positive flow change. In other words, flow almost always
-increased. However, for increases in forest cover, this is not the case,
-and flow can both increase and decrease. However in both cases the
-variability in the reported change in flow increases with the increase
-in forest cover change.
-
-\hypertarget{the-initial-relationship-between-change-in-forest-cover-and-streamflow}{%
-\subsection{The initial relationship between change in forest cover and
-streamflow}\label{the-initial-relationship-between-change-in-forest-cover-and-streamflow}}
-
-Following Zhang et al. (2017), the first step is to use a linear
-regression to investigate the percent change in flow as a result in the
-percent change forestry and modulated by the direction of the change,
-either an increase in forest cover, or decrease in forest cover:
+This shows that for the data related to forest decreases, there is almost always a positive flow change.
+In other words, flow almost always increased.
+However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
+However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
+
+\hypertarget{the-general-relationship-between-change-in-forest-cover-and-streamflow}{%
+\subsection{The general relationship between change in forest cover and streamflow}\label{the-general-relationship-between-change-in-forest-cover-and-streamflow}}
+
+Following Zhang et al.
+(2017), the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
 \end{aligned}\]
 
-\begin{longtable}[]{@{}ccccc@{}}
-\caption{Summary results of the first regression model predicting change
-in streamflow from change in forest cover and accounting for the
-direction of the change}\tabularnewline
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-8.65\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.56\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.56\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.12\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.45\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.26\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--29.17\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.79\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.04\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\caption{Summary results of the first regression model predicting change
-in streamflow from change in forest cover and accounting for the
-direction of the change including the new data sets}\tabularnewline
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
+\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
+\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:tabmodel1b} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-9.43\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.54\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.7\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.09\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.44\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-5.12\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--36.54\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.59\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.53\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-While the overall variance explained in this model is not high with an
-adjusted \emph{r\textsuperscript{2}} of 0.18, it clearly supports the
-hypothesized relationship between the change in forest cover and the
-change in flow. The model suggests that for every 1\% change in forest
-cover, on the average, the flow changes 0.45\%. However the change in
-flow is different for forest cover decreases compared to forest cover
-increases. In fact, forest cover increases decrease flow by 29\% less
-than a similar decrease in forest cover causes flow to increase. So
-roughly speaking, a 1\% forest cover increase on the average decreases
-flow by \((1 - 0.29)*0.45\%\), while a the percentage forest cover
-decrease will increase flow by 0.45\%.
-
-Checking this result against including the new data indicates that this
-mainly strengthens the difference between the forest cover increases and
-decreases, and decreases the mean decrease in flow as a result of forest
-cover change
-
-It is however it is clear from the lack of explaining power for the
-model, that there could be confounding factors, as alluded to in the
-methods. The obvious ones being watershed dryness and area (following
-Zhang et al. (2017)):
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 9.43 & 5.54 & 1.7 & 0.09 \\
+\textbf{DeltaF\_perc\_pos} & 0.44 & 0.09 & 5.12 & 0 \\
+\textbf{Forest\_Signincrease} & -36.54 & 5.59 & -6.53 & 0 \\
+\bottomrule
+\end{longtable}
+
+The overall variance explained in this model is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
+However the change in flow is different for forest cover decreases compared to forest cover increases.
+In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase.
+So roughly speaking, a 1\% forest cover increase on the average decreases flow by \((1 - 0.29)*0.45\%\), while a the percentage forest cover decrease will increase flow by 0.45\%.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{residual_plot_model1} \caption{Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution}\label{fig:gamcheck}
+\end{figure}
+
+Of importance here is to highlight the residuals of this regression.
+These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \ref{fig:gamcheck}).
+In other words, the distribution of the residuals is somewhat fat-tailed.
+We will discuss this later.
+
+Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1} and Table \ref{tab:tabmodel1b}), and the result indication a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
+
+It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
+The obvious ones being watershed dryness and area (following Zhang et al.
+(2017)):
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
 \end{aligned}\]
 
-Where \(Pa_mm\) is the annual average rainfall in mm, and a log base 10
-transformation is applied to the area variable given that the
-distribution of Area is higly skewed (Figure 1)
-
-\begin{longtable}[]{@{}ccccc@{}}
-\caption{Summary of the second model, taking into account the annual
-rainfall and the area of the watershed}\tabularnewline
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
+Where \(Pa_mm\) is the annual average rainfall in mm, and a log base 10 transformation is applied to the area variable given that the distribution of Area is higly skewed (Figure 1)
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:out-model2} Summary of the second model, taking into account the annual rainfall and the area of the watershed}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-22.55\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-9.16\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.46\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.34\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.26\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--35.52\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.67\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.27\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{log10(Area\_km2)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--3.12\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-1.72\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1.81\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.32\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-Including area and annual precipitation slightly improves the overall
-explaining power of the model. Annual precipitation is in fact not
-significant. Relative to earlier reported studies (Filoso et al., 2017;
-Zhang et al., 2017), the log base 10 transformed watershed area
-indicates a p-value of only 0.07, suggesting a marginal impact on the
-change in stream flow. This supports our approach (in contrast to Zhang
-et al. (2017)) to consider watershed area as a continuous variable and
-making no separation between larger and smaller watersheds The main
-effects remain the change in forest cover and whether this is an
-increase or decrease.
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 22.55 & 9.16 & 2.46 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.34 & 0.1 & 3.26 & 0 \\
+\textbf{Forest\_Signincrease} & -35.52 & 5.67 & -6.27 & 0 \\
+\textbf{log10(Area\_km2)} & -3.12 & 1.72 & -1.81 & 0.07 \\
+\textbf{Pa\_mm} & 0 & 0 & -1 & 0.32 \\
+\bottomrule
+\end{longtable}
+
+Including area and annual precipitation slightly improves the overall explaining power of the model (Table \ref{tab:out-model2}).
+Annual precipitation is in fact not significant.
+Relative to earlier reported studies (Filoso et al., 2017; Zhang et al., 2017), the log base 10 transformed watershed area indicates a p-value of only 0.07, suggesting a marginal impact on the change in stream flow.
+This supports our approach (in contrast to Zhang et al.
+(2017)) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds.
+The main effects in the model remain the change in forest cover and whether this is an increase or decrease.
 
 \hypertarget{the-effect-of-location-on-the-globe}{%
-\subsection{The effect of location on the
-globe}\label{the-effect-of-location-on-the-globe}}
+\subsection{The effect of location on the globe}\label{the-effect-of-location-on-the-globe}}
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:global_map}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
 \end{figure}
 
-As indicated, a further hypothesis relates to whether there is a strong
-spatial global gradient as captured by latitude and longitude. As the
-global map (@ref(fig:global\_map)) shows, the distribution of case study
-watersheds covers multiple continents and shows some distinct clustering
-in parts of the world. Of interest is whether the spatial clustering
-also indicates a difference in response to forest cover change:
+As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
+As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
+Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
 
@@
 \begin{aligned}
-\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
+\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}\]
 
-\begin{longtable}[]{@{}ccccc@{}}
-\caption{Results of the model including Latitude and Longitude including
-new data}\tabularnewline
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:out-model3} Results of the model including Latitude and Longitude including new data}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-23.29\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-9.96\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.34\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.35\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-3.32\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--37.21\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.2\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{log10(Area\_km2)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--3.28\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-1.73\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1.89\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.06\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Pa\_mm}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.83\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.41\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Latitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--0.05\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.09\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--0.55\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.58\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Longitude}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.03\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.23\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.82\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-There appears to be no significant latitudinal gradient. The total
-explaining power of the model is still low with an adjusted
-\emph{r\textsuperscript{2}} of 0.22 suggesting further factors that are
-currently not included in the model.
-
-Note that in this case the significance of the Area variable increases,
-and generally indicates that larger watersheds would be expected to have
-a lower change in streamflow as also indicated in Zhang et al. (2017).
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 23.29 & 9.96 & 2.34 & 0.02 \\
+\textbf{DeltaF\_perc\_pos} & 0.35 & 0.1 & 3.32 & 0 \\
+\textbf{Forest\_Signincrease} & -37.21 & 6 & -6.2 & 0 \\
+\textbf{log10(Area\_km2)} & -3.28 & 1.73 & -1.89 & 0.06 \\
+\textbf{Pa\_mm} & 0 & 0 & -0.83 & 0.41 \\
+\textbf{Latitude} & -0.05 & 0.09 & -0.55 & 0.58 \\
+\textbf{Longitude} & 0.01 & 0.03 & 0.23 & 0.82 \\
+\bottomrule
+\end{longtable}
+
+There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the watersheds across the globe has little influence on the overall result.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+
+Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow, as also indicated in Zhang et al.
+(2017).
 
 \hypertarget{impact-of-the-dryness-index}{%
-\subsection{Impact of the dryness
-index}\label{impact-of-the-dryness-index}}
-
-Climate, and in particular evapotranspiration would be expected to have
-a significant effect on the streamflow change as represented by the
-dryness index, which is also indicated in Zhang et al. (2017). Increased
-evapotranspiration could lead to drier watersheds, unless balanced by
-rainfall (such as possibly in the tropics). This model introduces the
-dryness index as a linear variable and drops the annual average
-precipitation as a variable, as dryness is calculated from the
-precipitation. It also drops the Latitude and Longitude as these are
-indicated not to be significant.
-
-\begin{longtable}[]{@{}ccccc@{}}
-\caption{Results of the model replacing the annual precipitation with
-the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
+\subsection{Impact of the dryness index}\label{impact-of-the-dryness-index}}
+
+Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both Zhang et al.
+(2017) and Jackson et al.
+(2005).
+Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
+This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from the precipitation.
+It also drops the Latitude and Longitude as these are indicated not to be significant.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:out-model4} Results of the model replacing the annual precipitation with the dryness index}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-10.46\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-7.66\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1.36\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.17\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.31\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.93\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--35.67\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-5.73\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.23\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{log10(Area\_km2)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--3.5\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-1.76\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--1.99\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.05\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Dryness}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-6.54\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-3.08\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.12\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.03\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-The results from this model confirm that dryness is a significant
-confounding factor of the change in streamflow as function of the change
-in forest cover change. In fact if the dryness index doubles
-(remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa,
-which is very dry), the change in runoff is \textasciitilde14\% greater.
-However, more interesting, Latitude remains a significant predictor with
-each degree in latitude causing an -0.31\% change in runoff. This
-indicates that Dryness (i.e.~an increase in radiation) alone does not
-explain the trend in the Latitude and some other unknown confounding
-factor is captured by Latitude.
-
-However, the result also indicates possible issues with the data, some
-of the Dryness values are very large (\textgreater{} 4) and these values
-have high leverage in the data. These watersheds are listed in Table 5:
-
-Quitting from lines 471-472 (Forest\_and\_Water.Rmd) Error in filter(.,
-Dryness \textgreater{} 4) : object `Zhang\_all2' not found Calls:
-\ldots{} withVisible -\textgreater{} eval -\textgreater{} eval
--\textgreater{} pander -\textgreater{} \%\textgreater\% -\textgreater{}
-select -\textgreater{} filter In addition: Warning messages: 1: In
-eval(expr, envir, enclos) : NAs introduced by coercion 2: In eval(expr,
-envir, enclos) : NAs introduced by coercion
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 10.46 & 7.66 & 1.36 & 0.17 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.11 & 2.93 & 0 \\
+\textbf{Forest\_Signincrease} & -35.67 & 5.73 & -6.23 & 0 \\
+\textbf{log10(Area\_km2)} & -3.5 & 1.76 & -1.99 & 0.05 \\
+\textbf{Dryness} & 6.54 & 3.08 & 2.12 & 0.03 \\
+\bottomrule
+\end{longtable}
+
+The results from this model (Table \ref{tab:out-model4}) confirm that dryness is a significant confounding factor related to the change in streamflow as a function of the change in forest cover change.
+In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is \textasciitilde14\% greater.
+However, more interesting, Latitude remains a significant predictor with each degree in latitude causing an -0.31\% change in runoff.
+This indicates that Dryness (i.e.~an increase in radiation) alone does not explain the trend in the Latitude and some other unknown confounding factor is captured by Latitude.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.17}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.46}}@{}}
+\caption{\label{tab:drytable} Watersheds for which the dryness index \textgreater{} 4}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+Latitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Longitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Watershed name
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+Latitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Longitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Watershed name
+\end{minipage} \\
+\midrule
+\endhead
+34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
+36.4 & -120.4 & Cantua \\
+34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+32.74 & -111.5 & Natural DRDages, Ariz., U.S.A,
+A \\
+\bottomrule
+\end{longtable}
+
+However, the result also indicates possible issues with the data, some of the Dryness values are very large (\textgreater{} 4) and these values have high leverage in the data.
+These watersheds are listed in Table \ref{tab:drytable}.
 
 \hypertarget{are-some-of-the-variables-possibly-non-linear}{%
-\subsubsection{Are some of the variables possibly
-non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
-
-The work by Filoso et al. (2017) and earlier by Jackson et al. (2005)
-has indicated that the length of the study might influence the response.
-This links to the idea from Kuczera (1987) that the effect of logging or
-deforestation or reforestation reduces with the length of time post
-intervention (see also Jackson et al. (2005)). In addition to adding the
-length as a variable, all continuous variables (Dryness, Area, length
-and Latitude) were considered non-linear in this model and as indicated
-a shrinkage smoothing spline (Wood, 2006) was applied to these
-variables.
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(log10(Area_km2), 
-##     bs = "ts") + s(Dryness, bs = "ts") + s(length, bs = "ts")
-## 
-## Parametric coefficients:
-##                     Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)          17.0942     6.0776   2.813  0.00526 ** 
-## DeltaF_perc_pos       0.2606     0.1067   2.442  0.01521 *  
-## Forest_Signincrease -34.5961     6.1052  -5.667 3.59e-08 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                       edf Ref.df     F p-value   
-## s(log10(Area_km2)) 1.7399      9 0.782 0.01465 * 
-## s(Dryness)         0.8731      9 0.570 0.01514 * 
-## s(length)          8.8276      9 3.097 0.00114 **
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.287   Deviance explained =   32%
-## GCV = 2171.2  Scale est. = 2065.7    n = 297
-\end{verbatim}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.31\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-17.09\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.08\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.81\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.26\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-0.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-2.44\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.31\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--34.6\strut
-\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
-6.11\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--5.67\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-edf\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-Ref.df\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-F\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-p-value\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(log10(Area\_km2))}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-1.74\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.78\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(Dryness)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.87\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.57\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(length)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-8.83\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-3.1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-Including non-linearity increases the overall explaining power of the
-model to an adjusted \emph{r\textsuperscript{2}} of 0.29, but creates
-few changes in the significance of the variables. However, it also can
-create a chance of overfitting, as the smoothing splines allow
-significant flexibility, which will be investigated below.
-
-However, first a final full model which includes the remaining
-categorical variables (Precipitation data type, Assessment technique,
-Forest type and Hydrological regime) are included in the model
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.36\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--27.3\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-19.84\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--1.38\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.17\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.31\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.01\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--21.54\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-7.29\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--2.96\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Precip\_data\_typeOB}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--8.13\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-15.17\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.54\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.59\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Precip\_data\_typeSG}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-17.33\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-17.29\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.32\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueEA, HM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-13.93\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-46.74\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-0.3\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.77\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueHM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-32.96\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-13.92\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-2.37\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniquePWE}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-53.24\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-12.81\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-4.16\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniquePWE, HM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-33.54\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-47.07\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-0.71\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.48\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueQPW}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-39.8\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-22.06\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.8\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueQPW, EA}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-45.97\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-26.25\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.75\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.08\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueSH}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-44.89\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-13.49\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.33\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_typeCF}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--2.2\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-8.52\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.26\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.8\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_typeMF}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--6.56\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-8.96\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.73\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.46\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Hydrological\_regimeSD}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-4.59\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-10.25\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-0.45\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.65\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-edf\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-Ref.df\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-F\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-p-value\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(log10(Area\_km2))}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.31\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.07\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.16\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(Dryness)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-3.45\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-1.39\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(length)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-8.73\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-3.04\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model6_all-1.pdf}
-
-Clearly Latitude is masking other factors including the assessment
-technique and the forest type
-
-\begin{verbatim}
-## Warning: Removed 5 rows containing non-finite values (stat_boxplot).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-20-1.pdf}
-
-\begin{verbatim}
-## # A tibble: 8 x 2
-##   Assessment_technique     n
-##   <chr>                <int>
-## 1 EA                      32
-## 2 EA, HM                   1
-## 3 HM                      53
-## 4 PWE                    185
-## 5 PWE, HM                  1
-## 6 QPW                      7
-## 7 QPW, EA                  4
-## 8 SH                      26
-\end{verbatim}
-
-Maybe remove data for low numbers of assessment techniques and rerun the
-analysis (\textless10)
+\subsubsection{Are some of the variables possibly non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
+
+The work by Filoso et al.
+(2017) and earlier by Jackson et al.
+(2005) has indicated that the length of the study might influence the response.
+This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al.
+(2005)).
+In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 17.09 & 6.08 & 2.81 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
+\textbf{Forest\_Signincrease} & -34.6 & 6.11 & -5.67 & 0 \\
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(log10(Area\_km2))} & 1.74 & 9 & 0.78 & 0.01 \\
+\textbf{s(Dryness)} & 0.87 & 9 & 0.57 & 0.02 \\
+\textbf{s(length)} & 8.83 & 9 & 3.1 & 0 \\
+\bottomrule
+\end{longtable}
+
+Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.29 and deviance explained of 0.32, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
+
+A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.42}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{Statistical summary for the linear terms the full model}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & -27.3 & 19.84 & -1.38 & 0.17 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 3.01 & 0 \\
+\textbf{Forest\_Signincrease} & -21.54 & 7.29 & -2.96 & 0 \\
+\textbf{Precip\_data\_typeOB} & -8.13 & 15.17 & -0.54 & 0.59 \\
+\textbf{Precip\_data\_typeSG} & 17.33 & 17.29 & 1 & 0.32 \\
+\textbf{Assessment\_techniqueEA, HM} & 13.93 & 46.74 & 0.3 & 0.77 \\
+\textbf{Assessment\_techniqueHM} & 32.96 & 13.92 & 2.37 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 53.24 & 12.81 & 4.16 & 0 \\
+\textbf{Assessment\_techniquePWE,
+HM} & 33.54 & 47.07 & 0.71 & 0.48 \\
+\textbf{Assessment\_techniqueQPW} & 39.8 & 22.06 & 1.8 & 0.07 \\
+\textbf{Assessment\_techniqueQPW,
+EA} & 45.97 & 26.25 & 1.75 & 0.08 \\
+\textbf{Assessment\_techniqueSH} & 44.89 & 13.49 & 3.33 & 0 \\
+\textbf{Forest\_typeCF} & -2.2 & 8.52 & -0.26 & 0.8 \\
+\textbf{Forest\_typeMF} & -6.56 & 8.96 & -0.73 & 0.46 \\
+\textbf{Hydrological\_regimeSD} & 4.59 & 10.25 & 0.45 & 0.65 \\
+\bottomrule
+\end{longtable}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{pander}\NormalTok{(}\FunctionTok{round}\NormalTok{(}\FunctionTok{summary}\NormalTok{(model6\_all)}\SpecialCharTok{$}\NormalTok{s.table,}\DecValTok{2}\NormalTok{), }\AttributeTok{caption =} \StringTok{"Statistical summary for the smooth terms for the full model"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{Statistical summary for the smooth terms for the full model}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(log10(Area\_km2))} & 0.31 & 9 & 0.07 & 0.16 \\
+\textbf{s(Dryness)} & 3.45 & 9 & 1.39 & 0.01 \\
+\textbf{s(length)} & 8.73 & 9 & 3.04 & 0 \\
+\bottomrule
+\end{longtable}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#plot(model6\_all)}
+\end{Highlighting}
+\end{Shaded}
+
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.38.
+This indicates that the categorical variables explain a limited amount of the variance.
+However, interesting to note from Table \ref{tab:msix-linear} that several of the assessment methods are significant.
+In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments}\label{fig:assessment}
+\includegraphics[width=0.9\linewidth]{model6_smooths} \caption{Visualisation of the smooth variables in the model}\label{fig:smoothsmodel6}
 \end{figure}
 
-Clearly all have at least some relationship with Latitude, therefore are
-being masked if Latitude is included in the model.
-
-The flexible nature of the splines and the shrinkage applied to the
-splines means that the length variable can capture variation in the
-data, but it is unclear what exactly is captured. The shape of the
-resulting conditional response does not reflect the reponses highlighted
-in Filoso et al. (2017) and Jackson et al. (2005). Reducing the
-flexibility of the splines, or fitting a linear term results in
-``length'' not being significant, which suggests that this variable is
-not helpful in explaining the variation in the data.
-
-\begin{verbatim}
-## Warning: Removed 5 rows containing missing values (geom_point).
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/unnamed-chunk-25-1.pdf}
-
-In drier watersheds, changes in forest cover have greater impact on
-flow, which is similar to Zhang et al. (2017). This is most likely
-because in these watersheds the overall flow is surface flow dominated
-and therefore the buffering that is afforded by the groundwater inputs
-is not as great. As we don't have a separate variable for groundwater
-inputs (although this effect is estimated in many studies), we cannot
-analyse this effect separately.
-
-In contrast to Filoso et al. (2017), we also did not identify an effect
-of the
-
-Given how skewed Dryness is due tot he few watersheds that have very
-high dryness values, it is worth investigating what excluding these 4
-watersheds from the data means for the relationships.
-
-\newpage
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.36\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--12.59\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-18.25\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.69\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.49\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.28\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-2.71\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--23.5\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-6.91\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--3.4\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Precip\_data\_typeOB}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--11.87\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-14.16\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.84\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.4\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Precip\_data\_typeSG}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-16.94\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-16.33\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.04\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.3\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueEA, HM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--8.27\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-3.51\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--2.36\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueHM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-28.02\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-12.3\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-2.28\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.02\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniquePWE}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-39.24\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-12.08\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.25\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniquePWE, HM}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--14.48\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-5.56\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--2.6\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueQPW}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-33.8\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-20.78\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.63\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.11\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueQPW, EA}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-35.58\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-25.22\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.41\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.16\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Assessment\_techniqueSH}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-42.81\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-12.71\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.37\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_typeCF}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--3.86\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-8.17\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.47\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.64\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Forest\_typeMF}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
--4.02\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-8.65\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.46\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.64\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.36\columnwidth}\centering
-\textbf{Hydrological\_regimeSD}\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-6.48\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
-10.06\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-0.64\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.52\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-edf\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-Ref.df\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-F\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-p-value\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(Dryness)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-3.59\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-1.67\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(log10(Area\_km2))}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.71\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.29\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.05\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(length)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-1\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-1.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-2.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model7_noLatb-3.pdf}
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between log10(Area) and the change in flow is essentially linear and does not need to be smoothed (this is the value of using penalized smooths following Wood (2006)).
+It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
+In addition, both the length and Dryness variables show non-linearity, but this does not show a clear trend.
+
+Quitting from lines 645-646 (Forest\_and\_Water.Rmd)
+Error in summary(model7\_nolongLength) :
+object `model7\_nolongLength' not found
+Calls: \ldots{} withCallingHandlers -\textgreater{} withVisible -\textgreater{} eval -\textgreater{} eval -\textgreater{} pander -\textgreater{} summary
+In addition: Warning messages:
+1: In eval(expr, envir, enclos) : NAs introduced by coercion
+2: In eval(expr, envir, enclos) : NAs introduced by coercion
+
+The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
+The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) also does not reflect the type of response highlighted in Filoso et al.
+(2017) and Jackson et al.
+(2005).
+One reason could be the few data points with very long data series, but highly variable responses (Figure \ref{fig:smoothsmodel6}).
+Reducing the flexibility of the splines, removing any studies longer than 60 years (Table \ref{tab:restrictlength}), or fitting a linear term results in ``length'' not being significant, suggesting that this variable is not helpful in explaining the variation in the data.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Zhang\_all2 }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(length)) }\SpecialCharTok{+} \FunctionTok{geom\_histogram}\NormalTok{(}\AttributeTok{fill=}\StringTok{"blue"}\NormalTok{, }\AttributeTok{bins =}\DecValTok{50}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Essentially, the analysis shows at the moment that in contrast to Zhang et al.
+(2017) there is no evidence that the size of a watershed influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds.
+In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
+
+As shown earlier, most of the smaller watersheds are ``real observed data'' using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+This is a confounding factor that is not easily addressed in the regression modelling attempted here.
+Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
 
 \hypertarget{remove-the-assessment-techniques-with-very-small-numbers}{%
-\subsubsection{remove the assessment techniques with very small
-numbers}\label{remove-the-assessment-techniques-with-very-small-numbers}}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.34\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Estimate\strut
-\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
-Std. Error\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-t value\strut
-\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
-Pr(\textgreater\textbar t\textbar)\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{(Intercept)}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--18.71\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-19.02\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.98\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.33\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{DeltaF\_perc\_pos}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.29\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-0.1\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-2.83\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Forest\_Signincrease}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--23.26\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-7.15\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--3.25\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Precip\_data\_typeOB}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--11.25\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-14.19\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.79\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.43\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Precip\_data\_typeSG}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-18.43\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-16.43\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.12\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.26\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Assessment\_techniqueHM}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-32.02\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-12.76\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-2.51\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.01\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Assessment\_techniquePWE}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-44.48\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-12.97\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.43\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Assessment\_techniqueQPW}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-38.12\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-21.21\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-1.8\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.07\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Assessment\_techniqueSH}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-46.79\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-13.17\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-3.55\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Forest\_typeCF}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--3.3\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-8.23\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.4\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.69\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Forest\_typeMF}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
--3.88\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-8.69\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
--0.45\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.66\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.34\columnwidth}\centering
-\textbf{Hydrological\_regimeSD}\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-5.73\strut
-\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
-10.11\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-0.57\strut
-\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
-0.57\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}ccccc@{}}
-\toprule
-\begin{minipage}[b]{0.30\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-edf\strut
-\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
-Ref.df\strut
-\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
-F\strut
-\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
-p-value\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(Dryness)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-3.51\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-1.96\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(log10(Area\_km2))}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.7\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0.24\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.08\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.30\columnwidth}\centering
-\textbf{s(length)}\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
-9\strut
-\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
-0\strut
-\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
-0.99\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-1.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-2.pdf}
-\includegraphics{Forest_and_Water_files/figure-latex/model8_reduc_assess-3.pdf}
+\subsubsection{remove the assessment techniques with very small numbers}\label{remove-the-assessment-techniques-with-very-small-numbers}}
+
+One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.32}}
+  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.08}}@{}}
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+Assessment\_technique
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+n
+\end{minipage} \\
+\midrule
+\endhead
+PWE & 185 \\
+HM & 53 \\
+EA & 32 \\
+SH & 26 \\
+QPW & 7 \\
+QPW, EA & 4 \\
+EA, HM & 1 \\
+PWE, HM & 1 \\
+\bottomrule
+\end{longtable}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Longitude, Latitude, }\DataTypeTok{colour =}\NormalTok{ DeltaF_perc, }\DataTypeTok{size =}\NormalTok{ DeltaQf_perc}\OperatorTok{/}\DecValTok{100}\NormalTok{ )) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.5}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(Area_km2)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_x_log10}\NormalTok{()}
-\NormalTok{total <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(Zhang_all)}
-\KeywordTok{length}\NormalTok{(Zhang_all}\OperatorTok{$}\NormalTok{Area_km2[Zhang_all}\OperatorTok{$}\NormalTok{Area_km2}\OperatorTok{<}\DecValTok{10}\NormalTok{])}\OperatorTok{/}\NormalTok{total}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang_all2 }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(length)) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom_histogram}\NormalTok{(}\DataTypeTok{fill=}\StringTok{"blue"}\NormalTok{, }\DataTypeTok{bins =}\DecValTok{50}\NormalTok{) }
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Essentially, the analysis shows at the moment that in contrast to Zhang
-et al. (2017) there is no evidence that the size of a watershed
-influences the change in the streamflow as a result of changes in
-forestry. If anything the scatter in the data (in the change in flow) is
-greater for the smaller watersheds then for the larger watersheds. In
-other words, the response to changes in forest cover is more consistent
-for larger watersheds than it is for smaller watersheds.
-
-As shown earlier, most of the smaller watersheds are ``real observed
-data'' using paired watershed studies, while for larger watersheds, the
-analysis are mostly based on modelling approximations using either
-elasticity analysis (EA), Hydrological modelling (HM) or a combined use
-of statistical methods (SH) or quasi paired watershed analysis (QPW),
-thus all providing an approximation of the effect of forestry on
-streamflow rather than a direct comparison of watersheds. This is a
-confounding factor that is not easily addressed in the regression
-modelling attempted here. Furthermore, the catchments analysed using EA,
-are concentrated in the drier end of the Dryness index scale compared to
-the other methods, with only the paired watershed experiment (PWE)
-assessment technique covering the full range of dryness indices.
-
-There are further confounding factors in the data, which were also
-classified by Filoso et al. (2017) and these create biases in the data
-set that can impact the overall assessment. For example, snow dominated
-hydrological regimes (SD), which are weakly significant, are dominated
-by Coniferous Forests (CF), while the majority of the rain dominated
-regimes are all broadleaf forests (BF). However, the forest type
-classification is very coarse and does not fully capture possible
-physiological differences that could affect evapotranspiration and
-therefore changes in streamflow.
-
-Apart from a difficulty of analysing complex confounding factors in the
-data, a general limitation of the type of analysis presented is that
-this work does not consider the spatial arrangement of the forest
-clearing in the catchments. While for fully or almost fully cleared
-smaller catchments this might not be an issue, it is perceivable that
-for larger catchments being partially cleared, a interaction between
-spatial location and clearing could be a factor in determining the
-change in streamflow. Clearing head water catchments on shallower soils
-might have a larger impact than clearing in downstream areas on deeper
-soils.
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments}\label{fig:assessment}
+\end{figure}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.40}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & -18.71 & 19.02 & -0.98 & 0.33 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.83 & 0 \\
+\textbf{Forest\_Signincrease} & -23.26 & 7.15 & -3.25 & 0 \\
+\textbf{Precip\_data\_typeOB} & -11.25 & 14.19 & -0.79 & 0.43 \\
+\textbf{Precip\_data\_typeSG} & 18.43 & 16.43 & 1.12 & 0.26 \\
+\textbf{Assessment\_techniqueHM} & 32.02 & 12.76 & 2.51 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 44.48 & 12.97 & 3.43 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 38.12 & 21.21 & 1.8 & 0.07 \\
+\textbf{Assessment\_techniqueSH} & 46.79 & 13.17 & 3.55 & 0 \\
+\textbf{Forest\_typeCF} & -3.3 & 8.23 & -0.4 & 0.69 \\
+\textbf{Forest\_typeMF} & -3.88 & 8.69 & -0.45 & 0.66 \\
+\textbf{Hydrological\_regimeSD} & 5.73 & 10.11 & 0.57 & 0.57 \\
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(Dryness)} & 3.51 & 9 & 1.96 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.7 & 9 & 0.24 & 0.08 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.99 \\
+\bottomrule
+\end{longtable}
+
+In drier watersheds, changes in forest cover have greater impact on flow, which is similar to Zhang et al.
+(2017).
+This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
+
+In contrast to Filoso et al. (2017), we also did not identify an effect of the
+
+Given how skewed Dryness is due to the few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
+
+\hypertarget{other-possible-bias-in-the-data}{%
+\subsubsection{Other possible bias in the data}\label{other-possible-bias-in-the-data}}
+
+There are further confounding factors in the data, which were also classified by Filoso et al.
+(2017) and these create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow.
+
+Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
+While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
+Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-andreassian2004}{}%
-Andréassian, V., 2004. Waters and forests: From historical controversy
-to scientific debate. Journal of Hydrology 291, 1--27.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2003.12.015}{https://doi.org/10.1016/j.jhydrol.2003.12.015}
-
-\leavevmode\hypertarget{ref-borg1988}{}%
-Borg, H., Bell, R.W., Loh, I.C., 1988. Streamflow and stream salinity in
-a small water supply catchment in southwest western australia after
-reforestation. Journal of Hydrology 103, 323--333.
-doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(88)90141-2}{https://doi.org/10.1016/0022-1694(88)90141-2}
-
-\leavevmode\hypertarget{ref-hewlett1984}{}%
-Bosch, J.M., Hewlett, J.D., 1982. A review of catchment experiments to
-determine the effect of vegetation changes on water yield and
-evapotranspiration. Journal of Hydrology 55, 3--23.
-
-\leavevmode\hypertarget{ref-brown2013}{}%
-Brown, A.E., Western, A.W., McMahon, T.A., Zhang, L., 2013. Impact of
-forest cover changes on annual streamflow and flow duration curves.
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-andreassian2004}{}}%
+Andréassian, V., 2004. Waters and forests: From historical controversy to scientific debate. Journal of Hydrology 291, 1--27. doi:\url{https://doi.org/10.1016/j.jhydrol.2003.12.015}
+
+\leavevmode\vadjust pre{\hypertarget{ref-borg1988}{}}%
+Borg, H., Bell, R.W., Loh, I.C., 1988.
+Streamflow and stream salinity in a small water supply catchment in southwest western australia after reforestation.
+Journal of Hydrology 103, 323--333.
+doi:\url{https://doi.org/10.1016/0022-1694(88)90141-2}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hewlett1984}{}}%
+Bosch, J.M., Hewlett, J.D., 1982. A review of catchment experiments to determine the effect of vegetation changes on water yield and evapotranspiration. Journal of Hydrology 55, 3--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-brown2013}{}}%
+Brown, A.E., Western, A.W., McMahon, T.A., Zhang, L., 2013.
+Impact of forest cover changes on annual streamflow and flow duration curves.
 Journal of Hydrology 483, 39--50.
-doi:\href{https://doi.org/http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}
-
-\leavevmode\hypertarget{ref-brown2005}{}%
-Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A.,
-2005. A review of paired catchment studies for determining changes in
-water yield resulting from alterations in vegetation. Journal of
-Hydrology 310, 28--61.
-
-\leavevmode\hypertarget{ref-cosandey2005}{}%
-Cosandey, C., Andréassian, V., Martin, C., Didon-Lescot, J.F., Lavabre,
-J., Folton, N., Mathys, N., Richard, D., 2005. The hydrological impact
-of the mediterranean forest: A review of french research. Journal of
-Hydrology 301, 235--249.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2004.06.040}{https://doi.org/10.1016/j.jhydrol.2004.06.040}
-
-\leavevmode\hypertarget{ref-filoso2017}{}%
-Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017. Impacts of
-forest restoration on water yield: A systematic review. PLOS ONE 12,
-e0183210.
+doi:\url{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}
+
+\leavevmode\vadjust pre{\hypertarget{ref-brown2005}{}}%
+Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A., 2005.
+A review of paired catchment studies for determining changes in water yield resulting from alterations in vegetation.
+Journal of Hydrology 310, 28--61.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cosandey2005}{}}%
+Cosandey, C., Andréassian, V., Martin, C., Didon-Lescot, J.F., Lavabre, J., Folton, N., Mathys, N., Richard, D., 2005.
+The hydrological impact of the mediterranean forest: A review of french research.
+Journal of Hydrology 301, 235--249.
+doi:\url{https://doi.org/10.1016/j.jhydrol.2004.06.040}
+
+\leavevmode\vadjust pre{\hypertarget{ref-filoso2017}{}}%
+Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017.
+Impacts of forest restoration on water yield: A systematic review.
+PLOS ONE 12, e0183210.
 doi:\href{https://doi.org/10.1371/journal.pone.0183210}{10.1371/journal.pone.0183210}
 
-\leavevmode\hypertarget{ref-jackson2005}{}%
-Jackson, R.B., Jobbagy, E.G., Avissar, R., Roy, S.B., Barrett, D.J.,
-Cook, C.W., Farley, K.A., Maitre, D.C. le, McCarl, B.A., Murray, B.C.,
-2005. Trading water for carbon with biological carbon sequestration.
+\leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
+Jackson, R.B., Jobbagy, E.G., Avissar, R., Roy, S.B., Barrett, D.J., Cook, C.W., Farley, K.A., Maitre, D.C.
+le, McCarl, B.A., Murray, B.C., 2005.
+Trading water for carbon with biological carbon sequestration.
 Science 310, 1944--1947.
@@
 
-\leavevmode\hypertarget{ref-kuczera1987}{}%
-Kuczera, G., 1987. Prediction of water yield reductions following a
-bushfire in ash-mixed species eucalypt forest. Journal of Hydrology 94,
-215--236.
+\leavevmode\vadjust pre{\hypertarget{ref-kuczera1987}{}}%
+Kuczera, G., 1987.
+Prediction of water yield reductions following a bushfire in ash-mixed species eucalypt forest.
+Journal of Hydrology 94, 215--236.
 doi:\href{https://doi.org/Doi:\%2010.1016/0022-1694(87)90054-0}{Doi: 10.1016/0022-1694(87)90054-0}
 
-\leavevmode\hypertarget{ref-pena-arancibia2012}{}%
-Peña-Arancibia, J.L., Dijk, A.I.J.M. van, Guerschman, J.P., Mulligan,
-M., Bruijnzeel, L.A., McVicar, T.R., 2012. Detecting changes in
-streamflow after partial woodland clearing in two large catchments in
-the seasonal tropics. Journal of Hydrology 416-417, 60--71.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2011.11.036}{https://doi.org/10.1016/j.jhydrol.2011.11.036}
-
-\leavevmode\hypertarget{ref-roche1981}{}%
-Roche, M., 1981. Watershed investigations for development of forest
-resources of the amazon region in french guyana. Tropical Agricultural
-Hydrology. J 75--82.
-
-\leavevmode\hypertarget{ref-rodriguez2010}{}%
-Rodriguez, D.A., Tomasella, J., Linhares, C., 2010. Is the forest
-conversion to pasture affecting the hydrological response of amazonian
-catchments? Signals in the ji-paraná basin. Hydrological Processes 24,
-1254--1269.
-doi:\href{https://doi.org/https://doi.org/10.1002/hyp.7586}{https://doi.org/10.1002/hyp.7586}
-
-\leavevmode\hypertarget{ref-ruprechtetal1991}{}%
-Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A.,
-Stoneman, G.L., 1991. Early hydrological response to intense forest
-thinning in southwestern australia. Journal of Hydrology 127, 261--277.
-doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(91)90118-2}{https://doi.org/10.1016/0022-1694(91)90118-2}
-
-\leavevmode\hypertarget{ref-thornton2007}{}%
-Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007. The
-brigalow catchment study: II*. Clearing brigalow (acacia harpophylla)
-for cropping or pasture increases runoff. Australian Journal of Soil
-Research 45, 496--511.
+\leavevmode\vadjust pre{\hypertarget{ref-pena-arancibia2012}{}}%
+Peña-Arancibia, J.L., Dijk, A.I.J.M.
+van, Guerschman, J.P., Mulligan, M., Bruijnzeel, L.A., McVicar, T.R., 2012.
+Detecting changes in streamflow after partial woodland clearing in two large catchments in the seasonal tropics.
+Journal of Hydrology 416-417, 60--71.
+doi:\url{https://doi.org/10.1016/j.jhydrol.2011.11.036}
+
+\leavevmode\vadjust pre{\hypertarget{ref-roche1981}{}}%
+Roche, M., 1981. Watershed investigations for development of forest resources of the amazon region in french guyana. Tropical Agricultural Hydrology. J 75--82.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rodriguez2010}{}}%
+Rodriguez, D.A., Tomasella, J., Linhares, C., 2010.
+Is the forest conversion to pasture affecting the hydrological response of amazonian catchments?
+Signals in the ji-paraná basin.
+Hydrological Processes 24, 1254--1269.
+doi:\url{https://doi.org/10.1002/hyp.7586}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ruprechtetal1991}{}}%
+Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A., Stoneman, G.L., 1991.
+Early hydrological response to intense forest thinning in southwestern australia.
+Journal of Hydrology 127, 261--277.
+doi:\url{https://doi.org/10.1016/0022-1694(91)90118-2}
+
+\leavevmode\vadjust pre{\hypertarget{ref-thornton2007}{}}%
+Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007.
+The brigalow catchment study: II*.
+Clearing brigalow (acacia harpophylla) for cropping or pasture increases runoff.
+Australian Journal of Soil Research 45, 496--511.
 doi:\href{https://doi.org/doi:10.1071/SR07064}{doi:10.1071/SR07064}
 
-\leavevmode\hypertarget{ref-trabucco2018}{}%
-Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential
-evapo-transpiration (et0) climate database v2. CGIAR consortium for
-spatial information(CGIAR-csi).
-
-\leavevmode\hypertarget{ref-wood2006}{}%
-Wood, S., 2006. Generalized additive models: An introduction with r. CRC
-Press, Boca Raton, FL.
-
-\leavevmode\hypertarget{ref-zhang2011}{}%
-Zhang, L., Zhao, F., Chen, Y., Dixon, R.N.M., 2011. Estimating effects
-of plantation expansion and climate variability on streamflow for
-catchments in australia. Water Resources Research 47, W12539.
+\leavevmode\vadjust pre{\hypertarget{ref-trabucco2018}{}}%
+Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential evapo-transpiration (ET0) climate database v2. CGIAR consortium for spatial information(CGIAR-CSI).
+
+\leavevmode\vadjust pre{\hypertarget{ref-wood2006}{}}%
+Wood, S., 2006. Generalized additive models: An introduction with r. CRC Press, Boca Raton, FL.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zhang2011}{}}%
+Zhang, L., Zhao, F., Chen, Y., Dixon, R.N.M., 2011.
+Estimating effects of plantation expansion and climate variability on streamflow for catchments in australia.
+Water Resources Research 47, W12539.
 doi:\href{https://doi.org/10.1029/2011wr010711}{10.1029/2011wr010711}
 
-\leavevmode\hypertarget{ref-zhang2017}{}%
-Zhang, M., Liu, N., Harper, R., Li, Q., Liu, K., Wei, X., Ning, D., Hou,
-Y., Liu, S., 2017. A global review on hydrological responses to forest
-change across multiple spatial scales: Importance of scale, climate,
-forest type and hydrological regime. Journal of Hydrology 546, 44--59.
-doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2016.12.040}{https://doi.org/10.1016/j.jhydrol.2016.12.040}
-
-\leavevmode\hypertarget{ref-zhao2010}{}%
-Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010. Evaluation of methods
-for estimating the effects of vegetation change and climate variability
-on streamflow. Water Resources Research 46, W03505.
+\leavevmode\vadjust pre{\hypertarget{ref-zhang2017}{}}%
+Zhang, M., Liu, N., Harper, R., Li, Q., Liu, K., Wei, X., Ning, D., Hou, Y., Liu, S., 2017.
+A global review on hydrological responses to forest change across multiple spatial scales: Importance of scale, climate, forest type and hydrological regime.
+Journal of Hydrology 546, 44--59.
+doi:\url{https://doi.org/10.1016/j.jhydrol.2016.12.040}
+
+\leavevmode\vadjust pre{\hypertarget{ref-zhao2010}{}}%
+Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010.
+Evaluation of methods for estimating the effects of vegetation change and climate variability on streamflow.
+Water Resources Research 46, W03505.
 doi:\href{https://doi.org/10.1029/2009wr007702}{10.1029/2009wr007702}
 
-\leavevmode\hypertarget{ref-zhou2015}{}%
-Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G.,
-Scott, D.F., Zhou, S., Han, L., Su, Y., 2015. Global pattern for the
-effect of climate and land cover on water yield. Nature Communications
-6, 5918.
+\leavevmode\vadjust pre{\hypertarget{ref-zhou2015}{}}%
+Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G., Scott, D.F., Zhou, S., Han, L., Su, Y., 2015.
+Global pattern for the effect of climate and land cover on water yield.
+Nature Communications 6, 5918.
 doi:\href{https://doi.org/10.1038/ncomms6918}{10.1038/ncomms6918}
 
-\leavevmode\hypertarget{ref-zhou2010}{}%
-Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang,
-C., 2010. Forest recovery and river discharge at the regional scale of
-guangdong province, china. Water Resources Research 46.
-doi:\href{https://doi.org/https://doi.org/10.1029/2009WR008829}{https://doi.org/10.1029/2009WR008829}
+\leavevmode\vadjust pre{\hypertarget{ref-zhou2010}{}}%
+Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang, C., 2010.
+Forest recovery and river discharge at the regional scale of guangdong province, china.
+Water Resources Research 46.
+doi:\url{https://doi.org/10.1029/2009WR008829}
+
+\end{CSLReferences}
 

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.Rmd
updates on paper 12012022

@@
 ---
-title: Do larger watersheds respond different to forest cover change? Re-analysing a global data set.
+title: Factors determining how catchments respond to forest cover change: Re-analysing global data sets.
 author:
@@
   - code: INIA
-    address: Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
+    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
   - code: IMFIA
@@
 abstract: |
-  This is the abstract.
-  It consists of two paragraphs.
+  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
+However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
+The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+Generalised additive models were used to run flexible regressions including multiple variables.
+  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
+Deforestation causes a 32\% greater change in flow compared to reforestation.
+Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies.
+Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2.
+Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
+Despite these findings, overall explained variance (38/%) is low due the quality of the inputs and additional unknown confounding factors.
 journal: "Journal of Hydrology"
@@
 The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian watersheds) a 100% decrease in streamflow for watersheds with 100% forest cover.
-However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
+The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
+However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchmensts after deforestation.
 
@@
 
-Several review papers have summarized different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
+Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
 These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover.
-The @zhang2017 dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover.
+The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
 In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+There is some overlap between the two data sets, but there are also some studies unique to both sets.
 
@@
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of quantitative data (37) they showed a relationship between watershed size and decline in streamflow.
+For a subset of quantitative data (37) they showed a log-linear relationship between decreasing watershed size and an increasing decline in streamflow.
+In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
 
@@
 
-Encouraged by the work presented by @zhang2017 and @filoso2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion.
-In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.
+Encouraged by the work presented by @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
+In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.
 
-In particular, the main method in the work by @zhang2017 is using simple linear regression, and in @filoso2017 the focus is mainly on classification.
-As @zhang2017 points out, the main assumption in their work is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds, but the subset of data in @filoso2017 does not appear to support this.
-And while te work @filoso2017 provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact.
-This is because the work had a strict criterion to select quantitative studies.
-However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.
+In particular, the main method in the work by @zhang2017 is a single covariate linear regression, and in @filoso2017 the focus is mainly on classification and there is again some single covariate linear regression.
+As @zhang2017 points out, a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
+However, the subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum.
+And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.
 
-As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
+As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further catchments (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
 
 Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
-@jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
+However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
+@jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
 
@@
 While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
-We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watershed data sets in our analysis.
+We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) catchment data sets in our analysis.
 
@@
 This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for watershed #76 was corrected to 157% based on the original publication [@baker1984].
-Finally, on review of all the data in @zhang2017, 29 potential duplicates were identified and flagged in the data.
+Furthermore, the change in flow for watershed #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
+Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.
 
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}$$
@@
 The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` observations.
+In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` watersheds and the @filoso2017 study used 37 watersheds (Table S2 in @filoso2017).
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
@@
       "DeltaF_perc" = "Delta Forest (%)",
-      "DeltaQf_perc" = "Delta Flow (%)",
+      "DeltaQf_perc" = "Delta Qf (%)",
       "E0" = "Annual Potential E (mm)",
@@
 
-(ref:datagraphscaption) foo
+(ref:datagraphscaption) Overview of the distribution of the data set for five of the included variables.
 
@@
                       values = c("increase" = "blue", "decrease" = "red")) +
-  xlab("Change in forest cover %") + ylab("Change in flow %")
+  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
 dev.off()
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
 \end{aligned}$$
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}$$
@@
 This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005).
-In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.
+In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (*Dryness, Area, From*) were considered non-linear in this model.
+This is also indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.
+*From* represents the starting date of the data collection.
+
+$$\tag{5}
+\begin{aligned}
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
+\end{aligned}$$
+
+<!--
+```{r model4b, echo = F, eval = F}
+All_data2 <- All_data2 %>%
+  mutate(length = To - From)
+
+model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
+                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
+                 s(length, bs = "ts"),
+                  data = All_data2 %>% filter(length < 60))
+summary(model4b)
+#gam.check(model5_all)
+#plot(model5_all)
+```
+-->
 
@@
                     s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
-                    s(length, bs="ts") ,
+                    s(length, bs="ts") + s(From, bs = "ts") ,
                   data = All_data2)
@@
 
-```{r mfive-linear, echo = F}
-pander(round(summary(model5_all)$p.table,2), caption = "Statistical summary for the linear terms in the model with non-linear terms")
+```{r , echo = F}
+pander(round(summary(model5_all)$p.table,2), caption = "(\\#tab:mfive-linear) Statistical summary for the linear terms in the model with non-linear terms")
 ```
 
-```{r mfive-smooth, echo = F}
-pander(round(summary(model5_all)$s.table,2), caption = "Statistical summary for the smooth terms in the model with non-linear terms")
+```{r , echo = F}
+pander(round(summary(model5_all)$s.table,2), caption = "(\\#tab:mfive-smooth) Statistical summary for the smooth terms in the model with non-linear terms")
 
 ```
-
-<!-- model 2c include smooth variables -->
 
@@
 
-A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.
+A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).
 
@@
                     s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts" ) + 
-                     s(length, bs="ts") +
+                     s(length, bs="ts") + s(From, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
@@
 
-```{r msix-linear, echo = F}
-pander(round(summary(model6_all)$p.table,2), caption = "Statistical summary for the linear terms the full model")
+```{r, echo = F}
+pander(round(summary(model6_all)$p.table,2), caption = "(\\#tab:msix-linear) Statistical summary for the linear terms the full model")
 #plot(model6_all)
@@
 
-```{r msix-smooth}
-pander(round(summary(model6_all)$s.table,2), caption = "Statistical summary for the smooth terms for the full model")
+```{r , echo = F}
+pander(round(summary(model6_all)$s.table,2), caption = "(\\#tab:msix-smooth) Statistical summary for the smooth terms for the full model")
 #plot(model6_all)
@@
 D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
+A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
 L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
-A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-D / (L + A) 
+From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
+(D + A) / (L + From) 
 # visreg(model6_all,"Dryness",gg=T) + theme_bw()
@@
 
-Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area) and the change in flow is essentially linear and does not need to be smoothed (this is the value of using penalized smooths following @wood2006).
+Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006).
 It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
-In addition, both the length and Dryness variables show non-linearity, but this does not show a clear trend.
+Similarly, the length variable is not significant.
+However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
 
@@
 <!--filter lenght < 60 -->
-```{r model7_noLatb, echo = F, eval = F}
-model7_nolongLength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+```{r model7_nolongLength, echo = F}
+NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
+model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                     s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
+                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data2 %>% filter(length< 60))
-#summary(model7_noLatb)
-#plot(model7_noLatb)
-#pander(round(summary(model7_nolongLength)$p.table,2))
-#plot(model7_noLatb)
-#summary(model7_nolongLength)
+                    Hydrological_regime, data = NoLong_Data)
+#summary(model7)
 ```
-```{r restrictlength, echo = F}
-pander(round(summary(model7_nolongLength)$s.table,2), caption = "Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
+```{r, echo = F}
+pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
+```
+```{r visualisesmooths7,echo =F, eval = F}
+#require(visreg)
+png("model7_smooths.png", width = 2500, height = 1800, res = 300)
+#par(mfrow=c(2,2))
+D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
+A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
+L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
+From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
+(D + A) / (L + From) 
+# visreg(model6_all,"Dryness",gg=T) + theme_bw()
+# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
+# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
+#par(mfrow=c(1,1))
+dev.off()
+```
+```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}
+
+include_graphics("model7_smooths.png")
 ```
@@
 The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005.
-One reason could be the few data points with very long data series, but highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Reducing the flexibility of the splines, removing any studies longer than 60 years (Table \@ref(tab:restrictlength)), or fitting a linear term results in "length" not being significant, suggesting that this variable is not helpful in explaining the variation in the data.
+One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)).
+Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness > 4 (Table \@ref(tab:restrictlength)) result in log10(Area) showing a significant effect with larger catchments showing less impact of changes in forest cover.
+Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
+Finally, more recent studies also show greater impacts of changes in forest cover (Figure \@ref(fig:smoothsmodel7)).
 
@@
 ```
-```{r, eval =F}
+```{r, eval =F, echo = F}
 Zhang_all2 %>%
@@
 
+```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
+
+include_graphics("flow_forest_byArea.png")
+
+```
+
 Discussion
@@
 
-Essentially, the analysis shows at the moment that in contrast to @zhang2017 there is no evidence that the size of a watershed influences the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds.
+Essentially, the analysis shows that in contrast to @zhang2017 there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds (Figure \@ref(fig:overview)).
 In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
 
-As shown earlier, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+However, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 
-### remove the assessment techniques with very small numbers
+## The effect of assessment techniques with very small numbers of observations
 
@@
 This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).
+The following model removes these from the data set.
 
@@
 
-pander(Table_assessment)
+pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
 ```
 
-```{r, eval = T, echo =F}
+```{r, eval = F, echo =F}
 png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
@@
 
-```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F}
+```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F, eval =F}
 
@@
                     s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
+                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
                     Hydrological_regime, data = All_data3 %>% filter(length< 60))
-#summary(model7_noLatb)
-#plot(model7_noLatb)
-pander(round(summary(model8_assess)$p.table,2))
-pander(round(summary(model8_assess)$s.table,2))
 #plot(model8_assess)
@@
 ```
+```{r, echo = F}
+pander(round(summary(model8_assess)$p.table,2), caption= "(\\#tab:model8-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
+pander(round(summary(model8_assess)$s.table,2), caption = "(\\#tab:model8-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
 
-In drier watersheds, changes in forest cover have greater impact on flow, which is similar to @zhang2017.
+```
+
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results.
+It strengthens the significance of the assessment technqiues but slightly reduces the importance of some of the smooth variables (*log10(Area (km^2^))* and *From*, Table \@ref(tab:modl8-linear) and \@ref(tab:model8-smooth)).
+Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes
+
+In drier watersheds, changes in forest cover have greater impact on flow, which is similar to the observations from @zhang2017.
 This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
-As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
+As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect cannot be analysed separately.
 
-In contrast to @filoso2017, we also did not identify an effect of the
+In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
+What is puzzling is that the effect of the start date (*From*) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
+At this point, it is unclear what this trend signifies.
+It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
 
-Given how skewed Dryness is due to the few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
+Excluding the few watersheds that have very high dryness values, clarified the *Dryness* trend, agreeing with earlier studies.
 
-### Other possible bias in the data
+## Other possible bias in the data
 
-There are further confounding factors in the data, which were also classified by @filoso2017 and these create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow.
+There are further confounding factors in the data, which were also classified by @filoso2017 and these might create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
 
@@
 
+Conclusions
+============
+
 References {#references .unnumbered}

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.tex
updates on paper 12012022

@@
 
-% Pandoc syntax highlighting
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-
 % tightlist command for lists without linebreak
@@
    \ead{willem.vervoort@sydney.edu.au} 
-    \author[INIA]{Eliana Nervi\corref{2}}
+    \author[INIA]{Eliana Nervi}
    \ead{eliananervi@gmail.com} 
-    \author[IMFIA]{Jimena Alonso\corref{2}}
+    \author[IMFIA]{Jimena Alonso}
    \ead{jalonso@fing.edu.uy} 
@@
     \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
-    \address[INIA]{Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
     \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
       \cortext[1]{Corresponding Author}
-    \cortext[2]{Equal contribution}
 
   \begin{abstract}
-  This is the abstract.
-  It consists of two paragraphs.
+  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
+However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
+The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+Generalised additive models were used to run flexible regressions including multiple variables.
+  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
+Deforestation causes a 32\% greater change in flow compared to reforestation.
+Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies.
+Smaller studies dominate the database with 42\% of the data \textless{} 1 km2 and 65\% of the data \textless{} 10 km2.
+Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
+Despite these findings, overall explained variance (38/\%) is low due the quality of the inputs and additional unknown confounding factors.
   \end{abstract}
@@
 \section{Introduction}\label{introduction}}
-
-\hypertarget{introduction-1}{%
-\subsection{Introduction}\label{introduction-1}}
 
@@
 The most dramatic result in relation to this, is Figure 5 in Zhang et al.
-(2011) indicating (for Australian watersheds) a 100\% decrease in streamflow for watersheds with 100\% forest cover.
-However, on the other end of the spectrum, in a series of French watersheds (Cosandey et al., 2005), there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.
-
-Several review papers have summarized different studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017).
+(2011) indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
+However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchmensts after deforestation.
+
+Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017).
 These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover.
+The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover.
 The Zhang et al.
-(2017) dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+(2017) dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by Filoso et al.
@@
 In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+There is some overlap between the two data sets, but there are also some studies unique to both sets.
 
@@
 Encouraged by the work presented by Zhang et al.
-(2017) and Filoso et al.
-(2017) and the fantastic database of studies presented by these authors, we believe we can add to the discussion.
-In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.
+(2017), Filoso et al.
+(2017) and Zhou et al.
+(2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
+In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.
 
 In particular, the main method in the work by Zhang et al.
-(2017) is using simple linear regression, and in Filoso et al.
-(2017) the focus is mainly on classification.
+(2017) is a single covariate linear regression, and in Filoso et al.
+(2017) the focus is mainly on classification and there is again some single covariate linear regression.
 As Zhang et al.
-(2017) points out, the main assumption in their work is that the threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' watersheds, but the subset of data in Filoso et al.
-(2017) does not appear to support this.
-And while te work Filoso et al.
-(2017) provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact.
-This is because the work had a strict criterion to select quantitative studies.
-However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km\textsuperscript{2}.
+(2017) points out, a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments.
+However, the subset of 37 data points in Filoso et al.
+(2017) (their Figure 9) does not appear to support this, suggesting a continuum.
+And while the work Filoso et al.
+(2017) provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.
 
 As a result the objective of this paper is to 1) enhance the data set from Zhang et al.
-(2017) with further watersheds (such as from Filoso et al.
-(2017)) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).
+(2017) with further catchments (such as from Filoso et al.
+(2017)) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).
 
@@
 (2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths Zhou et al.
+However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths Zhou et al.
 (2015)), the length of the study (c.f.
 Jackson et al.
-(2005)) and possibly the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al.
+(2005); Filoso et al.
+(2017)) and the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al.
 (2017); Zhou et al.
@@
 While Zhang et al. (2017) use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
-We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) watershed data sets in our analysis.
+We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) catchment data sets in our analysis.
 
@@
 This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
+Furthermore, the change in flow for watershed \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984).
+Finally, on review of all the data in Zhang et al.
+(2017) and Filoso et al.
+(2017), 29 potential duplicates were identified and flagged in the data.
 
@@
 
-Removing possible duplicates removes 29 watersheds from the overall data set
-
 To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) (Wood, 2006).
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
 \end{aligned}\]
@@
 In contrast, the original dataset from Zhang et al.
-(2017) contained 312 observations.
+(2017) contained 312 watersheds and the Filoso et al.
+(2017) study used 37 watersheds (Table S2 in Filoso et al.
+(2017)).
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{foo}\label{fig:datagraphs}
+\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables.}\label{fig:datagraphs}
 \end{figure}
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
 \end{aligned}\]
@@
 \begin{aligned}
-\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
 \end{aligned}\]
@@
 (2005)).
-In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
+In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (\emph{Dryness, Area, From}) were considered non-linear in this model.
+This is also indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
+\emph{From} represents the starting date of the data collection.
+
+\[\tag{5}
+\begin{aligned}
+\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
+\end{aligned}\]
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 17.09 & 6.08 & 2.81 & 0.01 \\
+\caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 17.37 & 5.91 & 2.94 & 0 \\
 \textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
-\textbf{Forest\_Signincrease} & -34.6 & 6.11 & -5.67 & 0 \\
+\textbf{Forest\_Signincrease} & -34.93 & 5.84 & -5.98 & 0 \\
 \bottomrule
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
+\caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 1.74 & 9 & 0.78 & 0.01 \\
-\textbf{s(Dryness)} & 0.87 & 9 & 0.57 & 0.02 \\
-\textbf{s(length)} & 8.83 & 9 & 3.1 & 0 \\
-\bottomrule
-\end{longtable}
-
-Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.29 and deviance explained of 0.32, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+\textbf{s(log10(Area\_km2))} & 2.48 & 9 & 0.55 & 0.1 \\
+\textbf{s(Dryness)} & 0.89 & 9 & 0.62 & 0.01 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.66 \\
+\textbf{s(From)} & 8.1 & 9 & 3.97 & 0 \\
+\bottomrule
+\end{longtable}
+
+Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.31 and deviance explained of 0.34, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
 
-A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.
+A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{Statistical summary for the linear terms the full model}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & -27.3 & 19.84 & -1.38 & 0.17 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 3.01 & 0 \\
-\textbf{Forest\_Signincrease} & -21.54 & 7.29 & -2.96 & 0 \\
-\textbf{Precip\_data\_typeOB} & -8.13 & 15.17 & -0.54 & 0.59 \\
-\textbf{Precip\_data\_typeSG} & 17.33 & 17.29 & 1 & 0.32 \\
-\textbf{Assessment\_techniqueEA, HM} & 13.93 & 46.74 & 0.3 & 0.77 \\
-\textbf{Assessment\_techniqueHM} & 32.96 & 13.92 & 2.37 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 53.24 & 12.81 & 4.16 & 0 \\
+\caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & -21.06 & 19.16 & -1.1 & 0.27 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 2.99 & 0 \\
+\textbf{Forest\_Signincrease} & -23.11 & 7.13 & -3.24 & 0 \\
+\textbf{Precip\_data\_typeOB} & -7.65 & 14.51 & -0.53 & 0.6 \\
+\textbf{Precip\_data\_typeSG} & 18.92 & 16.17 & 1.17 & 0.24 \\
+\textbf{Assessment\_techniqueEA, HM} & 17.73 & 45.71 & 0.39 & 0.7 \\
+\textbf{Assessment\_techniqueHM} & 29.82 & 12.66 & 2.36 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 44.69 & 12.68 & 3.53 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 33.54 & 47.07 & 0.71 & 0.48 \\
-\textbf{Assessment\_techniqueQPW} & 39.8 & 22.06 & 1.8 & 0.07 \\
+HM} & 39.72 & 46.79 & 0.85 & 0.4 \\
+\textbf{Assessment\_techniqueQPW} & 37.54 & 21.54 & 1.74 & 0.08 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 45.97 & 26.25 & 1.75 & 0.08 \\
-\textbf{Assessment\_techniqueSH} & 44.89 & 13.49 & 3.33 & 0 \\
-\textbf{Forest\_typeCF} & -2.2 & 8.52 & -0.26 & 0.8 \\
-\textbf{Forest\_typeMF} & -6.56 & 8.96 & -0.73 & 0.46 \\
-\textbf{Hydrological\_regimeSD} & 4.59 & 10.25 & 0.45 & 0.65 \\
-\bottomrule
-\end{longtable}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{pander}\NormalTok{(}\FunctionTok{round}\NormalTok{(}\FunctionTok{summary}\NormalTok{(model6\_all)}\SpecialCharTok{$}\NormalTok{s.table,}\DecValTok{2}\NormalTok{), }\AttributeTok{caption =} \StringTok{"Statistical summary for the smooth terms for the full model"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+EA} & 24.9 & 27.45 & 0.91 & 0.37 \\
+\textbf{Assessment\_techniqueSH} & 43.68 & 13.27 & 3.29 & 0 \\
+\textbf{Forest\_typeCF} & -4.06 & 8.71 & -0.47 & 0.64 \\
+\textbf{Forest\_typeMF} & -8.33 & 8.7 & -0.96 & 0.34 \\
+\textbf{Hydrological\_regimeSD} & 9.41 & 10.57 & 0.89 & 0.37 \\
+\bottomrule
+\end{longtable}
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{Statistical summary for the smooth terms for the full model}\tabularnewline
+\caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.31 & 9 & 0.07 & 0.16 \\
-\textbf{s(Dryness)} & 3.45 & 9 & 1.39 & 0.01 \\
-\textbf{s(length)} & 8.73 & 9 & 3.04 & 0 \\
-\bottomrule
-\end{longtable}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#plot(model6\_all)}
-\end{Highlighting}
-\end{Shaded}
-
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.38.
+\textbf{s(log10(Area\_km2))} & 0.44 & 9 & 0.08 & 0.21 \\
+\textbf{s(Dryness)} & 3.5 & 9 & 1.76 & 0 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.45 \\
+\textbf{s(From)} & 8.17 & 9 & 4.13 & 0 \\
+\bottomrule
+\end{longtable}
+
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.4.
 This indicates that the categorical variables explain a limited amount of the variance.
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between log10(Area) and the change in flow is essentially linear and does not need to be smoothed (this is the value of using penalized smooths following Wood (2006)).
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between log10(Area km\textsuperscript{2}) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following Wood (2006)).
 It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
-In addition, both the length and Dryness variables show non-linearity, but this does not show a clear trend.
-
-Quitting from lines 645-646 (Forest\_and\_Water.Rmd)
-Error in summary(model7\_nolongLength) :
-object `model7\_nolongLength' not found
-Calls: \ldots{} withCallingHandlers -\textgreater{} withVisible -\textgreater{} eval -\textgreater{} eval -\textgreater{} pander -\textgreater{} summary
-In addition: Warning messages:
-1: In eval(expr, envir, enclos) : NAs introduced by coercion
-2: In eval(expr, envir, enclos) : NAs introduced by coercion
+Similarly, the length variable is not significant.
+However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(Dryness)} & 2.62 & 9 & 3.4 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.76 & 9 & 0.35 & 0.04 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.4 \\
+\textbf{s(From)} & 0.99 & 9 & 0.46 & 0.04 \\
+\bottomrule
+\end{longtable}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model wiht reduced data for dryness and length}\label{fig:smoothsmodel7}
+\end{figure}
 
@@
 (2005).
-One reason could be the few data points with very long data series, but highly variable responses (Figure \ref{fig:smoothsmodel6}).
-Reducing the flexibility of the splines, removing any studies longer than 60 years (Table \ref{tab:restrictlength}), or fitting a linear term results in ``length'' not being significant, suggesting that this variable is not helpful in explaining the variation in the data.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Zhang\_all2 }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(length)) }\SpecialCharTok{+} \FunctionTok{geom\_histogram}\NormalTok{(}\AttributeTok{fill=}\StringTok{"blue"}\NormalTok{, }\AttributeTok{bins =}\DecValTok{50}\NormalTok{) }
-\end{Highlighting}
-\end{Shaded}
+One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \ref{fig:smoothsmodel6}).
+Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness \textgreater{} 4 (Table \ref{tab:restrictlength}) result in log10(Area) showing a significant effect with larger catchments showing less impact of changes in forest cover.
+Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
+Finally, more recent studies also show greater impacts of changes in forest cover (Figure \ref{fig:smoothsmodel7}).
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{flow_forest_byArea} \caption{Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data}\label{fig:overview}
+\end{figure}
 
@@
 
-Essentially, the analysis shows at the moment that in contrast to Zhang et al.
-(2017) there is no evidence that the size of a watershed influences the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds.
+Essentially, the analysis shows that in contrast to Zhang et al.
+(2017) there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds (Figure \ref{fig:overview}).
 In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
 
-As shown earlier, most of the smaller watersheds are ``real observed data'' using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+However, most of the smaller watersheds are ``real observed data'' using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 
-\hypertarget{remove-the-assessment-techniques-with-very-small-numbers}{%
-\subsubsection{remove the assessment techniques with very small numbers}\label{remove-the-assessment-techniques-with-very-small-numbers}}
+\hypertarget{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}{%
+\subsubsection{The effect of assessment techniques with very small numbers of observations}\label{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}}
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.08}}@{}}
+\caption{\label{tab:tableassess} Distribution of assessment techniques in the data set}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+Assessment\_technique
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+n
+\end{minipage} \\
+\midrule
+\endfirsthead
 \toprule
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments}\label{fig:assessment}
-\end{figure}
-
 \begin{longtable}[]{@{}
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & -18.71 & 19.02 & -0.98 & 0.33 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.83 & 0 \\
-\textbf{Forest\_Signincrease} & -23.26 & 7.15 & -3.25 & 0 \\
-\textbf{Precip\_data\_typeOB} & -11.25 & 14.19 & -0.79 & 0.43 \\
-\textbf{Precip\_data\_typeSG} & 18.43 & 16.43 & 1.12 & 0.26 \\
-\textbf{Assessment\_techniqueHM} & 32.02 & 12.76 & 2.51 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 44.48 & 12.97 & 3.43 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 38.12 & 21.21 & 1.8 & 0.07 \\
-\textbf{Assessment\_techniqueSH} & 46.79 & 13.17 & 3.55 & 0 \\
-\textbf{Forest\_typeCF} & -3.3 & 8.23 & -0.4 & 0.69 \\
-\textbf{Forest\_typeMF} & -3.88 & 8.69 & -0.45 & 0.66 \\
-\textbf{Hydrological\_regimeSD} & 5.73 & 10.11 & 0.57 & 0.57 \\
+\caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & -16.88 & 19.02 & -0.89 & 0.38 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.85 & 0 \\
+\textbf{Forest\_Signincrease} & -22.24 & 7.15 & -3.11 & 0 \\
+\textbf{Precip\_data\_typeOB} & -13.73 & 14.27 & -0.96 & 0.34 \\
+\textbf{Precip\_data\_typeSG} & 17.94 & 16.44 & 1.09 & 0.28 \\
+\textbf{Assessment\_techniqueHM} & 31.06 & 12.73 & 2.44 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 44.97 & 13.06 & 3.44 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 34.75 & 21.26 & 1.63 & 0.1 \\
+\textbf{Assessment\_techniqueSH} & 48.06 & 13.16 & 3.65 & 0 \\
+\textbf{Forest\_typeCF} & -6.9 & 8.5 & -0.81 & 0.42 \\
+\textbf{Forest\_typeMF} & -4.53 & 8.68 & -0.52 & 0.6 \\
+\textbf{Hydrological\_regimeSD} & 10.3 & 10.46 & 0.99 & 0.33 \\
 \bottomrule
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{\label{tab:model8-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
 \toprule
@@
 \midrule
-\endhead
-\textbf{s(Dryness)} & 3.51 & 9 & 1.96 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.7 & 9 & 0.24 & 0.08 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.99 \\
-\bottomrule
-\end{longtable}
-
-In drier watersheds, changes in forest cover have greater impact on flow, which is similar to Zhang et al.
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(Dryness)} & 3.55 & 9 & 1.82 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.75 & 9 & 0.27 & 0.07 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.37 \\
+\textbf{s(From)} & 0.79 & 9 & 0.26 & 0.08 \\
+\bottomrule
+\end{longtable}
+
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results.
+It strengthens the significance of the assessment technqiues but slightly reduces the importance of some of the smooth variables (\emph{log10(Area (km\textsuperscript{2}))} and \emph{From}, Table \ref{tab:modl8-linear} and \ref{tab:model8-smooth}).
+Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes
+
+In drier watersheds, changes in forest cover have greater impact on flow, which is similar to the observations from Zhang et al.
 (2017).
 This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
-As we don't have a separate variable for groundwater inputs (although this effect is estimated in many studies), we cannot analyse this effect separately.
-
-In contrast to Filoso et al. (2017), we also did not identify an effect of the
-
-Given how skewed Dryness is due to the few watersheds that have very high dryness values, it is worth investigating what excluding these 4 watersheds from the data means for the relationships.
+As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect cannot be analysed separately.
+
+In contrast to Filoso et al.
+(2017), we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
+What is puzzling is that the effect of the start date (\emph{From}) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
+At this point, it is unclear what this trend signifies.
+It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
+
+Excluding the few watersheds that have very high dryness values, clarified the \emph{Dryness} trend, agreeing with earlier studies.
 
@@
 There are further confounding factors in the data, which were also classified by Filoso et al.
-(2017) and these create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD), which are weakly significant, are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf forests (BF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow.
+(2017) and these might create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow (Vervoort et al., 2021).
 
@@
 Andréassian, V., 2004. Waters and forests: From historical controversy to scientific debate. Journal of Hydrology 291, 1--27. doi:\url{https://doi.org/10.1016/j.jhydrol.2003.12.015}
+
+\leavevmode\vadjust pre{\hypertarget{ref-baker1984}{}}%
+Baker Jr., M.B., 1984. Changes in streamflow in an herbicide-treated pinyon-juniper watershed in arizona. Water Resources Research 20, 1639--1642. doi:\url{https://doi.org/10.1029/WR020i011p01639}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vervoort2021}{}}%
+Vervoort, R.W., Dolk, M.M., Ogtrop, F.F.
+van, 2021.
+Climate change and other trends in streamflow observations in australian forested catchments since 1970.
+Hydrological Processes 35, e13999.
+doi:\url{https://doi.org/10.1002/hyp.13999}
+
 \leavevmode\vadjust pre{\hypertarget{ref-wood2006}{}}%

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.Rmd
moving towards a final version'

@@
 ---
-title: Factors determining how catchments respond to forest cover change: Re-analysing global data sets.
+title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
 author:
@@
 Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38/%) is low due the quality of the inputs and additional unknown confounding factors.
+Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
 journal: "Journal of Hydrology"
@@
 <!-- Include Filoso paper -->
+For the purpose of this paper, *watershed* and *catchment* are interchangeable terms.
+Many of the US studies use *watershed*, while European and Australian studies use *catchment*.
+In particular, we retained the term "paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.
 
@@
 
-The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as < 1000 km^2^ and large watersheds > 1000 km^2^.
-While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed [@zhang2017].
+The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^.
+While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
+Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].
 
@@
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of quantitative data (37) they showed a log-linear relationship between decreasing watershed size and an increasing decline in streamflow.
+For a subset of quantitative data (37) they showed a log-linear relationship between decreasing catchment size and an increasing decline in streamflow.
 In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
@@
 The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
 Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
@@
 
-While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
-We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) catchment data sets in our analysis.
+While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
+In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
+We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) in our analysis.
 
@@
 
-In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
+In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
 Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
-For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+
+Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+
+\begin{equation}
+D = \frac{E_{0}}{Pa} (\#eq:eq1)
+\end{equation}
 
 The length of the study can be a variable influencing the change in flow [e.g.
-@jackson2005], as for example, more mature plantations are thought to have smaller impacts on flow.
+@jackson2005; @filoso2017], as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a "Kuczera curve" [@kuczera1987].
 Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
+The length of the study was already included in the data from @filoso2017, but these were checked against the original publications.
 
-Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
+Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
 In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
 Table 6 in @zhang2011).
-We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017.
-This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for watershed #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
+We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017.
+This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
+Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
 Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.
@@
 The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
-
-Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
-$$\tag{1}
-\begin{aligned}
-D = \frac{E_{0}}{Pa}
-\end{aligned}$$
 
@@
 
-$$\tag{2}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
-\end{aligned}$$
+\begin{align}
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
+& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
+\end{align}
 
@@
 
-For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the results.
-Subsequently the analysis was repeated and the additionally identified watersheds were added.
+For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results.
+Subsequently the analysis was repeated and the additionally identified catchments were added.
 
@@
 The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` watersheds and the @filoso2017 study used 37 watersheds (Table S2 in @filoso2017).
+In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
@@
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller watersheds dominate the database with 42\% of the data from watersheds < 1 km^2^ and 65\% of the data for watersheds < 10 km^2^.
+Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^.
 
@@
 
-$$\tag{3}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
-\end{aligned}$$
+\begin{equation}
+\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon (\#eq:eq3)
+\end{equation}
 
@@
 ```{r, echo =F}
-pander(round(summary(linear_model_simple)$p.table,2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
+pander(round(rbind(summary(linear_model_simple)$p.table,
+                       summary(linear_model_simple_b)$p.table),2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
+The first three rows relate to the model using the original data base from Zhang et al.
+(2017).
+The bottom three rows are the results of the model including the new data")
 ```
 
-```{r, echo =F}
-pander(round(summary(linear_model_simple_b)$p.table,2), caption="(\\#tab:tabmodel1b) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets")
-```
-
-The overall variance explained in this model is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \@ref(eq:eq3)) is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 ```
+
 ```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
@@
 
-Of importance here is to highlight the residuals of this regression.
+Of importance here is to highlight the residuals of this regression (equation \@ref(eq:eq3) and Figure \@ref(fig:gamcheck)).
 These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)).
@@
 
-Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1) and Table \@ref(tab:tabmodel1b)), and the result indication a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
+Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
 
 It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being watershed dryness and area (following @zhang2017):
-
-$$\tag{4}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
-\end{aligned}$$
-
-Where $Pa_mm$ is the annual average rainfall in mm, and a log base 10 transformation is applied to the area variable given that the distribution of Area is higly skewed (Figure 1)
-
-<!-- At the moment using watershed rainfall, we probably want to straight away use P/E0, and explain that rainfall itself might not be a good predictor.
-Alternatively we can first show that Latitude is a predictor -->
-
-```{r model2, echo = F}
-
-linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm, data = All_data2)
-#summary(linear_model2)
-#gam.check(linear_model2)
-```
-
-```{r, echo = F}
-pander(round(summary(linear_model2)$p.table,2), caption="(\\#tab:out-model2) Summary of the second model, taking into account the annual rainfall and the area of the watershed")
-```
-
-Including area and annual precipitation slightly improves the overall explaining power of the model (Table \@ref(tab:out-model2)).
-Annual precipitation is in fact not significant.
-Relative to earlier reported studies [@zhang2017; @filoso2017], the log base 10 transformed watershed area indicates a p-value of only `r round(summary(linear_model2)$p.table[4,4],2)`, suggesting a marginal impact on the change in stream flow.
-This supports our approach (in contrast to @zhang2017) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds.
-The main effects in the model remain the change in forest cover and whether this is an increase or decrease.
+The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.
 
@@
 
-```{r globalmap, fig.cap="Distribution of included watersheds across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
+```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
 
@@
 
-As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
+As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
+These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps.
+In the dataset, an additional column is added to indicate the source of the location data.
+As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
 Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
 
-$$\tag{5}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
-\end{aligned}$$
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon (\#eq:eq4)
+\end{align}
 
 ```{r model3, echo = F}
-model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) + Pa_mm + Latitude + Longitude, data = All_data2)
+model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Latitude + Longitude, data = All_data2)
 #summary(model3_includeLatLong)
@@
 
-There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the watersheds across the globe has little influence on the overall result.
+There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
 The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
-Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow, as also indicated in @zhang2017.
-
-## Impact of the dryness index
+## Impact of climate
 
 Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both @zhang2017 and @jackson2005.
-Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
-This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from the precipitation.
-It also drops the Latitude and Longitude as these are indicated not to be significant.
+Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+Initially, we tested models using annual average precipitation (*Pa (mm)*), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable.
+Given that Latitude and Longitude were not significant, we dropped these from the model.
+
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness + \varepsilon (\#eq:eq5)
+\end{align}
 
@@
 #  mutate(length = To - From)
-model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + log10(Area_km2) +  
+model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                         Dryness, data = All_data2)
@@
 ```{r, echo =F}
-pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model replacing the annual precipitation with the dryness index")
+pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
 ```
 
-The results from this model (Table \@ref(tab:out-model4)) confirm that dryness is a significant confounding factor related to the change in streamflow as a function of the change in forest cover change.
-In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is ~14\% greater.
-However, more interesting, Latitude remains a significant predictor with each degree in latitude causing an -0.31\% change in runoff.
-This indicates that Dryness (i.e.
-an increase in radiation) alone does not explain the trend in the Latitude and some other unknown confounding factor is captured by Latitude.
+The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
+In this case the evidence is not very strong (p = `r round(summary(model4_dryness)$p.table[4,4],2)`.
+However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is ~6\% greater.
+Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
 
 ```{r, echo = F}
-pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) Watersheds for which the dryness index > 4")
+pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
 ```
 
-However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data.
-These watersheds are listed in Table \@ref(tab:drytable).
+There possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
+These catchments are listed in Table \@ref(tab:drytable).
 
-### Are some of the variables possibly non-linear?
+## Is there a distinct effect of area?
+
+The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017.
+Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
+
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon (\#eq:eq6)
+\end{align}
+
+```{r modelArea, echo = F}
+#Zhang_all2 <- Zhang_all2 %>%
+#  mutate(length = To - From)
+model_Area <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
+                        Dryness + log10(Area_km2), data = All_data2)
+#summary(model_Area)
+#gam.check(model_Area)
+```
+
+```{r, echo =F}
+pander(round(summary(model_Area)$p.table,2), caption = "(\\#tab:out-modelArea) Results of the model including Area and the dryness index")
+```
+
+The results of this model (Equation \@ref(eq:eq6)) clearly indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)).
+In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%.
+Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
+Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
+
+```{r modelArea_b, echo = F, eval =F}
+#Zhang_all2 <- Zhang_all2 %>%
+#  mutate(length = To - From)
+model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
+                        Dryness*log10(Area_km2), data = All_data2)
+summary(model_Area_b)
+#gam.check(model_Area)
+```
+
+## Are some of the variables possibly non-linear?
 
@@
 
-$$\tag{5}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
-\end{aligned}$$
-
-<!--
-```{r model4b, echo = F, eval = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From)
-
-model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
-                 s(length, bs = "ts"),
-                  data = All_data2 %>% filter(length < 60))
-summary(model4b)
-#gam.check(model5_all)
-#plot(model5_all)
-```
--->
+\begin{align}
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon (\#eq:eq7)
+\end{align}
 
@@
 
-Including non-linearity increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
+Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
+For example, the smoothed variable for Area (km^2^) is no longer a strong effect explaining variations in changes in stream flow.
+Clearly *length* is also not explaining the variation.
+In contrast, *From*, which indicate the start date of the study (and therefore the age of the study) is a significant effect.
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
+Including interactions into the smooths is possible, but the results are even more difficult to interpret given the high flexibility of the two-dimensional smooth.
 
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
+In this case, *Area (km^2)* is no longer a significant predictor.
 
@@
 Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006).
-It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
+It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
 Similarly, the length variable is not significant.
@@
 
-<!--
-```{r model6_reduc, eval = F, echo = F}
-model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" )  + 
-                      s(log10(Area_km2), bs="ts") +  length +
-                    Assessment_technique +
-                    Hydrological_regime, data = Zhang_all2)
-summary(model6_reduc)
-
-```
-
-```{r, eval =F, echo =F}
-All_data2 %>%
-  ggplot(aes(Assessment_technique, length)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
-```
-
-Maybe remove data for low numbers of assessment techniques and rerun the analysis (<10)
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Forest_type, Latitude)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
-```
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Hydrological_regime, Latitude)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
-```
- -->
-
-<!--
-```{r, eval =F, echo =F}
-All_data2 %>%
-  ggplot(aes(DeltaQf_perc, length)) + 
-  geom_point(colour = "blue") + theme_bw()
-```
--->
-
-<!--filter lenght < 60 -->
+<!--filter length < 60 -->
 ```{r model7_nolongLength, echo = F}
@@
 One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness > 4 (Table \@ref(tab:restrictlength)) result in log10(Area) showing a significant effect with larger catchments showing less impact of changes in forest cover.
+Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness > 4 (Table \@ref(tab:restrictlength)) results in log10(Area) once again showing a significant effect with larger catchments having less impact of changes in forest cover.
 Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
@@
 ```
+
+```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
+
+include_graphics("flow_forest_byArea.png")
+
+```
+
+Discussion
+======================
+
+## Catchment size
+
+Essentially, the analysis shows that there is a clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), even though, in contrast to @zhang2017, there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \@ref(fig:overview)).
+In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
+
+An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change [@navas2019].
+Therefore, specifically if the  forest cover change is small relative to the catchment size, the effect of this change will be buffered.
+
+```{r, eval =T, echo =F}
+# All_data2 %>%
+#   ggplot(aes(log10(Area_km2), DeltaF_perc_pos)) + 
+#   geom_point(alpha = 0.5) + theme_bw()
+
+Area_deltaF <- lm(DeltaF_perc_pos~log10(Area_km2), data = All_data2)
+```
+
+There are two caveats on this explanation.
+The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average the recorded forest cover change is 10% lower.
+This is basically a result of the fact that large changes in forest cover are difficult to "implement" in an experiment.
+This is also reflected in the second caveat most of the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
+These techniques for larger catchments all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
+This is a confounding factor that is not easily addressed in the regression modelling attempted here.
+Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
+
+In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
+
+Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
+While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
+Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
+
+## Model residuals
+
+As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
+These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \@ref(fig:overview)), which show very high variability.
+The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \@ref(fig:gamcheck7)).
+
+```{r, eval = F, echo = F}
+png("residual_plot_model7.png", width = 2500, height = 1800, res = 300)
+par(mfrow=c(2,2))
+gam.check(model7)
+par(mfrow=c(1,1))
+dev.off()
+```
+```{r gamcheck7, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
+include_graphics("residual_plot_model1.png")
+```
+
+The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
+Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
+In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
+This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \@ref(fig:increasedecrease)).
+
+## The effect of assessment techniques with very small numbers of observations
+
+```{r, eval =T, echo =F}
+Table_assessment <- All_data2 %>%
+  group_by(Assessment_technique) %>%
+  summarise(n = n()) %>%
+  arrange(desc(n))
+```
+```{r, echo=F}
+pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
+```
+
+```{r, eval = F, echo =F}
+png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
+All_data2 %>%
+  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") + theme_bw() +
+    theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)))
+dev.off()
+```
+
+```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =T}
+
+include_graphics("AssessmentTechnique_byArea.png")
+```
+
+One concern is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
+
+```{r model8_reduc_assess, echo = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
+
+model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
+#plot(model8_assess)
+#summary(model8_assess)
+
+```
+```{r, echo = F}
+pander(round(summary(model8_assess)$p.table,2), caption= "(\\#tab:model8-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
+pander(round(summary(model8_assess)$s.table,2), caption = "(\\#tab:model8-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
+
+```
+
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:model8-linear) and \@ref(tab:model8-smooth)).
+It strengthens the significance of the different assessment techniques but slightly reduces the importance of some of the smooth variables (*log10(Area (km^2^))* and *From*).
+Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
+
+However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
+Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
+In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
+However, as indicated earlier, the EA studies are all on the drier end of the Dryness spectrum, highlighting another unresolved interaction in the data.
+
+## The effect of climate
+
+In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies [@zhang2017; @zhou2015; @filoso2017].
+This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
+
+Excluding the few catchments that have very high dryness values, clarified the *Dryness* trend, agreeing with earlier studies and showing a increase in the change in flow for drier catchments.
+However, this really only starts to have an effect for Dryness > 2.
+
+## Further considerations
+
+In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
+What is puzzling is that the effect of the start date (*From*) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
+At this point, it is unclear what this trend signifies.
+It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
+
+There are further confounding factors in the data, which were also classified by @filoso2017 and these might create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
+
+Conclusions
+============
+
+More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
+
+Results of this statistical analysis need to be considered "conditional on the data"
+Conditional on the data, the impact of forestry on streamflow is:  
+- Greater for forest clearing then for reforestation  
+- Reduced for larger watersheds  
+- Increases for drier watersheds  
+- Sensitive to the assessment method used in the historical data
+
+References {#references .unnumbered}
+==========
 
@@
 ```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  mutate(size_class = round(log10(Area_km2),0)) %>%
-  ggplot(aes(log10(Area_km2), DeltaQf_perc, group = size_class)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
+All_data2 %>%
+  ggplot(aes(log10(Area_km2), Dryness)) + 
+  geom_point(colour = "blue") + theme_bw()
 ```
@@
 
-```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
+<!--
+```{r model4b, echo = F, eval = F}
+All_data2 <- All_data2 %>%
+  mutate(length = To - From)
 
-include_graphics("flow_forest_byArea.png")
+model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
+                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
+                 s(length, bs = "ts"),
+                  data = All_data2 %>% filter(length < 60))
+summary(model4b)
 
 ```
+-->
+```{r model5b, echo = F, eval = F}
+All_data2 <- All_data2 %>%
+  mutate(length = To - From)
 
-Discussion
-======================
-
-Essentially, the analysis shows that in contrast to @zhang2017 there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds (Figure \@ref(fig:overview)).
-In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
-
-However, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
-This is a confounding factor that is not easily addressed in the regression modelling attempted here.
-Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
-
-## The effect of assessment techniques with very small numbers of observations
-
-One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).
-The following model removes these from the data set.
-
-```{r tableassess, eval =T, echo =F}
-Table_assessment <- All_data2 %>%
-  group_by(Assessment_technique) %>%
-  summarise(n = n()) %>%
-  arrange(desc(n))
-
-pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
+model5_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    te(log10(Area_km2),Dryness, bs="ts") +  
+                  s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
+                    s(length, bs="ts") + s(From, bs = "ts") ,
+                  data = All_data2)
+summary(model5_b)
+plot(model5_b)
 ```
-
-```{r, eval = F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw() +
-    theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)))
-dev.off()
-```
-
-```{r assessment, fig.cap="Boxplot of the log base 10 of the watershed area (in km2) for the different assessment techniques, showing the dominance of small watersheds in the paired watershed experiments", out.width="90%", echo=F, eval =F}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
-
-```{r model8_reduc_assess, echo = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
-#plot(model8_assess)
-#summary(model8_assess)
-
-```
-```{r, echo = F}
-pander(round(summary(model8_assess)$p.table,2), caption= "(\\#tab:model8-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
-pander(round(summary(model8_assess)$s.table,2), caption = "(\\#tab:model8-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
-
-```
-
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results.
-It strengthens the significance of the assessment technqiues but slightly reduces the importance of some of the smooth variables (*log10(Area (km^2^))* and *From*, Table \@ref(tab:modl8-linear) and \@ref(tab:model8-smooth)).
-Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes
-
-In drier watersheds, changes in forest cover have greater impact on flow, which is similar to the observations from @zhang2017.
-This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
-As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect cannot be analysed separately.
-
-In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
-What is puzzling is that the effect of the start date (*From*) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
-At this point, it is unclear what this trend signifies.
-It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
-
-Excluding the few watersheds that have very high dryness values, clarified the *Dryness* trend, agreeing with earlier studies.
-
-## Other possible bias in the data
-
-There are further confounding factors in the data, which were also classified by @filoso2017 and these might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
-
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-
-Conclusions
-============
-
-References {#references .unnumbered}
-==========

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.tex
moving towards a final version'

@@
             pdfauthor={},
-            pdftitle={Do larger watersheds respond different to forest cover change? Re-analysing a global data set.},
+            pdftitle={Factors determining how catchments respond to forest cover change. Re-analysing global data sets.},
             colorlinks=false,
@@
 
-  \title{Do larger watersheds respond different to forest cover change? Re-analysing a global data set.}
+  \title{Factors determining how catchments respond to forest cover change. Re-analysing global data sets.}
     \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
@@
 Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38/\%) is low due the quality of the inputs and additional unknown confounding factors.
+Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
   \end{abstract}
@@
 However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchmensts after deforestation.
+
+For the purpose of this paper, \emph{watershed} and \emph{catchment} are interchangeable terms.
+Many of the US studies use \emph{watershed}, while European and Australian studies use \emph{catchment}.
+In particular, we retained the term ``paired watershed studies'' as this is the most common terminology, but further mostly use the term catchment.
 
@@
 
-The conclusions of the first paper (Zhang et al., 2017) suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds, defined as \textless{} 1000 km\textsuperscript{2} and large watersheds \textgreater{} 1000 km\textsuperscript{2}.
-While for small watersheds there was no real change in runoff with changes in cover, for large watersheds there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited watersheds relative to energy limited watershed (Zhang et al., 2017).
+The conclusions of the first paper (Zhang et al., 2017) suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as \textless{} 1000 km\textsuperscript{2} and large watersheds (catchments) \textgreater{} 1000 km\textsuperscript{2}.
+While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
+Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments (Zhang et al., 2017).
 
@@
 (2017) as supplementary material.
-The columns in this data set are the watershed number, the watershed name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in Zhang et al.
+The columns in this data set are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in Zhang et al.
 (2017)), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
@@
 
-While Zhang et al. (2017) use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
-We combined the tables for small (\textless{} 1000 km\textsuperscript{2}) and large (\textgreater= 1000 km\textsuperscript{2}) catchment data sets in our analysis.
+While Zhang et al.
+(2017) use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
+In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
+We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) in our analysis.
 
@@
 
-In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location.
+In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
 Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
-For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
-
-The length of the study can be a variable influencing the change in flow (e.g.
-Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow.
+For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+
+Similar to Zhang et al. (2017), the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+
+\begin{equation}
+D = \frac{E_{0}}{Pa} \label{eq:eq1}
+\end{equation}
+
+The length of the study can be a variable influencing the change in flow (Filoso et al., 2017; e.g.
+Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987).
 Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al.
 (2017).
-
-Several additional data points from watershed studies were extracted from Zhang et al.
+The length of the study was already included in the data from Filoso et al.
+(2017), but these were checked against the original publications.
+
+Several additional data points from catchment studies were extracted from Zhang et al.
 (2011), Zhao et al.
@@
 (2011)).
-We also removed one data point from the analysis, which corresponds to Watershed \#1 (Amazon) in Zhang et al.
+We also removed one data point from the analysis, which corresponds to catchment \#1 (Amazon) in Zhang et al.
 (2017).
-This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for watershed \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984).
+This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
+Furthermore, the change in flow for catchment \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984).
 Finally, on review of all the data in Zhang et al.
@@
 
-Similar to Zhang et al. (2017), the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
-\[\tag{1}
-\begin{aligned}
-D = \frac{E_{0}}{Pa}
-\end{aligned}\]
-
 \hypertarget{statistical-modelling}{%
@@
 
-\[\tag{2}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
-\end{aligned}\]
+\begin{align}
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
+& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq2}
+\end{align}
 
@@
 For the model in equation 2, we initially only used the data from Zhang et al.
-(2017) to make sure that the additional watersheds added to the data set did not influence the results.
-Subsequently the analysis was repeated and the additionally identified watersheds were added.
+(2017) to make sure that the additional catchments added to the data set did not influence the results.
+Subsequently the analysis was repeated and the additionally identified catchments were added.
 
@@
 In contrast, the original dataset from Zhang et al.
-(2017) contained 312 watersheds and the Filoso et al.
-(2017) study used 37 watersheds (Table S2 in Filoso et al.
+(2017) contained 312 catchments and the Filoso et al.
+(2017) study used 37 catchments (Table S2 in Filoso et al.
 (2017)).
@@
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller watersheds dominate the database with 42\% of the data from watersheds \textless{} 1 km\textsuperscript{2} and 65\% of the data for watersheds \textless{} 10 km\textsuperscript{2}.
+Smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2}.
 
@@
 
-\[\tag{3}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
-\end{aligned}\]
+\begin{equation}
+\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon \label{eq:eq3}
+\end{equation}
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change}\tabularnewline
+\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
+The first three rows relate to the model using the original data base from Zhang et al.~(2017).
+The bottom three rows are the results of the model including the new data}\tabularnewline
 \toprule
@@
 \textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:tabmodel1b} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change including the new data sets}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
 \textbf{(Intercept)} & 9.43 & 5.54 & 1.7 & 0.09 \\
@@
 
-The overall variance explained in this model is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 
-Of importance here is to highlight the residuals of this regression.
+Of importance here is to highlight the residuals of this regression (equation \eqref{eq:eq3} and Figure \ref{fig:gamcheck}).
 These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \ref{fig:gamcheck}).
@@
 
-Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1} and Table \ref{tab:tabmodel1b}), and the result indication a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
+Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
 
 It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being watershed dryness and area (following Zhang et al.
-(2017)):
-
-\[\tag{4}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
-\end{aligned}\]
-
-Where \(Pa_mm\) is the annual average rainfall in mm, and a log base 10 transformation is applied to the area variable given that the distribution of Area is higly skewed (Figure 1)
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:out-model2} Summary of the second model, taking into account the annual rainfall and the area of the watershed}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 22.55 & 9.16 & 2.46 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.34 & 0.1 & 3.26 & 0 \\
-\textbf{Forest\_Signincrease} & -35.52 & 5.67 & -6.27 & 0 \\
-\textbf{log10(Area\_km2)} & -3.12 & 1.72 & -1.81 & 0.07 \\
-\textbf{Pa\_mm} & 0 & 0 & -1 & 0.32 \\
-\bottomrule
-\end{longtable}
-
-Including area and annual precipitation slightly improves the overall explaining power of the model (Table \ref{tab:out-model2}).
-Annual precipitation is in fact not significant.
-Relative to earlier reported studies (Filoso et al., 2017; Zhang et al., 2017), the log base 10 transformed watershed area indicates a p-value of only 0.07, suggesting a marginal impact on the change in stream flow.
-This supports our approach (in contrast to Zhang et al.
-(2017)) to consider watershed area as a continuous variable and making no separation between larger and smaller watersheds.
-The main effects in the model remain the change in forest cover and whether this is an increase or decrease.
+The obvious ones being catchment dryness and area (following Zhang et al.
+(2017)), which we will analyse later.
 
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included watersheds across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
 \end{figure}
 
-As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world.
+As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
+These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps.
+In the dataset, an additional column is added to indicate the source of the location data.
+As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
 Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
 
-\[\tag{5}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
-\end{aligned}\]
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon \label{eq:eq4}
+\end{align}
 
@@
 \endhead
-\textbf{(Intercept)} & 23.29 & 9.96 & 2.34 & 0.02 \\
-\textbf{DeltaF\_perc\_pos} & 0.35 & 0.1 & 3.32 & 0 \\
-\textbf{Forest\_Signincrease} & -37.21 & 6 & -6.2 & 0 \\
-\textbf{log10(Area\_km2)} & -3.28 & 1.73 & -1.89 & 0.06 \\
-\textbf{Pa\_mm} & 0 & 0 & -0.83 & 0.41 \\
-\textbf{Latitude} & -0.05 & 0.09 & -0.55 & 0.58 \\
-\textbf{Longitude} & 0.01 & 0.03 & 0.23 & 0.82 \\
+\textbf{(Intercept)} & 10.13 & 6.01 & 1.69 & 0.09 \\
+\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.22 & 0 \\
+\textbf{Forest\_Signincrease} & -38.15 & 5.95 & -6.41 & 0 \\
+\textbf{Latitude} & -0.04 & 0.09 & -0.4 & 0.69 \\
+\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
 \bottomrule
@@
 
-There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the watersheds across the globe has little influence on the overall result.
+There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
 The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
-Note that in this case the significance of the Area variable increases, and generally indicates that larger watersheds would be expected to have a lower change in streamflow, as also indicated in Zhang et al.
-(2017).
-
-\hypertarget{impact-of-the-dryness-index}{%
-\subsection{Impact of the dryness index}\label{impact-of-the-dryness-index}}
+\hypertarget{impact-of-climate}{%
+\subsection{Impact of climate}\label{impact-of-climate}}
 
@@
 (2005).
-Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics).
-This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from the precipitation.
-It also drops the Latitude and Longitude as these are indicated not to be significant.
+Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+Initially, we tested models using annual average precipitation (\emph{Pa (mm)}), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable.
+Given that Latitude and Longitude were not significant, we dropped these from the model.
+
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness + \varepsilon \label{eq:eq5}
+\end{align}
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:out-model4} Results of the model replacing the annual precipitation with the dryness index}\tabularnewline
+\caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{(Intercept)} & 10.46 & 7.66 & 1.36 & 0.17 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.11 & 2.93 & 0 \\
-\textbf{Forest\_Signincrease} & -35.67 & 5.73 & -6.23 & 0 \\
-\textbf{log10(Area\_km2)} & -3.5 & 1.76 & -1.99 & 0.05 \\
-\textbf{Dryness} & 6.54 & 3.08 & 2.12 & 0.03 \\
+\textbf{(Intercept)} & 3.18 & 6.77 & 0.47 & 0.64 \\
+\textbf{DeltaF\_perc\_pos} & 0.43 & 0.09 & 4.91 & 0 \\
+\textbf{Forest\_Signincrease} & -37.43 & 5.68 & -6.59 & 0 \\
+\textbf{Dryness} & 5.56 & 3.06 & 1.82 & 0.07 \\
 \bottomrule
@@
 
-The results from this model (Table \ref{tab:out-model4}) confirm that dryness is a significant confounding factor related to the change in streamflow as a function of the change in forest cover change.
-In fact if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2*Pa, which is very dry), the change in runoff is \textasciitilde14\% greater.
-However, more interesting, Latitude remains a significant predictor with each degree in latitude causing an -0.31\% change in runoff.
-This indicates that Dryness (i.e.~an increase in radiation) alone does not explain the trend in the Latitude and some other unknown confounding factor is captured by Latitude.
+The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
+In this case the evidence is not very strong (p = 0.07.
+However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is \textasciitilde6\% greater.
+Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.46}}@{}}
-\caption{\label{tab:drytable} Watersheds for which the dryness index \textgreater{} 4}\tabularnewline
+\caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
 \toprule
@@
 
-However, the result also indicates possible issues with the data, some of the Dryness values are very large (\textgreater{} 4) and these values have high leverage in the data.
-These watersheds are listed in Table \ref{tab:drytable}.
+There possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
+These catchments are listed in Table \ref{tab:drytable}.
+
+\hypertarget{is-there-a-distinct-effect-of-area}{%
+\subsection{Is there a distinct effect of area?}\label{is-there-a-distinct-effect-of-area}}
+
+The major hypothesis to test is the effect of area on the change in flow, following the analysis by Zhang et al.
+(2017) and Filoso et al.
+(2017).
+Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.
+
+\begin{align}
+\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon \label{eq:eq6}
+\end{align}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 10.46 & 7.66 & 1.36 & 0.17 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.11 & 2.93 & 0 \\
+\textbf{Forest\_Signincrease} & -35.67 & 5.73 & -6.23 & 0 \\
+\textbf{Dryness} & 6.54 & 3.08 & 2.12 & 0.03 \\
+\textbf{log10(Area\_km2)} & -3.5 & 1.76 & -1.99 & 0.05 \\
+\bottomrule
+\end{longtable}
+
+The results of this model (Equation \eqref{eq:eq6}) clearly indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.5\%.
+Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
+Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
 
 \hypertarget{are-some-of-the-variables-possibly-non-linear}{%
-\subsubsection{Are some of the variables possibly non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
+\subsection{Are some of the variables possibly non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
 
@@
 
-\[\tag{5}
-\begin{aligned}
-\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
-\end{aligned}\]
+\begin{align}
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon \label{eq:eq7}
+\end{align}
 
@@
 
-Including non-linearity increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.31 and deviance explained of 0.34, but creates few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.31 and deviance explained of 0.34, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, the smoothed variable for Area (km\textsuperscript{2}) is no longer a strong effect explaining variations in changes in stream flow.
+Clearly \emph{length} is also not explaining the variation.
+In contrast, \emph{From}, which indicate the start date of the study (and therefore the age of the study) is a significant effect.\\
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
+Including interactions into the smooths is possible, but the results are even more difficult to interpret given the high flexibility of the two-dimensional smooth.
 
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
+In this case, \emph{Area (km\^{}2)} is no longer a significant predictor.
 
@@
 Figure \ref{fig:smoothsmodel6} highlights that the relationship between log10(Area km\textsuperscript{2}) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following Wood (2006)).
-It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds.
+It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
 Similarly, the length variable is not significant.
@@
 One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \ref{fig:smoothsmodel6}).
-Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness \textgreater{} 4 (Table \ref{tab:restrictlength}) result in log10(Area) showing a significant effect with larger catchments showing less impact of changes in forest cover.
+Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness \textgreater{} 4 (Table \ref{tab:restrictlength}) results in log10(Area) once again showing a significant effect with larger catchments having less impact of changes in forest cover.
 Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
@@
 
-Essentially, the analysis shows that in contrast to Zhang et al.
-(2017) there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds (Figure \ref{fig:overview}).
-In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.
-
-However, most of the smaller watersheds are ``real observed data'' using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds.
+\hypertarget{catchment-size}{%
+\subsection{Catchment size}\label{catchment-size}}
+
+Essentially, the analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), even though, in contrast to Zhang et al.
+(2017), there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
+In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
+
+An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change (Navas et al., 2019).
+Therefore, specifically if the forest cover change is small relative to the catchment size, the effect of this change will be buffered.
+
+There are two caveats on this explanation.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.34 with a slope of -9.63) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is 10\% lower.
+This is basically a result of the fact that large changes in forest cover are difficult to ``implement'' in an experiment.
+This is also reflected in the second caveat most of the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
+These techniques for larger catchments all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 
+In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
+
+Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
+While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
+Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
+
+\hypertarget{model-residuals}{%
+\subsection{Model residuals}\label{model-residuals}}
+
+As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
+These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \ref{fig:overview}), which show very high variability.
+The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheck7}).
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{residual_plot_model1} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheck7}
+\end{figure}
+
+The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
+Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
+In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
+This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \ref{fig:increasedecrease}).
+
 \hypertarget{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}{%
-\subsubsection{The effect of assessment techniques with very small numbers of observations}\label{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}}
-
-One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).
-The following model removes these from the data set.
+\subsection{The effect of assessment techniques with very small numbers of observations}\label{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}}
 
@@
 \end{longtable}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments}\label{fig:assessment}
+\end{figure}
+
+One concern is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
 
@@
 
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results.
-It strengthens the significance of the assessment technqiues but slightly reduces the importance of some of the smooth variables (\emph{log10(Area (km\textsuperscript{2}))} and \emph{From}, Table \ref{tab:modl8-linear} and \ref{tab:model8-smooth}).
-Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes
-
-In drier watersheds, changes in forest cover have greater impact on flow, which is similar to the observations from Zhang et al.
-(2017).
-This is most likely because in these watersheds the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
-As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect cannot be analysed separately.
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \ref{tab:model8-linear} and \ref{tab:model8-smooth}).
+It strengthens the significance of the different assessment techniques but slightly reduces the importance of some of the smooth variables (\emph{log10(Area (km\textsuperscript{2}))} and \emph{From}).
+Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
+
+However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
+Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
+In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in Zhou et al.
+(2015)) appears to suggest a much smaller effect on the change in flow.
+However, as indicated earlier, the EA studies are all on the drier end of the Dryness spectrum, highlighting another unresolved interaction in the data.
+
+\hypertarget{the-effect-of-climate}{%
+\subsection{The effect of climate}\label{the-effect-of-climate}}
+
+In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies (Filoso et al., 2017; Zhang et al., 2017; Zhou et al., 2015).
+This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
+
+Excluding the few catchments that have very high dryness values, clarified the \emph{Dryness} trend, agreeing with earlier studies and showing a increase in the change in flow for drier catchments.
+However, this really only starts to have an effect for Dryness \textgreater{} 2.
+
+\hypertarget{further-considerations}{%
+\subsection{Further considerations}\label{further-considerations}}
 
@@
 
-Excluding the few watersheds that have very high dryness values, clarified the \emph{Dryness} trend, agreeing with earlier studies.
-
-\hypertarget{other-possible-bias-in-the-data}{%
-\subsubsection{Other possible bias in the data}\label{other-possible-bias-in-the-data}}
-
 There are further confounding factors in the data, which were also classified by Filoso et al.
@@
 
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
+
+Results of this statistical analysis need to be considered ``conditional on the data''
+Conditional on the data, the impact of forestry on streamflow is:\\
+- Greater for forest clearing then for reforestation\\
+- Reduced for larger watersheds\\
+- Increases for drier watersheds\\
+- Sensitive to the assessment method used in the historical data
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-navas2019}{}}%
+Navas, R., Alonso, J., Gorgoglione, A., Vervoort, R.W., 2019. Identifying climate and human impact trends in streamflow: A case study in uruguay. Water 11, 1433.
+
 \leavevmode\vadjust pre{\hypertarget{ref-pena-arancibia2012}{}}%

### 2022-02-01 paper/Forest_and_Water/Forest_and_Water.Rmd
Figure 2 label corrected figure 4 modified with new data and all catchments under the same category

@@
 abstract: |
-  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
-Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
-Generalised additive models were used to run flexible regressions including multiple variables.
   Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
@@
 Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
+  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
+However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
+The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+Generalised additive models were used to run flexible regressions including multiple variables.
 journal: "Journal of Hydrology"
@@
 <!-- analysis -->
-## description of the data
+## Description of the data
 
@@
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^.
+Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
 
@@
 
+In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2).
+In other words, flow almost always increased.
+However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
+However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
+
 ```{r, echo = F, warning = F, message=F, eval =F}
@@
   scale_colour_manual(name = "Forest change direction", 
-                      values = c("increase" = "blue", "decrease" = "red")) +
+                      values = c("decrease" = "red","increase" = "blue")) +
   xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
@@
 ```
-
-This shows that for the data related to forest decreases, there is almost always a positive flow change.
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
 
@@
 (2017).
-The bottom three rows are the results of the model including the new data")
+The bottom three rows are the results of the model including the new data.")
 ```
@@
 
-include_graphics("../../data/FAOET0data2.png")
+include_graphics("../../data/FAOET0data_final_2022.png")
 ```

### 2022-02-01 paper/Forest_and_Water/Forest_and_Water.tex
Figure 2 label corrected figure 4 modified with new data and all catchments under the same category

@@
   \linenumbers % turns line numbering on
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
 \usepackage{graphicx}
+\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-
 % Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+% Pandoc header
 
 \begin{document}
-
 \begin{frontmatter}
@@
   \begin{abstract}
-  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
-Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
-Generalised additive models were used to run flexible regressions including multiple variables.
   Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
@@
 Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
+  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
+However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
+The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+Generalised additive models were used to run flexible regressions including multiple variables.
   \end{abstract}
@@
 \toprule
-Factor & Abbreviation & Definition \\
+Factor & Abbreviation & Definition\tabularnewline
 \midrule
 \endhead
-forest type & CF & coniferous forest \\
-& BF & broadleaf forest \\
-& MF & mixed forest \\
-hydrological regime & RD & rain dominated \\
-& SD & snow dominated \\
-climate type & EL & energy limited \\
-& WL & water limited \\
-& EQ & equitant \\
-precipitation data type & OB & observed \\
-& SG & spatial gridded \\
-& MD & modelled \\
-assessment technique & PWE & paired watershed experiment \\
-& QPW & quasi-paired watershed experiment \\
-& HM & hydrological modelling \\
-& EA & elastictity analysis \\
-& SH & combined use of statistical methods \\
-& & and hydrographs \\
+forest type & CF & coniferous forest\tabularnewline
+& BF & broadleaf forest\tabularnewline
+& MF & mixed forest\tabularnewline
+hydrological regime & RD & rain dominated\tabularnewline
+& SD & snow dominated\tabularnewline
+climate type & EL & energy limited\tabularnewline
+& WL & water limited\tabularnewline
+& EQ & equitant\tabularnewline
+precipitation data type & OB & observed\tabularnewline
+& SG & spatial gridded\tabularnewline
+& MD & modelled\tabularnewline
+assessment technique & PWE & paired watershed experiment\tabularnewline
+& QPW & quasi-paired watershed experiment\tabularnewline
+& HM & hydrological modelling\tabularnewline
+& EA & elastictity analysis\tabularnewline
+& SH & combined use of statistical methods\tabularnewline
+& & and hydrographs\tabularnewline
 \bottomrule
@@
 
-The length of the study can be a variable influencing the change in flow (Filoso et al., 2017; e.g.
-Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987).
+The length of the study can be a variable influencing the change in flow (e.g.~Jackson et al., 2005; Filoso et al., 2017), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987).
 Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al.
@@
 \hypertarget{description-of-the-data}{%
-\subsection{description of the data}\label{description-of-the-data}}
+\subsection{Description of the data}\label{description-of-the-data}}
 
@@
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2}.
+Smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).
 
@@
 
+In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2).
+In other words, flow almost always increased.
+However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
+However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
+
 \begin{figure}
@@
 
-This shows that for the data related to forest decreases, there is almost always a positive flow change.
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
-
 \hypertarget{the-general-relationship-between-change-in-forest-cover-and-streamflow}{%
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
 The first three rows relate to the model using the original data base from Zhang et al.~(2017).
-The bottom three rows are the results of the model including the new data}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+The bottom three rows are the results of the model including the new data.}\tabularnewline
+\toprule
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
-\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
-\textbf{(Intercept)} & 9.43 & 5.54 & 1.7 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.09 & 5.12 & 0 \\
-\textbf{Forest\_Signincrease} & -36.54 & 5.59 & -6.53 & 0 \\
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+8.65\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.56\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.56\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.12\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.45\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.26\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-29.17\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.79\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.04\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+9.43\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.54\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.7\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.09\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.44\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-36.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.59\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.53\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:out-model3} Results of the model including Latitude and Longitude including new data}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & 10.13 & 6.01 & 1.69 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.22 & 0 \\
-\textbf{Forest\_Signincrease} & -38.15 & 5.95 & -6.41 & 0 \\
-\textbf{Latitude} & -0.04 & 0.09 & -0.4 & 0.69 \\
-\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+10.13\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.01\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.69\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.09\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.45\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+5.22\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-38.15\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.95\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.41\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Latitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-0.04\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.69\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Longitude}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.01\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.03\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.26\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.79\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & 3.18 & 6.77 & 0.47 & 0.64 \\
-\textbf{DeltaF\_perc\_pos} & 0.43 & 0.09 & 4.91 & 0 \\
-\textbf{Forest\_Signincrease} & -37.43 & 5.68 & -6.59 & 0 \\
-\textbf{Dryness} & 5.56 & 3.06 & 1.82 & 0.07 \\
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+3.18\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+6.77\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.47\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.64\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.43\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.09\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+4.91\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-37.43\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.68\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.59\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Dryness}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+5.56\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+3.06\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.82\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.07\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.46}}@{}}
+\begin{longtable}[]{@{}ccc@{}}
 \caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-Latitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Longitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Watershed name
-\end{minipage} \\
+\begin{minipage}[b]{0.14\columnwidth}\centering
+Latitude\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Longitude\strut
+\end{minipage} & \begin{minipage}[b]{0.42\columnwidth}\centering
+Watershed name\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-Latitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Longitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Watershed name
-\end{minipage} \\
+\begin{minipage}[b]{0.14\columnwidth}\centering
+Latitude\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Longitude\strut
+\end{minipage} & \begin{minipage}[b]{0.42\columnwidth}\centering
+Watershed name\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
-36.4 & -120.4 & Cantua \\
-34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
-32.74 & -111.5 & Natural DRDages, Ariz., U.S.A,
-A \\
+\begin{minipage}[t]{0.14\columnwidth}\centering
+34.67\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-111.7\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Beaver Creek, AZ \#3-2\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+36.4\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-120.4\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Cantua\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+34.43\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-112.3\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+White Spar, Ariz., U.S.A, B\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\centering
+32.74\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+-111.5\strut
+\end{minipage} & \begin{minipage}[t]{0.42\columnwidth}\centering
+Natural DRDages, Ariz., U.S.A,
+A\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & 10.46 & 7.66 & 1.36 & 0.17 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.11 & 2.93 & 0 \\
-\textbf{Forest\_Signincrease} & -35.67 & 5.73 & -6.23 & 0 \\
-\textbf{Dryness} & 6.54 & 3.08 & 2.12 & 0.03 \\
-\textbf{log10(Area\_km2)} & -3.5 & 1.76 & -1.99 & 0.05 \\
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+10.46\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+7.66\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.36\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.17\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.93\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-35.67\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.73\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-6.23\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Dryness}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+6.54\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+3.08\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.12\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.03\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{log10(Area\_km2)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-3.5\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+1.76\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.99\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.05\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.31\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & 17.37 & 5.91 & 2.94 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
-\textbf{Forest\_Signincrease} & -34.93 & 5.84 & -5.98 & 0 \\
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+17.37\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.91\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.94\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.26\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+0.11\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.44\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.31\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-34.93\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+5.84\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-5.98\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{s(log10(Area\_km2))} & 2.48 & 9 & 0.55 & 0.1 \\
-\textbf{s(Dryness)} & 0.89 & 9 & 0.62 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.66 \\
-\textbf{s(From)} & 8.1 & 9 & 3.97 & 0 \\
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+2.48\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.55\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.89\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.62\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.01\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.66\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(From)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+8.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.97\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.42}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.36\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.36\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & -21.06 & 19.16 & -1.1 & 0.27 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 2.99 & 0 \\
-\textbf{Forest\_Signincrease} & -23.11 & 7.13 & -3.24 & 0 \\
-\textbf{Precip\_data\_typeOB} & -7.65 & 14.51 & -0.53 & 0.6 \\
-\textbf{Precip\_data\_typeSG} & 18.92 & 16.17 & 1.17 & 0.24 \\
-\textbf{Assessment\_techniqueEA, HM} & 17.73 & 45.71 & 0.39 & 0.7 \\
-\textbf{Assessment\_techniqueHM} & 29.82 & 12.66 & 2.36 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 44.69 & 12.68 & 3.53 & 0 \\
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-21.06\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+19.16\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-1.1\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.27\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.31\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.99\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-23.11\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+7.13\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-3.24\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeOB}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-7.65\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+14.51\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.53\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.6\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Precip\_data\_typeSG}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+18.92\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+16.17\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.17\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.24\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueEA, HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+17.73\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+45.71\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.39\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.7\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueHM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+29.82\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.66\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.36\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniquePWE}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+44.69\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12.68\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.53\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
 \textbf{Assessment\_techniquePWE,
-HM} & 39.72 & 46.79 & 0.85 & 0.4 \\
-\textbf{Assessment\_techniqueQPW} & 37.54 & 21.54 & 1.74 & 0.08 \\
+HM}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+39.72\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+46.79\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.85\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+37.54\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+21.54\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.74\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.08\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
 \textbf{Assessment\_techniqueQPW,
-EA} & 24.9 & 27.45 & 0.91 & 0.37 \\
-\textbf{Assessment\_techniqueSH} & 43.68 & 13.27 & 3.29 & 0 \\
-\textbf{Forest\_typeCF} & -4.06 & 8.71 & -0.47 & 0.64 \\
-\textbf{Forest\_typeMF} & -8.33 & 8.7 & -0.96 & 0.34 \\
-\textbf{Hydrological\_regimeSD} & 9.41 & 10.57 & 0.89 & 0.37 \\
+EA}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+24.9\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+27.45\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.91\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.37\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Assessment\_techniqueSH}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+43.68\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+13.27\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.29\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeCF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-4.06\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.71\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.47\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.64\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Forest\_typeMF}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-8.33\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+8.7\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.96\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.34\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.36\columnwidth}\centering
+\textbf{Hydrological\_regimeSD}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+9.41\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+10.57\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.89\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.37\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.44 & 9 & 0.08 & 0.21 \\
-\textbf{s(Dryness)} & 3.5 & 9 & 1.76 & 0 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.45 \\
-\textbf{s(From)} & 8.17 & 9 & 4.13 & 0 \\
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.44\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.08\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.21\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.5\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.76\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.45\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(From)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+8.17\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+4.13\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{s(Dryness)} & 2.62 & 9 & 3.4 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.76 & 9 & 0.35 & 0.04 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.4 \\
-\textbf{s(From)} & 0.99 & 9 & 0.46 & 0.04 \\
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+2.62\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.4\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.76\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.35\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.04\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(From)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.99\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.46\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.04\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.32}}
-  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.08}}@{}}
+\begin{longtable}[]{@{}cc@{}}
 \caption{\label{tab:tableassess} Distribution of assessment techniques in the data set}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-Assessment\_technique
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-n
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+Assessment\_technique\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+n\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-Assessment\_technique
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-n
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+Assessment\_technique\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+n\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-PWE & 185 \\
-HM & 53 \\
-EA & 32 \\
-SH & 26 \\
-QPW & 7 \\
-QPW, EA & 4 \\
-EA, HM & 1 \\
-PWE, HM & 1 \\
+\begin{minipage}[t]{0.30\columnwidth}\centering
+PWE\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+185\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+HM\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+53\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+EA\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+32\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+SH\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+26\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+QPW\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+7\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+QPW, EA\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+EA, HM\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+PWE, HM\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.40}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.34\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
+\begin{minipage}[b]{0.34\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Estimate\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Std. Error\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+t value\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Pr(\textgreater\textbar t\textbar)\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{(Intercept)} & -16.88 & 19.02 & -0.89 & 0.38 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.85 & 0 \\
-\textbf{Forest\_Signincrease} & -22.24 & 7.15 & -3.11 & 0 \\
-\textbf{Precip\_data\_typeOB} & -13.73 & 14.27 & -0.96 & 0.34 \\
-\textbf{Precip\_data\_typeSG} & 17.94 & 16.44 & 1.09 & 0.28 \\
-\textbf{Assessment\_techniqueHM} & 31.06 & 12.73 & 2.44 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 44.97 & 13.06 & 3.44 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 34.75 & 21.26 & 1.63 & 0.1 \\
-\textbf{Assessment\_techniqueSH} & 48.06 & 13.16 & 3.65 & 0 \\
-\textbf{Forest\_typeCF} & -6.9 & 8.5 & -0.81 & 0.42 \\
-\textbf{Forest\_typeMF} & -4.53 & 8.68 & -0.52 & 0.6 \\
-\textbf{Hydrological\_regimeSD} & 10.3 & 10.46 & 0.99 & 0.33 \\
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{(Intercept)}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-16.88\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+19.02\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.89\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.38\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{DeltaF\_perc\_pos}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.29\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+0.1\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.85\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_Signincrease}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-22.24\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+7.15\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-3.11\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Precip\_data\_typeOB}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-13.73\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+14.27\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.96\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.34\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Precip\_data\_typeSG}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+17.94\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+16.44\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.09\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.28\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueHM}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+31.06\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+12.73\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2.44\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.02\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniquePWE}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+44.97\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+13.06\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.44\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueQPW}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+34.75\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+21.26\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1.63\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Assessment\_techniqueSH}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+48.06\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+13.16\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3.65\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_typeCF}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-6.9\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+8.5\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.81\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.42\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Forest\_typeMF}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+-4.53\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+8.68\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.52\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.6\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.34\columnwidth}\centering
+\textbf{Hydrological\_regimeSD}\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+10.3\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+10.46\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.99\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+0.33\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\begin{longtable}[]{@{}ccccc@{}}
 \caption{\label{tab:model8-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+edf\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+Ref.df\strut
+\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\centering
+F\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+p-value\strut
+\end{minipage}\tabularnewline
 \midrule
 \endhead
-\textbf{s(Dryness)} & 3.55 & 9 & 1.82 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.75 & 9 & 0.27 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.37 \\
-\textbf{s(From)} & 0.79 & 9 & 0.26 & 0.08 \\
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(Dryness)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+3.55\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+1.82\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(log10(Area\_km2))}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.75\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.27\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.07\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(length)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.37\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{s(From)}\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.79\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+9\strut
+\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\centering
+0.26\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.08\strut
+\end{minipage}\tabularnewline
 \bottomrule
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-andreassian2004}{}}%
-Andréassian, V., 2004. Waters and forests: From historical controversy to scientific debate. Journal of Hydrology 291, 1--27. doi:\url{https://doi.org/10.1016/j.jhydrol.2003.12.015}
-
-\leavevmode\vadjust pre{\hypertarget{ref-baker1984}{}}%
-Baker Jr., M.B., 1984. Changes in streamflow in an herbicide-treated pinyon-juniper watershed in arizona. Water Resources Research 20, 1639--1642. doi:\url{https://doi.org/10.1029/WR020i011p01639}
-
-\leavevmode\vadjust pre{\hypertarget{ref-borg1988}{}}%
+\leavevmode\hypertarget{ref-andreassian2004}{}%
+Andréassian, V., 2004.
+Waters and forests: From historical controversy to scientific debate.
+Journal of Hydrology 291, 1--27.
+doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2003.12.015}{https://doi.org/10.1016/j.jhydrol.2003.12.015}
+
+\leavevmode\hypertarget{ref-baker1984}{}%
+Baker Jr., M.B., 1984.
+Changes in streamflow in an herbicide-treated pinyon-juniper watershed in arizona.
+Water Resources Research 20, 1639--1642.
+doi:\href{https://doi.org/https://doi.org/10.1029/WR020i011p01639}{https://doi.org/10.1029/WR020i011p01639}
+
+\leavevmode\hypertarget{ref-borg1988}{}%
 Borg, H., Bell, R.W., Loh, I.C., 1988.
@@
 Journal of Hydrology 103, 323--333.
-doi:\url{https://doi.org/10.1016/0022-1694(88)90141-2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hewlett1984}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(88)90141-2}{https://doi.org/10.1016/0022-1694(88)90141-2}
+
+\leavevmode\hypertarget{ref-hewlett1984}{}%
 Bosch, J.M., Hewlett, J.D., 1982. A review of catchment experiments to determine the effect of vegetation changes on water yield and evapotranspiration. Journal of Hydrology 55, 3--23.
 
-\leavevmode\vadjust pre{\hypertarget{ref-brown2013}{}}%
+\leavevmode\hypertarget{ref-brown2013}{}%
 Brown, A.E., Western, A.W., McMahon, T.A., Zhang, L., 2013.
@@
 Journal of Hydrology 483, 39--50.
-doi:\url{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}
-
-\leavevmode\vadjust pre{\hypertarget{ref-brown2005}{}}%
+doi:\href{https://doi.org/http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}
+
+\leavevmode\hypertarget{ref-brown2005}{}%
 Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A., 2005.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cosandey2005}{}}%
+\leavevmode\hypertarget{ref-cosandey2005}{}%
 Cosandey, C., Andréassian, V., Martin, C., Didon-Lescot, J.F., Lavabre, J., Folton, N., Mathys, N., Richard, D., 2005.
@@
 Journal of Hydrology 301, 235--249.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2004.06.040}
-
-\leavevmode\vadjust pre{\hypertarget{ref-filoso2017}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2004.06.040}{https://doi.org/10.1016/j.jhydrol.2004.06.040}
+
+\leavevmode\hypertarget{ref-filoso2017}{}%
 Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
+\leavevmode\hypertarget{ref-jackson2005}{}%
 Jackson, R.B., Jobbagy, E.G., Avissar, R., Roy, S.B., Barrett, D.J., Cook, C.W., Farley, K.A., Maitre, D.C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kuczera1987}{}}%
+\leavevmode\hypertarget{ref-kuczera1987}{}%
 Kuczera, G., 1987.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-navas2019}{}}%
+\leavevmode\hypertarget{ref-navas2019}{}%
 Navas, R., Alonso, J., Gorgoglione, A., Vervoort, R.W., 2019. Identifying climate and human impact trends in streamflow: A case study in uruguay. Water 11, 1433.
 
-\leavevmode\vadjust pre{\hypertarget{ref-pena-arancibia2012}{}}%
+\leavevmode\hypertarget{ref-pena-arancibia2012}{}%
 Peña-Arancibia, J.L., Dijk, A.I.J.M.
@@
 Journal of Hydrology 416-417, 60--71.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2011.11.036}
-
-\leavevmode\vadjust pre{\hypertarget{ref-roche1981}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2011.11.036}{https://doi.org/10.1016/j.jhydrol.2011.11.036}
+
+\leavevmode\hypertarget{ref-roche1981}{}%
 Roche, M., 1981. Watershed investigations for development of forest resources of the amazon region in french guyana. Tropical Agricultural Hydrology. J 75--82.
 
-\leavevmode\vadjust pre{\hypertarget{ref-rodriguez2010}{}}%
+\leavevmode\hypertarget{ref-rodriguez2010}{}%
 Rodriguez, D.A., Tomasella, J., Linhares, C., 2010.
@@
 Hydrological Processes 24, 1254--1269.
-doi:\url{https://doi.org/10.1002/hyp.7586}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ruprechtetal1991}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1002/hyp.7586}{https://doi.org/10.1002/hyp.7586}
+
+\leavevmode\hypertarget{ref-ruprechtetal1991}{}%
 Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A., Stoneman, G.L., 1991.
@@
 Journal of Hydrology 127, 261--277.
-doi:\url{https://doi.org/10.1016/0022-1694(91)90118-2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-thornton2007}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(91)90118-2}{https://doi.org/10.1016/0022-1694(91)90118-2}
+
+\leavevmode\hypertarget{ref-thornton2007}{}%
 Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-trabucco2018}{}}%
-Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential evapo-transpiration (ET0) climate database v2. CGIAR consortium for spatial information(CGIAR-CSI).
-
-\leavevmode\vadjust pre{\hypertarget{ref-vervoort2021}{}}%
+\leavevmode\hypertarget{ref-trabucco2018}{}%
+Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential evapo-transpiration (et0) climate database v2. CGIAR consortium for spatial information(CGIAR-csi).
+
+\leavevmode\hypertarget{ref-vervoort2021}{}%
 Vervoort, R.W., Dolk, M.M., Ogtrop, F.F.
@@
 Hydrological Processes 35, e13999.
-doi:\url{https://doi.org/10.1002/hyp.13999}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2006}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1002/hyp.13999}{https://doi.org/10.1002/hyp.13999}
+
+\leavevmode\hypertarget{ref-wood2006}{}%
 Wood, S., 2006. Generalized additive models: An introduction with r. CRC Press, Boca Raton, FL.
 
-\leavevmode\vadjust pre{\hypertarget{ref-zhang2011}{}}%
+\leavevmode\hypertarget{ref-zhang2011}{}%
 Zhang, L., Zhao, F., Chen, Y., Dixon, R.N.M., 2011.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zhang2017}{}}%
+\leavevmode\hypertarget{ref-zhang2017}{}%
 Zhang, M., Liu, N., Harper, R., Li, Q., Liu, K., Wei, X., Ning, D., Hou, Y., Liu, S., 2017.
@@
 Journal of Hydrology 546, 44--59.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2016.12.040}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhao2010}{}}%
+doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2016.12.040}{https://doi.org/10.1016/j.jhydrol.2016.12.040}
+
+\leavevmode\hypertarget{ref-zhao2010}{}%
 Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zhou2015}{}}%
+\leavevmode\hypertarget{ref-zhou2015}{}%
 Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G., Scott, D.F., Zhou, S., Han, L., Su, Y., 2015.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zhou2010}{}}%
+\leavevmode\hypertarget{ref-zhou2010}{}%
 Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang, C., 2010.
@@
 Water Resources Research 46.
-doi:\url{https://doi.org/10.1029/2009WR008829}
-
-\end{CSLReferences}
+doi:\href{https://doi.org/https://doi.org/10.1029/2009WR008829}{https://doi.org/10.1029/2009WR008829}
 

### 2022-02-02 paper/Forest_and_Water/Forest_and_Water.Rmd
updates paper 02022022

--- 2022-01-12 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
 
-In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
-Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
+In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location.
+Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018].
 For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
-Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.
+Finally, on review of all the data in @zhang2017 and @filoso2017 and the additional studies, 29 potential duplicates were identified and flagged in the data.
 
@@
 <!-- read in the ET data -->
-
+<!-- no longer needed as this is now included in the data
 ```{r, echo = F, warning = F, message=F}
@@
 ```
+--->
 
@@
 
+```{r, eval = F, echo = F}
+# checking missing values
+summary(All_data)
+
+```
+
 <!--now combine with E0 values -->
+<!-- No longer needed as now included in the data -->
 ```{r, echo = F}
-All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
-All_data <- All_data %>%
-  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
-  select(-c(E0.x,E0.y)) #%>%
-  #rename(Latitude = Latitude.x,
-#         Longitude = Longitude.x)
+# All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
+# All_data <- All_data %>%
+#   mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
+#   select(-c(E0.x,E0.y)) #%>%
+#   #rename(Latitude = Latitude.x,
+# #         Longitude = Longitude.x)
 
@@
       "DeltaQf_perc" = "Delta Qf (%)",
-      "E0" = "Annual Potential E (mm)",
+      "E0" = "Annual E0 (mm)",
       "Pa_mm" = "Annual Rainfall (mm)"))) +   
@@
   mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
-                              "decrease", "increase"),
+                              "Decrease", "Increase"),
          DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
@@
 All_data2 %>%
+  filter(is.na(DeltaF_perc_pos) == F) %>%
   ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
@@
   scale_colour_manual(name = "Forest change direction", 
-                      values = c("increase" = "blue", "decrease" = "red")) +
+                      values = c("Decrease" = "red", "Increase" = "blue")) +
   xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
@@
 
-include_graphics("../../data/FAOET0data2.png")
+include_graphics("../../data/FAOET0data.png")
 ```
@@
 ```
+```{r testing_dryness, echo = F, eval= F}
+All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
+```
+```{r testing_PCA, eval = F, echo = F, warning = F}
+require(FactoMineR)
+require(factoextra)
+ res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
+ # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
+ #  scree.plot
+ #  
+ #   var.plot <- fviz_pca_var(res.pca, axes = c(1, 2), col.var = "black")
+ #  var.plot
+ png("PCA_biplot_Alldata.png", width = 800, height = 800)
+  biplot_test <- fviz_pca_biplot(res.pca, axes = c(1, 2), repel = TRUE,
+                                                      col.var = "darkblue", # Variables color
+                            label="var", #"var" 
+                            col.ind = "red", # Individuals color
+                            title = "",labelsize = 8, pointsize = 4, arrowsize = 1) +
+        theme(text = element_text(size = 14, face = "bold"),
+              axis.title = element_text(size = rel(2), face = "bold"),
+        axis.text = element_text(size = rel(2), face = "bold"))
+ print(biplot_test)
+ dev.off()
 
-The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
-In this case the evidence is not very strong (p = `r round(summary(model4_dryness)$p.table[4,4],2)`.
-However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is ~6\% greater.
+```
+
+The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results [@zhang2017; @filoso2017].
+In this case the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
+However, it is very well possible that there is a further interaction in the data that this simpler version of the model cannot identify.
 Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
@@
 
-There possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
+There are also possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \@ref(tab:drytable).
@@
 Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
+The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.
 
@@
 
-The results of this model (Equation \@ref(eq:eq6)) clearly indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)).
+The results of this model (Equation \@ref(eq:eq6)) indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)).
 In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%.
@@
 ```{r model7_nolongLength, echo = F}
-NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
+NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1930)
 model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
 ```
-```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}
+```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model with reduced data for dryness and length", out.width="90%", echo=F}
 
@@
 ```
+
+```{r model9_interact, echo = F, eval = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
+
+model9_interact <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
+                      ti(Dryness,From, bs="ts" ) + #s(Latitude, bs="ts") + 
+                     # ti(log10(Area_km2),From, bs="ts" ) +
+                     # log10(Area_km2):Assessment_technique +
+                      s(length, bs = "ts") + 
+                      s(From, bs = "ts") +
+                    Precip_data_type + Forest_type + Assessment_technique +
+                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4))
+#gam.check(model9_interact)
+# plot(model9_interact)
+summary(model9_interact)
+
+```

### 2022-02-02 paper/Forest_and_Water/Forest_and_Water.tex
updates paper 02022022

--- 2022-01-12 paper/Forest_and_Water/Forest_and_Water.tex
@@
 
-In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
-Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
+In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location.
+Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018).
 For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 (2017) and Filoso et al.
-(2017), 29 potential duplicates were identified and flagged in the data.
+(2017) and the additional studies, 29 potential duplicates were identified and flagged in the data.
 
@@
 
-The overall dataset contains 309 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+The overall dataset contains 355 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
 In contrast, the original dataset from Zhang et al.
@@
 \endhead
-\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
-\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
-\textbf{(Intercept)} & 9.43 & 5.54 & 1.7 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.09 & 5.12 & 0 \\
-\textbf{Forest\_Signincrease} & -36.54 & 5.59 & -6.53 & 0 \\
+\textbf{(Intercept)} & 8.66 & 5.52 & 1.57 & 0.12 \\
+\textbf{DeltaF\_perc\_pos} & 0.5 & 0.09 & 5.81 & 0 \\
+\textbf{Forest\_SignIncrease} & -31 & 5.87 & -5.28 & 0 \\
+\textbf{(Intercept)} & 10.74 & 5.29 & 2.03 & 0.04 \\
+\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.6 & 0 \\
+\textbf{Forest\_SignIncrease} & -39.7 & 5.3 & -7.49 & 0 \\
 \bottomrule
@@
 
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.23, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data2} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
 \end{figure}
@@
 \endhead
-\textbf{(Intercept)} & 10.13 & 6.01 & 1.69 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.22 & 0 \\
-\textbf{Forest\_Signincrease} & -38.15 & 5.95 & -6.41 & 0 \\
-\textbf{Latitude} & -0.04 & 0.09 & -0.4 & 0.69 \\
-\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
+\textbf{(Intercept)} & 12.02 & 5.66 & 2.12 & 0.03 \\
+\textbf{DeltaF\_perc\_pos} & 0.47 & 0.08 & 5.67 & 0 \\
+\textbf{Forest\_SignIncrease} & -41.77 & 5.53 & -7.55 & 0 \\
+\textbf{Latitude} & -0.05 & 0.09 & -0.55 & 0.58 \\
+\textbf{Longitude} & 0.02 & 0.03 & 0.56 & 0.58 \\
 \bottomrule
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.22 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.24 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 \endhead
-\textbf{(Intercept)} & 3.18 & 6.77 & 0.47 & 0.64 \\
-\textbf{DeltaF\_perc\_pos} & 0.43 & 0.09 & 4.91 & 0 \\
-\textbf{Forest\_Signincrease} & -37.43 & 5.68 & -6.59 & 0 \\
-\textbf{Dryness} & 5.56 & 3.06 & 1.82 & 0.07 \\
+\textbf{(Intercept)} & 8.43 & 6.39 & 1.32 & 0.19 \\
+\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.58 & 0 \\
+\textbf{Forest\_SignIncrease} & -40.3 & 5.37 & -7.51 & 0 \\
+\textbf{Dryness} & 1.81 & 2.73 & 0.66 & 0.51 \\
 \bottomrule
@@
 
-The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
-In this case the evidence is not very strong (p = 0.07.
-However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is \textasciitilde6\% greater.
-Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
+The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+In this case the evidence is highly doubtful (p = 0.51).
+However, it is very well possible that there is a further interaction in the data that this simpler version of the model cannot identify.
+Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
 
@@
 36.4 & -120.4 & Cantua \\
-34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
 32.74 & -111.5 & Natural DRDages, Ariz., U.S.A,
 A \\
+32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, C \\
+34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+-25.75 & 28.23 & Queens river \\
 \bottomrule
@@
 
-There possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
+There are also possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \ref{tab:drytable}.
@@
 Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.
+The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.
 
@@
 \endhead
-\textbf{(Intercept)} & 10.46 & 7.66 & 1.36 & 0.17 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.11 & 2.93 & 0 \\
-\textbf{Forest\_Signincrease} & -35.67 & 5.73 & -6.23 & 0 \\
-\textbf{Dryness} & 6.54 & 3.08 & 2.12 & 0.03 \\
-\textbf{log10(Area\_km2)} & -3.5 & 1.76 & -1.99 & 0.05 \\
+\textbf{(Intercept)} & 13.86 & 7.2 & 1.92 & 0.06 \\
+\textbf{DeltaF\_perc\_pos} & 0.38 & 0.1 & 3.77 & 0 \\
+\textbf{Forest\_SignIncrease} & -38.44 & 5.48 & -7.02 & 0 \\
+\textbf{Dryness} & 2.45 & 2.75 & 0.89 & 0.37 \\
+\textbf{log10(Area\_km2)} & -2.75 & 1.7 & -1.62 & 0.11 \\
 \bottomrule
@@
 
-The results of this model (Equation \eqref{eq:eq6}) clearly indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+The results of this model (Equation \eqref{eq:eq6}) indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
 In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.5\%.
@@
 \endhead
-\textbf{(Intercept)} & 17.37 & 5.91 & 2.94 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.26 & 0.11 & 2.44 & 0.02 \\
-\textbf{Forest\_Signincrease} & -34.93 & 5.84 & -5.98 & 0 \\
+\textbf{(Intercept)} & 18.46 & 5.47 & 3.38 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.24 & 0.1 & 2.43 & 0.02 \\
+\textbf{Forest\_SignIncrease} & -38.44 & 5.71 & -6.74 & 0 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 2.48 & 9 & 0.55 & 0.1 \\
-\textbf{s(Dryness)} & 0.89 & 9 & 0.62 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.66 \\
-\textbf{s(From)} & 8.1 & 9 & 3.97 & 0 \\
+\textbf{s(log10(Area\_km2))} & 3.6 & 9 & 0.88 & 0.05 \\
+\textbf{s(Dryness)} & 0 & 9 & 0 & 0.74 \\
+\textbf{s(length)} & 1.87 & 9 & 0.44 & 0.09 \\
+\textbf{s(From)} & 7.22 & 9 & 1.84 & 0.01 \\
 \bottomrule
@@
 
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.31 and deviance explained of 0.34, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.26 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
 For example, the smoothed variable for Area (km\textsuperscript{2}) is no longer a strong effect explaining variations in changes in stream flow.
@@
 \endhead
-\textbf{(Intercept)} & -21.06 & 19.16 & -1.1 & 0.27 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.1 & 2.99 & 0 \\
-\textbf{Forest\_Signincrease} & -23.11 & 7.13 & -3.24 & 0 \\
-\textbf{Precip\_data\_typeOB} & -7.65 & 14.51 & -0.53 & 0.6 \\
-\textbf{Precip\_data\_typeSG} & 18.92 & 16.17 & 1.17 & 0.24 \\
-\textbf{Assessment\_techniqueEA, HM} & 17.73 & 45.71 & 0.39 & 0.7 \\
-\textbf{Assessment\_techniqueHM} & 29.82 & 12.66 & 2.36 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 44.69 & 12.68 & 3.53 & 0 \\
+\textbf{(Intercept)} & -23.61 & 18.27 & -1.29 & 0.2 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.16 & 0 \\
+\textbf{Forest\_SignIncrease} & -26.28 & 6.64 & -3.96 & 0 \\
+\textbf{Precip\_data\_typeOB} & -7.03 & 13.51 & -0.52 & 0.6 \\
+\textbf{Precip\_data\_typeSG} & 13.98 & 15.5 & 0.9 & 0.37 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.74 & 45.33 & 0.41 & 0.68 \\
+\textbf{Assessment\_techniqueHM} & 25.27 & 11.91 & 2.12 & 0.03 \\
+\textbf{Assessment\_techniquePWE} & 54.21 & 11.81 & 4.59 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 39.72 & 46.79 & 0.85 & 0.4 \\
-\textbf{Assessment\_techniqueQPW} & 37.54 & 21.54 & 1.74 & 0.08 \\
+HM} & 31.99 & 46.11 & 0.69 & 0.49 \\
+\textbf{Assessment\_techniqueQPW} & 29.16 & 20.96 & 1.39 & 0.17 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 24.9 & 27.45 & 0.91 & 0.37 \\
-\textbf{Assessment\_techniqueSH} & 43.68 & 13.27 & 3.29 & 0 \\
-\textbf{Forest\_typeCF} & -4.06 & 8.71 & -0.47 & 0.64 \\
-\textbf{Forest\_typeMF} & -8.33 & 8.7 & -0.96 & 0.34 \\
-\textbf{Hydrological\_regimeSD} & 9.41 & 10.57 & 0.89 & 0.37 \\
+EA} & 34.71 & 34.04 & 1.02 & 0.31 \\
+\textbf{Assessment\_techniqueSH} & 37.64 & 11.87 & 3.17 & 0 \\
+\textbf{Forest\_typeCF} & -11.48 & 8.07 & -1.42 & 0.16 \\
+\textbf{Forest\_typeMF} & -8.65 & 7.72 & -1.12 & 0.26 \\
+\textbf{Hydrological\_regimeSD} & 13.8 & 10.15 & 1.36 & 0.18 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.44 & 9 & 0.08 & 0.21 \\
-\textbf{s(Dryness)} & 3.5 & 9 & 1.76 & 0 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.45 \\
-\textbf{s(From)} & 8.17 & 9 & 4.13 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.32 & 9 & 0.05 & 0.25 \\
+\textbf{s(Dryness)} & 0.48 & 9 & 0.09 & 0.19 \\
+\textbf{s(length)} & 0.88 & 9 & 0.69 & 0.01 \\
+\textbf{s(From)} & 1.79 & 9 & 0.96 & 0.01 \\
 \bottomrule
@@
 
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.4.
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.31.
 This indicates that the categorical variables explain a limited amount of the variance.
@@
 \endhead
-\textbf{s(Dryness)} & 2.62 & 9 & 3.4 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.76 & 9 & 0.35 & 0.04 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.4 \\
-\textbf{s(From)} & 0.99 & 9 & 0.46 & 0.04 \\
+\textbf{s(Dryness)} & 0.2 & 9 & 0.03 & 0.25 \\
+\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.33 \\
+\textbf{s(length)} & 3.81 & 9 & 1.19 & 0.02 \\
+\textbf{s(From)} & 3.94 & 9 & 1.31 & 0.01 \\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model wiht reduced data for dryness and length}\label{fig:smoothsmodel7}
+\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model with reduced data for dryness and length}\label{fig:smoothsmodel7}
 \end{figure}
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.34 with a slope of -9.63) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.33 with a slope of -9.56) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is 10\% lower.
 This is basically a result of the fact that large changes in forest cover are difficult to ``implement'' in an experiment.
@@
 \endhead
-PWE & 185 \\
-HM & 53 \\
-EA & 32 \\
-SH & 26 \\
+PWE & 208 \\
+HM & 57 \\
+SH & 43 \\
+EA & 34 \\
 QPW & 7 \\
@@
 \endhead
-\textbf{(Intercept)} & -16.88 & 19.02 & -0.89 & 0.38 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.85 & 0 \\
-\textbf{Forest\_Signincrease} & -22.24 & 7.15 & -3.11 & 0 \\
-\textbf{Precip\_data\_typeOB} & -13.73 & 14.27 & -0.96 & 0.34 \\
-\textbf{Precip\_data\_typeSG} & 17.94 & 16.44 & 1.09 & 0.28 \\
-\textbf{Assessment\_techniqueHM} & 31.06 & 12.73 & 2.44 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 44.97 & 13.06 & 3.44 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 34.75 & 21.26 & 1.63 & 0.1 \\
-\textbf{Assessment\_techniqueSH} & 48.06 & 13.16 & 3.65 & 0 \\
-\textbf{Forest\_typeCF} & -6.9 & 8.5 & -0.81 & 0.42 \\
-\textbf{Forest\_typeMF} & -4.53 & 8.68 & -0.52 & 0.6 \\
-\textbf{Hydrological\_regimeSD} & 10.3 & 10.46 & 0.99 & 0.33 \\
+\textbf{(Intercept)} & -32.33 & 18.4 & -1.76 & 0.08 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.22 & 0 \\
+\textbf{Forest\_SignIncrease} & -26.07 & 6.79 & -3.84 & 0 \\
+\textbf{Precip\_data\_typeOB} & 0.12 & 13.84 & 0.01 & 0.99 \\
+\textbf{Precip\_data\_typeSG} & 21.54 & 15.96 & 1.35 & 0.18 \\
+\textbf{Assessment\_techniqueHM} & 23.08 & 12.26 & 1.88 & 0.06 \\
+\textbf{Assessment\_techniquePWE} & 56.89 & 11.19 & 5.09 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 29.72 & 21.04 & 1.41 & 0.16 \\
+\textbf{Assessment\_techniqueSH} & 35.41 & 11.7 & 3.03 & 0 \\
+\textbf{Forest\_typeCF} & -9.41 & 8.06 & -1.17 & 0.24 \\
+\textbf{Forest\_typeMF} & -6.14 & 7.96 & -0.77 & 0.44 \\
+\textbf{Hydrological\_regimeSD} & 11.07 & 10.07 & 1.1 & 0.27 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 3.55 & 9 & 1.82 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.75 & 9 & 0.27 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.37 \\
-\textbf{s(From)} & 0.79 & 9 & 0.26 & 0.08 \\
+\textbf{s(Dryness)} & 0 & 9 & 0 & 0.31 \\
+\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.35 \\
+\textbf{s(length)} & 4.58 & 9 & 1.59 & 0.01 \\
+\textbf{s(From)} & 0.87 & 9 & 0.64 & 0.01 \\
 \bottomrule

### 2022-02-02 paper/Forest_and_Water/Forest_and_Water.Rmd
Update Forest_and_Water.Rmd
Figure 2 label corrected
Figure 4 modified with new data and all catchments

@@
 
-This shows that for the data related to forest decreases, there is almost always a positive flow change.
+In addition, Figure 2 shows that for the data related to forest decreases, there is almost always a positive flow change.
 In other words, flow almost always increased.
@@
 
-include_graphics("../../data/FAOET0data.png")
+include_graphics("../../data/FAOET0data_final_2022.png")
 ```

### 2022-02-02 paper/Forest_and_Water/Forest_and_Water.Rmd
Merge branch 'main' of https://github.com/WillemVervoort/Forest_and_water

--- 2022-02-01 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
   scale_colour_manual(name = "Forest change direction", 
-                      values = c("decrease" = "red","increase" = "blue")) +
+                      values = c("Decrease" = "red","Increase" = "blue")) +
   xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")

### 2022-02-03 paper/Forest_and_Water/Forest_and_Water.Rmd
update paper 20220203

--- 2022-02-02 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
 
-Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both @zhang2017 and @jackson2005.
+Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
 Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
@@
 ```
-```{r testing_dryness, echo = F, eval= F}
-All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
-```
+<!-- running PCA, this is used later -->
+
 ```{r testing_PCA, eval = F, echo = F, warning = F}
@@
 require(factoextra)
- res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
+ res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length, Latitude, Longitude), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
  # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
@@
 
-The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results [@zhang2017; @filoso2017].
-In this case the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
-However, it is very well possible that there is a further interaction in the data that this simpler version of the model cannot identify.
-Again, overall variance explained is not very much, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
+The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results [@zhang2017; @filoso2017].
+In this case, the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
+However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
 
@@
 
-There are also possible issues with the data, a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
+There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \@ref(tab:drytable).
@@
 
-The results of this model (Equation \@ref(eq:eq6)) indicate a reduction in the effect of forest cover change with Area (km^2^) (Table \@ref(tab:out-modelArea)).
-In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by 3.5\%.
-Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
-Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
-
-```{r modelArea_b, echo = F, eval =F}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
+```{r modelArea_int, echo = F, eval =T}
 model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                         Dryness*log10(Area_km2), data = All_data2)
-summary(model_Area_b)
+#summary(model_Area_b)
 #gam.check(model_Area)
+```
+
+The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to Area (km^2^) (Table \@ref(tab:out-modelArea)).
+In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by `r abs(round(summary(model_Area)$p.table[5,1],1))`\%.
+Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
+Including this interaction in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.
+
+```{r, echo =F}
+pander(round(summary(model_Area_b)$p.table,2), caption = "(\\#tab:out-modelArea-int) Results of the model including an interaction between Area and the dryness index")
 ```
@@
 In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (*Dryness, Area, From*) were considered non-linear in this model.
-This is also indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.
+As a result a shrinkage smoothing spline [@wood2006] was applied to these variables.
 *From* represents the starting date of the data collection.
@@
 Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
-For example, the smoothed variable for Area (km^2^) is no longer a strong effect explaining variations in changes in stream flow.
-Clearly *length* is also not explaining the variation.
-In contrast, *From*, which indicate the start date of the study (and therefore the age of the study) is a significant effect.
+For example, the smoothed variable for Area (km^2^) (p = `r round(summary(model5_all)$s.table[4,1],2)`)) and Dryness (p = `r round(summary(model5_all)$s.table[4,2],2)`))still provides reasonable explanation of changes in stream flow.
+However *length* does not explain any of the variation.
+In contrast, *From*, which indicates the start date of the study (and therefore the age of the study) has an effect (p = `r round(summary(model5_all)$s.table[4,4],2)`)).
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
@@
 
-A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).
+We now also include the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) i.e. Equation \@ref(eq:eq2).
 
@@
 
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of `r round(summary(model6_all)$dev.expl,2)`.
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
 This indicates that the categorical variables explain a limited amount of the variance.
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
-In this case, *Area (km^2)* is no longer a significant predictor.
+In this case, *Area (km^2)* is no longer a significant predictor, the reasons for this will be discussed later.
 
@@
 
-Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006).
+Figure \@ref(fig:smoothsmodel6) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the reason why using penalized smooths following @wood2006 is useful).
 It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
-Similarly, the length variable is not significant.
-However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+Similarly, the *length* variable is not significant.
+However, both *From* and *Dryness* variables show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+For example, in the case of *From* the relationship is clearly strongly influenced by the two data points before 1920.
+Similarly, *length* and *Dryness* have points with high leverage.
 
@@
 ```{r model7_nolongLength, echo = F}
-NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1930)
+NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1920)
 model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
 ```{r, echo = F}
-pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
+pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness > 4 and studies commencing after 1920")
 ```
@@
 The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005.
-One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness > 4 (Table \@ref(tab:restrictlength)) results in log10(Area) once again showing a significant effect with larger catchments having less impact of changes in forest cover.
-Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
-Finally, more recent studies also show greater impacts of changes in forest cover (Figure \@ref(fig:smoothsmodel7)).
+One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)).
+Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables.
+The next model therefore removes the following data: *Dryness* > 4, *length* > 60 years, *From* > 1920.
+This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
+The model has more explaining power with an adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
+This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \@ref(fig:smoothsmodel7)).
+However the other 3 variables are no longer significant due to the scatter in the data, but *length* and *From* show some strong local non-linear behaviour.
 
@@
 
-Essentially, the analysis shows that there is a clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), even though, in contrast to @zhang2017, there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+Essentially, the analysis shows that there is no clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), and, in contrast to @zhang2017, there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
 If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \@ref(fig:overview)).
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average the recorded forest cover change is 10% lower.
-This is basically a result of the fact that large changes in forest cover are difficult to "implement" in an experiment.
+The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average the recorded forest cover change is approximately  10\% lower.
+This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to "implement" in an experiment.
+
 This is also reflected in the second caveat most of the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
-These techniques for larger catchments all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
+For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 ```{r gamcheck7, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
-include_graphics("residual_plot_model1.png")
+include_graphics("residual_plot_model7.png")
 ```
@@
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60))
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4, From > 1920))
 #plot(model8_assess)
@@
 Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:model8-linear) and \@ref(tab:model8-smooth)).
-It strengthens the significance of the different assessment techniques but slightly reduces the importance of some of the smooth variables (*log10(Area (km^2^))* and *From*).
+It strengthens the significance of the different assessment techniques and *Dryness* but generally results in the same interpretation.
 Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
@@
 In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies are all on the drier end of the Dryness spectrum, highlighting another unresolved interaction in the data.
+However, as indicated earlier, the EA studies are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.
 
@@
 
+## Interactions
+
+Generally this study did not consider interactions, but the above discussion suggest that there are possible several interactions.
+The relationships between forest cover change and *Area (km^2^)* and between *Area (km^2^)* and assessment technique have already been highlighted.
+However there are further unexplored interactions between the study length and some of the variables and *From* and the other variables.
+
+A principle component analysis of the numeric data reveals some of these interactions (Figure \@ref(fig:pcaplot)), such as between *length* and *Dryness* and between *From* and *Area (km^2^)*.
+Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
+
+```{r pcaplot, fig.cap="Biplot of the first two principle components using a principle component analysis on the numerical values of the ", out.width="90%", echo=F, eval =T}
+
+include_graphics("PCA_biplot_Alldata.png")
+```
+
 ## Further considerations
 
-In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
-What is puzzling is that the effect of the start date (*From*) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
-At this point, it is unclear what this trend signifies.
-It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
-
+In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable in our final model.
 There are further confounding factors in the data, which were also classified by @filoso2017 and these might create biases in the data set that can impact the overall assessment.
@@
 However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
+This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
+By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
 
@@
 More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
+It demonstrates that analysis of large databases of essentially "aggregated data" should be considered carefully.
 
@@
 Conditional on the data, the impact of forestry on streamflow is:
+
 - Greater for forest clearing then for reforestation
-- Reduced for larger watersheds  
+
+- Possibly reduced for larger watersheds, but this is only in interaction with other factors such as climate
+
 - Increases for drier watersheds
+
 - Sensitive to the assessment method used in the historical data
+
+Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area.
+Furthermore, the current study analyses a large global dataset.
+This analysis does not exlude more local and regional effects that cannot be identified in the global data.
 
@@
   ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
+```
+
+<!-- testing whether dryness is related to Area -->
+
+```{r testing_dryness, echo = F, eval= F}
+All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
 ```
@@
                       ti(Dryness,From, bs="ts" ) + #s(Latitude, bs="ts") + 
-                     # ti(log10(Area_km2),From, bs="ts" ) +
+                      ti(log10(Area_km2),length, bs="ts" ) +
                      # log10(Area_km2):Assessment_technique +
@@
                     Precip_data_type + Forest_type + Assessment_technique +
-                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4))
+                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4, From > 1920))
 #gam.check(model9_interact)

### 2022-02-03 paper/Forest_and_Water/Forest_and_Water.tex
update paper 20220203

@@
 
-The overall dataset contains 355 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+The overall dataset contains 330 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
 In contrast, the original dataset from Zhang et al.
@@
 \endhead
-\textbf{(Intercept)} & 8.66 & 5.52 & 1.57 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.5 & 0.09 & 5.81 & 0 \\
-\textbf{Forest\_SignIncrease} & -31 & 5.87 & -5.28 & 0 \\
-\textbf{(Intercept)} & 10.74 & 5.29 & 2.03 & 0.04 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.6 & 0 \\
-\textbf{Forest\_SignIncrease} & -39.7 & 5.3 & -7.49 & 0 \\
+\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
+\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
+\textbf{Forest\_SignIncrease} & -29.17 & 5.79 & -5.04 & 0 \\
+\textbf{(Intercept)} & 10.76 & 5.34 & 2.02 & 0.04 \\
+\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.04 & 0 \\
+\textbf{Forest\_SignIncrease} & -38.03 & 5.26 & -7.23 & 0 \\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
+\includegraphics[width=0.9\linewidth]{../../data/FAOET0data_final_2022} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
 \end{figure}
@@
 \endhead
-\textbf{(Intercept)} & 12.02 & 5.66 & 2.12 & 0.03 \\
-\textbf{DeltaF\_perc\_pos} & 0.47 & 0.08 & 5.67 & 0 \\
-\textbf{Forest\_SignIncrease} & -41.77 & 5.53 & -7.55 & 0 \\
-\textbf{Latitude} & -0.05 & 0.09 & -0.55 & 0.58 \\
-\textbf{Longitude} & 0.02 & 0.03 & 0.56 & 0.58 \\
+\textbf{(Intercept)} & 10.99 & 5.75 & 1.91 & 0.06 \\
+\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.13 & 0 \\
+\textbf{Forest\_SignIncrease} & -38.99 & 5.54 & -7.04 & 0 \\
+\textbf{Latitude} & -0.01 & 0.09 & -0.13 & 0.9 \\
+\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
 \bottomrule
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.24 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.23 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 
-Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlghted by both Zhang et al.
+Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al.
 (2017) and Jackson et al.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
+\caption{\label{tab:out-model4} Results of the model including the dryness index
+}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{(Intercept)} & 8.43 & 6.39 & 1.32 & 0.19 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.58 & 0 \\
-\textbf{Forest\_SignIncrease} & -40.3 & 5.37 & -7.51 & 0 \\
-\textbf{Dryness} & 1.81 & 2.73 & 0.66 & 0.51 \\
-\bottomrule
-\end{longtable}
-
-The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case the evidence is highly doubtful (p = 0.51).
-However, it is very well possible that there is a further interaction in the data that this simpler version of the model cannot identify.
-Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
+\textbf{(Intercept)} & 9.53 & 6.38 & 1.49 & 0.14 \\
+\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 4.86 & 0 \\
+\textbf{Forest\_SignIncrease} & -39.4 & 5.42 & -7.26 & 0 \\
+\textbf{Dryness} & 1.59 & 2.7 & 0.59 & 0.56 \\
+\bottomrule
+\end{longtable}
+
+The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+In this case, the evidence is highly doubtful (p = 0.56).
+However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.46}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.42}}@{}}
 \caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
@@
 36.4 & -120.4 & Cantua \\
-32.74 & -111.5 & Natural DRDages, Ariz., U.S.A,
-A \\
 32.74 & -111.5 & Natural Drainages, Ariz.,
 U.S.A, C \\
+32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, A \\
 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
@@
 
-There are also possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
+There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \ref{tab:drytable}.
@@
 \endhead
-\textbf{(Intercept)} & 13.86 & 7.2 & 1.92 & 0.06 \\
-\textbf{DeltaF\_perc\_pos} & 0.38 & 0.1 & 3.77 & 0 \\
-\textbf{Forest\_SignIncrease} & -38.44 & 5.48 & -7.02 & 0 \\
-\textbf{Dryness} & 2.45 & 2.75 & 0.89 & 0.37 \\
-\textbf{log10(Area\_km2)} & -2.75 & 1.7 & -1.62 & 0.11 \\
-\bottomrule
-\end{longtable}
-
-The results of this model (Equation \eqref{eq:eq6}) indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.5\%.
-Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
-Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
+\textbf{(Intercept)} & 16 & 7.32 & 2.18 & 0.03 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 2.96 & 0 \\
+\textbf{Forest\_SignIncrease} & -37.69 & 5.49 & -6.86 & 0 \\
+\textbf{Dryness} & 2.39 & 2.73 & 0.87 & 0.38 \\
+\textbf{log10(Area\_km2)} & -3.03 & 1.7 & -1.78 & 0.08 \\
+\bottomrule
+\end{longtable}
+
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.08) that there is a reduction in the effect of forest cover change on streamflow related to Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3\%.
+Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
+Including this interaction in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.99)) as an interaction with Dryness.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.40}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\caption{\label{tab:out-modelArea-int} Results of the model including an interaction between Area and the dryness index}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{(Intercept)} & 12.72 & 7.52 & 1.69 & 0.09 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.86 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.76 & 5.57 & -6.42 & 0 \\
+\textbf{Dryness} & 5.28 & 3.16 & 1.67 & 0.1 \\
+\textbf{log10(Area\_km2)} & 0.03 & 2.39 & 0.01 & 0.99 \\
+\textbf{Dryness:log10(Area\_km2)} & -2.52 & 1.39 & -1.81 & 0.07 \\
+\bottomrule
+\end{longtable}
 
@@
 In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (\emph{Dryness, Area, From}) were considered non-linear in this model.
-This is also indicated a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
+As a result a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
 \emph{From} represents the starting date of the data collection.
@@
 \endhead
-\textbf{(Intercept)} & 18.46 & 5.47 & 3.38 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.24 & 0.1 & 2.43 & 0.02 \\
-\textbf{Forest\_SignIncrease} & -38.44 & 5.71 & -6.74 & 0 \\
+\textbf{(Intercept)} & 14.77 & 5.75 & 2.57 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.49 & 0 \\
+\textbf{Forest\_SignIncrease} & -38.53 & 5.61 & -6.87 & 0 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 3.6 & 9 & 0.88 & 0.05 \\
-\textbf{s(Dryness)} & 0 & 9 & 0 & 0.74 \\
-\textbf{s(length)} & 1.87 & 9 & 0.44 & 0.09 \\
-\textbf{s(From)} & 7.22 & 9 & 1.84 & 0.01 \\
-\bottomrule
-\end{longtable}
-
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.26 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, the smoothed variable for Area (km\textsuperscript{2}) is no longer a strong effect explaining variations in changes in stream flow.
-Clearly \emph{length} is also not explaining the variation.
-In contrast, \emph{From}, which indicate the start date of the study (and therefore the age of the study) is a significant effect.\\
+\textbf{s(log10(Area\_km2))} & 0.59 & 9 & 0.2 & 0.08 \\
+\textbf{s(Dryness)} & 4.19 & 9 & 0.86 & 0.09 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.55 \\
+\textbf{s(From)} & 5.35 & 9 & 1.05 & 0.09 \\
+\bottomrule
+\end{longtable}
+
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.27 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, the smoothed variable for Area (km\textsuperscript{2}) (p = 5.35)) and Dryness (p = 9))still provides reasonable explanation of changes in stream flow.
+However \emph{length} does not explain any of the variation.
+In contrast, \emph{From}, which indicates the start date of the study (and therefore the age of the study) has an effect (p = 0.09)).\\
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
@@
 
-A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).
+We now also include the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) i.e.~Equation \eqref{eq:eq2}.
 
@@
 \endhead
-\textbf{(Intercept)} & -23.61 & 18.27 & -1.29 & 0.2 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.16 & 0 \\
-\textbf{Forest\_SignIncrease} & -26.28 & 6.64 & -3.96 & 0 \\
-\textbf{Precip\_data\_typeOB} & -7.03 & 13.51 & -0.52 & 0.6 \\
-\textbf{Precip\_data\_typeSG} & 13.98 & 15.5 & 0.9 & 0.37 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.74 & 45.33 & 0.41 & 0.68 \\
-\textbf{Assessment\_techniqueHM} & 25.27 & 11.91 & 2.12 & 0.03 \\
-\textbf{Assessment\_techniquePWE} & 54.21 & 11.81 & 4.59 & 0 \\
+\textbf{(Intercept)} & -26.15 & 18.59 & -1.41 & 0.16 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 3.05 & 0 \\
+\textbf{Forest\_SignIncrease} & -23.79 & 7.04 & -3.38 & 0 \\
+\textbf{Precip\_data\_typeOB} & -6.89 & 14.24 & -0.48 & 0.63 \\
+\textbf{Precip\_data\_typeSG} & 19.59 & 15.77 & 1.24 & 0.22 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.33 & 45.34 & 0.4 & 0.69 \\
+\textbf{Assessment\_techniqueHM} & 31.53 & 12.7 & 2.48 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 53.37 & 11.93 & 4.47 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 31.99 & 46.11 & 0.69 & 0.49 \\
-\textbf{Assessment\_techniqueQPW} & 29.16 & 20.96 & 1.39 & 0.17 \\
+HM} & 46.51 & 46.35 & 1 & 0.32 \\
+\textbf{Assessment\_techniqueQPW} & 41.17 & 21.62 & 1.9 & 0.06 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 34.71 & 34.04 & 1.02 & 0.31 \\
-\textbf{Assessment\_techniqueSH} & 37.64 & 11.87 & 3.17 & 0 \\
-\textbf{Forest\_typeCF} & -11.48 & 8.07 & -1.42 & 0.16 \\
-\textbf{Forest\_typeMF} & -8.65 & 7.72 & -1.12 & 0.26 \\
-\textbf{Hydrological\_regimeSD} & 13.8 & 10.15 & 1.36 & 0.18 \\
+EA} & 32.24 & 27.41 & 1.18 & 0.24 \\
+\textbf{Assessment\_techniqueSH} & 42.47 & 12.58 & 3.38 & 0 \\
+\textbf{Forest\_typeCF} & -4.11 & 8.38 & -0.49 & 0.62 \\
+\textbf{Forest\_typeMF} & -10.52 & 8.14 & -1.29 & 0.2 \\
+\textbf{Hydrological\_regimeSD} & 7.68 & 10.4 & 0.74 & 0.46 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.32 & 9 & 0.05 & 0.25 \\
-\textbf{s(Dryness)} & 0.48 & 9 & 0.09 & 0.19 \\
-\textbf{s(length)} & 0.88 & 9 & 0.69 & 0.01 \\
-\textbf{s(From)} & 1.79 & 9 & 0.96 & 0.01 \\
-\bottomrule
-\end{longtable}
-
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a deviance explained of 0.31.
+\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.41 \\
+\textbf{s(Dryness)} & 3.86 & 9 & 1.64 & 0 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.31 \\
+\textbf{s(From)} & 8.8 & 9 & 3.53 & 0 \\
+\bottomrule
+\end{longtable}
+
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.34.
 This indicates that the categorical variables explain a limited amount of the variance.
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
-In this case, \emph{Area (km\^{}2)} is no longer a significant predictor.
+In this case, \emph{Area (km\^{}2)} is no longer a significant predictor, the reasons for this will be discussed later.
 
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between log10(Area km\textsuperscript{2}) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following Wood (2006)).
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the reason why using penalized smooths following Wood (2006) is useful).
 It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
-Similarly, the length variable is not significant.
-However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+Similarly, the \emph{length} variable is not significant.
+However, both \emph{From} and \emph{Dryness} variables show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+For example, in the case of \emph{From} the relationship is clearly strongly influenced by the two data points before 1920.
+Similarly, \emph{length} and \emph{Dryness} have points with high leverage.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years}\tabularnewline
+\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness \textgreater{} 4 and studies commencing after 1920}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{s(Dryness)} & 0.2 & 9 & 0.03 & 0.25 \\
-\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.33 \\
-\textbf{s(length)} & 3.81 & 9 & 1.19 & 0.02 \\
-\textbf{s(From)} & 3.94 & 9 & 1.31 & 0.01 \\
+\textbf{s(Dryness)} & 2.85 & 9 & 2.39 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.43 & 9 & 0.09 & 0.15 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.37 \\
+\textbf{s(From)} & 3.91 & 9 & 0.66 & 0.17 \\
 \bottomrule
@@
 (2005).
-One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \ref{fig:smoothsmodel6}).
-Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness \textgreater{} 4 (Table \ref{tab:restrictlength}) results in log10(Area) once again showing a significant effect with larger catchments having less impact of changes in forest cover.
-Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover.
-Finally, more recent studies also show greater impacts of changes in forest cover (Figure \ref{fig:smoothsmodel7}).
+One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \ref{fig:smoothsmodel6}).
+Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables.
+The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4, \emph{length} \textgreater{} 60 years, \emph{From} \textgreater{} 1920.
+This result in a reduction of the data set from 330 to 298 catchments.
+The model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.34.
+This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \ref{fig:smoothsmodel7}).
+However the other 3 variables are no longer significant due to the scatter in the data, but \emph{length} and \emph{From} show some strong local non-linear behaviour.
 
@@
 
-Essentially, the analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), even though, in contrast to Zhang et al.
+Essentially, the analysis shows that there is no clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), and, in contrast to Zhang et al.
 (2017), there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.33 with a slope of -9.56) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is 10\% lower.
-This is basically a result of the fact that large changes in forest cover are difficult to ``implement'' in an experiment.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.76) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is approximately 10\% lower.
+This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
+
 This is also reflected in the second caveat most of the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
-These techniques for larger catchments all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
+For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
 This is a confounding factor that is not easily addressed in the regression modelling attempted here.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model1} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheck7}
+\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheck7}
 \end{figure}
@@
 \endhead
-PWE & 208 \\
+PWE & 183 \\
 HM & 57 \\
@@
 \endhead
-\textbf{(Intercept)} & -32.33 & 18.4 & -1.76 & 0.08 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.22 & 0 \\
-\textbf{Forest\_SignIncrease} & -26.07 & 6.79 & -3.84 & 0 \\
-\textbf{Precip\_data\_typeOB} & 0.12 & 13.84 & 0.01 & 0.99 \\
-\textbf{Precip\_data\_typeSG} & 21.54 & 15.96 & 1.35 & 0.18 \\
-\textbf{Assessment\_techniqueHM} & 23.08 & 12.26 & 1.88 & 0.06 \\
-\textbf{Assessment\_techniquePWE} & 56.89 & 11.19 & 5.09 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 29.72 & 21.04 & 1.41 & 0.16 \\
-\textbf{Assessment\_techniqueSH} & 35.41 & 11.7 & 3.03 & 0 \\
-\textbf{Forest\_typeCF} & -9.41 & 8.06 & -1.17 & 0.24 \\
-\textbf{Forest\_typeMF} & -6.14 & 7.96 & -0.77 & 0.44 \\
-\textbf{Hydrological\_regimeSD} & 11.07 & 10.07 & 1.1 & 0.27 \\
+\textbf{(Intercept)} & -24.6 & 18.66 & -1.32 & 0.19 \\
+\textbf{DeltaF\_perc\_pos} & 0.27 & 0.1 & 2.76 & 0.01 \\
+\textbf{Forest\_SignIncrease} & -22.57 & 7.02 & -3.21 & 0 \\
+\textbf{Precip\_data\_typeOB} & -13.54 & 14.15 & -0.96 & 0.34 \\
+\textbf{Precip\_data\_typeSG} & 18.35 & 16.1 & 1.14 & 0.26 \\
+\textbf{Assessment\_techniqueHM} & 34.7 & 12.65 & 2.74 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 57.63 & 12.48 & 4.62 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 43.25 & 21.29 & 2.03 & 0.04 \\
+\textbf{Assessment\_techniqueSH} & 48.72 & 12.48 & 3.9 & 0 \\
+\textbf{Forest\_typeCF} & -6.27 & 8.31 & -0.75 & 0.45 \\
+\textbf{Forest\_typeMF} & -5.53 & 8.24 & -0.67 & 0.5 \\
+\textbf{Hydrological\_regimeSD} & 8.79 & 10.39 & 0.85 & 0.4 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 0 & 9 & 0 & 0.31 \\
-\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.35 \\
-\textbf{s(length)} & 4.58 & 9 & 1.59 & 0.01 \\
-\textbf{s(From)} & 0.87 & 9 & 0.64 & 0.01 \\
+\textbf{s(Dryness)} & 2.84 & 9 & 2.33 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.41 & 9 & 0.09 & 0.16 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.36 \\
+\textbf{s(From)} & 3.91 & 9 & 0.65 & 0.18 \\
 \bottomrule
@@
 Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \ref{tab:model8-linear} and \ref{tab:model8-smooth}).
-It strengthens the significance of the different assessment techniques but slightly reduces the importance of some of the smooth variables (\emph{log10(Area (km\textsuperscript{2}))} and \emph{From}).
+It strengthens the significance of the different assessment techniques and \emph{Dryness} but generally results in the same interpretation.
 Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
@@
 (2015)) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies are all on the drier end of the Dryness spectrum, highlighting another unresolved interaction in the data.
+However, as indicated earlier, the EA studies are all on the drier end of the \emph{Dryness} spectrum, highlighting another unresolved interaction in the data.
 
@@
 
+\hypertarget{interactions}{%
+\subsection{Interactions}\label{interactions}}
+
+Generally this study did not consider interactions, but the above discussion suggest that there are possible several interactions.
+The relationships between forest cover change and \emph{Area (km\textsuperscript{2})} and between \emph{Area (km\textsuperscript{2})} and assessment technique have already been highlighted.
+However there are further unexplored interactions between the study length and some of the variables and \emph{From} and the other variables.
+
+A principle component analysis of the numeric data reveals some of these interactions (Figure \ref{fig:pcaplot}), such as between \emph{length} and \emph{Dryness} and between \emph{From} and \emph{Area (km\textsuperscript{2})}.
+Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{PCA_biplot_Alldata} \caption{Biplot of the first two principle components using a principle component analysis on the numerical values of the }\label{fig:pcaplot}
+\end{figure}
+
 \hypertarget{further-considerations}{%
@@
 In contrast to Filoso et al.
-(2017), we did not identify the length of the observation period as a significant variable, however, we did find that the start date of the observations was significant.
-What is puzzling is that the effect of the start date (\emph{From}) had a positive slope indicating that more recent studies measured a greater effect on stream flow of forest cover change.
-At this point, it is unclear what this trend signifies.
-It is also possible that there is an interaction between the start date of the observations and the length of the study, as later start dates would never have been able to present long observed data series.
-
+(2017), we did not identify the length of the observation period as a significant variable in our final model.
 There are further confounding factors in the data, which were also classified by Filoso et al.
@@
 However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow (Vervoort et al., 2021).
+This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.\\
+By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
 
@@
 More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
+It demonstrates that analysis of large databases of essentially ``aggregated data'' should be considered carefully.
 
 Results of this statistical analysis need to be considered ``conditional on the data''
-Conditional on the data, the impact of forestry on streamflow is:\\
-- Greater for forest clearing then for reforestation\\
-- Reduced for larger watersheds\\
-- Increases for drier watersheds\\
-- Sensitive to the assessment method used in the historical data
+Conditional on the data, the impact of forestry on streamflow is:
+
+\begin{itemize}
+\item
+  Greater for forest clearing then for reforestation
+\item
+  Possibly reduced for larger watersheds, but this is only in interaction with other factors such as climate
+\item
+  Increases for drier watersheds
+\item
+  Sensitive to the assessment method used in the historical data
+\end{itemize}
+
+Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area.
+Furthermore, the current study analyses a large global dataset.
+This analysis does not exlude more local and regional effects that cannot be identified in the global data.
 

### 2022-02-03 paper/Forest_and_Water/Forest_and_Water.tex

@@
     \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
-    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
     \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
@@
 
-There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017).
+There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017).
 The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
@@
 (2011) indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
-However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchmensts after deforestation.
+However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchments after deforestation.
 
@@
 
-In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location.
-Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018).
+In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
+Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
 For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 (2017) and Filoso et al.
-(2017) and the additional studies, 29 potential duplicates were identified and flagged in the data.
+(2017), 29 potential duplicates were identified and flagged in the data.
 
@@
 \end{figure}
-
-This shows that for the data related to forest decreases, there is almost always a positive flow change.
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
 
@@
 The first three rows relate to the model using the original data base from Zhang et al.~(2017).
-The bottom three rows are the results of the model including the new data}\tabularnewline
+The bottom three rows are the results of the model including the new data.}\tabularnewline
 \toprule
@@
 \textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
-\textbf{Forest\_SignIncrease} & -29.17 & 5.79 & -5.04 & 0 \\
+\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
 \textbf{(Intercept)} & 10.76 & 5.34 & 2.02 & 0.04 \\
 \textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.04 & 0 \\
-\textbf{Forest\_SignIncrease} & -38.03 & 5.26 & -7.23 & 0 \\
+\textbf{Forest\_Signincrease} & -38.03 & 5.26 & -7.23 & 0 \\
 \bottomrule
@@
 \textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.13 & 0 \\
-\textbf{Forest\_SignIncrease} & -38.99 & 5.54 & -7.04 & 0 \\
+\textbf{Forest\_Signincrease} & -38.99 & 5.54 & -7.04 & 0 \\
 \textbf{Latitude} & -0.01 & 0.09 & -0.13 & 0.9 \\
@@
 \caption{\label{tab:out-model4} Results of the model including the dryness index
+\textless\textless\textless\textless\textless\textless\textless{} HEAD
 }\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 9.53 & 6.38 & 1.49 & 0.14 \\
-\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 4.86 & 0 \\
-\textbf{Forest\_SignIncrease} & -39.4 & 5.42 & -7.26 & 0 \\
-\textbf{Dryness} & 1.59 & 2.7 & 0.59 & 0.56 \\
-\bottomrule
-\end{longtable}
-
-The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.56).
+\textbf{(Intercept)} & 8.83 & 6.3 & 1.4 & 0.16 \\
+\textbf{DeltaF\_perc\_pos} & 0.4 & 0.08 & 4.88 & 0 \\
+\textbf{Forest\_Signincrease} & -38.26 & 5.32 & -7.19 & 0 \\
+\textbf{Dryness} & 1.7 & 2.67 & 0.64 & 0.52 \\
+\bottomrule
+\end{longtable}
+
+\hypertarget{the-results-from-this-model-equation-refeqeq5-and-table-reftabout-model4-interestingly-indicate-no-impact-of-dryness-modulating-the-change-in-streamflow-as-a-function-of-the-change-in-forest-cover-change-which-is-surprising-in-light-of-earlier-reported-results-zhang2017-filoso2017.-in-this-case-the-evidence-is-highly-doubtful-p-0.52.-however-it-is-very-well-possible-that-there-is-a-further-interaction-in-the-data-with-other-variables-or-unknown-variables-that-this-simpler-version-of-the-model-cannot-identify.-this-is-partly-evidenced-by-the-fact-that-the-overall-variance-explained-is-still-low-with-an-adjusted-r2-of-0.22.}{%
+\section{\texorpdfstring{The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+In this case, the evidence is highly doubtful (p = 0.52).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.22.}{The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+In this case, the evidence is highly doubtful (p = 0.52).
+However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted r2 of 0.22.}}\label{the-results-from-this-model-equation-refeqeq5-and-table-reftabout-model4-interestingly-indicate-no-impact-of-dryness-modulating-the-change-in-streamflow-as-a-function-of-the-change-in-forest-cover-change-which-is-surprising-in-light-of-earlier-reported-results-zhang2017-filoso2017.-in-this-case-the-evidence-is-highly-doubtful-p-0.52.-however-it-is-very-well-possible-that-there-is-a-further-interaction-in-the-data-with-other-variables-or-unknown-variables-that-this-simpler-version-of-the-model-cannot-identify.-this-is-partly-evidenced-by-the-fact-that-the-overall-variance-explained-is-still-low-with-an-adjusted-r2-of-0.22.}}
+
+The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
+In this case the evidence is not very strong (p = 0.52.
+However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is \textasciitilde6\% greater.
+Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
+\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater{} 70692a2b708e51dd658780f5487e0fb00aaf13a7
 
@@
 
+\textless\textless\textless\textless\textless\textless\textless{} HEAD
 There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \ref{tab:drytable}.
+=======
+There possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
+These catchments are listed in Table \ref{tab:drytable}.
+\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater{} 70692a2b708e51dd658780f5487e0fb00aaf13a7
 
@@
 Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.
-The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.
 
@@
 \endhead
-\textbf{(Intercept)} & 16 & 7.32 & 2.18 & 0.03 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 2.96 & 0 \\
-\textbf{Forest\_SignIncrease} & -37.69 & 5.49 & -6.86 & 0 \\
-\textbf{Dryness} & 2.39 & 2.73 & 0.87 & 0.38 \\
-\textbf{log10(Area\_km2)} & -3.03 & 1.7 & -1.78 & 0.08 \\
-\bottomrule
-\end{longtable}
-
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.08) that there is a reduction in the effect of forest cover change on streamflow related to Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+\textbf{(Intercept)} & 15.37 & 7.24 & 2.12 & 0.03 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 2.94 & 0 \\
+\textbf{Forest\_Signincrease} & -36.61 & 5.38 & -6.8 & 0 \\
+\textbf{Dryness} & 2.48 & 2.69 & 0.92 & 0.36 \\
+\textbf{log10(Area\_km2)} & -3.03 & 1.67 & -1.81 & 0.07 \\
+\bottomrule
+\end{longtable}
+
+The results of this model (Equation \eqref{eq:eq6}) clearly indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.5\%.
+Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
+Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
+
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.07) that there is a reduction in the effect of forest cover change on streamflow related to Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
 In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including this interaction in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.99)) as an interaction with Dryness.
+Including this interaction in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.95)) as an interaction with Dryness.
 
@@
 \endhead
-\textbf{(Intercept)} & 12.72 & 7.52 & 1.69 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.86 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.76 & 5.57 & -6.42 & 0 \\
-\textbf{Dryness} & 5.28 & 3.16 & 1.67 & 0.1 \\
-\textbf{log10(Area\_km2)} & 0.03 & 2.39 & 0.01 & 0.99 \\
-\textbf{Dryness:log10(Area\_km2)} & -2.52 & 1.39 & -1.81 & 0.07 \\
+\textbf{(Intercept)} & 11.95 & 7.43 & 1.61 & 0.11 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.85 & 0 \\
+\textbf{Forest\_Signincrease} & -34.62 & 5.46 & -6.34 & 0 \\
+\textbf{Dryness} & 5.46 & 3.1 & 1.76 & 0.08 \\
+\textbf{log10(Area\_km2)} & 0.15 & 2.35 & 0.06 & 0.95 \\
+\textbf{Dryness:log10(Area\_km2)} & -2.62 & 1.37 & -1.91 & 0.06 \\
 \bottomrule
@@
 \endhead
-\textbf{(Intercept)} & 14.77 & 5.75 & 2.57 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.49 & 0 \\
-\textbf{Forest\_SignIncrease} & -38.53 & 5.61 & -6.87 & 0 \\
+\textbf{(Intercept)} & 13.99 & 5.68 & 2.46 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.51 & 0 \\
+\textbf{Forest\_Signincrease} & -37.18 & 5.48 & -6.78 & 0 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.59 & 9 & 0.2 & 0.08 \\
-\textbf{s(Dryness)} & 4.19 & 9 & 0.86 & 0.09 \\
+\textbf{s(log10(Area\_km2))} & 0.68 & 9 & 0.28 & 0.05 \\
+\textbf{s(Dryness)} & 4.28 & 9 & 0.94 & 0.07 \\
 \textbf{s(length)} & 0 & 9 & 0 & 0.55 \\
-\textbf{s(From)} & 5.35 & 9 & 1.05 & 0.09 \\
-\bottomrule
-\end{longtable}
-
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.27 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, the smoothed variable for Area (km\textsuperscript{2}) (p = 5.35)) and Dryness (p = 9))still provides reasonable explanation of changes in stream flow.
+\textbf{s(From)} & 4.62 & 9 & 0.9 & 0.1 \\
+\bottomrule
+\end{longtable}
+
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.26 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, the smoothed variable for Area (km\textsuperscript{2}) (p = 4.62)) and Dryness (p = 9))still provides reasonable explanation of changes in stream flow.
 However \emph{length} does not explain any of the variation.
-In contrast, \emph{From}, which indicates the start date of the study (and therefore the age of the study) has an effect (p = 0.09)).\\
+In contrast, \emph{From}, which indicates the start date of the study (and therefore the age of the study) has an effect (p = 0.1)).\\
 However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
@@
 \endhead
-\textbf{(Intercept)} & -26.15 & 18.59 & -1.41 & 0.16 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 3.05 & 0 \\
-\textbf{Forest\_SignIncrease} & -23.79 & 7.04 & -3.38 & 0 \\
-\textbf{Precip\_data\_typeOB} & -6.89 & 14.24 & -0.48 & 0.63 \\
-\textbf{Precip\_data\_typeSG} & 19.59 & 15.77 & 1.24 & 0.22 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.33 & 45.34 & 0.4 & 0.69 \\
-\textbf{Assessment\_techniqueHM} & 31.53 & 12.7 & 2.48 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 53.37 & 11.93 & 4.47 & 0 \\
+\textbf{(Intercept)} & -29.23 & 18.18 & -1.61 & 0.11 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.29 & 0 \\
+\textbf{Forest\_Signincrease} & -23.48 & 6.84 & -3.43 & 0 \\
+\textbf{Precip\_data\_typeOB} & -5.03 & 13.94 & -0.36 & 0.72 \\
+\textbf{Precip\_data\_typeSG} & 19.41 & 15.38 & 1.26 & 0.21 \\
+\textbf{Assessment\_techniqueEA, HM} & 19.56 & 45.02 & 0.43 & 0.66 \\
+\textbf{Assessment\_techniqueHM} & 33.57 & 12.36 & 2.72 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 52.79 & 11.72 & 4.5 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 46.51 & 46.35 & 1 & 0.32 \\
-\textbf{Assessment\_techniqueQPW} & 41.17 & 21.62 & 1.9 & 0.06 \\
+HM} & 43.58 & 45.93 & 0.95 & 0.34 \\
+\textbf{Assessment\_techniqueQPW} & 39.98 & 21.36 & 1.87 & 0.06 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 32.24 & 27.41 & 1.18 & 0.24 \\
-\textbf{Assessment\_techniqueSH} & 42.47 & 12.58 & 3.38 & 0 \\
-\textbf{Forest\_typeCF} & -4.11 & 8.38 & -0.49 & 0.62 \\
-\textbf{Forest\_typeMF} & -10.52 & 8.14 & -1.29 & 0.2 \\
-\textbf{Hydrological\_regimeSD} & 7.68 & 10.4 & 0.74 & 0.46 \\
+EA} & 31.36 & 27.12 & 1.16 & 0.25 \\
+\textbf{Assessment\_techniqueSH} & 42.63 & 12.41 & 3.44 & 0 \\
+\textbf{Forest\_typeCF} & -2.33 & 8.23 & -0.28 & 0.78 \\
+\textbf{Forest\_typeMF} & -10.36 & 8 & -1.3 & 0.2 \\
+\textbf{Hydrological\_regimeSD} & 6.99 & 10.25 & 0.68 & 0.5 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.41 \\
-\textbf{s(Dryness)} & 3.86 & 9 & 1.64 & 0 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.31 \\
-\textbf{s(From)} & 8.8 & 9 & 3.53 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.37 \\
+\textbf{s(Dryness)} & 3.96 & 9 & 1.68 & 0 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.36 \\
+\textbf{s(From)} & 8.8 & 9 & 3.64 & 0 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 2.85 & 9 & 2.39 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.43 & 9 & 0.09 & 0.15 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.37 \\
-\textbf{s(From)} & 3.91 & 9 & 0.66 & 0.17 \\
+\textbf{s(Dryness)} & 3.21 & 9 & 3.22 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.46 & 9 & 0.11 & 0.13 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.44 \\
+\textbf{s(From)} & 4.06 & 9 & 1.12 & 0.05 \\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model with reduced data for dryness and length}\label{fig:smoothsmodel7}
+\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model wiht reduced data for dryness and length}\label{fig:smoothsmodel7}
 \end{figure}
@@
 The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4, \emph{length} \textgreater{} 60 years, \emph{From} \textgreater{} 1920.
-This result in a reduction of the data set from 330 to 298 catchments.
+This result in a reduction of the data set from 330 to 304 catchments.
 The model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.34.
@@
 \endhead
-\textbf{(Intercept)} & -24.6 & 18.66 & -1.32 & 0.19 \\
-\textbf{DeltaF\_perc\_pos} & 0.27 & 0.1 & 2.76 & 0.01 \\
-\textbf{Forest\_SignIncrease} & -22.57 & 7.02 & -3.21 & 0 \\
-\textbf{Precip\_data\_typeOB} & -13.54 & 14.15 & -0.96 & 0.34 \\
-\textbf{Precip\_data\_typeSG} & 18.35 & 16.1 & 1.14 & 0.26 \\
-\textbf{Assessment\_techniqueHM} & 34.7 & 12.65 & 2.74 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 57.63 & 12.48 & 4.62 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 43.25 & 21.29 & 2.03 & 0.04 \\
-\textbf{Assessment\_techniqueSH} & 48.72 & 12.48 & 3.9 & 0 \\
-\textbf{Forest\_typeCF} & -6.27 & 8.31 & -0.75 & 0.45 \\
-\textbf{Forest\_typeMF} & -5.53 & 8.24 & -0.67 & 0.5 \\
-\textbf{Hydrological\_regimeSD} & 8.79 & 10.39 & 0.85 & 0.4 \\
+\textbf{(Intercept)} & -27.03 & 18.34 & -1.47 & 0.14 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.96 & 0 \\
+\textbf{Forest\_Signincrease} & -22.23 & 6.83 & -3.25 & 0 \\
+\textbf{Precip\_data\_typeOB} & -11.91 & 13.85 & -0.86 & 0.39 \\
+\textbf{Precip\_data\_typeSG} & 18.07 & 15.65 & 1.15 & 0.25 \\
+\textbf{Assessment\_techniqueHM} & 36.48 & 12.3 & 2.97 & 0 \\
+\textbf{Assessment\_techniquePWE} & 56.52 & 12.33 & 4.58 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 42.07 & 21.03 & 2 & 0.05 \\
+\textbf{Assessment\_techniqueSH} & 48.82 & 12.33 & 3.96 & 0 \\
+\textbf{Forest\_typeCF} & -4.64 & 8.18 & -0.57 & 0.57 \\
+\textbf{Forest\_typeMF} & -5.22 & 8.08 & -0.65 & 0.52 \\
+\textbf{Hydrological\_regimeSD} & 8.29 & 10.25 & 0.81 & 0.42 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 2.84 & 9 & 2.33 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.41 & 9 & 0.09 & 0.16 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.36 \\
-\textbf{s(From)} & 3.91 & 9 & 0.65 & 0.18 \\
+\textbf{s(Dryness)} & 2.93 & 9 & 2.41 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.45 & 9 & 0.1 & 0.14 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.4 \\
+\textbf{s(From)} & 3.84 & 9 & 0.66 & 0.17 \\
 \bottomrule
@@
 Furthermore, the current study analyses a large global dataset.
-This analysis does not exlude more local and regional effects that cannot be identified in the global data.
+This analysis does not exclude more local and regional effects that cannot be identified in the global data.
 

### 2022-02-04 paper/Forest_and_Water/Forest_and_Water.Rmd
fixing page breaks and merge issues

@@
   - code: INIA
-    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
+    address: Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
   - code: IMFIA
@@
 
-There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017].
+There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017].
 The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
 The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
-However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchmensts after deforestation.
+However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
 
@@
 
-In addition, additional variables added were the latitude and longitude for the centre of the catchment as an approximation of its spatial location.
-Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018].
+In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
+Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
 For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
@@
 Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
-Finally, on review of all the data in @zhang2017 and @filoso2017 and the additional studies, 29 potential duplicates were identified and flagged in the data.
+Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.
 
@@
 ## Statistical modelling
-
+```{r, include=FALSE}
+options(tinytex.verbose = TRUE)
+```
 <!----
@@
 <!-- read in the ET data -->
-<!-- no longer needed as this is now included in the data
+
 ```{r, echo = F, warning = F, message=F}
@@
 ```
---->
 
@@
 
-```{r, eval = F, echo = F}
-# checking missing values
-summary(All_data)
-
-```
-
 <!--now combine with E0 values -->
-<!-- No longer needed as now included in the data -->
 ```{r, echo = F}
-# All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
-# All_data <- All_data %>%
-#   mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
-#   select(-c(E0.x,E0.y)) #%>%
-#   #rename(Latitude = Latitude.x,
-# #         Longitude = Longitude.x)
+All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
+All_data <- All_data %>%
+  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
+  select(-c(E0.x,E0.y)) #%>%
+  #rename(Latitude = Latitude.x,
+#         Longitude = Longitude.x)
 
@@
 Therefore all the change in forest cover data is converted to positive values, and an additional column ($sign_{forest cover}$) is added that indicates whether it was a forest cover increase or decrease.
+
 A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Qf \%$.
@@
 In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
-The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2*.
 The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^.
+Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
 
@@
       "DeltaQf_perc" = "Delta Qf (%)",
-      "E0" = "Annual E0 (mm)",
+      "E0" = "Annual Potential E (mm)",
       "Pa_mm" = "Annual Rainfall (mm)"))) +   
@@
   mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
-                              "Decrease", "Increase"),
+                              "decrease", "increase"),
          DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
@@
 ```
+
+In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2).
+In other words, flow almost always increased.
+However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
+However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
 
@@
 All_data2 %>%
-  filter(is.na(DeltaF_perc_pos) == F) %>%
   ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
@@
   scale_colour_manual(name = "Forest change direction", 
-                      values = c("Decrease" = "red", "Increase" = "blue")) +
+                      values = c("Decrease" = "red","Increase" = "blue")) +
   xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
@@
 \end{align}
+
+<!--\newpage-->
 
@@
 ```
+
 <!-- running PCA, this is used later -->
@@
 Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
-The effect of Dryness remains in the model, given that there could be interactions that are worth exploring.
 
@@
 
-```{r modelArea_int, echo = F, eval =T}
+```{r modelArea_b, echo = F, eval =T}
+#Zhang_all2 <- Zhang_all2 %>%
+#  mutate(length = To - From)
+
 model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
-In this case, *Area (km^2)* is no longer a significant predictor, the reasons for this will be discussed later.
+In this case, *log10(Area (km^2^))* is no longer a significant predictor, the reasons for this will be discussed later.
 
@@
 ```{r model7_nolongLength, echo = F}
+
 NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1920)
+
 model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
 ```
+
 ```{r, echo = F}
@@
 ```
+
 ```{r visualisesmooths7,echo =F, eval = F}
@@
 ```
-```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model with reduced data for dryness and length", out.width="90%", echo=F}
+
+```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}
 
@@
 ```
+
+<!--\newpage-->
 
@@
 This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
+
 The model has more explaining power with an adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
@@
 Furthermore, the current study analyses a large global dataset.
-This analysis does not exlude more local and regional effects that cannot be identified in the global data.
-
-References {#references .unnumbered}
-==========
+This analysis does not exclude more local and regional effects that cannot be identified in the global data.
 
@@
 ```
+
+References {#references .unnumbered}
+==========

### 2022-02-04 paper/Forest_and_Water/Forest_and_Water.tex
fixing page breaks and merge issues

@@
 Therefore all the change in forest cover data is converted to positive values, and an additional column (\(sign_{forest cover}\)) is added that indicates whether it was a forest cover increase or decrease.
+
 A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have a linear or non-linear relationship with \(\Delta Qf \%\).
@@
 (2017)).
-The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\^{}2}.
 The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:out-model4} Results of the model including the dryness index
-\textless\textless\textless\textless\textless\textless\textless{} HEAD
-}\tabularnewline
+\caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
 \toprule
@@
 
-\hypertarget{the-results-from-this-model-equation-refeqeq5-and-table-reftabout-model4-interestingly-indicate-no-impact-of-dryness-modulating-the-change-in-streamflow-as-a-function-of-the-change-in-forest-cover-change-which-is-surprising-in-light-of-earlier-reported-results-zhang2017-filoso2017.-in-this-case-the-evidence-is-highly-doubtful-p-0.52.-however-it-is-very-well-possible-that-there-is-a-further-interaction-in-the-data-with-other-variables-or-unknown-variables-that-this-simpler-version-of-the-model-cannot-identify.-this-is-partly-evidenced-by-the-fact-that-the-overall-variance-explained-is-still-low-with-an-adjusted-r2-of-0.22.}{%
-\section{\texorpdfstring{The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
 In this case, the evidence is highly doubtful (p = 0.52).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.22.}{The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.52).
-However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted r2 of 0.22.}}\label{the-results-from-this-model-equation-refeqeq5-and-table-reftabout-model4-interestingly-indicate-no-impact-of-dryness-modulating-the-change-in-streamflow-as-a-function-of-the-change-in-forest-cover-change-which-is-surprising-in-light-of-earlier-reported-results-zhang2017-filoso2017.-in-this-case-the-evidence-is-highly-doubtful-p-0.52.-however-it-is-very-well-possible-that-there-is-a-further-interaction-in-the-data-with-other-variables-or-unknown-variables-that-this-simpler-version-of-the-model-cannot-identify.-this-is-partly-evidenced-by-the-fact-that-the-overall-variance-explained-is-still-low-with-an-adjusted-r2-of-0.22.}}
-
-The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) confirm that dryness is a clear confounding factor related to the change in streamflow as a function of the change in forest cover change.
-In this case the evidence is not very strong (p = 0.52.
-However, if the dryness index doubles (remembering that Dryness = 1 when E0 = Pa, so in this case E0 = 2×Pa, which is very dry), the change in runoff is \textasciitilde6\% greater.
-Again, overall variance explained is not very much, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
-\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater{} 70692a2b708e51dd658780f5487e0fb00aaf13a7
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
 
@@
 
-\textless\textless\textless\textless\textless\textless\textless{} HEAD
 There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
 These catchments are listed in Table \ref{tab:drytable}.
-=======
-There possible issues with the data, a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
-These catchments are listed in Table \ref{tab:drytable}.
-\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater\textgreater{} 70692a2b708e51dd658780f5487e0fb00aaf13a7
 
@@
 \end{longtable}
-
-The results of this model (Equation \eqref{eq:eq6}) clearly indicate a reduction in the effect of forest cover change with Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.5\%.
-Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes more significant, possibly suggesting an interaction between Dryness and Area.
-Including this interaction suggest that the interaction term (log10(Area) by Dryness) would be significant, but this replaces the effect of Area (results not shown).
 
@@
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
-In this case, \emph{Area (km\^{}2)} is no longer a significant predictor, the reasons for this will be discussed later.
+In this case, \emph{log10(Area (km\textsuperscript{2}))} is no longer a significant predictor, the reasons for this will be discussed later.
 
@@
 This result in a reduction of the data set from 330 to 304 catchments.
+
 The model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.34.

### 2022-02-04 paper/Forest_and_Water/Forest_and_Water.Rmd
Update Forest_and_Water.Rmd

@@
   - code: INIA
-    address: Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
+    address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
   - code: IMFIA
@@
 However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
+Otherwise, in both cases the variability in the reported change in flow increases with the increase in forest cover change.
 
@@
 
-It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
+However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
 The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.
@@
 
-Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
+Climate, and in particular evapotranspiration, can have a significant effect on the streamflow change as represented by the dryness index.
+This is also highlighted by both @zhang2017 and @jackson2005.
 Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).

### 2022-03-08 paper/Forest_and_Water/Forest_and_Water.Rmd
addressing WBs comments

--- 2022-02-04 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
     email: willem.vervoort@sydney.edu.au
-    affiliation: DARE, The University of Sydney
+    affiliation: DARE
     footnote: 1
@@
   - code: DARE
-    address: ARC Training Centre Data Analytics for Resources and the Environment
+    address: ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.
   - code: The University of Sydney
-    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
+    address: The University of Sydney, Sydney, NSW 2006, Australia
   - code: INIA
@@
 abstract: |
-  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
-Deforestation causes a 32\% greater change in flow compared to reforestation.
-Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies.
-Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2.
-Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
   Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
@@
 Generalised additive models were used to run flexible regressions including multiple variables.
+  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
+Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover.
+Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data.
+Smaller catchment dominate the database with 42\% of the data < 1 km^2^ and 65\% of the data < 10 km^2^.
+Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
+Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
 journal: "Journal of Hydrology"
@@
 Many of the US studies use *watershed*, while European and Australian studies use *catchment*.
-In particular, we retained the term "paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.
+In particular, we retained the term "paired watershed studies" and "quasi-paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.
 
@@
 
-The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
+The second study [@filoso2017] is a systematic review of reforestation studies (only studies in which forest cover increased).
+This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of quantitative data (37) they showed a log-linear relationship between decreasing catchment size and an increasing decline in streamflow.
+For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with @zhang2017), but they did not identify a distinct threshold and fitted a log-linear relationship.
 In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
@@
 A final summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
-However, this paper aims to explain the variation in the data using the Fuh model, and in particular aims to link the variation in the observed data to variations in the exponent *m* in the model.
-A key observation is that in drier environments, the effects of deforestation are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
-
-Encouraged by the work presented by @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
-In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.
+However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
+In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model.
+A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
+
+Encouraged by the work from @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
+In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.
 
@@
 While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
-In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
-We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) in our analysis.
+In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET.
+In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).
+In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
+We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.
 
@@
 To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from @filoso2017.
-In particular, we focussed on the 37 data points included in the quantitative analysis in @filoso2017.
-
-In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
-Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
-For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+In particular, we focussed on including the 37 data points related to the quantitative analysis in @filoso2017.
+
+In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location.
+These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps^TM^.
+In the dataset, an additional column has been added to indicate the source of the location data.
+
+Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
+Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
+For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
 
@@
 @jackson2005; @filoso2017], as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a "Kuczera curve" [@kuczera1987].
+It is not clear if this is an effect of increased water use in growth [@vertessy2001] or due to changes in interception [@stoof2012].
 Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
@@
 Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
-In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
+In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
 Table 6 in @zhang2011).
@@
 Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
-Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.
-
-The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
+Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
+
+The final column in the improved data set is a "notes" column, which we added, but is not further used in the analysis.
+It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
+This will allow future research to further sructinise our input for errors.
 
@@
 ```{r, include=FALSE}
-options(tinytex.verbose = TRUE)
+options(tinytex.verbose = FALSE)
 ```
@@
 --->
-```{r, echo = F, message = F}
+```{r, echo = F, message = F, warning=FALSE}
 require(tidyverse)
@@
 
-```{r, echo = F, warning = F}
+```{r, echo = F, warning = F, message = F}
 Zhang_small$From <- as.numeric(Zhang_small$From)
@@
 Here $X_i$ are factorial variables, while $Z_i$ are continuous variables.
-The model assumes no direct interactions and all variables are additive.
+The model intially assumes no direct interactions and all variables are additive.
+We will comment in this assumption in the discussion.
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation.
+In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data is converted to positive values, and an additional column ($sign_{forest cover}$) is added that indicates whether it was a forest cover increase or decrease.
-
-A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Qf \%$.
-This means that a smooth function $s()$ is applied to the $Z_i$ variables.
-For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant [@wood2006].
-
-For the model in equation 2, we initially only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results.
+Therefore all the change in forest cover data was converted to positive values, and an additional factorial column ($sign_{forest cover}$) is included indicating whether it was a forest cover increase or decrease.
+
+A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with $\Delta Qf \%$.
+This means that a smooth function $s()$ can be applied to the $Z_i$ variables.
+For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty.
+The result of this approach is that for high enough smoothing parameters (i.e.
+if the data is very "wiggly") the smooth term can be shrunk to 0 and thus will be no longer significant [@wood2006].
+This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship.
+In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.
+
+The over arching test focuses on  the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
+Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.
+
+As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results).
 Subsequently the analysis was repeated and the additionally identified catchments were added.
-
-More generally the results were analysed to identify:  
-1. the significance of the different variables  
-2. the direction of the categorical or shape of the smooth variables
 
@@
 In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
-The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2*.
+The current number of catchments is the result of the removal of duplicates and our modifications and additions.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*.
 The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
+Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
 
 ```{r, echo = F, warning = F, message=F, eval =F}
+require(latex2exp)
 png("DataExploration.png", width = 2500, height = 1800, res = 300)
 All_data %>%
-  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
+  mutate(log10Area = log10(Area_km2)) %>%
+  pivot_longer(c(DeltaF_perc, DeltaQf_perc, log10Area, Pa_mm, E0), 
                names_to = "variable", values_to = "value") %>%
@@
   facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
-    c("Area_km2" = "Area (km2)",
+    c("log10Area" =  "log10(Area km2)", #latex2exp::TeX("$\\log_{10}({Area~km^2})$"),
       "DeltaF_perc" = "Delta Forest (%)",
@@
       "E0" = "Annual Potential E (mm)",
-      "Pa_mm" = "Annual Rainfall (mm)"))) +   
+      "Pa_mm" = "Annual Rainfall (mm)"),default = label_parsed)) +   
   theme(axis.title = element_text(size = rel(1.5)),
@@
 (ref:datagraphscaption) Overview of the distribution of the data set for five of the included variables.
+Note that the first panel (showing the distribution of the catchment areas) indicates the distribution of the *log_10_* transformed Area (in km^2^).
 
@@
 
-In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2).
+```{r, eval = F, echo = F}
+All_data2 %>%
+  filter(is.na(Forest_Sign)==T)
+```
+
+Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
 In other words, flow almost always increased.
@@
 All_data2 %>%
+  filter(is.na(Forest_Sign)==F) %>%
   ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
@@
         strip.text = element_text(size = rel(1.5))) + 
-  scale_colour_manual(name = "Forest change direction", 
+  scale_colour_manual(name = "Forest cover direction", 
                       values = c("Decrease" = "red","Increase" = "blue")) +
@@
 
-```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of increases and decreases in forest cover", out.width="90%"}
+```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of decreases (top) and increases (bottom) in forest cover", out.width="90%"}
 include_graphics("Increase_decrease.png")
+```
+
+```{r, eval = F, echo = F}
+linear_model_E0 <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
+                             Forest_Sign + E0, 
+                           data = All_data2)
+#summary(linear_model_E0)
+#gam.check(linear_model_simple)
+
 ```
@@
 (2017).
-The bottom three rows are the results of the model including the new data.")
+The bottom three rows are the results of the model including the new data.
+Clearly there is no major change arising from the additional data.")
 ```
@@
 Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
+Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total @zhang2017 data set.
+But this also means that our re-analysis of the data can be directly compared to the original study.
 
@@
 
-As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps.
-In the dataset, an additional column is added to indicate the source of the location data.
+Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients.
 As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
@@
 ```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model including Latitude and Longitude including new data")
-```
-
-There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
+pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model based on the complete dataset and including Latitude and Longitude")
+```
+
+There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence.
 The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.
@@
 
-Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
-Initially, we tested models using annual average precipitation (*Pa (mm)*), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable.
+While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
+Potential evapotranspiration (*E0*) by itself was  not significant in the.
+Initially, we also tested models using only the annual average precipitation (*Pa (mm)*), but  interactions between precipitation and evapotranspiraton might be captured by the dryness index.
+Both dryness index  and *Pa (mm)* were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
+As a result only the dryness index was retained as a climate indicator to align with the earlier work by @zhang2017.
 Given that Latitude and Longitude were not significant, we dropped these from the model.
@@
 
-The results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results [@zhang2017; @filoso2017].
+Similar to *E0* or *Pa_mm*, the results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
+This might seem suprising in light of earlier reported results [@zhang2017; @filoso2017].
 In this case, the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
@@
 This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
+As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
 
 ```{r, echo = F}
-pander(All_data2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
+pander(All_data2 %>% filter(Dryness > 4) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
+         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
 ```
@@
 
-The major hypothesis to test is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017.
+The second major variable is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017.
 Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
@@
 
-The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to Area (km^2^) (Table \@ref(tab:out-modelArea)).
+The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km^2^)) (Table \@ref(tab:out-modelArea)).
 In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by `r abs(round(summary(model_Area)$p.table[5,1],1))`\%.
 Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including this interaction in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km^2^)) in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.
 
@@
 
+```{r, eval = F, echo = F}
+All_data2 %>%
+  ggplot(aes(log10(Area_km2),Dryness, size = DeltaF_perc_pos)) + geom_point(colour = "blue", alpha =0.5)
+summary(lm(Dryness~log10(Area_km2),data = All_data2))
+```
+
 ## Are some of the variables possibly non-linear?
@@
 This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005).
-In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (*Dryness, Area, From*) were considered non-linear in this model.
+In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (*Dryness* and *Area*) were considered non-linear.
 As a result a shrinkage smoothing spline [@wood2006] was applied to these variables.
-*From* represents the starting date of the data collection.
+
+<!-- *From* represents the starting date of the data collection.
+The idea is that this last variable gives some indication of the "age" of the experiment.
+Improvements in data collection could introduce a positive or negative bias between more recent and older studies -->
 
 \begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon (\#eq:eq7)
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon (\#eq:eq7)
 \end{align}
@@
 All_data2 <- All_data2 %>%
-  mutate(length = To - From)
+  mutate(length = To - From,
+         mid_year = From + (To - From)/2)
 
 model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
-                    s(length, bs="ts") + s(From, bs = "ts") ,
+                    s(log10(Area_km2), k = 6, bs="ts") + 
+                    s(Dryness, k = 10, bs="ts") + 
+                    s(length, k = 50, bs="ts"), #+ 
+                    #s(From, k = 30, bs = "ts") ,
                   data = All_data2)
-#summary(model5_all)
-#gam.check(model5_all)
-#plot(model5_all)
+# summary(model5_all)
+# gam.check(model5_all)
+# plot(model5_all)
 ```
@@
 Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
-For example, the smoothed variable for Area (km^2^) (p = `r round(summary(model5_all)$s.table[4,1],2)`)) and Dryness (p = `r round(summary(model5_all)$s.table[4,2],2)`))still provides reasonable explanation of changes in stream flow.
-However *length* does not explain any of the variation.
-In contrast, *From*, which indicates the start date of the study (and therefore the age of the study) has an effect (p = `r round(summary(model5_all)$s.table[4,4],2)`)).
-However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
-Including interactions into the smooths is possible, but the results are even more difficult to interpret given the high flexibility of the two-dimensional smooth.
-
-We now also include the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) i.e. Equation \@ref(eq:eq2).
+For example, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(model5_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(model5_all)$s.table[2,4],2)`)) and *length* (p = `r round(summary(model5_all)$s.table[3,4],2)`)) explain significant variation in the data.
+
+However, including the non-linearity also increases the chance of over fitting, as the smoothing splines allow significant flexibility.
+Including interactions between the smooth variables is also possible, but the results are difficult to interpret given the high flexibility of the two-dimensional smooth.
+Given the overall variability in the data we did not attempt this.
+
+Finally the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included i.e. Equation \@ref(eq:eq2).
 
@@
 model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts" ) + 
-                     s(length, bs="ts") + s(From, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
+                    s(log10(Area_km2), k = 5, bs="ts") + 
+                    s(Dryness, k = 10, bs="ts" ) + 
+                     s(length, k = 35, bs="ts") +# s(From, k = 40, bs = "ts") +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
                     Hydrological_regime
@@
 This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
-This indicates that the categorical variables explain a limited amount of the variance.
-However, interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant.
+This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
+However, it is interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant.
 In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
@@
 L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
-From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
-(D + A) / (L + From) 
+#From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
+(D + A) / L 
 # visreg(model6_all,"Dryness",gg=T) + theme_bw()
@@
 
-```{r smoothsmodel6, fig.cap="Visualisation of the smooth variables in the model", out.width="90%", echo=F}
+```{r smoothsmodel6, fig.cap="Visualisation of the smooth variables in the model.
+The shaded areas are the 95% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationshiop.
+The data are plotted as individual points", out.width="90%", echo=F}
 
@@
 
-Figure \@ref(fig:smoothsmodel6) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the reason why using penalized smooths following @wood2006 is useful).
-It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
-Similarly, the *length* variable is not significant.
-However, both *From* and *Dryness* variables show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
-For example, in the case of *From* the relationship is clearly strongly influenced by the two data points before 1920.
-Similarly, *length* and *Dryness* have points with high leverage.
+Figure \@ref(fig:smoothsmodel6) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear, but, given all the data, not significant at p = `r round(summary(model6_all)$s.table[1,4],2)`, likely due to the high variance in the data.
+It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+Both the *length* and *Dryness* variables are significant and show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+For example, *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the data.
 
@@
 
-NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4, From > 1920)
+NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
 
 model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
+                    s(Dryness, k = 10, bs="ts" ) + 
+                #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), k = 5, bs = "ts") +
+                #length +
+                s(length, k = 10, bs = "ts") + #s(From, k=40, bs = "ts") +
+                    Precip_data_type +  Assessment_technique +
+                Forest_type +
                     Hydrological_regime, data = NoLong_Data)
-#summary(model7)
+# summary(model7)
+# gam.check(model7)
+# plot(model7)
 ```
@@
 ```{r, echo = F}
-pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness > 4 and studies commencing after 1920")
+pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness < 4.")
 ```
@@
 L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
-From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
-(D + A) / (L + From) 
+#From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
+(D + A) / L#(L + From) 
 # visreg(model6_all,"Dryness",gg=T) + theme_bw()
@@
 The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005.
-One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables.
-The next model therefore removes the following data: *Dryness* > 4, *length* > 60 years, *From* > 1920.
+The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) does not reflect a similar response to @filoso2017 and @jackson2005.
+One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \@ref(fig:smoothsmodel6)).
+Therefore it can be important to investigate what removing these few data points has on the overall model and the significance of the variables.
+The next model therefore removes the following data: *Dryness* > 4 and *length* > 60 years.
 This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
 
-The model has more explaining power with an adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
-This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \@ref(fig:smoothsmodel7)).
-However the other 3 variables are no longer significant due to the scatter in the data, but *length* and *From* show some strong local non-linear behaviour.
+This last model has more explaining power with an adjusted *r^2^* of `r round(summary(model7)$r.sq,2)`.
+* The results indicate that *Dryness* indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodel7)).
+Catchment area (*log10(Area (km^2^))*) also shows reasonable evidence of having an impact on flow with p = `r round(summary(model7)$s.table[2,4],2)`, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+The variable *length* no longer is significant, after removal of the two studies with very long lengths.
 
@@
 
-Essentially, the analysis shows that there is no clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), and, in contrast to @zhang2017, there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), however, in contrast to @zhang2017, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
 If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \@ref(fig:overview)).
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average the recorded forest cover change is approximately  10\% lower.
+The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average, the forest cover change data is approximately  10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to "implement" in an experiment.
 
-This is also reflected in the second caveat most of the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
+This is also reflected in the second caveat.
+Most of the data from the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
 For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
@@
 Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
+As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamdlow.
 
@@
 
-One concern is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+One concern with the result presented heer, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
 This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
@@
 model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
+                    s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), k = 10, bs = "ts") + s(length, bs = "ts") + #s(From, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4, From > 1920))
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4))
 #plot(model8_assess)
-#summary(model8_assess)
-
+# summary(model8_assess)
+# gam.check(model8_assess)
 ```
@@
 In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.
+However, as indicated earlier, the EA studies in the database are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.
 
@@
 In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies [@zhang2017; @zhou2015; @filoso2017].
-This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
 As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
-
-Excluding the few catchments that have very high dryness values, clarified the *Dryness* trend, agreeing with earlier studies and showing a increase in the change in flow for drier catchments.
-However, this really only starts to have an effect for Dryness > 2.
+This points to a need for future studies that unravel this interaction.
 
@@
 The relationships between forest cover change and *Area (km^2^)* and between *Area (km^2^)* and assessment technique have already been highlighted.
-However there are further unexplored interactions between the study length and some of the variables and *From* and the other variables.
-
-A principle component analysis of the numeric data reveals some of these interactions (Figure \@ref(fig:pcaplot)), such as between *length* and *Dryness* and between *From* and *Area (km^2^)*.
+However there are further unexplored interactions between the study length and some of the variables.
+
+A principle component analysis of the numeric data reveals some of these interactions (Figure \@ref(fig:pcaplot)), such as between *length* and *Dryness*.
 Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
 
-```{r pcaplot, fig.cap="Biplot of the first two principle components using a principle component analysis on the numerical values of the ", out.width="90%", echo=F, eval =T}
+```{r pcaplot, fig.cap="Biplot of the first two principle components using a principle component analysis on the numerical values of the data set", out.width="90%", echo=F, eval =T}
 
@@
 
-In contrast to @filoso2017, we did not identify the length of the observation period as a significant variable in our final model.
-There are further confounding factors in the data, which were also classified by @filoso2017 and these might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
+In contrast to @filoso2017, we did not identify that the length of the observation period is a significant variable in our final model.
+However, there are further confounding factors in the data, which were not analysed in this study.
+These were also classified by @filoso2017 and these factors might create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
 However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
 This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
+
+Large databases based on historical studies, such as used here, also have significant uncertainty.
+While we have reviewed a large number of the studies in more detail, we have generally assumed that the assessments of past authors of the changes in streamflow and changes in forest cover are correct.
+More generally a lot of the data in the database are "summary data" extracted from the paper and this often neglects a lot of possibly important detail in the original studies.
+This introduces additional uncertainty in the analysis.
+
 By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
@@
 
-More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
-It demonstrates that analysis of large databases of essentially "aggregated data" should be considered carefully.
-
-Results of this statistical analysis need to be considered "conditional on the data"
-Conditional on the data, the impact of forestry on streamflow is:
-
-- Greater for forest clearing then for reforestation
-
-- Possibly reduced for larger watersheds, but this is only in interaction with other factors such as climate
-
-- Increases for drier watersheds
-
-- Sensitive to the assessment method used in the historical data
-
-Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area.
-Furthermore, the current study analyses a large global dataset.
+More rigorous checking of an existing database on catchment studies relating to changes in forest cover to changes in flow and more detailed statistical analysis results in both agreement and disagreement with older studies.
+It demonstrates that analysis of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often fail to capture the complexity in the data.
+The variablility in the aggregated historical data is simply too large.
+
+As with any analysis, the results of the statistical analysis in this paper need to be considered "conditional on the data". 
+Conditional on the data, it can be determined that the impact of forestry on streamflow:
+
+- is greater for forest clearing then for reforestation;
+
+- is reduced for larger watersheds;
+
+- Increases for drier watersheds; and
+
+- is sensitive to the assessment method used in the historical data.
+
+Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area, especially in relation to larger catchments.
+Furthermore, the current study analyses a large global dataset of aggregated data.
 This analysis does not exclude more local and regional effects that cannot be identified in the global data.
+In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
 

### 2022-03-08 paper/Forest_and_Water/Forest_and_Water.tex
addressing WBs comments

@@
   \title{Factors determining how catchments respond to forest cover change. Re-analysing global data sets.}
-    \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
+    \author[DARE]{R. Willem Vervoort\corref{1}}
    \ead{willem.vervoort@sydney.edu.au} 
@@
    \ead{jalonso@fing.edu.uy} 
-      \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
-    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
+      \address[DARE]{ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
+    \address[The University of Sydney]{The University of Sydney, Sydney, NSW 2006, Australia}
     \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
@@
   \begin{abstract}
-  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
-Deforestation causes a 32\% greater change in flow compared to reforestation.
-Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies.
-Smaller studies dominate the database with 42\% of the data \textless{} 1 km2 and 65\% of the data \textless{} 10 km2.
-Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
   Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
@@
 Generalised additive models were used to run flexible regressions including multiple variables.
+  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
+Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover.
+Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data.
+Smaller catchment dominate the database with 42\% of the data \textless{} 1 km\textsuperscript{2} and 65\% of the data \textless{} 10 km\textsuperscript{2}.
+Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
+Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
   \end{abstract}
@@
 Many of the US studies use \emph{watershed}, while European and Australian studies use \emph{catchment}.
-In particular, we retained the term ``paired watershed studies'' as this is the most common terminology, but further mostly use the term catchment.
+In particular, we retained the term ``paired watershed studies'' and ``quasi-paired watershed studies'' as this is the most common terminology, but further mostly use the term catchment.
 
@@
 
-The second study (Filoso et al., 2017) was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
+The second study (Filoso et al., 2017) is a systematic review of reforestation studies (only studies in which forest cover increased).
+This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of quantitative data (37) they showed a log-linear relationship between decreasing catchment size and an increasing decline in streamflow.
+For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with Zhang et al.
+(2017)), but they did not identify a distinct threshold and fitted a log-linear relationship.
 In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
@@
 (2017).
-However, this paper aims to explain the variation in the data using the Fuh model, and in particular aims to link the variation in the observed data to variations in the exponent \emph{m} in the model.
-A key observation is that in drier environments, the effects of deforestation are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al.
+However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
+In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model.
+A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al.
 (2017).
 
-Encouraged by the work presented by Zhang et al.
+Encouraged by the work from Zhang et al.
 (2017), Filoso et al.
@@
 (2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
-In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.
+In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.
 
@@
 (2017) use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
-In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
-We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) in our analysis.
+In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET.
+In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).
+In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
+We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) from Zhang et al. (2017) in our analysis.
 
@@
 (2017).
-In particular, we focussed on the 37 data points included in the quantitative analysis in Filoso et al.
+In particular, we focussed on including the 37 data points related to the quantitative analysis in Filoso et al.
 (2017).
 
-In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location.
-Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
-For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.
+In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location.
+These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps\textsuperscript{TM}.
+In the dataset, an additional column has been added to indicate the source of the location data.
+
+Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al.
+(2017) and Jackson et al.
+(2005).
+Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
+For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
 
@@
 Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987).
+It is not clear if this is an effect of increased water use in growth (Vertessy et al., 2001) or due to changes in interception (Stoof et al., 2012).
 Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al.
@@
 (2012), and these were checked against the existing studies to prevent overlap.
-In the citation column in the data set, in general the main reference for the calculated change in streamflow was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al.
+In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al.
 (2011)).
@@
 (2017) and Filoso et al.
-(2017), 29 potential duplicates were identified and flagged in the data.
-
-The final column in the improved data set is a ``notes'' column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al.
+(2017), 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
+
+The final column in the improved data set is a ``notes'' column, which we added, but is not further used in the analysis.
+It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al.
 (2017).
+This will allow future research to further sructinise our input for errors.
 
@@
 Here \(X_i\) are factorial variables, while \(Z_i\) are continuous variables.
-The model assumes no direct interactions and all variables are additive.
+The model intially assumes no direct interactions and all variables are additive.
+We will comment in this assumption in the discussion.
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
 In Zhang et al.
-(2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation.
+(2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data is converted to positive values, and an additional column (\(sign_{forest cover}\)) is added that indicates whether it was a forest cover increase or decrease.
-
-A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have a linear or non-linear relationship with \(\Delta Qf \%\).
-This means that a smooth function \(s()\) is applied to the \(Z_i\) variables.
-For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant (Wood, 2006).
-
-For the model in equation 2, we initially only used the data from Zhang et al.
-(2017) to make sure that the additional catchments added to the data set did not influence the results.
+Therefore all the change in forest cover data was converted to positive values, and an additional factorial column (\(sign_{forest cover}\)) is included indicating whether it was a forest cover increase or decrease.
+
+A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with \(\Delta Qf \%\).
+This means that a smooth function \(s()\) can be applied to the \(Z_i\) variables.
+For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty.
+The result of this approach is that for high enough smoothing parameters (i.e.~if the data is very ``wiggly'') the smooth term can be shrunk to 0 and thus will be no longer significant (Wood, 2006).
+This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship.
+In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.
+
+The over arching test focuses on the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: Zhang et al.
+(2017); Filoso et al.
+(2017); Zhou et al.
+(2015)): climate, size of catchment and length of study.
+Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.
+
+As an initial approach we only used the data from Zhang et al.
+(2017) to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results).
 Subsequently the analysis was repeated and the additionally identified catchments were added.
-
-More generally the results were analysed to identify:\\
-1. the significance of the different variables\\
-2. the direction of the categorical or shape of the smooth variables
 
@@
 
-The overall dataset contains 330 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+The overall dataset contains 327 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
 In contrast, the original dataset from Zhang et al.
@@
 (2017)).
-The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\^{}2}.
+The current number of catchments is the result of the removal of duplicates and our modifications and additions.
+The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\textsuperscript{2}}.
 The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
 Note also the large number of studies with 100\% forest cover reduction.
-Smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).
+Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables.}\label{fig:datagraphs}
+\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables.
+Note that the first panel (showing the distribution of the catchment areas) indicates the distribution of the \emph{log\_10\_} transformed Area (in km\textsuperscript{2}).}\label{fig:datagraphs}
 \end{figure}
 
-In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2).
+Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
 In other words, flow almost always increased.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of increases and decreases in forest cover}\label{fig:increasedecrease}
+\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of decreases (top) and increases (bottom) in forest cover}\label{fig:increasedecrease}
 \end{figure}
@@
 The first three rows relate to the model using the original data base from Zhang et al.~(2017).
-The bottom three rows are the results of the model including the new data.}\tabularnewline
+The bottom three rows are the results of the model including the new data.
+Clearly there is no major change arising from the additional data.}\tabularnewline
 \toprule
@@
 \textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
-\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
-\textbf{(Intercept)} & 10.76 & 5.34 & 2.02 & 0.04 \\
-\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.04 & 0 \\
-\textbf{Forest\_Signincrease} & -38.03 & 5.26 & -7.23 & 0 \\
+\textbf{Forest\_SignIncrease} & -29.17 & 5.79 & -5.04 & 0 \\
+\textbf{(Intercept)} & 10.12 & 5.32 & 1.9 & 0.06 \\
+\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.24 & 0 \\
+\textbf{Forest\_SignIncrease} & -36.8 & 5.26 & -6.99 & 0 \\
 \bottomrule
@@
 Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
+Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total Zhang et al.
+(2017) data set.
+But this also means that our re-analysis of the data can be directly compared to the original study.
 
@@
 
-As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude.
-These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps.
-In the dataset, an additional column is added to indicate the source of the location data.
+Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients.
 As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
-\caption{\label{tab:out-model3} Results of the model including Latitude and Longitude including new data}\tabularnewline
+\caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{(Intercept)} & 10.99 & 5.75 & 1.91 & 0.06 \\
-\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.13 & 0 \\
-\textbf{Forest\_Signincrease} & -38.99 & 5.54 & -7.04 & 0 \\
-\textbf{Latitude} & -0.01 & 0.09 & -0.13 & 0.9 \\
-\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
+\textbf{(Intercept)} & 14.74 & 5.35 & 2.76 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.5 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.72 & 5.47 & -6.53 & 0 \\
+\textbf{Latitude} & -0.18 & 0.05 & -3.45 & 0 \\
+\textbf{Longitude} & -0.01 & 0.01 & -1.89 & 0.06 \\
 \bottomrule
@@
 
-There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence on the overall result.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.23 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.27 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 
-Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al.
-(2017) and Jackson et al.
-(2005).
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
-Initially, we tested models using annual average precipitation (\emph{Pa (mm)}), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable.
+While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
+Potential evapotranspiration (\emph{E0}) by itself was not significant in the.
+Initially, we also tested models using only the annual average precipitation (\emph{Pa (mm)}), but interactions between precipitation and evapotranspiraton might be captured by the dryness index.
+Both dryness index and \emph{Pa (mm)} were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
+As a result only the dryness index was retained as a climate indicator to align with the earlier work by Zhang et al.
+(2017).
 Given that Latitude and Longitude were not significant, we dropped these from the model.
@@
 \endhead
-\textbf{(Intercept)} & 8.83 & 6.3 & 1.4 & 0.16 \\
-\textbf{DeltaF\_perc\_pos} & 0.4 & 0.08 & 4.88 & 0 \\
-\textbf{Forest\_Signincrease} & -38.26 & 5.32 & -7.19 & 0 \\
-\textbf{Dryness} & 1.7 & 2.67 & 0.64 & 0.52 \\
+\textbf{(Intercept)} & 2.11 & 6.08 & 0.35 & 0.73 \\
+\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.06 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.69 & 5.26 & -6.78 & 0 \\
+\textbf{Dryness} & 7.09 & 2.58 & 2.75 & 0.01 \\
 \bottomrule
@@
 
-The results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness modulating the change in streamflow as a function of the change in forest cover change, which is surprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.52).
+Similar to \emph{E0} or \emph{Pa\_mm}, the results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
+This might seem suprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
+In this case, the evidence is highly doubtful (p = 0.01).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.22.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.24.
+As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
 
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.42}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.17}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.42}}@{}}
 \caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
@@
 \begin{minipage}[b]{\linewidth}\centering
+Number
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
 Latitude
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Watershed name
+Catchment name
 \end{minipage} \\
@@
 \endhead
-34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
-36.4 & -120.4 & Cantua \\
-32.74 & -111.5 & Natural Drainages, Ariz.,
+76 & 34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
+90 & 36.4 & -120.4 & Cantua, CA, U.S.A. \\
+295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, A \\
+226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
 U.S.A, C \\
-32.74 & -111.5 & Natural Drainages, Ariz.,
-U.S.A, A \\
-34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
--25.75 & 28.23 & Queens river \\
 \bottomrule
@@
 
-The major hypothesis to test is the effect of area on the change in flow, following the analysis by Zhang et al.
+The second major variable is the effect of area on the change in flow, following the analysis by Zhang et al.
 (2017) and Filoso et al.
@@
 \endhead
-\textbf{(Intercept)} & 15.37 & 7.24 & 2.12 & 0.03 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 2.94 & 0 \\
-\textbf{Forest\_Signincrease} & -36.61 & 5.38 & -6.8 & 0 \\
-\textbf{Dryness} & 2.48 & 2.69 & 0.92 & 0.36 \\
-\textbf{log10(Area\_km2)} & -3.03 & 1.67 & -1.81 & 0.07 \\
+\textbf{(Intercept)} & 9.28 & 7.18 & 1.29 & 0.2 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.09 & 0 \\
+\textbf{Forest\_SignIncrease} & -33.78 & 5.34 & -6.32 & 0 \\
+\textbf{Dryness} & 7.35 & 2.57 & 2.86 & 0 \\
+\textbf{log10(Area\_km2)} & -3.04 & 1.64 & -1.86 & 0.06 \\
 \bottomrule
@@
 
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.07) that there is a reduction in the effect of forest cover change on streamflow related to Area (km\textsuperscript{2}) (Table \ref{tab:out-modelArea}).
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.06) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
 In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including this interaction in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.95)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.59)) as an interaction with Dryness.
 
@@
 \endhead
-\textbf{(Intercept)} & 11.95 & 7.43 & 1.61 & 0.11 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.85 & 0 \\
-\textbf{Forest\_Signincrease} & -34.62 & 5.46 & -6.34 & 0 \\
-\textbf{Dryness} & 5.46 & 3.1 & 1.76 & 0.08 \\
-\textbf{log10(Area\_km2)} & 0.15 & 2.35 & 0.06 & 0.95 \\
-\textbf{Dryness:log10(Area\_km2)} & -2.62 & 1.37 & -1.91 & 0.06 \\
+\textbf{(Intercept)} & 6.61 & 7.39 & 0.9 & 0.37 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.01 & 0 \\
+\textbf{Forest\_SignIncrease} & -32.58 & 5.39 & -6.04 & 0 \\
+\textbf{Dryness} & 9.79 & 3.05 & 3.21 & 0 \\
+\textbf{log10(Area\_km2)} & -1.11 & 2.09 & -0.53 & 0.59 \\
+\textbf{Dryness:log10(Area\_km2)} & -1.91 & 1.29 & -1.49 & 0.14 \\
 \bottomrule
@@
 (2005)).
-In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (\emph{Dryness, Area, From}) were considered non-linear in this model.
+In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (\emph{Dryness} and \emph{Area}) were considered non-linear.
 As a result a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
-\emph{From} represents the starting date of the data collection.
 
 \begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon \label{eq:eq7}
+\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon \label{eq:eq7}
 \end{align}
@@
 \endhead
-\textbf{(Intercept)} & 13.99 & 5.68 & 2.46 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.51 & 0 \\
-\textbf{Forest\_Signincrease} & -37.18 & 5.48 & -6.78 & 0 \\
+\textbf{(Intercept)} & 13.22 & 5.7 & 2.32 & 0.02 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.94 & 0 \\
+\textbf{Forest\_SignIncrease} & -29.88 & 6.03 & -4.96 & 0 \\
 \bottomrule
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.11}}
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.68 & 9 & 0.28 & 0.05 \\
-\textbf{s(Dryness)} & 4.28 & 9 & 0.94 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.55 \\
-\textbf{s(From)} & 4.62 & 9 & 0.9 & 0.1 \\
+\textbf{s(log10(Area\_km2))} & 2.32 & 5 & 1.42 & 0.02 \\
+\textbf{s(Dryness)} & 4.69 & 9 & 1.92 & 0 \\
+\textbf{s(length)} & 28.53 & 49 & 1.17 & 0 \\
 \bottomrule
@@
 
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.26 and deviance explained of 0.29, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, the smoothed variable for Area (km\textsuperscript{2}) (p = 4.62)) and Dryness (p = 9))still provides reasonable explanation of changes in stream flow.
-However \emph{length} does not explain any of the variation.
-In contrast, \emph{From}, which indicates the start date of the study (and therefore the age of the study) has an effect (p = 0.1)).\\
-However, it also increases the chance of over fitting, as the smoothing splines allow significant flexibility, which will be investigated later.
-Including interactions into the smooths is possible, but the results are even more difficult to interpret given the high flexibility of the two-dimensional smooth.
-
-We now also include the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) i.e.~Equation \eqref{eq:eq2}.
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.35 and deviance explained of 0.42, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
+
+However, including the non-linearity also increases the chance of over fitting, as the smoothing splines allow significant flexibility.
+Including interactions between the smooth variables is also possible, but the results are difficult to interpret given the high flexibility of the two-dimensional smooth.
+Given the overall variability in the data we did not attempt this.
+
+Finally the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included i.e.~Equation \eqref{eq:eq2}.
 
@@
 \endhead
-\textbf{(Intercept)} & -29.23 & 18.18 & -1.61 & 0.11 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.29 & 0 \\
-\textbf{Forest\_Signincrease} & -23.48 & 6.84 & -3.43 & 0 \\
-\textbf{Precip\_data\_typeOB} & -5.03 & 13.94 & -0.36 & 0.72 \\
-\textbf{Precip\_data\_typeSG} & 19.41 & 15.38 & 1.26 & 0.21 \\
-\textbf{Assessment\_techniqueEA, HM} & 19.56 & 45.02 & 0.43 & 0.66 \\
-\textbf{Assessment\_techniqueHM} & 33.57 & 12.36 & 2.72 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 52.79 & 11.72 & 4.5 & 0 \\
+\textbf{(Intercept)} & -17.49 & 20.4 & -0.86 & 0.39 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.28 & 0 \\
+\textbf{Forest\_SignIncrease} & -23.92 & 7.11 & -3.37 & 0 \\
+\textbf{Precip\_data\_typeOB} & -9.76 & 14.34 & -0.68 & 0.5 \\
+\textbf{Precip\_data\_typeSG} & 18.48 & 16.99 & 1.09 & 0.28 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.69 & 46.67 & 0.4 & 0.69 \\
+\textbf{Assessment\_techniqueHM} & 24.58 & 14.5 & 1.7 & 0.09 \\
+\textbf{Assessment\_techniquePWE} & 42.73 & 14.9 & 2.87 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 43.58 & 45.93 & 0.95 & 0.34 \\
-\textbf{Assessment\_techniqueQPW} & 39.98 & 21.36 & 1.87 & 0.06 \\
+HM} & 27.27 & 47.04 & 0.58 & 0.56 \\
+\textbf{Assessment\_techniqueQPW} & 35.11 & 22.51 & 1.56 & 0.12 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 31.36 & 27.12 & 1.16 & 0.25 \\
-\textbf{Assessment\_techniqueSH} & 42.63 & 12.41 & 3.44 & 0 \\
-\textbf{Forest\_typeCF} & -2.33 & 8.23 & -0.28 & 0.78 \\
-\textbf{Forest\_typeMF} & -10.36 & 8 & -1.3 & 0.2 \\
-\textbf{Hydrological\_regimeSD} & 6.99 & 10.25 & 0.68 & 0.5 \\
+EA} & 35.56 & 26.84 & 1.33 & 0.19 \\
+\textbf{Assessment\_techniqueSH} & 41.11 & 13.23 & 3.11 & 0 \\
+\textbf{Forest\_typeCF} & -4.51 & 8.34 & -0.54 & 0.59 \\
+\textbf{Forest\_typeMF} & -6.26 & 8.47 & -0.74 & 0.46 \\
+\textbf{Hydrological\_regimeSD} & 5.88 & 10.16 & 0.58 & 0.56 \\
 \bottomrule
 \end{longtable}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.11}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+\caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Ref.df
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+F
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+p-value
+\end{minipage} \\
+\midrule
+\endhead
+\textbf{s(log10(Area\_km2))} & 0.69 & 4 & 0.33 & 0.15 \\
+\textbf{s(Dryness)} & 4.27 & 9 & 1.45 & 0.01 \\
+\textbf{s(length)} & 17.14 & 34 & 1.05 & 0 \\
+\bottomrule
+\end{longtable}
+
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.34.
+This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
+However, it is interesting to note from Table \ref{tab:msix-linear} that several of the assessment methods are significant.
+In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
+In this case, \emph{log10(Area (km\textsuperscript{2}))} is no longer a significant predictor, the reasons for this will be discussed later.
+
+\textbackslash begin\{figure\}
+\includegraphics[width=0.9\linewidth]{model6_smooths} \textbackslash caption\{Visualisation of the smooth variables in the model.
+The shaded areas are the 95\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationshiop.
+The data are plotted as individual points\}\label{fig:smoothsmodel6}
+\textbackslash end\{figure\}
+
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.15, likely due to the high variance in the data.
+It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+Both the \emph{length} and \emph{Dryness} variables are significant and show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
+For example, \emph{length} and \emph{Dryness} have several points with very high leverage that determine much of the non-linearity in the data.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
+\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness \textless{} 4.}\tabularnewline
 \toprule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0 & 9 & 0 & 0.37 \\
-\textbf{s(Dryness)} & 3.96 & 9 & 1.68 & 0 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.36 \\
-\textbf{s(From)} & 8.8 & 9 & 3.64 & 0 \\
-\bottomrule
-\end{longtable}
-
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.34.
-This indicates that the categorical variables explain a limited amount of the variance.
-However, interesting to note from Table \ref{tab:msix-linear} that several of the assessment methods are significant.
-In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
-In this case, \emph{log10(Area (km\textsuperscript{2}))} is no longer a significant predictor, the reasons for this will be discussed later.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{model6_smooths} \caption{Visualisation of the smooth variables in the model}\label{fig:smoothsmodel6}
-\end{figure}
-
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the reason why using penalized smooths following Wood (2006) is useful).
-It still has a negative slope, indicating that in larger catchments the impact of changes in forest cover on streamflow is less than for smaller catchments.
-Similarly, the \emph{length} variable is not significant.
-However, both \emph{From} and \emph{Dryness} variables show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
-For example, in the case of \emph{From} the relationship is clearly strongly influenced by the two data points before 1920.
-Similarly, \emph{length} and \emph{Dryness} have points with high leverage.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
-\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years, Dryness \textgreater{} 4 and studies commencing after 1920}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{s(Dryness)} & 3.21 & 9 & 3.22 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.46 & 9 & 0.11 & 0.13 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.44 \\
-\textbf{s(From)} & 4.06 & 9 & 1.12 & 0.05 \\
+\textbf{s(Dryness)} & 2.9 & 9 & 2.8 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.71 & 4 & 0.59 & 0.06 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.91 \\
 \bottomrule
@@
 The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) also does not reflect the type of response highlighted in Filoso et al.
+The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) does not reflect a similar response to Filoso et al.
 (2017) and Jackson et al.
 (2005).
-One reason could be the few data points with very long data series, and very old studies (before 1920 essentially), and highly variable responses (Figure \ref{fig:smoothsmodel6}).
-Therefore it can be important to investigate what removing these factors has on the overall model and the significance of the variables.
-The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4, \emph{length} \textgreater{} 60 years, \emph{From} \textgreater{} 1920.
-This result in a reduction of the data set from 330 to 304 catchments.
-
-The model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.34.
-This results in Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover on streamflow (Figure \ref{fig:smoothsmodel7}).
-However the other 3 variables are no longer significant due to the scatter in the data, but \emph{length} and \emph{From} show some strong local non-linear behaviour.
+One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \ref{fig:smoothsmodel6}).
+Therefore it can be important to investigate what removing these few data points has on the overall model and the significance of the variables.
+The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4 and \emph{length} \textgreater{} 60 years.
+This result in a reduction of the data set from 327 to 304 catchments.
+
+This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.29.
+* The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7}).
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.06, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
 
@@
 
-Essentially, the analysis shows that there is no clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), and, in contrast to Zhang et al.
-(2017), there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry.
+Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to Zhang et al.
+(2017), there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
 If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.76) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average the recorded forest cover change is approximately 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.81) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
 
-This is also reflected in the second caveat most of the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
+This is also reflected in the second caveat.
+Most of the data from the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
 For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
@@
 HM & 57 \\
-SH & 43 \\
-EA & 34 \\
+SH & 42 \\
+EA & 32 \\
 QPW & 7 \\
@@
 
-One concern is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+One concern with the result presented heer, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
 This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
@@
 \endhead
-\textbf{(Intercept)} & -27.03 & 18.34 & -1.47 & 0.14 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.96 & 0 \\
-\textbf{Forest\_Signincrease} & -22.23 & 6.83 & -3.25 & 0 \\
-\textbf{Precip\_data\_typeOB} & -11.91 & 13.85 & -0.86 & 0.39 \\
-\textbf{Precip\_data\_typeSG} & 18.07 & 15.65 & 1.15 & 0.25 \\
-\textbf{Assessment\_techniqueHM} & 36.48 & 12.3 & 2.97 & 0 \\
-\textbf{Assessment\_techniquePWE} & 56.52 & 12.33 & 4.58 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 42.07 & 21.03 & 2 & 0.05 \\
-\textbf{Assessment\_techniqueSH} & 48.82 & 12.33 & 3.96 & 0 \\
-\textbf{Forest\_typeCF} & -4.64 & 8.18 & -0.57 & 0.57 \\
-\textbf{Forest\_typeMF} & -5.22 & 8.08 & -0.65 & 0.52 \\
-\textbf{Hydrological\_regimeSD} & 8.29 & 10.25 & 0.81 & 0.42 \\
+\textbf{(Intercept)} & -18.96 & 18.8 & -1.01 & 0.31 \\
+\textbf{DeltaF\_perc\_pos} & 0.28 & 0.1 & 2.83 & 0 \\
+\textbf{Forest\_SignIncrease} & -25.46 & 6.8 & -3.75 & 0 \\
+\textbf{Precip\_data\_typeOB} & -9.75 & 13.25 & -0.74 & 0.46 \\
+\textbf{Precip\_data\_typeSG} & 19.37 & 15.64 & 1.24 & 0.22 \\
+\textbf{Assessment\_techniqueHM} & 31.9 & 12.97 & 2.46 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 44.99 & 13.59 & 3.31 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 40.11 & 21.12 & 1.9 & 0.06 \\
+\textbf{Assessment\_techniqueSH} & 46.08 & 12.43 & 3.71 & 0 \\
+\textbf{Forest\_typeCF} & -3.42 & 7.94 & -0.43 & 0.67 \\
+\textbf{Forest\_typeMF} & -3.02 & 8.08 & -0.37 & 0.71 \\
+\textbf{Hydrological\_regimeSD} & 2.46 & 9.87 & 0.25 & 0.8 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 2.93 & 9 & 2.41 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.45 & 9 & 0.1 & 0.14 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.4 \\
-\textbf{s(From)} & 3.84 & 9 & 0.66 & 0.17 \\
+\textbf{s(Dryness)} & 2.87 & 9 & 2.77 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.7 & 9 & 0.26 & 0.07 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.93 \\
 \bottomrule
@@
 (2015)) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies are all on the drier end of the \emph{Dryness} spectrum, highlighting another unresolved interaction in the data.
+However, as indicated earlier, the EA studies in the database are all on the drier end of the \emph{Dryness} spectrum, highlighting another unresolved interaction in the data.
 
@@
 In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies (Filoso et al., 2017; Zhang et al., 2017; Zhou et al., 2015).
-This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by the groundwater inputs is not as great.
+This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
 As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
-
-Excluding the few catchments that have very high dryness values, clarified the \emph{Dryness} trend, agreeing with earlier studies and showing a increase in the change in flow for drier catchments.
-However, this really only starts to have an effect for Dryness \textgreater{} 2.
+This points to a need for future studies that unravel this interaction.
 
@@
 The relationships between forest cover change and \emph{Area (km\textsuperscript{2})} and between \emph{Area (km\textsuperscript{2})} and assessment technique have already been highlighted.
-However there are further unexplored interactions between the study length and some of the variables and \emph{From} and the other variables.
-
-A principle component analysis of the numeric data reveals some of these interactions (Figure \ref{fig:pcaplot}), such as between \emph{length} and \emph{Dryness} and between \emph{From} and \emph{Area (km\textsuperscript{2})}.
+However there are further unexplored interactions between the study length and some of the variables.
+
+A principle component analysis of the numeric data reveals some of these interactions (Figure \ref{fig:pcaplot}), such as between \emph{length} and \emph{Dryness}.
 Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{PCA_biplot_Alldata} \caption{Biplot of the first two principle components using a principle component analysis on the numerical values of the }\label{fig:pcaplot}
+\includegraphics[width=0.9\linewidth]{PCA_biplot_Alldata} \caption{Biplot of the first two principle components using a principle component analysis on the numerical values of the data set}\label{fig:pcaplot}
 \end{figure}
@@
 In contrast to Filoso et al.
-(2017), we did not identify the length of the observation period as a significant variable in our final model.
-There are further confounding factors in the data, which were also classified by Filoso et al.
-(2017) and these might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD), are dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
+(2017), we did not identify that the length of the observation period is a significant variable in our final model.
+However, there are further confounding factors in the data, which were not analysed in this study.
+These were also classified by Filoso et al.
+(2017) and these factors might create biases in the data set that can impact the overall assessment.
+For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
 However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow (Vervoort et al., 2021).
-This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.\\
+This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
+
+Large databases based on historical studies, such as used here, also have significant uncertainty.
+While we have reviewed a large number of the studies in more detail, we have generally assumed that the assessments of past authors of the changes in streamflow and changes in forest cover are correct.
+More generally a lot of the data in the database are ``summary data'' extracted from the paper and this often neglects a lot of possibly important detail in the original studies.
+This introduces additional uncertainty in the analysis.
+
 By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
@@
 
-More rigorous checking of data and statistical analysis results in both agreement and disagreement with older studies.
-It demonstrates that analysis of large databases of essentially ``aggregated data'' should be considered carefully.
-
-Results of this statistical analysis need to be considered ``conditional on the data''
-Conditional on the data, the impact of forestry on streamflow is:
+More rigorous checking of an existing database on catchment studies relating to changes in forest cover to changes in flow and more detailed statistical analysis results in both agreement and disagreement with older studies.
+It demonstrates that analysis of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often fail to capture the complexity in the data.
+The variablility in the aggregated historical data is simply too large.
+
+As with any analysis, the results of the statistical analysis in this paper need to be considered ``conditional on the data.''
+Conditional on the data, it can be determined that the impact of forestry on streamflow:
 
@@
 \item
-  Greater for forest clearing then for reforestation
+  is greater for forest clearing then for reforestation;
 \item
-  Possibly reduced for larger watersheds, but this is only in interaction with other factors such as climate
+  is reduced for larger watersheds;
 \item
-  Increases for drier watersheds
+  Increases for drier watersheds; and
 \item
-  Sensitive to the assessment method used in the historical data
+  is sensitive to the assessment method used in the historical data.
 \end{itemize}
 
-Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area.
-Furthermore, the current study analyses a large global dataset.
+Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area, especially in relation to larger catchments.
+Furthermore, the current study analyses a large global dataset of aggregated data.
 This analysis does not exclude more local and regional effects that cannot be identified in the global data.
+In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-stoof2012}{}}%
+Stoof, C.R., Vervoort, R.W., Iwema, J., Elsen, E.
+van den, Ferreira, A.J.D., Ritsema, C.J., 2012.
+Hydrological response of a small catchment burned by experimental fire.
+Hydrol.
+Earth Syst.
+Sci.
+16, 267--285.
+doi:\href{https://doi.org/10.5194/hess-16-267-2012}{10.5194/hess-16-267-2012}
+
 \leavevmode\vadjust pre{\hypertarget{ref-thornton2007}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vertessy2001}{}}%
+Vertessy, R.A., Watson, F.G.R., O'Sullivan, S.K., 2001.
+Factors determining relations between stand age and catchment water balance in mountain ash forests.
+Forest Ecology and Management 143, 13--26.
+doi:\url{https://doi.org/10.1016/S0378-1127(00)00501-6}
+
 \leavevmode\vadjust pre{\hypertarget{ref-vervoort2021}{}}%

### 2022-03-08 paper/Forest_and_Water/Forest_and_Water.tex

@@
     \address[The University of Sydney]{The University of Sydney, Sydney, NSW 2006, Australia}
-    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \address[INIA]{Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
     \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
@@
 
-It is however it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
+However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
 The obvious ones being catchment dryness and area (following Zhang et al.

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.Rmd
version to publish 20220330

@@
 
+All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water) on the "publish" branch.
+
 ## Statistical modelling
@@
 
-Zhang_small <- read_csv("../../data/TablesZhangetal2017_small.csv")
-Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv") %>%
-  mutate(Complete = 1,
-         `Possible duplicate` = 0)
-new_data <- read_csv("../../data/TableNewData.csv")
-```
-
-<!-- read in the ET data -->
-
-```{r, echo = F, warning = F, message=F}
-PET_large <- read_csv("../../data/largeET.csv")
-PET_small <- read_csv("../../data/smallET.csv")
-
-# drop columns and make sure the titles are the same
-PET_large <- PET_large %>%
-  rename(Latitude = Latitiude,
-         E0 = rvalue_1,
-         `Watershed #` = Watershed)
-
-PET_small <- PET_small %>%
-  select(site_name, lat, long, rvalue_1) %>%
-  rename(Latitude = lat,
-         Longitude = long,
-         `Watershed #` = site_name,
-         E0 = rvalue_1)
-
-# 
-PET_all <- bind_rows(PET_large, PET_small)
-# There are 7 missing E0 values
-# > PET_all$`Watershed #`[is.na(PET_all$E0)==T]
-# [1]  38 278 269 243 236 299 250
+Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv")
+Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv")
+new_data <- read_csv("../../data/NewCatchments_T3.csv")
 ```
@@
 ```{r, echo = F}
-All_data <- left_join(All_data ,PET_all %>% select(-c(Longitude, Latitude)), by = "Watershed #")
-All_data <- All_data %>%
-  mutate(E0 = ifelse(is.na(E0.x)==T, E0.y, E0.x)) %>%
-  select(-c(E0.x,E0.y)) #%>%
-  #rename(Latitude = Latitude.x,
-#         Longitude = Longitude.x)
-
 # calculate dryness index

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.tex
version to publish 20220330

@@
 
+All the final data and analysis for this paper are located on github: \url{https://github.com/WillemVervoort/Forest_and_water} on the ``publish'' branch.
+
 \hypertarget{statistical-modelling}{%
@@
 
-The overall dataset contains 327 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+The overall dataset contains 350 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
 In contrast, the original dataset from Zhang et al.
-(2017) contained 312 catchments and the Filoso et al.
+(2017) contained 340 catchments and the Filoso et al.
 (2017) study used 37 catchments (Table S2 in Filoso et al.
@@
 \endhead
-\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
-\textbf{Forest\_SignIncrease} & -29.17 & 5.79 & -5.04 & 0 \\
-\textbf{(Intercept)} & 10.12 & 5.32 & 1.9 & 0.06 \\
-\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.24 & 0 \\
-\textbf{Forest\_SignIncrease} & -36.8 & 5.26 & -6.99 & 0 \\
+\textbf{(Intercept)} & 7.44 & 5.44 & 1.37 & 0.17 \\
+\textbf{DeltaF\_perc\_pos} & 0.48 & 0.08 & 5.66 & 0 \\
+\textbf{Forest\_SignIncrease} & -28.17 & 5.73 & -4.92 & 0 \\
+\textbf{(Intercept)} & 8.93 & 5.22 & 1.71 & 0.09 \\
+\textbf{DeltaF\_perc\_pos} & 0.45 & 0.08 & 5.63 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.81 & 5.21 & -6.88 & 0 \\
 \bottomrule
@@
 
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.23, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 \endhead
-\textbf{(Intercept)} & 14.74 & 5.35 & 2.76 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.5 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.72 & 5.47 & -6.53 & 0 \\
-\textbf{Latitude} & -0.18 & 0.05 & -3.45 & 0 \\
-\textbf{Longitude} & -0.01 & 0.01 & -1.89 & 0.06 \\
+\textbf{(Intercept)} & 11.31 & 5.65 & 2 & 0.05 \\
+\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.66 & 0 \\
+\textbf{Forest\_SignIncrease} & -38.1 & 5.49 & -6.95 & 0 \\
+\textbf{Latitude} & -0.1 & 0.09 & -1.05 & 0.3 \\
+\textbf{Longitude} & 0 & 0.03 & 0.14 & 0.89 \\
 \bottomrule
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.27 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.23 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 \endhead
-\textbf{(Intercept)} & 2.11 & 6.08 & 0.35 & 0.73 \\
-\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.06 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.69 & 5.26 & -6.78 & 0 \\
-\textbf{Dryness} & 7.09 & 2.58 & 2.75 & 0.01 \\
+\textbf{(Intercept)} & 6.34 & 6.17 & 1.03 & 0.3 \\
+\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.69 & 0 \\
+\textbf{Forest\_SignIncrease} & -36.27 & 5.22 & -6.95 & 0 \\
+\textbf{Dryness} & 1.87 & 2.6 & 0.72 & 0.47 \\
 \bottomrule
@@
 This might seem suprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.01).
+In this case, the evidence is highly doubtful (p = 0.47).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.24.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
 As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
@@
 76 & 34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
-90 & 36.4 & -120.4 & Cantua, CA, U.S.A. \\
-295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+90 & 36.4 & -120.4 & Cantua \\
+226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, C \\
 225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
 U.S.A, A \\
-226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
-U.S.A, C \\
+295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
+356 & -25.75 & 28.23 & Queens river \\
 \bottomrule
@@
 \endhead
-\textbf{(Intercept)} & 9.28 & 7.18 & 1.29 & 0.2 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.09 & 0 \\
-\textbf{Forest\_SignIncrease} & -33.78 & 5.34 & -6.32 & 0 \\
-\textbf{Dryness} & 7.35 & 2.57 & 2.86 & 0 \\
-\textbf{log10(Area\_km2)} & -3.04 & 1.64 & -1.86 & 0.06 \\
+\textbf{(Intercept)} & 12.41 & 7 & 1.77 & 0.08 \\
+\textbf{DeltaF\_perc\_pos} & 0.36 & 0.1 & 3.76 & 0 \\
+\textbf{Forest\_SignIncrease} & -34.4 & 5.31 & -6.48 & 0 \\
+\textbf{Dryness} & 2.48 & 2.61 & 0.95 & 0.34 \\
+\textbf{log10(Area\_km2)} & -2.97 & 1.64 & -1.81 & 0.07 \\
 \bottomrule
@@
 
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.06) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.07) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
 In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.59)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.92)) as an interaction with Dryness.
 
@@
 \endhead
-\textbf{(Intercept)} & 6.61 & 7.39 & 0.9 & 0.37 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.01 & 0 \\
-\textbf{Forest\_SignIncrease} & -32.58 & 5.39 & -6.04 & 0 \\
-\textbf{Dryness} & 9.79 & 3.05 & 3.21 & 0 \\
-\textbf{log10(Area\_km2)} & -1.11 & 2.09 & -0.53 & 0.59 \\
-\textbf{Dryness:log10(Area\_km2)} & -1.91 & 1.29 & -1.49 & 0.14 \\
+\textbf{(Intercept)} & 9.15 & 7.18 & 1.27 & 0.2 \\
+\textbf{DeltaF\_perc\_pos} & 0.35 & 0.1 & 3.65 & 0 \\
+\textbf{Forest\_SignIncrease} & -32.09 & 5.42 & -5.92 & 0 \\
+\textbf{Dryness} & 5.15 & 2.96 & 1.74 & 0.08 \\
+\textbf{log10(Area\_km2)} & 0.24 & 2.35 & 0.1 & 0.92 \\
+\textbf{Dryness:log10(Area\_km2)} & -2.64 & 1.39 & -1.9 & 0.06 \\
 \bottomrule
@@
 \endhead
-\textbf{(Intercept)} & 13.22 & 5.7 & 2.32 & 0.02 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.94 & 0 \\
-\textbf{Forest\_SignIncrease} & -29.88 & 6.03 & -4.96 & 0 \\
+\textbf{(Intercept)} & 13.42 & 5.42 & 2.48 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.16 & 0 \\
+\textbf{Forest\_SignIncrease} & -29.28 & 5.82 & -5.03 & 0 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 2.32 & 5 & 1.42 & 0.02 \\
-\textbf{s(Dryness)} & 4.69 & 9 & 1.92 & 0 \\
-\textbf{s(length)} & 28.53 & 49 & 1.17 & 0 \\
+\textbf{s(log10(Area\_km2))} & 3.39 & 5 & 2.23 & 0.01 \\
+\textbf{s(Dryness)} & 3.74 & 9 & 0.77 & 0.09 \\
+\textbf{s(length)} & 27.88 & 49 & 1.7 & 0 \\
 \bottomrule
@@
 
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.35 and deviance explained of 0.42, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.37 and deviance explained of 0.44, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.01)), \emph{Dryness} (p = 0.09)) and \emph{length} (p = 0)) explain significant variation in the data.
 
@@
 \endhead
-\textbf{(Intercept)} & -17.49 & 20.4 & -0.86 & 0.39 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.1 & 3.28 & 0 \\
-\textbf{Forest\_SignIncrease} & -23.92 & 7.11 & -3.37 & 0 \\
-\textbf{Precip\_data\_typeOB} & -9.76 & 14.34 & -0.68 & 0.5 \\
-\textbf{Precip\_data\_typeSG} & 18.48 & 16.99 & 1.09 & 0.28 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.69 & 46.67 & 0.4 & 0.69 \\
-\textbf{Assessment\_techniqueHM} & 24.58 & 14.5 & 1.7 & 0.09 \\
-\textbf{Assessment\_techniquePWE} & 42.73 & 14.9 & 2.87 & 0 \\
+\textbf{(Intercept)} & -21.82 & 19.28 & -1.13 & 0.26 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.16 & 0 \\
+\textbf{Forest\_SignIncrease} & -19.45 & 6.78 & -2.87 & 0 \\
+\textbf{Precip\_data\_typeOB} & -13.49 & 14.09 & -0.96 & 0.34 \\
+\textbf{Precip\_data\_typeSG} & 15.05 & 18.39 & 0.82 & 0.41 \\
+\textbf{Assessment\_techniqueEA, HM} & 12.1 & 45.85 & 0.26 & 0.79 \\
+\textbf{Assessment\_techniqueHM} & 35.6 & 13.06 & 2.73 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 50.09 & 14.04 & 3.57 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 27.27 & 47.04 & 0.58 & 0.56 \\
-\textbf{Assessment\_techniqueQPW} & 35.11 & 22.51 & 1.56 & 0.12 \\
+HM} & 36.02 & 45.37 & 0.79 & 0.43 \\
+\textbf{Assessment\_techniqueQPW} & 38.76 & 21.66 & 1.79 & 0.07 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 35.56 & 26.84 & 1.33 & 0.19 \\
-\textbf{Assessment\_techniqueSH} & 41.11 & 13.23 & 3.11 & 0 \\
-\textbf{Forest\_typeCF} & -4.51 & 8.34 & -0.54 & 0.59 \\
-\textbf{Forest\_typeMF} & -6.26 & 8.47 & -0.74 & 0.46 \\
-\textbf{Hydrological\_regimeSD} & 5.88 & 10.16 & 0.58 & 0.56 \\
+EA} & 48.73 & 26.18 & 1.86 & 0.06 \\
+\textbf{Assessment\_techniqueSH} & 45.52 & 12.82 & 3.55 & 0 \\
+\textbf{Forest\_typeCF} & -4.81 & 7.85 & -0.61 & 0.54 \\
+\textbf{Forest\_typeMF} & -3.75 & 8.27 & -0.45 & 0.65 \\
+\textbf{Hydrological\_regimeSD} & 8.89 & 9.63 & 0.92 & 0.36 \\
 \bottomrule
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.69 & 4 & 0.33 & 0.15 \\
-\textbf{s(Dryness)} & 4.27 & 9 & 1.45 & 0.01 \\
-\textbf{s(length)} & 17.14 & 34 & 1.05 & 0 \\
+\textbf{s(log10(Area\_km2))} & 3.43 & 4 & 1.66 & 0.1 \\
+\textbf{s(Dryness)} & 3.66 & 9 & 1.42 & 0.01 \\
+\textbf{s(length)} & 19.84 & 34 & 2.15 & 0 \\
 \bottomrule
@@
 
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.34.
+This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.15, likely due to the high variance in the data.
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.1, likely due to the high variance in the data.
 It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
@@
 \endhead
-\textbf{s(Dryness)} & 2.9 & 9 & 2.8 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.71 & 4 & 0.59 & 0.06 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.91 \\
+\textbf{s(Dryness)} & 2.39 & 9 & 1.68 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.7 & 4 & 0.55 & 0.07 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.87 \\
 \bottomrule
@@
 The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4 and \emph{length} \textgreater{} 60 years.
-This result in a reduction of the data set from 327 to 304 catchments.
-
-This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.29.
+This result in a reduction of the data set from 350 to 327 catchments.
+
+This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.28.
 * The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.06, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.07, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
 The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.81) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.33 with a slope of -9.69) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
@@
 \endhead
-PWE & 183 \\
-HM & 57 \\
+PWE & 204 \\
+HM & 59 \\
 SH & 42 \\
@@
 \endhead
-\textbf{(Intercept)} & -18.96 & 18.8 & -1.01 & 0.31 \\
-\textbf{DeltaF\_perc\_pos} & 0.28 & 0.1 & 2.83 & 0 \\
-\textbf{Forest\_SignIncrease} & -25.46 & 6.8 & -3.75 & 0 \\
-\textbf{Precip\_data\_typeOB} & -9.75 & 13.25 & -0.74 & 0.46 \\
-\textbf{Precip\_data\_typeSG} & 19.37 & 15.64 & 1.24 & 0.22 \\
-\textbf{Assessment\_techniqueHM} & 31.9 & 12.97 & 2.46 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 44.99 & 13.59 & 3.31 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 40.11 & 21.12 & 1.9 & 0.06 \\
-\textbf{Assessment\_techniqueSH} & 46.08 & 12.43 & 3.71 & 0 \\
-\textbf{Forest\_typeCF} & -3.42 & 7.94 & -0.43 & 0.67 \\
-\textbf{Forest\_typeMF} & -3.02 & 8.08 & -0.37 & 0.71 \\
-\textbf{Hydrological\_regimeSD} & 2.46 & 9.87 & 0.25 & 0.8 \\
+\textbf{(Intercept)} & -23.29 & 17.91 & -1.3 & 0.19 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.09 & 3.08 & 0 \\
+\textbf{Forest\_SignIncrease} & -21.92 & 6.47 & -3.39 & 0 \\
+\textbf{Precip\_data\_typeOB} & -10.35 & 13.18 & -0.79 & 0.43 \\
+\textbf{Precip\_data\_typeSG} & 18.84 & 15.62 & 1.21 & 0.23 \\
+\textbf{Assessment\_techniqueHM} & 35.63 & 11.98 & 2.98 & 0 \\
+\textbf{Assessment\_techniquePWE} & 48.19 & 12.44 & 3.87 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 42.14 & 20.57 & 2.05 & 0.04 \\
+\textbf{Assessment\_techniqueSH} & 46.49 & 12.29 & 3.78 & 0 \\
+\textbf{Forest\_typeCF} & -3.33 & 7.54 & -0.44 & 0.66 \\
+\textbf{Forest\_typeMF} & -2.73 & 7.97 & -0.34 & 0.73 \\
+\textbf{Hydrological\_regimeSD} & 4.63 & 9.38 & 0.49 & 0.62 \\
 \bottomrule
@@
 \endhead
-\textbf{s(Dryness)} & 2.87 & 9 & 2.77 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.7 & 9 & 0.26 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.93 \\
+\textbf{s(Dryness)} & 2.38 & 9 & 1.64 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.68 & 9 & 0.24 & 0.07 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.85 \\
 \bottomrule

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.Rmd
update link to github repos

@@
 
-All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water) on the "publish" branch.
+All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
 

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.tex
update link to github repos

@@
 
-All the final data and analysis for this paper are located on github: \url{https://github.com/WillemVervoort/Forest_and_water} on the ``publish'' branch.
+All the final data and analysis for this paper are located on github: \href{https://github.com/WillemVervoort/Forest_and_water/tree/publish}{https://github.com/WillemVervoort/Forest\_and\_water} on the ``publish'' branch.
 

### 2022-03-30 paper/Forest_and_Water/Abstract.Rmd
update funding

--- 2022-03-30 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
 ---
-title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
-author:
-  - name: R. Willem Vervoort
-    email: willem.vervoort@sydney.edu.au
-    affiliation: DARE
-    footnote: 1
-  - name: Eliana Nervi
-    email: eliananervi@gmail.com
-    affiliation: INIA 
-  - name: Jimena Alonso
-    email: jalonso@fing.edu.uy
-    affiliation: IMFIA
-address:
-  - code: DARE
-    address: ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.
-  - code: The University of Sydney
-    address: The University of Sydney, Sydney, NSW 2006, Australia
-  - code: INIA
-    address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
-  - code: IMFIA
-    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
-footnote:
-  - code: 1
-    text: "Corresponding Author"
-abstract: |
+title: "Abstract"
+output: pdf_document
+---
+
   Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
@@
 Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
-journal: "Journal of Hydrology"
-date: "`r Sys.Date()`"
-bibliography: forest_and_water.bib
-linenumbers: true
-#numbersections: true
-csl: elsevier-harvard.csl
-output:   
-  bookdown::pdf_book:
-    base_format: rticles::elsevier_article
-editor_options: 
-  chunk_output_type: inline
----
-
-Introduction
-==========================
-
-There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017].
-The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
-However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-
-<!-- Include Filoso paper -->
-For the purpose of this paper, *watershed* and *catchment* are interchangeable terms.
-Many of the US studies use *watershed*, while European and Australian studies use *catchment*.
-In particular, we retained the term "paired watershed studies" and "quasi-paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.
-
-Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
-These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover.
-The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
-In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
-In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
-There is some overlap between the two data sets, but there are also some studies unique to both sets.
-
-The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^.
-While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].
-
-The second study [@filoso2017] is a systematic review of reforestation studies (only studies in which forest cover increased).
-This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
-Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with @zhang2017), but they did not identify a distinct threshold and fitted a log-linear relationship.
-In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
-
-A final summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
-However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
-In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model.
-A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
-
-Encouraged by the work from @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
-In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.
-
-In particular, the main method in the work by @zhang2017 is a single covariate linear regression, and in @filoso2017 the focus is mainly on classification and there is again some single covariate linear regression.
-As @zhang2017 points out, a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
-However, the subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum.
-And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.
-
-As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further catchments (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
-
-Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
-@jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
-
-However, there could be further confounding factors, which are eluded to by @filoso2017:
-
-* the type of analysis, i.e. paired watershed studies, modelling, time series analysis etc.
-
-* the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
-
-Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
-
-Methods
-==========
-
-## The original data sets
-
-The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
-
-Table 1  Summary of abbreviations of factors used in the @zhang2017 data set
-
-|Factor | Abbreviation | Definition |
-|:------------|:-------|:----------|
-| forest type | CF |coniferous forest | 
-|  | BF | broadleaf forest |
-|  | MF |mixed forest  |
-|hydrological regime|RD|rain dominated|
-| |SD|snow dominated|
-| climate type| EL| energy limited|
-| |WL| water limited|
-| |EQ| equitant|
-|precipitation data type|OB|observed|
-| |SG|spatial gridded|
-| |MD|modelled|
-|assessment technique|PWE|paired watershed experiment|
-| |QPW|quasi-paired watershed experiment|
-| |HM|hydrological modelling|
-| |EA|elastictity analysis|
-| |SH|combined use of statistical methods|
-| | |and hydrographs|
-
-While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
-In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET.
-In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).
-In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
-We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.
-
-## Additional data collection
-
-To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from @filoso2017.
-In particular, we focussed on including the 37 data points related to the quantitative analysis in @filoso2017.
-
-In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location.
-These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps^TM^.
-In the dataset, an additional column has been added to indicate the source of the location data.
-
-Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
-Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
-For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
-
-Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
-
-\begin{equation}
-D = \frac{E_{0}}{Pa} (\#eq:eq1)
-\end{equation}
-
-The length of the study can be a variable influencing the change in flow [e.g.
-@jackson2005; @filoso2017], as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a "Kuczera curve" [@kuczera1987].
-It is not clear if this is an effect of increased water use in growth [@vertessy2001] or due to changes in interception [@stoof2012].
-Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017.
-The length of the study was already included in the data from @filoso2017, but these were checked against the original publications.
-
-Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
-Table 6 in @zhang2011).
-We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017.
-This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
-Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
-
-The final column in the improved data set is a "notes" column, which we added, but is not further used in the analysis.
-It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
-This will allow future research to further sructinise our input for errors.
-
-All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water) on the "publish" branch.
-
-## Statistical modelling
-```{r, include=FALSE}
-options(tinytex.verbose = FALSE)
-```
-<!----
-The starting point consists of the two tables in the supplementary data from Zhang et al.
-(2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2).
-We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
---->
-```{r, echo = F, message = F, warning=FALSE}
-require(tidyverse)
-require(lubridate)
-require(mgcv)
-require(visreg)
-require(patchwork)
-suppressWarnings(require(knitr))
-suppressWarnings(require(bookdown))
-suppressWarnings(require(pander))
-
-Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv")
-Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv")
-new_data <- read_csv("../../data/NewCatchments_T3.csv")
-```
-
-<!--- ### change names of columns
-the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple --->
-
-```{r, echo = F}
-names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
-names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
-names(new_data)[3:4] <- c("Area_km2", "Pa_mm")
-names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
-names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
-names(new_data)[5:6] <- c("Forest_type", "Hydrological_regime")
-names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
-names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
-names(new_data)[9:10] <- c("Precip_data_type", "Assessment_technique")
-
-```
-
-<!-- combine both tables -->
-
-```{r, echo = F, warning = F, message = F}
-Zhang_small$From <- as.numeric(Zhang_small$From)
-Zhang_small$To <- as.numeric(Zhang_small$To)
-Zhang_all <- bind_rows(Zhang_large,Zhang_small) %>%
-  mutate(dataset = "original Zhang et al data")
-new_data <- new_data %>%
-  mutate(dataset = "new data")
-All_data <- bind_rows(Zhang_all, new_data)
-```
-
-<!--filter duplicates -->
-```{r, echo = F}
-All_data <- All_data %>%
-  mutate(`Possible duplicate` = 
-           ifelse(is.na(`Possible duplicate`)==T,0,`Possible duplicate`),
-         `Possible duplicate` = as.numeric(`Possible duplicate`)) %>%
-  filter(`Possible duplicate` != 1)
-```
-
-<!--now combine with E0 values -->
-```{r, echo = F}
-# calculate dryness index
-All_data <- All_data %>%
-  mutate(Dryness = E0/Pa_mm)
-```
-
-<!--remove the Amazon from the analysis -->
-```{r, echo = F}
-All_data <- All_data %>%
-  filter(`Watershed #` != 1)
-```
-
-<!-- correct watershed #76 (Baker1984) to be 157 for DeltaQf_perc -->
-
-```{r,echo = F}
-All_data <- All_data %>%
-  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
-```
-
-To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
-
-The general model tested is:
-
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
-& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
-\end{align}
-
-Here $X_i$ are factorial variables, while $Z_i$ are continuous variables.
-The model intially assumes no direct interactions and all variables are additive.
-We will comment in this assumption in the discussion.
-The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data was converted to positive values, and an additional factorial column ($sign_{forest cover}$) is included indicating whether it was a forest cover increase or decrease.
-
-A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with $\Delta Qf \%$.
-This means that a smooth function $s()$ can be applied to the $Z_i$ variables.
-For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty.
-The result of this approach is that for high enough smoothing parameters (i.e.
-if the data is very "wiggly") the smooth term can be shrunk to 0 and thus will be no longer significant [@wood2006].
-This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship.
-In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.
-
-The over arching test focuses on  the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
-Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.
-
-As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results).
-Subsequently the analysis was repeated and the additionally identified catchments were added.
-
-Results
-===========
-
-<!-- analysis -->
-## Description of the data
-
-The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
-The current number of catchments is the result of the removal of duplicates and our modifications and additions.
-The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*.
-The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
-Note also the large number of studies with 100\% forest cover reduction.
-Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
-
-```{r, echo = F, warning = F, message=F, eval =F}
-require(latex2exp)
-png("DataExploration.png", width = 2500, height = 1800, res = 300)
-All_data %>%
-  mutate(log10Area = log10(Area_km2)) %>%
-  pivot_longer(c(DeltaF_perc, DeltaQf_perc, log10Area, Pa_mm, E0), 
-               names_to = "variable", values_to = "value") %>%
-  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
-  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
-    c("log10Area" =  "log10(Area km2)", #latex2exp::TeX("$\\log_{10}({Area~km^2})$"),
-      "DeltaF_perc" = "Delta Forest (%)",
-      "DeltaQf_perc" = "Delta Qf (%)",
-      "E0" = "Annual Potential E (mm)",
-      "Pa_mm" = "Annual Rainfall (mm)"),default = label_parsed)) +   
-  theme(axis.title = element_text(size = rel(1.5)),
-        axis.text = element_text(size = rel(1.2)),
-        strip.text = element_text(size = rel(1.5)),
-        axis.text.x = element_text(angle = 45, hjust = 1))
-dev.off()
-```
-
-(ref:datagraphscaption) Overview of the distribution of the data set for five of the included variables.
-Note that the first panel (showing the distribution of the catchment areas) indicates the distribution of the *log_10_* transformed Area (in km^2^).
-
-```{r datagraphs, echo=F, out.width="90%", fig.cap="(ref:datagraphscaption)"}
-
-include_graphics("./DataExploration.png")
-```
-
-```{r, echo = F}
-All_data2 <- All_data %>%
-  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
-                              "Decrease", "Increase"),
-         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
-                                  -1*DeltaF_perc,
-                                  DeltaF_perc))
-```
-
-```{r, eval = F, echo = F}
-All_data2 %>%
-  filter(is.na(Forest_Sign)==T)
-```
-
-Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
-
-```{r, echo = F, warning = F, message=F, eval =F}
-# plot to show the difference
-png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  filter(is.na(Forest_Sign)==F) %>%
-  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
-  geom_point() +  
-  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
-  theme_bw() + 
-    theme(axis.title = element_text(size = rel(1.5)),
-        axis.text = element_text(size = rel(1.2)),
-        legend.text = element_text(size = rel(1.2)),
-        legend.title = element_text(size = rel(1.2)),
-        strip.text = element_text(size = rel(1.5))) + 
-  scale_colour_manual(name = "Forest cover direction", 
-                      values = c("Decrease" = "red","Increase" = "blue")) +
-  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
-dev.off()
-
-```
-
-```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of decreases (top) and increases (bottom) in forest cover", out.width="90%"}
-include_graphics("Increase_decrease.png")
-```
-
-```{r, eval = F, echo = F}
-linear_model_E0 <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign + E0, 
-                           data = All_data2)
-#summary(linear_model_E0)
-#gam.check(linear_model_simple)
-
-```
-
-## The general relationship between change in forest cover and streamflow
-
-Following @zhang2017, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
-
-\begin{equation}
-\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon (\#eq:eq3)
-\end{equation}
-
-```{r model1, echo = F}
-linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign, 
-                           data = All_data2 %>% filter(dataset == "original Zhang et al data"))
-#summary(linear_model_simple)
-#gam.check(linear_model_simple)
-
-```
-
-```{r model1b, echo = F}
-linear_model_simple_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign, 
-                           data = All_data2)
-#summary(linear_model_simple)
-r_sq <- round(summary(linear_model_simple_b)$r.sq,2)
-```
-
-```{r, echo =F}
-pander(round(rbind(summary(linear_model_simple)$p.table,
-                       summary(linear_model_simple_b)$p.table),2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
-The first three rows relate to the model using the original data base from Zhang et al.
-(2017).
-The bottom three rows are the results of the model including the new data.
-Clearly there is no major change arising from the additional data.")
-```
-
-The overall variance explained in this model (equation \@ref(eq:eq3)) is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
-The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
-However the change in flow is different for forest cover decreases compared to forest cover increases.
-In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase.
-So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.
-
-```{r, eval = F, echo = F}
-png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
-par(mfrow=c(2,2))
-gam.check(linear_model_simple)
-par(mfrow=c(1,1))
-dev.off()
-```
-
-```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
-include_graphics("residual_plot_model1.png")
-```
-
-Of importance here is to highlight the residuals of this regression (equation \@ref(eq:eq3) and Figure \@ref(fig:gamcheck)).
-These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)).
-In other words, the distribution of the residuals is somewhat fat-tailed.
-We will discuss this later.
-
-Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
-Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total @zhang2017 data set.
-But this also means that our re-analysis of the data can be directly compared to the original study.
-
-However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.
-
-## The effect of location on the globe
-<!-- include Lat and Long (later E0) -->
-
-```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
-
-include_graphics("../../data/FAOET0data_final_2022.png")
-```
-
-Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients.
-As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
-Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon (\#eq:eq4)
-\end{align}
-
-```{r model3, echo = F}
-model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Latitude + Longitude, data = All_data2)
-#summary(model3_includeLatLong)
-#gam.check(model3_includeLatLong)
-```
-
-```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model based on the complete dataset and including Latitude and Longitude")
-```
-
-There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.
-
-## Impact of climate
-
-While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
-Potential evapotranspiration (*E0*) by itself was  not significant in the.
-Initially, we also tested models using only the annual average precipitation (*Pa (mm)*), but  interactions between precipitation and evapotranspiraton might be captured by the dryness index.
-Both dryness index  and *Pa (mm)* were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
-As a result only the dryness index was retained as a climate indicator to align with the earlier work by @zhang2017.
-Given that Latitude and Longitude were not significant, we dropped these from the model.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness  + \varepsilon (\#eq:eq5)
-\end{align}
-
-<!--\newpage-->
-
-<!-- include Dryness index -->
-
-```{r model4, echo = F}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness, data = All_data2)
-#summary(model4_dryness)
-#gam.check(model4_dryness)
-```
-
-```{r, echo =F}
-pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
-```
-
-<!-- running PCA, this is used later -->
-
-```{r testing_PCA, eval = F, echo = F, warning = F}
-require(FactoMineR)
-require(factoextra)
- res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length, Latitude, Longitude), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
- # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
- #  scree.plot
- #  
- #   var.plot <- fviz_pca_var(res.pca, axes = c(1, 2), col.var = "black")
- #  var.plot
- png("PCA_biplot_Alldata.png", width = 800, height = 800)
-  biplot_test <- fviz_pca_biplot(res.pca, axes = c(1, 2), repel = TRUE,
-                                                      col.var = "darkblue", # Variables color
-                            label="var", #"var" 
-                            col.ind = "red", # Individuals color
-                            title = "",labelsize = 8, pointsize = 4, arrowsize = 1) +
-        theme(text = element_text(size = 14, face = "bold"),
-              axis.title = element_text(size = rel(2), face = "bold"),
-        axis.text = element_text(size = rel(2), face = "bold"))
- print(biplot_test)
- dev.off()
-
-```
-
-Similar to *E0* or *Pa_mm*, the results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
-This might seem suprising in light of earlier reported results [@zhang2017; @filoso2017].
-In this case, the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
-However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
-As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
-
-```{r, echo = F}
-pander(All_data2 %>% filter(Dryness > 4) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
-         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
-```
-
-There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
-These catchments are listed in Table \@ref(tab:drytable).
-
-## Is there a distinct effect of area?
-
-The second major variable is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017.
-Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon (\#eq:eq6)
-\end{align}
-
-```{r modelArea, echo = F}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-model_Area <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness + log10(Area_km2), data = All_data2)
-#summary(model_Area)
-#gam.check(model_Area)
-```
-
-```{r, echo =F}
-pander(round(summary(model_Area)$p.table,2), caption = "(\\#tab:out-modelArea) Results of the model including Area and the dryness index")
-```
-
-```{r modelArea_b, echo = F, eval =T}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-
-model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness*log10(Area_km2), data = All_data2)
-#summary(model_Area_b)
-#gam.check(model_Area)
-```
-
-The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km^2^)) (Table \@ref(tab:out-modelArea)).
-In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by `r abs(round(summary(model_Area)$p.table[5,1],1))`\%.
-Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km^2^)) in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.
-
-```{r, echo =F}
-pander(round(summary(model_Area_b)$p.table,2), caption = "(\\#tab:out-modelArea-int) Results of the model including an interaction between Area and the dryness index")
-```
-
-```{r, eval = F, echo = F}
-All_data2 %>%
-  ggplot(aes(log10(Area_km2),Dryness, size = DeltaF_perc_pos)) + geom_point(colour = "blue", alpha =0.5)
-summary(lm(Dryness~log10(Area_km2),data = All_data2))
-```
-
-## Are some of the variables possibly non-linear?
-
-The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response.
-This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005).
-In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (*Dryness* and *Area*) were considered non-linear.
-As a result a shrinkage smoothing spline [@wood2006] was applied to these variables.
-
-<!-- *From* represents the starting date of the data collection.
-The idea is that this last variable gives some indication of the "age" of the experiment.
-Improvements in data collection could introduce a positive or negative bias between more recent and older studies -->
-
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon (\#eq:eq7)
-\end{align}
-
-```{r model5, echo = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From,
-         mid_year = From + (To - From)/2)
-
-model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), k = 6, bs="ts") + 
-                    s(Dryness, k = 10, bs="ts") + 
-                    s(length, k = 50, bs="ts"), #+ 
-                    #s(From, k = 30, bs = "ts") ,
-                  data = All_data2)
-# summary(model5_all)
-# gam.check(model5_all)
-# plot(model5_all)
-```
-
-```{r , echo = F}
-pander(round(summary(model5_all)$p.table,2), caption = "(\\#tab:mfive-linear) Statistical summary for the linear terms in the model with non-linear terms")
-```
-
-```{r , echo = F}
-pander(round(summary(model5_all)$s.table,2), caption = "(\\#tab:mfive-smooth) Statistical summary for the smooth terms in the model with non-linear terms")
-
-```
-
-Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
-For example, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(model5_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(model5_all)$s.table[2,4],2)`)) and *length* (p = `r round(summary(model5_all)$s.table[3,4],2)`)) explain significant variation in the data.
-
-However, including the non-linearity also increases the chance of over fitting, as the smoothing splines allow significant flexibility.
-Including interactions between the smooth variables is also possible, but the results are difficult to interpret given the high flexibility of the two-dimensional smooth.
-Given the overall variability in the data we did not attempt this.
-
-Finally the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included i.e. Equation \@ref(eq:eq2).
-
-```{r model6_all, echo = F}
-
-model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), k = 5, bs="ts") + 
-                    s(Dryness, k = 10, bs="ts" ) + 
-                     s(length, k = 35, bs="ts") +# s(From, k = 40, bs = "ts") +
-                    Precip_data_type +  Assessment_technique +
-                    Forest_type +
-                    Hydrological_regime
-                    , data = All_data2)
-#summary(model6_all)
-#gam.check(model6_all)
-#plot(model6_all)
-```
-
-```{r, echo = F}
-pander(round(summary(model6_all)$p.table,2), caption = "(\\#tab:msix-linear) Statistical summary for the linear terms the full model")
-#plot(model6_all)
-```
-
-```{r , echo = F}
-pander(round(summary(model6_all)$s.table,2), caption = "(\\#tab:msix-smooth) Statistical summary for the smooth terms for the full model")
-#plot(model6_all)
-```
-
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
-This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
-However, it is interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant.
-In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
-In this case, *log10(Area (km^2^))* is no longer a significant predictor, the reasons for this will be discussed later.
-
-```{r visualisesmooths,echo =F, eval = F}
-#require(visreg)
-png("model6_smooths.png", width = 2500, height = 1800, res = 300)
-#par(mfrow=c(2,2))
-D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
-A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
-#From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
-(D + A) / L 
-# visreg(model6_all,"Dryness",gg=T) + theme_bw()
-# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
-# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
-#par(mfrow=c(1,1))
-dev.off()
-```
-
-```{r smoothsmodel6, fig.cap="Visualisation of the smooth variables in the model.
-The shaded areas are the 95% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationshiop.
-The data are plotted as individual points", out.width="90%", echo=F}
-
-include_graphics("model6_smooths.png")
-```
-
-Figure \@ref(fig:smoothsmodel6) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear, but, given all the data, not significant at p = `r round(summary(model6_all)$s.table[1,4],2)`, likely due to the high variance in the data.
-It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
-Both the *length* and *Dryness* variables are significant and show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
-For example, *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the data.
-
-<!--filter length < 60 -->
-```{r model7_nolongLength, echo = F}
-
-NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
-
-model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, k = 10, bs="ts" ) + 
-                #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), k = 5, bs = "ts") +
-                #length +
-                s(length, k = 10, bs = "ts") + #s(From, k=40, bs = "ts") +
-                    Precip_data_type +  Assessment_technique +
-                Forest_type +
-                    Hydrological_regime, data = NoLong_Data)
-# summary(model7)
-# gam.check(model7)
-# plot(model7)
-```
-
-```{r, echo = F}
-pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness < 4.")
-```
-
-```{r visualisesmooths7,echo =F, eval = F}
-#require(visreg)
-png("model7_smooths.png", width = 2500, height = 1800, res = 300)
-#par(mfrow=c(2,2))
-D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
-A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
-#From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
-(D + A) / L#(L + From) 
-# visreg(model6_all,"Dryness",gg=T) + theme_bw()
-# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
-# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
-#par(mfrow=c(1,1))
-dev.off()
-```
-
-```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}
-
-include_graphics("model7_smooths.png")
-```
-
-<!--\newpage-->
-
-The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) does not reflect a similar response to @filoso2017 and @jackson2005.
-One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Therefore it can be important to investigate what removing these few data points has on the overall model and the significance of the variables.
-The next model therefore removes the following data: *Dryness* > 4 and *length* > 60 years.
-This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
-
-This last model has more explaining power with an adjusted *r^2^* of `r round(summary(model7)$r.sq,2)`.
-* The results indicate that *Dryness* indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodel7)).
-Catchment area (*log10(Area (km^2^))*) also shows reasonable evidence of having an impact on flow with p = `r round(summary(model7)$s.table[2,4],2)`, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
-The variable *length* no longer is significant, after removal of the two studies with very long lengths.
-
-```{r, echo = F, eval =F}
-png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
-             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
-  theme_bw() + ylab("% change in flow") +
-  theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5))) +
-  xlab("% change in forest cover") + #scale_y_log10() +
-  scale_size_continuous(name = "log10(Area in km2)") +
-  scale_colour_discrete(name = "Assessment Technique") +
-  scale_shape_discrete(name = "Forest cover direction")
-dev.off()
-```
-
-```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
-
-include_graphics("flow_forest_byArea.png")
-
-```
-
-Discussion
-======================
-
-## Catchment size
-
-Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), however, in contrast to @zhang2017, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \@ref(fig:overview)).
-In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
-
-An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change [@navas2019].
-Therefore, specifically if the  forest cover change is small relative to the catchment size, the effect of this change will be buffered.
-
-```{r, eval =T, echo =F}
-# All_data2 %>%
-#   ggplot(aes(log10(Area_km2), DeltaF_perc_pos)) + 
-#   geom_point(alpha = 0.5) + theme_bw()
-
-Area_deltaF <- lm(DeltaF_perc_pos~log10(Area_km2), data = All_data2)
-```
-
-There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average, the forest cover change data is approximately  10\% lower.
-This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to "implement" in an experiment.
-
-This is also reflected in the second caveat.
-Most of the data from the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
-For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
-This is a confounding factor that is not easily addressed in the regression modelling attempted here.
-Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
-
-In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
-
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamdlow.
-
-## Model residuals
-
-As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
-These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \@ref(fig:overview)), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \@ref(fig:gamcheck7)).
-
-```{r, eval = F, echo = F}
-png("residual_plot_model7.png", width = 2500, height = 1800, res = 300)
-par(mfrow=c(2,2))
-gam.check(model7)
-par(mfrow=c(1,1))
-dev.off()
-```
-```{r gamcheck7, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
-include_graphics("residual_plot_model7.png")
-```
-
-The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
-Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
-In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
-This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \@ref(fig:increasedecrease)).
-
-## The effect of assessment techniques with very small numbers of observations
-
-```{r, eval =T, echo =F}
-Table_assessment <- All_data2 %>%
-  group_by(Assessment_technique) %>%
-  summarise(n = n()) %>%
-  arrange(desc(n))
-```
-```{r, echo=F}
-pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
-```
-
-```{r, eval = F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw() +
-    theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)))
-dev.off()
-```
-
-```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =T}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
-
-One concern with the result presented heer, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
-
-```{r model8_reduc_assess, echo = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), k = 10, bs = "ts") + s(length, bs = "ts") + #s(From, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4))
-#plot(model8_assess)
-# summary(model8_assess)
-# gam.check(model8_assess)
-```
-```{r, echo = F}
-pander(round(summary(model8_assess)$p.table,2), caption= "(\\#tab:model8-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
-pander(round(summary(model8_assess)$s.table,2), caption = "(\\#tab:model8-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
-
-```
-
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:model8-linear) and \@ref(tab:model8-smooth)).
-It strengthens the significance of the different assessment techniques and *Dryness* but generally results in the same interpretation.
-Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
-
-However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
-Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
-In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies in the database are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.
-
-## The effect of climate
-
-In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies [@zhang2017; @zhou2015; @filoso2017].
-This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
-As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
-This points to a need for future studies that unravel this interaction.
-
-## Interactions
-
-Generally this study did not consider interactions, but the above discussion suggest that there are possible several interactions.
-The relationships between forest cover change and *Area (km^2^)* and between *Area (km^2^)* and assessment technique have already been highlighted.
-However there are further unexplored interactions between the study length and some of the variables.
-
-A principle component analysis of the numeric data reveals some of these interactions (Figure \@ref(fig:pcaplot)), such as between *length* and *Dryness*.
-Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
-
-```{r pcaplot, fig.cap="Biplot of the first two principle components using a principle component analysis on the numerical values of the data set", out.width="90%", echo=F, eval =T}
-
-include_graphics("PCA_biplot_Alldata.png")
-```
-
-## Further considerations
-
-In contrast to @filoso2017, we did not identify that the length of the observation period is a significant variable in our final model.
-However, there are further confounding factors in the data, which were not analysed in this study.
-These were also classified by @filoso2017 and these factors might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
-This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
-
-Large databases based on historical studies, such as used here, also have significant uncertainty.
-While we have reviewed a large number of the studies in more detail, we have generally assumed that the assessments of past authors of the changes in streamflow and changes in forest cover are correct.
-More generally a lot of the data in the database are "summary data" extracted from the paper and this often neglects a lot of possibly important detail in the original studies.
-This introduces additional uncertainty in the analysis.
-
-By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
-
-Conclusions
-============
-
-More rigorous checking of an existing database on catchment studies relating to changes in forest cover to changes in flow and more detailed statistical analysis results in both agreement and disagreement with older studies.
-It demonstrates that analysis of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often fail to capture the complexity in the data.
-The variablility in the aggregated historical data is simply too large.
-
-As with any analysis, the results of the statistical analysis in this paper need to be considered "conditional on the data". 
-Conditional on the data, it can be determined that the impact of forestry on streamflow:
-
-- is greater for forest clearing then for reforestation;
-
-- is reduced for larger watersheds;
-
-- Increases for drier watersheds; and
-
-- is sensitive to the assessment method used in the historical data.
-
-Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area, especially in relation to larger catchments.
-Furthermore, the current study analyses a large global dataset of aggregated data.
-This analysis does not exclude more local and regional effects that cannot be identified in the global data.
-In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
-```
-
-<!-- testing whether dryness is related to Area -->
-
-```{r testing_dryness, echo = F, eval= F}
-All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
-```
-
-<!--Check the size distribution of the watersheds -->
-
-```{r, eval =F, echo =F}
-Zhang_all %>%
-  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
-  scale_x_log10()
-total <- nrow(Zhang_all)
-length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
-```
-```{r, eval =F, echo = F}
-Zhang_all2 %>%
-  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
-```
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(log10(Area_km2), DeltaQf_perc, colour = DeltaF_perc)) + 
-  geom_point(alpha = 0.5) + theme_bw()
-```
-
-```{r, eval =F, echo =F}
-All_data2 %>%
-  ggplot(aes(log10(Area_km2), Dryness)) + 
-  geom_point(colour = "blue") + theme_bw()
-```
-
-<!--- Analysis of Methods by dryness -->
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Assessment_technique, Dryness)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
-```
-
-<!--
-```{r model4b, echo = F, eval = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From)
-
-model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
-                 s(length, bs = "ts"),
-                  data = All_data2 %>% filter(length < 60))
-summary(model4b)
-
-```
--->
-```{r model5b, echo = F, eval = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From)
-
-model5_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    te(log10(Area_km2),Dryness, bs="ts") +  
-                  s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
-                    s(length, bs="ts") + s(From, bs = "ts") ,
-                  data = All_data2)
-summary(model5_b)
-plot(model5_b)
-```
-
-```{r model9_interact, echo = F, eval = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model9_interact <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
-                      ti(Dryness,From, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      ti(log10(Area_km2),length, bs="ts" ) +
-                     # log10(Area_km2):Assessment_technique +
-                      s(length, bs = "ts") + 
-                      s(From, bs = "ts") +
-                    Precip_data_type + Forest_type + Assessment_technique +
-                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4, From > 1920))
-#gam.check(model9_interact)
-# plot(model9_interact)
-summary(model9_interact)
-
-```
-
-References {#references .unnumbered}
-==========

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.Rmd
update funding

@@
   - name: Eliana Nervi
-    email: eliananervi@gmail.com
+    email: eliananervif@gmail.com
     affiliation: INIA 
@@
 
+Acknowledgements
+================
+This work was funded through project FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay.
+
 References {#references .unnumbered}

### 2022-03-30 paper/Forest_and_Water/Forest_and_Water.Rmd
double spacing

@@
   chunk_output_type: inline
+header-includes:
+  - \usepackage{setspace}
 ---
+
+\doublespacing
 

### 2022-04-01 paper/Forest_and_Water/Forest_and_Water.Rmd
fixed errors in pdf and added cover letter

@@
 
-All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
+All the final data and analysis for this paper are located on github:  
+[https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
 
@@
 
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
+This model (Tables \@ref(tab:msix-linear) and \@ref(tab:msix-smooth)) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-```{r smoothsmodel6, fig.cap="Visualisation of the smooth variables in the model.
-The shaded areas are the 95% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationshiop.
-The data are plotted as individual points", out.width="90%", echo=F}
-
+```{r smoothsmodel6, echo = F, fig.cap="Visualisation of the smooth variables in the model, the shaded areas are the 95\\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points", out.width="90%"}
 include_graphics("model6_smooths.png")
 ```
+
+\newpage
 
@@
 This last model has more explaining power with an adjusted *r^2^* of `r round(summary(model7)$r.sq,2)`.
-* The results indicate that *Dryness* indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodel7)).
+The results indicate that *Dryness* indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodel7) and Table \@ref(tab:restrictlength)).
 Catchment area (*log10(Area (km^2^))*) also shows reasonable evidence of having an impact on flow with p = `r round(summary(model7)$s.table[2,4],2)`, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
@@
 
-One concern with the result presented heer, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
 This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).

### 2022-04-01 paper/Forest_and_Water/Forest_and_Water.tex
update funding
double spacing
fixed errors in pdf and added cover letter

--- 2022-03-30 paper/Forest_and_Water/Forest_and_Water.tex
@@
 
+\usepackage{setspace}
+
 \begin{document}
@@
     \author[INIA]{Eliana Nervi}
-   \ead{eliananervi@gmail.com} 
+   \ead{eliananervif@gmail.com} 
     \author[IMFIA]{Jimena Alonso}
@@
  \end{frontmatter}
+
+\doublespacing
 
@@
 
-All the final data and analysis for this paper are located on github: \url{https://github.com/WillemVervoort/Forest_and_water} on the ``publish'' branch.
+All the final data and analysis for this paper are located on github:\\
+\href{https://github.com/WillemVervoort/Forest_and_water/tree/publish}{https://github.com/WillemVervoort/Forest\_and\_water} on the ``publish'' branch.
 
@@
 
-This model explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
+This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-\textbackslash begin\{figure\}
-\includegraphics[width=0.9\linewidth]{model6_smooths} \textbackslash caption\{Visualisation of the smooth variables in the model.
-The shaded areas are the 95\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationshiop.
-The data are plotted as individual points\}\label{fig:smoothsmodel6}
-\textbackslash end\{figure\}
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{model6_smooths} \caption{Visualisation of the smooth variables in the model, the shaded areas are the 95\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points}\label{fig:smoothsmodel6}
+\end{figure}
+
+\newpage
 
@@
 This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.28.
-* The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7}).
+The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7} and Table \ref{tab:restrictlength}).
 Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.07, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
@@
 
-One concern with the result presented heer, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
 This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
@@
 
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+This work was funded through project FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay.
+
 \hypertarget{references}{%

### 2022-07-02 paper/Forest_and_Water/Forest_and_Water.tex
changes since publication

@@
 
-The overall dataset contains 350 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+The overall dataset contains 327 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
 In contrast, the original dataset from Zhang et al.
-(2017) contained 340 catchments and the Filoso et al.
+(2017) contained 312 catchments and the Filoso et al.
 (2017) study used 37 catchments (Table S2 in Filoso et al.
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
 \caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
@@
 \endhead
-\textbf{(Intercept)} & 7.44 & 5.44 & 1.37 & 0.17 \\
-\textbf{DeltaF\_perc\_pos} & 0.48 & 0.08 & 5.66 & 0 \\
-\textbf{Forest\_SignIncrease} & -28.17 & 5.73 & -4.92 & 0 \\
-\textbf{(Intercept)} & 8.93 & 5.22 & 1.71 & 0.09 \\
-\textbf{DeltaF\_perc\_pos} & 0.45 & 0.08 & 5.63 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.81 & 5.21 & -6.88 & 0 \\
+\textbf{(Intercept)} & 6.24 & 5.19 & 1.2 & 0.23 \\
+\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.71 & 0 \\
+\textbf{Forest\_SignIncrease} & -26.95 & 5.4 & -4.99 & 0 \\
+\textbf{(Intercept)} & 7.79 & 5.01 & 1.56 & 0.12 \\
+\textbf{DeltaF\_perc\_pos} & 0.43 & 0.08 & 5.64 & 0 \\
+\textbf{Forest\_SignIncrease} & -34.65 & 4.95 & -7 & 0 \\
 \bottomrule
@@
 
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.22, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.24, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
 \caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 11.31 & 5.65 & 2 & 0.05 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.66 & 0 \\
-\textbf{Forest\_SignIncrease} & -38.1 & 5.49 & -6.95 & 0 \\
-\textbf{Latitude} & -0.1 & 0.09 & -1.05 & 0.3 \\
-\textbf{Longitude} & 0 & 0.03 & 0.14 & 0.89 \\
+\textbf{(Intercept)} & 9.88 & 5.42 & 1.82 & 0.07 \\
+\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.68 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.72 & 5.21 & -6.86 & 0 \\
+\textbf{Latitude} & -0.1 & 0.09 & -1.17 & 0.24 \\
+\textbf{Longitude} & -0.02 & 0.03 & -0.62 & 0.54 \\
 \bottomrule
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.23 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.24 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
 \caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 6.34 & 6.17 & 1.03 & 0.3 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.69 & 0 \\
-\textbf{Forest\_SignIncrease} & -36.27 & 5.22 & -6.95 & 0 \\
-\textbf{Dryness} & 1.87 & 2.6 & 0.72 & 0.47 \\
+\textbf{(Intercept)} & 4.2 & 5.86 & 0.72 & 0.47 \\
+\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.7 & 0 \\
+\textbf{Forest\_SignIncrease} & -35.37 & 4.97 & -7.12 & 0 \\
+\textbf{Dryness} & 2.76 & 2.5 & 1.11 & 0.27 \\
 \bottomrule
@@
 This might seem suprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.47).
+In this case, the evidence is highly doubtful (p = 0.27).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.23.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.24.
 As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.42}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4167}}@{}}
 \caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
@@
 295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
-295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
 356 & -25.75 & 28.23 & Queens river \\
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
 \caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 12.41 & 7 & 1.77 & 0.08 \\
-\textbf{DeltaF\_perc\_pos} & 0.36 & 0.1 & 3.76 & 0 \\
-\textbf{Forest\_SignIncrease} & -34.4 & 5.31 & -6.48 & 0 \\
-\textbf{Dryness} & 2.48 & 2.61 & 0.95 & 0.34 \\
-\textbf{log10(Area\_km2)} & -2.97 & 1.64 & -1.81 & 0.07 \\
+\textbf{(Intercept)} & 11.15 & 6.73 & 1.66 & 0.1 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.09 & 3.49 & 0 \\
+\textbf{Forest\_SignIncrease} & -33.52 & 5.02 & -6.67 & 0 \\
+\textbf{Dryness} & 3.52 & 2.51 & 1.4 & 0.16 \\
+\textbf{log10(Area\_km2)} & -3.23 & 1.56 & -2.07 & 0.04 \\
 \bottomrule
@@
 
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.07) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3\%.
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.04) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.2\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.92)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.77)) as an interaction with Dryness.
 
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.40}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4000}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1333}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}@{}}
 \caption{\label{tab:out-modelArea-int} Results of the model including an interaction between Area and the dryness index}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 9.15 & 7.18 & 1.27 & 0.2 \\
-\textbf{DeltaF\_perc\_pos} & 0.35 & 0.1 & 3.65 & 0 \\
-\textbf{Forest\_SignIncrease} & -32.09 & 5.42 & -5.92 & 0 \\
-\textbf{Dryness} & 5.15 & 2.96 & 1.74 & 0.08 \\
-\textbf{log10(Area\_km2)} & 0.24 & 2.35 & 0.1 & 0.92 \\
-\textbf{Dryness:log10(Area\_km2)} & -2.64 & 1.39 & -1.9 & 0.06 \\
+\textbf{(Intercept)} & 7.13 & 6.86 & 1.04 & 0.3 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
+\textbf{Forest\_SignIncrease} & -30.74 & 5.1 & -6.02 & 0 \\
+\textbf{Dryness} & 7.05 & 2.86 & 2.46 & 0.01 \\
+\textbf{log10(Area\_km2)} & 0.64 & 2.19 & 0.29 & 0.77 \\
+\textbf{Dryness:log10(Area\_km2)} & -3.23 & 1.29 & -2.5 & 0.01 \\
 \bottomrule
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
 \caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & 13.42 & 5.42 & 2.48 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.16 & 0 \\
-\textbf{Forest\_SignIncrease} & -29.28 & 5.82 & -5.03 & 0 \\
+\textbf{(Intercept)} & 12.09 & 5.31 & 2.28 & 0.02 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.27 & 0 \\
+\textbf{Forest\_SignIncrease} & -30.5 & 5.62 & -5.43 & 0 \\
 \bottomrule
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.11}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
 \caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 3.39 & 5 & 2.23 & 0.01 \\
-\textbf{s(Dryness)} & 3.74 & 9 & 0.77 & 0.09 \\
-\textbf{s(length)} & 27.88 & 49 & 1.7 & 0 \\
+\textbf{s(log10(Area\_km2))} & 1.95 & 5 & 1.43 & 0.02 \\
+\textbf{s(Dryness)} & 4.62 & 9 & 1.81 & 0 \\
+\textbf{s(length)} & 28.4 & 49 & 1.2 & 0 \\
 \bottomrule
@@
 Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.37 and deviance explained of 0.44, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.01)), \emph{Dryness} (p = 0.09)) and \emph{length} (p = 0)) explain significant variation in the data.
+For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
 
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.42}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
 \caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & -21.82 & 19.28 & -1.13 & 0.26 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 3.16 & 0 \\
-\textbf{Forest\_SignIncrease} & -19.45 & 6.78 & -2.87 & 0 \\
-\textbf{Precip\_data\_typeOB} & -13.49 & 14.09 & -0.96 & 0.34 \\
-\textbf{Precip\_data\_typeSG} & 15.05 & 18.39 & 0.82 & 0.41 \\
-\textbf{Assessment\_techniqueEA, HM} & 12.1 & 45.85 & 0.26 & 0.79 \\
-\textbf{Assessment\_techniqueHM} & 35.6 & 13.06 & 2.73 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 50.09 & 14.04 & 3.57 & 0 \\
+\textbf{(Intercept)} & -22.77 & 18.14 & -1.26 & 0.21 \\
+\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
+\textbf{Forest\_SignIncrease} & -21.42 & 6.58 & -3.25 & 0 \\
+\textbf{Precip\_data\_typeOB} & -11.41 & 13.41 & -0.85 & 0.4 \\
+\textbf{Precip\_data\_typeSG} & 3.29 & 16.54 & 0.2 & 0.84 \\
+\textbf{Assessment\_techniqueEA, HM} & 12.7 & 43.18 & 0.29 & 0.77 \\
+\textbf{Assessment\_techniqueHM} & 37.34 & 12.44 & 3 & 0 \\
+\textbf{Assessment\_techniquePWE} & 49.63 & 12.95 & 3.83 & 0 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 36.02 & 45.37 & 0.79 & 0.43 \\
-\textbf{Assessment\_techniqueQPW} & 38.76 & 21.66 & 1.79 & 0.07 \\
+HM} & 39.85 & 42.28 & 0.94 & 0.35 \\
+\textbf{Assessment\_techniqueQPW} & 46.12 & 20.21 & 2.28 & 0.02 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 48.73 & 26.18 & 1.86 & 0.06 \\
-\textbf{Assessment\_techniqueSH} & 45.52 & 12.82 & 3.55 & 0 \\
-\textbf{Forest\_typeCF} & -4.81 & 7.85 & -0.61 & 0.54 \\
-\textbf{Forest\_typeMF} & -3.75 & 8.27 & -0.45 & 0.65 \\
-\textbf{Hydrological\_regimeSD} & 8.89 & 9.63 & 0.92 & 0.36 \\
+EA} & 50.81 & 24.17 & 2.1 & 0.04 \\
+\textbf{Assessment\_techniqueSH} & 42.84 & 12.22 & 3.51 & 0 \\
+\textbf{Forest\_typeCF} & -11.16 & 7.54 & -1.48 & 0.14 \\
+\textbf{Forest\_typeMF} & -1.81 & 7.8 & -0.23 & 0.82 \\
+\textbf{Hydrological\_regimeSD} & 11.44 & 9.16 & 1.25 & 0.21 \\
 \bottomrule
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.11}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
 \caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 3.43 & 4 & 1.66 & 0.1 \\
-\textbf{s(Dryness)} & 3.66 & 9 & 1.42 & 0.01 \\
-\textbf{s(length)} & 19.84 & 34 & 2.15 & 0 \\
+\textbf{s(log10(Area\_km2))} & 1.66 & 4 & 0.74 & 0.14 \\
+\textbf{s(Dryness)} & 4.3 & 9 & 2.42 & 0 \\
+\textbf{s(length)} & 21.69 & 34 & 1.56 & 0 \\
 \bottomrule
@@
 
-This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
+This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.4.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.1, likely due to the high variance in the data.
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.14, likely due to the high variance in the data.
 It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
 \caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness \textless{} 4.}\tabularnewline
@@
 \endhead
-\textbf{s(Dryness)} & 2.39 & 9 & 1.68 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.7 & 4 & 0.55 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.87 \\
+\textbf{s(Dryness)} & 3.12 & 9 & 2.65 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.53 & 4 & 0.3 & 0.13 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.98 \\
 \bottomrule
@@
 The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4 and \emph{length} \textgreater{} 60 years.
-This result in a reduction of the data set from 350 to 327 catchments.
-
-This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.28.
+This result in a reduction of the data set from 327 to 306 catchments.
+
+This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.33.
 The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7} and Table \ref{tab:restrictlength}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.07, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.13, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
 The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.33 with a slope of -9.69) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.81) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.32}}
-  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.08}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3194}}
+  >{\centering\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.0833}}@{}}
 \caption{\label{tab:tableassess} Distribution of assessment techniques in the data set}\tabularnewline
@@
 \endhead
-PWE & 204 \\
-HM & 59 \\
+PWE & 183 \\
+HM & 57 \\
 SH & 42 \\
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.40}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.17}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.13}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4000}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1333}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}@{}}
 \caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
@@
 \endhead
-\textbf{(Intercept)} & -23.29 & 17.91 & -1.3 & 0.19 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.09 & 3.08 & 0 \\
-\textbf{Forest\_SignIncrease} & -21.92 & 6.47 & -3.39 & 0 \\
-\textbf{Precip\_data\_typeOB} & -10.35 & 13.18 & -0.79 & 0.43 \\
-\textbf{Precip\_data\_typeSG} & 18.84 & 15.62 & 1.21 & 0.23 \\
-\textbf{Assessment\_techniqueHM} & 35.63 & 11.98 & 2.98 & 0 \\
-\textbf{Assessment\_techniquePWE} & 48.19 & 12.44 & 3.87 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 42.14 & 20.57 & 2.05 & 0.04 \\
-\textbf{Assessment\_techniqueSH} & 46.49 & 12.29 & 3.78 & 0 \\
-\textbf{Forest\_typeCF} & -3.33 & 7.54 & -0.44 & 0.66 \\
-\textbf{Forest\_typeMF} & -2.73 & 7.97 & -0.34 & 0.73 \\
-\textbf{Hydrological\_regimeSD} & 4.63 & 9.38 & 0.49 & 0.62 \\
+\textbf{(Intercept)} & -25.35 & 16.81 & -1.51 & 0.13 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.09 & 3.25 & 0 \\
+\textbf{Forest\_SignIncrease} & -22.67 & 6.23 & -3.64 & 0 \\
+\textbf{Precip\_data\_typeOB} & -8.99 & 12.42 & -0.72 & 0.47 \\
+\textbf{Precip\_data\_typeSG} & 9 & 14.41 & 0.62 & 0.53 \\
+\textbf{Assessment\_techniqueHM} & 37.79 & 11.38 & 3.32 & 0 \\
+\textbf{Assessment\_techniquePWE} & 51.85 & 11.55 & 4.49 & 0 \\
+\textbf{Assessment\_techniqueQPW} & 43.98 & 19.18 & 2.29 & 0.02 \\
+\textbf{Assessment\_techniqueSH} & 42.26 & 11.51 & 3.67 & 0 \\
+\textbf{Forest\_typeCF} & -9.09 & 7.17 & -1.27 & 0.21 \\
+\textbf{Forest\_typeMF} & -1.9 & 7.47 & -0.25 & 0.8 \\
+\textbf{Hydrological\_regimeSD} & 5.54 & 8.88 & 0.62 & 0.53 \\
 \bottomrule
@@
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.35}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.10}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
 \caption{\label{tab:model8-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
@@
 \endhead
-\textbf{s(Dryness)} & 2.38 & 9 & 1.64 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.68 & 9 & 0.24 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.85 \\
+\textbf{s(Dryness)} & 3.11 & 9 & 2.59 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.51 & 9 & 0.13 & 0.13 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.96 \\
 \bottomrule
@@
 
-As with any analysis, the results of the statistical analysis in this paper need to be considered ``conditional on the data.''
+As with any analysis, the results of the statistical analysis in this paper need to be considered ``conditional on the data''.
 Conditional on the data, it can be determined that the impact of forestry on streamflow:
@@
 Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A., 2005.
-A review of paired catchment studies for determining changes in water yield resulting from alterations in vegetation.
+\href{http://www.sciencedirect.com/science/article/B6V6C-4G05MM9-1/2/bbc5fc0e958a8f34bcb7c1cc7fa57b48}{A review of paired catchment studies for determining changes in water yield resulting from alterations in vegetation}.
 Journal of Hydrology 310, 28--61.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-navas2019}{}}%
-Navas, R., Alonso, J., Gorgoglione, A., Vervoort, R.W., 2019. Identifying climate and human impact trends in streamflow: A case study in uruguay. Water 11, 1433.
+Navas, R., Alonso, J., Gorgoglione, A., Vervoort, R.W., 2019.
+\href{https://www.mdpi.com/2073-4441/11/7/1433}{Identifying climate and human impact trends in streamflow: A case study in uruguay}.
+Water 11, 1433.
 

### 2022-07-02 paper/Forest_and_Water/Reply_first_review_JoH.Rmd

---
title: "Replies to reviewer JoH"
author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
date: "`r Sys.Date()`"
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forest_and_water.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document records the replies to the reviewer for the first submission of "Factors determining how catchments respond to forest cover change. Re-analysing global data sets" to Journal of Hydrology which was rejected.

Reviewer comments in \textcolor{blue}{blue} responses in normal text

The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.

## Editor comments

\textcolor{blue}{The first is to streamline the statistics.  As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}

We politely disagree with this, as we outline in our reply to the AE. If the statistical modelling aimed at developing the best predictive model, then this would be the right approach. However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.

\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.
The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.  Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.  At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.
The final - and perhaps in truth the most problematic issue that may stand in the way of the MS being published - relates to novelty.  The AE highlights several recent works with similar questions, approaches and findings at global scales.  I could add to that list with more regional studies (e.g. Levy et al 2018).  So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}

\textcolor{blue}{These are substantial changes and go beyond a major revision.  For this reason, we're rejecting the MS at present.  If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}

\textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.  Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}

## Associate Editor:

\textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields. The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}

\textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
- As apparent from the diagnostic plots, the model assumptions may be violated in many cases. This can make the results of the fitting (and hence the manuscript conclusions) incorrect. I urge the authors to double check if this is indeed the case and consider ways to address the problem. It is also good practice to check the relevance of outliers (of data with high VIF) and set them aside before model fitting. It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}

Thank you for raising these important points in relation to the validity of the statistical model. 
A first point that arises from this is that we clarify better what the aim of our statistical modelling is.

There are in essence two "modes" in relation statistical modelling. Generally a model is developed to be used in predictive mode: using a model to predict unknown values. In this case the model should be reduced to its most efficient version that minimises the bias - variance trade-off. In this case, automatic variable selection and potentially validation on independent data are important.

However, a second reason for fitting a statistical model is purely to explain the variance in the data. In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data. This is subsequently followed by a step by step analysis to test the different causal relationship, either as single variables (as in @zhang2017) or jointly (as in our approach). In this case there is no attempt to find the best "predictive model", the focus is on the additional amount of explained variance from adding each variable.

**Include a section in the paper to explain this clearly**

Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript. Including such diagnostic plots for single variable regression is often ommitted. For example, both @zhang2017 and @filoso2017 do not present any diagnostics for their relationships and simply looking at their regressions qualitatively indicates issues with the residuals.

In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables.

Because our modelling is focused on explaining the maximum variance, excluding values that are outliers is probably not a good idea. However, identifying outliers (particularly values with high leverage), and investigating why these might be different from the overall distribution is a great suggestion.

**Check Venables and Ripley p119**
"Outliers are sample values that cause surprise in relation to the majority of the sample. This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors."

Analyse outliers:  
- check 298 (duplicate): small catchments had rows 286 - 312 repeated, fixed  
- 124, D3, -50, 250, Amatya and Skaggs, 2008: clearly 250\% change is wrong. The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively. So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%. corrected to 16\%
- 3, Baker Creek, -42.2, 201.1, Zhang and Wei, 2012. Also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected

A VIF analysis is also a great suggestion as this will indicate variables which are highly correlated within the dataset, such as the Precipitation and Dryness (which we already indicate).

```{r}
require(tidyverse)
require(lubridate)
require(mgcv)
require(car)

All_data <- read_csv("All_data.csv")
All_data2 <- read_csv("All_data2.csv")
All_data2 <- All_data2 %>%
  mutate(length = To - From,
         mid_year = From + (To - From)/2)

#Maybe first simply
# https://stackoverflow.com/questions/52554336/plot-the-equivalent-of-correlation-matrix-for-factors-categorical-data-and-mi
library(ggcorrplot)
png("testcorr_plot.png", width = 960, height = 960)
model.matrix(~0+DeltaF_perc_pos + 
                    Forest_Sign + 
                   log10(Area_km2) + 
                    Dryness + 
                    length +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime, data=All_data2) %>% 
  cor(use="pairwise.complete.obs") %>% 
  ggcorrplot(show.diag = F, type="lower", lab=TRUE, lab_size=4)
dev.off()

model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + 
                    Forest_Sign + 
                   log10(Area_km2) + 
                    Dryness + 
                    length +# s(From, k = 40, bs = "ts") +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data2)
summary(model6_all)
gam.check(model6_all)
#plot(model6_all)
inf1 = influence.gam(model6_all)
hist(inf1)
hist(cooks.distance(model6_all))
which.max(cooks.distance(model6_all))

regclass::VIF(model6_all)

# Can only do car::vif on linear model and only on numerical values
car::vif(gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    log10(Area_km2) + 
                    Dryness + 
                    length, data = All_data2))
```

So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.

However, being too focused on just the residual distribution can also lead to over fitting or over smoothing of the variables. This means that important characteristics of the data might be lost.

\textcolor{blue}{- The manuscript presents a number of alternative statistical models, differing by candidate explanatory variables. Each model is designed considering the key shortcomings of the previous one. The end result of such an approach is a complex and somewhat non systematic exploration of predictors and their explanatory power, where it is easy to get lost. I suggest restructuring the manuscript around a well-designed and robustly formalized model selection. One way to proceed could be to start with the most complex model suggested by the extant understanding of the processes at play, and then proceed with a model simplification, according to some consistent criteria (AIC, dropping non significant terms, or similar; high r2 is not a good criterion because it does not consider the number of parameters). A full blown model selection would also allow to retain or discard the interaction terms, which could be important (as also recognized by the authors; Section 4.5) and should not be discarded a priori. Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}

As outlined in the last part of the reply, the focus of the statistical modelling

\textcolor{blue}{- The novelty of this work needs to emerge more clearly in the introduction. As it looks now, the manuscript could be easily considered somewhat confirmatory, with respect to most data, approaches and conclusions reached by Zhang et al 2017 and Filoso et al 2017. Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}

\textcolor{blue}{There are also some typos and unfinished sentences (e.g., L 142, L 298). Some units are missing (for example those of length of the experiment in the figures) and symbols are not defined at their first appearance (E0/Pa in L 99; Dryness Index). Also: how is Table 1 used? These are not big issues per se but are nonetheless distracting.}

\textcolor{blue}{I would also like to provide the authors with a couple of suggestions regarding the statistical model and their interpretation. 
- The models used in the manuscript consider the absolute value of the forest cover change and then its sign, but this choice is not well justified. It implicitly assumes that the status corresponding to no change distinguishes two ‘realms’. Yet, I would expect (and it is also hinted at at some point in the manuscript) that what really matters is the \%forested area (possibly in relation to the climatic conditions) and how it changes. So, I would suggest the authors to consider whether a model nearer to our understanding of the phenomena at play would be one including, for the forest part, \%change in forested area (with sign) and \%forested area, with the latter possibly as random effect, if not of interested.
- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode. While it is important to present also the r2, even a model with low r2 square we learn which factors significantly affect the change in streamflow and which do not do so.}

### 2022-07-06 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
changes to data and changes to reply to reviewers

@@
 
-The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.
+**The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.**
 
@@
 
-\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.
-The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
+\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
+
+\textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
 Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.
-At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.
-The final - and perhaps in truth the most problematic issue that may stand in the way of the MS being published - relates to novelty.
+At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
+
+\textcolor{blue}{The final - and perhaps in truth the most problematic issue that may stand in the way of the MS being published - relates to novelty.
 The AE highlights several recent works with similar questions, approaches and findings at global scales.
@@
 So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}
+
+Thank you for the suggestion of the @levy2018 paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change.
+This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
+Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as a "verfication" of our analysis.
+
+Impact of forest cover on streamflow: it is not that simple.
+While easily hypothesised, the research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+In terms of the "novelty" of our work, we provide two key insights.  
+- While global databases can be interesting, we need to be careful with drawing major conclusions (as in @zhang2017, @filoso2017, @zhou2015 and @jackson2005) based on basic regression analysis.
+In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
+This is without considering the number of errors that existed in the data.
+This is particularly important, since results from these global analyses are used to build further models to analyse impacts [e.g.
+@hoekvandijke2022], leading to possible wrong policy or management responses.
+- Cumulative and average values of change can be misleading, especially when extracted
 
@@
 There are in essence two "modes" in relation statistical modelling.
-Generally a model is developed to be used in predictive mode: using a model to predict unknown values.
+Generally a model is developed to be used in predictive mode: using a model to predict unknown values, either within or beyond the current data set (forecasting).
 In this case the model should be reduced to its most efficient version that minimises the bias - variance trade-off.
-In this case, automatic variable selection and potentially validation on independent data are important.
+In this case, automatic variable selection and potentially validation on independent data are important, as the aim is to develop the most robust model for prediction.
 
@@
 Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript.
-Including such diagnostic plots for single variable regression is often ommitted.
+Including such diagnostic plots for single variable regression is often omitted.
 For example, both @zhang2017 and @filoso2017 do not present any diagnostics for their relationships and simply looking at their regressions qualitatively indicates issues with the residuals.
@@
 
-Analyse outliers:  
+However, another careful review of the data identified many further errors in the data, which were all originally in @zhang2017.
+A particular problem was that many catchments had the wrong sign for the change in forest cover.
+There are many catchments with reported positive change in cover and a large increase in flow.
+These were all checked and corrected:
 - check 298 (duplicate): small catchments had rows 286 - 312 repeated, fixed  
@@
 - 3, Baker Creek, -42.2, 201.1, Zhang and Wei, 2012. Also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
+- 67, April rd, 100, 167.7 originally attributed to Ruprecht and Schofield (1991) in @zhang2017.
+This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates "clearfelling".
+As a result the change in forest cover was changed to -100\% rather than +100\%.
+- 210, March rd, 100, 147.6. Same problem as 67, Bari et al. (1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
+- 213, 214 and 215, Monda 1, 2 and 3.
+This is a tricky one.
+The original paper [@oshaughnessy1979] only reports on the control period and indicates that the catchments will be cleared.
+The later summary paper [@watson2001] shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
+The further complication is that the treatment included clearing and reseeding and regrowth.
+This suggest that the records should be removed from the database, or only the first few years of the experiment used.
 
@@
 
-model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + 
-                    Forest_Sign + 
-                   log10(Area_km2) + 
-                    Dryness + 
-                    length +# s(From, k = 40, bs = "ts") +
-                    Precip_data_type +  Assessment_technique +
-                    Forest_type +
-                    Hydrological_regime
-                    , data = All_data2)
-summary(model6_all)
-gam.check(model6_all)
-#plot(model6_all)
-inf1 = influence.gam(model6_all)
-hist(inf1)
-hist(cooks.distance(model6_all))
-which.max(cooks.distance(model6_all))
-
-regclass::VIF(model6_all)
-
-# Can only do car::vif on linear model and only on numerical values
-car::vif(gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    log10(Area_km2) + 
-                    Dryness + 
-                    length, data = All_data2))
+# model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + 
+#                     Forest_Sign + 
+#                    log10(Area_km2) + 
+#                     Dryness + 
+#                     length +# s(From, k = 40, bs = "ts") +
+#                     Precip_data_type +  Assessment_technique +
+#                     Forest_type +
+#                     Hydrological_regime
+#                     , data = All_data2)
+# summary(model6_all)
+# gam.check(model6_all)
+# #plot(model6_all)
+# inf1 = influence.gam(model6_all)
+# hist(inf1)
+# hist(cooks.distance(model6_all))
+# which.max(cooks.distance(model6_all))
+# 
+# regclass::VIF(model6_all)
+# 
+# # Can only do car::vif on linear model and only on numerical values
+# car::vif(gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+#                     log10(Area_km2) + 
+#                     Dryness + 
+#                     length, data = All_data2))
 ```
@@
 
-As outlined in the last part of the reply, the focus of the statistical modelling
+As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data
 
@@
 Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}
+
+This is a good point and we have rewritten the paper to strengthen this point.
 
@@
 Yet, I would expect (and it is also hinted at at some point in the manuscript) that what really matters is the \%forested area (possibly in relation to the climatic conditions) and how it changes.
-So, I would suggest the authors to consider whether a model nearer to our understanding of the phenomena at play would be one including, for the forest part, \%change in forested area (with sign) and \%forested area, with the latter possibly as random effect, if not of interested.
-- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode.
+So, I would suggest the authors to consider whether a model nearer to our understanding of the phenomena at play would be one including, for the forest part, \%change in forested area (with sign) and \%forested area, with the latter possibly as random effect, if not of interested.}
+
+\textcolor{blue}{- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode.
 While it is important to present also the r2, even a model with low r2 square we learn which factors significantly affect the change in streamflow and which do not do so.}

### 2022-07-06 paper/Forest_and_Water/Forest_and_Water.Rmd
changes since publication
changes to data and changes to reply to reviewers

@@
 However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-
-<!-- Include Filoso paper -->
+In contrast, a more recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, showing also quite a dramatic variation in responses to deforestation.
+Understanding the impact of
+
 For the purpose of this paper, *watershed* and *catchment* are interchangeable terms.
@@
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
+The more regionally concentrated and detailed study by @levy2018 is a further independent dataset with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover would need to be derived again.
 
@@
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
+
+In contrast to the basic linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+The analysis specifically accounted for differences between catchments and differences due to variations in climate
 
@@
   filter(is.na(Forest_Sign)==T)
+```
+
+```{r, echo = F}
+# temporary write out of data files for review reply
+write_csv(All_data, "All_data.csv")
+write_csv(All_data2, "All_data2.csv")
+
 ```

### 2022-07-07 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
further updates on paper, response and data

@@
 These were all checked and corrected:
-- check 298 (duplicate): small catchments had rows 286 - 312 repeated, fixed  
-- 124, D3, -50, 250, Amatya and Skaggs, 2008: clearly 250\% change is wrong.
+- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
@@
 corrected to 16\%
-- 3, Baker Creek, -42.2, 201.1, Zhang and Wei, 2012. Also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
-- 67, April rd, 100, 167.7 originally attributed to Ruprecht and Schofield (1991) in @zhang2017.
+- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
+- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in @zhang2017.
 This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates "clearfelling".
@@
 - 213, 214 and 215, Monda 1, 2 and 3.
-This is a tricky one.
+These catchments are tricky.
 The original paper [@oshaughnessy1979] only reports on the control period and indicates that the catchments will be cleared.
@@
 This suggest that the records should be removed from the database, or only the first few years of the experiment used.
+In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).
+- 230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta_F_perc to -85 from 90.  
+- 312, Yerraminup South.
+The original publication for this catchment is a Western Australian Water Authority report from 1987, which is hard to find, but we have added a copy in the "Papers" folder on github.
+In this report, in Table 2 on page 11, for the catchment a "Crown cover" decrease of 60\% is given.
+Changed the sign of the change in forest cover: -60\%.
+- 100 Coachwood, 103 Corkwood, and 117 Bollygum, as cited by @cornish1993 and @cornish2001.
+In the database from @zhang2017, the forest change for all these catchments is positive.
+However, the paper highlights that these catchments were all logged and either naturally regenerated or were planted with a plantation species.
+So, similar to the the earlier mentioned Monda catchments, the reported change probably only refers to the first couple of years after clearing (before regrowth).
+In any case, the reported change in forest cover should be negative (clearing) rather than positive.
+Corrected for all three catchments.
+- 78, Black Spur 1, the treatments and effects are only reported in a conference paper [@jayasuriya1988] and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in @zhang2017).
+Corrected.
+Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.
 

### 2022-07-07 paper/Forest_and_Water/Reply_first_review_JoH.tex
further updates on paper, response and data

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

\usepackage{lineno} % add

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
            pdftitle={Replies to reviewer JoH},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\begin{document}

\begin{frontmatter}

  \title{Replies to reviewer JoH}
    \author[]{}

  \begin{abstract}

  \end{abstract}

 \end{frontmatter}

This document records the replies to the reviewer for the first submission of ``Factors determining how catchments respond to forest cover change. Re-analysing global data sets'' to Journal of Hydrology which was rejected.

Reviewer comments in \textcolor{blue}{blue} responses in normal text

\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}

\hypertarget{editor-comments}{%
\subsection{Editor comments}\label{editor-comments}}

\textcolor{blue}{The first is to streamline the statistics.  As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}

We politely disagree with this, as we outline in our reply to the AE. If the statistical modelling aimed at developing the best predictive model, then this would be the right approach. However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.

\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}

\textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.  Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.  At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}

\textcolor{blue}{The final - and perhaps in truth the most problematic issue that may stand in the way of the MS being published - relates to novelty.  The AE highlights several recent works with similar questions, approaches and findings at global scales.  I could add to that list with more regional studies (e.g. Levy et al 2018).  So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}

Thank you for the suggestion of the Levy et al. (2018) paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change. This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data. Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use Levy et al. (2018) as a ``verfication'' of our analysis.

Impact of forest cover on streamflow: it is not that simple. While easily hypothesised, the research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
In terms of the ``novelty'' of our work, we provide two key insights.\\
- While global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al. (2017), Filoso et al. (2017), Zhou et al. (2015) and Jackson et al. (2005)) based on basic regression analysis. In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships. This is without considering the number of errors that existed in the data. This is particularly important, since results from these global analyses are used to build further models to analyse impacts (e.g. Hoek van Dijke et al. 2022), leading to possible wrong policy or management responses.\\
- Cumulative and average values of change can be misleading, especially when extracted

\textcolor{blue}{These are substantial changes and go beyond a major revision.  For this reason, we're rejecting the MS at present.  If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}

\textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.  Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}

\hypertarget{associate-editor}{%
\subsection{Associate Editor:}\label{associate-editor}}

\textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields. The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}

\textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
- As apparent from the diagnostic plots, the model assumptions may be violated in many cases. This can make the results of the fitting (and hence the manuscript conclusions) incorrect. I urge the authors to double check if this is indeed the case and consider ways to address the problem. It is also good practice to check the relevance of outliers (of data with high VIF) and set them aside before model fitting. It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}

Thank you for raising these important points in relation to the validity of the statistical model.
A first point that arises from this is that we clarify better what the aim of our statistical modelling is.

There are in essence two ``modes'' in relation statistical modelling. Generally a model is developed to be used in predictive mode: using a model to predict unknown values, either within or beyond the current data set (forecasting). In this case the model should be reduced to its most efficient version that minimises the bias - variance trade-off. In this case, automatic variable selection and potentially validation on independent data are important, as the aim is to develop the most robust model for prediction.

However, a second reason for fitting a statistical model is purely to explain the variance in the data. In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data. This is subsequently followed by a step by step analysis to test the different causal relationship, either as single variables (as in Zhang et al. (2017)) or jointly (as in our approach). In this case there is no attempt to find the best ``predictive model'', the focus is on the additional amount of explained variance from adding each variable.

\textbf{Include a section in the paper to explain this clearly}

Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript. Including such diagnostic plots for single variable regression is often omitted. For example, both Zhang et al. (2017) and Filoso et al. (2017) do not present any diagnostics for their relationships and simply looking at their regressions qualitatively indicates issues with the residuals.

In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables.

Because our modelling is focused on explaining the maximum variance, excluding values that are outliers is probably not a good idea. However, identifying outliers (particularly values with high leverage), and investigating why these might be different from the overall distribution is a great suggestion.

\textbf{Check Venables and Ripley p119}
``Outliers are sample values that cause surprise in relation to the majority of the sample. This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors.''

However, another careful review of the data identified many further errors in the data, which were all originally in Zhang et al. (2017). A particular problem was that many catchments had the wrong sign for the change in forest cover. There are many catchments with reported positive change in cover and a large increase in flow. These were all checked and corrected:\\
- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by Zhang et al. (2017) is clearly wrong. The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively. So value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%. corrected to 16\%
- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by Zhang et al. (2017) is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected\\
- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in Zhang et al. (2017). This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''. As a result the change in forest cover was changed to -100\% rather than +100\%.
- 210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
- 213, 214 and 215, Monda 1, 2 and 3. These catchments are tricky. The original paper (O'Shaughnessy et al. 1979) only reports on the control period and indicates that the catchments will be cleared. The later summary paper (Watson et al. 2001) shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries. The further complication is that the treatment included clearing and reseeding and regrowth. This suggest that the records should be removed from the database, or only the first few years of the experiment used. In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).\\
- 230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.\\
- 312, Yerraminup South. The original publication for this catchment is a Western Australian Water Authority report from 1987, which is hard to find, but we have added a copy in the ``Papers'' folder on github. In this report, in Table 2 on page 11, for the catchment a ``Crown cover'' decrease of 60\% is given. Changed the sign of the change in forest cover: -60\%.\\
- 100 Coachwood, 103 Corkwood, and 117 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001). In the database from Zhang et al. (2017), the forest change for all these catchments is positive. However, the paper highlights that these catchments were all logged and either naturally regenerated or were planted with a plantation species. So, similar to the the earlier mentioned Monda catchments, the reported change probably only refers to the first couple of years after clearing (before regrowth). In any case, the reported change in forest cover should be negative (clearing) rather than positive. Corrected for all three catchments.\\
- 78, Black Spur 1, the treatments and effects are only reported in a conference paper (Jayasuriya and O'Shaughnessy 1988) and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in Zhang et al. (2017)). Corrected. Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.

A VIF analysis is also a great suggestion as this will indicate variables which are highly correlated within the dataset, such as the Precipitation and Dryness (which we already indicate).

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: tidyverse
\end{verbatim}

\begin{verbatim}
## Warning: package 'tidyverse' was built under R version 4.1.3
\end{verbatim}

\begin{verbatim}
## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
\end{verbatim}

\begin{verbatim}
## v ggplot2 3.3.6     v purrr   0.3.4
## v tibble  3.1.6     v dplyr   1.0.9
## v tidyr   1.2.0     v stringr 1.4.0
## v readr   2.1.1     v forcats 0.5.1
\end{verbatim}

\begin{verbatim}
## Warning: package 'ggplot2' was built under R version 4.1.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'tidyr' was built under R version 4.1.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'dplyr' was built under R version 4.1.3
\end{verbatim}

\begin{verbatim}
## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
## x dplyr::filter() masks stats::filter()
## x dplyr::lag()    masks stats::lag()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(lubridate)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: lubridate
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'lubridate'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:base':
## 
##     date, intersect, setdiff, union
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(mgcv)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: mgcv
\end{verbatim}

\begin{verbatim}
## Warning: package 'mgcv' was built under R version 4.1.3
\end{verbatim}

\begin{verbatim}
## Loading required package: nlme
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'nlme'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     collapse
\end{verbatim}

\begin{verbatim}
## This is mgcv 1.8-40. For overview type 'help("mgcv-package")'.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(car)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: car
\end{verbatim}

\begin{verbatim}
## Loading required package: carData
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'car'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     recode
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:purrr':
## 
##     some
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"All\_data.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Rows: 327 Columns: 24
\end{verbatim}

\begin{verbatim}
## -- Column specification --------------------------------------------------------
## Delimiter: ","
## chr (10): Watershed name, Forest_type, Hydrological_regime, Precip_data_type...
## dbl (14): Watershed #, Area_km2, Pa_mm, DeltaF_perc, DeltaQf_perc, E0, Latit...
## 
## i Use `spec()` to retrieve the full column specification for this data.
## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data2 }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"All\_data2.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Rows: 327 Columns: 26
## -- Column specification --------------------------------------------------------
## Delimiter: ","
## chr (11): Watershed name, Forest_type, Hydrological_regime, Precip_data_type...
## dbl (15): Watershed #, Area_km2, Pa_mm, DeltaF_perc, DeltaQf_perc, E0, Latit...
## 
## i Use `spec()` to retrieve the full column specification for this data.
## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data2 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{length =}\NormalTok{ To }\SpecialCharTok{{-}}\NormalTok{ From,}
         \AttributeTok{mid\_year =}\NormalTok{ From }\SpecialCharTok{+}\NormalTok{ (To }\SpecialCharTok{{-}}\NormalTok{ From)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}

\CommentTok{\#Maybe first simply}
\CommentTok{\# https://stackoverflow.com/questions/52554336/plot{-}the{-}equivalent{-}of{-}correlation{-}matrix{-}for{-}factors{-}categorical{-}data{-}and{-}mi}
\FunctionTok{library}\NormalTok{(ggcorrplot)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: package 'ggcorrplot' was built under R version 4.1.3
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{png}\NormalTok{(}\StringTok{"testcorr\_plot.png"}\NormalTok{, }\AttributeTok{width =} \DecValTok{960}\NormalTok{, }\AttributeTok{height =} \DecValTok{960}\NormalTok{)}
\FunctionTok{model.matrix}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\DecValTok{0}\SpecialCharTok{+}\NormalTok{DeltaF\_perc\_pos }\SpecialCharTok{+} 
\NormalTok{                    Forest\_Sign }\SpecialCharTok{+} 
                   \FunctionTok{log10}\NormalTok{(Area\_km2) }\SpecialCharTok{+} 
\NormalTok{                    Dryness }\SpecialCharTok{+} 
\NormalTok{                    length }\SpecialCharTok{+}
\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
\NormalTok{                    Forest\_type }\SpecialCharTok{+}
\NormalTok{                    Hydrological\_regime, }\AttributeTok{data=}\NormalTok{All\_data2) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{cor}\NormalTok{(}\AttributeTok{use=}\StringTok{"pairwise.complete.obs"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ggcorrplot}\NormalTok{(}\AttributeTok{show.diag =}\NormalTok{ F, }\AttributeTok{type=}\StringTok{"lower"}\NormalTok{, }\AttributeTok{lab=}\ConstantTok{TRUE}\NormalTok{, }\AttributeTok{lab\_size=}\DecValTok{4}\NormalTok{)}
\FunctionTok{dev.off}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# model6\_all \textless{}{-} gam(DeltaQf\_perc \textasciitilde{} DeltaF\_perc\_pos + }
\CommentTok{\#                     Forest\_Sign + }
\CommentTok{\#                    log10(Area\_km2) + }
\CommentTok{\#                     Dryness + }
\CommentTok{\#                     length +\# s(From, k = 40, bs = "ts") +}
\CommentTok{\#                     Precip\_data\_type +  Assessment\_technique +}
\CommentTok{\#                     Forest\_type +}
\CommentTok{\#                     Hydrological\_regime}
\CommentTok{\#                     , data = All\_data2)}
\CommentTok{\# summary(model6\_all)}
\CommentTok{\# gam.check(model6\_all)}
\CommentTok{\# \#plot(model6\_all)}
\CommentTok{\# inf1 = influence.gam(model6\_all)}
\CommentTok{\# hist(inf1)}
\CommentTok{\# hist(cooks.distance(model6\_all))}
\CommentTok{\# which.max(cooks.distance(model6\_all))}
\CommentTok{\# }
\CommentTok{\# regclass::VIF(model6\_all)}
\CommentTok{\# }
\CommentTok{\# \# Can only do car::vif on linear model and only on numerical values}
\CommentTok{\# car::vif(gam(DeltaQf\_perc \textasciitilde{} DeltaF\_perc\_pos + Forest\_Sign + }
\CommentTok{\#                     log10(Area\_km2) + }
\CommentTok{\#                     Dryness + }
\CommentTok{\#                     length, data = All\_data2))}
\end{Highlighting}
\end{Shaded}

So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.

However, being too focused on just the residual distribution can also lead to over fitting or over smoothing of the variables. This means that important characteristics of the data might be lost.

\textcolor{blue}{- The manuscript presents a number of alternative statistical models, differing by candidate explanatory variables. Each model is designed considering the key shortcomings of the previous one. The end result of such an approach is a complex and somewhat non systematic exploration of predictors and their explanatory power, where it is easy to get lost. I suggest restructuring the manuscript around a well-designed and robustly formalized model selection. One way to proceed could be to start with the most complex model suggested by the extant understanding of the processes at play, and then proceed with a model simplification, according to some consistent criteria (AIC, dropping non significant terms, or similar; high r2 is not a good criterion because it does not consider the number of parameters). A full blown model selection would also allow to retain or discard the interaction terms, which could be important (as also recognized by the authors; Section 4.5) and should not be discarded a priori. Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}

As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data

\textcolor{blue}{- The novelty of this work needs to emerge more clearly in the introduction. As it looks now, the manuscript could be easily considered somewhat confirmatory, with respect to most data, approaches and conclusions reached by Zhang et al 2017 and Filoso et al 2017. Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}

This is a good point and we have rewritten the paper to strengthen this point.

\textcolor{blue}{There are also some typos and unfinished sentences (e.g., L 142, L 298). Some units are missing (for example those of length of the experiment in the figures) and symbols are not defined at their first appearance (E0/Pa in L 99; Dryness Index). Also: how is Table 1 used? These are not big issues per se but are nonetheless distracting.}

\textcolor{blue}{I would also like to provide the authors with a couple of suggestions regarding the statistical model and their interpretation. 
- The models used in the manuscript consider the absolute value of the forest cover change and then its sign, but this choice is not well justified. It implicitly assumes that the status corresponding to no change distinguishes two ‘realms’. Yet, I would expect (and it is also hinted at at some point in the manuscript) that what really matters is the \%forested area (possibly in relation to the climatic conditions) and how it changes. So, I would suggest the authors to consider whether a model nearer to our understanding of the phenomena at play would be one including, for the forest part, \%change in forested area (with sign) and \%forested area, with the latter possibly as random effect, if not of interested.}

\textcolor{blue}{- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode. While it is important to present also the r2, even a model with low r2 square we learn which factors significantly affect the change in streamflow and which do not do so.}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-cornish1993}{}}%
Cornish, P. M. 1993. {``The Effects of Logging and Forest Regeneration on Water Yields in a Moist Eucalypt Forest in New South Wales, Australia.''} Journal Article. \emph{Journal of Hydrology} 150 (2-4): 301--22. \href{http://www.sciencedirect.com/science/article/B6V6C-487D3Y2-9J/2/73c981ba76284d9d629f6b221d6fd6c6\%20}{http://www.sciencedirect.com/science/article/B6V6C-487D3Y2-9J/2/73c981ba76284d9d629f6b221d6fd6c6 }.

\leavevmode\vadjust pre{\hypertarget{ref-cornish2001}{}}%
Cornish, P. M., and R. A. Vertessy. 2001. {``Forest Age-Induced Changes in Evapotranspiration and Water Yield in a Eucalypt Forest.''} Journal Article. \emph{Journal of Hydrology} 242 (1-2): 43--63. \href{http://www.sciencedirect.com/science/article/B6V6C-429910G-3/2/0158b1f89ff436f338a9e688a47f06c4\%20}{http://www.sciencedirect.com/science/article/B6V6C-429910G-3/2/0158b1f89ff436f338a9e688a47f06c4 }.

\leavevmode\vadjust pre{\hypertarget{ref-filoso2017}{}}%
Filoso, Solange, Maíra Ometto Bezerra, Katherine C. B. Weiss, and Margaret A. Palmer. 2017. {``Impacts of Forest Restoration on Water Yield: A Systematic Review.''} Journal Article. \emph{PLOS ONE} 12 (8): e0183210. \url{https://doi.org/10.1371/journal.pone.0183210}.

\leavevmode\vadjust pre{\hypertarget{ref-hoekvandijke2022}{}}%
Hoek van Dijke, Anne J., Martin Herold, Kaniska Mallick, Imme Benedict, Miriam Machwitz, Martin Schlerf, Agnes Pranindita, Jolanda J. E. Theeuwen, Jean-François Bastin, and Adriaan J. Teuling. 2022. {``Shifts in Regional Water Availability Due to Global Tree Restoration.''} Journal Article. \emph{Nature Geoscience} 15 (5): 363--68. \url{https://doi.org/10.1038/s41561-022-00935-0}.

\leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
Jackson, Robert B., Esteban G. Jobbagy, Roni Avissar, Somnath Baidya Roy, Damian J. Barrett, Charles W. Cook, Kathleen A. Farley, David C. le Maitre, Bruce A. McCarl, and Brian C. Murray. 2005. {``Trading Water for Carbon with Biological Carbon Sequestration.''} Journal Article. \emph{Science} 310 (5756): 1944--47. \url{https://doi.org/10.1126/science.1119282}.

\leavevmode\vadjust pre{\hypertarget{ref-jayasuriya1988}{}}%
Jayasuriya, M. D. A., and P. J. O'Shaughnessy. 1988. {``The Use of Mathematical Models in Evaluating Forest Treatment Effects on Streamflow.''} Book Section. In \emph{Hydrology and Water Resources Symposium 1988: Preprints of Papers,} 135--39. Hydrology; Water Resources Symposium. \url{https://doi.org/10.3316/informit.692214289455295}.

\leavevmode\vadjust pre{\hypertarget{ref-levy2018}{}}%
Levy, M. C., A. V. Lopes, A. Cohn, L. G. Larsen, and S. E. Thompson. 2018. {``Land Use Change Increases Streamflow Across the Arc of Deforestation in Brazil.''} Journal Article. \emph{Geophysical Research Letters} 45 (8): 3520--30. https://doi.org/\url{https://doi.org/10.1002/2017GL076526}.

\leavevmode\vadjust pre{\hypertarget{ref-oshaughnessy1979}{}}%
O'Shaughnessy, P. J., K. J. Langford, H. P. Duncan, and R. J. Moran. 1979. {``Catchment Experiments in Mountain Ash Forests at North Maroondah.''} Journal Article. \emph{Australian Forestry} 42 (3): 150--60. \url{https://doi.org/10.1080/00049158.1979.10674220}.

\leavevmode\vadjust pre{\hypertarget{ref-watson2001}{}}%
Watson, Fred, Rob Vertessy, Tom McMahon, Bruce Rhodes, and Ian Watson. 2001. {``Improved Methods to Assess Water Yield Changes from Paired-Catchment Studies: Application to the Maroondah Catchments.''} Journal Article. \emph{Forest Ecology and Management} 143 (1): 189--204. https://doi.org/\url{https://doi.org/10.1016/S0378-1127(00)00517-X}.

\leavevmode\vadjust pre{\hypertarget{ref-zhang2017}{}}%
Zhang, Mingfang, Ning Liu, Richard Harper, Qiang Li, Kuan Liu, Xiaohua Wei, Dingyuan Ning, Yiping Hou, and Shirong Liu. 2017. {``A Global Review on Hydrological Responses to Forest Change Across Multiple Spatial Scales: Importance of Scale, Climate, Forest Type and Hydrological Regime.''} Journal Article. \emph{Journal of Hydrology} 546: 44--59. https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2016.12.040}.

\leavevmode\vadjust pre{\hypertarget{ref-zhou2015}{}}%
Zhou, Guoyi, Xiaohua Wei, Xiuzhi Chen, Ping Zhou, Xiaodong Liu, Yin Xiao, Ge Sun, et al. 2015. {``Global Pattern for the Effect of Climate and Land Cover on Water Yield.''} Journal Article. \emph{Nature Communications} 6 (1): 5918. \url{https://doi.org/10.1038/ncomms6918}.

\end{CSLReferences}

\end{document}

### 2022-07-10 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
further changes to data and document

@@
 
-Reviewer comments in \textcolor{blue}{blue} responses in normal text
+Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text
 
@@
 \textcolor{blue}{The first is to streamline the statistics.
-As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}
+As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.} \
 
-We politely disagree with this, as we outline in our reply to the AE.
+We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
 If the statistical modelling aimed at developing the best predictive model, then this would be the right approach.
 However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
-
-\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
-
+\
+\
+\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}  \
+\
+Please see our comments in reply to the AE below. \
+\  
 \textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
 Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.
-At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
+At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}  \
+\
+Please see our comments in reply to the AE below.  \
+\
 
@@
 Levy et al 2018).
-So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}
+So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}    \
+\
 
@@
 
-Impact of forest cover on streamflow: it is not that simple.
-While easily hypothesised, the research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
-In terms of the "novelty" of our work, we provide two key insights.  
-- While global databases can be interesting, we need to be careful with drawing major conclusions (as in @zhang2017, @filoso2017, @zhou2015 and @jackson2005) based on basic regression analysis.
+You are quite right that we need to be clearer about the objectives of the paper and how this is different from the existing work.
+We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017).
+\
+\
+As a result of this, we have rewritten the scope of the paper and changed the title to better reflect the main findings and message coming from this paper.   
+\
+\
+The new title of the paper is:
+**Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.**  
+\
+The key contribution of this paper is to highligt the gap that exists in trying to extrapolate multiple local studies to a global scale.
+While the impact of forest cover on streamflow is easily hypothesised (e.g.
+@zhou2015; @hoekvandijke2022), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of  forestation or deforestation, and in many cases this becomes a qualitative assessment.
+As such, we provide three key insights.  
+- While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in @zhang2017, @filoso2017, @zhou2015 and @jackson2005) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
 In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
+In addition, the equilibirum analysis is based on the assumption of water balance closure, which might not always be the case in arid and semi-arid climates.
+It can easily become a case of 'correlation without causation'.
 This is without considering the number of errors that existed in the data.
-This is particularly important, since results from these global analyses are used to build further models to analyse impacts [e.g.
+This is particularly important, since results from these global analyses are used to build further models to analyse global impacts [e.g.
 @hoekvandijke2022], leading to possible wrong policy or management responses.
-- Cumulative and average values of change can be misleading, especially when extracted
+- Cumulative and average values of change can be misleading, especially when extracted from published field studies which originally had different objectives.
+This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
+Many of the Paired Watershed Experiments in Australia [e.g.
+@cornish2001; @webb2009; @webb2012; @webb2013; @watson2001] fall in this category, and therefore easily be classified as either forestation or deforestation.
+In principle a decision needs to be made how many years post clearing needs to be considered, and whether the remainder of the timeseries should be classified as regeneration.
+We demonstrate how removing these experiments from the total changes the results of the analysis.
+- In general, the size of the catchment and the length of the study play a huge role in the interpretation of the results.
+The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
+In terms of the catchment size, one of the key issues we originally had with the @zhang2017 paper was the arbitrary split between catchment > 1000 km^2^ and catchment < 1000 km^2^.
+Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of methods used to analyse small catchments (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
+The paper by @beck2013 is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
+However, as we argue, this results of this paper are misrepresented in the database, as none of the analyse catchments had a significant change in the streamflow.
+This means the change should be recorded as 0.
+\
 
@@
 If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}
-
+\
+We acknowledge this, and we hope that the current revised version and our responses adress these concerns.   
+\
 \textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.
 Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}
+\
+We acknowledge this, and we hope that the current revised version and our responses adress these concerns. 
+\
 
@@
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected:
+These were all checked and corrected.
+A final issue was the inclusion of the catchments from the study by @beck2013.
+This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:
 - 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.
@@
 Changed the sign of the change in forest cover: -60\%.
-- 100 Coachwood, 103 Corkwood, and 117 Bollygum, as cited by @cornish1993 and @cornish2001.
+- 72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by @cornish1993 and @cornish2001.
 In the database from @zhang2017, the forest change for all these catchments is positive.
@@
 Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.
+- 104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta_Q_f to be negative.  
+- 102, Cold Spring. Checking the original paper [@schneider1961] indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta_Q_f to be negative.  
+- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
+These are all catchments in Puerto Rico from the study from @beck2013.
+They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
+The values that are cited in the database should all be set to "not significant from 0", so might be included as 0.
+Including them with positive or negative values is misleading.
+This not judging the study by @beck2013, which is a thorough and excellent modelling study.
+- 85 Bosboukloof.
+This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
+In any case, the data should reflect a decrease in forest cover: changed the sign of Delta_f_perc to -80\%.
+- 259 Shackam Brook.
+There were a few issues with this catchment in the original database.
+The name was misspelled and it was incorrectly attributed to @brown2005.
+The original paper is the same as 102 [@schneider1961].
+Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases.
+Corrected Delta_Qf_perc to -20.7\%.
+- 95 Sage Brook. Similar to 259 and 102, originates from @schneider1961. Reforestation so Delta_Qf_perc corrected to -19.8\%.  
+- 101 Coalburn.
+Original publication (Robinson, 1993) which is a symposium paper, is not available.
+The best summary of the reseearch is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment.
+It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period.
+Changed to -20.3\%.
 

### 2022-07-10 paper/Forest_and_Water/Reply_first_review_JoH.tex
further changes to data and document

@@
 \textcolor{blue}{The first is to streamline the statistics.
-As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}
+As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.} ~
 
-We politely disagree with this, as we outline in our reply to the AE.
+We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
 If the statistical modelling aimed at developing the best predictive model, then this would be the right approach.
 However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
-
-\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
-
+~\\
+\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}\\
+\strut \\
+Please see our comments in reply to the AE below.\\
+\hspace*{0.333em}
 \textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
 Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.
-At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
+At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}\\
+\strut \\
+Please see our comments in reply to the AE below.\\
+\strut \\
 
@@
 Levy et al 2018).
-So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}
+So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}\\
+\strut \\
 
@@
 
-Impact of forest cover on streamflow: it is not that simple.
-While easily hypothesised, the research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
-In terms of the ``novelty'' of our work, we provide two key insights.\\
-- While global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al.
+You are quite right that we need to be clearer about the objectives of the paper and how this is different from the existing work.
+We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in Zhou et al.
+(2015); Jackson et al.
+(2005); Filoso et al.
+(2017) and Zhang et al.
+(2017)).\\
+\strut \\
+\strut \\
+As a result of this, we have rewritten the scope of the paper and changed the title to better reflect the main findings and message coming from this paper.\\
+\strut \\
+\strut \\
+The new title of the paper is:
+\textbf{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}\\
+\strut \\
+The key contribution of this paper is to highligt the gap that exists in trying to extrapolate multiple local studies to a global scale.
+While the impact of forest cover on streamflow is easily hypothesised (e.g.
+Zhou et al.
+(2015); Hoek van Dijke et al.
+(2022)), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of forestation or deforestation, and in many cases this becomes a qualitative assessment.\\
+As such, we provide three key insights.\\
+- While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al.
 (2017), Filoso et al.
@@
 (2015) and Jackson et al.
-(2005)) based on basic regression analysis.
+(2005)) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
 In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
+In addition, the equilibirum analysis is based on the assumption of water balance closure, which might not always be the case in arid and semi-arid climates.
+It can easily become a case of `correlation without causation'.
 This is without considering the number of errors that existed in the data.
-This is particularly important, since results from these global analyses are used to build further models to analyse impacts (e.g.
+This is particularly important, since results from these global analyses are used to build further models to analyse global impacts (e.g.
 Hoek van Dijke et al.
 2022), leading to possible wrong policy or management responses.\\
-- Cumulative and average values of change can be misleading, especially when extracted
+- Cumulative and average values of change can be misleading, especially when extracted from published field studies which originally had different objectives.
+This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
+Many of the Paired Watershed Experiments in Australia (e.g.
+Cornish and Vertessy 2001; Webb 2009; Webb and Kathuria 2012; Webb and Jarrett 2013; Watson et al.
+2001) fall in this category, and therefore easily be classified as either forestation or deforestation.
+In principle a decision needs to be made how many years post clearing needs to be considered, and whether the remainder of the timeseries should be classified as regeneration.
+We demonstrate how removing these experiments from the total changes the results of the analysis.
+- In general, the size of the catchment and the length of the study play a huge role in the interpretation of the results.
+The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
+In terms of the catchment size, one of the key issues we originally had with the Zhang et al.
+(2017) paper was the arbitrary split between catchment \textgreater{} 1000 km\textsuperscript{2} and catchment \textless{} 1000 km\textsuperscript{2}.
+Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of methods used to analyse small catchments (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
+The paper by Beck et al.
+(2013) is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
+However, as we argue, this results of this paper are misrepresented in the database, as none of the analyse catchments had a significant change in the streamflow.
+This means the change should be recorded as 0.\\
+\strut \\
 
@@
 For this reason, we're rejecting the MS at present.
-If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}
-
+If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}\\
+We acknowledge this, and we hope that the current revised version and our responses adress these concerns.\\
+\strut \\
 \textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.
-Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}
+Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}\\
+We acknowledge this, and we hope that the current revised version and our responses adress these concerns.\\
 
@@
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected:\\
+These were all checked and corrected.
+A final issue was the inclusion of the catchments from the study by Beck et al.
+(2013).
+This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:\\
 - 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by Zhang et al.
@@
 Changed the sign of the change in forest cover: -60\%.\\
-- 100 Coachwood, 103 Corkwood, and 117 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001).
+- 72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001).
 In the database from Zhang et al.
@@
 Corrected.
-Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.
+Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.\\
+- 104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
+- 102, Cold Spring.
+Checking the original paper (Schneider and Ayer 1961) indicates that this is in fact a reduction in flow as a result of reforestation.
+Changed the sign of Delta\_Q\_f to be negative.\\
+- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
+These are all catchments in Puerto Rico from the study from Beck et al.
+(2013).
+They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
+The values that are cited in the database should all be set to ``not significant from 0'', so might be included as 0.
+Including them with positive or negative values is misleading.
+This not judging the study by Beck et al.
+(2013), which is a thorough and excellent modelling study.\\
+- 85 Bosboukloof.
+This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
+In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.\\
+- 259 Shackam Brook.
+There were a few issues with this catchment in the original database.
+The name was misspelled and it was incorrectly attributed to Brown et al.
+(2005).
+The original paper is the same as 102 (Schneider and Ayer 1961).
+Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases.
+Corrected Delta\_Qf\_perc to -20.7\%.\\
+- 95 Sage Brook. Similar to 259 and 102, originates from Schneider and Ayer (1961). Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
+- 101 Coalburn.
+Original publication (Robinson, 1993) which is a symposium paper, is not available.
+The best summary of the reseearch is in Birkinshaw, Bathurst, and Robinson (2014) which summaries 45 years of resaerch in the Coalburn catchment.
+It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period.
+Changed to -20.3\%.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-beck2013}{}}%
+Beck, H.
+E., L.
+A.
+Bruijnzeel, A.
+I.
+J.
+M.
+van Dijk, T.
+R.
+McVicar, F.
+N.
+Scatena, and J.
+Schellekens.
+2013.
+{``The Impact of Forest Regeneration on Streamflow in 12 Mesoscale Humid Tropical Catchments.''} Journal Article.
+\emph{Hydrol.
+Earth Syst.
+Sci.} 17 (7): 2613--35.
+\url{https://doi.org/10.5194/hess-17-2613-2013}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-birkinshaw2014}{}}%
+Birkinshaw, Stephen J., James C.
+Bathurst, and Mark Robinson.
+2014.
+{``45 Years of Non-Stationary Hydrology over a Forest Plantation Growth Cycle, Coalburn Catchment, Northern England.''} Journal Article.
+\emph{Journal of Hydrology} 519: 559--73.
+https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2014.07.050}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-brown2005}{}}%
+Brown, Alice E., Lu Zhang, Thomas A.
+McMahon, Andrew W.
+Western, and Robert A.
+Vertessy.
+2005.
+{``A Review of Paired Catchment Studies for Determining Changes in Water Yield Resulting from Alterations in Vegetation.''} Journal Article.
+\emph{Journal of Hydrology} 310 (1-4): 28--61.
+\url{http://www.sciencedirect.com/science/article/B6V6C-4G05MM9-1/2/bbc5fc0e958a8f34bcb7c1cc7fa57b48}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-cornish1993}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-schneider1961}{}}%
+Schneider, William Joseph, and Gordon Roundy Ayer. 1961. {``Effect of Reforestation on Streamflow in Central New York.''} Report 1602. \url{https://doi.org/10.3133/wsp1602}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-watson2001}{}}%
@@
 https://doi.org/\url{https://doi.org/10.1016/S0378-1127(00)00517-X}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-webb2009}{}}%
+Webb, Ashley A.
+2009.
+{``Streamflow Response to Pinus Plantation Harvesting: Canobolas State Forest, Southeastern Australia.''} Journal Article.
+\emph{Hydrological Processes} 23 (12): 1679--89.
+https://doi.org/\url{https://doi.org/10.1002/hyp.7301}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-webb2013}{}}%
+Webb, Ashley A., and Brad W.
+Jarrett.
+2013.
+{``Hydrological Response to Wildfire, Integrated Logging and Dry Mixed Species Eucalypt Forest Regeneration: The Yambulla Experiment.''} Journal Article.
+\emph{Forest Ecology and Management} 306: 107--17.
+https://doi.org/\url{https://doi.org/10.1016/j.foreco.2013.06.020}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-webb2012}{}}%
+Webb, Ashley A., and Amrit Kathuria.
+2012.
+{``Response of Streamflow to Afforestation and Thinning at Red Hill, Murray Darling Basin, Australia.''} Journal Article.
+\emph{Journal of Hydrology} 412-413: 133--40.
+https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2011.05.033}.
 

### 2022-07-17 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
furtehr updates on reply to reviewers

@@
 ---
-title: "Replies to reviewer JoH"
+title: "Replies to editor and associate editor Journal of Hydrology"
 author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, echo =F}
 knitr::opts_chunk$set(echo = TRUE)
+require(tidyverse)
+require(lubridate)
+suppressWarnings(require(knitr))
+suppressWarnings(require(bookdown))
+suppressWarnings(require(pander))
 ```
 
-This document records the replies to the reviewer for the first submission of "Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets" to Journal of Hydrology which was rejected.
+# Introduction
 
-Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text
+This document records the replies to the reviewer for the first submission of *"Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets"* to Journal of Hydrology, which was rejected before review.
+\
+Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text.  
+\
+**The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.**  
+\
 
-**The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.**
-
-## Editor comments
+# Editor comments
 
 \textcolor{blue}{The first is to streamline the statistics.
-As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.} \
-
+As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}
+\
+**check this**
 We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
@@
 \
+\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
 \
-\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}  \
-\
-Please see our comments in reply to the AE below. \
+Please see our comments in reply to the AE below.    
 \  
@@
 Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.
-At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}  \
+At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
 \
@@
 \
-The key contribution of this paper is to highligt the gap that exists in trying to extrapolate multiple local studies to a global scale.
-While the impact of forest cover on streamflow is easily hypothesised (e.g.
-@zhou2015; @hoekvandijke2022), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
+While the impact of forest cover on streamflow is easily hypothesised [e.g.
+@zhou2015; @hoekvandijke2022], our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
 In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of  forestation or deforestation, and in many cases this becomes a qualitative assessment.
 As such, we provide three key insights.
+
 - While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in @zhang2017, @filoso2017, @zhou2015 and @jackson2005) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
@@
 The paper by @beck2013 is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
-However, as we argue, this results of this paper are misrepresented in the database, as none of the analyse catchments had a significant change in the streamflow.
+However, as we argue, the results of this paper are misrepresented in the database, as none of the analysed catchments had a significant change in the streamflow.
 This means the change should be recorded as 0.
@@
 \
-We acknowledge this, and we hope that the current revised version and our responses adress these concerns.   
+We acknowledge this, and we hope that the current revised version and our responses address these concerns.   
 \
@@
 \
-We acknowledge this, and we hope that the current revised version and our responses adress these concerns. 
+We acknowledge this, and we hope that the current revised version and our responses address these concerns. 
 \
 
-## Associate Editor:
+# Associate Editor:
 
@@
 The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}
-
+\
+<!-- answer -->
+Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.  
+\
 \textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
@@
 It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}
-
+\
+### error distributions
+<!-- answer -->
 Thank you for raising these important points in relation to the validity of the statistical model. 
 A first point that arises from this is that we clarify better what the aim of our statistical modelling is.  
-
-There are in essence two "modes" in relation statistical modelling.
+There are in essence two approaches to statistical modelling.
 Generally a model is developed to be used in predictive mode: using a model to predict unknown values, either within or beyond the current data set (forecasting).
 In this case the model should be reduced to its most efficient version that minimises the bias - variance trade-off.
-In this case, automatic variable selection and potentially validation on independent data are important, as the aim is to develop the most robust model for prediction.
+Automatic variable selection and potentially validation on independent data are therefore important, as the aim is to develop the most robust model for prediction.
+However, a second reason for to use a statistical model is to explain the maximum variance in the data.
+In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
+This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in @zhang2017) or jointly (as in our approach).
+In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from adding each variable.
+This explains why we build the model starting from the most simple model, rather than starting from the most complex model.
 
-However, a second reason for fitting a statistical model is purely to explain the variance in the data.
-In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
-This is subsequently followed by a step by step analysis to test the different causal relationship, either as single variables (as in @zhang2017) or jointly (as in our approach).
-In this case there is no attempt to find the best "predictive model", the focus is on the additional amount of explained variance from adding each variable.
-
+\
 **Include a section in the paper to explain this clearly**  
+\
 
 Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript.
-Including such diagnostic plots for single variable regression is often omitted.
-For example, both @zhang2017 and @filoso2017 do not present any diagnostics for their relationships and simply looking at their regressions qualitatively indicates issues with the residuals.
+In many cases, including such diagnostic plots for single variable regressions is often omitted.
+For example, both @zhang2017 and @filoso2017 do not present any diagnostics for their relationships and the regressions (Fig 2 - Fig 4 in @zhang2017 and Fig 9 in @filoso2017) qualitatively indicate issues with the residuals.
 
-In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables.
+In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, or explaining why we did not take any further steps.
 
-Because our modelling is focused on explaining the maximum variance, excluding values that are outliers is probably not a good idea.
-However, identifying outliers (particularly values with high leverage), and investigating why these might be different from the overall distribution is a great suggestion.
+### the issue of outliers
 
@@
 This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors."
-
+Because our modelling is focused on explaining the maximum variance, we believe that excluding values that are outliers is probably not a good idea.  
+\
 However, another careful review of the data identified many further errors in the data, which were all originally in @zhang2017.
@@
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected.
-A final issue was the inclusion of the catchments from the study by @beck2013.
-This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:
+These were all checked and corrected if needed and a full list of all these changes is below:
+
 - 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.
@@
 - 102, Cold Spring. Checking the original paper [@schneider1961] indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta_Q_f to be negative.  
-- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
-These are all catchments in Puerto Rico from the study from @beck2013.
-They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
-The values that are cited in the database should all be set to "not significant from 0", so might be included as 0.
-Including them with positive or negative values is misleading.
-This not judging the study by @beck2013, which is a thorough and excellent modelling study.
 - 85 Bosboukloof.
@@
 Original publication (Robinson, 1993) which is a symposium paper, is not available.
-The best summary of the reseearch is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment.
+The best summary of the research is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment.
 It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period.
 Changed to -20.3\%.
+
+A final issue was the inclusion of the results of several catchments, for example from the study by @beck2013, which had no significant change in flow.
+Despite this, the "average" change in flow was reported int he database.
+We don't believe that this is correct and the results from such studies should be set to 0.
+A full list of changes is provided below:
+
+- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
+These are all catchments in Puerto Rico from the study from @beck2013.
+They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
+The values that are cited in the database should all be set to "not significant from 0", so might be included as 0.
+Including them with positive or negative values is misleading.
+This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation.
+Values for all 12 studied catchments set to 0 in the database.
+- 188 Kimakia
+- 221	N. Creek, Babinda, Queensland
+- 254 Sambret
+
+There were similar studies that we identified that
+
+## VIF analysis and understanding cross correlations between the variables
 
@@
 ```{r}
-require(tidyverse)
-require(lubridate)
 require(mgcv)
@@
 
+```{r corgraphs, echo=F, out.width="90%", fig.cap="Correlation matrix for all variables"}
+
+include_graphics("./testcorr_plot.png")
+```
 So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.
@@
 Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}
-
-As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data
+\
+<!-- answer -->
+As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data.   
+\
 
@@
 Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}
-
+\
+<!-- answer -->
 This is a good point and we have rewritten the paper to strengthen this point.  
+\
 

### 2022-07-17 paper/Forest_and_Water/Reply_first_review_JoH.tex
furtehr updates on reply to reviewers

@@
             pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
-            pdftitle={Replies to reviewer JoH},
+            pdftitle={Replies to editor and associate editor Journal of Hydrology},
             colorlinks=false,
@@
 
-  \title{Replies to reviewer JoH}
+  \title{Replies to editor and associate editor Journal of Hydrology}
     \author[]{}
@@
 
-This document records the replies to the reviewer for the first submission of ``Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets'' to Journal of Hydrology which was rejected.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text
-
-\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}
+This document records the replies to the reviewer for the first submission of \emph{``Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets''} to Journal of Hydrology, which was rejected before review.\\
+\strut \\
+Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text.\\
+\strut \\
+\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}\\
+\strut \\
 
 \hypertarget{editor-comments}{%
-\subsection{Editor comments}\label{editor-comments}}
+\section{Editor comments}\label{editor-comments}}
 
 \textcolor{blue}{The first is to streamline the statistics.
-As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.} ~
-
+As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}\\
+\strut \\
+\textbf{check this}
 We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
 If the statistical modelling aimed at developing the best predictive model, then this would be the right approach.
-However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
-~\\
+However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.\\
+\strut \\
 \textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}\\
@@
 \strut \\
-The key contribution of this paper is to highligt the gap that exists in trying to extrapolate multiple local studies to a global scale.
+The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
 While the impact of forest cover on streamflow is easily hypothesised (e.g.
 Zhou et al.
-(2015); Hoek van Dijke et al.
-(2022)), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+2015; Hoek van Dijke et al.
+2022), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
 In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of forestation or deforestation, and in many cases this becomes a qualitative assessment.\\
-As such, we provide three key insights.\\
-- While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al.
+As such, we provide three key insights.
+
+\begin{itemize}
+\tightlist
+\item
+  While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al.
 (2017), Filoso et al.
@@
 2022), leading to possible wrong policy or management responses.\\
-- Cumulative and average values of change can be misleading, especially when extracted from published field studies which originally had different objectives.
+\item
+  Cumulative and average values of change can be misleading, especially when extracted from published field studies which originally had different objectives.
 This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
@@
 We demonstrate how removing these experiments from the total changes the results of the analysis.
-- In general, the size of the catchment and the length of the study play a huge role in the interpretation of the results.
+\item
+  In general, the size of the catchment and the length of the study play a huge role in the interpretation of the results.
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
@@
   \strut \\
+\end{itemize}
 
@@
 If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}\\
-We acknowledge this, and we hope that the current revised version and our responses adress these concerns.\\
+We acknowledge this, and we hope that the current revised version and our responses address these concerns.\\
 \strut \\
@@
 Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}\\
-We acknowledge this, and we hope that the current revised version and our responses adress these concerns.\\
+We acknowledge this, and we hope that the current revised version and our responses address these concerns.\\
 
 \hypertarget{associate-editor}{%
-\subsection{Associate Editor:}\label{associate-editor}}
+\section{Associate Editor:}\label{associate-editor}}
 
 \textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields.
-The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}
-
+The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}\\
+\strut \\
+Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.\\
+\strut \\
 \textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
@@
 It is also good practice to check the relevance of outliers (of data with high VIF) and set them aside before model fitting.
-It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}
-
+It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}\\
+\strut \\
+\#\#\# error distributions
 Thank you for raising these important points in relation to the validity of the statistical model.
-A first point that arises from this is that we clarify better what the aim of our statistical modelling is.
-
-There are in essence two ``modes'' in relation statistical modelling.
+A first point that arises from this is that we clarify better what the aim of our statistical modelling is.\\
+There are in essence two approaches to statistical modelling.
 Generally a model is developed to be used in predictive mode: using a model to predict unknown values, either within or beyond the current data set (forecasting).
 In this case the model should be reduced to its most efficient version that minimises the bias - variance trade-off.
-In this case, automatic variable selection and potentially validation on independent data are important, as the aim is to develop the most robust model for prediction.
+Automatic variable selection and potentially validation on independent data are therefore important, as the aim is to develop the most robust model for prediction.\\
+However, a second reason for to use a statistical model is to explain the maximum variance in the data.
+In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
+This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in Zhang et al.
+(2017)) or jointly (as in our approach).
+In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from adding each variable.\\
+This explains why we build the model starting from the most simple model, rather than starting from the most complex model.
 
-However, a second reason for fitting a statistical model is purely to explain the variance in the data.
-In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
-This is subsequently followed by a step by step analysis to test the different causal relationship, either as single variables (as in Zhang et al.
-(2017)) or jointly (as in our approach).
-In this case there is no attempt to find the best ``predictive model'', the focus is on the additional amount of explained variance from adding each variable.
-
-\textbf{Include a section in the paper to explain this clearly}
+\hfill\break
+\textbf{Include a section in the paper to explain this clearly}\\
+\strut \\
 
 Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript.
-Including such diagnostic plots for single variable regression is often omitted.
+In many cases, including such diagnostic plots for single variable regressions is often omitted.
 For example, both Zhang et al.
 (2017) and Filoso et al.
-(2017) do not present any diagnostics for their relationships and simply looking at their regressions qualitatively indicates issues with the residuals.
+(2017) do not present any diagnostics for their relationships and the regressions (Fig 2 - Fig 4 in Zhang et al.
+(2017) and Fig 9 in Filoso et al.
+(2017)) qualitatively indicate issues with the residuals.
 
-In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables.
+In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, or explaining why we did not take any further steps.
+
+\hypertarget{the-issue-of-outliers}{%
+\subsubsection{the issue of outliers}\label{the-issue-of-outliers}}
 
@@
 (2013).
-This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:\\
-- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by Zhang et al.
+This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:
+
+\begin{itemize}
+\tightlist
+\item
+  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by Zhang et al.
 (2017) is clearly wrong.
@@
 corrected to 16\%
-- 3, Baker Creek, Zhang and Wei, 2012.
+\item
+  3, Baker Creek, Zhang and Wei, 2012.
 The original recorde 201.1\% change by Zhang et al.
@@
 corrected\\
-- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in Zhang et al.
+\item
+  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in Zhang et al.
 (2017).
 This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''.
-As a result the change in forest cover was changed to -100\% rather than +100\%.
-- 210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
-- 213, 214 and 215, Monda 1, 2 and 3.
+As a result the change in forest cover was changed to -100\% rather than +100\%.\\
+\item
+  210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
+\item
+  213, 214 and 215, Monda 1, 2 and 3.
 These catchments are tricky.
@@
 In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).\\
-- 230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.\\
-- 312, Yerraminup South.
+\item
+  230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.\\
+\item
+  312, Yerraminup South.
 The original publication for this catchment is a Western Australian Water Authority report from 1987, which is hard to find, but we have added a copy in the ``Papers'' folder on github.
@@
 Changed the sign of the change in forest cover: -60\%.\\
-- 72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001).
+\item
+  72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001).
 In the database from Zhang et al.
@@
 Corrected for all three catchments.\\
-- 78, Black Spur 1, the treatments and effects are only reported in a conference paper (Jayasuriya and O'Shaughnessy 1988) and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in Zhang et al.
+\item
+  78, Black Spur 1, the treatments and effects are only reported in a conference paper (Jayasuriya and O'Shaughnessy 1988) and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in Zhang et al.
 (2017)).
@@
 Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.\\
-- 104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
-- 102, Cold Spring.
+\item
+  104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
+\item
+  102, Cold Spring.
 Checking the original paper (Schneider and Ayer 1961) indicates that this is in fact a reduction in flow as a result of reforestation.
 Changed the sign of Delta\_Q\_f to be negative.\\
-- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
+\item
+  97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
 These are all catchments in Puerto Rico from the study from Beck et al.
@@
 (2013), which is a thorough and excellent modelling study.\\
-- 85 Bosboukloof.
+\item
+  85 Bosboukloof.
 This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
 In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.\\
-- 259 Shackam Brook.
+\item
+  259 Shackam Brook.
 There were a few issues with this catchment in the original database.
@@
 Corrected Delta\_Qf\_perc to -20.7\%.\\
-- 95 Sage Brook. Similar to 259 and 102, originates from Schneider and Ayer (1961). Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
-- 101 Coalburn.
+\item
+  95 Sage Brook. Similar to 259 and 102, originates from Schneider and Ayer (1961). Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
+\item
+  101 Coalburn.
 Original publication (Robinson, 1993) which is a symposium paper, is not available.
@@
 Changed to -20.3\%.
+\end{itemize}
 
 A VIF analysis is also a great suggestion as this will indicate variables which are highly correlated within the dataset, such as the Precipitation and Dryness (which we already indicate).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{require}\NormalTok{(tidyverse)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: tidyverse
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'tidyverse' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{verbatim}
-## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
-\end{verbatim}
-
-\begin{verbatim}
-## v ggplot2 3.3.6     v purrr   0.3.4
-## v tibble  3.1.6     v dplyr   1.0.9
-## v tidyr   1.2.0     v stringr 1.4.0
-## v readr   2.1.1     v forcats 0.5.1
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'ggplot2' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'tidyr' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: package 'dplyr' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{verbatim}
-## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
-## x dplyr::filter() masks stats::filter()
-## x dplyr::lag()    masks stats::lag()
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{require}\NormalTok{(lubridate)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: lubridate
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'lubridate'
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:base':
-## 
-##     date, intersect, setdiff, union
-\end{verbatim}
 
@@
 
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{./testcorr_plot} \caption{Correlation matrix for all variables}\label{fig:corgraphs}
+\end{figure}
+
 So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.
@@
 A full blown model selection would also allow to retain or discard the interaction terms, which could be important (as also recognized by the authors; Section 4.5) and should not be discarded a priori.
-Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}
-
-As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data
+Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}\\
+\strut \\
+As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data.\\
+\strut \\
 
@@
 As it looks now, the manuscript could be easily considered somewhat confirmatory, with respect to most data, approaches and conclusions reached by Zhang et al 2017 and Filoso et al 2017.
-Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}
-
-This is a good point and we have rewritten the paper to strengthen this point.
+Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}\\
+\strut \\
+This is a good point and we have rewritten the paper to strengthen this point.\\
+\strut \\
 

### 2022-07-29 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
first draft response review completed

@@
 date: "`r Sys.Date()`"
+linenumbers: true
+#numbersections: true
+csl: elsevier-harvard.csl
 output:   
@@
     base_format: rticles::elsevier_article
-bibliography: forest_and_water.bib
+bibliography: forestandwater.bib
+editor_options: 
+  chunk_output_type: inline
+header-includes:
+  - \usepackage{setspace}
+  - \usepackage{color}
 ---
@@
 
-This document records the replies to the reviewer for the first submission of *"Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets"* to Journal of Hydrology, which was rejected before review.
+This document records the replies to the reviewer for the first submission of *Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets* to Journal of Hydrology, which was rejected before review.
+
 \
-Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text.  
+Reviewer comments in \textcolor{blue}{blue} and our responses in normal text.  
 \
 **The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.**
-\
 
@@
 
+## Comment AE 1
+
 \textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields.
@@
 Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.
-\
+
+## Comment AE 2
+
 \textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
@@
 \
-### error distributions
+Our answer to this point from the associate editor is quite long and covers the following sub topics:
+
+- Error distributions  
+- The issue of outliers  
+- VIF analysis and understanding cross correlations between the variables
+
+### Error distributions
+
 <!-- answer -->
@@
 In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, or explaining why we did not take any further steps.
+Furthermore, the residual distributions are mostly well-behaved, it is only in the tails of the distribution (very high flow changes and very small flow changes) where the residual distribution diverges from normal.
+As the change in flow variable covers $\mathbb{R}$, we cannot use a log transformation on the predictant, which is the usual solution for such residual distributions, especially in Hydrology.
+We therefore chose not to transform and discuss the issue with the fat tails of the distribution in the discussion.
 
-### the issue of outliers
+### The issue of outliers
 
-**Check Venables and Ripley p119**
+We agree that outliers could affect the observed residual distribution as this would most likley be obvious in the tails of the distribution. As @venables2013modern outline on p119:
 "Outliers are sample values that cause surprise in relation to the majority of the sample.
-This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors."
-Because our modelling is focused on explaining the maximum variance, we believe that excluding values that are outliers is probably not a good idea.  
+This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors." As a result, we believe that excluding values that are outliers is probably not a good idea.
 \
@@
 
-A final issue was the inclusion of the results of several catchments, for example from the study by @beck2013, which had no significant change in flow.
-Despite this, the "average" change in flow was reported int he database.
+A further issue was the inclusion of the results of several catchments, for example from the study by @beck2013, which had no significant change in flow.
+Despite this, the "average" change in flow was reported in the database.
 We don't believe that this is correct and the results from such studies should be set to 0.
@@
 Values for all 12 studied catchments set to 0 in the database.
-- 188 Kimakia
-- 221	N. Creek, Babinda, Queensland
-- 254 Sambret
+- 188 Kimakia.
+and 254 Sambret.
+The data in the database from @zhang2017 appear to originate from @bruijnzeel1990 which gives 3 values for different lengths of studies.
+However, the values in the original study by @blackie1979kimakia and @blackie1979kericho do not seem to add up to the same values, and the specific values are not mentioned in the actual papers.
+In addition, as @bruijnzeel1990 mentions in the footnotes, the control for Kimakia is a bamboo catchment, while the control for Sambret is a tea plantation.
+Overall, this suggests that the data are probably not a clear deforestation/reforestation study and should be discarded from the analysis.
+- 221	N. Creek, Babinda, Queensland. The original paper from this study highlights that the differences between the catchments were insignificant.
 
-There were similar studies that we identified that
+### VIF analysis and understanding cross correlations between the variables
 
-## VIF analysis and understanding cross correlations between the variables
+We agree that a VIF analysis can be important to identify high correlation between variables.
+As we have already indicated, we consider Dryness and Precipitation to be highly correlated, and therefore we don't include both in the model.
+However, a VIF analysis is only useful when all the data are numeric, and in this case several of the variables are factors and not numeric.
 
-A VIF analysis is also a great suggestion as this will indicate variables which are highly correlated within the dataset, such as the Precipitation and Dryness (which we already indicate).
+```{r corgraphs, echo=F, out.width="90%", fig.cap="Correlation matrix for all variables"}
 
-```{r}
+include_graphics("./testcorr_plot.png")
+```
+
+As an alternative we created a correlation plot (Figure \@ref(fig:corgraphs)).
+This clearly shows the obvious correlations, but also shows that in general cross correlation is quite low between variables.
+Some interesting relationships, some of which were also highlighted in our models, appear in this graph:
+
+- the negative relationship between log10(Area) and change in forest area (DeltaF_perc_pos) indicating that in the data larger catchment tended to have (obviously) smaller areas of forest change.  
+- the weak positive relationship between log10(Area) and the assessment method using hydrological models.
+This is also obvious as it would be impossible to perform paired catchment studies at very large scales.
+- As we also indicate in the models, there is a strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
+There is of course no causal effect of the assessment methods, it is simply that smaller catchments used in paired watershed assessments are easier to fully clear or fully replant.
+
+Overall this analysis shows very clearly the danger of simply investigating single variable regressions as was done in @zhang2017 and @filoso2017 or even using simple modelling as in @jackson2005 and @zhou2015.
+It is too easy to miss the latent variables that are the underlying factors and influence the model results.
+
+```{r correlation, eval=F, echo = F}
 require(mgcv)
@@
 
-```{r corgraphs, echo=F, out.width="90%", fig.cap="Correlation matrix for all variables"}
-
-include_graphics("./testcorr_plot.png")
-```
-So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.
-
-However, being too focused on just the residual distribution can also lead to over fitting or over smoothing of the variables. This means that important characteristics of the data might be lost.
+## Comment AE 3
 
@@
 <!-- answer -->
-As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data.   
+The Associate editor raises several points that need further discussion.
+
+- a formalised model selection;  
+- the use of AIC rather than r^2^ for model selection;  
+- interaction terms.
+
+### A formalised model selection
+
+We agree that a classical statistical approach would involve a formal model selection.
+We acknowledge that in the manuscript we ended up mixing two approaches, where we should have stayed with a single approach.
+As outlined earlier, the current focus of the statistical modelling is on understanding the different covariates that explain the variation in the data, and to identify latent variables that cause the apparent relationships in the data.
+This means that there is no attempt to do a formal model selection, as we are not seeking the best predictive model.
+
+However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach.
+**FIX THIS UP IN THE PAPER and insert the relevant page and line numbers.**
+
+### AIC rather than r^2^
+
+We actually used the adjusted r^2^ in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
+However, as the GAM models also provide an AIC, we have for consistency changed our model performance measure to the AIC.
+
+### Interaction terms
+
+The issue of interaction terms is a tricky one.
+As we indicated in the manuscript, we did not include most of the interaction terms as it became a guessing game.
+While there clearly is cross correlation between the variables and there is potential interaction between terms, the question of clear causality remains unanswered.
+We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.
 \
+
+## Comment AE 4
 
@@
 <!-- answer -->
-This is a good point and we have rewritten the paper to strengthen this point.  
+We agree that this is a valid point and a weakness in the original manuscript.
+As we also outline in our reply to the Editor, we have rewritten the paper to strengthen the novelty of the work.
+In particular the paper now more directly focuses on the difficulties in analysing aggregated global data and the importance of latent variables.
+We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.
 \
+
+## Comment AE 5
 
@@
 These are not big issues per se but are nonetheless distracting.}
+\
+<!-- answer -->
+Thank you for pointing this out, we have reviewed the paper carefully and corrected these minor issues.
+
+## Comment AE 6
 
@@
 
+<!-- answer -->
+We agree with the AE that this would be an interesting idea and links back to our point about "latent variables".
+The excellent paper by @levy2018 includes \% forest area as a variable, but their data cannot be incorporated into the current paper.
+This is because their analysis focuses on a time progression rather than a "before and after" or a clear "control" and "treatment".
+Furthermore, extracting the total area of forest from the papers is not a trivial job, and the area is also not always mentioned in the papers.
+Finally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
+It is definitely an area of future research.
+
+## Comment AE 7
+
 \textcolor{blue}{- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode.
 While it is important to present also the r2, even a model with low r2 square we learn which factors significantly affect the change in streamflow and which do not do so.}
+
+<!--answer -->
+We totally agree with the AE, and this might be more a matter of symantics.
+We thought we had the same interpretation as the AE, but might have worded this correctly.
+The above comment from the AE actually points exactly to the way we are using models in the paper.
+We use the models to look at factors that significantly affect streamflow rather than looking at a predictive model.
+We have reviewed the text in the conclusion and have reworded to make sure we capture the above suggested meaning.

### 2022-07-29 paper/Forest_and_Water/Reply_first_review_JoH.tex
first draft response review completed

@@
 %%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+
 \usepackage[hyphens]{url}
@@
 \usepackage{lineno} % add
+  \linenumbers % turns line numbering on
 
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\bibliographystyle{elsarticle-harv}
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
+
 \ifxetex
@@
             pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
 
@@
 % Pandoc toggle for numbering sections (defaults to be off)
-
-% Pandoc syntax highlighting
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
@@
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{setspace}
+\usepackage{color}
 
@@
   \title{Replies to editor and associate editor Journal of Hydrology}
-    \author[]{}
+    \author[]{%
+  %
+  }
+
+      \cortext[cor1]{Corresponding author}
 
@@
 
-This document records the replies to the reviewer for the first submission of \emph{``Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets''} to Journal of Hydrology, which was rejected before review.\\
+This document records the replies to the reviewer for the first submission of \emph{Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets} to Journal of Hydrology, which was rejected before review.
+
+\hfill\break
+Reviewer comments in \textcolor{blue}{blue} and our responses in normal text.\\
 \strut \\
-Reviewer comments in \textcolor{blue}{blue}. Our responses in normal text.\\
-\strut \\
-\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}\\
-\strut \\
+\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}
 
@@
 
-Thank you for the suggestion of the Levy et al.
-(2018) paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change.
+Thank you for the suggestion of the \citet{levy2018} paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change.
 This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
-Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use Levy et al.
-(2018) as a ``verfication'' of our analysis.
+Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use \citet{levy2018} as a ``verfication'' of our analysis.
 
@@
 We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
-However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in Zhou et al.
-(2015); Jackson et al.
-(2005); Filoso et al.
-(2017) and Zhang et al.
-(2017)).\\
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).\\
 \strut \\
@@
 The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
-While the impact of forest cover on streamflow is easily hypothesised (e.g.
-Zhou et al.
-2015; Hoek van Dijke et al.
-2022), our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+While the impact of forest cover on streamflow is easily hypothesised \citep[e.g.][]{zhou2015, hoekvandijke2022}, our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
 In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of forestation or deforestation, and in many cases this becomes a qualitative assessment.\\
@@
 \item
-  While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in Zhang et al.
-(2017), Filoso et al.
-(2017), Zhou et al.
-(2015) and Jackson et al.
-(2005)) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
+  While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in \citet{zhang2017}, \citet{filoso2017}, \citet{zhou2015} and \citet{jackson2005}) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
 In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
@@
 This is without considering the number of errors that existed in the data.
-This is particularly important, since results from these global analyses are used to build further models to analyse global impacts (e.g.
-Hoek van Dijke et al.
-2022), leading to possible wrong policy or management responses.\\
+This is particularly important, since results from these global analyses are used to build further models to analyse global impacts \citep[e.g.][]{hoekvandijke2022}, leading to possible wrong policy or management responses.\\
 \item
@@
 This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
-Many of the Paired Watershed Experiments in Australia (e.g.
-Cornish and Vertessy 2001; Webb 2009; Webb and Kathuria 2012; Webb and Jarrett 2013; Watson et al.
-2001) fall in this category, and therefore easily be classified as either forestation or deforestation.
+Many of the Paired Watershed Experiments in Australia \citep[e.g.][]{cornish2001, webb2009, webb2012, webb2013, watson2001} fall in this category, and therefore easily be classified as either forestation or deforestation.
 In principle a decision needs to be made how many years post clearing needs to be considered, and whether the remainder of the timeseries should be classified as regeneration.
@@
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
-In terms of the catchment size, one of the key issues we originally had with the Zhang et al.
-(2017) paper was the arbitrary split between catchment \textgreater{} 1000 km\textsuperscript{2} and catchment \textless{} 1000 km\textsuperscript{2}.
+In terms of the catchment size, one of the key issues we originally had with the \citet{zhang2017} paper was the arbitrary split between catchment \textgreater{} 1000 km\textsuperscript{2} and catchment \textless{} 1000 km\textsuperscript{2}.
 Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of methods used to analyse small catchments (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
-The paper by Beck et al.
-(2013) is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
-However, as we argue, this results of this paper are misrepresented in the database, as none of the analyse catchments had a significant change in the streamflow.
+The paper by \citet{beck2013} is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
+However, as we argue, the results of this paper are misrepresented in the database, as none of the analysed catchments had a significant change in the streamflow.
 This means the change should be recorded as 0.\\
@@
 
+\hypertarget{comment-ae-1}{%
+\subsection{Comment AE 1}\label{comment-ae-1}}
+
 \textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields.
@@
 \strut \\
-Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.\\
-\strut \\
+Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.
+
+\hypertarget{comment-ae-2}{%
+\subsection{Comment AE 2}\label{comment-ae-2}}
+
 \textcolor{blue}{The main aspects that need to be addressed before the manuscript can be evaluated by experts in the field are listed here. 
@@
 \strut \\
-\#\#\# error distributions
+Our answer to this point from the associate editor is quite long and covers the following sub topics:
+
+\begin{itemize}
+\tightlist
+\item
+  Error distributions\\
+\item
+  The issue of outliers\\
+\item
+  VIF analysis and understanding cross correlations between the variables
+\end{itemize}
+
+\hypertarget{error-distributions}{%
+\subsubsection{Error distributions}\label{error-distributions}}
+
 Thank you for raising these important points in relation to the validity of the statistical model.
@@
 In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
-This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in Zhang et al.
-(2017)) or jointly (as in our approach).
+This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in \citet{zhang2017}) or jointly (as in our approach).
 In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from adding each variable.\\
@@
 In many cases, including such diagnostic plots for single variable regressions is often omitted.
-For example, both Zhang et al.
-(2017) and Filoso et al.
-(2017) do not present any diagnostics for their relationships and the regressions (Fig 2 - Fig 4 in Zhang et al.
-(2017) and Fig 9 in Filoso et al.
-(2017)) qualitatively indicate issues with the residuals.
+For example, both \citet{zhang2017} and \citet{filoso2017} do not present any diagnostics for their relationships and the regressions (Fig 2 - Fig 4 in \citet{zhang2017} and Fig 9 in \citet{filoso2017}) qualitatively indicate issues with the residuals.
 
 In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, or explaining why we did not take any further steps.
+Furthermore, the residual distributions are mostly well-behaved, it is only in the tails of the distribution (very high flow changes and very small flow changes) where the residual distribution diverges from normal.
+As the change in flow variable covers \(\mathbb{R}\), we cannot use a log transformation on the predictant, which is the usual solution for such residual distributions, especially in Hydrology.
+We therefore chose not to transform and discuss the issue with the fat tails of the distribution in the discussion.
 
 \hypertarget{the-issue-of-outliers}{%
-\subsubsection{the issue of outliers}\label{the-issue-of-outliers}}
+\subsubsection{The issue of outliers}\label{the-issue-of-outliers}}
 
-Because our modelling is focused on explaining the maximum variance, excluding values that are outliers is probably not a good idea.
-However, identifying outliers (particularly values with high leverage), and investigating why these might be different from the overall distribution is a great suggestion.
-
-\textbf{Check Venables and Ripley p119}
+We agree that outliers could affect the observed residual distribution as this would most likley be obvious in the tails of the distribution. As \citet{venables2013modern} outline on p119:
 ``Outliers are sample values that cause surprise in relation to the majority of the sample.
-This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors.''
-
-However, another careful review of the data identified many further errors in the data, which were all originally in Zhang et al.
-(2017).
+This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors.'' As a result, we believe that excluding values that are outliers is probably not a good idea.\\
+\strut \\
+However, another careful review of the data identified many further errors in the data, which were all originally in \citet{zhang2017}.
 A particular problem was that many catchments had the wrong sign for the change in forest cover.
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected.
-A final issue was the inclusion of the catchments from the study by Beck et al.
-(2013).
-This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation:
+These were all checked and corrected if needed and a full list of all these changes is below:
 
@@
 \item
-  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by Zhang et al.
-(2017) is clearly wrong.
+  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
@@
   3, Baker Creek, Zhang and Wei, 2012.
-The original recorde 201.1\% change by Zhang et al.
-(2017) is also wrong.
+The original recorde 201.1\% change by \citet{zhang2017} is also wrong.
 Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
@@
 \item
-  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in Zhang et al.
-(2017).
+  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in \citet{zhang2017}.
 This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''.
@@
 These catchments are tricky.
-The original paper (O'Shaughnessy et al.
-1979) only reports on the control period and indicates that the catchments will be cleared.
-The later summary paper (Watson et al.
-2001) shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
+The original paper \citep{oshaughnessy1979} only reports on the control period and indicates that the catchments will be cleared.
+The later summary paper \citep{watson2001} shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
 The further complication is that the treatment included clearing and reseeding and regrowth.
@@
 \item
-  72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by Cornish (1993) and Cornish and Vertessy (2001).
-In the database from Zhang et al.
-(2017), the forest change for all these catchments is positive.
+  72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by \citet{cornish1993} and \citet{cornish2001}.
+In the database from \citet{zhang2017}, the forest change for all these catchments is positive.
 However, the paper highlights that these catchments were all logged and either naturally regenerated or were planted with a plantation species.
@@
 \item
-  78, Black Spur 1, the treatments and effects are only reported in a conference paper (Jayasuriya and O'Shaughnessy 1988) and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in Zhang et al.
-(2017)).
+  78, Black Spur 1, the treatments and effects are only reported in a conference paper \citep{jayasuriya1988} and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in \citet{zhang2017}).
 Corrected.
@@
 \item
-  102, Cold Spring.
-Checking the original paper (Schneider and Ayer 1961) indicates that this is in fact a reduction in flow as a result of reforestation.
-Changed the sign of Delta\_Q\_f to be negative.\\
-\item
-  97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
-These are all catchments in Puerto Rico from the study from Beck et al.
-(2013).
-They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
-The values that are cited in the database should all be set to ``not significant from 0'', so might be included as 0.
-Including them with positive or negative values is misleading.
-This not judging the study by Beck et al.
-(2013), which is a thorough and excellent modelling study.\\
+  102, Cold Spring. Checking the original paper \citep{schneider1961} indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
 \item
@@
 There were a few issues with this catchment in the original database.
-The name was misspelled and it was incorrectly attributed to Brown et al.
-(2005).
-The original paper is the same as 102 (Schneider and Ayer 1961).
+The name was misspelled and it was incorrectly attributed to \citet{brown2005}.
+The original paper is the same as 102 \citep{schneider1961}.
 Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases.
@@
 \item
-  95 Sage Brook. Similar to 259 and 102, originates from Schneider and Ayer (1961). Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
+  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
 \item
@@
 Original publication (Robinson, 1993) which is a symposium paper, is not available.
-The best summary of the reseearch is in Birkinshaw, Bathurst, and Robinson (2014) which summaries 45 years of resaerch in the Coalburn catchment.
+The best summary of the research is in \citet{birkinshaw2014} which summaries 45 years of resaerch in the Coalburn catchment.
 It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period.
@@
 
-A VIF analysis is also a great suggestion as this will indicate variables which are highly correlated within the dataset, such as the Precipitation and Dryness (which we already indicate).
+A further issue was the inclusion of the results of several catchments, for example from the study by \citet{beck2013}, which had no significant change in flow.
+Despite this, the ``average'' change in flow was reported in the database.
+We don't believe that this is correct and the results from such studies should be set to 0.
+A full list of changes is provided below:
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{require}\NormalTok{(mgcv)}
-\end{Highlighting}
-\end{Shaded}
+\begin{itemize}
+\tightlist
+\item
+  97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati.
+These are all catchments in Puerto Rico from the study from \citet{beck2013}.
+They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation.
+The values that are cited in the database should all be set to ``not significant from 0'', so might be included as 0.
+Including them with positive or negative values is misleading.
+This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation.
+Values for all 12 studied catchments set to 0 in the database.\\
+\item
+  188 Kimakia.
+and 254 Sambret.
+The data in the database from \citet{zhang2017} appear to originate from \citet{bruijnzeel1990} which gives 3 values for different lengths of studies.
+However, the values in the original study by \citet{blackie1979kimakia} and \citet{blackie1979kericho} do not seem to add up to the same values, and the specific values are not mentioned in the actual papers.
+In addition, as \citet{bruijnzeel1990} mentions in the footnotes, the control for Kimakia is a bamboo catchment, while the control for Sambret is a tea plantation.
+Overall, this suggests that the data are probably not a clear deforestation/reforestation study and should be discarded from the analysis.
+\item
+  221 N. Creek, Babinda, Queensland. The original paper from this study highlights that the differences between the catchments were insignificant.
+\end{itemize}
 
-\begin{verbatim}
-## Loading required package: mgcv
-\end{verbatim}
+\hypertarget{vif-analysis-and-understanding-cross-correlations-between-the-variables}{%
+\subsubsection{VIF analysis and understanding cross correlations between the variables}\label{vif-analysis-and-understanding-cross-correlations-between-the-variables}}
 
-\begin{verbatim}
-## Warning: package 'mgcv' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: nlme
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'nlme'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:dplyr':
-## 
-##     collapse
-\end{verbatim}
-
-\begin{verbatim}
-## This is mgcv 1.8-40. For overview type 'help("mgcv-package")'.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{require}\NormalTok{(car)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: car
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: carData
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'car'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:dplyr':
-## 
-##     recode
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:purrr':
-## 
-##     some
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{All\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"All\_data.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Rows: 327 Columns: 24
-\end{verbatim}
-
-\begin{verbatim}
-## -- Column specification --------------------------------------------------------
-## Delimiter: ","
-## chr (10): Watershed name, Forest_type, Hydrological_regime, Precip_data_type...
-## dbl (14): Watershed #, Area_km2, Pa_mm, DeltaF_perc, DeltaQf_perc, E0, Latit...
-## 
-## i Use `spec()` to retrieve the full column specification for this data.
-## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{All\_data2 }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"All\_data2.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Rows: 327 Columns: 26
-## -- Column specification --------------------------------------------------------
-## Delimiter: ","
-## chr (11): Watershed name, Forest_type, Hydrological_regime, Precip_data_type...
-## dbl (15): Watershed #, Area_km2, Pa_mm, DeltaF_perc, DeltaQf_perc, E0, Latit...
-## 
-## i Use `spec()` to retrieve the full column specification for this data.
-## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data2 }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{length =}\NormalTok{ To }\SpecialCharTok{{-}}\NormalTok{ From,}
-         \AttributeTok{mid\_year =}\NormalTok{ From }\SpecialCharTok{+}\NormalTok{ (To }\SpecialCharTok{{-}}\NormalTok{ From)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
-
-\CommentTok{\#Maybe first simply}
-\CommentTok{\# https://stackoverflow.com/questions/52554336/plot{-}the{-}equivalent{-}of{-}correlation{-}matrix{-}for{-}factors{-}categorical{-}data{-}and{-}mi}
-\FunctionTok{library}\NormalTok{(ggcorrplot)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: package 'ggcorrplot' was built under R version 4.1.3
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{png}\NormalTok{(}\StringTok{"testcorr\_plot.png"}\NormalTok{, }\AttributeTok{width =} \DecValTok{960}\NormalTok{, }\AttributeTok{height =} \DecValTok{960}\NormalTok{)}
-\FunctionTok{model.matrix}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\DecValTok{0}\SpecialCharTok{+}\NormalTok{DeltaF\_perc\_pos }\SpecialCharTok{+} 
-\NormalTok{                    Forest\_Sign }\SpecialCharTok{+} 
-                   \FunctionTok{log10}\NormalTok{(Area\_km2) }\SpecialCharTok{+} 
-\NormalTok{                    Dryness }\SpecialCharTok{+} 
-\NormalTok{                    length }\SpecialCharTok{+}
-\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
-\NormalTok{                    Forest\_type }\SpecialCharTok{+}
-\NormalTok{                    Hydrological\_regime, }\AttributeTok{data=}\NormalTok{All\_data2) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{cor}\NormalTok{(}\AttributeTok{use=}\StringTok{"pairwise.complete.obs"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{ggcorrplot}\NormalTok{(}\AttributeTok{show.diag =}\NormalTok{ F, }\AttributeTok{type=}\StringTok{"lower"}\NormalTok{, }\AttributeTok{lab=}\ConstantTok{TRUE}\NormalTok{, }\AttributeTok{lab\_size=}\DecValTok{4}\NormalTok{)}
-\FunctionTok{dev.off}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# model6\_all \textless{}{-} gam(DeltaQf\_perc \textasciitilde{} DeltaF\_perc\_pos + }
-\CommentTok{\#                     Forest\_Sign + }
-\CommentTok{\#                    log10(Area\_km2) + }
-\CommentTok{\#                     Dryness + }
-\CommentTok{\#                     length +\# s(From, k = 40, bs = "ts") +}
-\CommentTok{\#                     Precip\_data\_type +  Assessment\_technique +}
-\CommentTok{\#                     Forest\_type +}
-\CommentTok{\#                     Hydrological\_regime}
-\CommentTok{\#                     , data = All\_data2)}
-\CommentTok{\# summary(model6\_all)}
-\CommentTok{\# gam.check(model6\_all)}
-\CommentTok{\# \#plot(model6\_all)}
-\CommentTok{\# inf1 = influence.gam(model6\_all)}
-\CommentTok{\# hist(inf1)}
-\CommentTok{\# hist(cooks.distance(model6\_all))}
-\CommentTok{\# which.max(cooks.distance(model6\_all))}
-\CommentTok{\# }
-\CommentTok{\# regclass::VIF(model6\_all)}
-\CommentTok{\# }
-\CommentTok{\# \# Can only do car::vif on linear model and only on numerical values}
-\CommentTok{\# car::vif(gam(DeltaQf\_perc \textasciitilde{} DeltaF\_perc\_pos + Forest\_Sign + }
-\CommentTok{\#                     log10(Area\_km2) + }
-\CommentTok{\#                     Dryness + }
-\CommentTok{\#                     length, data = All\_data2))}
-\end{Highlighting}
-\end{Shaded}
+We agree that a VIF analysis can be important to identify high correlation between variables.
+As we have already indicated, we consider Dryness and Precipitation to be highly correlated, and therefore we don't include both in the model.
+However, a VIF analysis is only useful when all the data are numeric, and in this case several of the variables are factors and not numeric.
 
@@
 
-So the VIF really does not work, only works for numerical values and not for factor values. So this suggestion is not really relevant either.
+As an alternative we created a correlation plot (Figure \ref{fig:corgraphs}).
+This clearly shows the obvious correlations, but also shows that in general cross correlation is quite low between variables.
+Some interesting relationships, some of which were also highlighted in our models, appear in this graph:
 
-However, being too focused on just the residual distribution can also lead to over fitting or over smoothing of the variables. This means that important characteristics of the data might be lost.
+\begin{itemize}
+\tightlist
+\item
+  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc\_pos) indicating that in the data larger catchment tended to have (obviously) smaller areas of forest change.\\
+\item
+  the weak positive relationship between log10(Area) and the assessment method using hydrological models.
+This is also obvious as it would be impossible to perform paired catchment studies at very large scales.\\
+\item
+  As we also indicate in the models, there is a strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
+There is of course no causal effect of the assessment methods, it is simply that smaller catchments used in paired watershed assessments are easier to fully clear or fully replant.
+\end{itemize}
+
+Overall this analysis shows very clearly the danger of simply investigating single variable regressions as was done in \citet{zhang2017} and \citet{filoso2017} or even using simple modelling as in \citet{jackson2005} and \citet{zhou2015}.
+It is too easy to miss the latent variables that are the underlying factors and influence the model results.
+
+\hypertarget{comment-ae-3}{%
+\subsection{Comment AE 3}\label{comment-ae-3}}
 
@@
 \strut \\
-As outlined in the last part of the reply, the focus of the statistical modelling in the paper is on understanding the different covariates that can explain the variation in the data.\\
-\strut \\
+The Associate editor raises several points that need further discussion.
+
+\begin{itemize}
+\tightlist
+\item
+  a formalised model selection;\\
+\item
+  the use of AIC rather than r\textsuperscript{2} for model selection;\\
+\item
+  interaction terms.
+\end{itemize}
+
+\hypertarget{a-formalised-model-selection}{%
+\subsubsection{A formalised model selection}\label{a-formalised-model-selection}}
+
+We agree that a classical statistical approach would involve a formal model selection.
+We acknowledge that in the manuscript we ended up mixing two approaches, where we should have stayed with a single approach.
+As outlined earlier, the current focus of the statistical modelling is on understanding the different covariates that explain the variation in the data, and to identify latent variables that cause the apparent relationships in the data.
+This means that there is no attempt to do a formal model selection, as we are not seeking the best predictive model.
+
+However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach.
+\textbf{FIX THIS UP IN THE PAPER and insert the relevant page and line numbers.}
+
+\hypertarget{aic-rather-than-r2}{%
+\subsubsection{\texorpdfstring{AIC rather than r\textsuperscript{2}}{AIC rather than r2}}\label{aic-rather-than-r2}}
+
+We actually used the adjusted r\textsuperscript{2} in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
+However, as the GAM models also provide an AIC, we have for consistency changed our model performance measure to the AIC.
+
+\hypertarget{interaction-terms}{%
+\subsubsection{Interaction terms}\label{interaction-terms}}
+
+The issue of interaction terms is a tricky one.
+As we indicated in the manuscript, we did not include most of the interaction terms as it became a guessing game.
+While there clearly is cross correlation between the variables and there is potential interaction between terms, the question of clear causality remains unanswered.
+We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.\\
+
+\hypertarget{comment-ae-4}{%
+\subsection{Comment AE 4}\label{comment-ae-4}}
 
@@
 \strut \\
-This is a good point and we have rewritten the paper to strengthen this point.\\
+We agree that this is a valid point and a weakness in the original manuscript.
+As we also outline in our reply to the Editor, we have rewritten the paper to strengthen the novelty of the work.
+In particular the paper now more directly focuses on the difficulties in analysing aggregated global data and the importance of latent variables.
+We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.\\
 \strut \\
+
+\hypertarget{comment-ae-5}{%
+\subsection{Comment AE 5}\label{comment-ae-5}}
 
@@
 Also: how is Table 1 used?
-These are not big issues per se but are nonetheless distracting.}
+These are not big issues per se but are nonetheless distracting.}\\
+Thank you for pointing this out, we have reviewed the paper carefully and corrected these minor issues.
+
+\hypertarget{comment-ae-6}{%
+\subsection{Comment AE 6}\label{comment-ae-6}}
 
@@
 
+We agree with the AE that this would be an interesting idea and links back to our point about ``latent variables''.
+The excellent paper by \citet{levy2018} includes \% forest area as a variable, but their data cannot be incorporated into the current paper.
+This is because their analysis focuses on a time progression rather than a ``before and after'' or a clear ``control'' and ``treatment''.
+Furthermore, extracting the total area of forest from the papers is not a trivial job, and the area is also not always mentioned in the papers.
+Finally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
+It is definitely an area of future research.
+
+\hypertarget{comment-ae-7}{%
+\subsection{Comment AE 7}\label{comment-ae-7}}
+
 \textcolor{blue}{- The fact that the explanatory power is low (low r2) does not necessarily make the results uninteresting (against conclusion on L 530), simply it suggests there are other factors, not included in the model, which have a large effect, and that the model presented cannot be used in a predictive mode.
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-beck2013}{}}%
-Beck, H.
-E., L.
-A.
-Bruijnzeel, A.
-I.
-J.
-M.
-van Dijk, T.
-R.
-McVicar, F.
-N.
-Scatena, and J.
-Schellekens.
-2013.
-{``The Impact of Forest Regeneration on Streamflow in 12 Mesoscale Humid Tropical Catchments.''} Journal Article.
-\emph{Hydrol.
-Earth Syst.
-Sci.} 17 (7): 2613--35.
-\url{https://doi.org/10.5194/hess-17-2613-2013}.
+We totally agree with the AE, and this might be more a matter of symantics.
+We thought we had the same interpretation as the AE, but might have worded this correctly.
+The above comment from the AE actually points exactly to the way we are using models in the paper.
+We use the models to look at factors that significantly affect streamflow rather than looking at a predictive model.
+We have reviewed the text in the conclusion and have reworded to make sure we capture the above suggested meaning.
 
-\leavevmode\vadjust pre{\hypertarget{ref-birkinshaw2014}{}}%
-Birkinshaw, Stephen J., James C.
-Bathurst, and Mark Robinson.
-2014.
-{``45 Years of Non-Stationary Hydrology over a Forest Plantation Growth Cycle, Coalburn Catchment, Northern England.''} Journal Article.
-\emph{Journal of Hydrology} 519: 559--73.
-https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2014.07.050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brown2005}{}}%
-Brown, Alice E., Lu Zhang, Thomas A.
-McMahon, Andrew W.
-Western, and Robert A.
-Vertessy.
-2005.
-{``A Review of Paired Catchment Studies for Determining Changes in Water Yield Resulting from Alterations in Vegetation.''} Journal Article.
-\emph{Journal of Hydrology} 310 (1-4): 28--61.
-\url{http://www.sciencedirect.com/science/article/B6V6C-4G05MM9-1/2/bbc5fc0e958a8f34bcb7c1cc7fa57b48}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cornish1993}{}}%
-Cornish, P.
-M.
-1993.
-{``The Effects of Logging and Forest Regeneration on Water Yields in a Moist Eucalypt Forest in New South Wales, Australia.''} Journal Article.
-\emph{Journal of Hydrology} 150 (2-4): 301--22.
-\href{http://www.sciencedirect.com/science/article/B6V6C-487D3Y2-9J/2/73c981ba76284d9d629f6b221d6fd6c6\%20}{http://www.sciencedirect.com/science/article/B6V6C-487D3Y2-9J/2/73c981ba76284d9d629f6b221d6fd6c6 }.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cornish2001}{}}%
-Cornish, P.
-M., and R.
-A.
-Vertessy.
-2001.
-{``Forest Age-Induced Changes in Evapotranspiration and Water Yield in a Eucalypt Forest.''} Journal Article.
-\emph{Journal of Hydrology} 242 (1-2): 43--63.
-\href{http://www.sciencedirect.com/science/article/B6V6C-429910G-3/2/0158b1f89ff436f338a9e688a47f06c4\%20}{http://www.sciencedirect.com/science/article/B6V6C-429910G-3/2/0158b1f89ff436f338a9e688a47f06c4 }.
-
-\leavevmode\vadjust pre{\hypertarget{ref-filoso2017}{}}%
-Filoso, Solange, Maíra Ometto Bezerra, Katherine C.
-B.
-Weiss, and Margaret A.
-Palmer.
-2017.
-{``Impacts of Forest Restoration on Water Yield: A Systematic Review.''} Journal Article.
-\emph{PLOS ONE} 12 (8): e0183210.
-\url{https://doi.org/10.1371/journal.pone.0183210}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hoekvandijke2022}{}}%
-Hoek van Dijke, Anne J., Martin Herold, Kaniska Mallick, Imme Benedict, Miriam Machwitz, Martin Schlerf, Agnes Pranindita, Jolanda J.
-E.
-Theeuwen, Jean-François Bastin, and Adriaan J.
-Teuling.
-2022.
-{``Shifts in Regional Water Availability Due to Global Tree Restoration.''} Journal Article.
-\emph{Nature Geoscience} 15 (5): 363--68.
-\url{https://doi.org/10.1038/s41561-022-00935-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
-Jackson, Robert B., Esteban G.
-Jobbagy, Roni Avissar, Somnath Baidya Roy, Damian J.
-Barrett, Charles W.
-Cook, Kathleen A.
-Farley, David C.
-le Maitre, Bruce A.
-McCarl, and Brian C.
-Murray.
-2005.
-{``Trading Water for Carbon with Biological Carbon Sequestration.''} Journal Article.
-\emph{Science} 310 (5756): 1944--47.
-\url{https://doi.org/10.1126/science.1119282}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jayasuriya1988}{}}%
-Jayasuriya, M.
-D.
-A., and P.
-J.
-O'Shaughnessy.
-1988.
-{``The Use of Mathematical Models in Evaluating Forest Treatment Effects on Streamflow.''} Book Section.
-In \emph{Hydrology and Water Resources Symposium 1988: Preprints of Papers,} 135--39.
-Hydrology; Water Resources Symposium.
-\url{https://doi.org/10.3316/informit.692214289455295}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-levy2018}{}}%
-Levy, M.
-C., A.
-V.
-Lopes, A.
-Cohn, L.
-G.
-Larsen, and S.
-E.
-Thompson.
-2018.
-{``Land Use Change Increases Streamflow Across the Arc of Deforestation in Brazil.''} Journal Article.
-\emph{Geophysical Research Letters} 45 (8): 3520--30.
-https://doi.org/\url{https://doi.org/10.1002/2017GL076526}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oshaughnessy1979}{}}%
-O'Shaughnessy, P.
-J., K.
-J.
-Langford, H.
-P.
-Duncan, and R.
-J.
-Moran.
-1979.
-{``Catchment Experiments in Mountain Ash Forests at North Maroondah.''} Journal Article.
-\emph{Australian Forestry} 42 (3): 150--60.
-\url{https://doi.org/10.1080/00049158.1979.10674220}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schneider1961}{}}%
-Schneider, William Joseph, and Gordon Roundy Ayer. 1961. {``Effect of Reforestation on Streamflow in Central New York.''} Report 1602. \url{https://doi.org/10.3133/wsp1602}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-watson2001}{}}%
-Watson, Fred, Rob Vertessy, Tom McMahon, Bruce Rhodes, and Ian Watson.
-2001.
-{``Improved Methods to Assess Water Yield Changes from Paired-Catchment Studies: Application to the Maroondah Catchments.''} Journal Article.
-\emph{Forest Ecology and Management} 143 (1): 189--204.
-https://doi.org/\url{https://doi.org/10.1016/S0378-1127(00)00517-X}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2009}{}}%
-Webb, Ashley A.
-2009.
-{``Streamflow Response to Pinus Plantation Harvesting: Canobolas State Forest, Southeastern Australia.''} Journal Article.
-\emph{Hydrological Processes} 23 (12): 1679--89.
-https://doi.org/\url{https://doi.org/10.1002/hyp.7301}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2013}{}}%
-Webb, Ashley A., and Brad W.
-Jarrett.
-2013.
-{``Hydrological Response to Wildfire, Integrated Logging and Dry Mixed Species Eucalypt Forest Regeneration: The Yambulla Experiment.''} Journal Article.
-\emph{Forest Ecology and Management} 306: 107--17.
-https://doi.org/\url{https://doi.org/10.1016/j.foreco.2013.06.020}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2012}{}}%
-Webb, Ashley A., and Amrit Kathuria.
-2012.
-{``Response of Streamflow to Afforestation and Thinning at Red Hill, Murray Darling Basin, Australia.''} Journal Article.
-\emph{Journal of Hydrology} 412-413: 133--40.
-https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2011.05.033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhang2017}{}}%
-Zhang, Mingfang, Ning Liu, Richard Harper, Qiang Li, Kuan Liu, Xiaohua Wei, Dingyuan Ning, Yiping Hou, and Shirong Liu.
-2017.
-{``A Global Review on Hydrological Responses to Forest Change Across Multiple Spatial Scales: Importance of Scale, Climate, Forest Type and Hydrological Regime.''} Journal Article.
-\emph{Journal of Hydrology} 546: 44--59.
-https://doi.org/\url{https://doi.org/10.1016/j.jhydrol.2016.12.040}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhou2015}{}}%
-Zhou, Guoyi, Xiaohua Wei, Xiuzhi Chen, Ping Zhou, Xiaodong Liu, Yin Xiao, Ge Sun, et al.
-2015.
-{``Global Pattern for the Effect of Climate and Land Cover on Water Yield.''} Journal Article.
-\emph{Nature Communications} 6 (1): 5918.
-\url{https://doi.org/10.1038/ncomms6918}.
-
-\end{CSLReferences}
+\bibliography{forestandwater.bib}
 

### 2022-07-29 paper/Forest_and_Water/Forest_and_Water.Rmd
first draft response review completed

@@
 Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
+However, this change is is potentially modulated by the area under consideration (affecting the length of the flow paths @zhou2015), the length of the study (c.f.
 @jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).

### 2022-07-29 paper/Forest_and_Water/Forest_and_Water.tex
first draft response review completed

@@
 However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchments after deforestation.
+In contrast, a more recent study in Brazil across 324 catchments, Levy et al.
+(2018) found a significant increase in streamflow, particular in the dry season, showing also quite a dramatic variation in responses to deforestation.
+Understanding the impact of
 
@@
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
+The more regionally concentrated and detailed study by Levy et al.
+(2018) is a further independent dataset with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover would need to be derived again.
 
@@
 
+In contrast to the basic linear regression in the earlier studies (Filoso et al., 2017; Zhang et al., 2017) and the top-down Budyko modelling (Hoek van Dijke et al., 2022; Zhou et al., 2015), the regional Brazilian Cerrado study (Levy et al., 2018) provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+The analysis specifically accounted for differences between catchments and differences due to variations in climate
+
 Encouraged by the work from Zhang et al.
@@
 (2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths Zhou et al.
+However, this change is is potentially modulated by the area under consideration (affecting the length of the flow paths Zhou et al.
 (2015)), the length of the study (c.f.
@@
 \begin{longtable}[]{@{}lll@{}}
-\toprule
+\toprule()
 Factor & Abbreviation & Definition \\
-\midrule
+\midrule()
 \endhead
@@
 & & and hydrographs \\
-\bottomrule
+\bottomrule()
 \end{longtable}
@@
 Clearly there is no major change arising from the additional data.}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 6.24 & 5.19 & 1.2 & 0.23 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.71 & 0 \\
-\textbf{Forest\_SignIncrease} & -26.95 & 5.4 & -4.99 & 0 \\
-\textbf{(Intercept)} & 7.79 & 5.01 & 1.56 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.43 & 0.08 & 5.64 & 0 \\
-\textbf{Forest\_SignIncrease} & -34.65 & 4.95 & -7 & 0 \\
-\bottomrule
+\textbf{(Intercept)} & 13.56 & 4.99 & 2.72 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.36 & 0.08 & 4.59 & 0 \\
+\textbf{Forest\_SignIncrease} & -43.11 & 5.46 & -7.89 & 0 \\
+\textbf{(Intercept)} & 15.14 & 4.76 & 3.18 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.07 & 4.48 & 0 \\
+\textbf{Forest\_SignIncrease} & -48.43 & 4.85 & -9.98 & 0 \\
+\bottomrule()
 \end{longtable}
 
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.24, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.32, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 \caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 9.88 & 5.42 & 1.82 & 0.07 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.68 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.72 & 5.21 & -6.86 & 0 \\
-\textbf{Latitude} & -0.1 & 0.09 & -1.17 & 0.24 \\
-\textbf{Longitude} & -0.02 & 0.03 & -0.62 & 0.54 \\
-\bottomrule
+\textbf{(Intercept)} & 16.53 & 5.04 & 3.28 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.34 & 0.07 & 4.56 & 0 \\
+\textbf{Forest\_SignIncrease} & -48.41 & 4.92 & -9.84 & 0 \\
+\textbf{Latitude} & -0.09 & 0.08 & -1.05 & 0.29 \\
+\textbf{Longitude} & -0.03 & 0.03 & -1.07 & 0.28 \\
+\bottomrule()
 \end{longtable}
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.24 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.33 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 \caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 4.2 & 5.86 & 0.72 & 0.47 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.7 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.37 & 4.97 & -7.12 & 0 \\
-\textbf{Dryness} & 2.76 & 2.5 & 1.11 & 0.27 \\
-\bottomrule
+\textbf{(Intercept)} & 9.49 & 5.47 & 1.73 & 0.08 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.07 & 4.47 & 0 \\
+\textbf{Forest\_SignIncrease} & -49.86 & 4.87 & -10.23 & 0 \\
+\textbf{Dryness} & 4.66 & 2.36 & 1.97 & 0.05 \\
+\bottomrule()
 \end{longtable}
@@
 This might seem suprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.27).
+In this case, the evidence is highly doubtful (p = 0.05).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.24.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.33.
 As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
@@
 \caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endfirsthead
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
@@
 90 & 36.4 & -120.4 & Cantua \\
+225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, A \\
 226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
 U.S.A, C \\
-225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
-U.S.A, A \\
 295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
 356 & -25.75 & 28.23 & Queens river \\
-\bottomrule
+\bottomrule()
 \end{longtable}
@@
 \caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 11.15 & 6.73 & 1.66 & 0.1 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.09 & 3.49 & 0 \\
-\textbf{Forest\_SignIncrease} & -33.52 & 5.02 & -6.67 & 0 \\
-\textbf{Dryness} & 3.52 & 2.51 & 1.4 & 0.16 \\
-\textbf{log10(Area\_km2)} & -3.23 & 1.56 & -2.07 & 0.04 \\
-\bottomrule
+\textbf{(Intercept)} & 12.99 & 6.29 & 2.07 & 0.04 \\
+\textbf{DeltaF\_perc\_pos} & 0.27 & 0.09 & 3.09 & 0 \\
+\textbf{Forest\_SignIncrease} & -48.57 & 5 & -9.71 & 0 \\
+\textbf{Dryness} & 5.02 & 2.38 & 2.11 & 0.04 \\
+\textbf{log10(Area\_km2)} & -1.68 & 1.49 & -1.13 & 0.26 \\
+\bottomrule()
 \end{longtable}
 
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.04) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.2\%.
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.26) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 1.7\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.77)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.51)) as an interaction with Dryness.
 
@@
 \caption{\label{tab:out-modelArea-int} Results of the model including an interaction between Area and the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 7.13 & 6.86 & 1.04 & 0.3 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
-\textbf{Forest\_SignIncrease} & -30.74 & 5.1 & -6.02 & 0 \\
-\textbf{Dryness} & 7.05 & 2.86 & 2.46 & 0.01 \\
-\textbf{log10(Area\_km2)} & 0.64 & 2.19 & 0.29 & 0.77 \\
-\textbf{Dryness:log10(Area\_km2)} & -3.23 & 1.29 & -2.5 & 0.01 \\
-\bottomrule
+\textbf{(Intercept)} & 9.96 & 6.42 & 1.55 & 0.12 \\
+\textbf{DeltaF\_perc\_pos} & 0.26 & 0.09 & 2.96 & 0 \\
+\textbf{Forest\_SignIncrease} & -46.37 & 5.09 & -9.12 & 0 \\
+\textbf{Dryness} & 7.77 & 2.71 & 2.87 & 0 \\
+\textbf{log10(Area\_km2)} & 1.35 & 2.07 & 0.65 & 0.51 \\
+\textbf{Dryness:log10(Area\_km2)} & -2.56 & 1.22 & -2.09 & 0.04 \\
+\bottomrule()
 \end{longtable}
@@
 \caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & 12.09 & 5.31 & 2.28 & 0.02 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.27 & 0 \\
-\textbf{Forest\_SignIncrease} & -30.5 & 5.62 & -5.43 & 0 \\
-\bottomrule
+\textbf{(Intercept)} & 19.11 & 5.07 & 3.77 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.27 & 0.09 & 3.12 & 0 \\
+\textbf{Forest\_SignIncrease} & -51.5 & 5.79 & -8.89 & 0 \\
+\bottomrule()
 \end{longtable}
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
@@
 \caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
-\textbf{s(log10(Area\_km2))} & 1.95 & 5 & 1.43 & 0.02 \\
-\textbf{s(Dryness)} & 4.62 & 9 & 1.81 & 0 \\
-\textbf{s(length)} & 28.4 & 49 & 1.2 & 0 \\
-\bottomrule
+\textbf{s(log10(Area\_km2))} & 1.58 & 5 & 0.73 & 0.08 \\
+\textbf{s(Dryness)} & 4.62 & 9 & 1.97 & 0 \\
+\textbf{s(length)} & 26.69 & 49 & 1.06 & 0 \\
+\bottomrule()
 \end{longtable}
 
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.37 and deviance explained of 0.44, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.43 and deviance explained of 0.49, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.08)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
 
@@
 \caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & -22.77 & 18.14 & -1.26 & 0.21 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
-\textbf{Forest\_SignIncrease} & -21.42 & 6.58 & -3.25 & 0 \\
-\textbf{Precip\_data\_typeOB} & -11.41 & 13.41 & -0.85 & 0.4 \\
-\textbf{Precip\_data\_typeSG} & 3.29 & 16.54 & 0.2 & 0.84 \\
-\textbf{Assessment\_techniqueEA, HM} & 12.7 & 43.18 & 0.29 & 0.77 \\
-\textbf{Assessment\_techniqueHM} & 37.34 & 12.44 & 3 & 0 \\
-\textbf{Assessment\_techniquePWE} & 49.63 & 12.95 & 3.83 & 0 \\
+\textbf{(Intercept)} & 7.09 & 17.5 & 0.4 & 0.69 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.35 & 0 \\
+\textbf{Forest\_SignIncrease} & -50.21 & 6.97 & -7.2 & 0 \\
+\textbf{Precip\_data\_typeOB} & -15.38 & 12.75 & -1.21 & 0.23 \\
+\textbf{Precip\_data\_typeSG} & -3.24 & 15.23 & -0.21 & 0.83 \\
+\textbf{Assessment\_techniqueEA, HM} & 15.67 & 40.83 & 0.38 & 0.7 \\
+\textbf{Assessment\_techniqueHM} & 29.29 & 11.9 & 2.46 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 25.49 & 12.47 & 2.04 & 0.04 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 39.85 & 42.28 & 0.94 & 0.35 \\
-\textbf{Assessment\_techniqueQPW} & 46.12 & 20.21 & 2.28 & 0.02 \\
+HM} & 12.55 & 40.66 & 0.31 & 0.76 \\
+\textbf{Assessment\_techniqueQPW} & 31.72 & 19.36 & 1.64 & 0.1 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 50.81 & 24.17 & 2.1 & 0.04 \\
-\textbf{Assessment\_techniqueSH} & 42.84 & 12.22 & 3.51 & 0 \\
-\textbf{Forest\_typeCF} & -11.16 & 7.54 & -1.48 & 0.14 \\
-\textbf{Forest\_typeMF} & -1.81 & 7.8 & -0.23 & 0.82 \\
-\textbf{Hydrological\_regimeSD} & 11.44 & 9.16 & 1.25 & 0.21 \\
-\bottomrule
+EA} & 38.53 & 23.04 & 1.67 & 0.1 \\
+\textbf{Assessment\_techniqueSH} & 35.07 & 11.61 & 3.02 & 0 \\
+\textbf{Forest\_typeCF} & -8.77 & 7.19 & -1.22 & 0.22 \\
+\textbf{Forest\_typeMF} & -0.05 & 7.38 & -0.01 & 0.99 \\
+\textbf{Hydrological\_regimeSD} & 5.25 & 8.71 & 0.6 & 0.55 \\
+\bottomrule()
 \end{longtable}
@@
 \caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
-\textbf{s(log10(Area\_km2))} & 1.66 & 4 & 0.74 & 0.14 \\
-\textbf{s(Dryness)} & 4.3 & 9 & 2.42 & 0 \\
-\textbf{s(length)} & 21.69 & 34 & 1.56 & 0 \\
-\bottomrule
+\textbf{s(log10(Area\_km2))} & 0.99 & 4 & 0.94 & 0.04 \\
+\textbf{s(Dryness)} & 4.41 & 9 & 2.45 & 0 \\
+\textbf{s(length)} & 18.55 & 34 & 1.25 & 0 \\
+\bottomrule()
 \end{longtable}
 
-This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.4.
+This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.44.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.14, likely due to the high variance in the data.
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.04, likely due to the high variance in the data.
 It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
@@
 \caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness \textless{} 4.}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
-\textbf{s(Dryness)} & 3.12 & 9 & 2.65 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.53 & 4 & 0.3 & 0.13 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.98 \\
-\bottomrule
+\textbf{s(Dryness)} & 3.51 & 9 & 3.72 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.83 & 4 & 1.08 & 0.02 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.55 \\
+\bottomrule()
 \end{longtable}
@@
 
-This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.33.
+This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.41.
 The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7} and Table \ref{tab:restrictlength}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.13, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.02, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
 The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.81) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.79) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
@@
 \caption{\label{tab:tableassess} Distribution of assessment techniques in the data set}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 PWE, HM & 1 \\
-\bottomrule
+\bottomrule()
 \end{longtable}
@@
 \caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
 \endhead
-\textbf{(Intercept)} & -25.35 & 16.81 & -1.51 & 0.13 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.09 & 3.25 & 0 \\
-\textbf{Forest\_SignIncrease} & -22.67 & 6.23 & -3.64 & 0 \\
-\textbf{Precip\_data\_typeOB} & -8.99 & 12.42 & -0.72 & 0.47 \\
-\textbf{Precip\_data\_typeSG} & 9 & 14.41 & 0.62 & 0.53 \\
-\textbf{Assessment\_techniqueHM} & 37.79 & 11.38 & 3.32 & 0 \\
-\textbf{Assessment\_techniquePWE} & 51.85 & 11.55 & 4.49 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 43.98 & 19.18 & 2.29 & 0.02 \\
-\textbf{Assessment\_techniqueSH} & 42.26 & 11.51 & 3.67 & 0 \\
-\textbf{Forest\_typeCF} & -9.09 & 7.17 & -1.27 & 0.21 \\
-\textbf{Forest\_typeMF} & -1.9 & 7.47 & -0.25 & 0.8 \\
-\textbf{Hydrological\_regimeSD} & 5.54 & 8.88 & 0.62 & 0.53 \\
-\bottomrule
+\textbf{(Intercept)} & 12.6 & 16.34 & 0.77 & 0.44 \\
+\textbf{DeltaF\_perc\_pos} & 0.24 & 0.09 & 2.81 & 0.01 \\
+\textbf{Forest\_SignIncrease} & -49.62 & 6.4 & -7.76 & 0 \\
+\textbf{Precip\_data\_typeOB} & -16.92 & 11.7 & -1.45 & 0.15 \\
+\textbf{Precip\_data\_typeSG} & -3.4 & 13.84 & -0.25 & 0.81 \\
+\textbf{Assessment\_techniqueHM} & 27.5 & 10.72 & 2.57 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 22.58 & 11.56 & 1.95 & 0.05 \\
+\textbf{Assessment\_techniqueQPW} & 26.66 & 18.13 & 1.47 & 0.14 \\
+\textbf{Assessment\_techniqueSH} & 35.48 & 10.8 & 3.29 & 0 \\
+\textbf{Forest\_typeCF} & -8.73 & 6.75 & -1.29 & 0.2 \\
+\textbf{Forest\_typeMF} & 1.53 & 6.99 & 0.22 & 0.83 \\
+\textbf{Hydrological\_regimeSD} & 2.6 & 8.3 & 0.31 & 0.75 \\
+\bottomrule()
 \end{longtable}
@@
 \caption{\label{tab:model8-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
-\toprule
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 \end{minipage} \\
-\midrule
+\midrule()
 \endhead
-\textbf{s(Dryness)} & 3.11 & 9 & 2.59 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.51 & 9 & 0.13 & 0.13 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.96 \\
-\bottomrule
+\textbf{s(Dryness)} & 3.48 & 9 & 3.64 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.81 & 9 & 0.47 & 0.02 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.52 \\
+\bottomrule()
 \end{longtable}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hoekvandijke2022}{}}%
+Hoek van Dijke, A.J., Herold, M., Mallick, K., Benedict, I., Machwitz, M., Schlerf, M., Pranindita, A., Theeuwen, J.J.E., Bastin, J.-F., Teuling, A.J., 2022.
+Shifts in regional water availability due to global tree restoration.
+Nature Geoscience 15, 363--368.
+doi:\href{https://doi.org/10.1038/s41561-022-00935-0}{10.1038/s41561-022-00935-0}
+
 \leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
@@
 doi:\href{https://doi.org/Doi:\%2010.1016/0022-1694(87)90054-0}{Doi: 10.1016/0022-1694(87)90054-0}
+
+\leavevmode\vadjust pre{\hypertarget{ref-levy2018}{}}%
+Levy, M.C., Lopes, A.V., Cohn, A., Larsen, L.G., Thompson, S.E., 2018.
+Land use change increases streamflow across the arc of deforestation in brazil.
+Geophysical Research Letters 45, 3520--3530.
+doi:\url{https://doi.org/10.1002/2017GL076526}
 

### 2022-08-09 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
progress in updating the paper 09082022

@@
   - \usepackage{color}
+abstract: |
+  This document records the replies to the reviewer for the first submission of *Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets* to Journal of Hydrology, which was rejected before review.
 ---
@@
 
-This document records the replies to the reviewer for the first submission of *Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets* to Journal of Hydrology, which was rejected before review.
-
-\
-Reviewer comments in \textcolor{blue}{blue} and our responses in normal text.  
+The reviewer comments are ordered with the Editor in Chief comments first and our responses.
+This is followed by the comments from the associate editor.
+The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.
 \
@@
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected if needed and a full list of all these changes is below:
+These were all checked and corrected if needed and a full list of all these changes is below and is now included in the paper as Supplementary Data Part 1:
 
+- 76, Beaver Creek, the flow was corrected from 600% to 157% after review of the original publication [@baker1984].
 - 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.

### 2022-08-09 paper/Forest_and_Water/Forest_and_Water.Rmd
progress in updating the paper 09082022

--- 2022-04-01 paper/Forest_and_Water/Forest_and_Water.Rmd
@@
 ---
-title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
+title: Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.
 author:
@@
 Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+\textcolor{red}{However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.} The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
 Generalised additive models were used to run flexible regressions including multiple variables.
@@
 
-There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017].
-The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
+The impacts of global deforestation and reforestation are important through their influence on streamflow and both blue and green water availability [@hoekvandijke2022; @schyns2019].
+The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
+The most dramatic result is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
 However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-
-<!-- Include Filoso paper -->
+For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007].
+Similarly a modelling study by @beck2013 found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
+In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
+This suggests that there can be significant variation across the different studies,methodologies and geographical regions.
+
 For the purpose of this paper, *watershed* and *catchment* are interchangeable terms.
@@
 
-Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
-These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover.
+As mentioned, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
+These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships.
+Others have used the understanding from these studies to extrapolate to global scales [@hoekvandijke2022].
+
+The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies with changes in streamflow due to changes in forest cover.
 The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
@@
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
-
-The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^.
+The more regionally concentrated and detailed study by @levy2018 is a further independent dataset with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.
+
+The conclusions of the first mentioned major review paper [@zhang2017] indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^.
 While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
@@
 
-A final summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
+A final earlier summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
 However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
@@
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
-
-Encouraged by the work from @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
-In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.
-
-In particular, the main method in the work by @zhang2017 is a single covariate linear regression, and in @filoso2017 the focus is mainly on classification and there is again some single covariate linear regression.
-As @zhang2017 points out, a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
-However, the subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum.
-And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.
-
-As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further catchments (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).
-
-Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f.
-@jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).
-
-However, there could be further confounding factors, which are eluded to by @filoso2017:
-
-* the type of analysis, i.e. paired watershed studies, modelling, time series analysis etc.
-
-* the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
-
-Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
+The Fuh model and variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2002 to interpret the global impact of reforestation.
+
+There are some clear limitations in these studies.
+The main method in the work by @zhang2017 is a single covariate linear regression.
+In contrast, the systematic review from @filoso2017 emphasises the classification and distributions of the study.
+@zhang2017 points out that a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
+However, a subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum.
+And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.
+
+In contrast to the single covariate linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+The analysis specifically accounted for differences between catchments and differences due to variations in climate.
+Their conclusion highlighted that in particular dry season streamflow was affected by deforestation.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
+There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales. 
+However, there is also a real danger in this process, which is what we will highlight in this paper. There are three potential errors (or limitations) in the mentioned global analyses:
+
+- Latent variables that are not included in the typical single covariate analysis;     
+- Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally    
+- Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
+
+The aim of this paper is to highlight examples of each of these limitations, how they have influenced past work, and provide suggestions of how we can overcome these limitations.
+In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
+Finally, we will highlight future research needs in this area.
+
+We are taking advantage of the earlier work by @zhang2017, @filoso2017 and @zhou2015 and the large database of studies these authors have shared.
 
@@
 
-## The original data sets
-
-The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
+## The original data set
+
+As indicated, the starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
+The columns in this data set (are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in \@ref(Table 1). These abbreviations will later be used in the models.
 
@@
 
-While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
-In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET.
-In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).
-In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
+@zhang2017 use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and use the dryness index to identify the climate type.
+However, the potential or reference ET was not originally included as part of the published data set.
 We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.
@@
 To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from @filoso2017.
-In particular, we focussed on including the 37 data points related to the quantitative analysis in @filoso2017.
-
-In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location.
-These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps^TM^.
-In the dataset, an additional column has been added to indicate the source of the location data.
-
-Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005.
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
+In particular, we focused on the 37 data points related to the quantitative regression analysis used in @filoso2017.
+
+In addition, a few additional variables were included to enhance the data set.
+We added latitude and longitude for the center of the catchment as an approximation of its spatial location.
+Mostly the  data reported by the authors was used, but in some cases the variables had to be approximated from the location of the centre of the catchment using Google Maps^TM^.
+In the dataset, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
+
+As highlighted, @zhang2015 did not provide values for evapotranspiration in the data base.
 Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
@@
 
-Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
+Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
 In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
 Table 6 in @zhang2011).
-We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017.
+
+We conducted a thorough review of all the studies mentioned in the data base of @zhang2015 and sourced all the original papers.
+As a result of this we made several changes to the data base, which are all recorded in Supplementary Data part 1.
+Overall 36 data points were changed and the most common problem was a change in the sign for the change in forest cover or the change in flow.
+We assume that these were transcription errors.
+We also removed one data point from the data set, which corresponds to catchment #1 (Amazon) in @zhang2017.
 This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984].
 Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
 
-The final column in the improved data set is a "notes" column, which we added, but is not further used in the analysis.
+The final column in the improved data base is a "notes" column, which we added, but is not further used in the analysis.
 It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
-This will allow future research to further sructinise our input for errors.
-
-All the final data and analysis for this paper are located on github:  
+This will allow future research to scrutinise our input for errors.
+
+To make all the data and code publicly available, all the final data and analysis for this paper are located on github:  
 [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
@@
 
-Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv")
-Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv")
-new_data <- read_csv("../../data/NewCatchments_T3.csv")
+Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv") %>%
+  mutate(table = "small")
+Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv") %>%
+  mutate(table = "large")
+new_data <- read_csv("../../data/NewCatchments_T3.csv") %>%
+  mutate(table = "new")
 ```
@@
 
-<!--now combine with E0 values -->
+<!--now calculate dryness -->
 ```{r, echo = F}
@@
 
-<!-- correct watershed #76 (Baker1984) to be 157 for DeltaQf_perc -->
+<!-- remove data set 188 and 254 Kamakia and Sambret -->
 
@@
 All_data <- All_data %>%
-  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
-```
-
-To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
+  filter(`Watershed #` != 188) %>%
+  filter(`Watershed #` != 254)
+```
+
+The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change flow as a consequence of changes in forest cover.
+This first aim is similar to @zhang2017, but the main difference is that we start off with all variables in the data set in the model.
+Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and  how latent variables in the data set can be masked and result in relationships that might not really exist.
+Finally, the analysis will highlight how the results are conditional on the dataset.
+
+To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
 
@@
 Here $X_i$ are factorial variables, while $Z_i$ are continuous variables.
-The model intially assumes no direct interactions and all variables are additive.
-We will comment in this assumption in the discussion.
-The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data was converted to positive values, and an additional factorial column ($sign_{forest cover}$) is included indicating whether it was a forest cover increase or decrease.
-
+As a first step, the model assumes no direct interactions and that all variables are additive.
 A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with $\Delta Qf \%$.
@@
 
+The changes in forest cover contain both positive (forestation) and negative values (deforestation).
+In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
+
+However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
+The question becomes how best to analyse this.
+One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
+This can be tested by converting all the change in forest cover data to positive values, and an additional binary column ($sign_{forest cover}$) can be included indicating whether it was a forest cover increase or decrease.
+In the model, the parameter for $sign_{forest cover}$ will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover.
+The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).
+
+A second approach is to test the change in forest cover as a non-linear relationship in the GAM model.
+Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous.
+The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form.
+Both these approaches are tested in the results.
+
 The over arching test focuses on  the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
@@
 
-As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results).
+As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (This analysis is in supplementary material part 2).
 Subsequently the analysis was repeated and the additionally identified catchments were added.

### 2022-08-09 paper/Forest_and_Water/SupplementaryMaterialPart1.Rmd
progress in updating the paper 09082022

---
title: 'Supplementary Information part 1: Changes to original database'
author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
date: "`r Sys.Date()`"
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forestandwater.bib
editor_options: 
  chunk_output_type: inline
header-includes:
  - \usepackage{setspace}
  - \usepackage{color}
abstract: |
  This supplementary material file gives an overview of the changes that were made in the original data based on review of the original literature. Overall 36 data points were changed. The most common problem was a change in the sign of the forest change or the streamflow change.
---

This document list all the changes that were made to the original @zhang2017 database after review of all the original papers.

A particular problem was that many catchments had the wrong sign for the change in forest cover. There are many catchments with reported positive change in cover and a large increase in flow. These were all checked and corrected if needed and a full list of all these changes is below:

- 76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication [@baker1984].  
- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong. The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively. So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%. corrected to 16\%
- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in @zhang2017. This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates "clearfelling". As a result the change in forest cover was changed to -100\% rather than +100\%.  
- 210, March rd, 100, 147.6. Same problem as 67, Bari et al. (1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.  
- 213, 214 and 215, Monda 1, 2 and 3. These catchments are tricky. The original paper [@oshaughnessy1979] only reports on the control period and indicates that the catchments will be cleared. The later summary paper [@watson2001] shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries. The further complication is that the treatment included clearing and reseeding and regrowth. This suggest that the records should be removed from the database, or only the first few years of the experiment used. In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).  
- 230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta_F_perc to -85 from 90.  
- 312, Yerraminup South. The original publication for this catchment is a Western Australian Water Authority report from 1987, which is hard to find, but we have added a copy in the "Papers" folder on github. In this report, in Table 2 on page 11, for the catchment a "Crown cover" decrease of 60\% is given. Changed the sign of the change in forest cover: -60\%.  
- 72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by @cornish1993 and @cornish2001. In the database from @zhang2017, the forest change for all these catchments is positive. However, the paper highlights that these catchments were all logged and either naturally regenerated or were planted with a plantation species. So, similar to the the earlier mentioned Monda catchments, the reported change probably only refers to the first couple of years after clearing (before regrowth). In any case, the reported change in forest cover should be negative (clearing) rather than positive. Corrected for all three catchments.  
- 78, Black Spur 1, the treatments and effects are only reported in a conference paper [@jayasuriya1988] and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in @zhang2017). Corrected. Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.  
- 104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta_Q_f to be negative.  
- 102, Cold Spring. Checking the original paper [@schneider1961] indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta_Q_f to be negative.  
- 85 Bosboukloof. This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire. In any case, the data should reflect a decrease in forest cover: changed the sign of Delta_f_perc to -80\%.   
- 259 Shackam Brook. There were a few issues with this catchment in the original database. The name was misspelled and it was incorrectly attributed to @brown2005. The original paper is the same as 102 [@schneider1961]. Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases. Corrected Delta_Qf_perc to -20.7\%.  
- 95 Sage Brook. Similar to 259 and 102, originates from @schneider1961. Reforestation so Delta_Qf_perc corrected to -19.8\%.  
- 101 Coalburn. Original publication (Robinson, 1993) which is a symposium paper, is not available. The best summary of the research is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment. It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period. Changed to -20.3\%.

A further issue was the inclusion of the results of several catchments, for example from the study by @beck2013, which had no significant change in flow. Despite this, the "average" change in flow was reported in the database. We don't believe that this is correct and the results from such studies should be set to 0. A full list of changes is provided below:

- 97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati. These are all catchments in Puerto Rico from the study from @beck2013. They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation. The values that are cited in the database should all be set to "not significant from 0", so might be included as 0. Including them with positive or negative values is misleading. This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation. Values for all 12 studied catchments set to 0 in the database.  
- 188 Kimakia. and 254 Sambret. The data in the database from @zhang2017 appear to originate from @bruijnzeel1990 which gives 3 values for different lengths of studies. However, the values in the original study by @blackie1979kimakia and @blackie1979kericho do not seem to add up to the same values, and the specific values are not mentioned in the actual papers. In addition, as @bruijnzeel1990 mentions in the footnotes, the control for Kimakia is a bamboo catchment, while the control for Sambret is a tea plantation. Overall, this suggests that the data are probably not a clear deforestation/reforestation study and should be discarded from the analysis.
- 221	N. Creek, Babinda, Queensland. The original paper from this study highlights that the differences between the catchments were insignificant.

### 2022-08-09 paper/Forest_and_Water/SupplementaryMaterialPart1.tex
progress in updating the paper 09082022

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
            pdftitle={Supplementary Information: Changes to original database},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{setspace}
\usepackage{color}

\begin{document}

\begin{frontmatter}

  \title{Supplementary Information: Changes to original database}
    \author[]{%
  %
  }

      \cortext[cor1]{Corresponding author}

  \begin{abstract}
  This supplementary material file gives an overview of the changes that were made in the original data based on review of the original literature. Overall 36 data points were changed. The most common problem was a change in the sign of the forest change or the streamflow change.
  \end{abstract}

 \end{frontmatter}

This document list all the changes that were made to the original \citet{zhang2017} database after review of all the original papers.

A particular problem was that many catchments had the wrong sign for the change in forest cover. There are many catchments with reported positive change in cover and a large increase in flow. These were all checked and corrected if needed and a full list of all these changes is below:

\begin{itemize}
\tightlist
\item
  76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication \citep{baker1984}.\\
\item
  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong. The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively. So value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%. corrected to 16\%
\item
  3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by \citet{zhang2017} is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected\\
\item
  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in \citet{zhang2017}. This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''. As a result the change in forest cover was changed to -100\% rather than +100\%.\\
\item
  210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
\item
  213, 214 and 215, Monda 1, 2 and 3. These catchments are tricky. The original paper \citep{oshaughnessy1979} only reports on the control period and indicates that the catchments will be cleared. The later summary paper \citep{watson2001} shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries. The further complication is that the treatment included clearing and reseeding and regrowth. This suggest that the records should be removed from the database, or only the first few years of the experiment used. In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).\\
\item
  230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.\\
\item
  312, Yerraminup South. The original publication for this catchment is a Western Australian Water Authority report from 1987, which is hard to find, but we have added a copy in the ``Papers'' folder on github. In this report, in Table 2 on page 11, for the catchment a ``Crown cover'' decrease of 60\% is given. Changed the sign of the change in forest cover: -60\%.\\
\item
  72 Barratta, 100 Coachwood, 103 Corkwood, and 83 Bollygum, as cited by \citet{cornish1993} and \citet{cornish2001}. In the database from \citet{zhang2017}, the forest change for all these catchments is positive. However, the paper highlights that these catchments were all logged and either naturally regenerated or were planted with a plantation species. So, similar to the the earlier mentioned Monda catchments, the reported change probably only refers to the first couple of years after clearing (before regrowth). In any case, the reported change in forest cover should be negative (clearing) rather than positive. Corrected for all three catchments.\\
\item
  78, Black Spur 1, the treatments and effects are only reported in a conference paper \citep{jayasuriya1988} and once again indicated clearing, meaning that the change in forest cover should be negative rather than positive (as reported in \citet{zhang2017}). Corrected. Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.\\
\item
  104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
\item
  102, Cold Spring. Checking the original paper \citep{schneider1961} indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
\item
  85 Bosboukloof. This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire. In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.\\
\item
  259 Shackam Brook. There were a few issues with this catchment in the original database. The name was misspelled and it was incorrectly attributed to \citet{brown2005}. The original paper is the same as 102 \citep{schneider1961}. Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases. Corrected Delta\_Qf\_perc to -20.7\%.\\
\item
  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
\item
  101 Coalburn. Original publication (Robinson, 1993) which is a symposium paper, is not available. The best summary of the research is in \citet{birkinshaw2014} which summaries 45 years of resaerch in the Coalburn catchment. It was a reforestation experiment, and there was a decrease in the streamflow over the longer time period. Changed to -20.3\%.
\end{itemize}

A further issue was the inclusion of the results of several catchments, for example from the study by \citet{beck2013}, which had no significant change in flow. Despite this, the ``average'' change in flow was reported in the database. We don't believe that this is correct and the results from such studies should be set to 0. A full list of changes is provided below:

\begin{itemize}
\tightlist
\item
  97 Cibucio, 123 Culebrinas, 244 Portugues, 161 Grande de Loıza, 271 Tanama, 132 Fajardo, 89 Canovanas, 73, Bauta, 163 Grande de Patillas, 283 Valenciano, 181 Inabon, and 162 Grande de Manati. These are all catchments in Puerto Rico from the study from \citet{beck2013}. They should probably be removed from the database as the paper clearly indicates that there is no evidence of a change in flow due to reforestation. The values that are cited in the database should all be set to ``not significant from 0'', so might be included as 0. Including them with positive or negative values is misleading. This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation. Values for all 12 studied catchments set to 0 in the database.\\
\item
  188 Kimakia. and 254 Sambret. The data in the database from \citet{zhang2017} appear to originate from \citet{bruijnzeel1990} which gives 3 values for different lengths of studies. However, the values in the original study by \citet{blackie1979kimakia} and \citet{blackie1979kericho} do not seem to add up to the same values, and the specific values are not mentioned in the actual papers. In addition, as \citet{bruijnzeel1990} mentions in the footnotes, the control for Kimakia is a bamboo catchment, while the control for Sambret is a tea plantation. Overall, this suggests that the data are probably not a clear deforestation/reforestation study and should be discarded from the analysis.
\item
  221 N. Creek, Babinda, Queensland. The original paper from this study highlights that the differences between the catchments were insignificant.
\end{itemize}

\bibliography{forestandwater.bib}

\end{document}

### 2022-08-09 paper/Forest_and_Water/SupplementaryMaterialPart2.Rmd
progress in updating the paper 09082022

---
title: 'Supplementary Information part 2: Testing the improved data sets'
author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
date: "`r Sys.Date()`"
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forestandwater.bib
editor_options: 
  chunk_output_type: inline
header-includes:
  - \usepackage{setspace}
  - \usepackage{color}
abstract: |
  This supplementary material file compares whether the inclusion of additional catchments generates fundamentally different results as the original (but improved) data. Single variable regressions on the smaller (original) dataset are compared with the extended data set.
---

# Introduction

In this document we tested whether the fundamental conclusions in the single variable regressions with the improved data base differed from the original conclusions in @zhang2017. This is to check how much influence the changes to the data set and the additional data might have changed the original conclusions.

First we will read in the data

```{r setup, include=FALSE, echo =T}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))

Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv") %>%
  mutate(table = "Small catchments")
Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv") %>%
  mutate(table = "Large catchments")
new_data <- read_csv("../../data/NewCatchments_T3.csv") %>%
  mutate(table = "new catchments")

# change some column names
names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
names(new_data)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
names(new_data)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(new_data)[9:10] <- c("Precip_data_type", "Assessment_technique")

# combine the data into one dataset

```

We will combine the different tables, but will keep an indicator to see where the data are from.

```{r, echo = T, warning = F, message = F}
Zhang_small$From <- as.numeric(Zhang_small$From)
Zhang_small$To <- as.numeric(Zhang_small$To)
Zhang_all <- bind_rows(Zhang_large,Zhang_small) %>%
  mutate(dataset = "original Zhang et al data")
new_data <- new_data %>%
  mutate(dataset = "new data")
All_data <- bind_rows(Zhang_all, new_data)
```

### Implementing the changes to the overall data

The following code implements the changes described in the Supplementary data part 1. However, many of the changes were implemented manually into the data set. These are simply the remaining changes not implemented manually.

1. removing the duplicates.  
```{r, echo = T}
All_data <- All_data %>%
  mutate(`Possible duplicate` = 
           ifelse(is.na(`Possible duplicate`)==T,0,`Possible duplicate`),
         `Possible duplicate` = as.numeric(`Possible duplicate`)) %>%
  filter(`Possible duplicate` != 1)
```

2. calculating the dryness

```{r, echo = T}
# calculate dryness index
All_data <- All_data %>%
  mutate(Dryness = E0/Pa_mm)
```

3. remove watershed 1 (the Amazon) from the analysis  
```{r, echo = T}
All_data <- All_data %>%
  filter(`Watershed #` != 1)
```

4. remove data set 188 and 254 Kamakia and Sambret
```{r,echo = T}
All_data <- All_data %>%
  filter(`Watershed #` != 188) %>%
  filter(`Watershed #` != 254)
```

5. add a column that indicates forst loss of forest gain

```{r, echo = T}
All_data <- All_data %>%
  mutate(forest_sign = ifelse(DeltaF_perc < 0, "Forest Cover Loss", "Forest Cover Gain"))
```

# Methods

The approach is similar to @zhang2017. We run single variable regressions separating large (> 1000 km^2^) and small catchments (<= 1000 km^2^).

The paper by @zhang2017 calculates the sensitivity of runoff as a function of runoff as:

$\Delta Q_f = 100 \times \frac{\Delta Q_{f,mm}}{\bar{Q}}$

This first equation is superfluous in this case as the data (as extracted from @zhang2017) is already defined in terms of $\Delta Q_f$.

$S_f = \frac{\Delta Q_f}{\Delta F}$

```{r, echo = T}
All_data <- All_data %>%
  filter(is.na(DeltaF_perc) == F) %>%
  mutate(S_f = DeltaQf_perc/DeltaF_perc)
```

In sequence we analyse:

- the relationship between forest cover change and streamflow change for small and large catchments (i.e. Figure 2 in @zhang2017);  
- the relationship between catchment size and the sensitivity to runoff change (i.e. Figure 3 in @zhang2017); and  
- the sensitivity to forest loss as a function of dryness (i.e. Figure 4 in @zhang2017).

# Results

1. The change in flow as a function of change in forest cover

First the results for the original (improved) dataset

```{r, echo = F, message = F}
png("Fig2Zhang.png", width = 2500, height = 1800, res = 300)
All_data %>%
  filter(dataset == "original Zhang et al data") %>%
  ggplot(aes(DeltaF_perc,DeltaQf_perc, colour = forest_sign)) + 
  geom_point() +  geom_smooth(method = "lm") +
  facet_wrap(~table, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  # scale_colour_manual(name = "Forest cover direction", 
  #                     values = c("Decrease" = "red","Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

# and for all data sets
png("Fig2Zhang_all.png", width = 2500, height = 1800, res = 300)
All_data %>%
#  filter(dataset == "original Zhang et al data") %>%
  ggplot(aes(DeltaF_perc,DeltaQf_perc, colour = forest_sign, shape = table)) + 
  geom_point() +  geom_smooth(method = "lm") +
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  # scale_colour_manual(name = "Forest cover direction", 
  #                     values = c("Decrease" = "red","Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

```

```{r Fig2Zhang, echo =F, fig.cap="Changes in flow based on the catchments from the original data set", out.width="90%"}

include_graphics("Fig2Zhang.png")
```
And then the results for all the catchments including the new "added" catchments.

```{r Fig2Zhangnew, echo =F, fig.cap="Changes in flow based on the catchments from the extended data set", out.width="90%"}

include_graphics("Fig2Zhang_all.png")
```

### 2022-08-09 paper/Forest_and_Water/SupplementaryMaterialPart2.tex
progress in updating the paper 09082022

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
            pdftitle={Supplementary Information part 2: Testing the improved data sets},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{setspace}
\usepackage{color}

\begin{document}

\begin{frontmatter}

  \title{Supplementary Information part 2: Testing the improved data sets}
    \author[]{%
  %
  }

      \cortext[cor1]{Corresponding author}

  \begin{abstract}
  This supplementary material file compares whether the inclusion of additional catchments generates fundamentally different results as the original (but improved) data. Single variable regressions on the smaller (original) dataset are compared with the extended data set.
  \end{abstract}

 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

In this document we tested whether the fundamental conclusions in the single variable regressions with the improved data base differed from the original conclusions in \citet{zhang2017}. This is to check how much influence the changes to the data set and the additional data might have changed the original conclusions.

First we will read in the data

We will combine the different tables, but will keep an indicator to see where the data are from.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{From }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{From)}
\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{To }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{To)}
\NormalTok{Zhang\_all }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_large,Zhang\_small) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"original Zhang et al data"}\NormalTok{)}
\NormalTok{new\_data }\OtherTok{\textless{}{-}}\NormalTok{ new\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"new data"}\NormalTok{)}
\NormalTok{All\_data }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_all, new\_data)}
\end{Highlighting}
\end{Shaded}

\hypertarget{implementing-the-changes-to-the-overall-data}{%
\subsubsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}

The following code implements the changes described in the Supplementary data part 1. However, many of the changes were implemented manually into the data set. These are simply the remaining changes not implemented manually.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  removing the duplicates.
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} 
           \FunctionTok{ifelse}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)}\SpecialCharTok{==}\NormalTok{T,}\DecValTok{0}\NormalTok{,}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{),}
         \StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{as.numeric}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  calculating the dryness
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# calculate dryness index}
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Dryness =}\NormalTok{ E0}\SpecialCharTok{/}\NormalTok{Pa\_mm)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  remove watershed 1 (the Amazon) from the analysis
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  remove data set 188 and 254 Kamakia and Sambret
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{188}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{254}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  add a column that indicates forst loss of forest gain
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{forest\_sign =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{, }\StringTok{"Forest Cover Loss"}\NormalTok{, }\StringTok{"Forest Cover Gain"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

The approach is similar to \citet{zhang2017}. We run single variable regressions separating large (\textgreater{} 1000 km\textsuperscript{2}) and small catchments (\textless= 1000 km\textsuperscript{2}).

The paper by \citet{zhang2017} calculates the sensitivity of runoff as a function of runoff as:

\(\Delta Q_f = 100 \times \frac{\Delta Q_{f,mm}}{\bar{Q}}\)

This first equation is superfluous in this case as the data (as extracted from \citet{zhang2017}) is already defined in terms of \(\Delta Q_f\).

\(S_f = \frac{\Delta Q_f}{\Delta F}\)

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(DeltaF\_perc) }\SpecialCharTok{==}\NormalTok{ F) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{S\_f =}\NormalTok{ DeltaQf\_perc}\SpecialCharTok{/}\NormalTok{DeltaF\_perc)}
\end{Highlighting}
\end{Shaded}

In sequence we analyse:

\begin{itemize}
\tightlist
\item
  the relationship between forest cover change and streamflow change for small and large catchments (i.e.~Figure 2 in \citet{zhang2017});\\
\item
  the relationship between catchment size and the sensitivity to runoff change (i.e.~Figure 3 in \citet{zhang2017}); and\\
\item
  the sensitivity to forest loss as a function of dryness (i.e.~Figure 4 in \citet{zhang2017}).
\end{itemize}

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The change in flow as a function of change in forest cover
\end{enumerate}

First the results for the original (improved) dataset

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\begin{figure}
\includegraphics[width=0.9\linewidth]{Fig2Zhang} \caption{Changes in flow based on the catchments from the original data set}\label{fig:Fig2Zhang}
\end{figure}

And then the results for all the catchments including the new ``added'' catchments.

\begin{figure}
\includegraphics[width=0.9\linewidth]{Fig2Zhang_all} \caption{Changes in flow based on the catchments from the extended data set}\label{fig:Fig2Zhangnew}
\end{figure}

\bibliography{forestandwater.bib}

\end{document}

### 2022-08-11 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
further updates paper 11082022

@@
 title: "Replies to editor and associate editor Journal of Hydrology"
-author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
+author:
+  - name: R. Willem Vervoort
+  - name: Eliana Nervi
+  - name: Jimena Alonso
 date: "`r Sys.Date()`"

### 2022-08-11 paper/Forest_and_Water/Reply_first_review_JoH.tex
further updates paper 11082022

@@
             bookmarks=true,
-            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
+            pdfauthor={},
             pdftitle={Replies to editor and associate editor Journal of Hydrology},
@@
   \title{Replies to editor and associate editor Journal of Hydrology}
-    \author[]{%
+    \author[]{R. Willem Vervoort%
+  %
+  }
+
+    \author[]{Eliana Nervi%
+  %
+  }
+
+    \author[]{Jimena Alonso%
   %
@@
   \begin{abstract}
-
+  This document records the replies to the reviewer for the first submission of \emph{Factors determining how catchments respond to forest cover change.
+Re-analysing global data sets} to Journal of Hydrology, which was rejected before review.
   \end{abstract}
@@
 
-This document records the replies to the reviewer for the first submission of \emph{Factors determining how catchments respond to forest cover change.
-Re-analysing global data sets} to Journal of Hydrology, which was rejected before review.
-
-\hfill\break
-Reviewer comments in \textcolor{blue}{blue} and our responses in normal text.\\
+The reviewer comments are ordered with the Editor in Chief comments first and our responses.
+This is followed by the comments from the associate editor.
+The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.\\
 \strut \\
@@
 There are many catchments with reported positive change in cover and a large increase in flow.
-These were all checked and corrected if needed and a full list of all these changes is below:
+These were all checked and corrected if needed and a full list of all these changes is below and is now included in the paper as Supplementary Data Part 1:
 
@@
 \tightlist
+\item
+  76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication \citep{baker1984}.
 \item

### 2022-08-11 paper/Forest_and_Water/Forest_and_Water.Rmd
further updates paper 11082022

@@
 ---
-title: Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.
+title: Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.
 author:
@@
     email: willem.vervoort@sydney.edu.au
-    affiliation: DARE
-    footnote: 1
+    affiliation: a,b
+    correspondingauthor: true
+  - name: R. Willem Vervoort
+    affiliation:
   - name: Eliana Nervi
     email: eliananervif@gmail.com
-    affiliation: INIA 
+    affiliation: c
   - name: Jimena Alonso
     email: jalonso@fing.edu.uy
-    affiliation: IMFIA
+    affiliation: d
 address:
-  - code: DARE
-    address: ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.
-  - code: The University of Sydney
+  - code: b
     address: The University of Sydney, Sydney, NSW 2006, Australia
-  - code: INIA
+  - code: a
+    address: ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of  Agriculture, School of Life and Environmental Sciences.
+  - code: c
     address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria,  INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
-  - code: IMFIA
+  - code: d
     address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
-footnote:
-  - code: 1
-    text: "Corresponding Author"
 abstract: |
   Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
-Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
-\textcolor{red}{However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.} The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
+However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
+The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
 Generalised additive models were used to run flexible regressions including multiple variables.
@@
 
-Table 1  Summary of abbreviations of factors used in the @zhang2017 data set
+Table (\#table:tab1)  Summary of abbreviations of factors used in the @zhang2017 data set
 
@@
 
-To make all the data and code publicly available, all the final data and analysis for this paper are located on github:  
-[https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
-
 ## Statistical modelling
@@
 require(patchwork)
+library(ggcorrplot)
 suppressWarnings(require(knitr))
@@
 
-The over arching test focuses on  the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
-Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.
+The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
+In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref:(tab1)
 
 As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (This analysis is in supplementary material part 2).
-Subsequently the analysis was repeated and the additionally identified catchments were added.
+This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
+
+To make all the data and code publicly available, all the final data and analysis for this paper are located on github:  
+[https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.
 
@@
 Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
+This high skew in some of the data can create difficulties in the statistical modelling and further transformation of the data might be required.
 
@@
 
+<!--
 Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
@@
 ```
+-->
+
+### Geospatial location of the catchments
+
+```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
+
+include_graphics("FAOET0data_final_2022.png")
+```
+
+<!-- DO WE NEED TO REDO THE MAP? Eliana please provide code -->
+
+Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
+The catchments are spread across the world, and relative to @zhang2017, this dataset has a very similar geospatial distribution.
+The major climate gradients are represented in the data, but there appears to be some bias in the spatial locations of the data.
+As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents.
+There is some spatial clustering in the studies in North America, Australia and East Asia.
+
+### Cross correlation between the different variables
+
+A final data exploration is to identify potential cross correlations in the data, which can point to possible interactions or potential biases.
+This analysis can also provide further insight for the statistical modelling, highlighting potential latent variables in the data set.
+
+```{r correlation, eval=T, echo = F}
+All_data2 <- All_data2 %>%
+  mutate(length = To - From,
+         mid_year = From + (To - From)/2)
+
+png("variable_corr_plot.png", width = 960, height = 960)
+model.matrix(~0+DeltaF_perc + 
+                   log10(Area_km2) + 
+                    Dryness + 
+                    length +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
+                    Hydrological_regime, data=All_data2) %>% 
+  cor(use="pairwise.complete.obs") %>% 
+  ggcorrplot(show.diag = F, type="lower", lab=TRUE, lab_size=4)
+dev.off()
+
+```
+
+```{r corgraphs, echo=F, out.width="90%", fig.cap="Correlation matrix for all variables"}
+
+include_graphics("variable_corr_plot.png")
+```
+
+The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations, even though in general cross correlation is quite low between variables.
+Some interesting relationships appear in this graph:
+
+- the negative relationship between log10(Area) and change in forest area (DeltaF_perc) indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.  
+- the weak positive relationship between log10(Area) and the assessment method using hydrological models. This highlights that paired catchment studies mostly concentrate on smaller scales.  
+- A strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
+Smaller catchments used in paired watershed assessments are easier to fully clear or fully replant.
 
@@
 The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.
-
-## The effect of location on the globe
-<!-- include Lat and Long (later E0) -->
-
-```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}
-
-include_graphics("../../data/FAOET0data_final_2022.png")
-```
-
-Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients.
-As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
-Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon (\#eq:eq4)
-\end{align}
 

### 2022-08-11 paper/Forest_and_Water/Forest_and_Water.tex
further updates paper 11082022

--- 2022-07-02 paper/Forest_and_Water/Forest_and_Water.tex
@@
             pdfauthor={},
-            pdftitle={Factors determining how catchments respond to forest cover change. Re-analysing global data sets.},
+            pdftitle={Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.},
             colorlinks=false,
@@
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
 \usepackage{setspace}
@@
 
-  \title{Factors determining how catchments respond to forest cover change. Re-analysing global data sets.}
-    \author[DARE]{R. Willem Vervoort\corref{1}}
+  \title{Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.}
+    \author[a,b]{R. Willem Vervoort}
    \ead{willem.vervoort@sydney.edu.au} 
-    \author[INIA]{Eliana Nervi}
+    \author[]{R. Willem Vervoort}
+
+    \author[c]{Eliana Nervi}
    \ead{eliananervif@gmail.com} 
-    \author[IMFIA]{Jimena Alonso}
+    \author[d]{Jimena Alonso}
    \ead{jalonso@fing.edu.uy} 
-      \address[DARE]{ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
-    \address[The University of Sydney]{The University of Sydney, Sydney, NSW 2006, Australia}
-    \address[INIA]{Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
-    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
-      \cortext[1]{Corresponding Author}
+      \affiliation[b]{The University of Sydney, Sydney, NSW 2006, Australia}
+    \affiliation[a]{ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
+    \affiliation[c]{Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
+    \affiliation[d]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
+    \cortext[cor1]{Corresponding author}
 
@@
   Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
-Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
+Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
 However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
+The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
 Generalised additive models were used to run flexible regressions including multiple variables.
@@
 
-There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow (Andréassian, 2004; Brown et al., 2013, 2005; Filoso et al., 2017; Jackson et al., 2005; Zhang et al., 2017).
-The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result in relation to this, is Figure 5 in Zhang et al.
-(2011) indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
-However, on the other end of the spectrum, for three French catchments (Cosandey et al., 2005), there was no change in streamflow characteristics in two of the catchments after deforestation.
+The impacts of global deforestation and reforestation are important through their influence on streamflow and both blue and green water availability \citep{hoekvandijke2022, schyns2019}.
+The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
+The most dramatic result is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
+However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation.
+For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}.
+Similarly a modelling study by \citet{beck2013} found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
+In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
+This suggests that there can be significant variation across the different studies,methodologies and geographical regions.
 
@@
 
-Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies (Bosch and Hewlett, 1982; Brown et al., 2005), related to reforestation in particular (Filoso et al., 2017), and more generally (Jackson et al., 2005; Zhang et al., 2017).
-These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed.
-The most recent reviews (Filoso et al., 2017; Zhang et al., 2017) developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover.
-The Zhang et al.
-(2017) dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
-In contrast, the paper by Filoso et al.
-(2017) focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
+As mentioned, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies \citep{brown2005, hewlett1984}, related to reforestation in particular \citep{filoso2017}, and more generally \citep{jackson2005, zhang2017}.
+These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships.
+Others have used the understanding from these studies to extrapolate to global scales \citep{hoekvandijke2022}.
+
+The most recent reviews \citep{zhang2017, filoso2017} developed an impressive global database of catchment studies with changes in streamflow due to changes in forest cover.
+The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+In contrast, the paper by \citet{filoso2017} focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
 In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
-
-The conclusions of the first paper (Zhang et al., 2017) suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as \textless{} 1000 km\textsuperscript{2} and large watersheds (catchments) \textgreater{} 1000 km\textsuperscript{2}.
+The more regionally concentrated and detailed study by \citet{levy2018} is a further independent dataset with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.
+
+The conclusions of the first mentioned major review paper \citep{zhang2017} indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as \textless{} 1000 km\textsuperscript{2} and large watersheds (catchments) \textgreater{} 1000 km\textsuperscript{2}.
 While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments (Zhang et al., 2017).
-
-The second study (Filoso et al., 2017) is a systematic review of reforestation studies (only studies in which forest cover increased).
+Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments \citep{zhang2017}.
+
+The second study \citep{filoso2017} is a systematic review of reforestation studies (only studies in which forest cover increased).
 This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
 Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
-For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with Zhang et al.
-(2017)), but they did not identify a distinct threshold and fitted a log-linear relationship.
+For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with \citet{zhang2017}), but they did not identify a distinct threshold and fitted a log-linear relationship.
 In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
 
-A final summary paper that includes much of the same data as Zhang et al.
-(2017) and Filoso et al.
-(2017) is Zhou et al.
-(2015), which has one author in common with Zhang et al.
-(2017).
+A final earlier summary paper that includes much of the same data as \citet{zhang2017} and \citet{filoso2017} is \citet{zhou2015}, which has one author in common with \citet{zhang2017}.
 However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
 In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model.
-A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al.
-(2017).
-
-Encouraged by the work from Zhang et al.
-(2017), Filoso et al.
-(2017) and Zhou et al.
-(2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion.
-In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.
-
-In particular, the main method in the work by Zhang et al.
-(2017) is a single covariate linear regression, and in Filoso et al.
-(2017) the focus is mainly on classification and there is again some single covariate linear regression.
-As Zhang et al.
-(2017) points out, a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments.
-However, the subset of 37 data points in Filoso et al.
-(2017) (their Figure 9) does not appear to support this, suggesting a continuum.
-And while the work Filoso et al.
-(2017) provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.
-
-As a result the objective of this paper is to 1) enhance the data set from Zhang et al.
-(2017) with further catchments (such as from Filoso et al.
-(2017)) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM Wood (2006)).
-
-Building on the analyses by Zhang et al.
-(2017) and Filoso et al.
-(2017), and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover.
-However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths Zhou et al.
-(2015)), the length of the study (c.f.
-Jackson et al.
-(2005); Filoso et al.
-(2017)) and the climate (as indicated by either E0/Pa or latitude and longitude Filoso et al.
-(2017); Zhou et al.
-(2015)).
-
-However, there could be further confounding factors, which are eluded to by Filoso et al. (2017):
+A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in \citet{zhang2017}.
+The Fuh model and variations of the Budyko equilibrium modelling approach was also used by \citet{hoekvandijke2002} to interpret the global impact of reforestation.
+
+There are some clear limitations in these studies.
+The main method in the work by \citet{zhang2017} is a single covariate linear regression.
+In contrast, the systematic review from \citet{filoso2017} emphasises the classification and distributions of the study.
+\citet{zhang2017} points out that a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments.
+However, a subset of 37 data points in \citet{filoso2017} (their Figure 9) does not appear to support this, suggesting a continuum.
+And while the work \citet{filoso2017} provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.
+
+In contrast to the single covariate linear regression in the earlier studies \citep{zhang2017, filoso2017} and the top-down Budyko modelling \citep{zhou2015, hoekvandijke2022}, the regional Brazilian Cerrado study \citep{levy2018} provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+The analysis specifically accounted for differences between catchments and differences due to variations in climate.
+Their conclusion highlighted that in particular dry season streamflow was affected by deforestation.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
+There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales.
+However, there is also a real danger in this process, which is what we will highlight in this paper. There are three potential errors (or limitations) in the mentioned global analyses:
 
 \begin{itemize}
+\tightlist
 \item
-  the type of analysis, i.e.~paired watershed studies, modelling, time series analysis etc.
+  Latent variables that are not included in the typical single covariate analysis;\\
 \item
-  the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.
+  Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally\\
+\item
+  Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
 \end{itemize}
 
-Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.
+The aim of this paper is to highlight examples of each of these limitations, how they have influenced past work, and provide suggestions of how we can overcome these limitations.
+In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
+Finally, we will highlight future research needs in this area.
+
+We are taking advantage of the earlier work by \citet{zhang2017}, \citet{filoso2017} and \citet{zhou2015} and the large database of studies these authors have shared.
 
@@
 
-\hypertarget{the-original-data-sets}{%
-\subsection{The original data sets}\label{the-original-data-sets}}
-
-The starting point of this paper is the data base of studies which were included in Zhang et al.
-(2017) as supplementary material.
-The columns in this data set are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in Zhang et al.
-(2017)), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.
-
-Table 1 Summary of abbreviations of factors used in the Zhang et al. (2017) data set
+\hypertarget{the-original-data-set}{%
+\subsection{The original data set}\label{the-original-data-set}}
+
+As indicated, the starting point of this paper is the data base of studies which were included in \citet{zhang2017} as supplementary material.
+The columns in this data set (are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in \citet{zhang2017}), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in @ref(Table 1). These abbreviations will later be used in the models.
+
+Table (\#table:tab1) Summary of abbreviations of factors used in the \citet{zhang2017} data set
 
 \begin{longtable}[]{@{}lll@{}}
-\toprule
+\toprule()
 Factor & Abbreviation & Definition \\
-\midrule
+\midrule()
 \endhead
@@
 & & and hydrographs \\
-\bottomrule
-\end{longtable}
-
-While Zhang et al.
-(2017) use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set.
-In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET.
-In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).
-In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
-We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) from Zhang et al. (2017) in our analysis.
+\bottomrule()
+\end{longtable}
+
+\citet{zhang2017} use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET\textsubscript{0} or E\textsubscript{0}) in their analysis, and use the dryness index to identify the climate type.
+However, the potential or reference ET was not originally included as part of the published data set.
+We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) from \citet{zhang2017} in our analysis.
 
@@
 
-To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from Filoso et al.
-(2017).
-In particular, we focussed on including the 37 data points related to the quantitative analysis in Filoso et al.
-(2017).
-
-In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location.
-These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps\textsuperscript{TM}.
-In the dataset, an additional column has been added to indicate the source of the location data.
-
-Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al.
-(2017) and Jackson et al.
-(2005).
-Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics).
-Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers.
+To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from \citet{filoso2017}.
+In particular, we focused on the 37 data points related to the quantitative regression analysis used in \citet{filoso2017}.
+
+In addition, a few additional variables were included to enhance the data set.
+We added latitude and longitude for the center of the catchment as an approximation of its spatial location.
+Mostly the data reported by the authors was used, but in some cases the variables had to be approximated from the location of the centre of the catchment using Google Maps\textsuperscript{TM}.
+In the dataset, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
+
+As highlighted, \citet{zhang2015} did not provide values for evapotranspiration in the data base.
+Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 \citep{trabucco2018}, if a value of E\textsubscript{0} was not available from the original papers.
 For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
 
-Similar to Zhang et al. (2017), the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+Similar to \citet{zhang2017}, the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
 
@@
 
-The length of the study can be a variable influencing the change in flow (Filoso et al., 2017; e.g.
-Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987).
-It is not clear if this is an effect of increased water use in growth (Vertessy et al., 2001) or due to changes in interception (Stoof et al., 2012).
-Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al.
-(2017).
-The length of the study was already included in the data from Filoso et al.
-(2017), but these were checked against the original publications.
-
-Several additional data points from catchment studies were extracted from Zhang et al.
-(2011), Zhao et al.
-(2010), Borg et al.
-(1988), Thornton et al.
-(2007), Zhou et al.
-(2010), Rodriguez et al.
-(2010), Ruprecht et al.
-(1991) and Peña-Arancibia et al.
-(2012), and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al.
-(2011)).
-We also removed one data point from the analysis, which corresponds to catchment \#1 (Amazon) in Zhang et al.
-(2017).
-This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
-Furthermore, the change in flow for catchment \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984).
-Finally, on review of all the data in Zhang et al.
-(2017) and Filoso et al.
-(2017), 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
-
-The final column in the improved data set is a ``notes'' column, which we added, but is not further used in the analysis.
-It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al.
-(2017).
-This will allow future research to further sructinise our input for errors.
-
-All the final data and analysis for this paper are located on github:\\
-\href{https://github.com/WillemVervoort/Forest_and_water/tree/publish}{https://github.com/WillemVervoort/Forest\_and\_water} on the ``publish'' branch.
+The length of the study can be a variable influencing the change in flow \citep[e.g.][]{jackson2005, filoso2017}, as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' \citep{kuczera1987}.
+It is not clear if this is an effect of increased water use in growth \citep{vertessy2001} or due to changes in interception \citep{stoof2012}.
+Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by \citet{zhang2017}.
+The length of the study was already included in the data from \citet{filoso2017}, but these were checked against the original publications.
+
+Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{pena-arancibia2012}, and these were checked against the existing studies to prevent overlap.
+In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in \citet{zhang2011}).
+
+We conducted a thorough review of all the studies mentioned in the data base of \citet{zhang2015} and sourced all the original papers.
+As a result of this we made several changes to the data base, which are all recorded in Supplementary Data part 1.
+Overall 36 data points were changed and the most common problem was a change in the sign for the change in forest cover or the change in flow.
+We assume that these were transcription errors.\\
+We also removed one data point from the data set, which corresponds to catchment \#1 (Amazon) in \citet{zhang2017}.
+This is because the cited reference \citep{roche1981} only relates to 1 and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded.
+Finally, on review of all the data in \citet{zhang2017} and \citet{filoso2017}, 29 potential duplicates were identified and flagged in the data, and not used in the analysis.
+
+The final column in the improved data base is a ``notes'' column, which we added, but is not further used in the analysis.
+It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from \citet{zhang2017}.
+This will allow future research to scrutinise our input for errors.
 
@@
 
-To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) (Wood, 2006).
+The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change flow as a consequence of changes in forest cover.
+This first aim is similar to \citet{zhang2017}, but the main difference is that we start off with all variables in the data set in the model.
+Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and how latent variables in the data set can be masked and result in relationships that might not really exist.
+Finally, the analysis will highlight how the results are conditional on the dataset.
+
+To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) \citep{wood2006}.
 
@@
 Here \(X_i\) are factorial variables, while \(Z_i\) are continuous variables.
-The model intially assumes no direct interactions and all variables are additive.
-We will comment in this assumption in the discussion.
-The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In Zhang et al.
-(2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-Therefore all the change in forest cover data was converted to positive values, and an additional factorial column (\(sign_{forest cover}\)) is included indicating whether it was a forest cover increase or decrease.
-
+As a first step, the model assumes no direct interactions and that all variables are additive.
 A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with \(\Delta Qf \%\).
@@
 For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty.
-The result of this approach is that for high enough smoothing parameters (i.e.~if the data is very ``wiggly'') the smooth term can be shrunk to 0 and thus will be no longer significant (Wood, 2006).
+The result of this approach is that for high enough smoothing parameters (i.e.~if the data is very ``wiggly'') the smooth term can be shrunk to 0 and thus will be no longer significant \citep{wood2006}.
 This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship.
@@
 
-The over arching test focuses on the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: Zhang et al.
-(2017); Filoso et al.
-(2017); Zhou et al.
-(2015)): climate, size of catchment and length of study.
-Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.
-
-As an initial approach we only used the data from Zhang et al.
-(2017) to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results).
-Subsequently the analysis was repeated and the additionally identified catchments were added.
+The changes in forest cover contain both positive (forestation) and negative values (deforestation).
+In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
+
+However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
+The question becomes how best to analyse this.
+One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
+This can be tested by converting all the change in forest cover data to positive values, and an additional binary column (\(sign_{forest cover}\)) can be included indicating whether it was a forest cover increase or decrease.
+In the model, the parameter for \(sign_{forest cover}\) will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover.
+The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).
+
+A second approach is to test the change in forest cover as a non-linear relationship in the GAM model.
+Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous.
+The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form.
+Both these approaches are tested in the results.
+
+The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
+In addition to these earlier identified factors, this study also tested for the factors listed in Table @ref:(tab1)
+
+As an initial approach we only used the data from \citet{zhang2017} to make sure that the additional catchments added to the data set did not influence the results (This analysis is in supplementary material part 2).
+This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
+
+To make all the data and code publicly available, all the final data and analysis for this paper are located on github:\\
+\href{https://github.com/WillemVervoort/Forest_and_water/tree/publish}{https://github.com/WillemVervoort/Forest\_and\_water} on the ``publish'' branch.
 
@@
 
-The overall dataset contains 327 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from Zhang et al.
-(2017) contained 312 catchments and the Filoso et al.
-(2017) study used 37 catchments (Table S2 in Filoso et al.
-(2017)).
+The overall dataset contains 329 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+In contrast, the original dataset from \citet{zhang2017} contained 312 catchments and the \citet{filoso2017} study used 37 catchments (Table S2 in \citet{filoso2017}).
 The current number of catchments is the result of the removal of duplicates and our modifications and additions.
@@
 Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).
+This high skew in some of the data can create difficulties in the statistical modelling and further transformation of the data might be required.
 
@@
 
-Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
+\hypertarget{geospatial-location-of-the-catchments}{%
+\subsubsection{Geospatial location of the catchments}\label{geospatial-location-of-the-catchments}}
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of decreases (top) and increases (bottom) in forest cover}\label{fig:increasedecrease}
+\includegraphics[width=0.9\linewidth]{FAOET0data_final_2022} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
 \end{figure}
+
+Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
+The catchments are spread across the world, and relative to \citet{zhang2017}, this dataset has a very similar geospatial distribution.
+The major climate gradients are represented in the data, but there appears to be some bias in the spatial locations of the data.
+As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents.
+There is some spatial clustering in the studies in North America, Australia and East Asia.
+
+\hypertarget{cross-correlation-between-the-different-variables}{%
+\subsubsection{Cross correlation between the different variables}\label{cross-correlation-between-the-different-variables}}
+
+A final data exploration is to identify potential cross correlations in the data, which can point to possible interactions or potential biases.
+This analysis can also provide further insight for the statistical modelling, highlighting potential latent variables in the data set.
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{variable_corr_plot} \caption{Correlation matrix for all variables}\label{fig:corgraphs}
+\end{figure}
+
+The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations, even though in general cross correlation is quite low between variables.
+Some interesting relationships appear in this graph:
+
+\begin{itemize}
+\tightlist
+\item
+  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc) indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.\\
+\item
+  the weak positive relationship between log10(Area) and the assessment method using hydrological models. This highlights that paired catchment studies mostly concentrate on smaller scales.\\
+\item
+  A strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
+Smaller catchments used in paired watershed assessments are easier to fully clear or fully replant.
+\end{itemize}
 
@@
 
-Following Zhang et al.
-(2017), the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
+Following \citet{zhang2017}, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
 
@@
 Clearly there is no major change arising from the additional data.}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 6.24 & 5.19 & 1.2 & 0.23 \\
-\textbf{DeltaF\_perc\_pos} & 0.46 & 0.08 & 5.71 & 0 \\
-\textbf{Forest\_SignIncrease} & -26.95 & 5.4 & -4.99 & 0 \\
-\textbf{(Intercept)} & 7.79 & 5.01 & 1.56 & 0.12 \\
-\textbf{DeltaF\_perc\_pos} & 0.43 & 0.08 & 5.64 & 0 \\
-\textbf{Forest\_SignIncrease} & -34.65 & 4.95 & -7 & 0 \\
-\bottomrule
-\end{longtable}
-
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.24, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 13.65 & 5.04 & 2.71 & 0.01 \\
+\textbf{DeltaF\_perc\_pos} & 0.36 & 0.08 & 4.52 & 0 \\
+\textbf{Forest\_SignIncrease} & -43.21 & 5.53 & -7.82 & 0 \\
+\textbf{(Intercept)} & 17.39 & 5.2 & 3.34 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.32 & 0.08 & 4.04 & 0 \\
+\textbf{Forest\_SignIncrease} & -50.63 & 5.29 & -9.57 & 0 \\
+\bottomrule()
+\end{longtable}
+
+The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.3, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
 The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
@@
 Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
-Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total Zhang et al.
-(2017) data set.
+Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total \citet{zhang2017} data set.
 But this also means that our re-analysis of the data can be directly compared to the original study.
@@
 However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being catchment dryness and area (following Zhang et al.
-(2017)), which we will analyse later.
-
-\hypertarget{the-effect-of-location-on-the-globe}{%
-\subsection{The effect of location on the globe}\label{the-effect-of-location-on-the-globe}}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{../../data/FAOET0data_final_2022} \caption{Distribution of included catchments across the globe based on reported or estimated latitude and longitude}\label{fig:globalmap}
-\end{figure}
-
-Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients.
-As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world.
-Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon \label{eq:eq4}
-\end{align}
+The obvious ones being catchment dryness and area (following \citet{zhang2017}), which we will analyse later.
 
@@
 \caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 9.88 & 5.42 & 1.82 & 0.07 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.68 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.72 & 5.21 & -6.86 & 0 \\
-\textbf{Latitude} & -0.1 & 0.09 & -1.17 & 0.24 \\
-\textbf{Longitude} & -0.02 & 0.03 & -0.62 & 0.54 \\
-\bottomrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 20.28 & 5.45 & 3.72 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.08 & 4.1 & 0 \\
+\textbf{Forest\_SignIncrease} & -50.75 & 5.34 & -9.51 & 0 \\
+\textbf{Latitude} & -0.17 & 0.09 & -1.96 & 0.05 \\
+\textbf{Longitude} & -0.05 & 0.03 & -1.71 & 0.09 \\
+\bottomrule()
 \end{longtable}
@@
 There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.24 suggesting further factors influencing the change in streamflow that are currently not included in the model.
+The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.31 suggesting further factors influencing the change in streamflow that are currently not included in the model.
 
@@
 Both dryness index and \emph{Pa (mm)} were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
-As a result only the dryness index was retained as a climate indicator to align with the earlier work by Zhang et al.
-(2017).
+As a result only the dryness index was retained as a climate indicator to align with the earlier work by \citet{zhang2017}.
 Given that Latitude and Longitude were not significant, we dropped these from the model.
@@
 \caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 4.2 & 5.86 & 0.72 & 0.47 \\
-\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.7 & 0 \\
-\textbf{Forest\_SignIncrease} & -35.37 & 4.97 & -7.12 & 0 \\
-\textbf{Dryness} & 2.76 & 2.5 & 1.11 & 0.27 \\
-\bottomrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 12.33 & 5.98 & 2.06 & 0.04 \\
+\textbf{DeltaF\_perc\_pos} & 0.33 & 0.08 & 4.03 & 0 \\
+\textbf{Forest\_SignIncrease} & -51.91 & 5.33 & -9.75 & 0 \\
+\textbf{Dryness} & 4.17 & 2.58 & 1.61 & 0.11 \\
+\bottomrule()
 \end{longtable}
@@
 Similar to \emph{E0} or \emph{Pa\_mm}, the results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
-This might seem suprising in light of earlier reported results (Filoso et al., 2017; Zhang et al., 2017).
-In this case, the evidence is highly doubtful (p = 0.27).
+This might seem suprising in light of earlier reported results \citep{zhang2017, filoso2017}.
+In this case, the evidence is highly doubtful (p = 0.11).
 However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.24.
+This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.31.
 As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
@@
 \end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
+\midrule()
+\endfirsthead
+\toprule()
 \begin{minipage}[b]{\linewidth}\centering
@@
 
-The second major variable is the effect of area on the change in flow, following the analysis by Zhang et al.
-(2017) and Filoso et al.
-(2017).
+The second major variable is the effect of area on the change in flow, following the analysis by \citet{zhang2017} and \citet{filoso2017}.
 Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.
@@
 \caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 11.15 & 6.73 & 1.66 & 0.1 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.09 & 3.49 & 0 \\
-\textbf{Forest\_SignIncrease} & -33.52 & 5.02 & -6.67 & 0 \\
-\textbf{Dryness} & 3.52 & 2.51 & 1.4 & 0.16 \\
-\textbf{log10(Area\_km2)} & -3.23 & 1.56 & -2.07 & 0.04 \\
-\bottomrule
-\end{longtable}
-
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.04) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 3.2\%.
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 16.51 & 6.82 & 2.42 & 0.02 \\
+\textbf{DeltaF\_perc\_pos} & 0.26 & 0.1 & 2.68 & 0.01 \\
+\textbf{Forest\_SignIncrease} & -50.38 & 5.46 & -9.23 & 0 \\
+\textbf{Dryness} & 4.62 & 2.61 & 1.77 & 0.08 \\
+\textbf{log10(Area\_km2)} & -2.04 & 1.62 & -1.26 & 0.21 \\
+\bottomrule()
+\end{longtable}
+
+The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.21) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
+In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 2\%.
 Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.77)) as an interaction with Dryness.
+Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.73)) as an interaction with Dryness.
 
@@
 \caption{\label{tab:out-modelArea-int} Results of the model including an interaction between Area and the dryness index}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 7.13 & 6.86 & 1.04 & 0.3 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
-\textbf{Forest\_SignIncrease} & -30.74 & 5.1 & -6.02 & 0 \\
-\textbf{Dryness} & 7.05 & 2.86 & 2.46 & 0.01 \\
-\textbf{log10(Area\_km2)} & 0.64 & 2.19 & 0.29 & 0.77 \\
-\textbf{Dryness:log10(Area\_km2)} & -3.23 & 1.29 & -2.5 & 0.01 \\
-\bottomrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 13.7 & 6.98 & 1.96 & 0.05 \\
+\textbf{DeltaF\_perc\_pos} & 0.25 & 0.1 & 2.56 & 0.01 \\
+\textbf{Forest\_SignIncrease} & -48.35 & 5.56 & -8.7 & 0 \\
+\textbf{Dryness} & 7.19 & 2.97 & 2.42 & 0.02 \\
+\textbf{log10(Area\_km2)} & 0.79 & 2.26 & 0.35 & 0.73 \\
+\textbf{Dryness:log10(Area\_km2)} & -2.39 & 1.34 & -1.78 & 0.08 \\
+\bottomrule()
 \end{longtable}
@@
 
-The work by Filoso et al.
-(2017) and earlier by Jackson et al.
-(2005) has indicated that the length of the study might influence the response.
-This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al.
-(2005)).
+The work by \citet{filoso2017} and earlier by \citet{jackson2005} has indicated that the length of the study might influence the response.
+This links to the idea from \citet{kuczera1987} that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also \citet{jackson2005}).
 In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (\emph{Dryness} and \emph{Area}) were considered non-linear.
-As a result a shrinkage smoothing spline (Wood, 2006) was applied to these variables.
+As a result a shrinkage smoothing spline \citep{wood2006} was applied to these variables.
 
@@
 \caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & 12.09 & 5.31 & 2.28 & 0.02 \\
-\textbf{DeltaF\_perc\_pos} & 0.3 & 0.09 & 3.27 & 0 \\
-\textbf{Forest\_SignIncrease} & -30.5 & 5.62 & -5.43 & 0 \\
-\bottomrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 18.87 & 5.37 & 3.52 & 0 \\
+\textbf{DeltaF\_perc\_pos} & 0.32 & 0.09 & 3.77 & 0 \\
+\textbf{Forest\_SignIncrease} & -53.75 & 6.15 & -8.74 & 0 \\
+\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule
-\endhead
-\textbf{s(log10(Area\_km2))} & 1.95 & 5 & 1.43 & 0.02 \\
-\textbf{s(Dryness)} & 4.62 & 9 & 1.81 & 0 \\
-\textbf{s(length)} & 28.4 & 49 & 1.2 & 0 \\
-\bottomrule
-\end{longtable}
-
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.37 and deviance explained of 0.44, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0)) explain significant variation in the data.
+\midrule()
+\endhead
+\textbf{s(log10(Area\_km2))} & 0.22 & 5 & 0.08 & 0.18 \\
+\textbf{s(Dryness)} & 4.94 & 9 & 1.63 & 0.01 \\
+\textbf{s(length)} & 9.9 & 49 & 0.3 & 0.12 \\
+\bottomrule()
+\end{longtable}
+
+Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.35 and deviance explained of 0.39, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
+For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.18)), \emph{Dryness} (p = 0.01)) and \emph{length} (p = 0.12)) explain significant variation in the data.
 
@@
 \caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & -22.77 & 18.14 & -1.26 & 0.21 \\
-\textbf{DeltaF\_perc\_pos} & 0.31 & 0.09 & 3.33 & 0 \\
-\textbf{Forest\_SignIncrease} & -21.42 & 6.58 & -3.25 & 0 \\
-\textbf{Precip\_data\_typeOB} & -11.41 & 13.41 & -0.85 & 0.4 \\
-\textbf{Precip\_data\_typeSG} & 3.29 & 16.54 & 0.2 & 0.84 \\
-\textbf{Assessment\_techniqueEA, HM} & 12.7 & 43.18 & 0.29 & 0.77 \\
-\textbf{Assessment\_techniqueHM} & 37.34 & 12.44 & 3 & 0 \\
-\textbf{Assessment\_techniquePWE} & 49.63 & 12.95 & 3.83 & 0 \\
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
+\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
+\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
+\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
+\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
+\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
+\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 39.85 & 42.28 & 0.94 & 0.35 \\
-\textbf{Assessment\_techniqueQPW} & 46.12 & 20.21 & 2.28 & 0.02 \\
+HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
+\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 50.81 & 24.17 & 2.1 & 0.04 \\
-\textbf{Assessment\_techniqueSH} & 42.84 & 12.22 & 3.51 & 0 \\
-\textbf{Forest\_typeCF} & -11.16 & 7.54 & -1.48 & 0.14 \\
-\textbf{Forest\_typeMF} & -1.81 & 7.8 & -0.23 & 0.82 \\
-\textbf{Hydrological\_regimeSD} & 11.44 & 9.16 & 1.25 & 0.21 \\
-\bottomrule
+EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
+\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
+\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
+\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
+\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
+\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule
-\endhead
-\textbf{s(log10(Area\_km2))} & 1.66 & 4 & 0.74 & 0.14 \\
-\textbf{s(Dryness)} & 4.3 & 9 & 2.42 & 0 \\
-\textbf{s(length)} & 21.69 & 34 & 1.56 & 0 \\
-\bottomrule
-\end{longtable}
-
-This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.4.
+\midrule()
+\endhead
+\textbf{s(log10(Area\_km2))} & 0.7 & 4 & 0.4 & 0.12 \\
+\textbf{s(Dryness)} & 4.26 & 9 & 1.75 & 0 \\
+\textbf{s(length)} & 16.71 & 34 & 0.92 & 0.01 \\
+\bottomrule()
+\end{longtable}
+
+This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
 This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
@@
 
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.14, likely due to the high variance in the data.
+Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.12, likely due to the high variance in the data.
 It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
@@
 \end{minipage} \\
-\midrule
-\endhead
-\textbf{s(Dryness)} & 3.12 & 9 & 2.65 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.53 & 4 & 0.3 & 0.13 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.98 \\
-\bottomrule
+\midrule()
+\endhead
+\textbf{s(Dryness)} & 3.54 & 9 & 3.13 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.73 & 4 & 0.58 & 0.07 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.7 \\
+\bottomrule()
 \end{longtable}
@@
 The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) does not reflect a similar response to Filoso et al.
-(2017) and Jackson et al.
-(2005).
+The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) does not reflect a similar response to \citet{filoso2017} and \citet{jackson2005}.
 One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \ref{fig:smoothsmodel6}).
@@
 The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4 and \emph{length} \textgreater{} 60 years.
-This result in a reduction of the data set from 327 to 306 catchments.
-
-This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.33.
+This result in a reduction of the data set from 329 to 308 catchments.
+
+This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.37.
 The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7} and Table \ref{tab:restrictlength}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.13, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.07, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
 The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
@@
 
-Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to Zhang et al.
-(2017), there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
+Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to \citet{zhang2017}, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
 If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
@@
 
-An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change (Navas et al., 2019).
+An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change \citep{navas2019}.
 Therefore, specifically if the forest cover change is small relative to the catchment size, the effect of this change will be buffered.
@@
 There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.36 with a slope of -9.81) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.35 with a slope of -9.67) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
 This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
@@
 \end{minipage} \\
-\midrule
-\endhead
-PWE & 183 \\
+\midrule()
+\endhead
+PWE & 185 \\
 HM & 57 \\
@@
 \caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{(Intercept)} & -25.35 & 16.81 & -1.51 & 0.13 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.09 & 3.25 & 0 \\
-\textbf{Forest\_SignIncrease} & -22.67 & 6.23 & -3.64 & 0 \\
-\textbf{Precip\_data\_typeOB} & -8.99 & 12.42 & -0.72 & 0.47 \\
-\textbf{Precip\_data\_typeSG} & 9 & 14.41 & 0.62 & 0.53 \\
-\textbf{Assessment\_techniqueHM} & 37.79 & 11.38 & 3.32 & 0 \\
-\textbf{Assessment\_techniquePWE} & 51.85 & 11.55 & 4.49 & 0 \\
-\textbf{Assessment\_techniqueQPW} & 43.98 & 19.18 & 2.29 & 0.02 \\
-\textbf{Assessment\_techniqueSH} & 42.26 & 11.51 & 3.67 & 0 \\
-\textbf{Forest\_typeCF} & -9.09 & 7.17 & -1.27 & 0.21 \\
-\textbf{Forest\_typeMF} & -1.9 & 7.47 & -0.25 & 0.8 \\
-\textbf{Hydrological\_regimeSD} & 5.54 & 8.88 & 0.62 & 0.53 \\
-\bottomrule
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & 16.98 & 18.34 & 0.93 & 0.36 \\
+\textbf{DeltaF\_perc\_pos} & 0.23 & 0.1 & 2.39 & 0.02 \\
+\textbf{Forest\_SignIncrease} & -52.33 & 7.12 & -7.35 & 0 \\
+\textbf{Precip\_data\_typeOB} & -16.66 & 13.15 & -1.27 & 0.21 \\
+\textbf{Precip\_data\_typeSG} & -5.38 & 15.48 & -0.35 & 0.73 \\
+\textbf{Assessment\_techniqueHM} & 24.86 & 12.05 & 2.06 & 0.04 \\
+\textbf{Assessment\_techniquePWE} & 23.02 & 12.84 & 1.79 & 0.07 \\
+\textbf{Assessment\_techniqueQPW} & 24.65 & 20.36 & 1.21 & 0.23 \\
+\textbf{Assessment\_techniqueSH} & 33.31 & 12.13 & 2.75 & 0.01 \\
+\textbf{Forest\_typeCF} & -11.34 & 7.65 & -1.48 & 0.14 \\
+\textbf{Forest\_typeMF} & -0.46 & 7.85 & -0.06 & 0.95 \\
+\textbf{Hydrological\_regimeSD} & 1.41 & 9.36 & 0.15 & 0.88 \\
+\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule
-\endhead
-\textbf{s(Dryness)} & 3.11 & 9 & 2.59 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.51 & 9 & 0.13 & 0.13 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.96 \\
-\bottomrule
+\midrule()
+\endhead
+\textbf{s(Dryness)} & 3.46 & 9 & 2.71 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.72 & 9 & 0.32 & 0.04 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.74 \\
+\bottomrule()
 \end{longtable}
@@
 Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
-In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in Zhou et al.
-(2015)) appears to suggest a much smaller effect on the change in flow.
+In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in \citet{zhou2015}) appears to suggest a much smaller effect on the change in flow.
 However, as indicated earlier, the EA studies in the database are all on the drier end of the \emph{Dryness} spectrum, highlighting another unresolved interaction in the data.
@@
 
-In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies (Filoso et al., 2017; Zhang et al., 2017; Zhou et al., 2015).
+In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies \citep{zhang2017, zhou2015, filoso2017}.
 This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
@@
 
-In contrast to Filoso et al.
-(2017), we did not identify that the length of the observation period is a significant variable in our final model.
+In contrast to \citet{filoso2017}, we did not identify that the length of the observation period is a significant variable in our final model.
 However, there are further confounding factors in the data, which were not analysed in this study.
-These were also classified by Filoso et al.
-(2017) and these factors might create biases in the data set that can impact the overall assessment.
+These were also classified by \citet{filoso2017} and these factors might create biases in the data set that can impact the overall assessment.
 For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow (Vervoort et al., 2021).
+However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow \citep{vervoort2021}.
 This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-andreassian2004}{}}%
-Andréassian, V., 2004. Waters and forests: From historical controversy to scientific debate. Journal of Hydrology 291, 1--27. doi:\url{https://doi.org/10.1016/j.jhydrol.2003.12.015}
-
-\leavevmode\vadjust pre{\hypertarget{ref-baker1984}{}}%
-Baker Jr., M.B., 1984. Changes in streamflow in an herbicide-treated pinyon-juniper watershed in arizona. Water Resources Research 20, 1639--1642. doi:\url{https://doi.org/10.1029/WR020i011p01639}
-
-\leavevmode\vadjust pre{\hypertarget{ref-borg1988}{}}%
-Borg, H., Bell, R.W., Loh, I.C., 1988.
-Streamflow and stream salinity in a small water supply catchment in southwest western australia after reforestation.
-Journal of Hydrology 103, 323--333.
-doi:\url{https://doi.org/10.1016/0022-1694(88)90141-2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hewlett1984}{}}%
-Bosch, J.M., Hewlett, J.D., 1982. A review of catchment experiments to determine the effect of vegetation changes on water yield and evapotranspiration. Journal of Hydrology 55, 3--23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brown2013}{}}%
-Brown, A.E., Western, A.W., McMahon, T.A., Zhang, L., 2013.
-Impact of forest cover changes on annual streamflow and flow duration curves.
-Journal of Hydrology 483, 39--50.
-doi:\url{http://dx.doi.org/10.1016/j.jhydrol.2012.12.031}
-
-\leavevmode\vadjust pre{\hypertarget{ref-brown2005}{}}%
-Brown, A.E., Zhang, L., McMahon, T.A., Western, A.W., Vertessy, R.A., 2005.
-\href{http://www.sciencedirect.com/science/article/B6V6C-4G05MM9-1/2/bbc5fc0e958a8f34bcb7c1cc7fa57b48}{A review of paired catchment studies for determining changes in water yield resulting from alterations in vegetation}.
-Journal of Hydrology 310, 28--61.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cosandey2005}{}}%
-Cosandey, C., Andréassian, V., Martin, C., Didon-Lescot, J.F., Lavabre, J., Folton, N., Mathys, N., Richard, D., 2005.
-The hydrological impact of the mediterranean forest: A review of french research.
-Journal of Hydrology 301, 235--249.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2004.06.040}
-
-\leavevmode\vadjust pre{\hypertarget{ref-filoso2017}{}}%
-Filoso, S., Bezerra, M.O., Weiss, K.C.B., Palmer, M.A., 2017.
-Impacts of forest restoration on water yield: A systematic review.
-PLOS ONE 12, e0183210.
-doi:\href{https://doi.org/10.1371/journal.pone.0183210}{10.1371/journal.pone.0183210}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jackson2005}{}}%
-Jackson, R.B., Jobbagy, E.G., Avissar, R., Roy, S.B., Barrett, D.J., Cook, C.W., Farley, K.A., Maitre, D.C.
-le, McCarl, B.A., Murray, B.C., 2005.
-Trading water for carbon with biological carbon sequestration.
-Science 310, 1944--1947.
-doi:\href{https://doi.org/10.1126/science.1119282}{10.1126/science.1119282}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kuczera1987}{}}%
-Kuczera, G., 1987.
-Prediction of water yield reductions following a bushfire in ash-mixed species eucalypt forest.
-Journal of Hydrology 94, 215--236.
-doi:\href{https://doi.org/Doi:\%2010.1016/0022-1694(87)90054-0}{Doi: 10.1016/0022-1694(87)90054-0}
-
-\leavevmode\vadjust pre{\hypertarget{ref-navas2019}{}}%
-Navas, R., Alonso, J., Gorgoglione, A., Vervoort, R.W., 2019.
-\href{https://www.mdpi.com/2073-4441/11/7/1433}{Identifying climate and human impact trends in streamflow: A case study in uruguay}.
-Water 11, 1433.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pena-arancibia2012}{}}%
-Peña-Arancibia, J.L., Dijk, A.I.J.M.
-van, Guerschman, J.P., Mulligan, M., Bruijnzeel, L.A., McVicar, T.R., 2012.
-Detecting changes in streamflow after partial woodland clearing in two large catchments in the seasonal tropics.
-Journal of Hydrology 416-417, 60--71.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2011.11.036}
-
-\leavevmode\vadjust pre{\hypertarget{ref-roche1981}{}}%
-Roche, M., 1981. Watershed investigations for development of forest resources of the amazon region in french guyana. Tropical Agricultural Hydrology. J 75--82.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rodriguez2010}{}}%
-Rodriguez, D.A., Tomasella, J., Linhares, C., 2010.
-Is the forest conversion to pasture affecting the hydrological response of amazonian catchments?
-Signals in the ji-paraná basin.
-Hydrological Processes 24, 1254--1269.
-doi:\url{https://doi.org/10.1002/hyp.7586}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ruprechtetal1991}{}}%
-Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A., Stoneman, G.L., 1991.
-Early hydrological response to intense forest thinning in southwestern australia.
-Journal of Hydrology 127, 261--277.
-doi:\url{https://doi.org/10.1016/0022-1694(91)90118-2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-stoof2012}{}}%
-Stoof, C.R., Vervoort, R.W., Iwema, J., Elsen, E.
-van den, Ferreira, A.J.D., Ritsema, C.J., 2012.
-Hydrological response of a small catchment burned by experimental fire.
-Hydrol.
-Earth Syst.
-Sci.
-16, 267--285.
-doi:\href{https://doi.org/10.5194/hess-16-267-2012}{10.5194/hess-16-267-2012}
-
-\leavevmode\vadjust pre{\hypertarget{ref-thornton2007}{}}%
-Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007.
-The brigalow catchment study: II*.
-Clearing brigalow (acacia harpophylla) for cropping or pasture increases runoff.
-Australian Journal of Soil Research 45, 496--511.
-doi:\href{https://doi.org/doi:10.1071/SR07064}{doi:10.1071/SR07064}
-
-\leavevmode\vadjust pre{\hypertarget{ref-trabucco2018}{}}%
-Trabucco, A., Zomer, R.J., 2018. Global aridity index and potential evapo-transpiration (ET0) climate database v2. CGIAR consortium for spatial information(CGIAR-CSI).
-
-\leavevmode\vadjust pre{\hypertarget{ref-vertessy2001}{}}%
-Vertessy, R.A., Watson, F.G.R., O'Sullivan, S.K., 2001.
-Factors determining relations between stand age and catchment water balance in mountain ash forests.
-Forest Ecology and Management 143, 13--26.
-doi:\url{https://doi.org/10.1016/S0378-1127(00)00501-6}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vervoort2021}{}}%
-Vervoort, R.W., Dolk, M.M., Ogtrop, F.F.
-van, 2021.
-Climate change and other trends in streamflow observations in australian forested catchments since 1970.
-Hydrological Processes 35, e13999.
-doi:\url{https://doi.org/10.1002/hyp.13999}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2006}{}}%
-Wood, S., 2006. Generalized additive models: An introduction with r. CRC Press, Boca Raton, FL.
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhang2011}{}}%
-Zhang, L., Zhao, F., Chen, Y., Dixon, R.N.M., 2011.
-Estimating effects of plantation expansion and climate variability on streamflow for catchments in australia.
-Water Resources Research 47, W12539.
-doi:\href{https://doi.org/10.1029/2011wr010711}{10.1029/2011wr010711}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhang2017}{}}%
-Zhang, M., Liu, N., Harper, R., Li, Q., Liu, K., Wei, X., Ning, D., Hou, Y., Liu, S., 2017.
-A global review on hydrological responses to forest change across multiple spatial scales: Importance of scale, climate, forest type and hydrological regime.
-Journal of Hydrology 546, 44--59.
-doi:\url{https://doi.org/10.1016/j.jhydrol.2016.12.040}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhao2010}{}}%
-Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010.
-Evaluation of methods for estimating the effects of vegetation change and climate variability on streamflow.
-Water Resources Research 46, W03505.
-doi:\href{https://doi.org/10.1029/2009wr007702}{10.1029/2009wr007702}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhou2015}{}}%
-Zhou, G., Wei, X., Chen, X., Zhou, P., Liu, X., Xiao, Y., Sun, G., Scott, D.F., Zhou, S., Han, L., Su, Y., 2015.
-Global pattern for the effect of climate and land cover on water yield.
-Nature Communications 6, 5918.
-doi:\href{https://doi.org/10.1038/ncomms6918}{10.1038/ncomms6918}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhou2010}{}}%
-Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang, C., 2010.
-Forest recovery and river discharge at the regional scale of guangdong province, china.
-Water Resources Research 46.
-doi:\url{https://doi.org/10.1029/2009WR008829}
-
-\end{CSLReferences}
+\renewcommand\refname{References}
+\bibliography{forest\_and\_water.bib}
 

### 2022-08-11 paper/Forest_and_Water/SupplementaryMaterialPart2.Rmd
further updates paper 11082022

@@
 title: 'Supplementary Information part 2: Testing the improved data sets'
-author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
+author:
+  - name: R. Willem Vervoort
+    email: willem.vervoort@sydney.edu.au
+    footnote: 1
+  - name: Eliana Nervi
+    email: eliananervif@gmail.com
+  - name: Jimena Alonso
+    email: jalonso@fing.edu.uy
+footnote:
+  - code: 1
+    text: "Corresponding Author"    
 date: "`r Sys.Date()`"
@@
 
+# Methods
+
 First we will read in the data
@@
 new_data <- read_csv("../../data/NewCatchments_T3.csv") %>%
-  mutate(table = "new catchments")
+  mutate(table = ifelse(`Area(km2)` > 1000, "Large catchments", "Small catchments"))
 
@@
 
-### Implementing the changes to the overall data
+## Implementing the changes to the overall data
 
@@
 
-# Methods
+## Approach and analyses
 
@@
 
-$S_f = \frac{\Delta Q_f}{\Delta F}$
+$S_f = |\frac{\Delta Q_f}{\Delta F}|$
 
@@
   filter(is.na(DeltaF_perc) == F) %>%
-  mutate(S_f = DeltaQf_perc/DeltaF_perc)
+  mutate(S_f = abs(DeltaQf_perc/DeltaF_perc))
 ```
@@
 
-1. The change in flow as a function of change in forest cover
+## The change in stream flow as a function of change in forest cover
 
-First the results for the original (improved) dataset
+Figure \@ref(fig:Fig2Zhang) highlights that the overall relationship in the updated dataset is the same as in @zhang2017.
+This means that while the modifications have cleaned up the transcription errors in the data, they have not fundamentally changed the conclusions in the original paper.
 
-```{r, echo = F, message = F}
+The next figure (Figure \@ref(fig:Fig2Zhangnew)) is the same analysis, but this includes the new data that we identified in papers.
+Again, this figure highlights that the new datasets have not fundamentally changed the relationships found in @zhang2017.
+
+```{r, echo = F, message = F, warning = F}
 png("Fig2Zhang.png", width = 2500, height = 1800, res = 300)
@@
 #  filter(dataset == "original Zhang et al data") %>%
-  ggplot(aes(DeltaF_perc,DeltaQf_perc, colour = forest_sign, shape = table)) + 
+  ggplot(aes(DeltaF_perc,DeltaQf_perc, colour = forest_sign)) + 
   geom_point() +  geom_smooth(method = "lm") +
+  facet_wrap(~table, ncol=1, scales = "free") + 
   theme_bw() + 
@@
 ```
-And then the results for all the catchments including the new "added" catchments.
 
@@
 ```
+## The relationship between the area of the catchment and the sensitivity of streamflow to the change in forest cover.
+
+This analysis replicates Figure 3 in @zhang2017, which investigates for large and small catchments the sensitivity to runoff change from change in forest cover as a function of area.
+Note that in the original figure, the x-axis is on a log scale.
+In the original paper, the analysis is presented for all catchments as well as for large and small catchments.
+Here we only analyse the small and large catchments.
+
+We can see from Figure \@ref(fig:Fig3Zhang) that again the updated database for the original dataset results in little change in the relationships for both large and small catchments.
+However, when the additional new catchments are added to the database (Figure \@ref(fig:Fig3Zhangnew)), the relationships clearly change.
+In particular, for small catchments gaining forest cover, the sensitivity appears positively correlated with the logarithm of the size of the catchments.
+
+```{r, echo = F, message = F, warning = F}
+png("Fig3Zhang.png", width = 2500, height = 1800, res = 300)
+All_data %>%
+  filter(dataset == "original Zhang et al data") %>%
+  ggplot(aes(Area_km2, S_f, colour = forest_sign)) + 
+  geom_point() +  geom_smooth(method = "lm") +
+  scale_x_continuous(trans='log10') +
+  facet_wrap(~table, ncol=1, scales = "free") + 
+  theme_bw() + 
+    theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        legend.text = element_text(size = rel(1.2)),
+        legend.title = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5))) + 
+  # scale_colour_manual(name = "Forest cover direction", 
+  #                     values = c("Decrease" = "red","Increase" = "blue")) +
+  xlab("log10(Area in km2)") + ylab("Sf (sensivity)")
+dev.off()
+
+# and for all data sets
+png("Fig3Zhang_all.png", width = 2500, height = 1800, res = 300)
+All_data %>%
+#  filter(dataset == "original Zhang et al data") %>%
+  ggplot(aes(Area_km2,S_f, colour = forest_sign)) + 
+  geom_point() +  geom_smooth(method = "lm") +
+  scale_x_continuous(trans='log10') +
+  facet_wrap(~table, ncol=1, scales = "free") + 
+  theme_bw() + 
+    theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        legend.text = element_text(size = rel(1.2)),
+        legend.title = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5))) + 
+  # scale_colour_manual(name = "Forest cover direction", 
+  #                     values = c("Decrease" = "red","Increase" = "blue")) +
+  xlab("log10(Area in km2)") + ylab("Sf (sensitivity")
+dev.off()
+
+```
+
+```{r Fig3Zhang, echo =F, fig.cap="Changes in flow based on the catchments from the original data set", out.width="90%"}
+
+include_graphics("Fig3Zhang.png")
+```
+
+```{r Fig3Zhangnew, echo =F, fig.cap="Changes in flow based on the catchments from the extended data set", out.width="90%"}
+
+include_graphics("Fig3Zhang_all.png")
+```
+## The sensitivity to forest loss as a function of dryness
+
+The final analysis that we retest here is the relationship in Figure 4 in the original @zhang2017 paper, which highlights the sensitivity to forest loss as a function of dryness.
+We are again showing just the for the small and large catchments, similar to the original paper.
+
+Similar to earlier analyses in this document Figure \@ref(fig:Fig4Zhang) show that the updated database for the original dataset results in little change in the relationships for both large and small catchments.
+However, when the additional new catchments are added to the database (Figure \@ref(fig:Fig4Zhangnew)), the relationships clearly change.
+In particular, for small catchments both for forest gains and losses the relationship changes and appears stronger.
+
+```{r, echo = F, message = F, warning = F}
+png("Fig4Zhang.png", width = 2500, height = 1800, res = 300)
+All_data %>%
+  filter(dataset == "original Zhang et al data") %>%
+  ggplot(aes(Dryness, S_f, colour = forest_sign)) + 
+  geom_point() +  geom_smooth(method = "lm") +
+  #scale_x_continuous(trans='log10') +
+  facet_wrap(~table, ncol=1, scales = "free") + 
+  theme_bw() + 
+    theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        legend.text = element_text(size = rel(1.2)),
+        legend.title = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5))) + 
+  # scale_colour_manual(name = "Forest cover direction", 
+  #                     values = c("Decrease" = "red","Increase" = "blue")) +
+  xlab("Dryness (E0/P)") + ylab("Sf (sensivity)")
+dev.off()
+
+# and for all data sets
+png("Fig4Zhang_all.png", width = 2500, height = 1800, res = 300)
+All_data %>%
+#  filter(dataset == "original Zhang et al data") %>%
+  ggplot(aes(Dryness,S_f, colour = forest_sign)) + 
+  geom_point() +  geom_smooth(method = "lm") +
+  #scale_x_continuous(trans='log10') +
+  facet_wrap(~table, ncol=1, scales = "free") + 
+  theme_bw() + 
+    theme(axis.title = element_text(size = rel(1.5)),
+        axis.text = element_text(size = rel(1.2)),
+        legend.text = element_text(size = rel(1.2)),
+        legend.title = element_text(size = rel(1.2)),
+        strip.text = element_text(size = rel(1.5))) + 
+  # scale_colour_manual(name = "Forest cover direction", 
+  #                     values = c("Decrease" = "red","Increase" = "blue")) +
+  xlab("Dryness (E0/P)") + ylab("Sf (sensitivity")
+dev.off()
+
+```
+
+```{r Fig4Zhang, echo =F, fig.cap="Changes in flow based on the catchments from the original data set", out.width="90%"}
+
+include_graphics("Fig4Zhang.png")
+```
+
+```{r Fig4Zhangnew, echo =F, fig.cap="Changes in flow based on the catchments from the extended data set", out.width="90%"}
+
+include_graphics("Fig4Zhang_all.png")
+```

### 2022-08-11 paper/Forest_and_Water/SupplementaryMaterialPart2.tex
further updates paper 11082022

@@
             bookmarks=true,
-            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
+            pdfauthor={},
             pdftitle={Supplementary Information part 2: Testing the improved data sets},
@@
   \title{Supplementary Information part 2: Testing the improved data sets}
-    \author[]{%
+    \author[]{R. Willem Vervoort%
+  %
+  \fnref{1}}
+   \ead{willem.vervoort@sydney.edu.au} 
+    \author[]{Eliana Nervi%
   %
   }
-
+   \ead{eliananervif@gmail.com} 
+    \author[]{Jimena Alonso%
+  %
+  }
+   \ead{jalonso@fing.edu.uy} 
       \cortext[cor1]{Corresponding author}
+    \fntext[1]{Corresponding Author}
 
@@
 This is to check how much influence the changes to the data set and the additional data might have changed the original conclusions.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
@@
 \hypertarget{implementing-the-changes-to-the-overall-data}{%
-\subsubsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}
+\subsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}
 
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\hypertarget{approach-and-analyses}{%
+\subsection{Approach and analyses}\label{approach-and-analyses}}
 
@@
 
-\(S_f = \frac{\Delta Q_f}{\Delta F}\)
+\(S_f = |\frac{\Delta Q_f}{\Delta F}|\)
 
@@
   \FunctionTok{filter}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(DeltaF\_perc) }\SpecialCharTok{==}\NormalTok{ F) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{S\_f =}\NormalTok{ DeltaQf\_perc}\SpecialCharTok{/}\NormalTok{DeltaF\_perc)}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{S\_f =} \FunctionTok{abs}\NormalTok{(DeltaQf\_perc}\SpecialCharTok{/}\NormalTok{DeltaF\_perc))}
 \end{Highlighting}
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The change in flow as a function of change in forest cover
-\end{enumerate}
+\hypertarget{the-change-in-stream-flow-as-a-function-of-change-in-forest-cover}{%
+\subsection{The change in stream flow as a function of change in forest cover}\label{the-change-in-stream-flow-as-a-function-of-change-in-forest-cover}}
 
-First the results for the original (improved) dataset
+Figure \ref{fig:Fig2Zhang} highlights that the overall relationship in the updated dataset is the same as in \citet{zhang2017}.
+This means that while the modifications have cleaned up the transcription errors in the data, they have not fundamentally changed the conclusions in the original paper.
+
+The next figure (Figure \ref{fig:Fig2Zhangnew}) is the same analysis, but this includes the new data that we identified in papers.
+Again, this figure highlights that the new datasets have not fundamentally changed the relationships found in \citet{zhang2017}.
 
@@
 
-And then the results for all the catchments including the new ``added'' catchments.
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Fig2Zhang_all} \caption{Changes in flow based on the catchments from the extended data set}\label{fig:Fig2Zhangnew}
+\end{figure}
+
+\hypertarget{the-relationship-between-the-area-of-the-catchment-and-the-sensitivity-of-streamflow-to-the-change-in-forest-cover.}{%
+\subsection{The relationship between the area of the catchment and the sensitivity of streamflow to the change in forest cover.}\label{the-relationship-between-the-area-of-the-catchment-and-the-sensitivity-of-streamflow-to-the-change-in-forest-cover.}}
+
+This analysis replicates Figure 3 in \citet{zhang2017}, which investigates for large and small catchments the sensitivity to runoff change from change in forest cover as a function of area.
+Note that in the original figure, the x-axis is on a log scale.
+In the original paper, the analysis is presented for all catchments as well as for large and small catchments.
+Here we only analyse the small and large catchments.
+
+We can see from Figure \ref{fig:Fig3Zhang} that again the updated database for the original dataset results in little change in the relationships for both large and small catchments.
+However, when the additional new catchments are added to the database (Figure \ref{fig:Fig3Zhangnew}), the relationships clearly change.
+In particular, for small catchments gaining forest cover, the sensitivity appears positively correlated with the logarithm of the size of the catchments.
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{Fig2Zhang_all} \caption{Changes in flow based on the catchments from the extended data set}\label{fig:Fig2Zhangnew}
+\includegraphics[width=0.9\linewidth]{Fig3Zhang} \caption{Changes in flow based on the catchments from the original data set}\label{fig:Fig3Zhang}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Fig3Zhang_all} \caption{Changes in flow based on the catchments from the extended data set}\label{fig:Fig3Zhangnew}
+\end{figure}
+
+\hypertarget{the-sensitivity-to-forest-loss-as-a-function-of-dryness}{%
+\subsection{The sensitivity to forest loss as a function of dryness}\label{the-sensitivity-to-forest-loss-as-a-function-of-dryness}}
+
+The final analysis that we retest here is the relationship in Figure 4 in the original \citet{zhang2017} paper, which highlights the sensitivity to forest loss as a function of dryness.
+We are again showing just the for the small and large catchments, similar to the original paper.
+
+Similar to earlier analyses in this document Figure \ref{fig:Fig4Zhang} show that the updated database for the original dataset results in little change in the relationships for both large and small catchments.
+However, when the additional new catchments are added to the database (Figure \ref{fig:Fig4Zhangnew}), the relationships clearly change.
+In particular, for small catchments both for forest gains and losses the relationship changes and appears stronger.
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Fig4Zhang} \caption{Changes in flow based on the catchments from the original data set}\label{fig:Fig4Zhang}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Fig4Zhang_all} \caption{Changes in flow based on the catchments from the extended data set}\label{fig:Fig4Zhangnew}
 \end{figure}

### 2022-08-12 paper/Forest_and_Water/Forest_and_Water.Rmd
changes 12082022

@@
     correspondingauthor: true
-  - name: R. Willem Vervoort
-    affiliation:
   - name: Eliana Nervi
@@
 address:
+  - code: a
+    address: The University of Sydney, Sydney, NSW 2006, Australia
   - code: b
-    address: The University of Sydney, Sydney, NSW 2006, Australia
-  - code: a
     address: ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of  Agriculture, School of Life and Environmental Sciences.
@@
   - code: d
-    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
+    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la Republica, 11200 Montevideo, Uruguay
 abstract: |
@@
 Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
+keywords: 
+  - keyword1
+  - keyword2
 journal: "Journal of Hydrology"
 date: "`r Sys.Date()`"
-bibliography: forest_and_water.bib
+# classoption: preprint, 3p, authoryear
+bibliography: forestandwater.bib
 linenumbers: true
@@
     base_format: rticles::elsevier_article
-editor_options: 
-  chunk_output_type: inline
-header-includes:
-  - \usepackage{setspace}
 ---
-
-\doublespacing
 
@@
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
-The Fuh model and variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2002 to interpret the global impact of reforestation.
+The Fuh model and variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2022 to interpret the global impact of reforestation.
 
@@
 
-As highlighted, @zhang2015 did not provide values for evapotranspiration in the data base.
+As highlighted, @zhang2017 did not provide values for evapotranspiration in the data base.
 Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
@@
 
-Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap.
+Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @penaarancibia2012, and these were checked against the existing studies to prevent overlap.
 In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
@@
 
-We conducted a thorough review of all the studies mentioned in the data base of @zhang2015 and sourced all the original papers.
+We conducted a thorough review of all the studies mentioned in the data base of @zhang2017 and sourced all the original papers.
 As a result of this we made several changes to the data base, which are all recorded in Supplementary Data part 1.
@@
 ## Statistical modelling
+
 ```{r, include=FALSE}

### 2022-08-12 paper/Forest_and_Water/Forest_and_Water.tex
changes 12082022

@@
 
-\usepackage{setspace}
-
 \begin{document}
@@
    \ead{willem.vervoort@sydney.edu.au} 
-    \author[]{R. Willem Vervoort}
-
     \author[c]{Eliana Nervi}
@@
    \ead{jalonso@fing.edu.uy} 
-      \affiliation[b]{The University of Sydney, Sydney, NSW 2006, Australia}
-    \affiliation[a]{ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
+      \affiliation[a]{The University of Sydney, Sydney, NSW 2006, Australia}
+    \affiliation[b]{ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
     \affiliation[c]{Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
-    \affiliation[d]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
+    \affiliation[d]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la Republica, 11200 Montevideo, Uruguay}
     \cortext[cor1]{Corresponding author}
@@
   \end{abstract}
+    \begin{keyword}
+    keyword1 \sep 
+    keyword2
+  \end{keyword}
 
  \end{frontmatter}
-
-\doublespacing
 
@@
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in \citet{zhang2017}.
-The Fuh model and variations of the Budyko equilibrium modelling approach was also used by \citet{hoekvandijke2002} to interpret the global impact of reforestation.
+The Fuh model and variations of the Budyko equilibrium modelling approach was also used by \citet{hoekvandijke2022} to interpret the global impact of reforestation.
 
@@
 
-As highlighted, \citet{zhang2015} did not provide values for evapotranspiration in the data base.
+As highlighted, \citet{zhang2017} did not provide values for evapotranspiration in the data base.
 Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 \citep{trabucco2018}, if a value of E\textsubscript{0} was not available from the original papers.
@@
 
-Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{pena-arancibia2012}, and these were checked against the existing studies to prevent overlap.
+Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{penaarancibia2012}, and these were checked against the existing studies to prevent overlap.
 In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in \citet{zhang2011}).
@@
 \renewcommand\refname{References}
-\bibliography{forest\_and\_water.bib}
+\bibliography{forestandwater.bib}
 

### 2022-08-18 paper/Forest_and_Water/Forest_and_Water.Rmd
suppl3 adn other changes

@@
 ---
-title: Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.
+title: 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.'
 author:
@@
 - Latent variables that are not included in the typical single covariate analysis;
+- Interpretation errors due to incomplete descriptions of the experiments in the original papers;  
 - Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally    
@@
 The columns in this data set (are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in \@ref(Table 1). These abbreviations will later be used in the models.
-
-Table (\#table:tab1)  Summary of abbreviations of factors used in the @zhang2017 data set
-
-|Factor | Abbreviation | Definition |
-|:------------|:-------|:----------|
-| forest type | CF |coniferous forest | 
-|  | BF | broadleaf forest |
-|  | MF |mixed forest  |
-|hydrological regime|RD|rain dominated|
-| |SD|snow dominated|
-| climate type| EL| energy limited|
-| |WL| water limited|
-| |EQ| equitant|
-|precipitation data type|OB|observed|
-| |SG|spatial gridded|
-| |MD|modelled|
-|assessment technique|PWE|paired watershed experiment|
-| |QPW|quasi-paired watershed experiment|
-| |HM|hydrological modelling|
-| |EA|elastictity analysis|
-| |SH|combined use of statistical methods|
-| | |and hydrographs|
-
-@zhang2017 use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and use the dryness index to identify the climate type.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table @\ref(tab:table1). These abbreviations will later be used in the models.
+
+```{r table1}
+
+Table1 <- tibble("Factor" = c("forest type","","", "hydrological regime","", 
+                              "climate type","","","precipitation data type",
+                              "","","assessment technique", "","","",""),
+                 "Abbreviation" = c("CF","BF","MF","RD","SD", "EL", "WL", "EQ", 
+                                    "OB","SG","MD","PWE","QPW","HM","EA","SH"),
+                 "Definition" = c("coniferous forest", "broadleaf forest", "mixed forest",
+                                  "rain dominated", "snow dominated", "energy limited",
+                                  "water limited", "equitant", "observed", "spatial gridded",
+                                  "modelled", "paired watershed experiment", 
+                                  "quasi-paired watershed experiment", 
+                                  "hydrological modelling", "elasticity analysis",
+                                  "statistical modelling and hydrographs"))
+
+# |:------------|:-------|:----------|
+# | forest type | CF |coniferous forest | 
+# |  | BF | broadleaf forest |
+# |  | MF |mixed forest  |
+# |hydrological regime|RD|rain dominated|
+# | |SD|snow dominated|
+# | climate type| EL| energy limited|
+# | |WL| water limited|
+# | |EQ| equitant|
+# |precipitation data type|OB|observed|
+# | |SG|spatial gridded|
+# | |MD|modelled|
+# |assessment technique|PWE|paired watershed experiment|
+# | |QPW|quasi-paired watershed experiment|
+# | |HM|hydrological modelling|
+# | |EA|elastictity analysis|
+# | |SH|combined use of statistical methods|
+# | | |and hydrographs|
+
+pander(Table1, caption = "Summary of abbreviations of factors used in the Zhang et al. (2017) data set")
+```
+
+The paper by @zhang2017 use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and use the dryness index to identify the climate type.
 However, the potential or reference ET was not originally included as part of the published data set.
@@
 \begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
+\Delta Qf \% \sim &~ \Delta \% forest~cover + \notag \\ 
 & \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
@@
 
+## Statistical analysis
+
+The results of the overall statistical model that includes all the variables (but no interactions) reinforces some of the results from the correlation analysis.
+
+This includes introducing non-linearity (Equation \@ref(eq:eq2)) for the numerical variables in the model.
+While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
+The number of knots in the non-linear variables was based on assessment of the edf and k' **check wood 2006 and explain well**
+
+```{r model_all, echo = F}
+All_data2 <- All_data2 %>%
+  mutate(length = To - From,
+         mid_year = From + (To - From)/2)
+
+Forest_model_all <- gam(DeltaQf_perc ~ DeltaF_perc + 
+                    s(log10(Area_km2), k = 5, bs="ts") + 
+                    s(Dryness, k = 10, bs="ts" ) + 
+                     s(length, k = 35, bs="ts") +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
+                    Hydrological_regime
+                    , data = All_data2)
+#summary(Forest_model_all)
+#gam.check(Forest_model_all)
+#plot(model6_all)
+```
+```{r modelforestpos, eval =F, echo =F}
+Forest_model_all2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
+                    s(log10(Area_km2), k = 5, bs="ts") + 
+                    s(Dryness, k = 10, bs="ts" ) + 
+                     s(length, k = 35, bs="ts") +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
+                    Hydrological_regime
+                    , data = All_data2)
+summary(Forest_model_all2)
+
+```
+```{r modelforestnonlin, eval =F, echo =F}
+Forest_model_nonlin <- gam(DeltaQf_perc ~ s(DeltaF_perc, k=10, bs = "ts") + 
+                    s(log10(Area_km2), k = 5, bs="ts") + 
+                    s(Dryness, k = 10, bs="ts" ) + 
+                     s(length, k = 35, bs="ts") +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
+                    Hydrological_regime
+                    , data = All_data2)
+summary(Forest_model_nonlin)
+plot(Forest_model_nonlin)
+```
+
+```{r, echo = F}
+pander(round(summary(Forest_model_all)$p.table,2), caption = "(\\#tab:m_all-linear) Statistical summary for the linear terms the full model")
+#plot(model6_all)
+```
+
+```{r , echo = F}
+pander(round(summary(Forest_model_all)$s.table,2), caption = "(\\#tab:m_all-smooth) Statistical summary for the smooth terms for the full model")
+#plot(model6_all)
+```
+
+The overall explaining power of the model can be interpreted from the adjusted *r^2^* (which is penalised for the number of parameters).
+This indicates an *r^2^* of `r round(summary(Forest_model_all)$r.sq,2)` and deviance explained is `r round(summary(Forest_model_all)$dev.expl,2)`, suggesting the model only explains about 50% of the variance in the data.
+
+Inspecting the significance of the variables (Table \@ref(tab:m_all-linear) and Table \@ref(tab:m_all-smooth)) indicates some interesting features.
+
+The overall partial slope of the change in forest cover is `r round(summary(Forest_model_all)$p.table[2,1],2)`, if all other variables are kept constant.
+This suggest quite strong change in streamflow, moving from fully forested to fully cleared.
+Over the whole forest cover range, this is a change of `r round(summary(Forest_model_all)$p.table[2,1],2)*200` mm, with other variables held constant.
+This change is highly significant, as indicated by the low  p-value.
+
+In addition, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(Forest_model_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(Forest_model_all)$s.table[2,4],2)`)) and *length* (p = `r round(summary(Forest_model_all)$s.table[3,4],2)`)) explain variation in the data.
+For *length*, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
+
+Furthermore Table \@ref(tab:m_six_all-linear) indicates that several of the assessment methods are significant, which was also indicated in the correlation analysis.
+In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables ($p < 0.05$).
+
+```{r visualisesmooths,echo =F, eval = F}
+#require(visreg)
+png("Forest_model_allsmooths.png", width = 2500, height = 1800, res = 300)
+#par(mfrow=c(2,2))
+D <- visreg(Forest_model_all, "Dryness", gg = T, plot = T) + theme_bw()
+A <- visreg(Forest_model_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
+L <- visreg(Forest_model_all, "length", gg = T, plot = T) + theme_bw()
+#From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
+(D + A) / L 
+# visreg(model6_all,"Dryness",gg=T) + theme_bw()
+# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
+# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
+#par(mfrow=c(1,1))
+dev.off()
+```
+
+```{r smoothsmodelall, echo = F, fig.cap="Visualisation of the smooth variables in the model, the shaded areas are the 95\\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points", out.width="90%"}
+include_graphics("Forest_model_allsmooths.png")
+```
+
+Figure \@ref(fig:smoothsmodelall) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear.
+It indicates the negative slope that was also clear from @zhang2017, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+
+Both the *length* and *Dryness* variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
+A further problem appears to be that *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the relationship.
+
+As discussed in the methods, the single linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives.
+This tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods.
+The results of this showed
+
+<!-- plot the residuals -->
+```{r, eval = F, echo = F}
+png("residual_plot_model_all.png", width = 2500, height = 1800, res = 300)
+par(mfrow=c(2,2))
+gam.check(Forest_model_all)
+par(mfrow=c(1,1))
+dev.off()
+```
+
+```{r gamcheckmodelall, echo =F, fig.cap="Residual plots for the regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
+include_graphics("residual_plot_model_all.png")
+```
+
+As is not always shown, here we show the residual distribution of the model (Figure \@ref(fig:gamcheckmodelall).
+These are approximately normal, although there is a a skew in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall)).
+This is related to a limited number catchments that have very high changes in streamflow in the data set.In other words, the distribution of the residuals is somewhat fat-tailed.
+
+One solution could be to transform the data, however this is not that simple.
+As the data for the change in flow cover the domain $\mathbb{R}$ we can not use a simple log or Gamma transformation and more complex transformations make the results of the regression difficult to interpret.
+
+Given the majority of the residuals indicate a relatively well behaved distribution, we simply note the behaviour at the extremes and will discuss this later in the paper, and explain how this relates to the characteristics of the dataset.
+
+### Test Removal of studies of great lenght and very dry catchments
+
+```{r, echo = F}
+pander(All_data2 %>% filter(Dryness > 5) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
+         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 5")
+```
+
+The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
+The shape of the conditional response (Figure \@ref(fig:smoothsmodelall)) does not reflect a similar response to @filoso2017 and @jackson2005.
+One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \@ref(fig:smoothsmodelall)).
+
+The points related to catchments with very long studies (> 60 years) might be questionable, as changes other than forest cover change could affect streamflow.
+In addition, a few of the catchments have Dryness values that are very large (> 5) and these values have high leverage in the data, affecting the residual distribution.
+These catchments are listed in Table \@ref(tab:drytable), and are three catchments in Arizona and 1 catchment in South Africa.
+
+<!-- reduced length and dryness model -->
+
+```{r model_all_red_Length, echo = F}
+
+NoLong_Data <- All_data2 %>% filter(length<= 60, Dryness <= 5)
+
+model_redLength <- gam(DeltaQf_perc ~ DeltaF_perc + Forest_Sign + 
+                    s(Dryness, k = 10, bs="ts" ) + 
+                      s(log10(Area_km2), k = 5, bs = "ts") +
+                s(length, k = 10, bs = "ts") + 
+                    Precip_data_type +  Assessment_technique +
+                Forest_type +
+                    Hydrological_regime, data = NoLong_Data)
+summary(model_redLength)
+gam.check(model_redLength)
+# plot(model7)
+```
+
+```{r, echo = F}
+pander(round(summary(model_redLength)$p.table,2), caption = "(\\#tab:m_red-linear) Statistical summary for the linear terms the restricted model")
+#plot(model6_all)
+```
+
+```{r, echo = F}
+pander(round(summary(model_redLength)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness <= 5.")
+```
+
+```{r visualisesmoothsOther,echo =F, eval = F}
+#require(visreg)
+png("model_redLength_smooths.png", width = 2500, height = 1800, res = 300)
+#par(mfrow=c(2,2))
+D <- visreg(model_redLength, "Dryness", gg = T, plot = T) + theme_bw()
+A <- visreg(model_redLength, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
+L <- visreg(model_redLength, "length", gg = T, plot = T) + theme_bw()
+#From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
+(D + A) / L#(L + From) 
+# visreg(model6_all,"Dryness",gg=T) + theme_bw()
+# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
+# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
+#par(mfrow=c(1,1))
+dev.off()
+```
+
+```{r smoothsmodelredLength, fig.cap="Visualisation of the smooth variables in the model with reduced data for dryness and length", out.width="90%", echo=F}
+
+include_graphics("model_redLength_smooths.png")
+```
+
+Therefore it is worth investigating what removing these few data points has on the overall model and the significance of the variables.
+A model testing data that have *Dryness* <= 5 and *length* <= 60 years is therefore based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
+
+This last model explains only slightly less of the variation with an adjusted *r^2^* of `r round(summary(model_redLength)$r.sq,2)` and a deviance explained of `r round(summary(model_redLength)$dev.expl,2)`.
+
+Investigating the non-linear responses suggest that *Dryness* has a non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
+Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship visually looks almost linear.
+The variable *length* no longer is significant, after removal of the two studies with very long lengths.
+
+## So far the same as zhang et al. (2017), but is there a problem?
+
+At this point it seems like we are confirming the @zhang2017 and other results. But is there a problem hidden in the analysis?
+
+<!--
+
 ```{r, eval = F, echo = F}
@@
 
-<!--\newpage-->
-
-<!-- include Dryness index -->
-
 ```{r model4, echo = F}
@@
 ```
-
-<!-- running PCA, this is used later -->
 
@@
 
+<!--
 \begin{align}
@@
 
-```{r visualisesmooths,echo =F, eval = F}
+```{r visualisesmoothsblabla,echo =F, eval = F}
 #require(visreg)
@@
 For example, *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the data.
-
-<!--filter length < 60 -->
-```{r model7_nolongLength, echo = F}
-
-NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
-
-model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, k = 10, bs="ts" ) + 
-                #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), k = 5, bs = "ts") +
-                #length +
-                s(length, k = 10, bs = "ts") + #s(From, k=40, bs = "ts") +
-                    Precip_data_type +  Assessment_technique +
-                Forest_type +
-                    Hydrological_regime, data = NoLong_Data)
-# summary(model7)
-# gam.check(model7)
-# plot(model7)
-```
-
-```{r, echo = F}
-pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness < 4.")
-```
-
-```{r visualisesmooths7,echo =F, eval = F}
-#require(visreg)
-png("model7_smooths.png", width = 2500, height = 1800, res = 300)
-#par(mfrow=c(2,2))
-D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
-A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
-#From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
-(D + A) / L#(L + From) 
-# visreg(model6_all,"Dryness",gg=T) + theme_bw()
-# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
-# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
-#par(mfrow=c(1,1))
+-->
+
+```{r, eval = F, echo =F}
+png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
+All_data2 %>%
+  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") + theme_bw() +
+    theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)))
 dev.off()
@@
 
-```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}
-
-include_graphics("model7_smooths.png")
-```
-
-<!--\newpage-->
-
-The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) does not reflect a similar response to @filoso2017 and @jackson2005.
-One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \@ref(fig:smoothsmodel6)).
-Therefore it can be important to investigate what removing these few data points has on the overall model and the significance of the variables.
-The next model therefore removes the following data: *Dryness* > 4 and *length* > 60 years.
-This result in a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
-
-This last model has more explaining power with an adjusted *r^2^* of `r round(summary(model7)$r.sq,2)`.
-The results indicate that *Dryness* indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodel7) and Table \@ref(tab:restrictlength)).
-Catchment area (*log10(Area (km^2^))*) also shows reasonable evidence of having an impact on flow with p = `r round(summary(model7)$s.table[2,4],2)`, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
-The variable *length* no longer is significant, after removal of the two studies with very long lengths.
+```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =T}
+
+include_graphics("AssessmentTechnique_byArea.png")
+```
+
+One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
 
@@
 All_data2 %>%
-  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
+  filter(is.na(Forest_Sign) == F) %>%
+  ggplot(aes(DeltaF_perc, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
              shape = Forest_Sign)) + geom_point(alpha = 0.5) +
@@
 Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamdlow.
+As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamflow.
 
@@
 These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \@ref(fig:overview)), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \@ref(fig:gamcheck7)).
+The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \@ref(fig:gamcheckredLength)).
 
@@
 ```
-```{r gamcheck7, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
+```{r gamcheckredLength, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
 include_graphics("residual_plot_model7.png")
@@
 ```
-
-```{r, eval = F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw() +
-    theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)))
-dev.off()
-```
-
-```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =T}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
-
-One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
 

### 2022-08-18 paper/Forest_and_Water/Forest_and_Water.tex
suppl3 adn other changes

@@
 
+\usepackage{graphicx}
 \ifxetex
@@
 \item
-  Latent variables that are not included in the typical single covariate analysis;\\
+  Latent variables that are not included in the typical single covariate analysis;
+\item
+  Interpretation errors due to incomplete descriptions of the experiments in the original papers;\\
 \item
@@
 
-We conducted a thorough review of all the studies mentioned in the data base of \citet{zhang2015} and sourced all the original papers.
+We conducted a thorough review of all the studies mentioned in the data base of \citet{zhang2017} and sourced all the original papers.
 As a result of this we made several changes to the data base, which are all recorded in Supplementary Data part 1.
@@
 
-\hypertarget{the-general-relationship-between-change-in-forest-cover-and-streamflow}{%
-\subsection{The general relationship between change in forest cover and streamflow}\label{the-general-relationship-between-change-in-forest-cover-and-streamflow}}
-
-Following \citet{zhang2017}, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
-
-\begin{equation}
-\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon \label{eq:eq3}
-\end{equation}
+\hypertarget{statistical-analysis}{%
+\subsection{Statistical analysis}\label{statistical-analysis}}
+
+The results of the overall statistical model that includes all the variables (but no interactions) reinforces some of the results from the correlation analysis.
+
+This includes introducing non-linearity (Equation \eqref{eq:eq7}) for the numerical variables in the model.
+While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
+The number of knots in the non-linear variables was based on assessment of the edf and k' \textbf{check wood 2006 and explain well}
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc + s(log10(Area_km2), k = 5, bs = "ts") + 
+##     s(Dryness, k = 10, bs = "ts") + s(length, k = 35, bs = "ts") + 
+##     Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                              Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                  -7.47200   15.87043  -0.471  0.63814    
+## DeltaF_perc                  -0.59422    0.05459 -10.885  < 2e-16 ***
+## Precip_data_typeOB          -17.87257   12.90741  -1.385  0.16724    
+## Precip_data_typeSG            0.19950   14.82105   0.013  0.98927    
+## Assessment_techniqueEA, HM   18.66830   41.70211   0.448  0.65474    
+## Assessment_techniqueHM       26.60929   11.43273   2.327  0.02065 *  
+## Assessment_techniquePWE      30.77146   11.68272   2.634  0.00890 ** 
+## Assessment_techniquePWE, HM  15.79891   42.22036   0.374  0.70853    
+## Assessment_techniqueQPW      41.34637   19.66208   2.103  0.03636 *  
+## Assessment_techniqueQPW, EA  26.04728   23.83570   1.093  0.27542    
+## Assessment_techniqueSH       39.31195   11.53645   3.408  0.00075 ***
+## Forest_typeCF                -9.28294    7.41227  -1.252  0.21147    
+## Forest_typeMF                -6.34249    7.37916  -0.860  0.39078    
+## Hydrological_regimeSD         0.13379    8.93522   0.015  0.98806    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                       edf Ref.df     F  p-value    
+## s(log10(Area_km2)) 0.7738      4 0.846 0.032280 *  
+## s(Dryness)         4.7084      9 2.183 0.000615 ***
+## s(length)          4.4431     34 0.242 0.089106 .  
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.455   Deviance explained = 49.6%
+## GCV = 1766.8  Scale est. = 1629.1    n = 307
+\end{verbatim}
 
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
-\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
-The first three rows relate to the model using the original data base from Zhang et al.~(2017).
-The bottom three rows are the results of the model including the new data.
-Clearly there is no major change arising from the additional data.}\tabularnewline
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
+\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{(Intercept)} & 13.65 & 5.04 & 2.71 & 0.01 \\
-\textbf{DeltaF\_perc\_pos} & 0.36 & 0.08 & 4.52 & 0 \\
-\textbf{Forest\_SignIncrease} & -43.21 & 5.53 & -7.82 & 0 \\
-\textbf{(Intercept)} & 17.39 & 5.2 & 3.34 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.32 & 0.08 & 4.04 & 0 \\
-\textbf{Forest\_SignIncrease} & -50.63 & 5.29 & -9.57 & 0 \\
-\bottomrule()
-\end{longtable}
-
-The overall variance explained in this model (equation \eqref{eq:eq3}) is not high with an adjusted \emph{r\textsuperscript{2}} of 0.3, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
-The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
-However the change in flow is different for forest cover decreases compared to forest cover increases.
-In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase.
-So roughly speaking, a 1\% forest cover increase on the average decreases flow by \((1 - 0.29)*0.45\%\), while a the percentage forest cover decrease will increase flow by 0.45\%.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model1} \caption{Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution}\label{fig:gamcheck}
-\end{figure}
-
-Of importance here is to highlight the residuals of this regression (equation \eqref{eq:eq3} and Figure \ref{fig:gamcheck}).
-These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \ref{fig:gamcheck}).
-In other words, the distribution of the residuals is somewhat fat-tailed.
-We will discuss this later.
-
-Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
-Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total \citet{zhang2017} data set.
-But this also means that our re-analysis of the data can be directly compared to the original study.
-
-However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being catchment dryness and area (following \citet{zhang2017}), which we will analyse later.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
-\caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 20.28 & 5.45 & 3.72 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.08 & 4.1 & 0 \\
-\textbf{Forest\_SignIncrease} & -50.75 & 5.34 & -9.51 & 0 \\
-\textbf{Latitude} & -0.17 & 0.09 & -1.96 & 0.05 \\
-\textbf{Longitude} & -0.05 & 0.03 & -1.71 & 0.09 \\
-\bottomrule()
-\end{longtable}
-
-There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.31 suggesting further factors influencing the change in streamflow that are currently not included in the model.
-
-\hypertarget{impact-of-climate}{%
-\subsection{Impact of climate}\label{impact-of-climate}}
-
-While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
-Potential evapotranspiration (\emph{E0}) by itself was not significant in the.
-Initially, we also tested models using only the annual average precipitation (\emph{Pa (mm)}), but interactions between precipitation and evapotranspiraton might be captured by the dryness index.
-Both dryness index and \emph{Pa (mm)} were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
-As a result only the dryness index was retained as a climate indicator to align with the earlier work by \citet{zhang2017}.
-Given that Latitude and Longitude were not significant, we dropped these from the model.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness  + \varepsilon \label{eq:eq5}
-\end{align}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
-\caption{\label{tab:out-model4} Results of the model including the dryness index}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 12.33 & 5.98 & 2.06 & 0.04 \\
-\textbf{DeltaF\_perc\_pos} & 0.33 & 0.08 & 4.03 & 0 \\
-\textbf{Forest\_SignIncrease} & -51.91 & 5.33 & -9.75 & 0 \\
-\textbf{Dryness} & 4.17 & 2.58 & 1.61 & 0.11 \\
-\bottomrule()
-\end{longtable}
-
-Similar to \emph{E0} or \emph{Pa\_mm}, the results from this model (equation \eqref{eq:eq5} and Table \ref{tab:out-model4}) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
-This might seem suprising in light of earlier reported results \citep{zhang2017, filoso2017}.
-In this case, the evidence is highly doubtful (p = 0.11).
-However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted \emph{r\textsuperscript{2}} of 0.31.
-As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1250}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4167}}@{}}
-\caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 4}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-Number
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Latitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Longitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Catchment name
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-Number
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Latitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Longitude
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Catchment name
-\end{minipage} \\
-\midrule()
-\endhead
-76 & 34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
-90 & 36.4 & -120.4 & Cantua \\
-225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
-U.S.A, A \\
-226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
-U.S.A, C \\
-295 & 34.43 & -112.3 & White Spar, Ariz., U.S.A, B \\
-356 & -25.75 & 28.23 & Queens river \\
-\bottomrule()
-\end{longtable}
-
-There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (\textgreater{} 4) and these values have high leverage in the data, affecting the residual distribution.
-These catchments are listed in Table \ref{tab:drytable}.
-
-\hypertarget{is-there-a-distinct-effect-of-area}{%
-\subsection{Is there a distinct effect of area?}\label{is-there-a-distinct-effect-of-area}}
-
-The second major variable is the effect of area on the change in flow, following the analysis by \citet{zhang2017} and \citet{filoso2017}.
-Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon \label{eq:eq6}
-\end{align}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
-\caption{\label{tab:out-modelArea} Results of the model including Area and the dryness index}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 16.51 & 6.82 & 2.42 & 0.02 \\
-\textbf{DeltaF\_perc\_pos} & 0.26 & 0.1 & 2.68 & 0.01 \\
-\textbf{Forest\_SignIncrease} & -50.38 & 5.46 & -9.23 & 0 \\
-\textbf{Dryness} & 4.62 & 2.61 & 1.77 & 0.08 \\
-\textbf{log10(Area\_km2)} & -2.04 & 1.62 & -1.26 & 0.21 \\
-\bottomrule()
-\end{longtable}
-
-The results of this model (Equation \eqref{eq:eq6}) indicate there is at least some evidence (p = 0.21) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km\textsuperscript{2})) (Table \ref{tab:out-modelArea}).
-In fact, the results suggests that for every additional 10 km\textsuperscript{2} in catchment size the mean change in flow reduces by 2\%.
-Another interesting fact to note is that with the inclusion of Area (km\textsuperscript{2}) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km\textsuperscript{2})) in the model (Table \ref{tab:out-modelArea-int}) results in the increased evidence (p = 0) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = 0.73)) as an interaction with Dryness.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4000}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1333}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}@{}}
-\caption{\label{tab:out-modelArea-int} Results of the model including an interaction between Area and the dryness index}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 13.7 & 6.98 & 1.96 & 0.05 \\
-\textbf{DeltaF\_perc\_pos} & 0.25 & 0.1 & 2.56 & 0.01 \\
-\textbf{Forest\_SignIncrease} & -48.35 & 5.56 & -8.7 & 0 \\
-\textbf{Dryness} & 7.19 & 2.97 & 2.42 & 0.02 \\
-\textbf{log10(Area\_km2)} & 0.79 & 2.26 & 0.35 & 0.73 \\
-\textbf{Dryness:log10(Area\_km2)} & -2.39 & 1.34 & -1.78 & 0.08 \\
-\bottomrule()
-\end{longtable}
-
-\hypertarget{are-some-of-the-variables-possibly-non-linear}{%
-\subsection{Are some of the variables possibly non-linear?}\label{are-some-of-the-variables-possibly-non-linear}}
-
-The work by \citet{filoso2017} and earlier by \citet{jackson2005} has indicated that the length of the study might influence the response.
-This links to the idea from \citet{kuczera1987} that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also \citet{jackson2005}).
-In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (\emph{Dryness} and \emph{Area}) were considered non-linear.
-As a result a shrinkage smoothing spline \citep{wood2006} was applied to these variables.
-
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon \label{eq:eq7}
-\end{align}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}@{}}
-\caption{\label{tab:mfive-linear} Statistical summary for the linear terms in the model with non-linear terms}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 18.87 & 5.37 & 3.52 & 0 \\
-\textbf{DeltaF\_perc\_pos} & 0.32 & 0.09 & 3.77 & 0 \\
-\textbf{Forest\_SignIncrease} & -53.75 & 6.15 & -8.74 & 0 \\
+\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
+\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
+\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
+\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
+\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
+\textbf{Assessment\_techniquePWE,
+HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
+\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
+\textbf{Assessment\_techniqueQPW,
+EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
+\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
+\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
+\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
+\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
 \bottomrule()
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{\label{tab:mfive-smooth} Statistical summary for the smooth terms in the model with non-linear terms}\tabularnewline
+\caption{(\#tab:m\_all-smooth) Statistical summary for the smooth terms for the full model}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.22 & 5 & 0.08 & 0.18 \\
-\textbf{s(Dryness)} & 4.94 & 9 & 1.63 & 0.01 \\
-\textbf{s(length)} & 9.9 & 49 & 0.3 & 0.12 \\
+\textbf{s(log10(Area\_km2))} & 0.77 & 4 & 0.85 & 0.03 \\
+\textbf{s(Dryness)} & 4.71 & 9 & 2.18 & 0 \\
+\textbf{s(length)} & 4.44 & 34 & 0.24 & 0.09 \\
 \bottomrule()
@@
 
-Including non-linearity (Equation \eqref{eq:eq7}) increases the overall explaining power of the model to an adjusted \emph{r\textsuperscript{2}} of 0.35 and deviance explained of 0.39, but creates a few changes in the significance of the variables (Table \ref{tab:mfive-smooth}).
-For example, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.18)), \emph{Dryness} (p = 0.01)) and \emph{length} (p = 0.12)) explain significant variation in the data.
-
-However, including the non-linearity also increases the chance of over fitting, as the smoothing splines allow significant flexibility.
-Including interactions between the smooth variables is also possible, but the results are difficult to interpret given the high flexibility of the two-dimensional smooth.
-Given the overall variability in the data we did not attempt this.
-
-Finally the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included i.e.~Equation \eqref{eq:eq2}.
+The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters).
+This indicates an \emph{r\textsuperscript{2}} of 0.46 and deviance explained is 0.5, suggesting the model only explains about 50\% of the variance in the data.
+
+Inspecting the significance of the variables (Table @ref(tab:m\_all-linear) and Table @ref(tab:m\_all-smooth)) indicates some interesting features.
+
+The overall partial slope of the change in forest cover is -0.59, if all other variables are kept constant.
+This suggest quite strong change in streamflow, moving from fully forested to fully cleared.
+Over the whole forest cover range, this is a change of -118 mm, with other variables held constant.
+This change is highly significant, as indicated by the low p-value.
+
+In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.03)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.09)) explain variation in the data.
+For \emph{length}, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
+
+Furthermore Table @ref(tab:m\_six\_all-linear) indicates that several of the assessment methods are significant, which was also indicated in the correlation analysis.
+In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables (\(p < 0.05\)).
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{Forest_model_allsmooths} \caption{Visualisation of the smooth variables in the model, the shaded areas are the 95\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points}\label{fig:smoothsmodelall}
+\end{figure}
+
+Figure \ref{fig:smoothsmodelall} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear.
+It indicates the negative slope that was also clear from \citet{zhang2017}, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+
+Both the \emph{length} and \emph{Dryness} variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
+A further problem appears to be that \emph{length} and \emph{Dryness} have several points with very high leverage that determine much of the non-linearity in the relationship.
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{residual_plot_model_all} \caption{Residual plots for the regression model indicating a slightly fat-tailed residual distribution}\label{fig:gamcheckmodelkall}
+\end{figure}
+
+As is not always shown, here we show the residual distribution of the model (Figure \ref{fig:gamcheckmodelall}.
+These are approximately normal, although there is a a skew in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}).
+This is related to a limited number catchments that have very high changes in streamflow in the data set.In other words, the distribution of the residuals is somewhat fat-tailed.
+
+One solution could be to transform the data, however this is not that simple.
+As the data for the change in flow cover the domain \(\mathbb{R}\) we can not use a simple log or Gamma transformation and more complex transformations make the results of the regression difficult to interpret.
+
+Given the majority of the residuals indicate a relatively well behaved distribution, we simply note the behaviour at the extremes and will discuss this later in the paper, and explain how this relates to the characteristics of the dataset.
+
+\hypertarget{test-removal-of-studies-of-great-lenght-and-very-dry-catchments}{%
+\subsubsection{Test Removal of studies of great lenght and very dry catchments}\label{test-removal-of-studies-of-great-lenght-and-very-dry-catchments}}
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1250}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1528}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3889}}@{}}
+\caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 5}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Number
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Latitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Longitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Catchment name
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Number
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Latitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Longitude
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Catchment name
+\end{minipage} \\
+\midrule()
+\endhead
+76 & 34.67 & -111.7 & Beaver Creek, AZ \#3-2 \\
+225 & 32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, A \\
+226 & 32.74 & -111.5 & Natural Drainages, Ariz.,
+U.S.A, C \\
+356 & -25.75 & 28.23 & Queens river \\
+\bottomrule()
+\end{longtable}
+
+The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
+The shape of the conditional response (Figure \ref{fig:smoothsmodelall}) does not reflect a similar response to \citet{filoso2017} and \citet{jackson2005}.
+One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \ref{fig:smoothsmodelall}).
+
+The points related to catchments with very long studies (\textgreater{} 60 years) might be questionable, as changes other than forest cover change could affect streamflow.
+In addition, a few of the catchments have Dryness values that are very large (\textgreater{} 5) and these values have high leverage in the data, affecting the residual distribution.
+These catchments are listed in Table \ref{tab:drytable}, and are three catchments in Arizona and 1 catchment in South Africa.
+
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc + Forest_Sign + s(Dryness, k = 10, 
+##     bs = "ts") + s(log10(Area_km2), k = 5, bs = "ts") + s(length, 
+##     k = 10, bs = "ts") + Precip_data_type + Assessment_technique + 
+##     Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                              Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                 -10.27984   17.60496  -0.584 0.559747    
+## DeltaF_perc                  -0.58628    0.07722  -7.592 4.73e-13 ***
+## Forest_SignIncrease           0.92751    9.57534   0.097 0.922903    
+## Precip_data_typeOB          -12.54417   12.19935  -1.028 0.304714    
+## Precip_data_typeSG            5.90459   15.05545   0.392 0.695217    
+## Assessment_techniqueEA, HM   18.85724   39.89593   0.473 0.636824    
+## Assessment_techniqueHM       29.54406   11.08107   2.666 0.008119 ** 
+## Assessment_techniquePWE      24.55910   12.28800   1.999 0.046618 *  
+## Assessment_techniquePWE, HM  13.22185   40.95225   0.323 0.747043    
+## Assessment_techniqueQPW      44.21273   18.99855   2.327 0.020671 *  
+## Assessment_techniqueQPW, EA  25.53696   22.81154   1.119 0.263899    
+## Assessment_techniqueSH       40.88820   11.15628   3.665 0.000296 ***
+## Forest_typeCF               -10.21893    7.12542  -1.434 0.152648    
+## Forest_typeMF                -3.90403    7.14361  -0.547 0.585153    
+## Hydrological_regimeSD        -0.02434    8.65352  -0.003 0.997757    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                          edf Ref.df     F  p-value    
+## s(Dryness)         4.070e+00      9 2.127 0.000407 ***
+## s(log10(Area_km2)) 1.574e+00      4 1.865 0.009339 ** 
+## s(length)          6.223e-07      9 0.000 0.857556    
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.455   Deviance explained = 49.1%
+## GCV = 1612.2  Scale est. = 1501.2    n = 300
+\end{verbatim}
+
+\includegraphics{Forest_and_Water_files/figure-latex/model_all_red_Length-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Smoothing parameter selection converged after 19 iterations.
+## The RMS GCV score gradient at convergence was 7.893755e-05 .
+## The Hessian was positive definite.
+## Model rank =  37 / 37 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##                          k'      edf k-index p-value  
+## s(Dryness)         9.00e+00 4.07e+00    0.93   0.130  
+## s(log10(Area_km2)) 4.00e+00 1.57e+00    0.98   0.330  
+## s(length)          9.00e+00 6.22e-07    0.89   0.045 *
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
-\caption{\label{tab:msix-linear} Statistical summary for the linear terms the full model}\tabularnewline
+\caption{(\#tab:m\_red-linear) Statistical summary for the linear terms the restricted model}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
-\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
-\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
-\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
-\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
-\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
-\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
+\textbf{(Intercept)} & -10.28 & 17.6 & -0.58 & 0.56 \\
+\textbf{DeltaF\_perc} & -0.59 & 0.08 & -7.59 & 0 \\
+\textbf{Forest\_SignIncrease} & 0.93 & 9.58 & 0.1 & 0.92 \\
+\textbf{Precip\_data\_typeOB} & -12.54 & 12.2 & -1.03 & 0.3 \\
+\textbf{Precip\_data\_typeSG} & 5.9 & 15.06 & 0.39 & 0.7 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.86 & 39.9 & 0.47 & 0.64 \\
+\textbf{Assessment\_techniqueHM} & 29.54 & 11.08 & 2.67 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 24.56 & 12.29 & 2 & 0.05 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
-\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
+HM} & 13.22 & 40.95 & 0.32 & 0.75 \\
+\textbf{Assessment\_techniqueQPW} & 44.21 & 19 & 2.33 & 0.02 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
-\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
-\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
-\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
-\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
+EA} & 25.54 & 22.81 & 1.12 & 0.26 \\
+\textbf{Assessment\_techniqueSH} & 40.89 & 11.16 & 3.67 & 0 \\
+\textbf{Forest\_typeCF} & -10.22 & 7.13 & -1.43 & 0.15 \\
+\textbf{Forest\_typeMF} & -3.9 & 7.14 & -0.55 & 0.59 \\
+\textbf{Hydrological\_regimeSD} & -0.02 & 8.65 & 0 & 1 \\
 \bottomrule()
 \end{longtable}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{\label{tab:msix-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{s(log10(Area\_km2))} & 0.7 & 4 & 0.4 & 0.12 \\
-\textbf{s(Dryness)} & 4.26 & 9 & 1.75 & 0 \\
-\textbf{s(length)} & 16.71 & 34 & 0.92 & 0.01 \\
-\bottomrule()
-\end{longtable}
-
-This model (Tables \ref{tab:msix-linear} and \ref{tab:msix-smooth}) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted \emph{r\textsuperscript{2}} of 0.39.
-This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
-However, it is interesting to note from Table \ref{tab:msix-linear} that several of the assessment methods are significant.
-In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant (\(p < 0.05\)).
-In this case, \emph{log10(Area (km\textsuperscript{2}))} is no longer a significant predictor, the reasons for this will be discussed later.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{model6_smooths} \caption{Visualisation of the smooth variables in the model, the shaded areas are the 95\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points}\label{fig:smoothsmodel6}
-\end{figure}
-
-\newpage
-
-Figure \ref{fig:smoothsmodel6} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear, but, given all the data, not significant at p = 0.12, likely due to the high variance in the data.
-It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
-Both the \emph{length} and \emph{Dryness} variables are significant and show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
-For example, \emph{length} and \emph{Dryness} have several points with very high leverage that determine much of the non-linearity in the data.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness \textless{} 4.}\tabularnewline
+\caption{\label{tab:restrictlength} Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years and Dryness \textless= 5.}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{s(Dryness)} & 3.54 & 9 & 3.13 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.73 & 4 & 0.58 & 0.07 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.7 \\
+\textbf{s(Dryness)} & 4.07 & 9 & 2.13 & 0 \\
+\textbf{s(log10(Area\_km2))} & 1.57 & 4 & 1.86 & 0.01 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.86 \\
 \bottomrule()
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{model7_smooths} \caption{Visualisation of the smooth variables in the model wiht reduced data for dryness and length}\label{fig:smoothsmodel7}
+\includegraphics[width=0.9\linewidth]{model_redLength_smooths} \caption{Visualisation of the smooth variables in the model with reduced data for dryness and length}\label{fig:smoothsmodelredLength}
 \end{figure}
 
-The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \ref{fig:smoothsmodel6}) does not reflect a similar response to \citet{filoso2017} and \citet{jackson2005}.
-One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \ref{fig:smoothsmodel6}).
-Therefore it can be important to investigate what removing these few data points has on the overall model and the significance of the variables.
-The next model therefore removes the following data: \emph{Dryness} \textgreater{} 4 and \emph{length} \textgreater{} 60 years.
-This result in a reduction of the data set from 329 to 308 catchments.
-
-This last model has more explaining power with an adjusted \emph{r\textsuperscript{2}} of 0.37.
-The results indicate that \emph{Dryness} indicates a clear signifcant non-linear response where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodel7} and Table \ref{tab:restrictlength}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) also shows reasonable evidence of having an impact on flow with p = 0.07, and suggesting once again that changes in forest cover in larger catchments have less impact on streamflow.
+Therefore it is worth investigating what removing these few data points has on the overall model and the significance of the variables.
+A model testing data that have \emph{Dryness} \textless= 5 and \emph{length} \textless= 60 years is therefore based on a reduction of the data set from 329 to 310 catchments.
+
+This last model explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.46 and a deviance explained of 0.49.
+
+Investigating the non-linear responses suggest that \emph{Dryness} has a non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship visually looks almost linear.
 The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
+
+\hypertarget{so-far-the-same-as-zhang-et-al.-2017-but-is-there-a-problem}{%
+\subsection{So far the same as zhang et al.~(2017), but is there a problem?}\label{so-far-the-same-as-zhang-et-al.-2017-but-is-there-a-problem}}
+
+At this point it seems like we are confirming the \citet{zhang2017} and other results. But is there a problem hidden in the analysis?
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments}\label{fig:assessment}
+\end{figure}
+
+One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
 
@@
 These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \ref{fig:overview}), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheck7}).
+The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheckredLength}).
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheck7}
+\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheckredLength}
 \end{figure}
@@
 \end{longtable}
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments}\label{fig:assessment}
-\end{figure}
-
-One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
 

### 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart1.Rmd
suppl3 adn other changes

@@
 title: 'Supplementary Information part 1: Changes to original database'
-author: "Willem Vervoort/Eliana Nervi/Jimena Alonso"
+author:
+  - name: R. Willem Vervoort
+    email: willem.vervoort@sydney.edu.au
+    footnote: 1
+  - name: Eliana Nervi
+    email: eliananervif@gmail.com
+  - name: Jimena Alonso
+    email: jalonso@fing.edu.uy
+footnote:
+  - code: 1
+    text: "Corresponding Author"  
 date: "`r Sys.Date()`"
@@
 
+# Introduction
+
+This supplementary material is related to 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
+
 This document list all the changes that were made to the original @zhang2017 database after review of all the original papers.
+
+# Discussion of changes
 

### 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart2.Rmd
suppl3 adn other changes

@@
   - \usepackage{color}
+  - \newcommand{\beginsupplement}{  \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 abstract: |
@@
 Single variable regressions on the smaller (original) dataset are compared with the extended data set.
+
 ---
+\beginsupplement
 
 # Introduction
+
+This supplementary material is related to 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
 
@@
 
-## Implementing the changes to the overall data
+# Implementing the changes to the overall data
 

### 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart2.tex
suppl3 adn other changes

@@
 \usepackage{color}
+\newcommand{\beginsupplement}{  \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 
@@
    \ead{willem.vervoort@sydney.edu.au} 
-    \author[]{Eliana Nervi%
-  %
-  }
+    \author[]{Eliana Nervi}
    \ead{eliananervif@gmail.com} 
-    \author[]{Jimena Alonso%
-  %
-  }
+    \author[]{Jimena Alonso}
    \ead{jalonso@fing.edu.uy} 
@@
 
+\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}
+
 \hypertarget{introduction}{%
 \section{Introduction}\label{introduction}}
+
+This supplementary material is related to `Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
 
@@
 \hypertarget{implementing-the-changes-to-the-overall-data}{%
-\subsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}
+\section{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}
 

### 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart3.Rmd
suppl3 adn other changes

---
title: 'Supplementary Information part 3: Testing different forest cover relationships'
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    footnote: 1
  - name: Eliana Nervi
    email: eliananervif@gmail.com
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
footnote:
  - code: 1
    text: "Corresponding Author"    
date: "`r Sys.Date()`"
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forestandwater.bib
editor_options: 
  chunk_output_type: inline
header-includes:
  - \usepackage{setspace}
  - \usepackage{color}
  - \newcommand{\beginsupplement}{  \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}\setcounter{equation}{0} \renewcommand{\theequation}{S\arabic{equation}}}
abstract: |
  This supplementary material file compares different linear and non-linear relationships for the impact of forest cover on the change in stream flow.
---
\beginsupplement

# Introduction

This supplementary material is related to 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'

In this document we tested the following, which is outlined in the methods of the paper.
The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. Here we test two variation of this approach, leading to three approaches

1. Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following @zhang2017.

2. The impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyse this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.

3. A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).

# Methods

## Approach 1

To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].

The general model tested is:

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq1)
\end{align}

## Approach 2

Deforestation is different from reforestation. This can be tested by converting all the change in forest cover data to positive values, and an additional binary column ($sign_{forest cover}$) can be included indicating whether it was a forest cover increase or decrease. In the model, the parameter for $sign_{forest cover}$ will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover. The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).

This results in the following model:

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest cover} + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
\end{align}

## Approach 3

The relationship between the change in forest cover and change in streamflow is non-linear. The non-linear approach involves adding a smoothing function the the change in forestry. A shrinkage spline [@wood2006] is applied to allow the non-linear effect to be removed.  Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous. The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form. This leads to the following model:

\begin{align}
\Delta Qf \% \sim &~ s(\Delta \% forest~cover) + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq3)
\end{align}

## Preliminary data management

First we will read in the data

```{r setup, include=FALSE, echo =T}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
require(mgcv)
require(visreg)
require(patchwork)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))

Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv") %>%
  mutate(table = "Small catchments")
Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv") %>%
  mutate(table = "Large catchments")
new_data <- read_csv("../../data/NewCatchments_T3.csv") %>%
  mutate(table = ifelse(`Area(km2)` > 1000, "Large catchments", "Small catchments"))

# change some column names
names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
names(new_data)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
names(new_data)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(new_data)[9:10] <- c("Precip_data_type", "Assessment_technique")

# combine the data into one dataset

```

We will combine the different tables, but will keep an indicator to see where the data are from.

```{r, echo = T, warning = F, message = F}
Zhang_small$From <- as.numeric(Zhang_small$From)
Zhang_small$To <- as.numeric(Zhang_small$To)
Zhang_all <- bind_rows(Zhang_large,Zhang_small) %>%
  mutate(dataset = "original Zhang et al data")
new_data <- new_data %>%
  mutate(dataset = "new data")
All_data <- bind_rows(Zhang_all, new_data)
```

## Implementing the changes to the overall data

The following code implements the changes described in the Supplementary data part 1. However, many of the changes were implemented manually into the data set. These are simply the remaining changes not implemented manually.

1. removing the duplicates.  
```{r, echo = T}
All_data <- All_data %>%
  mutate(`Possible duplicate` = 
           ifelse(is.na(`Possible duplicate`)==T,0,`Possible duplicate`),
         `Possible duplicate` = as.numeric(`Possible duplicate`)) %>%
  filter(`Possible duplicate` != 1)
```

2. calculating the dryness

```{r, echo = T}
# calculate dryness index
All_data <- All_data %>%
  mutate(Dryness = E0/Pa_mm)
```

3. remove watershed 1 (the Amazon) from the analysis  
```{r, echo = T}
All_data <- All_data %>%
  filter(`Watershed #` != 1)
```

4. remove data set 188 and 254 Kamakia and Sambret
```{r,echo = T}
All_data <- All_data %>%
  filter(`Watershed #` != 188) %>%
  filter(`Watershed #` != 254)
```

5. add a column that indicates forst loss of forest gain

```{r, echo = T}
All_data <- All_data %>%
  mutate(forest_sign = ifelse(DeltaF_perc < 0, "Forest Cover Loss",
                              "Forest Cover Gain"))
```

# Results

## Approach 1: linear relationship across all changes in forest cover

First the variable `length` is added, representing the number of years that the data represents.

```{r model_all, echo = T}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "Decrease", "Increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
All_data2 <- All_data2 %>%
  mutate(length = To - From,
         mid_year = From + (To - From)/2)

Forest_model_all <- gam(DeltaQf_perc ~ DeltaF_perc + 
                    s(log10(Area_km2), k = 5, bs="ts") + 
                    s(Dryness, k = 10, bs="ts" ) + 
                     s(length, k = 35, bs="ts") +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data2)
#summary(Forest_model_all)
#gam.check(Forest_model_all)
#plot(model6_all)
```

```{r, echo = F}
pander(round(summary(Forest_model_all)$p.table,2), caption = "(\\#tab:m_all-linear) Statistical summary for the linear terms the full model")
#plot(model6_all)
```

This is the baseline result as it echoes @zhang2017 but in this case the relationship is tested together with other variables as is discussed in the main paper.

For this model, the focus can be on the linear variables. The variable `DeltaF_perc` clearly explains the variation in the change in flow with a p-value of `r round(summary(Forest_model_all)$p.table[2,4],2)`.  The overall model explains `r round(summary(Forest_model_all)$dev.expl,2)*100` \% of the variation in the data. For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used. The AIC of the model is `r AIC(Forest_model_all)`.

## The change in forest cover converted to positive values

```{r modelforestpos, eval =T, echo =T}
Forest_model_all2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), k = 5, bs="ts") + 
                    s(Dryness, k = 10, bs="ts" ) + 
                     s(length, k = 35, bs="ts") +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data2)
#summary(Forest_model_all2)
```

```{r, echo = F}
pander(round(summary(Forest_model_all2)$p.table,2), caption = "(\\#tab:m_all2-linear) Statistical summary for the linear terms the alternative model where the change is forestry is all positive")
#plot(model6_all)
```

Again the focus can be on the linear variables. In this case the variable `DeltaF_perc_pos` also is important to explain the variation in the change in flow with a p-value of `r round(summary(Forest_model_all2)$p.table[2,4],2)`. However, the results also show that in this case, the deforestation (`Forest_Sign` < 0) has a different effect than reforestation (`Forest_Sign` > 0). In fact, in absolute terms deforestration creates a `r round(summary(Forest_model_all2)$p.table[3,1],2)*100` \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).

However, overall this model explains less of the overall variation in the data with a deviance explained of `r round(summary(Forest_model_all2)$dev.expl,2)` and an AIC of `r AIC(Forest_model_all2)`.

## Assuming the change in forest cover is non-linear

The difference between reforestation and deforestation might suggest that the relationship between the change in forest cover and change in stream flow is non-linear. This is tested in the model where the change in forest cover is considered non-linear. The number of knots was decided based on checking the edf value relative to the apparent knots k' wiht the procedure `gam.check()` in the package `mgcv` [@wood2006].

```{r modelforestnonlin, eval =T, echo = T}
Forest_model_nonlin <- gam(DeltaQf_perc ~ s(DeltaF_perc, k=10, bs = "ts") + 
                    s(log10(Area_km2), k = 5, bs="ts") + 
                    s(Dryness, k = 10, bs="ts" ) + 
                     s(length, k = 35, bs="ts") +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data2)
#summary(Forest_model_nonlin)
```

```{r , echo = F}
pander(round(summary(Forest_model_nonlin)$s.table,2), caption = "(\\#tab:m_nonlin-smooth) Statistical summary for the smooth terms for the model that considers the change in forest cover to be a non-linear variable")
#plot(model6_all)
```

In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of `r round(summary(Forest_model_nonlin)$s.table[1,4],2)`. The total deviance explained for the model is `r round(summary(Forest_model_nonlin)$dev.expl,2)` and the model has an AIC of `r AIC(Forest_model_nonlin)`. These performance statistics are similar to the first (baseline) model.

One thing that is of interest is the fact that while a non-linear relationship between the change in forest cover and the change in streamflow was conceptualised, the actual relationship turns out to be linear. This can be shown from the plot and by running `gam.check`.

```{r}
D <- visreg(Forest_model_nonlin, "DeltaF_perc", gg = T, plot = T) + theme_bw()
A <- visreg(Forest_model_nonlin, "Area_km2", gg = T, plot = T) + theme_bw() + 
  scale_x_log10()

(A + D)

```
The procedure `gam.check` highlights that the p-value of the Basis dimension check of the smooth on `DeltaF_perc` is very small and the actual edf is close to 1 suggesting a linear relationship, similar to the smooth on log10(`Area_km2`).

The residual analysis shows that including this non-linearity makes no difference in terms of he distribution of the residuals. While there is some remaining weak trend in the data, the overall distribution of the residuals is approximately normal.

```{r}
par(mfrow= c(2,2))
gam.check(Forest_model_nonlin)
```

# References

### 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart3.tex
suppl3 adn other changes

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Supplementary Information part 3: Testing different forest cover relationships},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{setspace}
\usepackage{color}
\newcommand{\beginsupplement}{  \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}\setcounter{equation}{0} \renewcommand{\theequation}{S\arabic{equation}}}

\begin{document}

\begin{frontmatter}

  \title{Supplementary Information part 3: Testing different forest cover relationships}
    \author[]{R. Willem Vervoort%
  %
  \fnref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[]{Eliana Nervi}
   \ead{eliananervif@gmail.com} 
    \author[]{Jimena Alonso}
   \ead{jalonso@fing.edu.uy} 
      \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}

  \begin{abstract}
  This supplementary material file compares different linear and non-linear relationships for the impact of forest cover on the change in stream flow.
  \end{abstract}

 \end{frontmatter}

\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}\setcounter{equation}{0} \renewcommand{\theequation}{S\arabic{equation}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This supplementary material is related to `Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'

In this document we tested the following, which is outlined in the methods of the paper.
The changes in forest cover contain both positive (forestation) and negative values (deforestation). In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. Here we test two variation of this approach, leading to three approaches

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following \citet{zhang2017}.
\item
  The impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyse this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
\item
  A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).
\end{enumerate}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{approach-1}{%
\subsection{Approach 1}\label{approach-1}}

To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) \citep{wood2006}.

The general model tested is:

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq1}
\end{align}

\hypertarget{approach-2}{%
\subsection{Approach 2}\label{approach-2}}

Deforestation is different from reforestation. This can be tested by converting all the change in forest cover data to positive values, and an additional binary column (\(sign_{forest cover}\)) can be included indicating whether it was a forest cover increase or decrease. In the model, the parameter for \(sign_{forest cover}\) will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover. The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).

This results in the following model:

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest cover} + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq2}
\end{align}

\hypertarget{approach-3}{%
\subsection{Approach 3}\label{approach-3}}

The relationship between the change in forest cover and change in streamflow is non-linear. The non-linear approach involves adding a smoothing function the the change in forestry. A shrinkage spline \citep{wood2006} is applied to allow the non-linear effect to be removed. Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous. The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form. This leads to the following model:

\begin{align}
\Delta Qf \% \sim &~ s(\Delta \% forest~cover) + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq3}
\end{align}

\hypertarget{preliminary-data-management}{%
\subsection{Preliminary data management}\label{preliminary-data-management}}

First we will read in the data

We will combine the different tables, but will keep an indicator to see where the data are from.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{From }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{From)}
\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{To }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{To)}
\NormalTok{Zhang\_all }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_large,Zhang\_small) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"original Zhang et al data"}\NormalTok{)}
\NormalTok{new\_data }\OtherTok{\textless{}{-}}\NormalTok{ new\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"new data"}\NormalTok{)}
\NormalTok{All\_data }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_all, new\_data)}
\end{Highlighting}
\end{Shaded}

\hypertarget{implementing-the-changes-to-the-overall-data}{%
\subsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}

The following code implements the changes described in the Supplementary data part 1. However, many of the changes were implemented manually into the data set. These are simply the remaining changes not implemented manually.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  removing the duplicates.
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} 
           \FunctionTok{ifelse}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)}\SpecialCharTok{==}\NormalTok{T,}\DecValTok{0}\NormalTok{,}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{),}
         \StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{as.numeric}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  calculating the dryness
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# calculate dryness index}
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Dryness =}\NormalTok{ E0}\SpecialCharTok{/}\NormalTok{Pa\_mm)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  remove watershed 1 (the Amazon) from the analysis
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  remove data set 188 and 254 Kamakia and Sambret
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{188}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{254}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  add a column that indicates forst loss of forest gain
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{forest\_sign =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{, }\StringTok{"Forest Cover Loss"}\NormalTok{,}
                              \StringTok{"Forest Cover Gain"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{approach-1-linear-relationship-across-all-changes-in-forest-cover}{%
\subsection{Approach 1: linear relationship across all changes in forest cover}\label{approach-1-linear-relationship-across-all-changes-in-forest-cover}}

First the variable \texttt{length} is added, representing the number of years that the data represents.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Forest\_Sign =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{,}
                              \StringTok{"Decrease"}\NormalTok{, }\StringTok{"Increase"}\NormalTok{),}
         \AttributeTok{DeltaF\_perc\_pos =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{,}
                                  \SpecialCharTok{{-}}\DecValTok{1}\SpecialCharTok{*}\NormalTok{DeltaF\_perc,}
\NormalTok{                                  DeltaF\_perc))}
\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data2 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{length =}\NormalTok{ To }\SpecialCharTok{{-}}\NormalTok{ From,}
         \AttributeTok{mid\_year =}\NormalTok{ From }\SpecialCharTok{+}\NormalTok{ (To }\SpecialCharTok{{-}}\NormalTok{ From)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}

\NormalTok{Forest\_model\_all }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ DeltaF\_perc }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
\NormalTok{                    Forest\_type }\SpecialCharTok{+}
\NormalTok{                    Hydrological\_regime}
\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
\CommentTok{\#summary(Forest\_model\_all)}
\CommentTok{\#gam.check(Forest\_model\_all)}
\CommentTok{\#plot(model6\_all)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
\textbf{Assessment\_techniquePWE,
HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
\textbf{Assessment\_techniqueQPW,
EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
\bottomrule()
\end{longtable}

This is the baseline result as it echoes \citet{zhang2017} but in this case the relationship is tested together with other variables as is discussed in the main paper.

For this model, the focus can be on the linear variables. The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0. The overall model explains 50 \% of the variation in the data. For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used. The AIC of the model is 3166.6801428.

\hypertarget{the-change-in-forest-cover-converted-to-positive-values}{%
\subsection{The change in forest cover converted to positive values}\label{the-change-in-forest-cover-converted-to-positive-values}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Forest\_model\_all2 }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ DeltaF\_perc\_pos }\SpecialCharTok{+}\NormalTok{ Forest\_Sign }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
\NormalTok{                    Forest\_type }\SpecialCharTok{+}
\NormalTok{                    Hydrological\_regime}
\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
\CommentTok{\#summary(Forest\_model\_all2)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
\caption{(\#tab:m\_all2-linear) Statistical summary for the linear terms the alternative model where the change is forestry is all positive}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule()
\endhead
\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
\textbf{Assessment\_techniquePWE,
HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
\textbf{Assessment\_techniqueQPW,
EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
\bottomrule()
\end{longtable}

Again the focus can be on the linear variables. In this case the variable \texttt{DeltaF\_perc\_pos} also is important to explain the variation in the change in flow with a p-value of 0. However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0). In fact, in absolute terms deforestration creates a -5226 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).

However, overall this model explains less of the overall variation in the data with a deviance explained of 0.46 and an AIC of 3212.700798.

\hypertarget{assuming-the-change-in-forest-cover-is-non-linear}{%
\subsection{Assuming the change in forest cover is non-linear}\label{assuming-the-change-in-forest-cover-is-non-linear}}

The difference between reforestation and deforestation might suggest that the relationship between the change in forest cover and change in stream flow is non-linear. This is tested in the model where the change in forest cover is considered non-linear. The number of knots was decided based on checking the edf value relative to the apparent knots k' wiht the procedure \texttt{gam.check()} in the package \texttt{mgcv} \citep{wood2006}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Forest\_model\_nonlin }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{s}\NormalTok{(DeltaF\_perc, }\AttributeTok{k=}\DecValTok{10}\NormalTok{, }\AttributeTok{bs =} \StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+} 
                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
\NormalTok{                    Forest\_type }\SpecialCharTok{+}
\NormalTok{                    Hydrological\_regime}
\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
\CommentTok{\#summary(Forest\_model\_nonlin)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
\caption{(\#tab:m\_nonlin-smooth) Statistical summary for the smooth terms for the model that considers the change in forest cover to be a non-linear variable}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Ref.df
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p-value
\end{minipage} \\
\midrule()
\endhead
\textbf{s(DeltaF\_perc)} & 1.11 & 9 & 12.72 & 0 \\
\textbf{s(log10(Area\_km2))} & 0.76 & 4 & 0.8 & 0.03 \\
\textbf{s(Dryness)} & 4.7 & 9 & 2.2 & 0 \\
\textbf{s(length)} & 4.45 & 34 & 0.24 & 0.09 \\
\bottomrule()
\end{longtable}

In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0. The total deviance explained for the model is 0.5 and the model has an AIC of 3166.9317897. These performance statistics are similar to the first (baseline) model.

One thing that is of interest is the fact that while a non-linear relationship between the change in forest cover and the change in streamflow was conceptualised, the actual relationship turns out to be linear. This can be shown from the plot and by running \texttt{gam.check}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{D }\OtherTok{\textless{}{-}} \FunctionTok{visreg}\NormalTok{(Forest\_model\_nonlin, }\StringTok{"DeltaF\_perc"}\NormalTok{, }\AttributeTok{gg =}\NormalTok{ T, }\AttributeTok{plot =}\NormalTok{ T) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{()}
\NormalTok{A }\OtherTok{\textless{}{-}} \FunctionTok{visreg}\NormalTok{(Forest\_model\_nonlin, }\StringTok{"Area\_km2"}\NormalTok{, }\AttributeTok{gg =}\NormalTok{ T, }\AttributeTok{plot =}\NormalTok{ T) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+} 
  \FunctionTok{scale\_x\_log10}\NormalTok{()}

\NormalTok{(A }\SpecialCharTok{+}\NormalTok{ D)}
\end{Highlighting}
\end{Shaded}

\includegraphics{SupplementaryMaterialPart3_files/figure-latex/unnamed-chunk-10-1.pdf}
The procedure \texttt{gam.check} highlights that the p-value of the Basis dimension check of the smooth on \texttt{DeltaF\_perc} is very small and the actual edf is close to 1 suggesting a linear relationship, similar to the smooth on log10(\texttt{Area\_km2}).

The residual analysis shows that including this non-linearity makes no difference in terms of he distribution of the residuals. While there is some remaining weak trend in the data, the overall distribution of the residuals is approximately normal.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{par}\NormalTok{(}\AttributeTok{mfrow=} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\FunctionTok{gam.check}\NormalTok{(Forest\_model\_nonlin)}
\end{Highlighting}
\end{Shaded}

\includegraphics{SupplementaryMaterialPart3_files/figure-latex/unnamed-chunk-11-1.pdf}

\begin{verbatim}
## 
## Method: GCV   Optimizer: magic
## Smoothing parameter selection converged after 15 iterations.
## The RMS GCV score gradient at convergence was 0.004151441 .
## The Hessian was positive definite.
## Model rank =  69 / 69 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                        k'    edf k-index p-value  
## s(DeltaF_perc)      9.000  1.113    0.89    0.03 *
## s(log10(Area_km2))  4.000  0.756    0.96    0.20  
## s(Dryness)          9.000  4.703    0.94    0.13  
## s(length)          34.000  4.448    0.92    0.10 .
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\renewcommand\refname{References}
\bibliography{forestandwater.bib}

\end{document}

### 2022-08-21 paper/Forest_and_Water/Forest_and_Water.Rmd
further updates 210802022

@@
 There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales. 
-However, there is also a real danger in this process, which is what we will highlight in this paper. There are three potential errors (or limitations) in the mentioned global analyses:
+However, there is also a real danger in this process, which is what we will highlight in this paper. There are four potential errors (or limitations) in the mentioned global analyses:
 
@@
 
+```{r, include=FALSE}
+options(tinytex.verbose = FALSE)
+```
+```{r, echo = F, message = F, warning=FALSE}
+require(tidyverse)
+require(lubridate)
+require(mgcv)
+require(visreg)
+require(patchwork)
+library(ggcorrplot)
+suppressWarnings(require(knitr))
+suppressWarnings(require(bookdown))
+suppressWarnings(require(pander))
+```
 ## The original data set
@@
 
-```{r table1}
+```{r table1, echo =F}
 
@@
 Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @penaarancibia2012, and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.
-Table 6 in @zhang2011).
+In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow [i.e.
+Table 6 in @zhang2011].
 
@@
 
-```{r, include=FALSE}
-options(tinytex.verbose = FALSE)
-```
-<!----
-The starting point consists of the two tables in the supplementary data from Zhang et al.
-(2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2).
-We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
---->
-```{r, echo = F, message = F, warning=FALSE}
-require(tidyverse)
-require(lubridate)
-require(mgcv)
-require(visreg)
-require(patchwork)
-library(ggcorrplot)
-suppressWarnings(require(knitr))
-suppressWarnings(require(bookdown))
-suppressWarnings(require(pander))
-
+```{r read_data, echo = F, message = F}
 Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv") %>%
@@
 
-<!--
-Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2).
-In other words, flow almost always increased.
-However, for increases in forest cover, this is not the case, and flow can both increase and decrease.
-However in both cases the variability in the reported change in flow increases with the increase in forest cover change.
-
-```{r, echo = F, warning = F, message=F, eval =F}
-# plot to show the difference
-png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  filter(is.na(Forest_Sign)==F) %>%
-  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
-  geom_point() +  
-  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
-  theme_bw() + 
-    theme(axis.title = element_text(size = rel(1.5)),
-        axis.text = element_text(size = rel(1.2)),
-        legend.text = element_text(size = rel(1.2)),
-        legend.title = element_text(size = rel(1.2)),
-        strip.text = element_text(size = rel(1.5))) + 
-  scale_colour_manual(name = "Forest cover direction", 
-                      values = c("Decrease" = "red","Increase" = "blue")) +
-  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
-dev.off()
-
-```
-
-```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of decreases (top) and increases (bottom) in forest cover", out.width="90%"}
-include_graphics("Increase_decrease.png")
-```
--->
-
 ### Geospatial location of the catchments
@@
 
-<!-- DO WE NEED TO REDO THE MAP? Eliana please provide code -->
+** DO WE NEED TO REDO THE MAP? Eliana please provide code **
 
@@
 
-The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations, even though in general cross correlation is quite low between variables.
-Some interesting relationships appear in this graph:
+The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations thate are worth investigating, even though in general cross correlation is quite low between variables.
+Some interesting relationships that appear in this graph are:
 
@@
 ```
-```{r modelforestpos, eval =F, echo =F}
+```{r modelforestpos, eval =T, echo =F}
 Forest_model_all2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
@@
                     , data = All_data2)
-summary(Forest_model_all2)
-
-```
-```{r modelforestnonlin, eval =F, echo =F}
+#summary(Forest_model_all2)
+
+```
+```{r modelforestnonlin, eval =T, echo =F}
 Forest_model_nonlin <- gam(DeltaQf_perc ~ s(DeltaF_perc, k=10, bs = "ts") + 
@@
                     , data = All_data2)
-summary(Forest_model_nonlin)
-plot(Forest_model_nonlin)
+#summary(Forest_model_nonlin)
+#plot(Forest_model_nonlin)
 ```
@@
 ```{r, echo = F}
-pander(round(summary(Forest_model_all)$p.table,2), caption = "(\\#tab:m_all-linear) Statistical summary for the linear terms the full model")
+pander(round(summary(Forest_model_all)$p.table,2), caption = "(\\#tab:m-all-linear) Statistical summary for the linear terms the full model")
 #plot(model6_all)
@@
 ```{r , echo = F}
-pander(round(summary(Forest_model_all)$s.table,2), caption = "(\\#tab:m_all-smooth) Statistical summary for the smooth terms for the full model")
+pander(round(summary(Forest_model_all)$s.table,2), caption = "(\\#tab:m-all-smooth) Statistical summary for the smooth terms for the full model")
 #plot(model6_all)
@@
 
-Inspecting the significance of the variables (Table \@ref(tab:m_all-linear) and Table \@ref(tab:m_all-smooth)) indicates some interesting features.
+Inspecting the significance of the variables (Table \@ref(tab:m-all-linear) and Table \@ref(tab:m-all-smooth)) indicates some interesting features.
 
@@
 
-Furthermore Table \@ref(tab:m_six_all-linear) indicates that several of the assessment methods are significant, which was also indicated in the correlation analysis.
+Furthermore Table \@ref(tab:m-all-linear) indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
 In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables ($p < 0.05$).
+
+The remaining variables related to rainfall observation technique, forest type or hydrological regime don't appear to have an influence on the change in flow.
+
+```{r, echo = F}
+table_comparison <- tibble("Model for change in forest cover" = 
+                             c("linear across range",
+                               "different for forestation and deforestation",
+                               "non-linear across the range"),
+                           "Deviation explained" = 
+                             c(round(summary(Forest_model_all)$dev.expl,2),
+                               round(summary(Forest_model_all2)$dev.expl,2),
+                               round(summary(Forest_model_nonlin)$dev.expl,2)),
+                           "AIC" = c(AIC(Forest_model_all), AIC(Forest_model_all2),
+                                     AIC(Forest_model_nonlin)))
+pander(table_comparison, caption = "(\\#tab:tablecomparison) Comparison of alternative models for the relationship between the change in forest cover and the change in streamflow.
+(See Supplementary Material part 3)")
+```
+
+As discussed in the methods, the overall linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives and a check whether the relationship might be non-linear.
+This approach tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods.
+The detail of the comparison is highlighted in Supplementary material part 3.
+However, generally the results of the analysis showed two main points (Table \@ref(tab:tablecomparison):
+
+1.
+The model assuming a simple linear relationship between change in forest cover (both positive and negative) and the change in flow explained the most variation in the data and indicated the best performance in terms of the Aikaike Information Criterium (AIC); and
+2. There is no need to assume a non-linear relationship, as a linear relationship provides a better fit to the data.
+
+The smoothed variables in the model can be inspected visually to identify if there are any issues with the fit.
+This is in addition to using `gam.check()` in the R package `mgcv` to test whether the number of knots is adequate.
 
@@
 
-As discussed in the methods, the single linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives.
-This tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods.
-The results of this showed
-
 <!-- plot the residuals -->
@@
 
-As is not always shown, here we show the residual distribution of the model (Figure \@ref(fig:gamcheckmodelall).
-These are approximately normal, although there is a a skew in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall)).
-This is related to a limited number catchments that have very high changes in streamflow in the data set.In other words, the distribution of the residuals is somewhat fat-tailed.
+As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \@ref(fig:gamcheckmodelall).
+Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall)).
+This is related to a limited number catchments that have very high changes in streamflow in the data set.
+In other words, the distribution of the residuals is somewhat fat-tailed.
 
 One solution could be to transform the data, however this is not that simple.
-As the data for the change in flow cover the domain $\mathbb{R}$ we can not use a simple log or Gamma transformation and more complex transformations make the results of the regression difficult to interpret.
+As the data for the change in flow cover the domain $\mathbb{R}$, a simple log or Gamma transformation is not a solution.
+More complex transformations make the results of the regression difficult to interpret, and at some point can be slightly contrived.
 
@@
 
-### Test Removal of studies of great lenght and very dry catchments
+### Test removal of studies of great length and for very dry catchments
 
@@
 The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \@ref(fig:smoothsmodelall)) does not reflect a similar response to @filoso2017 and @jackson2005.
+The shape of the conditional response (Figure \@ref(fig:smoothsmodelall)) does not reflect a similar response as indicated by @filoso2017 and @jackson2005.
 One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \@ref(fig:smoothsmodelall)).
@@
 These catchments are listed in Table \@ref(tab:drytable), and are three catchments in Arizona and 1 catchment in South Africa.
+It is possible that catchments in these climate zones behave different from the rest of the catchments.
 
@@
                     Hydrological_regime, data = NoLong_Data)
-summary(model_redLength)
-gam.check(model_redLength)
+#summary(model_redLength)
+#gam.check(model_redLength)
 # plot(model7)
@@
 ```{r, echo = F}
-pander(round(summary(model_redLength)$p.table,2), caption = "(\\#tab:m_red-linear) Statistical summary for the linear terms the restricted model")
+pander(round(summary(model_redLength)$p.table,2), caption = "(\\#tab:m-red-linear) Statistical summary for the linear terms the restricted model")
 #plot(model6_all)
@@
 
-Therefore it is worth investigating what removing these few data points has on the overall model and the significance of the variables.
-A model testing data that have *Dryness* <= 5 and *length* <= 60 years is therefore based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments.
-
-This last model explains only slightly less of the variation with an adjusted *r^2^* of `r round(summary(model_redLength)$r.sq,2)` and a deviance explained of `r round(summary(model_redLength)$dev.expl,2)`.
+Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
+Data that have *Dryness* <= 5 and *length* <= 60 years were removed from the dataset and the model based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments is run again.
+
+This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted *r^2^* of `r round(summary(model_redLength)$r.sq,2)` and a deviance explained of `r round(summary(model_redLength)$dev.expl,2)`.
 
 Investigating the non-linear responses suggest that *Dryness* has a non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
-Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship visually looks almost linear.
-The variable *length* no longer is significant, after removal of the two studies with very long lengths.
-
-## So far the same as zhang et al. (2017), but is there a problem?
-
-At this point it seems like we are confirming the @zhang2017 and other results. But is there a problem hidden in the analysis?
+Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship looks almost linear.
+More importantly, the variable *length* is no longer significant, after removal of the two studies with very long lengths.
+
+**Possibly insert here a model to investigate total forest area as a random effect**
+
+```{r, eval =T, echo =F}
+Table_assessment <- All_data2 %>%
+  group_by(Assessment_technique) %>%
+  summarise(n = n()) %>%
+  arrange(desc(n))
+```
+```{r, echo=F}
+pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
+```
+
+One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess).
+<!-- and Figure \@ref(fig:assessment)).
+-->
+
+Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA).
+
+```{r model_reduc_assess, echo = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
+
+model_assess <- gam(DeltaQf_perc ~ DeltaF_perc  + 
+                    s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), k = 10, bs = "ts") + s(length, bs = "ts") + #s(From, bs = "ts") +
+                    Precip_data_type +  Assessment_technique + Forest_type +
+                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4))
+#plot(model8_assess)
+#summary(model_assess)
+#gam.check(model_assess)
+```
+
+```{r, echo = F}
+pander(round(summary(model_assess)$p.table,2), caption= "(\\#tab:modelassess-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
+pander(round(summary(model_assess)$s.table,2), caption = "(\\#tab:modelassess-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
+
+```
+
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:modelassess-linear) and \@ref(tab:modelassess-smooth)).
+It strengthens the significance of the different assessment techniques,  but generally results in the same interpretation.
+Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
+
+However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
+Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
+In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
+<!-- However, as indicated earlier, the EA studies in the database are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.-->
+
+Discussion
+======================
+
+The results presented so far, while using generalised additive modelling rather than single variable regression, end up with roughly the same conclusions as earlier papers [@zhang2017; @filoso2017].
+It appears that:
+
+1. Larger catchments show lower impact of forest cover change on streamflow;  
+2. Drier catchments show a greater impact of forest cover change on streamflow; and  
+3. There is a general linear relationship between the change in forest cover and the change in streamflow.
+
+```{r, echo = F, eval =F}
+png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
+All_data2 %>%
+  filter(is.na(Forest_Sign) == F) %>%
+  ggplot(aes(DeltaF_perc, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
+             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
+  theme_bw() + ylab("% change in flow") +
+  theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5))) +
+  xlab("% change in forest cover") + #scale_y_log10() +
+  scale_size_continuous(name = "log10(Area in km2)") +
+  scale_colour_discrete(name = "Assessment Technique") +
+  scale_shape_discrete(name = "Forest cover direction")
+dev.off()
+```
+
+```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
+
+include_graphics("flow_forest_byArea.png")
+
+```
+
+Figure \@ref(fig:overview) provides a further overview of the whole data set and the size of the catchments and the different assessment methods are highlighted.
+This figure clearly indicates that the data relating to high changes in forest cover are all small catchments and relate mostly to paired watershed experiments.
+In contrast, data related to large catchments are related to smaller changes in forest cover and different methods, such as hydrological modelling and elasticity analysis.
+This confirms the model results (Table \@ref(tab:modelassess-linear)) and the earlier correlation analysis (Figure \@ref(fig:corgraphs)).
+
+It is possible that one of the reasons why @zhang2017 separated their analysis in large (> 1000 km^2^) and small (< 1000 km^2^) catchments, is that they realised this difference in assessment methods and wanted to account for this.
+However, this is not explicitly identified, and there is no real physical explanation of the 1000 km^2^ threshold.
+
+The other interesting point in Figure \@ref(fig:overview) is that the variation in the data increases as the catchment size decreases and the change in forest cover increases.
+This also means that the overall variation in the data for paired watershed experiments (PWE) is much greater than for any of the other methods.
+
+## Is there a problem with extending local experimental data to larger scales?
+
+The overarching reason for combining past studies at a global scale is to infer relationships that can be used to make more general statements or develop more global scale modelling of impacts [i.e.
+@zhou2015; @jackson2005; @hoekvandijke2012].
+Therefore, the results fromthe analysis could be seen as a confirmation of the earlier research [@zhang2017; @filoso2017; @zhou2015; @jackson2005].
+However, as is highlighted in the introduction there are four major issues with this type of analysis, and the results from this paper can also highlight this.
+
+### Latent variables that are not included in the typical single covariate analysis;
+
+### Interpretation errors due to incomplete descriptions of the experiments in the original papers;
+
+### Aggregation of data that originates from different experiments with different objectives across a wide time period
+
+### Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
+
+```{r, eval = F, echo =F}
+png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
+All_data2 %>%
+  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  geom_boxplot(colour = "blue") + theme_bw() +
+    theme(axis.title = element_text(size = rel(2)),
+        axis.text = element_text(size = rel(1.5)))
+dev.off()
+```
+
+```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =F}
+
+include_graphics("AssessmentTechnique_byArea.png")
+```
 
@@
 
-```{r, eval = F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
-  geom_boxplot(colour = "blue") + theme_bw() +
-    theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5)))
-dev.off()
-```
-
-```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =T}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
-
-One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess) and Figure \@ref(fig:assessment)).
-
-```{r, echo = F, eval =F}
-png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
-All_data2 %>%
-  filter(is.na(Forest_Sign) == F) %>%
-  ggplot(aes(DeltaF_perc, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
-             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
-  theme_bw() + ylab("% change in flow") +
-  theme(axis.title = element_text(size = rel(2)),
-        axis.text = element_text(size = rel(1.5))) +
-  xlab("% change in forest cover") + #scale_y_log10() +
-  scale_size_continuous(name = "log10(Area in km2)") +
-  scale_colour_discrete(name = "Assessment Technique") +
-  scale_shape_discrete(name = "Forest cover direction")
-dev.off()
-```
-
-```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}
-
-include_graphics("flow_forest_byArea.png")
-
-```
-
-Discussion
-======================
-
 ## Catchment size
@@
 This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \@ref(fig:increasedecrease)).
-
-## The effect of assessment techniques with very small numbers of observations
-
-```{r, eval =T, echo =F}
-Table_assessment <- All_data2 %>%
-  group_by(Assessment_technique) %>%
-  summarise(n = n()) %>%
-  arrange(desc(n))
-```
-```{r, echo=F}
-pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
-```
-
-```{r model8_reduc_assess, echo = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model8_assess <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), k = 10, bs = "ts") + s(length, bs = "ts") + #s(From, bs = "ts") +
-                    Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4))
-#plot(model8_assess)
-# summary(model8_assess)
-# gam.check(model8_assess)
-```
-```{r, echo = F}
-pander(round(summary(model8_assess)$p.table,2), caption= "(\\#tab:model8-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
-pander(round(summary(model8_assess)$s.table,2), caption = "(\\#tab:model8-smooth) Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques")
-
-```
-
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:model8-linear) and \@ref(tab:model8-smooth)).
-It strengthens the significance of the different assessment techniques and *Dryness* but generally results in the same interpretation.
-Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
-
-However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
-Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
-In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in @zhou2015) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies in the database are all on the drier end of the *Dryness* spectrum, highlighting another unresolved interaction in the data.
 

### 2022-08-21 paper/Forest_and_Water/Forest_and_Water.tex
further updates 210802022

@@
 
-\usepackage{graphicx}
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.},
+            pdftitle={Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.},
             colorlinks=false,
@@
 
-  \title{Generalising the impact of forest cover on streamflow from experimental data, it is not that simple.}
+  \title{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}
     \author[a,b]{R. Willem Vervoort}
@@
 The columns in this data set (are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in \citet{zhang2017}), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in @ref(Table 1). These abbreviations will later be used in the models.
-
-Table (\#table:tab1) Summary of abbreviations of factors used in the \citet{zhang2017} data set
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule()
-Factor & Abbreviation & Definition \\
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table @\ref(tab:table1). These abbreviations will later be used in the models.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}@{}}
+\caption{Summary of abbreviations of factors used in the Zhang et al.~(2017) data set}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Factor
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Abbreviation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Definition
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Factor
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Abbreviation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Definition
+\end{minipage} \\
 \midrule()
@@
 assessment technique & PWE & paired watershed experiment \\
-& QPW & quasi-paired watershed experiment \\
+& QPW & quasi-paired watershed
+experiment \\
 & HM & hydrological modelling \\
-& EA & elastictity analysis \\
-& SH & combined use of statistical methods \\
-& & and hydrographs \\
+& EA & elasticity analysis \\
+& SH & statistical modelling and
+hydrographs \\
 \bottomrule()
@@
 
-\citet{zhang2017} use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET\textsubscript{0} or E\textsubscript{0}) in their analysis, and use the dryness index to identify the climate type.
+The paper by \citet{zhang2017} use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET\textsubscript{0} or E\textsubscript{0}) in their analysis, and use the dryness index to identify the climate type.
 However, the potential or reference ET was not originally included as part of the published data set.
@@
 Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{penaarancibia2012}, and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in \citet{zhang2011}).
+In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow \citep[i.e.~Table 6 in][]{zhang2011}.
 
@@
 \begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
+\Delta Qf \% \sim &~ \Delta \% forest~cover + \notag \\ 
 & \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq2}
@@
 
-The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations, even though in general cross correlation is quite low between variables.
-Some interesting relationships appear in this graph:
+The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations thate are worth investigating, even though in general cross correlation is quite low between variables.
+Some interesting relationships that appear in this graph are:
 
@@
 
-This includes introducing non-linearity (Equation \eqref{eq:eq7}) for the numerical variables in the model.
+This includes introducing non-linearity (Equation \eqref{eq:eq2}) for the numerical variables in the model.
 While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
 The number of knots in the non-linear variables was based on assessment of the edf and k' \textbf{check wood 2006 and explain well}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc + s(log10(Area_km2), k = 5, bs = "ts") + 
-##     s(Dryness, k = 10, bs = "ts") + s(length, k = 35, bs = "ts") + 
-##     Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime
-## 
-## Parametric coefficients:
-##                              Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                  -7.47200   15.87043  -0.471  0.63814    
-## DeltaF_perc                  -0.59422    0.05459 -10.885  < 2e-16 ***
-## Precip_data_typeOB          -17.87257   12.90741  -1.385  0.16724    
-## Precip_data_typeSG            0.19950   14.82105   0.013  0.98927    
-## Assessment_techniqueEA, HM   18.66830   41.70211   0.448  0.65474    
-## Assessment_techniqueHM       26.60929   11.43273   2.327  0.02065 *  
-## Assessment_techniquePWE      30.77146   11.68272   2.634  0.00890 ** 
-## Assessment_techniquePWE, HM  15.79891   42.22036   0.374  0.70853    
-## Assessment_techniqueQPW      41.34637   19.66208   2.103  0.03636 *  
-## Assessment_techniqueQPW, EA  26.04728   23.83570   1.093  0.27542    
-## Assessment_techniqueSH       39.31195   11.53645   3.408  0.00075 ***
-## Forest_typeCF                -9.28294    7.41227  -1.252  0.21147    
-## Forest_typeMF                -6.34249    7.37916  -0.860  0.39078    
-## Hydrological_regimeSD         0.13379    8.93522   0.015  0.98806    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                       edf Ref.df     F  p-value    
-## s(log10(Area_km2)) 0.7738      4 0.846 0.032280 *  
-## s(Dryness)         4.7084      9 2.183 0.000615 ***
-## s(length)          4.4431     34 0.242 0.089106 .  
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.455   Deviance explained = 49.6%
-## GCV = 1766.8  Scale est. = 1629.1    n = 307
-\end{verbatim}
 
@@
 
-Furthermore Table @ref(tab:m\_six\_all-linear) indicates that several of the assessment methods are significant, which was also indicated in the correlation analysis.
+Furthermore Table @ref(tab:m\_all-linear) indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
 In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables (\(p < 0.05\)).
+
+The remaining variables related to rainfall observation technique, forest type or hydrological regime don't appear to have an influence on the change in flow.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4444}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3056}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0972}}@{}}
+\caption{\label{tab:tablecomparison} Comparison of alternative models for the relationship between the change in forest cover and the change in streamflow.
+(See Supplementary Material part 3)}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Model for change in forest
+cover
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Deviation explained
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+AIC
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+Model for change in forest
+cover
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Deviation explained
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+AIC
+\end{minipage} \\
+\midrule()
+\endhead
+linear across range & 0.5 & 3167 \\
+different for forestation and
+deforestation & 0.46 & 3213 \\
+non-linear across the range & 0.5 & 3167 \\
+\bottomrule()
+\end{longtable}
+
+As discussed in the methods, the overall linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives and a check whether the relationship might be non-linear.
+This approach tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods.
+The detail of the comparison is highlighted in Supplementary material part 3.
+However, generally the results of the analysis showed two main points (Table \ref{tab:tablecomparison}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The model assuming a simple linear relationship between change in forest cover (both positive and negative) and the change in flow explained the most variation in the data and indicated the best performance in terms of the Aikaike Information Criterium (AIC); and\\
+\item
+  There is no need to assume a non-linear relationship, as a linear relationship provides a better fit to the data.
+\end{enumerate}
+
+The smoothed variables in the model can be inspected visually to identify if there are any issues with the fit.
+This is in addition to using \texttt{gam.check()} in the R package \texttt{mgcv} to test whether the number of knots is adequate.
 
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model_all} \caption{Residual plots for the regression model indicating a slightly fat-tailed residual distribution}\label{fig:gamcheckmodelkall}
+\includegraphics[width=0.9\linewidth]{residual_plot_model_all} \caption{Residual plots for the regression model indicating a slightly fat-tailed residual distribution}\label{fig:gamcheckmodelall}
 \end{figure}
 
-As is not always shown, here we show the residual distribution of the model (Figure \ref{fig:gamcheckmodelall}.
-These are approximately normal, although there is a a skew in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}).
-This is related to a limited number catchments that have very high changes in streamflow in the data set.In other words, the distribution of the residuals is somewhat fat-tailed.
+As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}.
+Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}).
+This is related to a limited number catchments that have very high changes in streamflow in the data set.
+In other words, the distribution of the residuals is somewhat fat-tailed.
 
 One solution could be to transform the data, however this is not that simple.
-As the data for the change in flow cover the domain \(\mathbb{R}\) we can not use a simple log or Gamma transformation and more complex transformations make the results of the regression difficult to interpret.
+As the data for the change in flow cover the domain \(\mathbb{R}\), a simple log or Gamma transformation is not a solution.
+More complex transformations make the results of the regression difficult to interpret, and at some point can be slightly contrived.
 
@@
 
-\hypertarget{test-removal-of-studies-of-great-lenght-and-very-dry-catchments}{%
-\subsubsection{Test Removal of studies of great lenght and very dry catchments}\label{test-removal-of-studies-of-great-lenght-and-very-dry-catchments}}
+\hypertarget{test-removal-of-studies-of-great-length-and-for-very-dry-catchments}{%
+\subsubsection{Test removal of studies of great length and for very dry catchments}\label{test-removal-of-studies-of-great-length-and-for-very-dry-catchments}}
 
@@
 The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
-The shape of the conditional response (Figure \ref{fig:smoothsmodelall}) does not reflect a similar response to \citet{filoso2017} and \citet{jackson2005}.
+The shape of the conditional response (Figure \ref{fig:smoothsmodelall}) does not reflect a similar response as indicated by \citet{filoso2017} and \citet{jackson2005}.
 One reason could be that the relationship is dominated by the few data points with very long data series, which show highly variable responses (Figure \ref{fig:smoothsmodelall}).
@@
 These catchments are listed in Table \ref{tab:drytable}, and are three catchments in Arizona and 1 catchment in South Africa.
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## DeltaQf_perc ~ DeltaF_perc + Forest_Sign + s(Dryness, k = 10, 
-##     bs = "ts") + s(log10(Area_km2), k = 5, bs = "ts") + s(length, 
-##     k = 10, bs = "ts") + Precip_data_type + Assessment_technique + 
-##     Forest_type + Hydrological_regime
-## 
-## Parametric coefficients:
-##                              Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)                 -10.27984   17.60496  -0.584 0.559747    
-## DeltaF_perc                  -0.58628    0.07722  -7.592 4.73e-13 ***
-## Forest_SignIncrease           0.92751    9.57534   0.097 0.922903    
-## Precip_data_typeOB          -12.54417   12.19935  -1.028 0.304714    
-## Precip_data_typeSG            5.90459   15.05545   0.392 0.695217    
-## Assessment_techniqueEA, HM   18.85724   39.89593   0.473 0.636824    
-## Assessment_techniqueHM       29.54406   11.08107   2.666 0.008119 ** 
-## Assessment_techniquePWE      24.55910   12.28800   1.999 0.046618 *  
-## Assessment_techniquePWE, HM  13.22185   40.95225   0.323 0.747043    
-## Assessment_techniqueQPW      44.21273   18.99855   2.327 0.020671 *  
-## Assessment_techniqueQPW, EA  25.53696   22.81154   1.119 0.263899    
-## Assessment_techniqueSH       40.88820   11.15628   3.665 0.000296 ***
-## Forest_typeCF               -10.21893    7.12542  -1.434 0.152648    
-## Forest_typeMF                -3.90403    7.14361  -0.547 0.585153    
-## Hydrological_regimeSD        -0.02434    8.65352  -0.003 0.997757    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                          edf Ref.df     F  p-value    
-## s(Dryness)         4.070e+00      9 2.127 0.000407 ***
-## s(log10(Area_km2)) 1.574e+00      4 1.865 0.009339 ** 
-## s(length)          6.223e-07      9 0.000 0.857556    
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.455   Deviance explained = 49.1%
-## GCV = 1612.2  Scale est. = 1501.2    n = 300
-\end{verbatim}
-
-\includegraphics{Forest_and_Water_files/figure-latex/model_all_red_Length-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 19 iterations.
-## The RMS GCV score gradient at convergence was 7.893755e-05 .
-## The Hessian was positive definite.
-## Model rank =  37 / 37 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                          k'      edf k-index p-value  
-## s(Dryness)         9.00e+00 4.07e+00    0.93   0.130  
-## s(log10(Area_km2)) 4.00e+00 1.57e+00    0.98   0.330  
-## s(length)          9.00e+00 6.22e-07    0.89   0.045 *
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{verbatim}
+It is possible that catchments in these climate zones behave different from the rest of the catchments.
 
@@
 
-Therefore it is worth investigating what removing these few data points has on the overall model and the significance of the variables.
-A model testing data that have \emph{Dryness} \textless= 5 and \emph{length} \textless= 60 years is therefore based on a reduction of the data set from 329 to 310 catchments.
-
-This last model explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.46 and a deviance explained of 0.49.
+Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
+Data that have \emph{Dryness} \textless= 5 and \emph{length} \textless= 60 years were removed from the dataset and the model based on a reduction of the data set from 329 to 310 catchments is run again.
+
+This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.46 and a deviance explained of 0.49.
 
 Investigating the non-linear responses suggest that \emph{Dryness} has a non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
-Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship visually looks almost linear.
-The variable \emph{length} no longer is significant, after removal of the two studies with very long lengths.
-
-\hypertarget{so-far-the-same-as-zhang-et-al.-2017-but-is-there-a-problem}{%
-\subsection{So far the same as zhang et al.~(2017), but is there a problem?}\label{so-far-the-same-as-zhang-et-al.-2017-but-is-there-a-problem}}
-
-At this point it seems like we are confirming the \citet{zhang2017} and other results. But is there a problem hidden in the analysis?
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byArea} \caption{Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments}\label{fig:assessment}
-\end{figure}
-
-One concern with the result presented here, is that there are a few assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess} and Figure \ref{fig:assessment}).
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{flow_forest_byArea} \caption{Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data}\label{fig:overview}
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{catchment-size}{%
-\subsection{Catchment size}\label{catchment-size}}
-
-Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to \citet{zhang2017}, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
-In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
-
-An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change \citep{navas2019}.
-Therefore, specifically if the forest cover change is small relative to the catchment size, the effect of this change will be buffered.
-
-There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.35 with a slope of -9.67) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
-This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
-
-This is also reflected in the second caveat.
-Most of the data from the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
-For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
-This is a confounding factor that is not easily addressed in the regression modelling attempted here.
-Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
-
-In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
-
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamflow.
-
-\hypertarget{model-residuals}{%
-\subsection{Model residuals}\label{model-residuals}}
-
-As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
-These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \ref{fig:overview}), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheckredLength}).
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheckredLength}
-\end{figure}
-
-The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
-Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
-In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
-This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \ref{fig:increasedecrease}).
-
-\hypertarget{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}{%
-\subsection{The effect of assessment techniques with very small numbers of observations}\label{the-effect-of-assessment-techniques-with-very-small-numbers-of-observations}}
+Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship looks almost linear.
+More importantly, the variable \emph{length} is no longer significant, after removal of the two studies with very long lengths.
+
+\textbf{Possibly insert here a model to investigate total forest area as a random effect}
 
@@
 
+One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess}.
+
+Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA).
+
 \begin{longtable}[]{@{}
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}@{}}
-\caption{\label{tab:model8-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
+\caption{\label{tab:modelassess-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{(Intercept)} & 16.98 & 18.34 & 0.93 & 0.36 \\
-\textbf{DeltaF\_perc\_pos} & 0.23 & 0.1 & 2.39 & 0.02 \\
-\textbf{Forest\_SignIncrease} & -52.33 & 7.12 & -7.35 & 0 \\
-\textbf{Precip\_data\_typeOB} & -16.66 & 13.15 & -1.27 & 0.21 \\
-\textbf{Precip\_data\_typeSG} & -5.38 & 15.48 & -0.35 & 0.73 \\
-\textbf{Assessment\_techniqueHM} & 24.86 & 12.05 & 2.06 & 0.04 \\
-\textbf{Assessment\_techniquePWE} & 23.02 & 12.84 & 1.79 & 0.07 \\
-\textbf{Assessment\_techniqueQPW} & 24.65 & 20.36 & 1.21 & 0.23 \\
-\textbf{Assessment\_techniqueSH} & 33.31 & 12.13 & 2.75 & 0.01 \\
-\textbf{Forest\_typeCF} & -11.34 & 7.65 & -1.48 & 0.14 \\
-\textbf{Forest\_typeMF} & -0.46 & 7.85 & -0.06 & 0.95 \\
-\textbf{Hydrological\_regimeSD} & 1.41 & 9.36 & 0.15 & 0.88 \\
+\textbf{(Intercept)} & -15.02 & 14.46 & -1.04 & 0.3 \\
+\textbf{DeltaF\_perc} & -0.58 & 0.05 & -10.99 & 0 \\
+\textbf{Precip\_data\_typeOB} & -12.13 & 12.03 & -1.01 & 0.31 \\
+\textbf{Precip\_data\_typeSG} & 7.25 & 14.02 & 0.52 & 0.61 \\
+\textbf{Assessment\_techniqueHM} & 31.88 & 10.86 & 2.94 & 0 \\
+\textbf{Assessment\_techniquePWE} & 30.77 & 11.07 & 2.78 & 0.01 \\
+\textbf{Assessment\_techniqueQPW} & 45.78 & 18.37 & 2.49 & 0.01 \\
+\textbf{Assessment\_techniqueSH} & 42.88 & 11.01 & 3.9 & 0 \\
+\textbf{Forest\_typeCF} & -8.78 & 7.06 & -1.24 & 0.21 \\
+\textbf{Forest\_typeMF} & -2.17 & 7.22 & -0.3 & 0.76 \\
+\textbf{Hydrological\_regimeSD} & -1.29 & 8.62 & -0.15 & 0.88 \\
 \bottomrule()
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{\label{tab:model8-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
+\caption{\label{tab:modelassess-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
 \toprule()
@@
 \endhead
-\textbf{s(Dryness)} & 3.46 & 9 & 2.71 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.72 & 9 & 0.32 & 0.04 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.74 \\
+\textbf{s(Dryness)} & 3.44 & 9 & 2.71 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.86 & 9 & 0.61 & 0.01 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.87 \\
 \bottomrule()
@@
 
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \ref{tab:model8-linear} and \ref{tab:model8-smooth}).
-It strengthens the significance of the different assessment techniques and \emph{Dryness} but generally results in the same interpretation.
+Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \ref{tab:modelassess-linear} and \ref{tab:modelassess-smooth}).
+It strengthens the significance of the different assessment techniques, but generally results in the same interpretation.
 Overall this suggests that although those observations have some impact on the overall relationships, they do not strongly bias the outcomes.
@@
 In other words, there is a distinct difference in the way the change in flow is assessed, and the EA method (for example in \citet{zhou2015}) appears to suggest a much smaller effect on the change in flow.
-However, as indicated earlier, the EA studies in the database are all on the drier end of the \emph{Dryness} spectrum, highlighting another unresolved interaction in the data.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The results presented so far, while using generalised additive modelling rather than single variable regression, end up with roughly the same conclusions as earlier papers \citep{zhang2017, filoso2017}.
+It appears that:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Larger catchments show lower impact of forest cover change on streamflow;\\
+\item
+  Drier catchments show a greater impact of forest cover change on streamflow; and\\
+\item
+  There is a general linear relationship between the change in forest cover and the change in streamflow.
+\end{enumerate}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{flow_forest_byArea} \caption{Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data}\label{fig:overview}
+\end{figure}
+
+Figure \ref{fig:overview} provides a further overview of the whole data set and the size of the catchments and the different assessment methods are highlighted.
+This figure clearly indicates that the data relating to high changes in forest cover are all small catchments and relate mostly to paired watershed experiments.
+In contrast, data related to large catchments are related to smaller changes in forest cover and different methods, such as hydrological modelling and elasticity analysis.
+This confirms the model results (Table @ref(tab:model\_assess-linear)) and the earlier correlation analysis (Figure \ref{fig:corgraphs}).
+
+It is possible that one of the reasons why \citet{zhang2017} separated their analysis in large (\textgreater{} 1000 km\textsuperscript{2}) and small (\textless{} 1000 km\textsuperscript{2}) catchments, is that they realised this difference in assessment methods.
+However, this is not explicitly identified.
+
+The other interesting point in Figure \ref{fig:overview} is that the variation in the data increases as the catchment size decreases and the change in forest cover increases.
+This also means that the overall variation in the data for paired watershed experiments (PWE) is much greater than for any of the other methods.
+
+\hypertarget{is-there-a-problem-with-extending-local-experimental-data-to-larger-scales}{%
+\subsection{Is there a problem with extending local experimental data to larger scales?}\label{is-there-a-problem-with-extending-local-experimental-data-to-larger-scales}}
+
+\hypertarget{catchment-size}{%
+\subsection{Catchment size}\label{catchment-size}}
+
+Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to \citet{zhang2017}, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
+If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
+In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
+
+An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change \citep{navas2019}.
+Therefore, specifically if the forest cover change is small relative to the catchment size, the effect of this change will be buffered.
+
+There are two caveats on this explanation.
+The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.35 with a slope of -9.67) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
+This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
+
+This is also reflected in the second caveat.
+Most of the data from the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
+For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
+This is a confounding factor that is not easily addressed in the regression modelling attempted here.
+Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
+
+In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
+
+Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
+While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
+Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
+As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamflow.
+
+\hypertarget{model-residuals}{%
+\subsection{Model residuals}\label{model-residuals}}
+
+As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
+These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \ref{fig:overview}), which show very high variability.
+The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheckredLength}).
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheckredLength}
+\end{figure}
+
+The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
+Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
+In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
+This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \ref{fig:increasedecrease}).
 

### 2022-09-01 paper/Forest_and_Water/Forest_and_Water.Rmd
further changes 01092022

@@
 @zhou2015; @jackson2005; @hoekvandijke2012].
-Therefore, the results fromthe analysis could be seen as a confirmation of the earlier research [@zhang2017; @filoso2017; @zhou2015; @jackson2005].
+Therefore, the results from the analysis could be seen as a confirmation of the earlier research [@zhang2017; @filoso2017; @zhou2015; @jackson2005].
 However, as is highlighted in the introduction there are four major issues with this type of analysis, and the results from this paper can also highlight this.
 
-### Latent variables that are not included in the typical single covariate analysis;
-
-### Interpretation errors due to incomplete descriptions of the experiments in the original papers;
-
-### Aggregation of data that originates from different experiments with different objectives across a wide time period
+### Issue 1: Latent variables are not included in the typical single covariate analysis
+
+The results show that it is simply impossible to analyse a single covariate relationship, as there are several latent variables in the data.
+An example of this is the general relationship of the change in flow as a function of the change in forest cover.
+Clearly the relationship is highly impacted by the fact that all the small catchments have large changes in forest cover and are all associated with paired watershed experiments.
+Without taking these factors into account, a definite answer about the impact of forest cover on the change in flow cannot be given.
+Furthermore, the large variability in the change in flow data for these small catchments (Figure \@ref(fig:overview)) indicates that there is a further (unknown) variable that explains the variation in the data.
+
+If the remaining variation in the residuals is small relative to the trend, then there is little need to identify further latent variables, but if the variation is large, then it is unclear if it is the latent variable that determines the trend, or the actual relationship in the data.
+
+Similarly, the data for the larger catchments containing smaller changes in forest cover are dominated by hydrological modelling studies, resulting in a further complication.
+If the reponse of the streamflow in the modelling studies is the result of the conceptualised relationship between streamflow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in streamflow is real, or simply a result of a pre-conceived model relationship.
+Is the smaller variation in the data for smaller changed in forest cover (Figure \@ref(fig:overview)) a result of similar conceptualised model relationships, or actual variation between catchments and climate types?
+Currently this question cannot be answered.
+
+This becomes problematic when extrapolated to larger scales.
+A clear example of this is the paper by @hoekvandijke2022 where the conceptualised relationship between forest cover and streamflow pre-determines the outcomes of the global modelling.
+
+The only  way to analyse changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in @levy2018).
+
+We are not arguing that there is no relationship between streamflow and forest cover, and there might indeed be a global relationship that can be discovered.
+But, this relationship can only be discovered if we are able to address some of the major other factors that explain the variability, and work with actual data and not model outputs.
+
+### Issue 2: Interpretation errors due to complex descriptions of the experiments in the original papers
+
+The second major issue that became clear from reviewing many of the original papers is that some of the variability might be an interpretation problem.
+In many cases the original description in the paper is interpreted to extract the \% change in streamflow from the \% change in forest cover.
+This seems like a simple activity, but this is not always the case.
+
+Two examples can be highlighted:
+
+- The papers from @almeida2016 and @ferreto2021 partly discuss the same experiment and the same catchment.
+In @almeida2016, the methods discuss how two experimental catchments of approximately 80ha in size which were harvested.
+One catchment was 100\% harvested and the other 30\% harvested.
+Throughout the paper the catchments are indicated as 100\% harvested and 30\% harvested.
+However, only after reading @ferreto2021, did we discover that in fact the 100\% and 30\% refer to the "eucalyptus plantation area", which was about 60\% of the total area.
+This is in fact mentioned in Table 1 in @almeida2016, but does not appear in the text.
+The question then becomes how to interpret this in the data base for this paper.
+Clearly it was a 100\% and 30\% change in forest cover, but only for the 60\% plantation cover, not for any of the other areas in the catchment, which included native vegetation and riparian vegetation.
+There are several other examples like this in the different papers (for example @blackie1979kimakia and @blackie1979kericho).
+- Another example is the paper by @waterloo2007.
+This modelling study in Fiji of the clearing of a catchment reports the changes in streamflow over parts of the year.
+For a period of 324 days the streamflow increased from 252 mm to 580 mm (a 230\% increase if calculated as $580/252*100$) and for a second period of 309 days the streamflow increased from 90 mm to 194 mm (a 215 \% increase).
+However, how we convert this to a change in annual flow (which most of the other data relate to) is difficult.
+The original data base listed a 50 \% change in flow, but it is difficult to identify how this is calculated.
+We suspect that is results from $252/580*100 ≈ 50$ and $90/194 ≈ 50$.
+
+Clearly, interpreting older papers can be difficult and this can result in variation and errors in the data that is being analysed.
+Similar to the last issue, if these errors only introduce small variation in the data, then it will not limit the interpolation to larger scales.
+At this point, it is not clear if this is indeed the case.
+The large variation in the experimental watershed data suggests that this might be a more serious problem.
+
+### Issue 3: Aggregation of data that originates from different experiments with different objectives across a wide time period
+
+The last issue leads into the next issue.
+For many of the catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary "forestation" or "deforestation".
+However, the real situation is much more complex.
 

### 2022-09-01 paper/Forest_and_Water/SupplementaryMaterialPart1.Rmd
further changes 01092022

@@
 
-A particular problem was that many catchments had the wrong sign for the change in forest cover.
+**Need to fix up references**
+
+A particular problem was that many catchments appeared to have the wrong sign for the change in forest cover.
 There are many catchments with reported positive change in cover and a large increase in flow.
@@
 corrected to 16\%
-- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
+- 3, Baker Creek, Zhang and Wei, 2012. The original recorded 201.1\% change by @zhang2017 also seems wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
 - 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in @zhang2017.

### 2022-09-04 paper/Forest_and_Water/Forest_and_Water.Rmd
update readme
almos there

@@
 The columns in this data set (are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table @\ref(tab:table1). These abbreviations will later be used in the models.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \@ref(tab:table1). These abbreviations will later be used in the models.
 
@@
 The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
-In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref:(tab1)
+In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref(tab:table1)
 
@@
 
-```{r correlation, eval=T, echo = F}
+```{r correlation, eval=F, echo = F}
 All_data2 <- All_data2 %>%
@@
 While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
-The number of knots in the non-linear variables was based on assessment of the edf and k' **check wood 2006 and explain well**
+Following @wood2006, the number of degrees of freedom *k* in the non-linear variables was based on assessment of the effective degrees of freedom in the model output.
+If the effective degrees of freedom were close to *k - 1* then *k* was increased and the model rerun.
+By using shrinkage splines, this also results in the whole term being shrunk to zero [@wood2006].
 
@@
 The smoothed variables in the model can be inspected visually to identify if there are any issues with the fit.
-This is in addition to using `gam.check()` in the R package `mgcv` to test whether the number of knots is adequate.
+This is in addition to the earlier mentioned checks using `gam.check()` in the R package `mgcv` to test whether the number of degrees of freedom *k* is adequate.
 
@@
 
-Investigating the non-linear responses suggest that *Dryness* has a non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
+Investigating the non-linear responses suggest that *Dryness* has a clear non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
 Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship looks almost linear.
@@
 The overarching reason for combining past studies at a global scale is to infer relationships that can be used to make more general statements or develop more global scale modelling of impacts [i.e.
-@zhou2015; @jackson2005; @hoekvandijke2012].
+@zhou2015; @jackson2005; @hoekvandijke2022].
 Therefore, the results from the analysis could be seen as a confirmation of the earlier research [@zhang2017; @filoso2017; @zhou2015; @jackson2005].
-However, as is highlighted in the introduction there are four major issues with this type of analysis, and the results from this paper can also highlight this.
+However, the explaining power of the developed model is quite low and a lot of variation in the data is unexplained.
+As is highlighted in the introduction there are four major issues with this type of analysis, and the results from this paper also highlight these issues.
+Here, these issues are further explained.
 
@@
 
-- The papers from @almeida2016 and @ferreto2021 partly discuss the same experiment and the same catchment.
+- The papers from @almeida2016 and @ferreto2020 partly discuss the same experiment and the same catchment.
 In @almeida2016, the methods discuss how two experimental catchments of approximately 80ha in size which were harvested.
@@
 Throughout the paper the catchments are indicated as 100\% harvested and 30\% harvested.
-However, only after reading @ferreto2021, did we discover that in fact the 100\% and 30\% refer to the "eucalyptus plantation area", which was about 60\% of the total area.
+However, only after reading @ferreto2020, did we discover that in fact the 100\% and 30\% refer to the "eucalyptus plantation area", which was about 60\% of the total area.
 This is in fact mentioned in Table 1 in @almeida2016, but does not appear in the text.
@@
 Clearly it was a 100\% and 30\% change in forest cover, but only for the 60\% plantation cover, not for any of the other areas in the catchment, which included native vegetation and riparian vegetation.
-There are several other examples like this in the different papers (for example @blackie1979kimakia and @blackie1979kericho).
+There are several other examples like this in the different papers [for example @blackie1979kimakia; @blackie1979kericho].
 - Another example is the paper by @waterloo2007.
@@
 The original data base listed a 50 \% change in flow, but it is difficult to identify how this is calculated.
-We suspect that is results from $252/580*100 ≈ 50$ and $90/194 ≈ 50$.
+We suspect that results from $252/580*100 \approx 50$ and $90/194 \approx 50$.
 
@@
 The last issue leads into the next issue.
-For many of the catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary "forestation" or "deforestation".
-However, the real situation is much more complex.
-
-### Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
+For many of the small catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary "forestation" or "deforestation".
+However, the real situation is often much more complex and fuzzy.
+
+Many of the paired watershed experiments included a harvesting and replanting or regrowth after  harvesting or fire experiment [e.g.
+@cornish1993; @cornish2001; @webb2013].
+As a result, it becomes difficult to assess how we interpret the change in flow as a result of a change in cover.
+In many cases we would expect the flow to change over time as a function of the recovery [@jones2017] and therefore the timeseries of the flow needs to be assessed over a longer time.
+
+Many of the papers in the database report early results (for example 1 or 3 years after harvesting), but some also report longer time periods.
+As earlier work [@cornish2001; @jones2017] has highlighted, we can always expect a larger effect directly after harvesting, but this effect diminishes over time (even if it does not always return to the original state).
+Comparing studies reporting results directly after treatment to longer term studies therefore becomes problematic.
+
+In our work, the variable *length* was used in the model to test for some of these effects, but this was insignificant in the model (Table \@ref(tab:restrictlength)).
+Given the other variation in the data, this does not necessarily mean that there is no effect.
+
+This is further complicated by the variation in different types of clearing and the different types of vegetation.
+In the original @zhang2017 a variable to describe the *forest type* was included (Table \@ref(tab:table1)), but in the model this is not significant (Table \@ref(tab:m-all-linear)).
+This is probably because the broad classification used does not capture the actual variation in runoff response.
+In addition, as Figure \@ref(fig:corgraphs) shows, there is a correlation between coniferous forests and snow dominated hydrological regimes, further complicating the analysis.
+
+An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
+For example, @heath2014 found that catchments with resprouting species around Sydney indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent @zhou2015bushfire.
+
+As a result, it can be difficult to exactly pinpoint the change in flow as a result of the change in cover, as well as being difficult to assess what the exact change in cover actually was.
+
+As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
+As @jones2017 indicate, this really needs time series analysis of the different experiments.
+Some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+
+### Issue 4: Transcription errors in the data
+
+This issue seems to mainly occur if data is collected from other review papers.
+This might be because some of the original papers are difficult to locate and therefore values from reporting papers are used.
+In supplementary data part 1, several changes to the original data sets have been documented, and as can be seen several of these are transcription errors.
+
+This does influence the results in @zhang2017, comparing the results in Supplementary material 2 with the original paper.
+The main example is that in this study the largest catchment (watershed #1 in @zhang2017) had to be removed, as this study actually involved paired watershed experiments on very small plots, for which the characteristics were not recoverable.
+
+Clearly, this is a problem for all reviews that attempt to bring together large numbers of results from published papers, and where actual results are copied rather than using some sort of automated text analysis.
+
+In the end, careful review of the data and the original papers can circumvent most of this issue.
+And, making the data available (as @zhang2017, @zhou2015 and @filoso2017 have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
 
 ```{r, eval = F, echo =F}
-png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
+png("AssessmentTechnique_byDeltaQf.png", width = 2500, height = 1800, res = 300)
 All_data2 %>%
-  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
+  ggplot(aes(Assessment_technique, DeltaQf_perc)) + 
   geom_boxplot(colour = "blue") + theme_bw() +
@@
 
-```{r assessment, fig.cap="Boxplot of the log base 10 of the catchment area (in km2) for the different assessment techniques, showing the dominance of small catchments in the paired watershed experiments", out.width="90%", echo=F, eval =F}
-
-include_graphics("AssessmentTechnique_byArea.png")
-```
-
-<!--
-
-```{r, eval = F, echo = F}
-linear_model_E0 <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign + E0, 
-                           data = All_data2)
-#summary(linear_model_E0)
-#gam.check(linear_model_simple)
-
-```
-
-## The general relationship between change in forest cover and streamflow
-
-Following @zhang2017, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:
-
-\begin{equation}
-\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon (\#eq:eq3)
-\end{equation}
-
-```{r model1, echo = F}
-linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign, 
-                           data = All_data2 %>% filter(dataset == "original Zhang et al data"))
-#summary(linear_model_simple)
-#gam.check(linear_model_simple)
-
-```
-
-```{r model1b, echo = F}
-linear_model_simple_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
-                             Forest_Sign, 
-                           data = All_data2)
-#summary(linear_model_simple)
-r_sq <- round(summary(linear_model_simple_b)$r.sq,2)
-```
-
-```{r, echo =F}
-pander(round(rbind(summary(linear_model_simple)$p.table,
-                       summary(linear_model_simple_b)$p.table),2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change.
-The first three rows relate to the model using the original data base from Zhang et al.
-(2017).
-The bottom three rows are the results of the model including the new data.
-Clearly there is no major change arising from the additional data.")
-```
-
-The overall variance explained in this model (equation \@ref(eq:eq3)) is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow.
-The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%.
-However the change in flow is different for forest cover decreases compared to forest cover increases.
-In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase.
-So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.
-
-```{r, eval = F, echo = F}
-png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
-par(mfrow=c(2,2))
-gam.check(linear_model_simple)
-par(mfrow=c(1,1))
-dev.off()
-```
-
-```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
-include_graphics("residual_plot_model1.png")
-```
-
-Of importance here is to highlight the residuals of this regression (equation \@ref(eq:eq3) and Figure \@ref(fig:gamcheck)).
-These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)).
-In other words, the distribution of the residuals is somewhat fat-tailed.
-We will discuss this later.
-
-Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.
-Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total @zhang2017 data set.
-But this also means that our re-analysis of the data can be directly compared to the original study.
-
-However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods.
-The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.
-
-```{r model3, echo = F}
-model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Latitude + Longitude, data = All_data2)
-#summary(model3_includeLatLong)
-#gam.check(model3_includeLatLong)
-```
-
-```{r, echo =F}
-pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model based on the complete dataset and including Latitude and Longitude")
-```
-
-There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence.
-The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.
-
-## Impact of climate
-
-While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
-Potential evapotranspiration (*E0*) by itself was  not significant in the.
-Initially, we also tested models using only the annual average precipitation (*Pa (mm)*), but  interactions between precipitation and evapotranspiraton might be captured by the dryness index.
-Both dryness index  and *Pa (mm)* were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable.
-As a result only the dryness index was retained as a climate indicator to align with the earlier work by @zhang2017.
-Given that Latitude and Longitude were not significant, we dropped these from the model.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness  + \varepsilon (\#eq:eq5)
-\end{align}
-
-```{r model4, echo = F}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness, data = All_data2)
-#summary(model4_dryness)
-#gam.check(model4_dryness)
-```
-
-```{r, echo =F}
-pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
-```
-
-```{r testing_PCA, eval = F, echo = F, warning = F}
-require(FactoMineR)
-require(factoextra)
- res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length, Latitude, Longitude), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
- # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
- #  scree.plot
- #  
- #   var.plot <- fviz_pca_var(res.pca, axes = c(1, 2), col.var = "black")
- #  var.plot
- png("PCA_biplot_Alldata.png", width = 800, height = 800)
-  biplot_test <- fviz_pca_biplot(res.pca, axes = c(1, 2), repel = TRUE,
-                                                      col.var = "darkblue", # Variables color
-                            label="var", #"var" 
-                            col.ind = "red", # Individuals color
-                            title = "",labelsize = 8, pointsize = 4, arrowsize = 1) +
-        theme(text = element_text(size = 14, face = "bold"),
-              axis.title = element_text(size = rel(2), face = "bold"),
-        axis.text = element_text(size = rel(2), face = "bold"))
- print(biplot_test)
- dev.off()
-
-```
-
-Similar to *E0* or *Pa_mm*, the results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change.
-This might seem suprising in light of earlier reported results [@zhang2017; @filoso2017].
-In this case, the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`).
-However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify.
-This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`.
-As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.
-
-```{r, echo = F}
-pander(All_data2 %>% filter(Dryness > 4) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
-         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
-```
-
-There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution.
-These catchments are listed in Table \@ref(tab:drytable).
-
-## Is there a distinct effect of area?
-
-The second major variable is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017.
-Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.
-
-\begin{align}
-\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon (\#eq:eq6)
-\end{align}
-
-```{r modelArea, echo = F}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-model_Area <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness + log10(Area_km2), data = All_data2)
-#summary(model_Area)
-#gam.check(model_Area)
-```
-
-```{r, echo =F}
-pander(round(summary(model_Area)$p.table,2), caption = "(\\#tab:out-modelArea) Results of the model including Area and the dryness index")
-```
-
-```{r modelArea_b, echo = F, eval =T}
-#Zhang_all2 <- Zhang_all2 %>%
-#  mutate(length = To - From)
-
-model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
-                        Dryness*log10(Area_km2), data = All_data2)
-#summary(model_Area_b)
-#gam.check(model_Area)
-```
-
-The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km^2^)) (Table \@ref(tab:out-modelArea)).
-In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by `r abs(round(summary(model_Area)$p.table[5,1],1))`\%.
-Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area.
-Including the interaction Dryness*log10(Area (km^2^)) in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.
-
-```{r, echo =F}
-pander(round(summary(model_Area_b)$p.table,2), caption = "(\\#tab:out-modelArea-int) Results of the model including an interaction between Area and the dryness index")
-```
-
-```{r, eval = F, echo = F}
-All_data2 %>%
-  ggplot(aes(log10(Area_km2),Dryness, size = DeltaF_perc_pos)) + geom_point(colour = "blue", alpha =0.5)
-summary(lm(Dryness~log10(Area_km2),data = All_data2))
-```
-
-## Are some of the variables possibly non-linear?
-
-The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response.
-This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005).
-In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (*Dryness* and *Area*) were considered non-linear.
-As a result a shrinkage smoothing spline [@wood2006] was applied to these variables.
-
-<!-- *From* represents the starting date of the data collection.
-The idea is that this last variable gives some indication of the "age" of the experiment.
-Improvements in data collection could introduce a positive or negative bias between more recent and older studies -->
-
-<!--
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon (\#eq:eq7)
-\end{align}
-
-```{r model5, echo = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From,
-         mid_year = From + (To - From)/2)
-
-model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), k = 6, bs="ts") + 
-                    s(Dryness, k = 10, bs="ts") + 
-                    s(length, k = 50, bs="ts"), #+ 
-                    #s(From, k = 30, bs = "ts") ,
-                  data = All_data2)
-# summary(model5_all)
-# gam.check(model5_all)
-# plot(model5_all)
-```
-
-```{r , echo = F}
-pander(round(summary(model5_all)$p.table,2), caption = "(\\#tab:mfive-linear) Statistical summary for the linear terms in the model with non-linear terms")
-```
-
-```{r , echo = F}
-pander(round(summary(model5_all)$s.table,2), caption = "(\\#tab:mfive-smooth) Statistical summary for the smooth terms in the model with non-linear terms")
-
-```
-
-Including non-linearity (Equation \@ref(eq:eq7)) increases the overall explaining power of the model to an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)` and deviance explained of `r round(summary(model5_all)$dev.expl,2)`, but creates a few changes in the significance of the variables (Table \@ref(tab:mfive-smooth)).
-For example, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(model5_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(model5_all)$s.table[2,4],2)`)) and *length* (p = `r round(summary(model5_all)$s.table[3,4],2)`)) explain significant variation in the data.
-
-However, including the non-linearity also increases the chance of over fitting, as the smoothing splines allow significant flexibility.
-Including interactions between the smooth variables is also possible, but the results are difficult to interpret given the high flexibility of the two-dimensional smooth.
-Given the overall variability in the data we did not attempt this.
-
-Finally the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) are included i.e. Equation \@ref(eq:eq2).
-
-```{r model6_all, echo = F}
-
-model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), k = 5, bs="ts") + 
-                    s(Dryness, k = 10, bs="ts" ) + 
-                     s(length, k = 35, bs="ts") +# s(From, k = 40, bs = "ts") +
-                    Precip_data_type +  Assessment_technique +
-                    Forest_type +
-                    Hydrological_regime
-                    , data = All_data2)
-#summary(model6_all)
-#gam.check(model6_all)
-#plot(model6_all)
-```
-
-```{r, echo = F}
-pander(round(summary(model6_all)$p.table,2), caption = "(\\#tab:msix-linear) Statistical summary for the linear terms the full model")
-#plot(model6_all)
-```
-
-```{r , echo = F}
-pander(round(summary(model6_all)$s.table,2), caption = "(\\#tab:msix-smooth) Statistical summary for the smooth terms for the full model")
-#plot(model6_all)
-```
-
-This model (Tables \@ref(tab:msix-linear) and \@ref(tab:msix-smooth)) explains more of the variance, but the improvement is marginal compared to the previous model with a adjusted *r^2^* of `r round(summary(model6_all)$r.sq,2)`.
-This indicates that the categorical variables explain a limited amount of the overall variance in the change in flow data.
-However, it is interesting to note from Table \@ref(tab:msix-linear) that several of the assessment methods are significant.
-In particular Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are strongly significant ($p < 0.05$).
-In this case, *log10(Area (km^2^))* is no longer a significant predictor, the reasons for this will be discussed later.
-
-```{r visualisesmoothsblabla,echo =F, eval = F}
-#require(visreg)
-png("model6_smooths.png", width = 2500, height = 1800, res = 300)
-#par(mfrow=c(2,2))
-D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
-A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
-#From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
-(D + A) / L 
-# visreg(model6_all,"Dryness",gg=T) + theme_bw()
-# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
-# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
-#par(mfrow=c(1,1))
-dev.off()
-```
-
-```{r smoothsmodel6, echo = F, fig.cap="Visualisation of the smooth variables in the model, the shaded areas are the 95\\% confidence intervals associated with the fit of the smooth, the blue line is the mean smoothed relationship, with data plotted as individual points", out.width="90%"}
-include_graphics("model6_smooths.png")
-```
-
-\newpage
-
-Figure \@ref(fig:smoothsmodel6) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear, but, given all the data, not significant at p = `r round(summary(model6_all)$s.table[1,4],2)`, likely due to the high variance in the data.
-It still has a negative slope, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
-Both the *length* and *Dryness* variables are significant and show strong non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.
-For example, *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the data.
--->
-
-## Catchment size
-
-Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \@ref(fig:smoothsmodel7)), however, in contrast to @zhang2017, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \@ref(fig:overview)).
-In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
-
-An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change [@navas2019].
-Therefore, specifically if the  forest cover change is small relative to the catchment size, the effect of this change will be buffered.
-
-```{r, eval =T, echo =F}
-# All_data2 %>%
-#   ggplot(aes(log10(Area_km2), DeltaF_perc_pos)) + 
-#   geom_point(alpha = 0.5) + theme_bw()
-
-Area_deltaF <- lm(DeltaF_perc_pos~log10(Area_km2), data = All_data2)
-```
-
-There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between $\Delta$Forest cover and log10(Area (km^2^)) (linear regression indicates an adjusted *r^2^* of `r round(summary(Area_deltaF)$r.sq,2)` with a slope of `r as.numeric(round(coefficients(Area_deltaF)[2],2))`) indicating that for every 10 km^2^ increase in catchment size on the average, the forest cover change data is approximately  10\% lower.
-This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to "implement" in an experiment.
-
-This is also reflected in the second caveat.
-Most of the data from the smaller catchments are "real observed data" using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)).
-For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
-This is a confounding factor that is not easily addressed in the regression modelling attempted here.
-Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
-
-In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
-
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamflow.
-
-## Model residuals
-
-As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
-These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \@ref(fig:overview)), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \@ref(fig:gamcheckredLength)).
-
-```{r, eval = F, echo = F}
-png("residual_plot_model7.png", width = 2500, height = 1800, res = 300)
-par(mfrow=c(2,2))
-gam.check(model7)
-par(mfrow=c(1,1))
-dev.off()
-```
-```{r gamcheckredLength, echo =F, fig.cap="Residual plots for the final model indicating a small improvement in the residual distribution towards normal", out.width="90%"}
-include_graphics("residual_plot_model7.png")
-```
-
-The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
-Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
-In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
-This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \@ref(fig:increasedecrease)).
-
-## The effect of climate
-
-In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies [@zhang2017; @zhou2015; @filoso2017].
-This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
-As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
-This points to a need for future studies that unravel this interaction.
-
-## Interactions
-
-Generally this study did not consider interactions, but the above discussion suggest that there are possible several interactions.
-The relationships between forest cover change and *Area (km^2^)* and between *Area (km^2^)* and assessment technique have already been highlighted.
-However there are further unexplored interactions between the study length and some of the variables.
-
-A principle component analysis of the numeric data reveals some of these interactions (Figure \@ref(fig:pcaplot)), such as between *length* and *Dryness*.
-Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
-
-```{r pcaplot, fig.cap="Biplot of the first two principle components using a principle component analysis on the numerical values of the data set", out.width="90%", echo=F, eval =T}
-
-include_graphics("PCA_biplot_Alldata.png")
-```
-
-## Further considerations
-
-In contrast to @filoso2017, we did not identify that the length of the observation period is a significant variable in our final model.
-However, there are further confounding factors in the data, which were not analysed in this study.
-These were also classified by @filoso2017 and these factors might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow [@vervoort2021].
-This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
-
-Large databases based on historical studies, such as used here, also have significant uncertainty.
-While we have reviewed a large number of the studies in more detail, we have generally assumed that the assessments of past authors of the changes in streamflow and changes in forest cover are correct.
-More generally a lot of the data in the database are "summary data" extracted from the paper and this often neglects a lot of possibly important detail in the original studies.
-This introduces additional uncertainty in the analysis.
-
-By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
+```{r assessment, fig.cap="Boxplot of the variation in the change in flow for the different assessment techniques, showing the dominance of the variation and the outliers in the dataset in the paired watershed experiments", out.width="90%", echo=F, eval =T}
+
+include_graphics("AssessmentTechnique_byDeltaQf.png")
+```
+
+**Maybe include a figure that shows the variation in the flow data by Assessment technique**
+
+## General discussion
+
+**Still to do**
+
+- studies are conditional on the data
+- residuals: should the data be further transformed  
+- no interactions  
+- implications for other "meta-analysis" studies  
+- future research needs
 
@@
 
-More rigorous checking of an existing database on catchment studies relating to changes in forest cover to changes in flow and more detailed statistical analysis results in both agreement and disagreement with older studies.
-It demonstrates that analysis of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often fail to capture the complexity in the data.
-The variablility in the aggregated historical data is simply too large.
-
-As with any analysis, the results of the statistical analysis in this paper need to be considered "conditional on the data". 
-Conditional on the data, it can be determined that the impact of forestry on streamflow:
-
-- is greater for forest clearing then for reforestation;
-
-- is reduced for larger watersheds;
-
-- Increases for drier watersheds; and
-
-- is sensitive to the assessment method used in the historical data.
-
-Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area, especially in relation to larger catchments.
-Furthermore, the current study analyses a large global dataset of aggregated data.
-This analysis does not exclude more local and regional effects that cannot be identified in the global data.
+This study demonstrates that analyses of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+
+There are four major interlinked reasons why this is the case, and this has implications for meta-analyses in Environmental Science and Hydrology in general:
+
+- The existence of latent variables in the data that create the appearance of a relationship that really does not exist;  
+- The difficulty in fully interpreting the specifics of different studies;  
+- The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and  
+- The chance of transcription errors influencing the data.
+
+Any statistical analysis, including the one in this paper, needs to be considered "conditional on the data", and given the issues indicated, extrapolation of the results of summary studies into global hydrological models has to be done with great care.
+Better would be to explicitly include uncertainty in the extrapolation of the results.
+
+This therefore has implications for the recent growth in meta-analysis review papers, which has been boosted by increased computational capacity and much better on-line accessible data bases with research data.
+Clearly, this requires careful definition of the search terms, and follow-up review of the harvested papers, as well as an understanding that the statistical relationships can be hiding other unknown factors.
+As the old adagium says: Correlation is not the same as causation.
+
+Future research should ...
+
  In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
-```
-
-<!-- testing whether dryness is related to Area -->
-
-```{r testing_dryness, echo = F, eval= F}
-All_data2 %>% ggplot(aes(Dryness,log10(Area_km2), colour = DeltaF_perc_pos)) + geom_point() + theme_bw()
-```
-
-<!--Check the size distribution of the watersheds -->
-
-```{r, eval =F, echo =F}
-Zhang_all %>%
-  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
-  scale_x_log10()
-total <- nrow(Zhang_all)
-length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
-```
-```{r, eval =F, echo = F}
-Zhang_all2 %>%
-  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
-```
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(log10(Area_km2), DeltaQf_perc, colour = DeltaF_perc)) + 
-  geom_point(alpha = 0.5) + theme_bw()
-```
-
-```{r, eval =F, echo =F}
-All_data2 %>%
-  ggplot(aes(log10(Area_km2), Dryness)) + 
-  geom_point(colour = "blue") + theme_bw()
-```
-
-<!--- Analysis of Methods by dryness -->
-
-```{r, eval =F, echo =F}
-Zhang_all2 %>%
-  ggplot(aes(Assessment_technique, Dryness)) + 
-  geom_boxplot(colour = "blue") + theme_bw()
-```
-
-<!--
-```{r model4b, echo = F, eval = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From)
-
-model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
-                 s(length, bs = "ts"),
-                  data = All_data2 %>% filter(length < 60))
-summary(model4b)
-
-```
--->
-```{r model5b, echo = F, eval = F}
-All_data2 <- All_data2 %>%
-  mutate(length = To - From)
-
-model5_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    te(log10(Area_km2),Dryness, bs="ts") +  
-                  s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
-                    s(length, bs="ts") + s(From, bs = "ts") ,
-                  data = All_data2)
-summary(model5_b)
-plot(model5_b)
-```
-
-```{r model9_interact, echo = F, eval = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
-model9_interact <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
-                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
-                      ti(Dryness,From, bs="ts" ) + #s(Latitude, bs="ts") + 
-                      ti(log10(Area_km2),length, bs="ts" ) +
-                     # log10(Area_km2):Assessment_technique +
-                      s(length, bs = "ts") + 
-                      s(From, bs = "ts") +
-                    Precip_data_type + Forest_type + Assessment_technique +
-                    Hydrological_regime, data = All_data3 %>% filter(length < 60, Dryness <= 4, From > 1920))
-#gam.check(model9_interact)
-# plot(model9_interact)
-summary(model9_interact)
-
-```
 

### 2022-09-04 paper/Forest_and_Water/Forest_and_Water.tex
almos there

@@
 There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales.
-However, there is also a real danger in this process, which is what we will highlight in this paper. There are three potential errors (or limitations) in the mentioned global analyses:
+However, there is also a real danger in this process, which is what we will highlight in this paper. There are four potential errors (or limitations) in the mentioned global analyses:
 
@@
 The columns in this data set (are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in \citet{zhang2017}), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table @\ref(tab:table1). These abbreviations will later be used in the models.
+Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \ref{tab:table1}. These abbreviations will later be used in the models.
 
@@
 The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
-In addition to these earlier identified factors, this study also tested for the factors listed in Table @ref:(tab1)
+In addition to these earlier identified factors, this study also tested for the factors listed in Table \ref{tab:table1}
 
@@
 
+** DO WE NEED TO REDO THE MAP? Eliana please provide code **
+
 Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
 \begin{figure}
@@
 While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
-The number of knots in the non-linear variables was based on assessment of the edf and k' \textbf{check wood 2006 and explain well}
+Following \citet{wood2006}, the number of degrees of freedom \emph{k} in the non-linear variables was based on assessment of the effective degrees of freedom in the model output.
+If the effective degrees of freedom were close to \emph{k - 1} then \emph{k} was increased and the model rerun.
+By using shrinkage splines, this also results in the whole term being shrunk to zero \citep{wood2006}.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
-\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
+\caption{\label{tab:m-all-linear} Statistical summary for the linear terms the full model}\tabularnewline
 \toprule()
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{(\#tab:m\_all-smooth) Statistical summary for the smooth terms for the full model}\tabularnewline
+\caption{\label{tab:m-all-smooth} Statistical summary for the smooth terms for the full model}\tabularnewline
 \toprule()
@@
 
-Inspecting the significance of the variables (Table @ref(tab:m\_all-linear) and Table @ref(tab:m\_all-smooth)) indicates some interesting features.
+Inspecting the significance of the variables (Table \ref{tab:m-all-linear} and Table \ref{tab:m-all-smooth}) indicates some interesting features.
 
@@
 
-Furthermore Table @ref(tab:m\_all-linear) indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
+Furthermore Table \ref{tab:m-all-linear} indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
 In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables (\(p < 0.05\)).
@@
 The smoothed variables in the model can be inspected visually to identify if there are any issues with the fit.
-This is in addition to using \texttt{gam.check()} in the R package \texttt{mgcv} to test whether the number of knots is adequate.
+This is in addition to the earlier mentioned checks using \texttt{gam.check()} in the R package \texttt{mgcv} to test whether the number of degrees of freedom \emph{k} is adequate.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
-\caption{(\#tab:m\_red-linear) Statistical summary for the linear terms the restricted model}\tabularnewline
+\caption{\label{tab:m-red-linear} Statistical summary for the linear terms the restricted model}\tabularnewline
 \toprule()
@@
 
-Investigating the non-linear responses suggest that \emph{Dryness} has a non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
+Investigating the non-linear responses suggest that \emph{Dryness} has a clear non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
 Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship looks almost linear.
@@
 In contrast, data related to large catchments are related to smaller changes in forest cover and different methods, such as hydrological modelling and elasticity analysis.
-This confirms the model results (Table @ref(tab:model\_assess-linear)) and the earlier correlation analysis (Figure \ref{fig:corgraphs}).
-
-It is possible that one of the reasons why \citet{zhang2017} separated their analysis in large (\textgreater{} 1000 km\textsuperscript{2}) and small (\textless{} 1000 km\textsuperscript{2}) catchments, is that they realised this difference in assessment methods.
-However, this is not explicitly identified.
+This confirms the model results (Table \ref{tab:modelassess-linear}) and the earlier correlation analysis (Figure \ref{fig:corgraphs}).
+
+It is possible that one of the reasons why \citet{zhang2017} separated their analysis in large (\textgreater{} 1000 km\textsuperscript{2}) and small (\textless{} 1000 km\textsuperscript{2}) catchments, is that they realised this difference in assessment methods and wanted to account for this.
+However, this is not explicitly identified, and there is no real physical explanation of the 1000 km\textsuperscript{2} threshold.
 
@@
 
-\hypertarget{catchment-size}{%
-\subsection{Catchment size}\label{catchment-size}}
-
-Essentially, the overall analysis shows that there is a clear effect of catchment size (Figure \ref{fig:smoothsmodel7}), however, in contrast to \citet{zhang2017}, there is no evidence of a distinct threshold in the size of the catchment that determines the change in the streamflow as a result of changes in forestry.
-If anything the scatter in the data (in the change in flow) is greater for the smaller catchments then for the larger catchments (Figure \ref{fig:overview}).
-In other words, the response to changes in forest cover is more consistent for larger catchments than it is for smaller catchments.
-
-An explanation for the catchment size effect might be that large catchments have more storage and longer flow paths and therefore have more opportunity to buffer the effects of forest cover change \citep{navas2019}.
-Therefore, specifically if the forest cover change is small relative to the catchment size, the effect of this change will be buffered.
-
-There are two caveats on this explanation.
-The first is that there is a distinct trend in the data between \(\Delta\)Forest cover and log10(Area (km\textsuperscript{2})) (linear regression indicates an adjusted \emph{r\textsuperscript{2}} of 0.35 with a slope of -9.67) indicating that for every 10 km\textsuperscript{2} increase in catchment size on the average, the forest cover change data is approximately 10\% lower.
-This is basically a result of the fact that large changes in forest cover in larger catchments are difficult to ``implement'' in an experiment.
-
-This is also reflected in the second caveat.
-Most of the data from the smaller catchments are ``real observed data'' using paired watershed studies, while for larger catchments, the data are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \ref{fig:overview}).
-For larger catchments, these techniques all provide an approximation of the effect of forestry on streamflow rather than a direct comparison of catchments.
-This is a confounding factor that is not easily addressed in the regression modelling attempted here.
-Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.
-
-In other words, the current data sets cannot resolve whether there actually is a non-linear catchment size × forest cover effect, which then feeds into the buffering in larger catchments.
-
-Apart from a difficulty of analysing complex confounding factors in the data, a general limitation of the type of analysis presented is that this work does not consider the spatial arrangement of the forest clearing in the catchments.
-While for fully or almost fully cleared smaller catchments this might not be an issue, it is perceivable that for larger catchments being partially cleared, a interaction between spatial location and clearing could be a factor in determining the change in streamflow.
-Clearing head water catchments on shallower soils might have a larger impact than clearing in downstream areas on deeper soils.
-As a result there is still a need for catchment scale studies related to the impact of changes in forest cover on streamflow.
-
-\hypertarget{model-residuals}{%
-\subsection{Model residuals}\label{model-residuals}}
-
-As pointed out earlier the residuals of the model diverge from the normal distribution for large positive and large negative residuals.
-These residuals are mainly associated with the small catchments from the paired watershed studies (Figure \ref{fig:overview}), which show very high variability.
-The final model removing the data with large values of Dryness and long study lengths has removed some of the spreading, mainly for the large negative residuals (Figure \ref{fig:gamcheckredLength}).
+The overarching reason for combining past studies at a global scale is to infer relationships that can be used to make more general statements or develop more global scale modelling of impacts \citep[i.e.][]{zhou2015, jackson2005, hoekvandijke2022}.
+Therefore, the results from the analysis could be seen as a confirmation of the earlier research \citep{zhang2017, filoso2017, zhou2015, jackson2005}.
+However, the explaining power of the developed model is quite low and a lot of variation in the data is unexplained.
+As is highlighted in the introduction there are four major issues with this type of analysis, and the results from this paper also highlight these issues.
+Here, these issues are further explained.
+
+\hypertarget{issue-1-latent-variables-are-not-included-in-the-typical-single-covariate-analysis}{%
+\subsubsection{Issue 1: Latent variables are not included in the typical single covariate analysis}\label{issue-1-latent-variables-are-not-included-in-the-typical-single-covariate-analysis}}
+
+The results show that it is simply impossible to analyse a single covariate relationship, as there are several latent variables in the data.
+An example of this is the general relationship of the change in flow as a function of the change in forest cover.
+Clearly the relationship is highly impacted by the fact that all the small catchments have large changes in forest cover and are all associated with paired watershed experiments.
+Without taking these factors into account, a definite answer about the impact of forest cover on the change in flow cannot be given.
+Furthermore, the large variability in the change in flow data for these small catchments (Figure \ref{fig:overview}) indicates that there is a further (unknown) variable that explains the variation in the data.
+
+If the remaining variation in the residuals is small relative to the trend, then there is little need to identify further latent variables, but if the variation is large, then it is unclear if it is the latent variable that determines the trend, or the actual relationship in the data.
+
+Similarly, the data for the larger catchments containing smaller changes in forest cover are dominated by hydrological modelling studies, resulting in a further complication.
+If the reponse of the streamflow in the modelling studies is the result of the conceptualised relationship between streamflow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in streamflow is real, or simply a result of a pre-conceived model relationship.
+Is the smaller variation in the data for smaller changed in forest cover (Figure \ref{fig:overview}) a result of similar conceptualised model relationships, or actual variation between catchments and climate types?
+Currently this question cannot be answered.
+
+This becomes problematic when extrapolated to larger scales.
+A clear example of this is the paper by \citet{hoekvandijke2022} where the conceptualised relationship between forest cover and streamflow pre-determines the outcomes of the global modelling.
+
+The only way to analyse changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in \citet{levy2018}).
+
+We are not arguing that there is no relationship between streamflow and forest cover, and there might indeed be a global relationship that can be discovered.
+But, this relationship can only be discovered if we are able to address some of the major other factors that explain the variability, and work with actual data and not model outputs.
+
+\hypertarget{issue-2-interpretation-errors-due-to-complex-descriptions-of-the-experiments-in-the-original-papers}{%
+\subsubsection{Issue 2: Interpretation errors due to complex descriptions of the experiments in the original papers}\label{issue-2-interpretation-errors-due-to-complex-descriptions-of-the-experiments-in-the-original-papers}}
+
+The second major issue that became clear from reviewing many of the original papers is that some of the variability might be an interpretation problem.
+In many cases the original description in the paper is interpreted to extract the \% change in streamflow from the \% change in forest cover.
+This seems like a simple activity, but this is not always the case.
+
+Two examples can be highlighted:
+
+\begin{itemize}
+\tightlist
+\item
+  The papers from \citet{almeida2016} and \citet{ferreto2020} partly discuss the same experiment and the same catchment.
+In \citet{almeida2016}, the methods discuss how two experimental catchments of approximately 80ha in size which were harvested.
+One catchment was 100\% harvested and the other 30\% harvested.
+Throughout the paper the catchments are indicated as 100\% harvested and 30\% harvested.
+However, only after reading \citet{ferreto2020}, did we discover that in fact the 100\% and 30\% refer to the ``eucalyptus plantation area'', which was about 60\% of the total area.
+This is in fact mentioned in Table 1 in \citet{almeida2016}, but does not appear in the text.
+The question then becomes how to interpret this in the data base for this paper.
+Clearly it was a 100\% and 30\% change in forest cover, but only for the 60\% plantation cover, not for any of the other areas in the catchment, which included native vegetation and riparian vegetation.
+There are several other examples like this in the different papers \citep[for example][]{blackie1979kimakia, blackie1979kericho}.
+\item
+  Another example is the paper by \citet{waterloo2007}.
+This modelling study in Fiji of the clearing of a catchment reports the changes in streamflow over parts of the year.
+For a period of 324 days the streamflow increased from 252 mm to 580 mm (a 230\% increase if calculated as \(580/252*100\)) and for a second period of 309 days the streamflow increased from 90 mm to 194 mm (a 215 \% increase).
+However, how we convert this to a change in annual flow (which most of the other data relate to) is difficult.
+The original data base listed a 50 \% change in flow, but it is difficult to identify how this is calculated.
+We suspect that results from \(252/580*100 \approx 50\) and \(90/194 \approx 50\).
+\end{itemize}
+
+Clearly, interpreting older papers can be difficult and this can result in variation and errors in the data that is being analysed.
+Similar to the last issue, if these errors only introduce small variation in the data, then it will not limit the interpolation to larger scales.
+At this point, it is not clear if this is indeed the case.
+The large variation in the experimental watershed data suggests that this might be a more serious problem.
+
+\hypertarget{issue-3-aggregation-of-data-that-originates-from-different-experiments-with-different-objectives-across-a-wide-time-period}{%
+\subsubsection{Issue 3: Aggregation of data that originates from different experiments with different objectives across a wide time period}\label{issue-3-aggregation-of-data-that-originates-from-different-experiments-with-different-objectives-across-a-wide-time-period}}
+
+The last issue leads into the next issue.
+For many of the small catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary ``forestation'' or ``deforestation''.
+However, the real situation is often much more complex and fuzzy.
+
+Many of the paired watershed experiments included a harvesting and replanting or regrowth after harvesting or fire experiment \citep[e.g.][]{cornish1993, cornish2001, webb2013}.
+As a result, it becomes difficult to assess how we interpret the change in flow as a result of a change in cover.
+In many cases we would expect the flow to change over time as a function of the recovery \citep{jones2017} and therefore the timeseries of the flow needs to be assessed over a longer time.
+
+Many of the papers in the database report early results (for example 1 or 3 years after harvesting), but some also report longer time periods.
+As earlier work \citep{cornish2001, jones2017} has highlighted, we can always expect a larger effect directly after harvesting, but this effect diminishes over time (even if it does not always return to the original state).
+Comparing studies reporting results directly after treatment to longer term studies therefore becomes problematic.
+
+In our work, the variable \emph{length} was used in the model to test for some of these effects, but this was insignificant in the model (Table \ref{tab:restrictlength}).
+Given the other variation in the data, this does not necessarily mean that there is no effect.
+
+This is further complicated by the variation in different types of clearing and the different types of vegetation.
+In the original \citet{zhang2017} a variable to describe the \emph{forest type} was included (Table \ref{tab:table1}), but in the model this is not significant (Table \ref{tab:m-all-linear}).
+This is probably because the broad classification used does not capture the actual variation in runoff response.
+In addition, as Figure \ref{fig:corgraphs} shows, there is a correlation between coniferous forests and snow dominated hydrological regimes, further complicating the analysis.
+
+An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
+For example, \citet{heath2014} found that catchments with resprouting species around Sydney indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent \citet{zhou2015bushfire}.
+
+As a result, it can be difficult to exactly pinpoint the change in flow as a result of the change in cover, as well as being difficult to assess what the exact change in cover actually was.
+
+As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
+As \citet{jones2017} indicate, this really needs time series analysis of the different experiments.
+Some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+
+\hypertarget{issue-4-transcription-errors-in-the-data}{%
+\subsubsection{Issue 4: Transcription errors in the data}\label{issue-4-transcription-errors-in-the-data}}
+
+This issue seems to mainly occur if data is collected from other review papers.
+This might be because some of the original papers are difficult to locate and therefore values from reporting papers are used.
+In supplementary data part 1, several changes to the original data sets have been documented, and as can be seen several of these are transcription errors.
+
+This does influence the results in \citet{zhang2017}, comparing the results in Supplementary material 2 with the original paper.
+The main example is that in this study the largest catchment (watershed \#1 in \citet{zhang2017}) had to be removed, as this study actually involved paired watershed experiments on very small plots, for which the characteristics were not recoverable.
+
+Clearly, this is a problem for all reviews that attempt to bring together large numbers of results from published papers, and where actual results are copied rather than using some sort of automated text analysis.
+
+In the end, careful review of the data and the original papers can circumvent most of this issue.
+And, making the data available (as \citet{zhang2017}, \citet{zhou2015} and \citet{filoso2017} have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
 
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{residual_plot_model7} \caption{Residual plots for the final model indicating a small improvement in the residual distribution towards normal}\label{fig:gamcheckredLength}
+\includegraphics[width=0.9\linewidth]{AssessmentTechnique_byDeltaQf} \caption{Boxplot of the variation in the change in flow for the different assessment techniques, showing the dominance of the variation and the outliers in the dataset in the paired watershed experiments}\label{fig:assessment}
 \end{figure}
 
-The reason why the regression model is better able to resolve the variance in the data for the negative residuals (generally related to increases in forest cover) compared the large positive residuals might link back to the issue of buffering and flow paths in the catchments.
-Small catchments that are stripped of most of the forest cover would provide little buffering, interception and infiltration, does leading to greater changes in flow.
-In contrast, revegetated catchments would have increased interception and buffering and therefore relatively smaller changes in flow.
-This also provides an explanation for the differences between forest cover removal and forest cover restoration (Figure \ref{fig:increasedecrease}).
-
-\hypertarget{the-effect-of-climate}{%
-\subsection{The effect of climate}\label{the-effect-of-climate}}
-
-In drier catchments, changes in forest cover have greater impact on flow, which is similar to the observations in earlier studies \citep{zhang2017, zhou2015, filoso2017}.
-This is most likely because in these catchments the overall flow is surface flow dominated and therefore the buffering that is afforded by groundwater flow is not as important.
-As the dataset currently does not include a separate variable for groundwater inputs (although this effect is estimated in several of the studies), the effect again cannot be analysed separately.
-This points to a need for future studies that unravel this interaction.
-
-\hypertarget{interactions}{%
-\subsection{Interactions}\label{interactions}}
-
-Generally this study did not consider interactions, but the above discussion suggest that there are possible several interactions.
-The relationships between forest cover change and \emph{Area (km\textsuperscript{2})} and between \emph{Area (km\textsuperscript{2})} and assessment technique have already been highlighted.
-However there are further unexplored interactions between the study length and some of the variables.
-
-A principle component analysis of the numeric data reveals some of these interactions (Figure \ref{fig:pcaplot}), such as between \emph{length} and \emph{Dryness}.
-Including these interactions into the smooths of the models (data not shown) increases the explained variance slightly but does not fundamentally change the significance of the different variables.
-
-\begin{figure}
-\includegraphics[width=0.9\linewidth]{PCA_biplot_Alldata} \caption{Biplot of the first two principle components using a principle component analysis on the numerical values of the data set}\label{fig:pcaplot}
-\end{figure}
-
-\hypertarget{further-considerations}{%
-\subsection{Further considerations}\label{further-considerations}}
-
-In contrast to \citet{filoso2017}, we did not identify that the length of the observation period is a significant variable in our final model.
-However, there are further confounding factors in the data, which were not analysed in this study.
-These were also classified by \citet{filoso2017} and these factors might create biases in the data set that can impact the overall assessment.
-For example, snow dominated hydrological regimes (SD) tend to be dominated by Coniferous Forests (CF), while the majority of the rain dominated regimes are all broadleaf of mixed type forests (BF or MF).
-However, the forest type classification is very coarse and does not fully capture possible physiological differences that could affect evapotranspiration and therefore changes in streamflow \citep{vervoort2021}.
-This is not further investigated in this study, but with more data available this might provide further opportunities for investigations.
-
-Large databases based on historical studies, such as used here, also have significant uncertainty.
-While we have reviewed a large number of the studies in more detail, we have generally assumed that the assessments of past authors of the changes in streamflow and changes in forest cover are correct.
-More generally a lot of the data in the database are ``summary data'' extracted from the paper and this often neglects a lot of possibly important detail in the original studies.
-This introduces additional uncertainty in the analysis.
-
-By making the updated the database of this study available, we hope that this provides further incentive to investigate the impact of land cover change on streamflow more generally.
+\textbf{Maybe include a figure that shows the variation in the flow data by Assessment technique}
+
+\hypertarget{general-discussion}{%
+\subsection{General discussion}\label{general-discussion}}
+
+\textbf{Still to do}
+
+\begin{itemize}
+\tightlist
+\item
+  studies are conditional on the data
+\item
+  residuals: should the data be further transformed\\
+\item
+  no interactions\\
+\item
+  implications for other ``meta-analysis'' studies\\
+\item
+  future research needs
+\end{itemize}
 
@@
 
-More rigorous checking of an existing database on catchment studies relating to changes in forest cover to changes in flow and more detailed statistical analysis results in both agreement and disagreement with older studies.
-It demonstrates that analysis of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often fail to capture the complexity in the data.
-The variablility in the aggregated historical data is simply too large.
-
-As with any analysis, the results of the statistical analysis in this paper need to be considered ``conditional on the data''.
-Conditional on the data, it can be determined that the impact of forestry on streamflow:
+This study demonstrates that analyses of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+
+There are four major interlinked reasons why this is the case, and this has implications for meta-analyses in Environmental Science and Hydrology in general:
 
 \begin{itemize}
-\item
-  is greater for forest clearing then for reforestation;
-\item
-  is reduced for larger watersheds;
-\item
-  Increases for drier watersheds; and
-\item
-  is sensitive to the assessment method used in the historical data.
+\tightlist
+\item
+  The existence of latent variables in the data that create the appearance of a relationship that really does not exist;\\
+\item
+  The difficulty in fully interpreting the specifics of different studies;\\
+\item
+  The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and\\
+\item
+  The chance of transcription errors influencing the data.
 \end{itemize}
 
-Stronger statements about the trends in the change in flow cannot be made until more data or better data becomes available in this area, especially in relation to larger catchments.
-Furthermore, the current study analyses a large global dataset of aggregated data.
-This analysis does not exclude more local and regional effects that cannot be identified in the global data.
+Any statistical analysis, including the one in this paper, needs to be considered ``conditional on the data'', and given the issues indicated, extrapolation of the results of summary studies into global hydrological models has to be done with great care.
+Better would be to explicitly include uncertainty in the extrapolation of the results.
+
+This therefore has implications for the recent growth in meta-analysis review papers, which has been boosted by increased computational capacity and much better on-line accessible data bases with research data.
+Clearly, this requires careful definition of the search terms, and follow-up review of the harvested papers, as well as an understanding that the statistical relationships can be hiding other unknown factors.
+As the old adagium says: Correlation is not the same as causation.
+
+Future research should \ldots{}
+
 In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.

### 2022-09-07 paper/Forest_and_Water/Forest_and_Water.Rmd
update ET0 figure script

@@
 
-The impacts of global deforestation and reforestation are important through their influence on streamflow and both blue and green water availability [@hoekvandijke2022; @schyns2019].
+There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects [@wang2020; @hoekvandijke2022].
+More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability [@hoekvandijke2022; @schyns2019].
 The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
@@
 However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007].
+For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% changes in streamflow as a result of reforestation.
 Similarly a modelling study by @beck2013 found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
@@
 
-The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies with changes in streamflow due to changes in forest cover.
+There has been a recent push in the hydrological community [@evaristo2017] to use 'meta-analysis' to summarise past studies.
+The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.
+
+From the most recent reviews on forest impact on streamflow [@zhang2017; @filoso2017], the study by @filoso2017 is a clear meta-analysis, while the other is not.
+However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data.
 The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
@@
 
-There are some clear limitations in these studies.
+There are some clear limitations in these studies, and some of this applies to meta-analyses in general.
 The main method in the work by @zhang2017 is a single covariate linear regression.
@@
 Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
-There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales. 
-However, there is also a real danger in this process, which is what we will highlight in this paper. There are four potential errors (or limitations) in the mentioned global analyses:
+There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales [@evaristo2017].
+However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:
 
-- Latent variables that are not included in the typical single covariate analysis;
+- Impact of latent variables that are not included in the typical single covariate analysis;
 - Interpretation errors due to incomplete descriptions of the experiments in the original papers;  
-- Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally    
+- Aggregation of data that originates from different experiments with different objectives across a wide time period, but have similar keywords; and, finally    
 - Transcription errors in the data, especially if data is collected from other review papers as some of the original papers are difficult to locate.
 
-The aim of this paper is to highlight examples of each of these limitations, how they have influenced past work, and provide suggestions of how we can overcome these limitations.
+The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation.
+This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
 In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
@@
 
-```{r table1, echo =F}
+```{r datasummary, echo =F}
 
@@
 
-pander(Table1, caption = "Summary of abbreviations of factors used in the Zhang et al. (2017) data set")
+```
+```{r table1, echo=F}
+pander(Table1, caption = "(\\#tab:table1) Summary of abbreviations of factors used in the Zhang et al. (2017) data set")
 ```

### 2022-09-07 scripts/worldETmap_paper.R

@@ -1,0 +1,64 @@
+######paper map
+
+library(readxl)
+library(readr)
+library(plotly)
+library(ggmap)
+library(sf)
+library(raster)
+library(rgdal)
+
+#read files
+LargeData_Catchments <- read_csv("C:/Users/nanin/Downloads/LargeData_Catchments.csv")
+LargeC<-LargeData_Catchments[,1:4]
+colnames(LargeC)<- c('Number',"Name",'latitude','longitude')
+
+SmallData_Catchments <- read_csv("C:/Users/nanin/Downloads/SmallData_Catchments.csv")
+SmallC<-SmallData_Catchments[,1:4]
+colnames(SmallC)<- c('Number',"Name",'latitude','longitude')
+
+NewData_Catchments <- read_csv("C:/Users/nanin/Downloads/NewData_Catchments.csv")
+NewC<-NewData_Catchments[,1:4]
+colnames(NewC)<- c('Number',"Name",'latitude','longitude')
+
+#combine datasets
+all_catchments<-bind_rows(LargeC,SmallC,NewC)
+
+#plot
+ETmap <- raster('./extraET0.tif')
+ETmapdf <- as.data.frame(ETmap)
+extaraET0 <- raster("./extraET0.tif")
+# regular sample of 500k cells
+
+ET0 <- sampleRegular(extaraET0, 
+                     size = 5e5, 
+                     asRaster = TRUE) %>% 
+  as.data.frame(xy = TRUE, 
+                na.rm = TRUE) %>% 
+  setNames(c("x", 
+             "y", 
+             "ET0_tc"))
+head(ET0)
+
+ET0 <- ET0 %>%
+  mutate(et0_band = cut(ET0_tc, breaks = 6))%>%
+  group_by(et0_band)
+#see bands
+ggplot() +
+  geom_bar(data = ET0, aes(et0_band))
+
+#plot to paper
+ggplot() +
+  theme(panel.grid.major = element_line(color = "grey"))+
+  xlab("Longitude") + ylab("Latitude")+
+  coord_sf(expand = FALSE) +
+  scale_y_continuous(limits = c(-79, 79))+
+  theme(panel.background = element_rect(fill = 'aliceblue')) +
+  geom_raster(data = ET0,aes(x = x,y = y,fill = et0_band))+
+  scale_fill_viridis_c()+
+  geom_point(data= all_catchments,aes(x=longitude, y=latitude, color="cacthment"),alpha=0.5)
+
[…]

### 2022-09-22 paper/Forest_and_Water/Forest_and_Water.Rmd
version for interbal review

@@
 Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
-Generalised additive models were used to run flexible regressions including multiple variables.
-  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
-Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover.
-Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data.
-Smaller catchment dominate the database with 42\% of the data < 1 km^2^ and 65\% of the data < 10 km^2^.
-Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
+The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling.
+The overall results of the modelling indicates similar results to previous papers.
+The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results.
+The area of the catchment is strongly related to the assessment technique and the variability in the response data.
+  A careful review of the data and the methodology in the previous papers highlights that there are  four interlinked reasons that are problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.
 keywords: 
-  - keyword1
-  - keyword2
+  - data analysis
+  - forest cover change
+  - global scales
+  - literature review
 journal: "Journal of Hydrology"
@@
     base_format: rticles::elsevier_article
+header-includes:
+    - \usepackage{setspace}\doublespacing
 ---
@@
 
-There has been a recent push in the hydrological community [@evaristo2017] to use 'meta-analysis' to summarise past studies.
+There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies.
 The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.
@@
 Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
-There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales [@evaristo2017].
+There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales [@evaristo2020metaanalysis].
 However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:
@@
 library(ggcorrplot)
+require(latex2exp)
 suppressWarnings(require(knitr))
@@
 ```{r, echo = F, warning = F, message=F, eval =F}
-require(latex2exp)
 png("DataExploration.png", width = 2500, height = 1800, res = 300)
@@
 ```
-
-** DO WE NEED TO REDO THE MAP? Eliana please provide code **
 
@@
 The detail of the comparison is highlighted in Supplementary material part 3.
-However, generally the results of the analysis showed two main points (Table \@ref(tab:tablecomparison):
+However, generally the results of the analysis showed two main points (Table \@ref(tab:tablecomparison)):
 
@@
 
-As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \@ref(fig:gamcheckmodelall).
+As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \@ref(fig:gamcheckmodelall)).
 Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall)).
@@
 One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess).
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \@ref(tab:tableassess)).
 <!-- and Figure \@ref(fig:assessment)).
@@
 
-Clearly, interpreting older papers can be difficult and this can result in variation and errors in the data that is being analysed.
+Clearly, interpreting older papers can be difficult and this can result in variation in the data that is being analysed.
 Similar to the last issue, if these errors only introduce small variation in the data, then it will not limit the interpolation to larger scales.
@@
 As @jones2017 indicate, this really needs time series analysis of the different experiments.
-Some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+However, some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+We will discuss this further below.
 
@@
 
-**Maybe include a figure that shows the variation in the flow data by Assessment technique**
-
 ## General discussion
 
-**Still to do**
+In this paper, a few studies have been singled out in the analysis.
+The choice of focus was mainly driven by the data that was made available by the authors of these papers [@zhang2017; @filoso2017], which provide a rich case study for the current paper.
 
-- studies are conditional on the data
-- residuals: should the data be further transformed  
-- no interactions  
-- implications for other "meta-analysis" studies  
-- future research needs
+Field research is by nature limited in space and time, due to the high costs involved of setting up experiments.
+This is particularly true for experiments in hydrology and forest hydrology, where field sites need to cover sufficient spatial and temporal variability.
+This means there is a general need to extrapolate the local results to larger scales to inform decision making and policy.
+
+However, as is demonstrated in this paper, there are multiple issues when this local scale data is extrapolated to larger scales.
+It clearly demonstrates that the results of any model (in this case a regression model) is highly dependent on the data, but also on the assumptions in the model.
+From the perspective of extrapolating local data to global scales for policy advice and decision making [i.e.
+@hoekvandijke2022; @jackson2005], this is an important point.
+
+### Residuals of the model
+
+The residuals of the final model presented in this paper (Figure \@ref(fig:gamcheckmodelall)) indicate that the residual distribution remains fat-tailed, causing deviations from an assumed $\epsilon \sim N(0,\sigma^2)$.
+This once again highlights that there is unexplained variation at the extremes of the distribution, once again related to the paired watershed experiments (Figure \@ref(fig:assessment)).
+Generally,in statistical models, the approach would be to further normalise the residuals through transformations.
+However, in this case this might be difficult and might not resolve all the issues due to the large variation in the data.
+
+### Interactions
+
+The current modelling approach does not consider any interactions between the variables, and this would offer another approach to understand the variation in the data.
+As already indicated in Figure \@ref(fig:corgraphs), there are interactions between different variables.
+This further complicates the extrapolation of the local scale experiment data to global scales and to extend historical data to current management and decisions.
+
+In this case, interactions were not included because, as was shown, there are bigger problems with trying to extrapolate the existing data, and the data itself can be problematic.
+To be able to model the interactions well, the nature of the variables and interactions need to be understood and or clearly hypothesized.
+Otherwise it becomes another case of correlation without causation.
+
+### Implications for other "meta-analysis" studies
+
+There has been a recent push to develop more meta-analysis studies in hydrology [@wang2020; @evaristo2020metaanalysis], and we strongly believe that developing new insights by combining historical data sets from reviewed papers is highly valuable.
+However, this paper highlights that there is considerable chance that large historical data sets include latent variables and are more complex than envisioned.
+This is particularly true for more historical work, as methods of observation and even apporaches to management have changed considerably.
+The same management description is not necessarily the same action on the ground.
+A carefully designed and systematic approach can prevent some of bigger problems as is demonstrated in @wang2020, where both the approach and the catchment area are investigated as latent variables.
+This is particularly relevant, where the results of meta-analyses are extrapolated to make global predictions without clearly quantified uncertainties (such as in @hoekvandijke2022 and @wang2020).
+
+A second potential danger is the extrapolation of the local small catchment results and conclusions tolarger scales, but beyond the original  scope of the studies.
+For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
+It is tempting to use the result of a large scale analysis of this data to make inferences about overall landuse change [@li2017; @wang2020], but this would not be valid, as the deforestation studies are generally not a transition to an agricultural landuse or pasture, but regrowing into forest.
+Similarly, using the plantation studies to extrapolate to "reforestation" (as in @filoso2017 and @hoekvandijke2022) is also tenuous.
+Plantation forests are generally fast growing hybrids that will have quite different ecophysiology, particularly in South America [@jones2017; @binkley2017], while other reforestation, for example for salinity control in Australia, might focus on a mix of native species.
+Given the link between ecophysiology and water and carbon budgets, care should be taken in extrapolation, introducing a further error.
+
+A final factor is the effect of climate change [@vervoort2021] on runoff, even if the effects are still minor.
+Earlier papers [@li2017; @wang2020] have analysed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
+Given that the database of studies now captures almost 100 years of work, we cannot ignore a climate change trend that is potentially hidden in the data.
+A simple inclusion of the start date of the experiment (*From*) in the GAM model does suggests an increase in change in the percentage of flow over time.
+However, this is once again not easy to analyse.
+As the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
+
+```{r climate_change, echo = F, eval = F}
+
+model_CC <- gam(DeltaQf_perc ~ DeltaF_perc  + 
+                    s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
+                      s(log10(Area_km2), k = 10, bs = "ts") + s(From, bs = "ts") + 
+                  #s(From, bs = "ts") +
+                    Precip_data_type  + Forest_type + Assessment_technique +
+                    Hydrological_regime, data = All_data3 %>% filter(From > 1910))
+summary(model_CC)
+plot(model_CC)
+gam.check(model_CC)
+# png("DeltaQf_time.png", width = 2500, height = 1800, res = 300)
+# All_data2 %>%
+#   filter(is.na(Forest_Sign)==F) %>%
+#   ggplot(aes(From, DeltaQf_perc)) + 
+#   geom_point(colour = "blue") + facet_wrap(~Forest_Sign, ncol = 1) +
+#   theme_bw() +
+#     theme(axis.title = element_text(size = rel(2)),
+#         axis.text = element_text(size = rel(1.5)))
+# dev.off()
+
+```
+
+### Future research needs (implications for forest hydrology)
+
+This study also points to several further opportunities and future research needs.
+
+A major focus of many of the papers related to forest hydrology have been specifically interested in the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
+As the paper by @jones2017 highlights this means there are opportunities to analyse the time evolution of the catchment response to forestation.
+Given the large number of studies that look at a time evolution of forest cover (i.e.
+either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in landcover.
+This will give further insights into the equilibrium state under forestry.
+As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
+
+More generally there is a clear need for a more in depth analysis of the data base of studies used here.
+In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, streamflow responses and responses of components of streamflow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
+
+There is also a clear need to understand the impact of the assessment methods with respect to scale.
+Extrapolating paired watershed experiment results into models can possibly overlook landscape interactions that are visible at larger scales, but do not occur on smaller scales.
+For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on landuse.
+A carefully designed simulation study that specifically investigates the change in stream flow repsonse with scale using local field data for verification can help solve this problem.
+
+At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
+A better approach is analyzing streamflow data at multiple spatial scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes appears to be the best approach.
+The highlighted paper by @levy2018 is currently the best example of a solid statistical approach to analyzing streamflow responses.
+Simulation modelling can be an approach to analyze different scenarios, if there is clear recognition of the potential impact of the model structure (the algorithms and parameters that describe for example plantation tree growth) on the simulation outcomes.
+
+We envision that in the future more innovative approaches to analyzing data at different scales will be developed.
 
@@
 
-There are four major interlinked reasons why this is the case, and this has implications for meta-analyses in Environmental Science and Hydrology in general:
+While the analysis reveals similar conclusions in relation to the response of streamflow to forest cover, there are four major interlinked reasons why these results should be considered carefully.
+This subsequently has implications for meta-analyses in Environmental Science and Hydrology in general.
+The reasons higlighted in this paper are:
 
@@
 
-Any statistical analysis, including the one in this paper, needs to be considered "conditional on the data", and given the issues indicated, extrapolation of the results of summary studies into global hydrological models has to be done with great care.
-Better would be to explicitly include uncertainty in the extrapolation of the results.
+Any statistical analysis, including the one in this paper, needs to be considered "conditional on the data", and given the issues indicated, extrapolation of the results of summary studies to larger scales and into global hydrological models has to be done with great care.
+Better would be to analyze observed data and explicitly include uncertainty in the extrapolation of the results.
 
@@
 As the old adagium says: Correlation is not the same as causation.
-
-Future research should ...
-
- In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
 

### 2022-09-22 paper/Forest_and_Water/Forest_and_Water.tex
version for interbal review

@@
 
+\usepackage{setspace}\doublespacing
+
 \begin{document}
@@
 Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for the recent three papers were reviewed, combined and re-analysed highlight the following: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
-Generalised additive models were used to run flexible regressions including multiple variables.
-  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
-Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover.
-Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data.
-Smaller catchment dominate the database with 42\% of the data \textless{} 1 km\textsuperscript{2} and 65\% of the data \textless{} 10 km\textsuperscript{2}.
-Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
+The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling.
+The overall results of the modelling indicates similar results to previous papers.
+The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results.
+The area of the catchment is strongly related to the assessment technique and the variability in the response data.
+  A careful review of the data and the methodology in the previous papers highlights that there are four interlinked reasons that are problematic: 1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.
   \end{abstract}
     \begin{keyword}
-    keyword1 \sep 
-    keyword2
+    data analysis \sep forest cover change \sep global scales \sep 
+    literature review
   \end{keyword}
@@
 
-The impacts of global deforestation and reforestation are important through their influence on streamflow and both blue and green water availability \citep{hoekvandijke2022, schyns2019}.
+There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects \citep{wang2020, hoekvandijke2022}.
+More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability \citep{hoekvandijke2022, schyns2019}.
 The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
@@
 However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}.
+For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% changes in streamflow as a result of reforestation.
 Similarly a modelling study by \citet{beck2013} found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
@@
 
-The most recent reviews \citep{zhang2017, filoso2017} developed an impressive global database of catchment studies with changes in streamflow due to changes in forest cover.
+There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarise past studies.
+The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.
+
+From the most recent reviews on forest impact on streamflow \citep{zhang2017, filoso2017}, the study by \citet{filoso2017} is a clear meta-analysis, while the other is not.
+However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data.
 The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
@@
 
-There are some clear limitations in these studies.
+There are some clear limitations in these studies, and some of this applies to meta-analyses in general.
 The main method in the work by \citet{zhang2017} is a single covariate linear regression.
@@
 Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
-There is a real attraction in the idea of quantitative analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales.
-However, there is also a real danger in this process, which is what we will highlight in this paper. There are four potential errors (or limitations) in the mentioned global analyses:
+There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales \citep{evaristo2020metaanalysis}.
+However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:
 
@@
 \item
-  Latent variables that are not included in the typical single covariate analysis;
+  Impact of latent variables that are not included in the typical single covariate analysis;
 \item
@@
 \item
-  Aggregation of data that originates from different experiments with different objectives across a wide time period; and, finally\\
+  Aggregation of data that originates from different experiments with different objectives across a wide time period, but have similar keywords; and, finally\\
 \item
@@
 
-The aim of this paper is to highlight examples of each of these limitations, how they have influenced past work, and provide suggestions of how we can overcome these limitations.
+The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation.
+This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
 In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}@{}}
-\caption{Summary of abbreviations of factors used in the Zhang et al.~(2017) data set}\tabularnewline
+\caption{\label{tab:table1} Summary of abbreviations of factors used in the Zhang et al.~(2017) data set}\tabularnewline
 \toprule()
@@
 
-** DO WE NEED TO REDO THE MAP? Eliana please provide code **
-
 Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
@@
 \endhead
-\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
-\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
-\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
-\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
-\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
+\textbf{(Intercept)} & -4.81 & 16.31 & -0.29 & 0.77 \\
+\textbf{DeltaF\_perc} & -0.6 & 0.06 & -10.71 & 0 \\
+\textbf{Precip\_data\_typeOB} & -21.4 & 13.23 & -1.62 & 0.11 \\
+\textbf{Precip\_data\_typeSG} & 9.36 & 15.17 & 0.62 & 0.54 \\
+\textbf{Assessment\_techniqueEA, HM} & 20.64 & 42.73 & 0.48 & 0.63 \\
+\textbf{Assessment\_techniqueHM} & 22.81 & 11.71 & 1.95 & 0.05 \\
+\textbf{Assessment\_techniquePWE} & 30.63 & 11.94 & 2.57 & 0.01 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
-\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
+HM} & 17.42 & 43.26 & 0.4 & 0.69 \\
+\textbf{Assessment\_techniqueQPW} & 39.52 & 20.15 & 1.96 & 0.05 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
-\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
-\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
-\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
-\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
+EA} & 24.39 & 24.42 & 1 & 0.32 \\
+\textbf{Assessment\_techniqueSH} & 45.3 & 11.83 & 3.83 & 0 \\
+\textbf{Forest\_typeCF} & -9.45 & 7.6 & -1.24 & 0.21 \\
+\textbf{Forest\_typeMF} & -8.05 & 7.56 & -1.06 & 0.29 \\
+\textbf{Hydrological\_regimeSD} & 3.57 & 9.16 & 0.39 & 0.7 \\
 \bottomrule()
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.77 & 4 & 0.85 & 0.03 \\
-\textbf{s(Dryness)} & 4.71 & 9 & 2.18 & 0 \\
-\textbf{s(length)} & 4.44 & 34 & 0.24 & 0.09 \\
+\textbf{s(log10(Area\_km2))} & 0.79 & 4 & 0.99 & 0.02 \\
+\textbf{s(Dryness)} & 4.64 & 9 & 2.26 & 0 \\
+\textbf{s(length)} & 4.45 & 34 & 0.22 & 0.12 \\
 \bottomrule()
@@
 The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters).
-This indicates an \emph{r\textsuperscript{2}} of 0.46 and deviance explained is 0.5, suggesting the model only explains about 50\% of the variance in the data.
+This indicates an \emph{r\textsuperscript{2}} of 0.45 and deviance explained is 0.49, suggesting the model only explains about 50\% of the variance in the data.
 
@@
 
-The overall partial slope of the change in forest cover is -0.59, if all other variables are kept constant.
+The overall partial slope of the change in forest cover is -0.6, if all other variables are kept constant.
 This suggest quite strong change in streamflow, moving from fully forested to fully cleared.
-Over the whole forest cover range, this is a change of -118 mm, with other variables held constant.
+Over the whole forest cover range, this is a change of -120 mm, with other variables held constant.
 This change is highly significant, as indicated by the low p-value.
 
-In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.03)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.09)) explain variation in the data.
+In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.12)) explain variation in the data.
 For \emph{length}, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
@@
 \endhead
-linear across range & 0.5 & 3167 \\
+linear across range & 0.49 & 3182 \\
 different for forestation and
-deforestation & 0.46 & 3213 \\
-non-linear across the range & 0.5 & 3167 \\
+deforestation & 0.45 & 3227 \\
+non-linear across the range & 0.5 & 3182 \\
 \bottomrule()
@@
 The detail of the comparison is highlighted in Supplementary material part 3.
-However, generally the results of the analysis showed two main points (Table \ref{tab:tablecomparison}:
+However, generally the results of the analysis showed two main points (Table \ref{tab:tablecomparison}):
 
@@
 
-As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}.
+As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}).
 Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}).
@@
 \endhead
-\textbf{(Intercept)} & -10.28 & 17.6 & -0.58 & 0.56 \\
-\textbf{DeltaF\_perc} & -0.59 & 0.08 & -7.59 & 0 \\
-\textbf{Forest\_SignIncrease} & 0.93 & 9.58 & 0.1 & 0.92 \\
-\textbf{Precip\_data\_typeOB} & -12.54 & 12.2 & -1.03 & 0.3 \\
-\textbf{Precip\_data\_typeSG} & 5.9 & 15.06 & 0.39 & 0.7 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.86 & 39.9 & 0.47 & 0.64 \\
-\textbf{Assessment\_techniqueHM} & 29.54 & 11.08 & 2.67 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 24.56 & 12.29 & 2 & 0.05 \\
+\textbf{(Intercept)} & -10.17 & 17.98 & -0.57 & 0.57 \\
+\textbf{DeltaF\_perc} & -0.59 & 0.08 & -7.45 & 0 \\
+\textbf{Forest\_SignIncrease} & 0.41 & 9.79 & 0.04 & 0.97 \\
+\textbf{Precip\_data\_typeOB} & -15.97 & 12.5 & -1.28 & 0.2 \\
+\textbf{Precip\_data\_typeSG} & 15.71 & 14.85 & 1.06 & 0.29 \\
+\textbf{Assessment\_techniqueEA, HM} & 20.38 & 41.03 & 0.5 & 0.62 \\
+\textbf{Assessment\_techniqueHM} & 26.42 & 11.4 & 2.32 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 28.51 & 12.15 & 2.35 & 0.02 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 13.22 & 40.95 & 0.32 & 0.75 \\
-\textbf{Assessment\_techniqueQPW} & 44.21 & 19 & 2.33 & 0.02 \\
+HM} & 17.4 & 42.05 & 0.41 & 0.68 \\
+\textbf{Assessment\_techniqueQPW} & 41.49 & 19.53 & 2.12 & 0.03 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 25.54 & 22.81 & 1.12 & 0.26 \\
-\textbf{Assessment\_techniqueSH} & 40.89 & 11.16 & 3.67 & 0 \\
-\textbf{Forest\_typeCF} & -10.22 & 7.13 & -1.43 & 0.15 \\
-\textbf{Forest\_typeMF} & -3.9 & 7.14 & -0.55 & 0.59 \\
-\textbf{Hydrological\_regimeSD} & -0.02 & 8.65 & 0 & 1 \\
+EA} & 24.81 & 23.32 & 1.06 & 0.29 \\
+\textbf{Assessment\_techniqueSH} & 47.26 & 11.49 & 4.11 & 0 \\
+\textbf{Forest\_typeCF} & -9.47 & 7.3 & -1.3 & 0.2 \\
+\textbf{Forest\_typeMF} & -6.01 & 7.35 & -0.82 & 0.41 \\
+\textbf{Hydrological\_regimeSD} & 2.5 & 8.89 & 0.28 & 0.78 \\
 \bottomrule()
@@
 \endhead
-\textbf{s(Dryness)} & 4.07 & 9 & 2.13 & 0 \\
-\textbf{s(log10(Area\_km2))} & 1.57 & 4 & 1.86 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.86 \\
+\textbf{s(Dryness)} & 4.02 & 9 & 2.16 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.87 & 4 & 1.53 & 0.01 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.98 \\
 \bottomrule()
@@
 
-This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.46 and a deviance explained of 0.49.
+This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.44 and a deviance explained of 0.48.
 
@@
 One concern with the results presented so far is that there are a few assessment techniques in the data set with a very low number of observations and could influence the results of the analysis.
-This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess}.
+This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM) (Table \ref{tab:tableassess}).
 
@@
 \endhead
-\textbf{(Intercept)} & -15.02 & 14.46 & -1.04 & 0.3 \\
-\textbf{DeltaF\_perc} & -0.58 & 0.05 & -10.99 & 0 \\
-\textbf{Precip\_data\_typeOB} & -12.13 & 12.03 & -1.01 & 0.31 \\
-\textbf{Precip\_data\_typeSG} & 7.25 & 14.02 & 0.52 & 0.61 \\
-\textbf{Assessment\_techniqueHM} & 31.88 & 10.86 & 2.94 & 0 \\
-\textbf{Assessment\_techniquePWE} & 30.77 & 11.07 & 2.78 & 0.01 \\
-\textbf{Assessment\_techniqueQPW} & 45.78 & 18.37 & 2.49 & 0.01 \\
-\textbf{Assessment\_techniqueSH} & 42.88 & 11.01 & 3.9 & 0 \\
-\textbf{Forest\_typeCF} & -8.78 & 7.06 & -1.24 & 0.21 \\
-\textbf{Forest\_typeMF} & -2.17 & 7.22 & -0.3 & 0.76 \\
-\textbf{Hydrological\_regimeSD} & -1.29 & 8.62 & -0.15 & 0.88 \\
+\textbf{(Intercept)} & -12.17 & 14.9 & -0.82 & 0.41 \\
+\textbf{DeltaF\_perc} & -0.58 & 0.05 & -10.79 & 0 \\
+\textbf{Precip\_data\_typeOB} & -16.53 & 12.37 & -1.34 & 0.18 \\
+\textbf{Precip\_data\_typeSG} & 15.32 & 14.39 & 1.06 & 0.29 \\
+\textbf{Assessment\_techniqueHM} & 28.74 & 11.16 & 2.57 & 0.01 \\
+\textbf{Assessment\_techniquePWE} & 31.37 & 11.3 & 2.78 & 0.01 \\
+\textbf{Assessment\_techniqueQPW} & 44.15 & 18.89 & 2.34 & 0.02 \\
+\textbf{Assessment\_techniqueSH} & 49.42 & 11.32 & 4.37 & 0 \\
+\textbf{Forest\_typeCF} & -8.72 & 7.25 & -1.2 & 0.23 \\
+\textbf{Forest\_typeMF} & -3.94 & 7.43 & -0.53 & 0.6 \\
+\textbf{Hydrological\_regimeSD} & 2.01 & 8.86 & 0.23 & 0.82 \\
 \bottomrule()
@@
 \endhead
-\textbf{s(Dryness)} & 3.44 & 9 & 2.71 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.86 & 9 & 0.61 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.87 \\
+\textbf{s(Dryness)} & 3.37 & 9 & 2.79 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.85 & 9 & 0.68 & 0.01 \\
+\textbf{s(length)} & 0 & 9 & 0 & 0.84 \\
 \bottomrule()
@@
 As \citet{jones2017} indicate, this really needs time series analysis of the different experiments.
-Some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+However, some of the time series data might not be recoverable from the older experiments, which will limit the opportunities for analysis.
+We will discuss this further below.
 
@@
 
-\textbf{Maybe include a figure that shows the variation in the flow data by Assessment technique}
-
 \hypertarget{general-discussion}{%
@@
 
-\textbf{Still to do}
+In this paper, a few studies have been singled out in the analysis.
+The choice of focus was mainly driven by the data that was made available by the authors of these papers \citep{zhang2017, filoso2017}, which provide a rich case study for the current paper.
+
+Field research is by nature limited in space and time, due to the high costs involved of setting up experiments.
+This is particularly true for experiments in hydrology and forest hydrology, where field sites need to cover sufficient spatial and temporal variability.
+This means there is a general need to extrapolate the local results to larger scales to inform decision making and policy.
+
+However, as is demonstrated in this paper, there are multiple issues when this local scale data is extrapolated to larger scales.
+It clearly demonstrates that the results of any model (in this case a regression model) is highly dependent on the data, but also on the assumptions in the model.
+From the perspective of extrapolating local data to global scales for policy advice and decision making \citep[i.e.][]{hoekvandijke2022, jackson2005}, this is an important point.
+
+\hypertarget{residuals-of-the-model}{%
+\subsubsection{Residuals of the model}\label{residuals-of-the-model}}
+
+The residuals of the final model presented in this paper (Figure \ref{fig:gamcheckmodelall}) indicate that the residual distribution remains fat-tailed, causing deviations from an assumed \(\epsilon \sim N(0,\sigma^2)\).
+This once again highlights that there is unexplained variation at the extremes of the distribution, once again related to the paired watershed experiments (Figure \ref{fig:assessment}).
+Generally,in statistical models, the approach would be to further normalise the residuals through transformations.
+However, in this case this might be difficult and might not resolve all the issues due to the large variation in the data.
+
+\hypertarget{interactions}{%
+\subsubsection{Interactions}\label{interactions}}
+
+The current modelling approach does not consider any interactions between the variables, and this would offer another approach to understand the variation in the data.
+As already indicated in Figure \ref{fig:corgraphs}, there are interactions between different variables.
+This further complicates the extrapolation of the local scale experiment data to global scales and to extend historical data to current management and decisions.
+
+In this case, interactions were not included because, as was shown, there are bigger problems with trying to extrapolate the existing data, and the data itself can be problematic.
+To be able to model the interactions well, the nature of the variables and interactions need to be understood and or clearly hypothesized.
+Otherwise it becomes another case of correlation without causation.
+
+\hypertarget{implications-for-other-meta-analysis-studies}{%
+\subsubsection{Implications for other ``meta-analysis'' studies}\label{implications-for-other-meta-analysis-studies}}
+
+There has been a recent push to develop more meta-analysis studies in hydrology \citep{wang2020, evaristo2020metaanalysis}, and we strongly believe that developing new insights by combining historical data sets from reviewed papers is highly valuable.
+However, this paper highlights that there is considerable chance that large historical data sets include latent variables and are more complex than envisioned.
+This is particularly true for more historical work, as methods of observation and even apporaches to management have changed considerably.
+The same management description is not necessarily the same action on the ground.
+A carefully designed and systematic approach can prevent some of bigger problems as is demonstrated in \citet{wang2020}, where both the approach and the catchment area are investigated as latent variables.
+This is particularly relevant, where the results of meta-analyses are extrapolated to make global predictions without clearly quantified uncertainties (such as in \citet{hoekvandijke2022} and \citet{wang2020}).
+
+A second potential danger is the extrapolation of the local small catchment results and conclusions tolarger scales, but beyond the original scope of the studies.
+For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
+It is tempting to use the result of a large scale analysis of this data to make inferences about overall landuse change \citep{li2017, wang2020}, but this would not be valid, as the deforestation studies are generally not a transition to an agricultural landuse or pasture, but regrowing into forest.
+Similarly, using the plantation studies to extrapolate to ``reforestation'' (as in \citet{filoso2017} and \citet{hoekvandijke2022}) is also tenuous.
+Plantation forests are generally fast growing hybrids that will have quite different ecophysiology, particularly in South America \citep{jones2017, binkley2017}, while other reforestation, for example for salinity control in Australia, might focus on a mix of native species.
+Given the link between ecophysiology and water and carbon budgets, care should be taken in extrapolation, introducing a further error.
+
+A final factor is the effect of climate change \citep{vervoort2021} on runoff, even if the effects are still minor.
+Earlier papers \citep{li2017, wang2020} have analysed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
+Given that the database of studies now captures almost 100 years of work, we cannot ignore a climate change trend that is potentially hidden in the data.
+A simple inclusion of the start date of the experiment (\emph{From}) in the GAM model does suggests an increase in change in the percentage of flow over time.
+However, this is once again not easy to analyse.
+As the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
+
+\hypertarget{future-research-needs-implications-for-forest-hydrology}{%
+\subsubsection{Future research needs (implications for forest hydrology)}\label{future-research-needs-implications-for-forest-hydrology}}
+
+This study also points to several further opportunities and future research needs.
+
+A major focus of many of the papers related to forest hydrology have been specifically interested in the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
+As the paper by \citet{jones2017} highlights this means there are opportunities to analyse the time evolution of the catchment response to forestation.
+Given the large number of studies that look at a time evolution of forest cover (i.e.~either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in landcover.
+This will give further insights into the equilibrium state under forestry.
+As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
+
+More generally there is a clear need for a more in depth analysis of the data base of studies used here.
+In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, streamflow responses and responses of components of streamflow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
+
+There is also a clear need to understand the impact of the assessment methods with respect to scale.
+Extrapolating paired watershed experiment results into models can possibly overlook landscape interactions that are visible at larger scales, but do not occur on smaller scales.
+For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on landuse.
+A carefully designed simulation study that specifically investigates the change in stream flow repsonse with scale using local field data for verification can help solve this problem.
+
+At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
+A better approach is analyzing streamflow data at multiple spatial scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes appears to be the best approach.
+The highlighted paper by \citet{levy2018} is currently the best example of a solid statistical approach to analyzing streamflow responses.
+Simulation modelling can be an approach to analyze different scenarios, if there is clear recognition of the potential impact of the model structure (the algorithms and parameters that describe for example plantation tree growth) on the simulation outcomes.
+
+We envision that in the future more innovative approaches to analyzing data at different scales will be developed.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+This study demonstrates that analyses of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+
+While the analysis reveals similar conclusions in relation to the response of streamflow to forest cover, there are four major interlinked reasons why these results should be considered carefully.
+This subsequently has implications for meta-analyses in Environmental Science and Hydrology in general.
+The reasons higlighted in this paper are:
 
@@
 \item
-  studies are conditional on the data
-\item
-  residuals: should the data be further transformed\\
-\item
-  no interactions\\
-\item
-  implications for other ``meta-analysis'' studies\\
-\item
-  future research needs
-\end{itemize}
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-This study demonstrates that analyses of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
-
-There are four major interlinked reasons why this is the case, and this has implications for meta-analyses in Environmental Science and Hydrology in general:
-
-\begin{itemize}
-\tightlist
-\item
   The existence of latent variables in the data that create the appearance of a relationship that really does not exist;\\
@@
 
-Any statistical analysis, including the one in this paper, needs to be considered ``conditional on the data'', and given the issues indicated, extrapolation of the results of summary studies into global hydrological models has to be done with great care.
-Better would be to explicitly include uncertainty in the extrapolation of the results.
+Any statistical analysis, including the one in this paper, needs to be considered ``conditional on the data'', and given the issues indicated, extrapolation of the results of summary studies to larger scales and into global hydrological models has to be done with great care.
+Better would be to analyze observed data and explicitly include uncertainty in the extrapolation of the results.
 
@@
 
-Future research should \ldots{}
-
-In addition, a more detailed analysis of the historical studies, in particular focussing on differences in flow components can further clarify some of the uncertainties highlighted here.
-
 \hypertarget{acknowledgements}{%

### 2022-09-22 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
version for interbal review 2

@@
 header-includes:
-  - \usepackage{setspace}
+  - \usepackage{setspace}\doublespacing
   - \usepackage{color}
@@
 The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.
-\
-**The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.**
 
@@
 As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}
-\
-**check this**
+
 We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
@@
 However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
-\
+
 \textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
-\
+
 Please see our comments in reply to the AE below.
-\  
+
 \textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
@@
 At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
-\
-Please see our comments in reply to the AE below.  \
-\
+
+Please see our comments in reply to the AE below.
 
@@
 Levy et al 2018).
-So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}    \
-\
+So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}
 
@@
 This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
-Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as a "verfication" of our analysis.
+Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as an example of how the analysis can be done well.
 
@@
 However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017).
-\
-\
+This is now extensively discussed in the discussion of the paper.
+
 As a result of this, we have rewritten the scope of the paper and changed the title to better reflect the main findings and message coming from this paper.
-\
-\
+
 The new title of the paper is:
+
 **Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.**
-\
+
 The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
@@
 This means the change should be recorded as 0.
-\
 
@@
 If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}
-\
+
 We acknowledge this, and we hope that the current revised version and our responses address these concerns.
-\
+
 \textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.
 Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}
-\
+
 We acknowledge this, and we hope that the current revised version and our responses address these concerns.
-\
 
@@
 The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}
-\
+
 <!-- answer -->
@@
 It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}
-\
+
 Our answer to this point from the associate editor is quite long and covers the following sub topics:
@@
 
-\
-**Include a section in the paper to explain this clearly**  
-\
-
 Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript.
@@
 
+We believe that our current discussion of the diagnostic plots in the paper clarifies this issue and also highlights the remaining non-normality in the residuals.
+
 ### The issue of outliers
@@
 This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors." As a result, we believe that excluding values that are outliers is probably not a good idea.
-\
+
 However, another careful review of the data identified many further errors in the data, which were all originally in @zhang2017.
@@
 corrected to 16\%
-- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
-- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in @zhang2017.
-This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates "clearfelling".
+- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected.  
+- 67, April rd, which is incorrectly attributed to @ruprecht1991 in @zhang2017.
+This is actually from @ruprecht1989 and the original paper clearly indicates "clearfelling".
 As a result the change in forest cover was changed to -100\% rather than +100\%.
-- 210, March rd, 100, 147.6. Same problem as 67, Bari et al. (1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.  
+- 210, March rd, 100, 147.6. Same problem as 67, @bari1996 clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.  
 - 213, 214 and 215, Monda 1, 2 and 3.
@@
 The original paper [@oshaughnessy1979] only reports on the control period and indicates that the catchments will be cleared.
-The later summary paper [@watson2001] shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
+The later summary paper [@watson2001] shows the time series of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
 The further complication is that the treatment included clearing and reseeding and regrowth.
@@
 - 101 Coalburn.
-Original publication (Robinson, 1993) which is a symposium paper, is not available.
+Original publication (Robinson, 1993) which is a symposium paper, is not available, even after contacting the original authors.
 The best summary of the research is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment.
@@
 Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}
-\
+
 <!-- answer -->
@@
 
-However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach.
-**FIX THIS UP IN THE PAPER and insert the relevant page and line numbers.**
+However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach (line 220 - 224 on page 11)
 
@@
 We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.
-\
 
@@
 Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}
-\
+
 <!-- answer -->
@@
 We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.
-\
 
@@
 These are not big issues per se but are nonetheless distracting.}
-\
+
 <!-- answer -->

### 2022-09-22 paper/Forest_and_Water/Reply_first_review_JoH.tex
version for interbal review 2

@@
 
-\usepackage{setspace}
+\usepackage{setspace}\doublespacing
 \usepackage{color}
@@
   \title{Replies to editor and associate editor Journal of Hydrology}
-    \author[]{R. Willem Vervoort%
-  %
-  }
+    \author[]{R. Willem Vervoort}
 
-    \author[]{Eliana Nervi%
-  %
-  }
+    \author[]{Eliana Nervi}
 
-    \author[]{Jimena Alonso%
-  %
-  }
+    \author[]{Jimena Alonso}
 
@@
 This is followed by the comments from the associate editor.
-The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.\\
-\strut \\
-\textbf{The most important point from the reviews is the fact that we have not communicated the need for this paper. We need to rewrite the introduction to do this better.}
+The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.
 
@@
 \textcolor{blue}{The first is to streamline the statistics.
-As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}\\
-\strut \\
-\textbf{check this}
+As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}
+
 We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
 If the statistical modelling aimed at developing the best predictive model, then this would be the right approach.
-However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.\\
-\strut \\
-\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}\\
-\strut \\
-Please see our comments in reply to the AE below.\\
-\hspace*{0.333em}
+However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
+
+\textcolor{blue}{In doing so, please carefully note the AEs' concerns about some of the statistical methodology - concerns which range from interpretation to the use of appropriate performance metrics across models of varying structure, to the selection of the appropriate metrics of forest cover change for analysis.}
+
+Please see our comments in reply to the AE below.
+
 \textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.
 Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.
-At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}\\
-\strut \\
-Please see our comments in reply to the AE below.\\
-\strut \\
+At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}
+
+Please see our comments in reply to the AE below.
 
@@
 Levy et al 2018).
-So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}\\
-\strut \\
+So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}
 
@@
 This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
-Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use \citet{levy2018} as a ``verfication'' of our analysis.
+Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use \citet{levy2018} as an example of how the analysis can be done well.
 
@@
 We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
-However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).\\
-\strut \\
-\strut \\
-As a result of this, we have rewritten the scope of the paper and changed the title to better reflect the main findings and message coming from this paper.\\
-\strut \\
-\strut \\
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).
+This is now extensively discussed in the discussion of the paper.
+
+As a result of this, we have rewritten the scope of the paper and changed the title to better reflect the main findings and message coming from this paper.
+
 The new title of the paper is:
-\textbf{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}\\
-\strut \\
+
+\textbf{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}
+
 The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
@@
 However, as we argue, the results of this paper are misrepresented in the database, as none of the analysed catchments had a significant change in the streamflow.
-This means the change should be recorded as 0.\\
-  \strut \\
+This means the change should be recorded as 0.
 \end{itemize}
@@
 For this reason, we're rejecting the MS at present.
-If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}\\
-We acknowledge this, and we hope that the current revised version and our responses address these concerns.\\
-\strut \\
+If the authors are able to address the 3 issues above in a substantial revision of the MS, we would be pleased to look at it again.}
+
+We acknowledge this, and we hope that the current revised version and our responses address these concerns.
+
 \textcolor{blue}{We would, however, consider as a new submission for review a substantially revised version of this paper that addresses all of the reviewers' comments.
-Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}\\
-We acknowledge this, and we hope that the current revised version and our responses address these concerns.\\
+Should you choose to submit such a revised manuscript please refer to the present manuscript number, provide a detailed point-by-point reply to all of the reviewers' comments, and state how the revised manuscript addresses these.}
+
+We acknowledge this, and we hope that the current revised version and our responses address these concerns.
 
@@
 \textcolor{blue}{The manuscript considers an enhanced dataset of streamflow and forest cover, to explore how deforestation/afforestation alters catchment water yields.
-The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}\\
-\strut \\
+The manuscript is potentially of interest of the JoH readership, but it is not ready for review in present form.}
+
 Thank you, we acknowledge this and we hope that our current revision and answers to your comments have improved the manuscript sufficiently to go out for further review.
@@
 It is also good practice to check the relevance of outliers (of data with high VIF) and set them aside before model fitting.
-It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}\\
-\strut \\
+It is also not correct to comment on models as if working better or worse in certain ranges, based on the residuals (P 29), because the residuals are the results of the data and fitted model, and the fitted model depends on all datapoints.}
+
 Our answer to this point from the associate editor is quite long and covers the following sub topics:
@@
 
-\hfill\break
-\textbf{Include a section in the paper to explain this clearly}\\
-\strut \\
-
 Either way, understanding the diagnostic plots and the residual distribution is important, which is why this was included in the manuscript.
@@
 
+We believe that our current discussion of the diagnostic plots in the paper clarifies this issue and also highlights the remaining non-normality in the residuals.
+
 \hypertarget{the-issue-of-outliers}{%
@@
 ``Outliers are sample values that cause surprise in relation to the majority of the sample.
-This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors.'' As a result, we believe that excluding values that are outliers is probably not a good idea.\\
-\strut \\
+This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors.'' As a result, we believe that excluding values that are outliers is probably not a good idea.
+
 However, another careful review of the data identified many further errors in the data, which were all originally in \citet{zhang2017}.
@@
 Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
-corrected\\
+corrected.\\
 \item
-  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in \citet{zhang2017}.
-This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''.
+  67, April rd, which is incorrectly attributed to \citet{ruprecht1991} in \citet{zhang2017}.
+This is actually from \citet{ruprecht1989} and the original paper clearly indicates ``clearfelling''.
 As a result the change in forest cover was changed to -100\% rather than +100\%.\\
 \item
-  210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
+  210, March rd, 100, 147.6. Same problem as 67, \citet{bari1996} clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
 \item
@@
 The original paper \citep{oshaughnessy1979} only reports on the control period and indicates that the catchments will be cleared.
-The later summary paper \citep{watson2001} shows the timeseries of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
+The later summary paper \citep{watson2001} shows the time series of the flow change, but does not report a single value, so the values in the database must have been estimated from the timeseries.
 The further complication is that the treatment included clearing and reseeding and regrowth.
@@
   101 Coalburn.
-Original publication (Robinson, 1993) which is a symposium paper, is not available.
+Original publication (Robinson, 1993) which is a symposium paper, is not available, even after contacting the original authors.
 The best summary of the research is in \citet{birkinshaw2014} which summaries 45 years of resaerch in the Coalburn catchment.
@@
 A full blown model selection would also allow to retain or discard the interaction terms, which could be important (as also recognized by the authors; Section 4.5) and should not be discarded a priori.
-Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}\\
-\strut \\
+Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}
+
 The Associate editor raises several points that need further discussion.
@@
 
-However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach.
-\textbf{FIX THIS UP IN THE PAPER and insert the relevant page and line numbers.}
+However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach (line 220 - 224 on page 11)
 
@@
 While there clearly is cross correlation between the variables and there is potential interaction between terms, the question of clear causality remains unanswered.
-We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.\\
+We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.
 
@@
 As it looks now, the manuscript could be easily considered somewhat confirmatory, with respect to most data, approaches and conclusions reached by Zhang et al 2017 and Filoso et al 2017.
-Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}\\
-\strut \\
+Furthermore, the introduction needs to be rearranged, starting with a clear statement of the problem, what we know about that based on previous results, what is missing/how these previous analyses can be improved, and, stemming from these knowledge gaps and/or our understanding of the mechanisms, the questions addressed in the work or the hypotheses tested.}
+
 We agree that this is a valid point and a weakness in the original manuscript.
@@
 In particular the paper now more directly focuses on the difficulties in analysing aggregated global data and the importance of latent variables.
-We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.\\
-\strut \\
+We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.
 
@@
 Also: how is Table 1 used?
-These are not big issues per se but are nonetheless distracting.}\\
+These are not big issues per se but are nonetheless distracting.}
+
 Thank you for pointing this out, we have reviewed the paper carefully and corrected these minor issues.

### 2022-09-22 paper/Forest_and_Water/Forest_and_Water.Rmd
version for interbal review 2

@@
 
+In the statistical analysis we are not necessarily seeking the best "predicitive" model, and as such do not perform a traditional variable selection process.
+Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.
+
 To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
@@
 
-As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (This analysis is in supplementary material part 2).
+As an initial approach we tested whether the additional catchments added to the original data from @zhang2017 did not majorly influenced the results (This analysis is in supplementary material part 2).
 This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
-
-To make all the data and code publicly available, all the final data and analysis for this paper are located on github:  
+However, this means that the results of the studies are still comparable.
+
+To make all the data and code used for the analysis publicly available, all the final data and analysis for this paper are located on github:  
 [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.

### 2022-09-22 paper/Forest_and_Water/Forest_and_Water.tex
version for interbal review 2

@@
 
+In the statistical analysis we are not necessarily seeking the best ``predicitive'' model, and as such do not perform a traditional variable selection process.
+Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.
+
 To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) \citep{wood2006}.
@@
 
-As an initial approach we only used the data from \citet{zhang2017} to make sure that the additional catchments added to the data set did not influence the results (This analysis is in supplementary material part 2).
+As an initial approach we tested whether the additional catchments added to the original data from \citet{zhang2017} did not majorly influenced the results (This analysis is in supplementary material part 2).
 This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
-
-To make all the data and code publicly available, all the final data and analysis for this paper are located on github:\\
+However, this means that the results of the studies are still comparable.
+
+To make all the data and code used for the analysis publicly available, all the final data and analysis for this paper are located on github:\\
 \href{https://github.com/WillemVervoort/Forest_and_water/tree/publish}{https://github.com/WillemVervoort/Forest\_and\_water} on the ``publish'' branch.

### 2022-09-28 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
revision of draft for internal review

@@
   - name: R. Willem Vervoort
+    correspondingauthor: true
   - name: Eliana Nervi
@@
 
-The reviewer comments are ordered with the Editor in Chief comments first and our responses.
-This is followed by the comments from the associate editor.
-The reviewer comments in \textcolor{blue}{blue} and our responses in normal text.
+The reviewer comments are ordered with the Editor in Chief comments listed first and our responses next.
+This is followed by the comments from the associate editor and our responsed.
+The reviewer comments are in \textcolor{blue}{blue} and our responses in normal text.
 
@@
 We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
-If the statistical modelling aimed at developing the best predictive model, then this would be the right approach.
+If the statistical modelling was aimed at developing the best predictive model, then this would be the right approach.
 However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.
@@
 
-Thank you for the suggestion of the @levy2018 paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change.
-This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
+Thank you for the suggestion of the @levy2018 paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic land use change.
+This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
 Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as an example of how the analysis can be done well.
@@
 We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
-However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017).
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalization of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017).
 This is now extensively discussed in the discussion of the paper.
@@
 - 76, Beaver Creek, the flow was corrected from 600% to 157% after review of the original publication [@baker1984].
-- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.
+- 124, D3, @amatya2008effects: The originally recorded 250\% change by @zhang2017 is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
@@
 corrected to 16\%
-- 3, Baker Creek, Zhang and Wei, 2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected.  
+- 3, Baker Creek, @zhangwei2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected.
 - 67, April rd, which is incorrectly attributed to @ruprecht1991 in @zhang2017.
@@
 - a formalised model selection;
-- the use of AIC rather than r^2^ for model selection;  
+- the use of AIC rather than *r*^2^ for model selection; and
 - interaction terms.
@@
 
-### AIC rather than r^2^
+### AIC rather than *r*^2^
 
-We actually used the adjusted r^2^ in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
+We actually used the adjusted *r*^2^ in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
 However, as the GAM models also provide an AIC, we have for consistency changed our model performance measure to the AIC.

### 2022-09-28 paper/Forest_and_Water/Reply_first_review_JoH.tex
revision of draft for internal review

@@
 
-Thank you for the suggestion of the \citet{levy2018} paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic landuse change.
-This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the landuse data.
+Thank you for the suggestion of the \citet{levy2018} paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic land use change.
+This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
 Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use \citet{levy2018} as an example of how the analysis can be done well.
@@
 We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction.
-However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalisation of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).
+However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalization of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).
 This is now extensively discussed in the discussion of the paper.
@@
 \item
-  Error distributions\\
+  Error distributions
 \item
-  The issue of outliers\\
+  The issue of outliers
 \item
@@
 \item
-  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong.
+  124, D3, \citet{amatya2008effects}: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
@@
 \item
-  3, Baker Creek, Zhang and Wei, 2012.
+  3, Baker Creek, \citet{zhangwei2012}.
 The original recorde 201.1\% change by \citet{zhang2017} is also wrong.
 Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
-corrected.\\
+corrected.
 \item
@@
 This is actually from \citet{ruprecht1989} and the original paper clearly indicates ``clearfelling''.
-As a result the change in forest cover was changed to -100\% rather than +100\%.\\
+As a result the change in forest cover was changed to -100\% rather than +100\%.
 \item
-  210, March rd, 100, 147.6. Same problem as 67, \citet{bari1996} clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
+  210, March rd, 100, 147.6. Same problem as 67, \citet{bari1996} clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
 \item
@@
 This suggest that the records should be removed from the database, or only the first few years of the experiment used.
-In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).\\
+In any case, if the values are kept, the sign of the change in forest cover needs to changed to negative (Clearing).
 \item
-  230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.\\
+  230, Oleolega catchment. The paper describes a removal of forest up to 85\%. changed Delta\_F\_perc to -85 from 90.
 \item
@@
 In this report, in Table 2 on page 11, for the catchment a ``Crown cover'' decrease of 60\% is given.
-Changed the sign of the change in forest cover: -60\%.\\
+Changed the sign of the change in forest cover: -60\%.
 \item
@@
 In any case, the reported change in forest cover should be negative (clearing) rather than positive.
-Corrected for all three catchments.\\
+Corrected for all three catchments.
 \item
@@
 Corrected.
-Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.\\
+Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.
 \item
@@
 This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
-In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.\\
+In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.
 \item
@@
 Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases.
-Corrected Delta\_Qf\_perc to -20.7\%.\\
+Corrected Delta\_Qf\_perc to -20.7\%.
 \item
-  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
+  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.
 \item
@@
 This study is a very detailed hydrological modelling study, but in the end finds no significant change in streamflow as a result of deforestation.
-Values for all 12 studied catchments set to 0 in the database.\\
+Values for all 12 studied catchments set to 0 in the database.
 \item
@@
 \item
-  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc\_pos) indicating that in the data larger catchment tended to have (obviously) smaller areas of forest change.\\
+  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc\_pos) indicating that in the data larger catchment tended to have (obviously) smaller areas of forest change.
 \item
   the weak positive relationship between log10(Area) and the assessment method using hydrological models.
-This is also obvious as it would be impossible to perform paired catchment studies at very large scales.\\
+This is also obvious as it would be impossible to perform paired catchment studies at very large scales.
 \item
@@
 \item
-  a formalised model selection;\\
+  a formalised model selection;
 \item
-  the use of AIC rather than r\textsuperscript{2} for model selection;\\
+  the use of AIC rather than \emph{r}\textsuperscript{2} for model selection; and
 \item
@@
 \hypertarget{aic-rather-than-r2}{%
-\subsubsection{\texorpdfstring{AIC rather than r\textsuperscript{2}}{AIC rather than r2}}\label{aic-rather-than-r2}}
+\subsubsection{\texorpdfstring{AIC rather than \emph{r}\textsuperscript{2}}{AIC rather than r2}}\label{aic-rather-than-r2}}
 
-We actually used the adjusted r\textsuperscript{2} in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
+We actually used the adjusted \emph{r}\textsuperscript{2} in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does.
 However, as the GAM models also provide an AIC, we have for consistency changed our model performance measure to the AIC.

### 2022-09-28 paper/Forest_and_Water/Forest_and_Water.Rmd
revision of draft for internal review

@@
   - code: a
-    address: The University of Sydney, Sydney, NSW 2006, Australia
+    address: Sydney Institute of  Agriculture and School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
   - code: b
-    address: ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of  Agriculture, School of Life and Environmental Sciences.
+    address: ARC ITTC in Data Analytics for Resources and Environments, The University of Sydney, NSW 2006, Sydney, Australia
   - code: c
@@
 abstract: |
-  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+  Three recent papers review and analyze large global datasets related to impacts of forest cover on streamflow.
 Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
-The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling.
-The overall results of the modelling indicates similar results to previous papers.
-The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results.
-The area of the catchment is strongly related to the assessment technique and the variability in the response data.
-  A careful review of the data and the methodology in the previous papers highlights that there are  four interlinked reasons that are problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
-Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.
+However, the results are problematic, the underlying data set is unbalanced, and there are correlations in the data that warrant further investigation as this would influence the results.
+For example, the area of the catchment is strongly related to the assessment technique and the variability in the response data.
+For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies.
+Subsequently, the data were re-analyzed using generalised additive modelling.
+  The results highlight that there are  four interlinked reasons that make the general outcomes from the previous papers problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalised and extrapolated to continental and global scales.
 keywords: 
@@
 The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
+The most dramatic example of this is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
 However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% changes in streamflow as a result of reforestation.
-Similarly a modelling study by @beck2013 found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
+For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% change in streamflow as a result of reforestation.
+Similarly a modelling study by @beck2013 found no significant change in streamflow in 12 catchments in Puerto Rico as a result of deforestation.
 In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
-This suggests that there can be significant variation across the different studies,methodologies and geographical regions.
+This suggests that there can be significant variation across the different studies, methodologies and geographical regions.
 
@@
 
-As mentioned, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
+There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies.
+The suggestion is that, because meta-analyses use clearly defined search terms and statistical methods to analyze the results, this will lead to more reliable summaries of past research.
+As a result, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].
 These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships.
-Others have used the understanding from these studies to extrapolate to global scales [@hoekvandijke2022].
-
-There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies.
-The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.
-
-From the most recent reviews on forest impact on streamflow [@zhang2017; @filoso2017], the study by @filoso2017 is a clear meta-analysis, while the other is not.
-However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data.
+Others have used the understanding from a global analysis to extrapolate to global scales [@hoekvandijke2022].
+
+The recent paper by @filoso2017 is a clear meta-analysis, but most others [@zhang2017; @hoekvandijke2022; @zhou2015] are not.
+However,  n impressive global database of catchment studies with changes in streamflow due to changes in forest cover has been developed [@zhang2017; @filoso2017] and statistical approaches are used to analyze the resulting data.
 The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
-In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+In this case the collected data is mostly coded as count data and only a subset of 37 studies was analyzed for actual water yield change.
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
@@
 The conclusions of the first mentioned major review paper [@zhang2017] indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^.
-While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].
+While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with increases in forest cover.
+The main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].
 
@@
 
-A final earlier summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017.
-However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
+An earlier paper, that includes much of the same data as @zhang2017 and @filoso2017, is @zhou2015, which has one author in common with @zhang2017.
+However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model, which is similar to well-known Budyko approaches [@zhang2018elasticity].
 In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model.
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
-The Fuh model and variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2022 to interpret the global impact of reforestation.
-
-There are some clear limitations in these studies, and some of this applies to meta-analyses in general.
+The Fuh model and the related variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2022 to interpret the global impact of reforestation.
+
+However, concerning is that there are some clear limitations in these studies, and some of this applies to meta-analyses in general.
 The main method in the work by @zhang2017 is a single covariate linear regression.
-In contrast, the systematic review from @filoso2017 emphasises the classification and distributions of the study.
+In contrast, the systematic review from @filoso2017 mainly emphasises the classification and distributions of the study.
 @zhang2017 points out that a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
@@
 
-In contrast to the single covariate linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+In contrast to the single covariate linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides an example of an carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
 The analysis specifically accounted for differences between catchments and differences due to variations in climate.
-Their conclusion highlighted that in particular dry season streamflow was affected by deforestation.
-
-Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
-There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales [@evaristo2020metaanalysis].
-However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:
+Not all datasets are however suitable for this kind of in-depth analysis.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another review paper on this topic?
+There is a real attraction in the concept of statistical analysis of past studies encapsulated in meta-analysis to be able to extrapolate findings to larger scales, and to identify factors across global scales [@evaristo2020metaanalysis].
+However, there are also some hidden complications in this that can invalidate results, which this paper aims to highlight.
+There are four potential errors (or limitations) in such global meta-analyses:
 
@@
 
-The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation.
+The aim of this paper is to first reanalyze the global dataset [@zhang2017; @filoso2017] using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
 This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
-In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
-Finally, we will highlight future research needs in this area.
+In addition, by applying more complex statistical models, we will highlight the conclusions that can be drawn from the data.
+Finally, we will highlight future research needs in the area of forest cover change impact on streamflow.
 
@@
 As indicated, the starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set (are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+The change in streamflow ($\Delta Qf\%$) is based on equation 1 in @zhang2017.
+
 Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \@ref(tab:table1). These abbreviations will later be used in the models.
@@
 
-The paper by @zhang2017 use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and use the dryness index to identify the climate type.
-However, the potential or reference ET was not originally included as part of the published data set.
+The paper by @zhang2017 also uses the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and have used the dryness index to identify the climate type.
+However, the potential or reference ET used for this calculation was originally not included in the published data set.
+We will discus below how we derived the dryness index in our data set.
 We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.
@@
 As highlighted, @zhang2017 did not provide values for evapotranspiration in the data base.
-Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
+Using the location information, reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers.
 For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
 
-Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+Similar to @zhang2017, the Dryness index was calculated from the catchment estimate of reference evapotranspiration and the catchment estimate of annual average rainfall (Pa) as:
 
 \begin{equation}
-D = \frac{E_{0}}{Pa} (\#eq:eq1)
+Dryness = \frac{E_{0}}{Pa} (\#eq:eq1)
 \end{equation}
@@
 
-The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change flow as a consequence of changes in forest cover.
+The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change in flow as a result of changes in forest cover.
 This first aim is similar to @zhang2017, but the main difference is that we start off with all variables in the data set in the model.
@@
 
-In the statistical analysis we are not necessarily seeking the best "predicitive" model, and as such do not perform a traditional variable selection process.
+In the statistical analysis we are not necessarily seeking the best "predictive" model, and as such do not perform a traditional variable selection process.
 Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.
@@
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
+In @zhang2017, these changes were jointly analyzed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
 
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-The question becomes how best to analyse this.
+The question becomes how best to analyze this.
 One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
@@
 The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form.
-Both these approaches are tested in the results.
-
-The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
+All three approaches are tested in this study.
+
+The overarching test focuses on identifying the change streamflow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
 In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref(tab:table1)
@@
 In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
-The current number of catchments is the result of the removal of duplicates and our modifications and additions.
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*.
-The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
+The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails in the change in flow distribution.
 Note also the large number of studies with 100\% forest cover reduction.
 Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).
-This high skew in some of the data can create difficulties in the statistical modelling and further transformation of the data might be required.
+This high skew in some of the data can create difficulties in the statistical modelling and this will be discussed later.
 
@@
         axis.text = element_text(size = rel(1.2)),
-        strip.text = element_text(size = rel(1.5)),
+        strip.text = element_text(size = rel(1.2)),
         axis.text.x = element_text(angle = 45, hjust = 1))
@@
 All_data2 <- All_data2 %>%
-  mutate(length = To - From,
+  mutate(Length = To - From,
          mid_year = From + (To - From)/2)
@@
                     Dryness + 
-                    length +
+                    Length +
                     Precip_data_type +  Assessment_technique +
@@
 
-The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations thate are worth investigating, even though in general cross correlation is quite low between variables.
+The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations thate are worth investigating, even though in general cross correlations between variables are quite low.
 Some interesting relationships that appear in this graph are:
 
-- the negative relationship between log10(Area) and change in forest area (DeltaF_perc) indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.  
+- the negative relationship between log10(Area) and change in forest area (DeltaF_perc), indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.
 - the weak positive relationship between log10(Area) and the assessment method using hydrological models. This highlights that paired catchment studies mostly concentrate on smaller scales.
-- A strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+- A strong inverse relationship between log10(Area) and the paired watershed assessment method (simply the inverse from the last point), which is also indicated by the negative relationship between the two assessment methods.
 This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
@@
 If the effective degrees of freedom were close to *k - 1* then *k* was increased and the model rerun.
-By using shrinkage splines, this also results in the whole term being shrunk to zero [@wood2006].
+By using shrinkage splines, this also results in the whole term being shrunk to zero if needed [@wood2006].
 
@@
                     s(Dryness, k = 10, bs="ts" ) + 
-                     s(length, k = 35, bs="ts") +
+                     s(Length, k = 35, bs="ts") +
                     Precip_data_type +  Assessment_technique +
@@
 The overall explaining power of the model can be interpreted from the adjusted *r^2^* (which is penalised for the number of parameters).
-This indicates an *r^2^* of `r round(summary(Forest_model_all)$r.sq,2)` and deviance explained is `r round(summary(Forest_model_all)$dev.expl,2)`, suggesting the model only explains about 50% of the variance in the data.
+This indicates an adjusted *r^2^* of `r round(summary(Forest_model_all)$r.sq,2)` and deviance explained is `r round(summary(Forest_model_all)$dev.expl,2)`, suggesting the model only explains about 50% of the variance in the data.
 
 Inspecting the significance of the variables (Table \@ref(tab:m-all-linear) and Table \@ref(tab:m-all-smooth)) indicates some interesting features.
-
 The overall partial slope of the change in forest cover is `r round(summary(Forest_model_all)$p.table[2,1],2)`, if all other variables are kept constant.
@@
 
-In addition, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(Forest_model_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(Forest_model_all)$s.table[2,4],2)`)) and *length* (p = `r round(summary(Forest_model_all)$s.table[3,4],2)`)) explain variation in the data.
-For *length*, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
+In addition, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(Forest_model_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(Forest_model_all)$s.table[2,4],2)`)) and *Length* (p = `r round(summary(Forest_model_all)$s.table[3,4],2)`)) explain variation in the data.
+For *Length*, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
 
 Furthermore Table \@ref(tab:m-all-linear) indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
-In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables ($p < 0.05$).
-
-The remaining variables related to rainfall observation technique, forest type or hydrological regime don't appear to have an influence on the change in flow.
+In particular, the assessment methods Paired Watersheds Experiments (PWE), Hydrological Modelling (HM) and Statistical modelling and hydrographs (SH) are important explaining variables ($p < 0.05$).
+
+The remaining variables related to rainfall observation technique, forest type, or hydrological regime don't appear to have an influence on the change in flow.
 
@@
 The model assuming a simple linear relationship between change in forest cover (both positive and negative) and the change in flow explained the most variation in the data and indicated the best performance in terms of the Aikaike Information Criterium (AIC); and
-2. There is no need to assume a non-linear relationship, as a linear relationship provides a better fit to the data.
+2. There is no need to assume a non-linear relationship, as a linear relationship provides a similar performance for the fit to the data.
 
@@
 A <- visreg(Forest_model_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(Forest_model_all, "length", gg = T, plot = T) + theme_bw()
+L <- visreg(Forest_model_all, "Length", gg = T, plot = T) + theme_bw()
 #From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
@@
 # visreg(model6_all,"Dryness",gg=T) + theme_bw()
-# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
+# visreg(model6_all,"Length",gg=T, add=T) + theme_bw()
 # visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
@@
 
-Both the *length* and *Dryness* variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
-A further problem appears to be that *length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the relationship.
+Both the *Length* and *Dryness* variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
+A further problem appears to be that *Length* and *Dryness* have several points with very high leverage that determine much of the non-linearity in the relationship.
 
@@
 As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \@ref(fig:gamcheckmodelall)).
-Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall)).
+Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall), top left).
 This is related to a limited number catchments that have very high changes in streamflow in the data set.
@@
 pander(All_data2 %>% filter(Dryness > 5) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
-         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 5")
-```
-
-The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
+         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) Catchments for which the dryness index > 5")
+```
+
+The flexible nature of the splines means that the Length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
 The shape of the conditional response (Figure \@ref(fig:smoothsmodelall)) does not reflect a similar response as indicated by @filoso2017 and @jackson2005.
@@
 
-NoLong_Data <- All_data2 %>% filter(length<= 60, Dryness <= 5)
+NoLong_Data <- All_data2 %>% filter(Length<= 60, Dryness <= 5)
 
@@
                       s(log10(Area_km2), k = 5, bs = "ts") +
-                s(length, k = 10, bs = "ts") + 
+                s(Length, k = 10, bs = "ts") + 
                     Precip_data_type +  Assessment_technique +
@@
 A <- visreg(model_redLength, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
-L <- visreg(model_redLength, "length", gg = T, plot = T) + theme_bw()
+L <- visreg(model_redLength, "Length", gg = T, plot = T) + theme_bw()
 #From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
@@
 Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
-Data that have *Dryness* <= 5 and *length* <= 60 years were removed from the dataset and the model based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments is run again.
+Data that have *Dryness* <= 5 and *Length* <= 60 years were removed from the dataset and the model based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments is run again.
 
@@
 Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship looks almost linear.
-More importantly, the variable *length* is no longer significant, after removal of the two studies with very long lengths.
+More importantly, the variable *Length* is no longer significant, after removal of the two studies with very long lengths.
 
@@
 -->
-
-Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA).
+```{r data3, echo = F}
+All_data3 <- All_data2 %>%
+  filter(Assessment_technique != "EA, HM") %>%
+  filter(Assessment_technique != "PWE, HM") %>%
+  filter(Assessment_technique != "QPW, EA") 
+```
+
+Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA) and the assessment technique QPW, which were all non-significant (Table \@ref(tab:tableassess)).
+This resulted in a data set of `r nrow(All_data3)` catchment studies.
 
 ```{r model_reduc_assess, echo = F}
-All_data3 <- All_data2 %>%
-  filter(Assessment_technique != "EA, HM" & Assessment_technique != "PWE, HM" & Assessment_technique != "QPW, EA" )
-
 model_assess <- gam(DeltaQf_perc ~ DeltaF_perc  + 
                     s(Dryness, k=10,bs="ts" ) + #s(Latitude, bs="ts") + 
-                      s(log10(Area_km2), k = 10, bs = "ts") + s(length, bs = "ts") + #s(From, bs = "ts") +
+                      s(log10(Area_km2), k = 10, bs = "ts") + s(Length, bs = "ts") + #s(From, bs = "ts") +
                     Precip_data_type +  Assessment_technique + Forest_type +
-                    Hydrological_regime, data = All_data3 %>% filter(length< 60, Dryness <= 4))
+                    Hydrological_regime, data = All_data3 %>% filter(Length< 60, Dryness <= 4))
 #plot(model8_assess)
@@
 ```
-
-```{r, echo = F}
+<!-- no longer include these tables -->
+```{r, echo = F, eval =F}
 pander(round(summary(model_assess)$p.table,2), caption= "(\\#tab:modelassess-linear) Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques")
@@
 
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \@ref(tab:modelassess-linear) and \@ref(tab:modelassess-smooth)).
+The model based on assessment techniques that have more than 10 observations in the data set does not change much in the results (results not shown).
 It strengthens the significance of the different assessment techniques,  but generally results in the same interpretation.
@@
 
-However, the model results also clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
+The overall model results clearly highlight that some of the assessment techniques (in particular paired watershed studies (PWE) and combined use of statistical methods and hydrographs (SH)), have a strong impact on the predicted change in flow.
 Particularly, relative to EA (elasticity approaches) all other assessment techniques have higher predicted changes in flow.
@@
 
-The results presented so far, while using generalised additive modelling rather than single variable regression, end up with roughly the same conclusions as earlier papers [@zhang2017; @filoso2017].
-It appears that:
+The generalised additive models appear to reach the same conclusions as the single variable regression in earlier papers [@zhang2017; @filoso2017]. It appears that:
 
@@
 3. There is a general linear relationship between the change in forest cover and the change in streamflow.
+
+This might suggest that the simpler models have reached the correct conclusion. However, this is somewhat premature. given that the other major point coming out of the results is:
+
+4.
+There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest % forest cover change and the largest variability in the flow response.
 
@@
 
-Figure \@ref(fig:overview) provides a further overview of the whole data set and the size of the catchments and the different assessment methods are highlighted.
+Figure \@ref(fig:overview) provides a clear overview of the whole data set, and in this figure the size of the catchments and the different assessment methods are highlighted.
 This figure clearly indicates that the data relating to high changes in forest cover are all small catchments and relate mostly to paired watershed experiments.
 In contrast, data related to large catchments are related to smaller changes in forest cover and different methods, such as hydrological modelling and elasticity analysis.
-This confirms the model results (Table \@ref(tab:modelassess-linear)) and the earlier correlation analysis (Figure \@ref(fig:corgraphs)).
+This confirms the model results (Table \@ref(tab:m-red-linear)) and the earlier correlation analysis (Figure \@ref(fig:corgraphs)).
 
@@
 
-The results show that it is simply impossible to analyse a single covariate relationship, as there are several latent variables in the data.
+The results show that it is simply impossible to analyze a single covariate relationship, as there are several latent variables in the data.
 An example of this is the general relationship of the change in flow as a function of the change in forest cover.
@@
 
-The only  way to analyse changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in @levy2018).
+The only  way to analyze changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in @levy2018).
 
@@
 
-Clearly, interpreting older papers can be difficult and this can result in variation in the data that is being analysed.
+Clearly, interpreting older papers can be difficult and this can result in variation in the data that is being analyzed.
 Similar to the last issue, if these errors only introduce small variation in the data, then it will not limit the interpolation to larger scales.
@@
 
-The last issue leads into the next issue.
 For many of the small catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary "forestation" or "deforestation".
@@
 
-In our work, the variable *length* was used in the model to test for some of these effects, but this was insignificant in the model (Table \@ref(tab:restrictlength)).
+In our work, the variable *Length* was used in the model to test for some of these effects, but this was insignificant in the model (Table \@ref(tab:restrictlength)).
 Given the other variation in the data, this does not necessarily mean that there is no effect.
@@
 An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
-For example, @heath2014 found that catchments with resprouting species around Sydney indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent @zhou2015bushfire.
+For example, @heath2014 found that catchments with resprouting species around Sydney, Australia, indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent [@zhou2015bushfire].
 
@@
 
-A second potential danger is the extrapolation of the local small catchment results and conclusions tolarger scales, but beyond the original  scope of the studies.
+A second potential danger is the extrapolation of the local small catchment results and conclusions to larger scales, but beyond the original  scope of the studies.
 For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
@@
 Plantation forests are generally fast growing hybrids that will have quite different ecophysiology, particularly in South America [@jones2017; @binkley2017], while other reforestation, for example for salinity control in Australia, might focus on a mix of native species.
-Given the link between ecophysiology and water and carbon budgets, care should be taken in extrapolation, introducing a further error.
-
-A final factor is the effect of climate change [@vervoort2021] on runoff, even if the effects are still minor.
-Earlier papers [@li2017; @wang2020] have analysed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
+Given the link between ecophysiology and water and carbon budgets [@jackson2005], care should be taken in extrapolation, introducing a further error.
+
+A final factor is ignoring the effect of climate change [@vervoort2021] on runoff, even if the effects are still minor.
+Earlier papers [@li2017; @wang2020] have analyzed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
 Given that the database of studies now captures almost 100 years of work, we cannot ignore a climate change trend that is potentially hidden in the data.
 A simple inclusion of the start date of the experiment (*From*) in the GAM model does suggests an increase in change in the percentage of flow over time.
-However, this is once again not easy to analyse.
-As the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
+However, as the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
 
@@
 
-This study also points to several further opportunities and future research needs.
-
-A major focus of many of the papers related to forest hydrology have been specifically interested in the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
-As the paper by @jones2017 highlights this means there are opportunities to analyse the time evolution of the catchment response to forestation.
+Beyond a more formal approach to investigating climate change effects in the data, this study also points to several further opportunities and future research needs.
+
+A major focus of many of the papers related to forest hydrology has been on the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
+As the paper by @jones2017 highlights this means there are opportunities to analyze the time evolution of the catchment response to forestation.
 Given the large number of studies that look at a time evolution of forest cover (i.e.
-either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in landcover.
-This will give further insights into the equilibrium state under forestry.
+either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in land cover.
 As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
@@
 At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
-A better approach is analyzing streamflow data at multiple spatial scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes appears to be the best approach.
+A better approach is analyzing streamflow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes.
 The highlighted paper by @levy2018 is currently the best example of a solid statistical approach to analyzing streamflow responses.
@@
 
-This study demonstrates that analyses of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+This study demonstrates that analyzes of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
 

### 2022-09-28 paper/Forest_and_Water/Forest_and_Water.tex
revision of draft for internal review

@@
    \ead{jalonso@fing.edu.uy} 
-      \affiliation[a]{The University of Sydney, Sydney, NSW 2006, Australia}
-    \affiliation[b]{ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
+      \affiliation[a]{Sydney Institute of Agriculture and School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
+    \affiliation[b]{ARC ITTC in Data Analytics for Resources and Environments, The University of Sydney, NSW 2006, Sydney, Australia}
     \affiliation[c]{Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
@@
   \begin{abstract}
-  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
+  Three recent papers review and analyze large global datasets related to impacts of forest cover on streamflow.
 Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
-The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling.
-The overall results of the modelling indicates similar results to previous papers.
-The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results.
-The area of the catchment is strongly related to the assessment technique and the variability in the response data.
-  A careful review of the data and the methodology in the previous papers highlights that there are four interlinked reasons that are problematic: 1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
-Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.
+However, the results are problematic, the underlying data set is unbalanced, and there are correlations in the data that warrant further investigation as this would influence the results.
+For example, the area of the catchment is strongly related to the assessment technique and the variability in the response data.
+For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies.
+Subsequently, the data were re-analyzed using generalised additive modelling.
+  The results highlight that there are four interlinked reasons that make the general outcomes from the previous papers problematic: 1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalised and extrapolated to continental and global scales.
   \end{abstract}
@@
 The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic result is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
+The most dramatic example of this is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
 However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% changes in streamflow as a result of reforestation.
-Similarly a modelling study by \citet{beck2013} found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation.
+For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% change in streamflow as a result of reforestation.
+Similarly a modelling study by \citet{beck2013} found no significant change in streamflow in 12 catchments in Puerto Rico as a result of deforestation.
 In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
-This suggests that there can be significant variation across the different studies,methodologies and geographical regions.
+This suggests that there can be significant variation across the different studies, methodologies and geographical regions.
 
@@
 
-As mentioned, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies \citep{brown2005, hewlett1984}, related to reforestation in particular \citep{filoso2017}, and more generally \citep{jackson2005, zhang2017}.
+There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarise past studies.
+The suggestion is that, because meta-analyses use clearly defined search terms and statistical methods to analyze the results, this will lead to more reliable summaries of past research.
+As a result, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies \citep{brown2005, hewlett1984}, related to reforestation in particular \citep{filoso2017}, and more generally \citep{jackson2005, zhang2017}.
 These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships.
-Others have used the understanding from these studies to extrapolate to global scales \citep{hoekvandijke2022}.
-
-There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarise past studies.
-The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.
-
-From the most recent reviews on forest impact on streamflow \citep{zhang2017, filoso2017}, the study by \citet{filoso2017} is a clear meta-analysis, while the other is not.
-However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data.
+Others have used the understanding from a global analysis to extrapolate to global scales \citep{hoekvandijke2022}.
+
+The recent paper by \citet{filoso2017} is a clear meta-analysis, but most others \citep{zhang2017, hoekvandijke2022, zhou2015} are not.
+However, n impressive global database of catchment studies with changes in streamflow due to changes in forest cover has been developed \citep{zhang2017, filoso2017} and statistical approaches are used to analyze the resulting data.
 The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by \citet{filoso2017} focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
-In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.
+In this case the collected data is mostly coded as count data and only a subset of 37 studies was analyzed for actual water yield change.
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
@@
 The conclusions of the first mentioned major review paper \citep{zhang2017} indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as \textless{} 1000 km\textsuperscript{2} and large watersheds (catchments) \textgreater{} 1000 km\textsuperscript{2}.
-While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover.
-Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments \citep{zhang2017}.
+While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with increases in forest cover.
+The main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments \citep{zhang2017}.
 
@@
 
-A final earlier summary paper that includes much of the same data as \citet{zhang2017} and \citet{filoso2017} is \citet{zhou2015}, which has one author in common with \citet{zhang2017}.
-However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model.
+An earlier paper, that includes much of the same data as \citet{zhang2017} and \citet{filoso2017}, is \citet{zhou2015}, which has one author in common with \citet{zhang2017}.
+However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model, which is similar to well-known Budyko approaches \citep{zhang2018elasticity}.
 In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model.
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in \citet{zhang2017}.
-The Fuh model and variations of the Budyko equilibrium modelling approach was also used by \citet{hoekvandijke2022} to interpret the global impact of reforestation.
-
-There are some clear limitations in these studies, and some of this applies to meta-analyses in general.
+The Fuh model and the related variations of the Budyko equilibrium modelling approach was also used by \citet{hoekvandijke2022} to interpret the global impact of reforestation.
+
+However, concerning is that there are some clear limitations in these studies, and some of this applies to meta-analyses in general.
 The main method in the work by \citet{zhang2017} is a single covariate linear regression.
-In contrast, the systematic review from \citet{filoso2017} emphasises the classification and distributions of the study.
+In contrast, the systematic review from \citet{filoso2017} mainly emphasises the classification and distributions of the study.
 \citet{zhang2017} points out that a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments.
@@
 
-In contrast to the single covariate linear regression in the earlier studies \citep{zhang2017, filoso2017} and the top-down Budyko modelling \citep{zhou2015, hoekvandijke2022}, the regional Brazilian Cerrado study \citep{levy2018} provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
+In contrast to the single covariate linear regression in the earlier studies \citep{zhang2017, filoso2017} and the top-down Budyko modelling \citep{zhou2015, hoekvandijke2022}, the regional Brazilian Cerrado study \citep{levy2018} provides an example of an carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation.
 The analysis specifically accounted for differences between catchments and differences due to variations in climate.
-Their conclusion highlighted that in particular dry season streamflow was affected by deforestation.
-
-Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
-There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales \citep{evaristo2020metaanalysis}.
-However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:
+Not all datasets are however suitable for this kind of in-depth analysis.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another review paper on this topic?
+There is a real attraction in the concept of statistical analysis of past studies encapsulated in meta-analysis to be able to extrapolate findings to larger scales, and to identify factors across global scales \citep{evaristo2020metaanalysis}.
+However, there are also some hidden complications in this that can invalidate results, which this paper aims to highlight.
+There are four potential errors (or limitations) in such global meta-analyses:
 
@@
 \item
-  Interpretation errors due to incomplete descriptions of the experiments in the original papers;\\
-\item
-  Aggregation of data that originates from different experiments with different objectives across a wide time period, but have similar keywords; and, finally\\
+  Interpretation errors due to incomplete descriptions of the experiments in the original papers;
+\item
+  Aggregation of data that originates from different experiments with different objectives across a wide time period, but have similar keywords; and, finally
 \item
@@
 
-The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation.
+The aim of this paper is to first reanalyze the global dataset \citep{zhang2017, filoso2017} using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
 This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
-In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow.
-Finally, we will highlight future research needs in this area.
+In addition, by applying more complex statistical models, we will highlight the conclusions that can be drawn from the data.
+Finally, we will highlight future research needs in the area of forest cover change impact on streamflow.
 
@@
 As indicated, the starting point of this paper is the data base of studies which were included in \citet{zhang2017} as supplementary material.
-The columns in this data set (are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), based on equation 1 in \citet{zhang2017}), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
+The columns in this data set are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
+The change in streamflow (\(\Delta Qf\%\)) is based on equation 1 in \citet{zhang2017}.
+
 Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \ref{tab:table1}. These abbreviations will later be used in the models.
@@
 
-The paper by \citet{zhang2017} use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET\textsubscript{0} or E\textsubscript{0}) in their analysis, and use the dryness index to identify the climate type.
-However, the potential or reference ET was not originally included as part of the published data set.
+The paper by \citet{zhang2017} also uses the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET\textsubscript{0} or E\textsubscript{0}) in their analysis, and have used the dryness index to identify the climate type.
+However, the potential or reference ET used for this calculation was originally not included in the published data set.
+We will discus below how we derived the dryness index in our data set.
 We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) from \citet{zhang2017} in our analysis.
@@
 As highlighted, \citet{zhang2017} did not provide values for evapotranspiration in the data base.
-Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 \citep{trabucco2018}, if a value of E\textsubscript{0} was not available from the original papers.
+Using the location information, reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 \citep{trabucco2018}, if a value of E\textsubscript{0} was not available from the original papers.
 For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.
 
-Similar to \citet{zhang2017}, the ``dryness index'' was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
+Similar to \citet{zhang2017}, the Dryness index was calculated from the catchment estimate of reference evapotranspiration and the catchment estimate of annual average rainfall (Pa) as:
 
 \begin{equation}
-D = \frac{E_{0}}{Pa} \label{eq:eq1}
+Dryness = \frac{E_{0}}{Pa} \label{eq:eq1}
 \end{equation}
@@
 
-The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change flow as a consequence of changes in forest cover.
+The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change in flow as a result of changes in forest cover.
 This first aim is similar to \citet{zhang2017}, but the main difference is that we start off with all variables in the data set in the model.
@@
 
-In the statistical analysis we are not necessarily seeking the best ``predicitive'' model, and as such do not perform a traditional variable selection process.
+In the statistical analysis we are not necessarily seeking the best ``predictive'' model, and as such do not perform a traditional variable selection process.
 Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.
@@
 The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
+In \citet{zhang2017}, these changes were jointly analyzed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
 
 However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-The question becomes how best to analyse this.
+The question becomes how best to analyze this.
 One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
@@
 The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form.
-Both these approaches are tested in the results.
-
-The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
+All three approaches are tested in this study.
+
+The overarching test focuses on identifying the change streamflow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
 In addition to these earlier identified factors, this study also tested for the factors listed in Table \ref{tab:table1}
@@
 In contrast, the original dataset from \citet{zhang2017} contained 312 catchments and the \citet{filoso2017} study used 37 catchments (Table S2 in \citet{filoso2017}).
-The current number of catchments is the result of the removal of duplicates and our modifications and additions.
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\textsuperscript{2}}.
-The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution.
+The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails in the change in flow distribution.
 Note also the large number of studies with 100\% forest cover reduction.
 Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).
-This high skew in some of the data can create difficulties in the statistical modelling and further transformation of the data might be required.
+This high skew in some of the data can create difficulties in the statistical modelling and this will be discussed later.
 
@@
 
-The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations thate are worth investigating, even though in general cross correlation is quite low between variables.
+The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations thate are worth investigating, even though in general cross correlations between variables are quite low.
 Some interesting relationships that appear in this graph are:
@@
 \item
-  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc) indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.\\
-\item
-  the weak positive relationship between log10(Area) and the assessment method using hydrological models. This highlights that paired catchment studies mostly concentrate on smaller scales.\\
-\item
-  A strong inverse relationship between log10(Area) and the paired watershed assessment method, which is simply the inverse from the last point, as also indicated by the negative relationship between the two assessment methods.
+  the negative relationship between log10(Area) and change in forest area (DeltaF\_perc), indicating that in the data set larger catchments tended to have (obviously) smaller areas of forest change.
+\item
+  the weak positive relationship between log10(Area) and the assessment method using hydrological models. This highlights that paired catchment studies mostly concentrate on smaller scales.
+\item
+  A strong inverse relationship between log10(Area) and the paired watershed assessment method (simply the inverse from the last point), which is also indicated by the negative relationship between the two assessment methods.
 This is further visible in the relationship between the change in forest cover and the paired watershed assessment method, showing the impact of the latent variable (log10(Area)).
@@
 If the effective degrees of freedom were close to \emph{k - 1} then \emph{k} was increased and the model rerun.
-By using shrinkage splines, this also results in the whole term being shrunk to zero \citep{wood2006}.
+By using shrinkage splines, this also results in the whole term being shrunk to zero if needed \citep{wood2006}.
 
@@
 \textbf{s(Dryness)} & 4.64 & 9 & 2.26 & 0 \\
-\textbf{s(length)} & 4.45 & 34 & 0.22 & 0.12 \\
+\textbf{s(Length)} & 4.45 & 34 & 0.22 & 0.12 \\
 \bottomrule()
@@
 The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters).
-This indicates an \emph{r\textsuperscript{2}} of 0.45 and deviance explained is 0.49, suggesting the model only explains about 50\% of the variance in the data.
+This indicates an adjusted \emph{r\textsuperscript{2}} of 0.45 and deviance explained is 0.49, suggesting the model only explains about 50\% of the variance in the data.
 
 Inspecting the significance of the variables (Table \ref{tab:m-all-linear} and Table \ref{tab:m-all-smooth}) indicates some interesting features.
-
 The overall partial slope of the change in forest cover is -0.6, if all other variables are kept constant.
@@
 
-In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{length} (p = 0.12)) explain variation in the data.
-For \emph{length}, the p-value is not strictly smaller, than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
+In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{Length} (p = 0.12)) explain variation in the data.
+For \emph{Length}, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
 
 Furthermore Table \ref{tab:m-all-linear} indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
-In particular, the assessment methods Paired Watersheds experiments (PWE), Hydrological modelling (HM) and Statistical techniques (SH) are important explaining variables (\(p < 0.05\)).
-
-The remaining variables related to rainfall observation technique, forest type or hydrological regime don't appear to have an influence on the change in flow.
+In particular, the assessment methods Paired Watersheds Experiments (PWE), Hydrological Modelling (HM) and Statistical modelling and hydrographs (SH) are important explaining variables (\(p < 0.05\)).
+
+The remaining variables related to rainfall observation technique, forest type, or hydrological regime don't appear to have an influence on the change in flow.
 
@@
 \item
-  The model assuming a simple linear relationship between change in forest cover (both positive and negative) and the change in flow explained the most variation in the data and indicated the best performance in terms of the Aikaike Information Criterium (AIC); and\\
-\item
-  There is no need to assume a non-linear relationship, as a linear relationship provides a better fit to the data.
+  The model assuming a simple linear relationship between change in forest cover (both positive and negative) and the change in flow explained the most variation in the data and indicated the best performance in terms of the Aikaike Information Criterium (AIC); and
+\item
+  There is no need to assume a non-linear relationship, as a linear relationship provides a similar performance for the fit to the data.
 \end{enumerate}
@@
 
-Both the \emph{length} and \emph{Dryness} variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
-A further problem appears to be that \emph{length} and \emph{Dryness} have several points with very high leverage that determine much of the non-linearity in the relationship.
+Both the \emph{Length} and \emph{Dryness} variables show strong non-linearity, but the relationships do not show a clear trend due to the scatter and the distribution of the data.
+A further problem appears to be that \emph{Length} and \emph{Dryness} have several points with very high leverage that determine much of the non-linearity in the relationship.
 
@@
 As this is not always shown in papers discussing regression relationship, the residual distribution is provided in more detail (Figure \ref{fig:gamcheckmodelall}).
-Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}).
+Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}, top left).
 This is related to a limited number catchments that have very high changes in streamflow in the data set.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3889}}@{}}
-\caption{\label{tab:drytable} catchments for which the dryness index \textgreater{} 5}\tabularnewline
+\caption{\label{tab:drytable} Catchments for which the dryness index \textgreater{} 5}\tabularnewline
 \toprule()
@@
 
-The flexible nature of the splines means that the length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
+The flexible nature of the splines means that the Length variable highlights substantial non-linearity in the data, but it is unclear what exactly is captured.
 The shape of the conditional response (Figure \ref{fig:smoothsmodelall}) does not reflect a similar response as indicated by \citet{filoso2017} and \citet{jackson2005}.
@@
 \textbf{s(log10(Area\_km2))} & 0.87 & 4 & 1.53 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.98 \\
+\textbf{s(Length)} & 0 & 9 & 0 & 0.98 \\
 \bottomrule()
@@
 Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
-Data that have \emph{Dryness} \textless= 5 and \emph{length} \textless= 60 years were removed from the dataset and the model based on a reduction of the data set from 329 to 310 catchments is run again.
+Data that have \emph{Dryness} \textless= 5 and \emph{Length} \textless= 60 years were removed from the dataset and the model based on a reduction of the data set from 329 to 310 catchments is run again.
 
@@
 Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship looks almost linear.
-More importantly, the variable \emph{length} is no longer significant, after removal of the two studies with very long lengths.
+More importantly, the variable \emph{Length} is no longer significant, after removal of the two studies with very long lengths.
 
@@
 
-Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4000}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1333}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1467}}@{}}
-\caption{\label{tab:modelassess-linear} Statistical overview of the linear components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & -12.17 & 14.9 & -0.82 & 0.41 \\
-\textbf{DeltaF\_perc} & -0.58 & 0.05 & -10.79 & 0 \\
-\textbf{Precip\_data\_typeOB} & -16.53 & 12.37 & -1.34 & 0.18 \\
-\textbf{Precip\_data\_typeSG} & 15.32 & 14.39 & 1.06 & 0.29 \\
-\textbf{Assessment\_techniqueHM} & 28.74 & 11.16 & 2.57 & 0.01 \\
-\textbf{Assessment\_techniquePWE} & 31.37 & 11.3 & 2.78 & 0.01 \\
-\textbf{Assessment\_techniqueQPW} & 44.15 & 18.89 & 2.34 & 0.02 \\
-\textbf{Assessment\_techniqueSH} & 49.42 & 11.32 & 4.37 & 0 \\
-\textbf{Forest\_typeCF} & -8.72 & 7.25 & -1.2 & 0.23 \\
-\textbf{Forest\_typeMF} & -3.94 & 7.43 & -0.53 & 0.6 \\
-\textbf{Hydrological\_regimeSD} & 2.01 & 8.86 & 0.23 & 0.82 \\
-\bottomrule()
-\end{longtable}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{\label{tab:modelassess-smooth} Statistical overview of the smooth components of the model removing studies with limited observations in the assessment techniques}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{s(Dryness)} & 3.37 & 9 & 2.79 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.85 & 9 & 0.68 & 0.01 \\
-\textbf{s(length)} & 0 & 9 & 0 & 0.84 \\
-\bottomrule()
-\end{longtable}
-
-Concentrating only on the assessment techniques that have more than 10 observations in the data set does not change much in the results (Table \ref{tab:modelassess-linear} and \ref{tab:modelassess-smooth}).
+Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA) and the assessment technique QPW, which were all non-significant (Table \ref{tab:tableassess}).
+This resulted in a data set of 323 catchment studies.
+
+The model based on assessment techniques that have more than 10 observations in the data set does not change much in the results (results not shown).
 It strengthens the significance of the different assessment techniques, but generally results in the same interpretation.
@@
 
-The results presented so far, while using generalised additive modelling rather than single variable regression, end up with roughly the same conclusions as earlier papers \citep{zhang2017, filoso2017}.
-It appears that:
+The generalised additive models appear to reach the same conclusions as the single variable regression in earlier papers \citep{zhang2017, filoso2017}. It appears that:
 
@@
 \item
-  Larger catchments show lower impact of forest cover change on streamflow;\\
-\item
-  Drier catchments show a greater impact of forest cover change on streamflow; and\\
+  Larger catchments show lower impact of forest cover change on streamflow;
+\item
+  Drier catchments show a greater impact of forest cover change on streamflow; and
 \item
   There is a general linear relationship between the change in forest cover and the change in streamflow.
+\end{enumerate}
+
+This might suggest that the simpler models have reached the correct conclusion. However, this is somewhat premature. given that the other major point coming out of the results is:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest \% forest cover change and the largest variability in the flow response.
 \end{enumerate}
@@
 
-Figure \ref{fig:overview} provides a further overview of the whole data set and the size of the catchments and the different assessment methods are highlighted.
+Figure \ref{fig:overview} provides a clear overview of the whole data set, and in this figure the size of the catchments and the different assessment methods are highlighted.
 This figure clearly indicates that the data relating to high changes in forest cover are all small catchments and relate mostly to paired watershed experiments.
 In contrast, data related to large catchments are related to smaller changes in forest cover and different methods, such as hydrological modelling and elasticity analysis.
-This confirms the model results (Table \ref{tab:modelassess-linear}) and the earlier correlation analysis (Figure \ref{fig:corgraphs}).
+This confirms the model results (Table \ref{tab:m-red-linear}) and the earlier correlation analysis (Figure \ref{fig:corgraphs}).
 
@@
 
-The only way to analyse changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in \citet{levy2018}).
+The only way to analyze changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in \citet{levy2018}).
 
@@
 
-The last issue leads into the next issue.
 For many of the small catchment studies listed in the database, the assumption is that the original experimental design can be interpreted in terms of a binary ``forestation'' or ``deforestation''.
@@
 
-In our work, the variable \emph{length} was used in the model to test for some of these effects, but this was insignificant in the model (Table \ref{tab:restrictlength}).
+In our work, the variable \emph{Length} was used in the model to test for some of these effects, but this was insignificant in the model (Table \ref{tab:restrictlength}).
 Given the other variation in the data, this does not necessarily mean that there is no effect.
@@
 An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
-For example, \citet{heath2014} found that catchments with resprouting species around Sydney indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent \citet{zhou2015bushfire}.
+For example, \citet{heath2014} found that catchments with resprouting species around Sydney, Australia, indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent \citep{zhou2015bushfire}.
 
@@
 
-A second potential danger is the extrapolation of the local small catchment results and conclusions tolarger scales, but beyond the original scope of the studies.
+A second potential danger is the extrapolation of the local small catchment results and conclusions to larger scales, but beyond the original scope of the studies.
 For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
@@
 Plantation forests are generally fast growing hybrids that will have quite different ecophysiology, particularly in South America \citep{jones2017, binkley2017}, while other reforestation, for example for salinity control in Australia, might focus on a mix of native species.
-Given the link between ecophysiology and water and carbon budgets, care should be taken in extrapolation, introducing a further error.
-
-A final factor is the effect of climate change \citep{vervoort2021} on runoff, even if the effects are still minor.
-Earlier papers \citep{li2017, wang2020} have analysed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
+Given the link between ecophysiology and water and carbon budgets \citep{jackson2005}, care should be taken in extrapolation, introducing a further error.
+
+A final factor is ignoring the effect of climate change \citep{vervoort2021} on runoff, even if the effects are still minor.
+Earlier papers \citep{li2017, wang2020} have analyzed climate effects relative to management effects in the data, but these studies did not explicitly test for climate change.
 Given that the database of studies now captures almost 100 years of work, we cannot ignore a climate change trend that is potentially hidden in the data.
 A simple inclusion of the start date of the experiment (\emph{From}) in the GAM model does suggests an increase in change in the percentage of flow over time.
-However, this is once again not easy to analyse.
-As the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
+However, as the data distribution is uneven in time, and consists of multiple assessment techniques there could be multiple complicating factors, and drawing a firm conclusion would be premature.
 
@@
 
-This study also points to several further opportunities and future research needs.
-
-A major focus of many of the papers related to forest hydrology have been specifically interested in the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
-As the paper by \citet{jones2017} highlights this means there are opportunities to analyse the time evolution of the catchment response to forestation.
-Given the large number of studies that look at a time evolution of forest cover (i.e.~either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in landcover.
-This will give further insights into the equilibrium state under forestry.
+Beyond a more formal approach to investigating climate change effects in the data, this study also points to several further opportunities and future research needs.
+
+A major focus of many of the papers related to forest hydrology has been on the impact of plantation forest operations on the catchment, rather than the transition of forestry to agriculture.
+As the paper by \citet{jones2017} highlights this means there are opportunities to analyze the time evolution of the catchment response to forestation.
+Given the large number of studies that look at a time evolution of forest cover (i.e.~either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in land cover.
 As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
@@
 At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
-A better approach is analyzing streamflow data at multiple spatial scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes appears to be the best approach.
+A better approach is analyzing streamflow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes.
 The highlighted paper by \citet{levy2018} is currently the best example of a solid statistical approach to analyzing streamflow responses.
@@
 
-This study demonstrates that analyses of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+This study demonstrates that analyzes of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
 

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart1.Rmd
almos there
revision of draft for internal review

@@
 - 76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication [@baker1984].
-- 124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by @zhang2017 is clearly wrong.
+- 124, D3, @amatya2008effects: The originally recorded 250\% change by @zhang2017 is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
 So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%.
-corrected to 16\%
-- 3, Baker Creek, Zhang and Wei, 2012. The original recorded 201.1\% change by @zhang2017 also seems wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
-- 67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in @zhang2017.
-This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates "clearfelling".
+corrected to 16\%.
+- 3, Baker Creek, @zhangwei2012. The original recorded 201.1\% change by @zhang2017 also seems wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  
+- 67, April rd, which is incorrectly attributed to @ruprecht1991 in @zhang2017.
+This is actually from @ruprecht1989 and the original paper clearly indicates "clearfelling".
 As a result the change in forest cover was changed to -100\% rather than +100\%.
-- 210, March rd, 100, 147.6. Same problem as 67, Bari et al. (1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.  
+- 210, March rd, 100, 147.6. Same problem as 67, @bari1996 clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
 - 213, 214 and 215, Monda 1, 2 and 3.
@@
 - 101 Coalburn.
-Original publication (Robinson, 1993) which is a symposium paper, is not available.
+Original publication (Robinson, 1993) which is a symposium paper, can not be located, even after contacting some of the authors.
 The best summary of the research is in @birkinshaw2014 which summaries 45 years of resaerch in the Coalburn catchment.

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart1.tex
revision of draft for internal review

@@
             bookmarks=true,
-            pdfauthor={Willem Vervoort/Eliana Nervi/Jimena Alonso},
-            pdftitle={Supplementary Information: Changes to original database},
+            pdfauthor={},
+            pdftitle={Supplementary Information part 1: Changes to original database},
             colorlinks=false,
@@
 
-  \title{Supplementary Information: Changes to original database}
-    \author[]{%
+  \title{Supplementary Information part 1: Changes to original database}
+    \author[]{R. Willem Vervoort%
   %
-  }
-
+  \fnref{1}}
+   \ead{willem.vervoort@sydney.edu.au} 
+    \author[]{Eliana Nervi}
+   \ead{eliananervif@gmail.com} 
+    \author[]{Jimena Alonso}
+   \ead{jalonso@fing.edu.uy} 
       \cortext[cor1]{Corresponding author}
+    \fntext[1]{Corresponding Author}
 
@@
 
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+This supplementary material is related to `Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
+
 This document list all the changes that were made to the original \citet{zhang2017} database after review of all the original papers.
 
-A particular problem was that many catchments had the wrong sign for the change in forest cover.
+\hypertarget{discussion-of-changes}{%
+\section{Discussion of changes}\label{discussion-of-changes}}
+
+\textbf{Need to fix up references}
+
+A particular problem was that many catchments appeared to have the wrong sign for the change in forest cover.
 There are many catchments with reported positive change in cover and a large increase in flow.
@@
 \item
-  76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication \citep{baker1984}.\\
+  76, Beaver Creek, the flow was corrected from 600\% to 157\% after review of the original publication \citep{baker1984}.
 \item
-  124, D3, Amatya and Skaggs, 2008: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong.
+  124, D3, \citet{amatya2008effects}: The originally recorded 250\% change by \citet{zhang2017} is clearly wrong.
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
 So value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%.
-corrected to 16\%
+corrected to 16\%.
 \item
-  3, Baker Creek, Zhang and Wei, 2012.
-The original recorde 201.1\% change by \citet{zhang2017} is also wrong.
+  3, Baker Creek, \citet{zhangwei2012}.
+The original recorded 201.1\% change by \citet{zhang2017} also seems wrong.
 Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
@@
 \item
-  67, April rd, which is incorrectly attributed to Ruprecht and Schofield (1991) in \citet{zhang2017}.
-This is actually from Ruprecht and Schofield (1989) and the original paper clearly indicates ``clearfelling''.
-As a result the change in forest cover was changed to -100\% rather than +100\%.\\
+  67, April rd, which is incorrectly attributed to \citet{ruprecht1991} in \citet{zhang2017}.
+This is actually from \citet{ruprecht1989} and the original paper clearly indicates ``clearfelling''.
+As a result the change in forest cover was changed to -100\% rather than +100\%.
 \item
-  210, March rd, 100, 147.6. Same problem as 67, Bari et al.~(1996) clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.\\
+  210, March rd, 100, 147.6. Same problem as 67, \citet{bari1996} clearly state that the catchment was cleared, so therefor the change in forest cover changed to -100\%.
 \item
@@
 Corrected.
-Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.\\
+Similar to other paired watershed experiments, only the first couple of years can be linked to the effect as later regrowth cancels out part of the increase in flow.
 \item
-  104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
+  104, Coshocton. Checking the original paper indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.
 \item
-  102, Cold Spring. Checking the original paper \citep{schneider1961} indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.\\
+  102, Cold Spring. Checking the original paper \citep{schneider1961} indicates that this is in fact a reduction in flow as a result of reforestation. Changed the sign of Delta\_Q\_f to be negative.
 \item
@@
 Finally, the catchments were all reforestation as the title of the original report indicates and the reported streamflows are all decreases.
-Corrected Delta\_Qf\_perc to -20.7\%.\\
+Corrected Delta\_Qf\_perc to -20.7\%.
 \item
-  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.\\
+  95 Sage Brook. Similar to 259 and 102, originates from \citet{schneider1961}. Reforestation so Delta\_Qf\_perc corrected to -19.8\%.
 \item
   101 Coalburn.
-Original publication (Robinson, 1993) which is a symposium paper, is not available.
+Original publication (Robinson, 1993) which is a symposium paper, can not be located, even after contacting some of the authors.
 The best summary of the research is in \citet{birkinshaw2014} which summaries 45 years of resaerch in the Coalburn catchment.

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart2.Rmd
revision of draft for internal review

@@
 
-```{r, echo = F, message = F, warning = F}
+```{r, echo = F, eval = F, message = F, warning = F}
 png("Fig2Zhang.png", width = 2500, height = 1800, res = 300)
@@
 
-```{r, echo = F, message = F, warning = F}
+```{r, echo = F, eval = F, message = F, warning = F}
 png("Fig3Zhang.png", width = 2500, height = 1800, res = 300)
@@
 
-```{r, echo = F, message = F, warning = F}
+```{r, echo = F, message = F, warning = F, eval = F}
 png("Fig4Zhang.png", width = 2500, height = 1800, res = 300)

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart3.Rmd
revision of draft for internal review

@@
 bibliography: forestandwater.bib
-editor_options: 
-  chunk_output_type: inline
+#editor_options: 
+#  chunk_output_type: inline
 header-includes:
@@
 In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-Here we test two variation of this approach, leading to three approaches
+Here we test two variation of this approach, leading to three approaches:
 
 1. Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following @zhang2017.
-
 2.
@@
 One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
-
 3. A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).

### 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart3.tex
revision of draft for internal review

@@
 In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-Here we test two variation of this approach, leading to three approaches
+Here we test two variation of this approach, leading to three approaches:
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
@@
 The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0.
-The overall model explains 50 \% of the variation in the data.
+The overall model explains 49 \% of the variation in the data.
 For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used.
-The AIC of the model is 3166.6801428.
+The AIC of the model is 3181.7600577.
 
@@
 \endhead
-\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
-\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
-\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
-\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
-\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
-\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
-\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
+\textbf{(Intercept)} & 14.05 & 20 & 0.7 & 0.48 \\
+\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.9 & 0 \\
+\textbf{Forest\_SignIncrease} & -51.91 & 7.79 & -6.67 & 0 \\
+\textbf{Precip\_data\_typeOB} & -19.43 & 14.56 & -1.33 & 0.18 \\
+\textbf{Precip\_data\_typeSG} & 4.43 & 17.22 & 0.26 & 0.8 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.4 & 46.56 & 0.4 & 0.69 \\
+\textbf{Assessment\_techniqueHM} & 21.99 & 13.63 & 1.61 & 0.11 \\
+\textbf{Assessment\_techniquePWE} & 27.8 & 13.85 & 2.01 & 0.05 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
-\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
+HM} & 17.58 & 46.6 & 0.38 & 0.71 \\
+\textbf{Assessment\_techniqueQPW} & 25.71 & 22.18 & 1.16 & 0.25 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
-\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
-\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
-\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
-\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
+EA} & 33.3 & 26.36 & 1.26 & 0.21 \\
+\textbf{Assessment\_techniqueSH} & 38.7 & 13.25 & 2.92 & 0 \\
+\textbf{Forest\_typeCF} & -12.71 & 8.25 & -1.54 & 0.12 \\
+\textbf{Forest\_typeMF} & -7.05 & 8.36 & -0.84 & 0.4 \\
+\textbf{Hydrological\_regimeSD} & 7.27 & 10 & 0.73 & 0.47 \\
 \bottomrule()
@@
 However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0).
-In fact, in absolute terms deforestration creates a -5226 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).
+In fact, in absolute terms deforestration creates a -5191 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).
 
-However, overall this model explains less of the overall variation in the data with a deviance explained of 0.46 and an AIC of 3212.700798.
+However, overall this model explains less of the overall variation in the data with a deviance explained of 0.45 and an AIC of 3227.3976077.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
@@
 \endhead
-\textbf{s(DeltaF\_perc)} & 1.11 & 9 & 12.72 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.76 & 4 & 0.8 & 0.03 \\
-\textbf{s(Dryness)} & 4.7 & 9 & 2.2 & 0 \\
-\textbf{s(length)} & 4.45 & 34 & 0.24 & 0.09 \\
+\textbf{s(DeltaF\_perc)} & 4.14 & 9 & 13.2 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.74 & 4 & 0.74 & 0.04 \\
+\textbf{s(Dryness)} & 4.73 & 9 & 2.4 & 0 \\
+\textbf{s(length)} & 4.68 & 34 & 0.26 & 0.08 \\
 \bottomrule()
@@
 In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0.
-The total deviance explained for the model is 0.5 and the model has an AIC of 3166.9317897.
+The total deviance explained for the model is 0.5 and the model has an AIC of 3182.1088731.
 These performance statistics are similar to the first (baseline) model.
@@
 ## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 15 iterations.
-## The RMS GCV score gradient at convergence was 0.004151441 .
+## Smoothing parameter selection converged after 11 iterations.
+## The RMS GCV score gradient at convergence was 0.0003791486 .
 ## The Hessian was positive definite.
@@
 ##                       k'   edf k-index p-value  
-## s(DeltaF_perc)      9.000  1.113    0.89    0.03 *
-## s(log10(Area_km2))  4.000  0.756    0.96    0.20  
-## s(Dryness)          9.000  4.703    0.94    0.13  
-## s(length)          34.000  4.448    0.92    0.10 .
+## s(DeltaF_perc)      9.00  4.14    0.91   0.075 .
+## s(log10(Area_km2))  4.00  0.74    0.95   0.140  
+## s(Dryness)          9.00  4.73    0.93   0.135  
+## s(length)          34.00  4.68    0.93   0.110  
 ## ---

### 2022-10-03 paper/Forest_and_Water/Reply_first_review_JoH.Rmd
resolve comment about percent total forest

@@
 This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
-Many of the Paired Watershed Experiments in Australia [e.g.
-@cornish2001; @webb2009; @webb2012; @webb2013; @watson2001] fall in this category, and therefore easily be classified as either forestation or deforestation.
+Many of the Paired Watershed Experiments [e.g.
+@cornish2001; @webb2009; @webb2012; @webb2013; @watson2001] fall in this category, and therefore can easily be classified as either forestation or deforestation.
 In principle a decision needs to be made how many years post clearing needs to be considered, and whether the remainder of the timeseries should be classified as regeneration.
-We demonstrate how removing these experiments from the total changes the results of the analysis.
+Or, as we discuss, the data should be analysed as a timeseries of change, rather than trying to pick a single point.
 - In general, the size of the catchment and the length of the study play a huge role in the interpretation of the results.
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
+Even though length is no significant in the data, does not mean that this effect exists in the data.
 In terms of the catchment size, one of the key issues we originally had with the @zhang2017 paper was the arbitrary split between catchment > 1000 km^2^ and catchment < 1000 km^2^.
@@
 
-Overall this analysis shows very clearly the danger of simply investigating single variable regressions as was done in @zhang2017 and @filoso2017 or even using simple modelling as in @jackson2005 and @zhou2015.
+Overall this analysis shows very clearly the challenges of simply investigating single variable regressions as was done in @zhang2017 and @filoso2017 or even using simple modelling as in @jackson2005 and @zhou2015.
 It is too easy to miss the latent variables that are the underlying factors and influence the model results.
@@
 As we also outline in our reply to the Editor, we have rewritten the paper to strengthen the novelty of the work.
-In particular the paper now more directly focuses on the difficulties in analysing aggregated global data and the importance of latent variables.
-We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this is fraught with danger.
+In particular the paper now focuses more directly on the difficulties in analysing aggregated global data and the importance of latent variables.
+We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this has significant challenges and can be easily lead to questionable results.
 
@@
 <!-- answer -->
-We agree with the AE that this would be an interesting idea and links back to our point about "latent variables".
+We agree with the AE that this was a logical idea and links back to our point about "latent variables".
 The excellent paper by @levy2018 includes \% forest area as a variable, but their data cannot be incorporated into the current paper.
 This is because their analysis focuses on a time progression rather than a "before and after" or a clear "control" and "treatment".
-Furthermore, extracting the total area of forest from the papers is not a trivial job, and the area is also not always mentioned in the papers.
-Finally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
-It is definitely an area of future research.
+Regrettably, extracting the total area of forest from the papers is not a trivial job, and also has some issues.
+The total forested area is also not always mentioned in the papers, or only a range is given.
+Additionally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
+
+```{r show_totalF_area, echo = F, message = F, warning=F}
+All_data <- read_csv("../../data/Data_TotForest_Area.csv")
+require(mgcv)
+Forest_model_all <- gam(DeltaQf_perc ~ DeltaF_perc + 
+                    log10(Area_km2) + 
+                    Dryness + 
+                    Perc_Farea_pre +
+                    Precip_data_type +  Assessment_technique +
+                    Forest_type +
+                    Hydrological_regime
+                    , data = All_data)
+```
+As a test, and to further address the AE's comments we have collected as many of the data for the total forest are for the larger catchments (the catchments > 1000 km^2^ from @zhang2017).
+This results in 60 data points (as not all forest areas were recoverable from the papers).
+This variable is called *Perc_Farea_pre* in Table \@ref(tab:modelforestarea).
+The results of a model that includes this variable and simply assumesa generalise linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
+
+Again, it is clear that the assessment technique (in this case there are no paired watershed experiments) is significant in explaining the variation in the data.
+
+```{r, echo = F, eval =T}
+pander(round(summary(Forest_model_all)$p.table,2), caption= "(\\#tab:modelforestarea) Statistical overview of the linear components of the model including percent forested area")
+```
+
+```{r, eval = F, echo = F}
+png("residual_plot_model_Farea.png", width = 2500, height = 1800, res = 300)
+par(mfrow=c(2,2))
+gam.check(Forest_model_all)
+par(mfrow=c(1,1))
+dev.off()
+```
+
+Inspecting the residuals of the model (Figure \@ref(fig:gamcheckmodelarea)) also indicates (as we have highlighted earlier) that the data for the larger catchments are much more well-behaved.
+In our opinion, this is because most of the larger catchment studies are hydrological modelling studies, and what is analysed here is the underlying conceptualised structure of the models rather than real responses of catchments to change in the forest cover.
+
+In the manuscript we have not repeated this analysis, but we have included the following text in the discussion (l524 - l537 on page 30):
+
+"One of these latent variables could be the total area of forest in a catchment, as was analysed in @levy2018.
+In this case, the total \% area of forest was not included in the data.
+As a test, the total \% area of forest for the larger catchments (> 1000 km^2^ in @zhang2017) were added to the dataset and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full dataset, it is complicated for two reasons:
+
+1. The area of forest is not always indicated in the original papers, or a range of values is given, complicating the data collection.
+2. Many of the small catchments have 100\% area covered in forest, introducing a strong skew in the data and complicating if total area of forest has an impact on the change in flow."
+
+```{r gamcheckmodelarea, echo =F, fig.cap="Residual plots for the regression model ", out.width="90%"}
+include_graphics("residual_plot_model_Farea.png")
+```
 
@@
 We totally agree with the AE, and this might be more a matter of symantics.
-We thought we had the same interpretation as the AE, but might have worded this correctly.
+We thought we had the same interpretation as the AE, but might have worded this incorrectly.
 The above comment from the AE actually points exactly to the way we are using models in the paper.

### 2022-10-03 paper/Forest_and_Water/Reply_first_review_JoH.tex
resolve comment about percent total forest

@@
 This is without considering the number of errors that existed in the data.
-This is particularly important, since results from these global analyses are used to build further models to analyse global impacts \citep[e.g.][]{hoekvandijke2022}, leading to possible wrong policy or management responses.\\
+This is particularly important, since results from these global analyses are used to build further models to analyse global impacts \citep[e.g.][]{hoekvandijke2022}, leading to possible wrong policy or management responses.
 \item
@@
 This is particularly true for quite a few studies which focused on regeneration of forests after wildfire or clear cutting followed by re-establishment of plantation of native forest.
-Many of the Paired Watershed Experiments in Australia \citep[e.g.][]{cornish2001, webb2009, webb2012, webb2013, watson2001} fall in this category, and therefore easily be classified as either forestation or deforestation.
+Many of the Paired Watershed Experiments \citep[e.g.][]{cornish2001, webb2009, webb2012, webb2013, watson2001} fall in this category, and therefore can easily be classified as either forestation or deforestation.
 In principle a decision needs to be made how many years post clearing needs to be considered, and whether the remainder of the timeseries should be classified as regeneration.
-We demonstrate how removing these experiments from the total changes the results of the analysis.
+Or, as we discuss, the data should be analysed as a timeseries of change, rather than trying to pick a single point.
 \item
@@
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
+Even though length is no significant in the data, does not mean that this effect exists in the data.
 In terms of the catchment size, one of the key issues we originally had with the \citet{zhang2017} paper was the arbitrary split between catchment \textgreater{} 1000 km\textsuperscript{2} and catchment \textless{} 1000 km\textsuperscript{2}.
@@
 
-Overall this analysis shows very clearly the danger of simply investigating single variable regressions as was done in \citet{zhang2017} and \citet{filoso2017} or even using simple modelling as in \citet{jackson2005} and \citet{zhou2015}.
+Overall this analysis shows very clearly the challenges of simply investigating single variable regressions as was done in \citet{zhang2017} and \citet{filoso2017} or even using simple modelling as in \citet{jackson2005} and \citet{zhou2015}.
 It is too easy to miss the latent variables that are the underlying factors and influence the model results.
@@
 
-We agree with the AE that this would be an interesting idea and links back to our point about ``latent variables''.
+We agree with the AE that this was a logical idea and links back to our point about ``latent variables''.
 The excellent paper by \citet{levy2018} includes \% forest area as a variable, but their data cannot be incorporated into the current paper.
 This is because their analysis focuses on a time progression rather than a ``before and after'' or a clear ``control'' and ``treatment''.
-Furthermore, extracting the total area of forest from the papers is not a trivial job, and the area is also not always mentioned in the papers.
-Finally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
-It is definitely an area of future research.
+Regrettably, extracting the total area of forest from the papers is not a trivial job, and also has some issues.
+The total forested area is also not always mentioned in the papers, or only a range is given.
+Additionally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
+
+As a test, and to further address the AE's comments we have collected as many of the data for the total forest are for the larger catchments (the catchments \textgreater{} 1000 km\textsuperscript{2} from \citet{zhang2017}).
+This results in 60 data points (as not all forest areas were recoverable from the papers).
+This variable is called \emph{Perc\_Farea\_pre} in Table \ref{tab:modelforestarea}.
+The results of a model that includes this variable and simply assumesa generalise linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
+
+Again, it is clear that the assessment technique (in this case there are no paired watershed experiments) is significant in explaining the variation in the data.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
+\caption{\label{tab:modelforestarea} Statistical overview of the linear components of the model including percent forested area}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\centering
+~
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Estimate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Std. Error
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+t value
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Pr(\textgreater\textbar t\textbar)
+\end{minipage} \\
+\midrule()
+\endhead
+\textbf{(Intercept)} & -16.05 & 15.08 & -1.06 & 0.3 \\
+\textbf{DeltaF\_perc} & -0.37 & 0.12 & -3.24 & 0 \\
+\textbf{log10(Area\_km2)} & 1.56 & 2.31 & 0.67 & 0.51 \\
+\textbf{Dryness} & 4.4 & 3.22 & 1.37 & 0.18 \\
+\textbf{Perc\_Farea\_pre} & 0.04 & 0.08 & 0.49 & 0.63 \\
+\textbf{Precip\_data\_typeOB} & -19.28 & 6.55 & -2.94 & 0.01 \\
+\textbf{Precip\_data\_typeSG} & -12 & 6.76 & -1.78 & 0.09 \\
+\textbf{Assessment\_techniqueEA, HM} & 1.37 & 9.86 & 0.14 & 0.89 \\
+\textbf{Assessment\_techniqueHM} & 15.44 & 6.17 & 2.5 & 0.02 \\
+\textbf{Assessment\_techniqueQPW} & 2.8 & 9.9 & 0.28 & 0.78 \\
+\textbf{Assessment\_techniqueSH} & 18.51 & 7.65 & 2.42 & 0.02 \\
+\textbf{Forest\_typeCF} & -9.47 & 6.17 & -1.54 & 0.14 \\
+\textbf{Forest\_typeMF} & 0.28 & 4.64 & 0.06 & 0.95 \\
+\textbf{Hydrological\_regimeSD} & 20.98 & 6.66 & 3.15 & 0 \\
+\bottomrule()
+\end{longtable}
+
+Inspecting the residuals of the model (Figure \ref{fig:gamcheckmodelarea}) also indicates (as we have highlighted earlier) that the data for the larger catchments are much more well-behaved.
+In our opinion, this is because most of the larger catchment studies are hydrological modelling studies, and what is analysed here is the underlying conceptualised structure of the models rather than real responses of catchments to change in the forest cover.
+
+In the manuscript we have not repeated this analysis, but we have included the following text in the discussion (l524 - l537 on page 30):
+
+``One of these latent variables could be the total area of forest in a catchment, as was analysed in \citet{levy2018}.
+In this case, the total \% area of forest was not included in the data.
+As a test, the total \% area of forest for the larger catchments (\textgreater{} 1000 km\textsuperscript{2} in \citet{zhang2017}) were added to the dataset and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full dataset, it is complicated for two reasons:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The area of forest is not always indicated in the original papers, or a range of values is given, complicating the data collection.
+\item
+  Many of the small catchments have 100\% area covered in forest, introducing a strong skew in the data and complicating if total area of forest has an impact on the change in flow.''
+\end{enumerate}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{residual_plot_model_Farea} \caption{Residual plots for the regression model }\label{fig:gamcheckmodelarea}
+\end{figure}
 

### 2022-10-03 paper/Forest_and_Water/Forest_and_Water.Rmd
resolve comment about percent total forest

@@
 
-### Test removal of studies of great length and for very dry catchments
+### Removal of studies of great length and for very dry catchments
 
@@
 
-**Possibly insert here a model to investigate total forest area as a random effect**
-
 ```{r, eval =T, echo =F}
@@
 The only  way to analyze changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in @levy2018).
+
+One of these latent variables could be the total area of forest in a catchment, as was analysed in @levy2018.
+In this case, the total \% area of forest was not included in the data.
+As a test, the total \% area of forest for the larger catchments (> 1000 km^2^ in @zhang2017) were added to the dataset and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full dataset, it is complicated for two reasons:
+
+1. The area of forest is not always indicated in the original papers, or a range of values is given, complicating the data collection.
+2. Many of the small catchments have 100\% area covered in forest, introducing a strong skew in the data and complicating if total area of forest has an impact on the change in flow.
 

### 2022-10-03 paper/Forest_and_Water/Forest_and_Water.tex
resolve comment about percent total forest

@@
 
-\hypertarget{test-removal-of-studies-of-great-length-and-for-very-dry-catchments}{%
-\subsubsection{Test removal of studies of great length and for very dry catchments}\label{test-removal-of-studies-of-great-length-and-for-very-dry-catchments}}
+\hypertarget{removal-of-studies-of-great-length-and-for-very-dry-catchments}{%
+\subsubsection{Removal of studies of great length and for very dry catchments}\label{removal-of-studies-of-great-length-and-for-very-dry-catchments}}
 
@@
 
-\textbf{Possibly insert here a model to investigate total forest area as a random effect}
-
 \begin{longtable}[]{@{}
@@
 
+One of these latent variables could be the total area of forest in a catchment, as was analysed in \citet{levy2018}.
+In this case, the total \% area of forest was not included in the data.
+As a test, the total \% area of forest for the larger catchments (\textgreater{} 1000 km\textsuperscript{2} in \citet{zhang2017}) were added to the dataset and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full dataset, it is complicated for two reasons:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The area of forest is not always indicated in the original papers, or a range of values is given, complicating the data collection.
+\item
+  Many of the small catchments have 100\% area covered in forest, introducing a strong skew in the data and complicating if total area of forest has an impact on the change in flow.
+\end{enumerate}
+
 We are not arguing that there is no relationship between streamflow and forest cover, and there might indeed be a global relationship that can be discovered.

### 2022-10-03 paper/Forest_and_Water/Test_model_totalForestArea.Rmd
resolve comment about percent total forest

---
title: "Testing Forest Area as a random effect"
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    footnote: 1
  - name: Eliana Nervi
    email: eliananervif@gmail.com
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
footnote:
  - code: 1
    text: "Corresponding Author"    
date: "`r Sys.Date()`"
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forestandwater.bib
#editor_options: 
#  chunk_output_type: inline
header-includes:
  - \usepackage{setspace}
  - \usepackage{color}
  - \newcommand{\beginsupplement}{  \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}\setcounter{equation}{0} \renewcommand{\theequation}{S\arabic{equation}}}
abstract: |
  This supplementary material file compares different linear and non-linear relationships for the impact of forest cover on the change in stream flow.
---
\beginsupplement

# Introduction

This supplementary material is related to 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'

This document outlines a test suggested in the first review of the manuscript. The AE of the Journal of Hydrology suggested that we test if including the total forest area as a variable would make sense. Not only the \% change in area is important, but what total area of forest was there in the first place.

This test focusses on the large catchment data (< 1000 km^2^) for two reasons:

1. This was a manageable subset and many of the papers are fairly accessible, in contrast to the small catchments.
2. Many of the small catchments have 100\% area covered in forest, so are not useful to identify if total area of forest has an impact on the change in flow.

## set up the packages

```{r setup, include=FALSE, echo =T}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
require(mgcv)
require(visreg)
require(patchwork)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))
```

## read in the data

Read in the data and massage column names to make columns useable in the model.

```{r data-w-forest}

Data_wF <- read_csv("../../data/LCdatawithFA.xlsx - LW_TotForest_Area.csv")

# change some column names
names(Data_wF)[3:4] <- c("Area_km2", "Pa_mm")
names(Data_wF)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Data_wF)[13:14] <- c("Precip_data_type", "Assessment_technique")

```

## Some changes to the overall data

1. calculating the dryness

```{r, echo = T}
# calculate dryness index
Data_wF <- Data_wF %>%
  mutate(Dryness = E0/Pa_mm)
```

2. remove watershed 1 (the Amazon) from the analysis  
```{r, echo = T}
Data_wF <- Data_wF %>%
  filter(`Watershed #` != 1)
```

3. include length as a variable

```{r length, echo =T}
All_data <- Data_wF %>%
  mutate(length = To - From,
         mid_year = From + (To - From)/2)

```

# Results

Run the model simply as a GLM, transform Area using log10, as in the manuscript.

```{r model-percent}

Forest_model_all <- gam(DeltaQf_perc ~ DeltaF_perc + 
                    log10(Area_km2) + 
                    Dryness + 
                    Perc_Farea_pre +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data)
summary(Forest_model_all)
```
This suggest that forest area is not at all significant in the model. In fact, the hydrological modelling technique is much more important. Overall the model explains roughly 82% of the variation.

```{r check-model-percent}
gam.check(Forest_model_all)
#plot(Forest_model_all)
```
The fit of the model is also well-behaved, as would be expected. The data are mostly from hydrological modelling, stabilising the response of the change in forest to change in streamflow.

## run the model on forest area rather than \%

Possibly, using the total \% forest is not taking into account the differences in total area of forest. So we can also test with total ha. Note that this can interact with total area: larger catchments will have larger total areas of forest.

We appear to have less data here as \% is reported more, but we can calculate total ha by multiplying \% by total area.

```{r model-area}
All_data <- Data_wF %>%
  mutate(Farea_km2 = ifelse(is.na(Farea_km2)==T,Area_km2*Perc_Farea_pre/100,Farea_km2))

#write_csv(All_data,"../../data/Data_TotForest_Area.csv")

Forest_model_area <- gam(DeltaQf_perc ~ DeltaF_perc + 
                    log10(Area_km2) + 
                    Dryness + 
                    Farea_km2 +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime
                    , data = All_data)
summary(Forest_model_area)
```
```{r check_resid_area}
gam.check(Forest_model_area)
#plot(Forest_model_all)
```

Again, not significant and the model is well-behaved. Note that the Hydrological modelling assessment technique (HM) is again highly significant.

# Conclusion

There is no effect of forest area in the relationship between change in flow  and percent change in forest cover.

### 2022-10-03 paper/Forest_and_Water/Test_model_totalForestArea.tex
resolve comment about percent total forest

--- 2022-09-28 paper/Forest_and_Water/SupplementaryMaterialPart3.tex
@@
             pdfauthor={},
-            pdftitle={Supplementary Information part 3: Testing different forest cover relationships},
+            pdftitle={Testing Forest Area as a random effect},
             colorlinks=false,
@@
 
-  \title{Supplementary Information part 3: Testing different forest cover relationships}
+  \title{Testing Forest Area as a random effect}
     \author[]{R. Willem Vervoort%
@@
 
-In this document we tested the following, which is outlined in the methods of the paper.
-The changes in forest cover contain both positive (forestation) and negative values (deforestation).
-In \citet{zhang2017}, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.
-Here we test two variation of this approach, leading to three approaches:
+This document outlines a test suggested in the first review of the manuscript.
+The AE of the Journal of Hydrology suggested that we test if including the total forest area as a variable would make sense.
+Not only the \% change in area is important, but what total area of forest was there in the first place.
+
+This test focusses on the large catchment data (\textless{} 1000 km\textsuperscript{2}) for two reasons:
 
@@
 \item
-  Joint analysis and assuming a linear relationship between the change in streamflow and the change in forest cover, following \citet{zhang2017}.
+  This was a manageable subset and many of the papers are fairly accessible, in contrast to the small catchments.
 \item
-  The impact of an increase in forest cover can be different from the same fractional decrease in forest cover.
-The question becomes how best to analyse this.
-One approach would be to allow a different slope and a different intercept for the decreases relative to the increases.
-\item
-  A second approach is to test the change in forest cover as a non-linear relationship in the Generalised additive model (GAM).
+  Many of the small catchments have 100\% area covered in forest, so are not useful to identify if total area of forest has an impact on the change in flow.
 \end{enumerate}
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\hypertarget{set-up-the-packages}{%
+\subsection{set up the packages}\label{set-up-the-packages}}
 
-\hypertarget{approach-1}{%
-\subsection{Approach 1}\label{approach-1}}
+\hypertarget{read-in-the-data}{%
+\subsection{read in the data}\label{read-in-the-data}}
 
-To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) \citep{wood2006}.
-
-The general model tested is:
-
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover + \notag \\ 
-& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq1}
-\end{align}
-
-\hypertarget{approach-2}{%
-\subsection{Approach 2}\label{approach-2}}
-
-Deforestation is different from reforestation.
-This can be tested by converting all the change in forest cover data to positive values, and an additional binary column (\(sign_{forest cover}\)) can be included indicating whether it was a forest cover increase or decrease.
-In the model, the parameter for \(sign_{forest cover}\) will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover.
-The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).
-
-This results in the following model:
-
-\begin{align}
-\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest cover} + \notag \\ 
-& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq2}
-\end{align}
-
-\hypertarget{approach-3}{%
-\subsection{Approach 3}\label{approach-3}}
-
-The relationship between the change in forest cover and change in streamflow is non-linear.
-The non-linear approach involves adding a smoothing function the the change in forestry.
-A shrinkage spline \citep{wood2006} is applied to allow the non-linear effect to be removed.
-Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous.
-The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form.
-This leads to the following model:
-
-\begin{align}
-\Delta Qf \% \sim &~ s(\Delta \% forest~cover) + \notag \\ 
-& \sum{X_i} + \sum{s(Z_i)} + \varepsilon \label{eq:eq3}
-\end{align}
-
-\hypertarget{preliminary-data-management}{%
-\subsection{Preliminary data management}\label{preliminary-data-management}}
-
-First we will read in the data
-
-We will combine the different tables, but will keep an indicator to see where the data are from.
+Read in the data and massage column names to make columns useable in the model.
 
@@
 \begin{Highlighting}[]
-\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{From }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{From)}
-\NormalTok{Zhang\_small}\SpecialCharTok{$}\NormalTok{To }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(Zhang\_small}\SpecialCharTok{$}\NormalTok{To)}
-\NormalTok{Zhang\_all }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_large,Zhang\_small) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"original Zhang et al data"}\NormalTok{)}
-\NormalTok{new\_data }\OtherTok{\textless{}{-}}\NormalTok{ new\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dataset =} \StringTok{"new data"}\NormalTok{)}
-\NormalTok{All\_data }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(Zhang\_all, new\_data)}
+\NormalTok{Data\_wF }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"../../data/LCdatawithFA.xlsx {-} LW\_TotForest\_Area.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 
-\hypertarget{implementing-the-changes-to-the-overall-data}{%
-\subsection{Implementing the changes to the overall data}\label{implementing-the-changes-to-the-overall-data}}
+\begin{verbatim}
+## Rows: 61 Columns: 25
+## -- Column specification --------------------------------------------------------
+## Delimiter: ","
+## chr (10): Watershed name, Forest type, Hydrological regime, yearFarea, Comme...
+## dbl (15): Watershed #, Area(km2), Pa(mm), Farea_km2, Perc_Farea_pre, DeltaF_...
+## 
+## i Use `spec()` to retrieve the full column specification for this data.
+## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+\end{verbatim}
 
-The following code implements the changes described in the Supplementary data part 1.
-However, many of the changes were implemented manually into the data set.
-These are simply the remaining changes not implemented manually.
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# change some column names}
+\FunctionTok{names}\NormalTok{(Data\_wF)[}\DecValTok{3}\SpecialCharTok{:}\DecValTok{4}\NormalTok{] }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Area\_km2"}\NormalTok{, }\StringTok{"Pa\_mm"}\NormalTok{)}
+\FunctionTok{names}\NormalTok{(Data\_wF)[}\DecValTok{5}\SpecialCharTok{:}\DecValTok{6}\NormalTok{] }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Forest\_type"}\NormalTok{, }\StringTok{"Hydrological\_regime"}\NormalTok{)}
+\FunctionTok{names}\NormalTok{(Data\_wF)[}\DecValTok{13}\SpecialCharTok{:}\DecValTok{14}\NormalTok{] }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Precip\_data\_type"}\NormalTok{, }\StringTok{"Assessment\_technique"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{some-changes-to-the-overall-data}{%
+\subsection{Some changes to the overall data}\label{some-changes-to-the-overall-data}}
 
@@
 \item
-  removing the duplicates.
+  calculating the dryness
 \end{enumerate}
@@
 \begin{Highlighting}[]
-\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} 
-           \FunctionTok{ifelse}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)}\SpecialCharTok{==}\NormalTok{T,}\DecValTok{0}\NormalTok{,}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{),}
-         \StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{as.numeric}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Possible duplicate}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
+\CommentTok{\# calculate dryness index}
+\NormalTok{Data\_wF }\OtherTok{\textless{}{-}}\NormalTok{ Data\_wF }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Dryness =}\NormalTok{ E0}\SpecialCharTok{/}\NormalTok{Pa\_mm)}
 \end{Highlighting}
@@
 \item
-  calculating the dryness
+  remove watershed 1 (the Amazon) from the analysis
 \end{enumerate}
@@
 \begin{Highlighting}[]
-\CommentTok{\# calculate dryness index}
-\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Dryness =}\NormalTok{ E0}\SpecialCharTok{/}\NormalTok{Pa\_mm)}
+\NormalTok{Data\_wF }\OtherTok{\textless{}{-}}\NormalTok{ Data\_wF }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
 \end{Highlighting}
@@
 \item
-  remove watershed 1 (the Amazon) from the analysis
+  include length as a variable
 \end{enumerate}
@@
 \begin{Highlighting}[]
-\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
-\item
-  remove data set 188 and 254 Kamakia and Sambret
-\end{enumerate}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{188}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Watershed \#}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \DecValTok{254}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{4}
-\tightlist
-\item
-  add a column that indicates forst loss of forest gain
-\end{enumerate}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{forest\_sign =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{, }\StringTok{"Forest Cover Loss"}\NormalTok{,}
-                              \StringTok{"Forest Cover Gain"}\NormalTok{))}
+\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ Data\_wF }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{length =}\NormalTok{ To }\SpecialCharTok{{-}}\NormalTok{ From,}
+         \AttributeTok{mid\_year =}\NormalTok{ From }\SpecialCharTok{+}\NormalTok{ (To }\SpecialCharTok{{-}}\NormalTok{ From)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
 \end{Highlighting}
@@
 
-\hypertarget{approach-1-linear-relationship-across-all-changes-in-forest-cover}{%
-\subsection{Approach 1: linear relationship across all changes in forest cover}\label{approach-1-linear-relationship-across-all-changes-in-forest-cover}}
-
-First the variable \texttt{length} is added, representing the number of years that the data represents.
+Run the model simply as a GLM, transform Area using log10, as in the manuscript.
 
@@
 \begin{Highlighting}[]
-\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Forest\_Sign =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{,}
-                              \StringTok{"Decrease"}\NormalTok{, }\StringTok{"Increase"}\NormalTok{),}
-         \AttributeTok{DeltaF\_perc\_pos =} \FunctionTok{ifelse}\NormalTok{(DeltaF\_perc }\SpecialCharTok{\textless{}} \DecValTok{0}\NormalTok{,}
-                                  \SpecialCharTok{{-}}\DecValTok{1}\SpecialCharTok{*}\NormalTok{DeltaF\_perc,}
-\NormalTok{                                  DeltaF\_perc))}
-\NormalTok{All\_data2 }\OtherTok{\textless{}{-}}\NormalTok{ All\_data2 }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{length =}\NormalTok{ To }\SpecialCharTok{{-}}\NormalTok{ From,}
-         \AttributeTok{mid\_year =}\NormalTok{ From }\SpecialCharTok{+}\NormalTok{ (To }\SpecialCharTok{{-}}\NormalTok{ From)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
-
 \NormalTok{Forest\_model\_all }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ DeltaF\_perc }\SpecialCharTok{+} 
-                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
-                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
+                    \FunctionTok{log10}\NormalTok{(Area\_km2) }\SpecialCharTok{+} 
+\NormalTok{                    Dryness }\SpecialCharTok{+} 
+\NormalTok{                    Perc\_Farea\_pre }\SpecialCharTok{+}
 \NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
@@
 \NormalTok{                    Hydrological\_regime}
-\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
-\CommentTok{\#summary(Forest\_model\_all)}
-\CommentTok{\#gam.check(Forest\_model\_all)}
-\CommentTok{\#plot(model6\_all)}
+\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data)}
+\FunctionTok{summary}\NormalTok{(Forest\_model\_all)}
 \end{Highlighting}
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
-\caption{(\#tab:m\_all-linear) Statistical summary for the linear terms the full model}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & -4.81 & 16.31 & -0.29 & 0.77 \\
-\textbf{DeltaF\_perc} & -0.6 & 0.06 & -10.71 & 0 \\
-\textbf{Precip\_data\_typeOB} & -21.4 & 13.23 & -1.62 & 0.11 \\
-\textbf{Precip\_data\_typeSG} & 9.36 & 15.17 & 0.62 & 0.54 \\
-\textbf{Assessment\_techniqueEA, HM} & 20.64 & 42.73 & 0.48 & 0.63 \\
-\textbf{Assessment\_techniqueHM} & 22.81 & 11.71 & 1.95 & 0.05 \\
-\textbf{Assessment\_techniquePWE} & 30.63 & 11.94 & 2.57 & 0.01 \\
-\textbf{Assessment\_techniquePWE,
-HM} & 17.42 & 43.26 & 0.4 & 0.69 \\
-\textbf{Assessment\_techniqueQPW} & 39.52 & 20.15 & 1.96 & 0.05 \\
-\textbf{Assessment\_techniqueQPW,
-EA} & 24.39 & 24.42 & 1 & 0.32 \\
-\textbf{Assessment\_techniqueSH} & 45.3 & 11.83 & 3.83 & 0 \\
-\textbf{Forest\_typeCF} & -9.45 & 7.6 & -1.24 & 0.21 \\
-\textbf{Forest\_typeMF} & -8.05 & 7.56 & -1.06 & 0.29 \\
-\textbf{Hydrological\_regimeSD} & 3.57 & 9.16 & 0.39 & 0.7 \\
-\bottomrule()
-\end{longtable}
+\begin{verbatim}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc + log10(Area_km2) + Dryness + Perc_Farea_pre + 
+##     Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                             Estimate Std. Error t value Pr(>|t|)   
+## (Intercept)                -16.04690   15.08196  -1.064  0.29676   
+## DeltaF_perc                 -0.37440    0.11561  -3.239  0.00318 **
+## log10(Area_km2)              1.55899    2.31470   0.674  0.50634   
+## Dryness                      4.40194    3.21852   1.368  0.18269   
+## Perc_Farea_pre               0.03722    0.07544   0.493  0.62572   
+## Precip_data_typeOB         -19.28048    6.55235  -2.943  0.00661 **
+## Precip_data_typeSG         -12.00304    6.75886  -1.776  0.08702 . 
+## Assessment_techniqueEA, HM   1.36550    9.86348   0.138  0.89092   
+## Assessment_techniqueHM      15.44379    6.16998   2.503  0.01866 * 
+## Assessment_techniqueQPW      2.79556    9.90330   0.282  0.77988   
+## Assessment_techniqueSH      18.51105    7.64697   2.421  0.02249 * 
+## Forest_typeCF               -9.46812    6.16583  -1.536  0.13628   
+## Forest_typeMF                0.27749    4.64399   0.060  0.95279   
+## Hydrological_regimeSD       20.97769    6.66044   3.150  0.00397 **
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.735   Deviance explained = 82.1%
+## GCV = 128.37  Scale est. = 84.534    n = 41
+\end{verbatim}
 
-This is the baseline result as it echoes \citet{zhang2017} but in this case the relationship is tested together with other variables as is discussed in the main paper.
-
-For this model, the focus can be on the linear variables.
-The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0.
-The overall model explains 49 \% of the variation in the data.
-For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used.
-The AIC of the model is 3181.7600577.
-
-\hypertarget{the-change-in-forest-cover-converted-to-positive-values}{%
-\subsection{The change in forest cover converted to positive values}\label{the-change-in-forest-cover-converted-to-positive-values}}
+This suggest that forest area is not at all significant in the model. In fact, the hydrological modelling technique is much more important. Overall the model explains roughly 82\% of the variation.
 
@@
 \begin{Highlighting}[]
-\NormalTok{Forest\_model\_all2 }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ DeltaF\_perc\_pos }\SpecialCharTok{+}\NormalTok{ Forest\_Sign }\SpecialCharTok{+}
-                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
-                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
-\NormalTok{                    Forest\_type }\SpecialCharTok{+}
-\NormalTok{                    Hydrological\_regime}
-\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
-\CommentTok{\#summary(Forest\_model\_all2)}
+\FunctionTok{gam.check}\NormalTok{(Forest\_model\_all)}
 \end{Highlighting}
@@
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4231}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1282}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1410}}@{}}
-\caption{(\#tab:m\_all2-linear) Statistical summary for the linear terms the alternative model where the change is forestry is all positive}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Estimate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Std. Error
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-t value
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Pr(\textgreater\textbar t\textbar)
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{(Intercept)} & 14.05 & 20 & 0.7 & 0.48 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.9 & 0 \\
-\textbf{Forest\_SignIncrease} & -51.91 & 7.79 & -6.67 & 0 \\
-\textbf{Precip\_data\_typeOB} & -19.43 & 14.56 & -1.33 & 0.18 \\
-\textbf{Precip\_data\_typeSG} & 4.43 & 17.22 & 0.26 & 0.8 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.4 & 46.56 & 0.4 & 0.69 \\
-\textbf{Assessment\_techniqueHM} & 21.99 & 13.63 & 1.61 & 0.11 \\
-\textbf{Assessment\_techniquePWE} & 27.8 & 13.85 & 2.01 & 0.05 \\
-\textbf{Assessment\_techniquePWE,
-HM} & 17.58 & 46.6 & 0.38 & 0.71 \\
-\textbf{Assessment\_techniqueQPW} & 25.71 & 22.18 & 1.16 & 0.25 \\
-\textbf{Assessment\_techniqueQPW,
-EA} & 33.3 & 26.36 & 1.26 & 0.21 \\
-\textbf{Assessment\_techniqueSH} & 38.7 & 13.25 & 2.92 & 0 \\
-\textbf{Forest\_typeCF} & -12.71 & 8.25 & -1.54 & 0.12 \\
-\textbf{Forest\_typeMF} & -7.05 & 8.36 & -0.84 & 0.4 \\
-\textbf{Hydrological\_regimeSD} & 7.27 & 10 & 0.73 & 0.47 \\
-\bottomrule()
-\end{longtable}
-
-Again the focus can be on the linear variables.
-In this case the variable \texttt{DeltaF\_perc\_pos} also is important to explain the variation in the change in flow with a p-value of 0.
-However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0).
-In fact, in absolute terms deforestration creates a -5191 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).
-
-However, overall this model explains less of the overall variation in the data with a deviance explained of 0.45 and an AIC of 3227.3976077.
-
-\hypertarget{assuming-the-change-in-forest-cover-is-non-linear}{%
-\subsection{Assuming the change in forest cover is non-linear}\label{assuming-the-change-in-forest-cover-is-non-linear}}
-
-The difference between reforestation and deforestation might suggest that the relationship between the change in forest cover and change in stream flow is non-linear.
-This is tested in the model where the change in forest cover is considered non-linear.
-The number of knots was decided based on checking the edf value relative to the apparent knots k' wiht the procedure \texttt{gam.check()} in the package \texttt{mgcv} \citep{wood2006}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{Forest\_model\_nonlin }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{s}\NormalTok{(DeltaF\_perc, }\AttributeTok{k=}\DecValTok{10}\NormalTok{, }\AttributeTok{bs =} \StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-                    \FunctionTok{s}\NormalTok{(}\FunctionTok{log10}\NormalTok{(Area\_km2), }\AttributeTok{k =} \DecValTok{5}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-                    \FunctionTok{s}\NormalTok{(Dryness, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{ ) }\SpecialCharTok{+} 
-                     \FunctionTok{s}\NormalTok{(length, }\AttributeTok{k =} \DecValTok{35}\NormalTok{, }\AttributeTok{bs=}\StringTok{"ts"}\NormalTok{) }\SpecialCharTok{+}
-\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
-\NormalTok{                    Forest\_type }\SpecialCharTok{+}
-\NormalTok{                    Hydrological\_regime}
-\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data2)}
-\CommentTok{\#summary(Forest\_model\_nonlin)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}@{}}
-\caption{(\#tab:m\_nonlin-smooth) Statistical summary for the smooth terms for the model that considers the change in forest cover to be a non-linear variable}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\centering
-~
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Ref.df
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-F
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-p-value
-\end{minipage} \\
-\midrule()
-\endhead
-\textbf{s(DeltaF\_perc)} & 4.14 & 9 & 13.2 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.74 & 4 & 0.74 & 0.04 \\
-\textbf{s(Dryness)} & 4.73 & 9 & 2.4 & 0 \\
-\textbf{s(length)} & 4.68 & 34 & 0.26 & 0.08 \\
-\bottomrule()
-\end{longtable}
-
-In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0.
-The total deviance explained for the model is 0.5 and the model has an AIC of 3182.1088731.
-These performance statistics are similar to the first (baseline) model.
-
-One thing that is of interest is the fact that while a non-linear relationship between the change in forest cover and the change in streamflow was conceptualised, the actual relationship turns out to be linear.
-This can be shown from the plot and by running \texttt{gam.check}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{D }\OtherTok{\textless{}{-}} \FunctionTok{visreg}\NormalTok{(Forest\_model\_nonlin, }\StringTok{"DeltaF\_perc"}\NormalTok{, }\AttributeTok{gg =}\NormalTok{ T, }\AttributeTok{plot =}\NormalTok{ T) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{()}
-\NormalTok{A }\OtherTok{\textless{}{-}} \FunctionTok{visreg}\NormalTok{(Forest\_model\_nonlin, }\StringTok{"Area\_km2"}\NormalTok{, }\AttributeTok{gg =}\NormalTok{ T, }\AttributeTok{plot =}\NormalTok{ T) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{scale\_x\_log10}\NormalTok{()}
-
-\NormalTok{(A }\SpecialCharTok{+}\NormalTok{ D)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{SupplementaryMaterialPart3_files/figure-latex/unnamed-chunk-10-1.pdf}
-The procedure \texttt{gam.check} highlights that the p-value of the Basis dimension check of the smooth on \texttt{DeltaF\_perc} is very small and the actual edf is close to 1 suggesting a linear relationship, similar to the smooth on log10(\texttt{Area\_km2}).
-
-The residual analysis shows that including this non-linearity makes no difference in terms of he distribution of the residuals.
-While there is some remaining weak trend in the data, the overall distribution of the residuals is approximately normal.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{par}\NormalTok{(}\AttributeTok{mfrow=} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\FunctionTok{gam.check}\NormalTok{(Forest\_model\_nonlin)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{SupplementaryMaterialPart3_files/figure-latex/unnamed-chunk-11-1.pdf}
+\includegraphics{Test_model_totalForestArea_files/figure-latex/check-model-percent-1.pdf}
 
@@
 ## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 11 iterations.
-## The RMS GCV score gradient at convergence was 0.0003791486 .
-## The Hessian was positive definite.
-## Model rank =  69 / 69 
+## Model required no smoothing parameter selectionModel rank =  14 / 14
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#plot(Forest\_model\_all)}
+\end{Highlighting}
+\end{Shaded}
+
+The fit of the model is also well-behaved, as would be expected. The data are mostly from hydrological modelling, stabilising the response of the change in forest to change in streamflow.
+
+\hypertarget{run-the-model-on-forest-area-rather-than}{%
+\subsection{run the model on forest area rather than \%}\label{run-the-model-on-forest-area-rather-than}}
+
+Possibly, using the total \% forest is not taking into account the differences in total area of forest.
+So we can also test with total ha.
+Note that this can interact with total area: larger catchments will have larger total areas of forest.
+
+We appear to have less data here as \% is reported more, but we can calculate total ha by multiplying \% by total area.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{All\_data }\OtherTok{\textless{}{-}}\NormalTok{ Data\_wF }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Farea\_km2 =} \FunctionTok{ifelse}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(Farea\_km2)}\SpecialCharTok{==}\NormalTok{T,Area\_km2}\SpecialCharTok{*}\NormalTok{Perc\_Farea\_pre}\SpecialCharTok{/}\DecValTok{100}\NormalTok{,Farea\_km2))}
+
+\NormalTok{Forest\_model\_area }\OtherTok{\textless{}{-}} \FunctionTok{gam}\NormalTok{(DeltaQf\_perc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ DeltaF\_perc }\SpecialCharTok{+} 
+                    \FunctionTok{log10}\NormalTok{(Area\_km2) }\SpecialCharTok{+} 
+\NormalTok{                    Dryness }\SpecialCharTok{+} 
+\NormalTok{                    Farea\_km2 }\SpecialCharTok{+}
+\NormalTok{                    Precip\_data\_type }\SpecialCharTok{+}\NormalTok{  Assessment\_technique }\SpecialCharTok{+}
+\NormalTok{                    Forest\_type }\SpecialCharTok{+}
+\NormalTok{                    Hydrological\_regime}
+\NormalTok{                    , }\AttributeTok{data =}\NormalTok{ All\_data)}
+\FunctionTok{summary}\NormalTok{(Forest\_model\_area)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
 ## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
+## Family: gaussian 
+## Link function: identity 
 ## 
-##                       k'   edf k-index p-value  
-## s(DeltaF_perc)      9.00  4.14    0.91   0.075 .
-## s(log10(Area_km2))  4.00  0.74    0.95   0.140  
-## s(Dryness)          9.00  4.73    0.93   0.135  
-## s(length)          34.00  4.68    0.93   0.110  
+## Formula:
+## DeltaQf_perc ~ DeltaF_perc + log10(Area_km2) + Dryness + Farea_km2 + 
+##     Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime
+## 
+## Parametric coefficients:
+##                              Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)                -3.021e+01  1.641e+01  -1.841 0.076621 .  
+## DeltaF_perc                -3.718e-01  9.294e-02  -4.000 0.000442 ***
+## log10(Area_km2)             6.084e+00  3.346e+00   1.818 0.080140 .  
+## Dryness                     3.255e+00  3.156e+00   1.031 0.311481    
+## Farea_km2                  -6.197e-05  3.922e-05  -1.580 0.125688    
+## Precip_data_typeOB         -1.740e+01  6.224e+00  -2.795 0.009431 ** 
+## Precip_data_typeSG         -8.667e+00  6.680e+00  -1.297 0.205467    
+## Assessment_techniqueEA, HM  4.465e-01  9.411e+00   0.047 0.962513    
+## Assessment_techniqueHM      1.436e+01  5.973e+00   2.404 0.023345 *  
+## Assessment_techniqueQPW    -6.118e-01  9.487e+00  -0.064 0.949055    
+## Assessment_techniqueSH      1.741e+01  7.243e+00   2.404 0.023362 *  
+## Forest_typeCF              -8.111e+00  5.993e+00  -1.353 0.187195    
+## Forest_typeMF              -4.063e-01  4.433e+00  -0.092 0.927641    
+## Hydrological_regimeSD       2.346e+01  6.605e+00   3.552 0.001428 ** 
 ## ---
 ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## 
+## R-sq.(adj) =  0.756   Deviance explained = 83.5%
+## GCV = 118.56  Scale est. = 78.075    n = 41
 \end{verbatim}
 
-\renewcommand\refname{References}
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{gam.check}\NormalTok{(Forest\_model\_area)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Test_model_totalForestArea_files/figure-latex/check_resid_area-1.pdf}
+
+\begin{verbatim}
+## 
+## Method: GCV   Optimizer: magic
+## Model required no smoothing parameter selectionModel rank =  14 / 14
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#plot(Forest\_model\_all)}
+\end{Highlighting}
+\end{Shaded}
+
+Again, not significant and the model is well-behaved. Note that the Hydrological modelling assessment technique (HM) is again highly significant.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+There is no effect of forest area in the relationship between change in flow and percent change in forest cover.
+
 \bibliography{forestandwater.bib}

### 2022-10-07 paper/Forest_and_Water/Reply_first_review_JoH.tex
version ready for resubmission

@@
 The reviewer comments are ordered with the Editor in Chief comments listed first and our responses next.
-This is followed by the comments from the associate editor and our responsed.
+This is followed by the comments from the associate editor and our responses.
 The reviewer comments are in \textcolor{blue}{blue} and our responses in normal text.
@@
 
-We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
+We agree that our statistics was too expansive and not very streamlined.
+However, we politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE.
 If the statistical modelling was aimed at developing the best predictive model, then this would be the right approach.
@@
 Thank you for the suggestion of the \citet{levy2018} paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic land use change.
-This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
+This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts streamflow.
+The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
 Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use \citet{levy2018} as an example of how the analysis can be done well.
@@
 However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalization of such studies (as attempted in \citet{zhou2015}; \citet{jackson2005}; \citet{filoso2017} and \citet{zhang2017}).
-This is now extensively discussed in the discussion of the paper.
+This is now extensively covered in the discussion of the paper.
 
@@
 
-\textbf{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}
+\textbf{Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple.}
 
 The key contribution of this paper is to highlight the knowledge gap that exists in the extrapolation of local studies to effects at the global scale.
-While the impact of forest cover on streamflow is easily hypothesised \citep[e.g.][]{zhou2015, hoekvandijke2022}, our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+While the impact of forest cover on streamflow is easily hypothesised \citep[e.g.][]{zhou2015, hoekvandijke2022} and measured at local scales, our research clearly shows that the global extrapolation of the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
 In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of forestation or deforestation, and in many cases this becomes a qualitative assessment.\\
@@
   While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in \citet{zhang2017}, \citet{filoso2017}, \citet{zhou2015} and \citet{jackson2005}) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
-In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
+In many cases statistical assumptions are violated and latent variables can hide or strengthen assumed relationships.
 In addition, the equilibirum analysis is based on the assumption of water balance closure, which might not always be the case in arid and semi-arid climates.
@@
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
-Even though length is no significant in the data, does not mean that this effect exists in the data.
+Even though length is not significant in the data, this does not mean that this effect does not exists in the data.
   In terms of the catchment size, one of the key issues we originally had with the \citet{zhang2017} paper was the arbitrary split between catchment \textgreater{} 1000 km\textsuperscript{2} and catchment \textless{} 1000 km\textsuperscript{2}.
-Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of methods used to analyse small catchments (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
+Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of assessment method reported for the smaller catchment studies (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
 The paper by \citet{beck2013} is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
@@
 Thank you for raising these important points in relation to the validity of the statistical model.
-A first point that arises from this is that we clarify better what the aim of our statistical modelling is.\\
+A first point that arises from this is that we improve the explanation of the aim of our statistical modelling.\\
 There are in essence two approaches to statistical modelling.
@@
 Automatic variable selection and potentially validation on independent data are therefore important, as the aim is to develop the most robust model for prediction.\\
-However, a second reason for to use a statistical model is to explain the maximum variance in the data.
-In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
+However, a second reason to use a statistical model is to explain the maximum variance in the data.
+In this case, it is important to develop an a-priori hypothesis about the causal relationships in the data.
 This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in \citet{zhang2017}) or jointly (as in our approach).
-In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from adding each variable.\\
-This explains why we build the model starting from the most simple model, rather than starting from the most complex model.
+In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from each variable.\\
+This explains why we originally built the model starting from the most simple model, rather than starting from the most complex model.
 
@@
 
-In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, or explaining why we did not take any further steps.
+In our case, we clearly indicate the steps we have taken to improve the quality of the regressions, such as transforming some of the variables, and explaining why we did not take any further steps.
 Furthermore, the residual distributions are mostly well-behaved, it is only in the tails of the distribution (very high flow changes and very small flow changes) where the residual distribution diverges from normal.
@@
 
-However, another careful review of the data identified many further errors in the data, which were all originally in \citet{zhang2017}.
+However, another careful review of the data identified many problematic values in the data, which were all originally in \citet{zhang2017}.
 A particular problem was that many catchments had the wrong sign for the change in forest cover.
@@
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
-So value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%.
-corrected to 16\%
+So the value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%.
+Corrected to 16\%
 \item
   3, Baker Creek, \citet{zhangwei2012}.
-The original recorde 201.1\% change by \citet{zhang2017} is also wrong.
+The original recorded 201.1\% change by \citet{zhang2017} is also wrong.
 Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
-corrected.
+This is now corrected.
 \item
@@
   85 Bosboukloof.
-This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
+This is esssentially a duplicate of catchment 184, but the cited paper analyses only 1 year of runoff after a major fire.
 In any case, the data should reflect a decrease in forest cover: changed the sign of Delta\_f\_perc to -80\%.
@@
 \item
-  a formalised model selection;
+  A formalised model selection;
 \item
-  the use of AIC rather than \emph{r}\textsuperscript{2} for model selection; and
+  The use of AIC rather than \emph{r}\textsuperscript{2} for model selection; and
 \item
-  interaction terms.
+  Interaction terms.
 \end{itemize}
@@
 While there clearly is cross correlation between the variables and there is potential interaction between terms, the question of clear causality remains unanswered.
-We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.
+We therefore believed that including these interaction terms in the model brought us back to the original point we were trying to make: we need to be careful in simply applying models to global data and assuming relationships that might be clouded by latent variables.As a result we have now dropped the interaction term analysis in the paper.
 
@@
 In particular the paper now focuses more directly on the difficulties in analysing aggregated global data and the importance of latent variables.
-We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this has significant challenges and can be easily lead to questionable results.
+We point out that while global databases seem to be a great opportunity to understand global trends and interactions, we show that this has significant challenges and can easily lead to questionable results.
 
@@
 
-Thank you for pointing this out, we have reviewed the paper carefully and corrected these minor issues.
+Thank you for pointing this out, we have reviewed the paper carefully and hopefully corrected these minor issues.
 
@@
 Regrettably, extracting the total area of forest from the papers is not a trivial job, and also has some issues.
-The total forested area is also not always mentioned in the papers, or only a range is given.
+Some papers actually don't provide the total forested area, or only a range is given.
 Additionally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
 
-As a test, and to further address the AE's comments we have collected as many of the data for the total forest are for the larger catchments (the catchments \textgreater{} 1000 km\textsuperscript{2} from \citet{zhang2017}).
+As a test, and to further address the AE's comments we have collected as many of the data for the total forest area for the larger catchments (the catchments \textgreater{} 1000 km\textsuperscript{2} from \citet{zhang2017}).
 This results in 60 data points (as not all forest areas were recoverable from the papers).
 This variable is called \emph{Perc\_Farea\_pre} in Table \ref{tab:modelforestarea}.
-The results of a model that includes this variable and simply assumesa generalise linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
+The results of a model that includes this variable and simply assumes a generalized linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
 
 Again, it is clear that the assessment technique (in this case there are no paired watershed experiments) is significant in explaining the variation in the data.
+Not surprisingly this is the Hydrological modelling technique.
 
@@
 As a test, the total \% area of forest for the larger catchments (\textgreater{} 1000 km\textsuperscript{2} in \citet{zhang2017}) were added to the dataset and the model for just the large catchments was tested.
-This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+This showed that the \% area of forest was not significant to explain the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
 While this might be an area of further research on the full dataset, it is complicated for two reasons:
@@
 We use the models to look at factors that significantly affect streamflow rather than looking at a predictive model.
-We have reviewed the text in the conclusion and have reworded to make sure we capture the above suggested meaning.
+We have reworded the text in the conclusion and have made sure we capture the above suggested meaning.
 

### 2022-10-07 paper/Forest_and_Water/SupplementaryMaterialPart1.tex
version ready for resubmission

@@
 
-This supplementary material is related to `Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
+This supplementary material is related to `Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
 
@@
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
-So value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%.
-corrected to 16\%.
+So the value should be \(0.64/0.55*100 - 100\) or \(0.5/044*100 - 100\): 16\% or 13\%.
+Corrected to 16\%.
 \item

### 2022-10-07 paper/Forest_and_Water/Reply_first_review_JoH.Rmd

@@
 Thank you for the suggestion of the @levy2018 paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic land use change.
-This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
+This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts streamflow.
+The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data.
 Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as an example of how the analysis can be done well.
@@
 However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalization of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017).
-This is now extensively discussed in the discussion of the paper.
+This is now extensively covered in the discussion of the paper.
 
@@
 
-**Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.**
+**Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple.**
 
@@
 While the impact of forest cover on streamflow is easily hypothesised [e.g.
-@zhou2015; @hoekvandijke2022], our research clearly shows that the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
+@zhou2015; @hoekvandijke2022] and measured at local scales, our research clearly shows that the global extrapolation of the causal relationship between change in forest cover and streamflow is complex and not as straight forward as shown in earlier literature.
 In addition, to this we highlight that it is very difficult to reinterpret older studies to isolate the effect of  forestation or deforestation, and in many cases this becomes a qualitative assessment.
@@
 - While analysing global databases can be interesting, we need to be careful with drawing major conclusions (as in @zhang2017, @filoso2017, @zhou2015 and @jackson2005) based on basic regression analysis or using equilibrium analysis (such as the Fu model).
-In many cases statistical assumptions are violated and confounding factors can hide or strengthen assumed relationships.
+In many cases statistical assumptions are violated and latent variables can hide or strengthen assumed relationships.
 In addition, the equilibirum analysis is based on the assumption of water balance closure, which might not always be the case in arid and semi-arid climates.
@@
 The length of the study relates to the last point, in all cases, there is large change in the streamflow in the first year, but this effect decreases with the length of the study due to either natural regeneration or some sort of other management, such as replanting.
-Even though length is no significant in the data, does not mean that this effect exists in the data.
+Even though length is not significant in the data, this does not mean that this effect does not exists in the data.
 In terms of the catchment size, one of the key issues we originally had with the @zhang2017 paper was the arbitrary split between catchment > 1000 km^2^ and catchment < 1000 km^2^.
-Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of methods used to analyse small catchments (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
+Our analysis demonstrates that there is no indication of a distinct split, but that, more importantly there is a distinct difference in the type of assessment method reported for the smaller catchment studies (mostly direct observation and paired catchment analysis) and large catchments (mostly hydrological modelling or some sort of statistical modelling).
 The paper by @beck2013 is an exception, focussing on hydrological model analysis of 12 small catchments in Puerto Rico.
@@
 Thank you for raising these important points in relation to the validity of the statistical model. 
-A first point that arises from this is that we clarify better what the aim of our statistical modelling is.  
+A first point that arises from this is that we improve the explanation of the aim of our statistical modelling.  
 There are in essence two approaches to statistical modelling.
@@
 Automatic variable selection and potentially validation on independent data are therefore important, as the aim is to develop the most robust model for prediction.
-However, a second reason for to use a statistical model is to explain the maximum variance in the data.
-In this case, it is important to develop a a-priori hypothesis about the causal relationships in the data.
+However, a second reason to use a statistical model is to explain the maximum variance in the data.
+In this case, it is important to develop an a-priori hypothesis about the causal relationships in the data.
 This is subsequently followed by a step by step analysis to test the different causal relationships, either as single variables (as was done in @zhang2017) or jointly (as in our approach).
-In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from adding each variable.
-This explains why we build the model starting from the most simple model, rather than starting from the most complex model.
+In this case there is no attempt to find the best predictive model, instead the focus is on the additional amount of explained variance from each variable.
+This explains why we originally built the model starting from the most simple model, rather than starting from the most complex model.
 
@@
 
-However, another careful review of the data identified many further errors in the data, which were all originally in @zhang2017.
+However, another careful review of the data identified many problematic values in the data, which were all originally in @zhang2017.
 A particular problem was that many catchments had the wrong sign for the change in forest cover.
@@
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
-So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%.
-corrected to 16\%
-- 3, Baker Creek, @zhangwei2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected.
+So the value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%.
+Corrected to 16\%
+- 3, Baker Creek, @zhangwei2012.
+The original recorded 201.1\% change by @zhang2017 is also wrong.
+Original paper says on page 2031: Annual mean flow has been increased by 47.6\%.
+This is now corrected.
 - 67, April rd, which is incorrectly attributed to @ruprecht1991 in @zhang2017.
@@
 - 85 Bosboukloof.
-This is esssentially a duplicate of 184, but the cited paper analyses only 1 year of runoff after a major fire.
+This is esssentially a duplicate of catchment 184, but the cited paper analyses only 1 year of runoff after a major fire.
 In any case, the data should reflect a decrease in forest cover: changed the sign of Delta_f_perc to -80\%.
@@
 
-- a formalised model selection;
-- the use of AIC rather than *r*^2^ for model selection; and
-- interaction terms.
+- A formalised model selection;
+- The use of AIC rather than *r*^2^ for model selection; and
+- Interaction terms.
 
@@
 <!-- answer -->
-Thank you for pointing this out, we have reviewed the paper carefully and corrected these minor issues.
+Thank you for pointing this out, we have reviewed the paper carefully and hopefully corrected these minor issues.
 
@@
 Regrettably, extracting the total area of forest from the papers is not a trivial job, and also has some issues.
-The total forested area is also not always mentioned in the papers, or only a range is given.
+Some papers actually don't provide the total forested area, or only a range is given.
 Additionally, the actual data is likely to show the same skew as the current information, with most of the small and paired watershed catchments having a 100\% cover and only the larger catchments having mix of landuses.
@@
 ```
-As a test, and to further address the AE's comments we have collected as many of the data for the total forest are for the larger catchments (the catchments > 1000 km^2^ from @zhang2017).
+As a test, and to further address the AE's comments we have collected as many of the data for the total forest area for the larger catchments (the catchments > 1000 km^2^ from @zhang2017).
 This results in 60 data points (as not all forest areas were recoverable from the papers).
 This variable is called *Perc_Farea_pre* in Table \@ref(tab:modelforestarea).
-The results of a model that includes this variable and simply assumesa generalise linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
+The results of a model that includes this variable and simply assumes a generalized linear structure, indicates that the percentage total area of forest in the catchment is insignificant, even though this model explains 82 \% of the variation in the data (for only the larger catchments).
 
 Again, it is clear that the assessment technique (in this case there are no paired watershed experiments) is significant in explaining the variation in the data.
+Not surprisingly this is the Hydrological modelling technique.
 
@@
 As a test, the total \% area of forest for the larger catchments (> 1000 km^2^ in @zhang2017) were added to the dataset and the model for just the large catchments was tested.
-This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+This showed that the \% area of forest was not significant to explain the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
 While this might be an area of further research on the full dataset, it is complicated for two reasons:
@@
 We use the models to look at factors that significantly affect streamflow rather than looking at a predictive model.
-We have reviewed the text in the conclusion and have reworded to make sure we capture the above suggested meaning.
+We have reworded the text in the conclusion and have made sure we capture the above suggested meaning.

### 2022-10-07 paper/Forest_and_Water/Abstract.Rmd

@@
 
-  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow.
-Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow.
-However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation.
-The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow?
-Generalised additive models were used to run flexible regressions including multiple variables.
-  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow.
-Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover.
-Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data.
-Smaller catchment dominate the database with 42\% of the data < 1 km^2^ and 65\% of the data < 10 km^2^.
-Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies.
-Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
+  Three recent papers review and analyze large global datasets related to impacts of forest cover on streamflow.
+Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow.
+However, the results are problematic, the underlying data set is unbalanced, and there are correlations in the data that warrant further investigation as this would influence the results.
+For example, the area of the catchment is strongly related to the assessment technique and the variability in the response data.
+For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies.
+Subsequently, the data were re-analyzed using generalized additive modelling.
+  The results highlight that there are  four interlinked reasons that make the general outcomes from the previous papers problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalized and extrapolated to continental and global scales.

### 2022-10-07 paper/Forest_and_Water/Forest_and_Water.Rmd

@@
 ---
-title: 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.'
+title: 'Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple.'
 author:
@@
 For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies.
-Subsequently, the data were re-analyzed using generalised additive modelling.
+Subsequently, the data were re-analyzed using generalized additive modelling.
   The results highlight that there are  four interlinked reasons that make the general outcomes from the previous papers problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
-Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalised and extrapolated to continental and global scales.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalized and extrapolated to continental and global scales.
 keywords: 
-  - data analysis
+  - meta analysis
   - forest cover change
   - global scales
-  - literature review
+  - statistical modelling
 journal: "Journal of Hydrology"
@@
 
-There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects [@wang2020; @hoekvandijke2022].
-More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability [@hoekvandijke2022; @schyns2019].
-The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic example of this is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover.
-However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% change in streamflow as a result of reforestation.
-Similarly a modelling study by @beck2013 found no significant change in streamflow in 12 catchments in Puerto Rico as a result of deforestation.
-In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
+There is a urgent need to identify the impacts of human intervention on stream flow at a global scale and to separate this from climate effects [@wang2020; @hoekvandijke2022].
+More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on stream flow and blue and green water availability [@hoekvandijke2022; @schyns2019].
+The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, stream flow decreases and vice-versa.
+The most dramatic example of this is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in stream flow for catchments with 100% forest cover.
+However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in stream flow characteristics in two of the catchments after deforestation.
+For reforestation, a modelling study across the 1 million km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% change in stream flow as a result of reforestation.
+Similarly a modelling study by @beck2013 found no significant change in stream flow in 12 catchments in Puerto Rico as a result of deforestation.
+In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in stream flow, particular in the dry season, as a result of deforestation.
 This suggests that there can be significant variation across the different studies, methodologies and geographical regions.
@@
 
-There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies.
+There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarize past studies.
 The suggestion is that, because meta-analyses use clearly defined search terms and statistical methods to analyze the results, this will lead to more reliable summaries of past research.
@@
 The recent paper by @filoso2017 is a clear meta-analysis, but most others [@zhang2017; @hoekvandijke2022; @zhou2015] are not.
-However,  n impressive global database of catchment studies with changes in streamflow due to changes in forest cover has been developed [@zhang2017; @filoso2017] and statistical approaches are used to analyze the resulting data.
-The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+However,  n impressive global database of catchment studies with changes in stream flow due to changes in forest cover has been developed [@zhang2017; @filoso2017] and statistical approaches are used to analyze the resulting data.
+The @zhang2017 data set, which covers over 312 studies, is described in terms of the change in stream flow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
@@
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
-The more regionally concentrated and detailed study by @levy2018 is a further independent dataset with no overlap with the other studies.
-However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.
+The more regionally concentrated and detailed study by @levy2018 is a further independent data set with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in land cover was derived from satellite data and was not made available.
 
@@
 This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
-Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
+Their main conclusion was also that reforestation decreases stream flow, but that there were many interacting factors.
 For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with @zhang2017), but they did not identify a distinct threshold and fitted a log-linear relationship.
-In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
+In addition, they identified that studies with shorter periods of data collection resulted in larger declines in stream flow.
 
@@
 However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model, which is similar to well-known Budyko approaches [@zhang2018elasticity].
-In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model.
+In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model, which represents vegetation cover.
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.
@@
 The main method in the work by @zhang2017 is a single covariate linear regression.
-In contrast, the systematic review from @filoso2017 mainly emphasises the classification and distributions of the study.
+In contrast, the systematic review from @filoso2017 mainly emphasizes the classification and distributions of the study.
 @zhang2017 points out that a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments.
@@
 The analysis specifically accounted for differences between catchments and differences due to variations in climate.
-Not all datasets are however suitable for this kind of in-depth analysis.
-
-Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another review paper on this topic?
+Not all data sets are however suitable for this kind of in-depth analysis.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on stream flow, the question is why another review paper on this topic?
 There is a real attraction in the concept of statistical analysis of past studies encapsulated in meta-analysis to be able to extrapolate findings to larger scales, and to identify factors across global scales [@evaristo2020metaanalysis].
@@
 
-The aim of this paper is to first reanalyze the global dataset [@zhang2017; @filoso2017] using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
+The aim of this paper is to first reanalyze the global data set [@zhang2017; @filoso2017] using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
 This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
 In addition, by applying more complex statistical models, we will highlight the conclusions that can be drawn from the data.
-Finally, we will highlight future research needs in the area of forest cover change impact on streamflow.
+Finally, we will highlight future research needs in the area of forest cover change impact on stream flow.
 
@@
 As indicated, the starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material.
-The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
-The change in streamflow ($\Delta Qf\%$) is based on equation 1 in @zhang2017.
-
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \@ref(tab:table1). These abbreviations will later be used in the models.
+The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in stream flow in \% $(\Delta Qf\%)$, the precipitation data type, the assessment technique, and the	source of the info, which is a citation.
+The change in stream flow ($\Delta Qf\%$) is based on equation 1 in @zhang2017.
+
+Several of these columns contain abbreviations to describe the different variables, which are summarized in Table \@ref(tab:table1). These abbreviations will later be used in the models.
 
@@
 Mostly the  data reported by the authors was used, but in some cases the variables had to be approximated from the location of the centre of the catchment using Google Maps^TM^.
-In the dataset, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
+In the data set, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
 
@@
 
-Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @penaarancibia2012, and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow [i.e.
+Several additional data points from catchment studies were extracted from @almeida2016, @ferreto2020, @iroume2013, @iroume2006, @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @penaarancibia2012, and these were checked against the existing studies to prevent overlap.
+In the citation column in the accompanying data set, the main reference for the calculated change in stream flow  was generally used, because sometimes the original study did not provide the quantification of the change in stream flow [i.e.
 Table 6 in @zhang2011].
@@
 It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017.
-This will allow future research to scrutinise our input for errors.
+This will allow future research to scrutinize our input for errors.
 
@@
 This first aim is similar to @zhang2017, but the main difference is that we start off with all variables in the data set in the model.
-Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and  how latent variables in the data set can be masked and result in relationships that might not really exist.
-Finally, the analysis will highlight how the results are conditional on the dataset.
+Subsequently the analysis will concentrate on how the individual variables in the data set relate to each other and  how latent variables in the data set can be masked and result in relationships that might not really exist.
+Finally, the analysis will highlight how the results are conditional on the data set.
 
@@
 
-To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].
+To estimate how the change in stream flow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalized additive modelling (GAM) [@wood2006].
 
@@
 This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship.
-In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.
+In other words, the approach balances finding a smooth non-linear relationship for the variable against over fitting the data.
 
@@
 
-The overarching test focuses on identifying the change streamflow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
+The overarching test focuses on identifying the change stream flow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study.
 In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref(tab:table1)
@@
 As an initial approach we tested whether the additional catchments added to the original data from @zhang2017 did not majorly influenced the results (This analysis is in supplementary material part 2).
-This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
+This analysis highlights that the newly added catchment and the changes to the data set create minor differences when repeating the analysis from the original paper.
 However, this means that the results of the studies are still comparable.
@@
 
-The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
+The overall data set contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+In contrast, the original data set from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017).
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*.
@@
 
-Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
-The catchments are spread across the world, and relative to @zhang2017, this dataset has a very similar geospatial distribution.
+Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analyzing the effect of climate.
+The catchments are spread across the world, and relative to @zhang2017, this data set has a very similar geospatial distribution.
 The major climate gradients are represented in the data, but there appears to be some bias in the spatial locations of the data.
@@
 
-The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations thate are worth investigating, even though in general cross correlations between variables are quite low.
+The correlation plot (Figure \@ref(fig:corgraphs)) highlights several correlations that are worth investigating, even though in general cross correlations between variables are quite low.
 Some interesting relationships that appear in this graph are:
@@
 This includes introducing non-linearity (Equation \@ref(eq:eq2)) for the numerical variables in the model.
-While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
+While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting over fitting.
 Following @wood2006, the number of degrees of freedom *k* in the non-linear variables was based on assessment of the effective degrees of freedom in the model output.
@@
 
-The overall explaining power of the model can be interpreted from the adjusted *r^2^* (which is penalised for the number of parameters).
+The overall explaining power of the model can be interpreted from the adjusted *r^2^* (which is penalized for the number of parameters).
 This indicates an adjusted *r^2^* of `r round(summary(Forest_model_all)$r.sq,2)` and deviance explained is `r round(summary(Forest_model_all)$dev.expl,2)`, suggesting the model only explains about 50% of the variance in the data.
@@
 The overall partial slope of the change in forest cover is `r round(summary(Forest_model_all)$p.table[2,1],2)`, if all other variables are kept constant.
-This suggest quite strong change in streamflow, moving from fully forested to fully cleared.
+This suggest quite strong change in stream flow, moving from fully forested to fully cleared.
 Over the whole forest cover range, this is a change of `r round(summary(Forest_model_all)$p.table[2,1],2)*200` mm, with other variables held constant.
@@
 In addition, all the smoothed variables *log10(Area (km^2^))* (p = `r round(summary(Forest_model_all)$s.table[1,4],2)`)), *Dryness* (p = `r round(summary(Forest_model_all)$s.table[2,4],2)`)) and *Length* (p = `r round(summary(Forest_model_all)$s.table[3,4],2)`)) explain variation in the data.
-For *Length*, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
-
-Furthermore Table \@ref(tab:m-all-linear) indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
+For *Length*, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in stream flow.
+
+Furthermore Table \@ref(tab:m-all-linear) indicates that several of the assessment methods explain variation in the change in stream flow, which was also indicated in the correlation analysis.
 In particular, the assessment methods Paired Watersheds Experiments (PWE), Hydrological Modelling (HM) and Statistical modelling and hydrographs (SH) are important explaining variables ($p < 0.05$).
@@
 As discussed in the methods, the overall linear response to the change in forest cover was compared to a transformation of the negative forest cover to positives and a check whether the relationship might be non-linear.
-This approach tests whether the impact on streamflow from removing forest cover is different from reforestation, as outlined in the methods.
+This approach tests whether the impact on stream flow from removing forest cover is different from reforestation, as outlined in the methods.
 The detail of the comparison is highlighted in Supplementary material part 3.
@@
 Figure \@ref(fig:smoothsmodelall) highlights that the relationship between *log10(Area km^2^)* and the change in flow is essentially linear.
-It indicates the negative slope that was also clear from @zhang2017, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+It indicates the negative slope that was also clear from @zhang2017, indicating that in larger catchments changes in forest cover have less impact on stream flow than for smaller catchments.
 
@@
 Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \@ref(fig:gamcheckmodelall), top left).
-This is related to a limited number catchments that have very high changes in streamflow in the data set.
+This is related to a limited number catchments that have very high changes in stream flow in the data set.
 In other words, the distribution of the residuals is somewhat fat-tailed.
@@
 
-Given the majority of the residuals indicate a relatively well behaved distribution, we simply note the behaviour at the extremes and will discuss this later in the paper, and explain how this relates to the characteristics of the dataset.
+Given the majority of the residuals indicate a relatively well behaved distribution, we simply note the behaviour at the extremes and will discuss this later in the paper, and explain how this relates to the characteristics of the data set.
 
@@
 
-The points related to catchments with very long studies (> 60 years) might be questionable, as changes other than forest cover change could affect streamflow.
+The points related to catchments with very long studies (> 60 years) might be questionable, as changes other than forest cover change could affect stream flow.
 In addition, a few of the catchments have Dryness values that are very large (> 5) and these values have high leverage in the data, affecting the residual distribution.
@@
 Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
-Data that have *Dryness* <= 5 and *Length* <= 60 years were removed from the dataset and the model based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments is run again.
+Data that have *Dryness* <= 5 and *Length* <= 60 years were removed from the data set and the model based on a reduction of the data set from `r nrow(All_data2)` to `r nrow(NoLong_Data)` catchments is run again.
 
@@
 
-Investigating the non-linear responses suggest that *Dryness* has a clear non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
+Investigating the non-linear responses suggest that *Dryness* has a clear non-linear response, which is significant,  where changes in forest cover in drier catchments having a greater impact on stream flow (Figure \@ref(fig:smoothsmodelredLength) and Table \@ref(tab:restrictlength)).
 Catchment area (*log10(Area (km^2^))*) still has an impact on flow with p = `r round(summary(model_redLength)$s.table[2,4],2)`, and the relationship looks almost linear.
@@
 
-The generalised additive models appear to reach the same conclusions as the single variable regression in earlier papers [@zhang2017; @filoso2017]. It appears that:
-
-1. Larger catchments show lower impact of forest cover change on streamflow;
-2. Drier catchments show a greater impact of forest cover change on streamflow; and
-3. There is a general linear relationship between the change in forest cover and the change in streamflow.
+The generalized additive models appear to reach the same conclusions as the single variable regression in earlier papers [@zhang2017; @filoso2017]. It appears that:
+
+1. Larger catchments show lower impact of forest cover change on stream flow;
+2. Drier catchments show a greater impact of forest cover change on stream flow; and
+3. There is a general linear relationship between the change in forest cover and the change in stream flow.
 
@@
 4.
-There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest % forest cover change and the largest variability in the flow response.
+There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest percent forest cover change and the largest variability in the flow response.
 
@@
 
-It is possible that one of the reasons why @zhang2017 separated their analysis in large (> 1000 km^2^) and small (< 1000 km^2^) catchments, is that they realised this difference in assessment methods and wanted to account for this.
+It is possible that one of the reasons why @zhang2017 separated their analysis in large (> 1000 km^2^) and small (< 1000 km^2^) catchments, is that they realized this difference in assessment methods and wanted to account for this.
 However, this is not explicitly identified, and there is no real physical explanation of the 1000 km^2^ threshold.
@@
 Similarly, the data for the larger catchments containing smaller changes in forest cover are dominated by hydrological modelling studies, resulting in a further complication.
-If the reponse of the streamflow in the modelling studies is the result of the conceptualised relationship between streamflow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in streamflow is real, or simply a result of a pre-conceived model relationship.
-Is the smaller variation in the data for smaller changed in forest cover (Figure \@ref(fig:overview)) a result of similar conceptualised model relationships, or actual variation between catchments and climate types?
+If the reponse of the stream flow in the modelling studies is the result of the conceptualized relationship between stream flow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in stream flow is real, or simply a result of a pre-conceived model relationship.
+Is the smaller variation in the data for smaller changed in forest cover (Figure \@ref(fig:overview)) a result of similar conceptualized model relationships, or actual variation between catchments and climate types?
 Currently this question cannot be answered.
@@
 This becomes problematic when extrapolated to larger scales.
-A clear example of this is the paper by @hoekvandijke2022 where the conceptualised relationship between forest cover and streamflow pre-determines the outcomes of the global modelling.
-
-The only  way to analyze changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in @levy2018).
+A clear example of this is the paper by @hoekvandijke2022 where the conceptualized relationship between forest cover and stream flow pre-determines the outcomes of the global modelling.
+
+The only  way to analyze changes in stream flow as a function of forest cover in larger catchments is to actually derive this from observed data of long term stream flow and forest cover (as was done in @levy2018).
 
@@
 In this case, the total \% area of forest was not included in the data.
-As a test, the total \% area of forest for the larger catchments (> 1000 km^2^ in @zhang2017) were added to the dataset and the model for just the large catchments was tested.
-This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
-While this might be an area of further research on the full dataset, it is complicated for two reasons:
+As a test, the total \% area of forest for the larger catchments (> 1000 km^2^ in @zhang2017) were added to the data set and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant to explain the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full data set, it is complicated for two reasons:
 
@@
 
-We are not arguing that there is no relationship between streamflow and forest cover, and there might indeed be a global relationship that can be discovered.
+We are not arguing that there is no relationship between stream flow and forest cover, and there might indeed be a global relationship that can be discovered.
 But, this relationship can only be discovered if we are able to address some of the major other factors that explain the variability, and work with actual data and not model outputs.
@@
 The second major issue that became clear from reviewing many of the original papers is that some of the variability might be an interpretation problem.
-In many cases the original description in the paper is interpreted to extract the \% change in streamflow from the \% change in forest cover.
+In many cases the original description in the paper is interpreted to extract the \% change in stream flow from the \% change in forest cover.
 This seems like a simple activity, but this is not always the case.
@@
 - Another example is the paper by @waterloo2007.
-This modelling study in Fiji of the clearing of a catchment reports the changes in streamflow over parts of the year.
-For a period of 324 days the streamflow increased from 252 mm to 580 mm (a 230\% increase if calculated as $580/252*100$) and for a second period of 309 days the streamflow increased from 90 mm to 194 mm (a 215 \% increase).
+This modelling study in Fiji of the clearing of a catchment reports the changes in stream flow over parts of the year.
+For a period of 324 days the stream flow increased from 252 mm to 580 mm (a 230\% increase if calculated as $580/252*100$) and for a second period of 309 days the stream flow increased from 90 mm to 194 mm (a 215 \% increase).
 However, how we convert this to a change in annual flow (which most of the other data relate to) is difficult.
@@
 We suspect that results from $252/580*100 \approx 50$ and $90/194 \approx 50$.
+- A final example is around the choice of control or treatment.
+In the data base the assumption is that the change in flow is relative to the original situation.
+This can be either a "before and after" analysis, which can be problematic by itself due to climate variation, or a comparison of a treatment with a control.
+But even in this case comparing across catchments can be tricky.
+For example, at one extreme, some controls in the database are a bamboo catchment and a tea plantation [@blackie1979kimakia; @blackie1979kericho].
+A clearer example is the Brigalow catchment study in Queensland, Australia [@thornton2007brigalow], catchment #336 in the data set.
+This is a paired watershed experiment of conversion of native Brigalow (an Acacia species) into cropland and pasture.
+We chose to use the cropped catchment (C2 @thornton2007brigalow) as the deforestation treatment resulting in a change of flow of 140\%, based on Table 4 in @thornton2007brigalow.
+However, had we used the pasture catchment as the treatment then the change in flow would have been up to 165\%.
 
@@
 As a result, it becomes difficult to assess how we interpret the change in flow as a result of a change in cover.
-In many cases we would expect the flow to change over time as a function of the recovery [@jones2017] and therefore the timeseries of the flow needs to be assessed over a longer time.
+In many cases we would expect the flow to change over time as a function of the recovery [@jones2017] and therefore the time series of the flow needs to be assessed over a longer time.
 
@@
 An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
-For example, @heath2014 found that catchments with resprouting species around Sydney, Australia, indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent [@zhou2015bushfire].
+For example, @heath2014 found that catchments with resprouting species around Sydney, Australia, indicated little change in the stream flow in comparison to species regrowing from seed further south on the continent [@zhou2015bushfire].
 
 As a result, it can be difficult to exactly pinpoint the change in flow as a result of the change in cover, as well as being difficult to assess what the exact change in cover actually was.
-
-As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
+In addition, using only the overall change in stream flow can discard a lot of information from observations in individual years.
+Many papers give stream flow values for multiple years, often showing significant variation.
+Summarizing this into one average value discards all the information on the variance.
+
+As indicated before, if the overall variation in the data due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
 As @jones2017 indicate, this really needs time series analysis of the different experiments.
@@
 In the end, careful review of the data and the original papers can circumvent most of this issue.
-And, making the data available (as @zhang2017, @zhou2015 and @filoso2017 have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
+And, making the data available (as @zhang2017, @zhou2015, @filoso2017 and this paper have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
 
@@
 
-In this paper, a few studies have been singled out in the analysis.
-The choice of focus was mainly driven by the data that was made available by the authors of these papers [@zhang2017; @filoso2017], which provide a rich case study for the current paper.
+In this paper, a few studies have been singled out to demonstrate the main point that extrapolating individual studies to global scales is not that simple in natural systems.
+Not only is there significant natural variation and latent variables, interpretation and aggregation can cause further unforeseen problems.
+The choice of papers to focus on was mainly driven by the data that was made available by the authors of these papers [@zhang2017; @filoso2017], which provide a rich case study for the current paper.
 
@@
 This once again highlights that there is unexplained variation at the extremes of the distribution, once again related to the paired watershed experiments (Figure \@ref(fig:assessment)).
-Generally,in statistical models, the approach would be to further normalise the residuals through transformations.
+Generally,in statistical models, the approach would be to further normalize the residuals through transformations.
 However, in this case this might be difficult and might not resolve all the issues due to the large variation in the data.
@@
 However, this paper highlights that there is considerable chance that large historical data sets include latent variables and are more complex than envisioned.
-This is particularly true for more historical work, as methods of observation and even apporaches to management have changed considerably.
+This is particularly true for work in natural systems and more historical work, as methods of observation and even approaches to management have changed considerably.
 The same management description is not necessarily the same action on the ground.
@@
 For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
-It is tempting to use the result of a large scale analysis of this data to make inferences about overall landuse change [@li2017; @wang2020], but this would not be valid, as the deforestation studies are generally not a transition to an agricultural landuse or pasture, but regrowing into forest.
+It is tempting to use the result of a large scale analysis of this data to make inferences about overall land use change [@li2017; @wang2020], but this would not be valid, as not all deforestation studies are a transition to an agricultural land use or pasture, as in @levy2018.
+Many are logging or bushfire studies regrowing into forest after the initial treatment.
 Similarly, using the plantation studies to extrapolate to "reforestation" (as in @filoso2017 and @hoekvandijke2022) is also tenuous.
@@
 Given the link between ecophysiology and water and carbon budgets [@jackson2005], care should be taken in extrapolation, introducing a further error.
+
+As highlighted summarizing highly variable observed time series into single mean values introduces further potential problems, as well as discarding information about the variance.
 
@@
 either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in land cover.
-As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
+In addition, this might allow analysis of the variance of the response in addition to the mean response.
+While some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
 
 More generally there is a clear need for a more in depth analysis of the data base of studies used here.
-In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, streamflow responses and responses of components of streamflow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
+In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, stream flow responses and responses of components of stream flow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
 
@@
 Extrapolating paired watershed experiment results into models can possibly overlook landscape interactions that are visible at larger scales, but do not occur on smaller scales.
-For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on landuse.
-A carefully designed simulation study that specifically investigates the change in stream flow repsonse with scale using local field data for verification can help solve this problem.
-
-At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
-A better approach is analyzing streamflow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes.
-The highlighted paper by @levy2018 is currently the best example of a solid statistical approach to analyzing streamflow responses.
+For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on land use.
+A carefully designed simulation study that specifically investigates the change in stream flow response with scale using local field data for verification can help solve this problem.
+
+At the moment, providing answers to the impact of stream flow at larger scales should generally not be approached by simulation modelling.
+A better approach is analyzing stream flow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include land use changes.
+The highlighted paper by @levy2018 is currently the best example of a solid statistical approach to analyzing stream flow responses.
 Simulation modelling can be an approach to analyze different scenarios, if there is clear recognition of the potential impact of the model structure (the algorithms and parameters that describe for example plantation tree growth) on the simulation outcomes.
@@
 
-This study demonstrates that analyzes of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
-
-While the analysis reveals similar conclusions in relation to the response of streamflow to forest cover, there are four major interlinked reasons why these results should be considered carefully.
-This subsequently has implications for meta-analyses in Environmental Science and Hydrology in general.
-The reasons higlighted in this paper are:
-
-- The existence of latent variables in the data that create the appearance of a relationship that really does not exist;  
-- The difficulty in fully interpreting the specifics of different studies;  
-- The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and  
+This study demonstrates that the analysis of large databases of essentially "aggregated data" should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+
+On first glance, stream flow will decrease with increases in forest cover, and increase with removal of forest cover.
+However, the conclusions on the exact relationship between stream flow change and forest cover change are highly uncertain.
+There are four major interlinked reasons why earlier conclusions should be considered carefully:
+
+- The existence of latent variables in the data that create the appearance of a relationship that really does not exist.
+In this case the assessment technique used to assess the change in stream flow is highly significant;
+- The difficulty in fully interpreting the specifics of different studies. In this case the definition of vegetation, stream flow volume, control or time period was difficult to assess;  
+- The difficulty of integrating data from seemingly similar studies, but with quite different objectives.
+This study highlighted the many logging studies followed by regrowth, or bushfire effects; and
 - The chance of transcription errors influencing the data.
 
+While some of these issues might be overcome with careful analysis and transcription, not all can be directly or easily resolved.
 Any statistical analysis, including the one in this paper, needs to be considered "conditional on the data", and given the issues indicated, extrapolation of the results of summary studies to larger scales and into global hydrological models has to be done with great care.
 Better would be to analyze observed data and explicitly include uncertainty in the extrapolation of the results.
-
 This therefore has implications for the recent growth in meta-analysis review papers, which has been boosted by increased computational capacity and much better on-line accessible data bases with research data.
-Clearly, this requires careful definition of the search terms, and follow-up review of the harvested papers, as well as an understanding that the statistical relationships can be hiding other unknown factors.
-As the old adagium says: Correlation is not the same as causation.
+This is particularly true for natural systems involving climate variability and therefore extrapolations of experimental work in Environmental Science and Hydrology to global scales in general.
 

### 2022-10-07 paper/Forest_and_Water/Forest_and_Water.tex

@@
             pdfauthor={},
-            pdftitle={Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.},
+            pdftitle={Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple.},
             colorlinks=false,
@@
 
-  \title{Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.}
+  \title{Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple.}
     \author[a,b]{R. Willem Vervoort}
@@
 For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies.
-Subsequently, the data were re-analyzed using generalised additive modelling.
+Subsequently, the data were re-analyzed using generalized additive modelling.
   The results highlight that there are four interlinked reasons that make the general outcomes from the previous papers problematic: 1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data.
-Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalised and extrapolated to continental and global scales.
+Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalized and extrapolated to continental and global scales.
   \end{abstract}
     \begin{keyword}
-    data analysis \sep forest cover change \sep global scales \sep 
-    literature review
+    meta analysis \sep forest cover change \sep global scales \sep 
+    statistical modelling
   \end{keyword}
@@
 
-There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects \citep{wang2020, hoekvandijke2022}.
-More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability \citep{hoekvandijke2022, schyns2019}.
-The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa.
-The most dramatic example of this is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in streamflow for catchments with 100\% forest cover.
-However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in streamflow characteristics in two of the catchments after deforestation.
-For reforestation, a modelling study across the 1 millon km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% change in streamflow as a result of reforestation.
-Similarly a modelling study by \citet{beck2013} found no significant change in streamflow in 12 catchments in Puerto Rico as a result of deforestation.
-In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in streamflow, particular in the dry season, as a result of deforestation.
+There is a urgent need to identify the impacts of human intervention on stream flow at a global scale and to separate this from climate effects \citep{wang2020, hoekvandijke2022}.
+More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on stream flow and blue and green water availability \citep{hoekvandijke2022, schyns2019}.
+The past work reviewing these impacts \citep{andreassian2004, jackson2005, zhang2017, brown2005, brown2013, filoso2017} highlights a general consensus that if forest areas increase, stream flow decreases and vice-versa.
+The most dramatic example of this is Figure 5 in \citet{zhang2011} indicating (for Australian catchments) a 100\% decrease in stream flow for catchments with 100\% forest cover.
+However, on the other end of the spectrum, for three French catchments \citep{cosandey2005}, there was no change in stream flow characteristics in two of the catchments after deforestation.
+For reforestation, a modelling study across the 1 million km\textsuperscript{2} Murray Darling Basin also found no major effect, especially in larger catchments \citep{vandijk2007}, but a recent study \citep{hoekvandijke2022} found an 8\% change in stream flow as a result of reforestation.
+Similarly a modelling study by \citet{beck2013} found no significant change in stream flow in 12 catchments in Puerto Rico as a result of deforestation.
+In contrast, in a recent study in Brazil across 324 catchments, \citet{levy2018} found a significant increase in stream flow, particular in the dry season, as a result of deforestation.
 This suggests that there can be significant variation across the different studies, methodologies and geographical regions.
@@
 
-There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarise past studies.
+There has been a recent push in the hydrological community \citep{evaristo2020metaanalysis} to use `meta-analysis' to summarize past studies.
 The suggestion is that, because meta-analyses use clearly defined search terms and statistical methods to analyze the results, this will lead to more reliable summaries of past research.
@@
 The recent paper by \citet{filoso2017} is a clear meta-analysis, but most others \citep{zhang2017, hoekvandijke2022, zhou2015} are not.
-However, n impressive global database of catchment studies with changes in streamflow due to changes in forest cover has been developed \citep{zhang2017, filoso2017} and statistical approaches are used to analyze the resulting data.
-The \citet{zhang2017} dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
+However, n impressive global database of catchment studies with changes in stream flow due to changes in forest cover has been developed \citep{zhang2017, filoso2017} and statistical approaches are used to analyze the resulting data.
+The \citet{zhang2017} data set, which covers over 312 studies, is described in terms of the change in stream flow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included.
 In contrast, the paper by \citet{filoso2017} focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review.
@@
 There is some overlap between the two data sets, but there are also some studies unique to both sets.
-The more regionally concentrated and detailed study by \citet{levy2018} is a further independent dataset with no overlap with the other studies.
-However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.
+The more regionally concentrated and detailed study by \citet{levy2018} is a further independent data set with no overlap with the other studies.
+However, for this study only the flow and rainfall data is available for the catchments, and the change in land cover was derived from satellite data and was not made available.
 
@@
 This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis.
-Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors.
+Their main conclusion was also that reforestation decreases stream flow, but that there were many interacting factors.
 For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with \citet{zhang2017}), but they did not identify a distinct threshold and fitted a log-linear relationship.
-In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.
+In addition, they identified that studies with shorter periods of data collection resulted in larger declines in stream flow.
 
@@
 However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model, which is similar to well-known Budyko approaches \citep{zhang2018elasticity}.
-In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model.
+In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model, which represents vegetation cover.
 A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in \citet{zhang2017}.
@@
 The main method in the work by \citet{zhang2017} is a single covariate linear regression.
-In contrast, the systematic review from \citet{filoso2017} mainly emphasises the classification and distributions of the study.
+In contrast, the systematic review from \citet{filoso2017} mainly emphasizes the classification and distributions of the study.
 \citet{zhang2017} points out that a main assumption in their work is that the catchment size threshold at 1000 km\textsuperscript{2} is a distinct separation between ``small'' and ``large'' catchments.
@@
 The analysis specifically accounted for differences between catchments and differences due to variations in climate.
-Not all datasets are however suitable for this kind of in-depth analysis.
-
-Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another review paper on this topic?
+Not all data sets are however suitable for this kind of in-depth analysis.
+
+Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on stream flow, the question is why another review paper on this topic?
 There is a real attraction in the concept of statistical analysis of past studies encapsulated in meta-analysis to be able to extrapolate findings to larger scales, and to identify factors across global scales \citep{evaristo2020metaanalysis}.
@@
 
-The aim of this paper is to first reanalyze the global dataset \citep{zhang2017, filoso2017} using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
+The aim of this paper is to first reanalyze the global data set \citep{zhang2017, filoso2017} using some more detailed statistical modelling and to use this to highlight examples of each of these limitations.
 This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations.
 In addition, by applying more complex statistical models, we will highlight the conclusions that can be drawn from the data.
-Finally, we will highlight future research needs in the area of forest cover change impact on streamflow.
+Finally, we will highlight future research needs in the area of forest cover change impact on stream flow.
 
@@
 As indicated, the starting point of this paper is the data base of studies which were included in \citet{zhang2017} as supplementary material.
-The columns in this data set are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in streamflow in \% \((\Delta Qf\%)\), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
-The change in streamflow (\(\Delta Qf\%\)) is based on equation 1 in \citet{zhang2017}.
-
-Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \ref{tab:table1}. These abbreviations will later be used in the models.
+The columns in this data set are the catchment number, the catchment name, the Area in km\textsuperscript{2}, the annual average precipitation (Pa) in mm, the forest type, hydrological regime, and climate type, the change in forest cover in \% (\(\Delta F\%\)) and the change in stream flow in \% \((\Delta Qf\%)\), the precipitation data type, the assessment technique, and the source of the info, which is a citation.
+The change in stream flow (\(\Delta Qf\%\)) is based on equation 1 in \citet{zhang2017}.
+
+Several of these columns contain abbreviations to describe the different variables, which are summarized in Table \ref{tab:table1}. These abbreviations will later be used in the models.
 
@@
 Mostly the data reported by the authors was used, but in some cases the variables had to be approximated from the location of the centre of the catchment using Google Maps\textsuperscript{TM}.
-In the dataset, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
+In the data set, an additional column has been added to indicate the source of the location data to indicate if this is directly from the paper or elsewhere.
 
@@
 
-Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{penaarancibia2012}, and these were checked against the existing studies to prevent overlap.
-In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow \citep[i.e.~Table 6 in][]{zhang2011}.
+Several additional data points from catchment studies were extracted from \citet{almeida2016}, \citet{ferreto2020}, \citet{iroume2013}, \citet{iroume2006}, \citet{zhang2011}, \citet{zhao2010}, \citet{borg1988}, \citet{thornton2007}, \citet{zhou2010}, \citet{rodriguez2010}, \citet{ruprechtetal1991} and \citet{penaarancibia2012}, and these were checked against the existing studies to prevent overlap.
+In the citation column in the accompanying data set, the main reference for the calculated change in stream flow was generally used, because sometimes the original study did not provide the quantification of the change in stream flow \citep[i.e.~Table 6 in][]{zhang2011}.
 
@@
 It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from \citet{zhang2017}.
-This will allow future research to scrutinise our input for errors.
+This will allow future research to scrutinize our input for errors.
 
@@
 This first aim is similar to \citet{zhang2017}, but the main difference is that we start off with all variables in the data set in the model.
-Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and how latent variables in the data set can be masked and result in relationships that might not really exist.
-Finally, the analysis will highlight how the results are conditional on the dataset.
+Subsequently the analysis will concentrate on how the individual variables in the data set relate to each other and how latent variables in the data set can be masked and result in relationships that might not really exist.
+Finally, the analysis will highlight how the results are conditional on the data set.
 
@@
 
-To estimate how the change in streamflow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalised additive modelling (GAM) \citep{wood2006}.
+To estimate how the change in stream flow is affected by the change in forest cover, while considering the effects of the other variables, we applied generalized additive modelling (GAM) \citep{wood2006}.
 
@@
 
-The overarching test focuses on identifying the change streamflow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
+The overarching test focuses on identifying the change stream flow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: \citet{zhang2017}; \citet{filoso2017}; \citet{zhou2015}): climate, size of catchment and length of study.
 In addition to these earlier identified factors, this study also tested for the factors listed in Table \ref{tab:table1}
@@
 As an initial approach we tested whether the additional catchments added to the original data from \citet{zhang2017} did not majorly influenced the results (This analysis is in supplementary material part 2).
-This analysis highlights that the newly added catchment and the changes to the dataset create minor differences when repeating the analysis from the original paper.
+This analysis highlights that the newly added catchment and the changes to the data set create minor differences when repeating the analysis from the original paper.
 However, this means that the results of the studies are still comparable.
@@
 
-The overall dataset contains 329 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
-In contrast, the original dataset from \citet{zhang2017} contained 312 catchments and the \citet{filoso2017} study used 37 catchments (Table S2 in \citet{filoso2017}).
+The overall data set contains 334 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data.
+In contrast, the original data set from \citet{zhang2017} contained 312 catchments and the \citet{filoso2017} study used 37 catchments (Table S2 in \citet{filoso2017}).
 The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\textsuperscript{2}}.
@@
 
-Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analysing the effect of climate.
-The catchments are spread across the world, and relative to \citet{zhang2017}, this dataset has a very similar geospatial distribution.
+Apart from looking at the distribution of the values, the spatial locations of the data can also be important, in particular when analyzing the effect of climate.
+The catchments are spread across the world, and relative to \citet{zhang2017}, this data set has a very similar geospatial distribution.
 The major climate gradients are represented in the data, but there appears to be some bias in the spatial locations of the data.
@@
 
-The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations thate are worth investigating, even though in general cross correlations between variables are quite low.
+The correlation plot (Figure \ref{fig:corgraphs}) highlights several correlations that are worth investigating, even though in general cross correlations between variables are quite low.
 Some interesting relationships that appear in this graph are:
@@
 This includes introducing non-linearity (Equation \eqref{eq:eq2}) for the numerical variables in the model.
-While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting overfitting.
+While increasing non-linearity in the model can increase the flexibility if the model, the shrinkage splines assist with limiting over fitting.
 Following \citet{wood2006}, the number of degrees of freedom \emph{k} in the non-linear variables was based on assessment of the effective degrees of freedom in the model output.
@@
 \endhead
-\textbf{(Intercept)} & -4.81 & 16.31 & -0.29 & 0.77 \\
-\textbf{DeltaF\_perc} & -0.6 & 0.06 & -10.71 & 0 \\
-\textbf{Precip\_data\_typeOB} & -21.4 & 13.23 & -1.62 & 0.11 \\
-\textbf{Precip\_data\_typeSG} & 9.36 & 15.17 & 0.62 & 0.54 \\
-\textbf{Assessment\_techniqueEA, HM} & 20.64 & 42.73 & 0.48 & 0.63 \\
-\textbf{Assessment\_techniqueHM} & 22.81 & 11.71 & 1.95 & 0.05 \\
-\textbf{Assessment\_techniquePWE} & 30.63 & 11.94 & 2.57 & 0.01 \\
+\textbf{(Intercept)} & -5.37 & 16.19 & -0.33 & 0.74 \\
+\textbf{DeltaF\_perc} & -0.61 & 0.06 & -11.03 & 0 \\
+\textbf{Precip\_data\_typeOB} & -21.34 & 13.16 & -1.62 & 0.11 \\
+\textbf{Precip\_data\_typeSG} & 9.57 & 15.16 & 0.63 & 0.53 \\
+\textbf{Assessment\_techniqueEA, HM} & 20.32 & 42.72 & 0.48 & 0.63 \\
+\textbf{Assessment\_techniqueHM} & 23.51 & 11.69 & 2.01 & 0.05 \\
+\textbf{Assessment\_techniquePWE} & 30.71 & 11.92 & 2.58 & 0.01 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 17.42 & 43.26 & 0.4 & 0.69 \\
-\textbf{Assessment\_techniqueQPW} & 39.52 & 20.15 & 1.96 & 0.05 \\
+HM} & 15.79 & 43.24 & 0.37 & 0.72 \\
+\textbf{Assessment\_techniqueQPW} & 41.29 & 20.14 & 2.05 & 0.04 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 24.39 & 24.42 & 1 & 0.32 \\
-\textbf{Assessment\_techniqueSH} & 45.3 & 11.83 & 3.83 & 0 \\
-\textbf{Forest\_typeCF} & -9.45 & 7.6 & -1.24 & 0.21 \\
-\textbf{Forest\_typeMF} & -8.05 & 7.56 & -1.06 & 0.29 \\
-\textbf{Hydrological\_regimeSD} & 3.57 & 9.16 & 0.39 & 0.7 \\
+EA} & 25.16 & 24.41 & 1.03 & 0.3 \\
+\textbf{Assessment\_techniqueSH} & 46.03 & 11.65 & 3.95 & 0 \\
+\textbf{Forest\_typeCF} & -7.76 & 7.52 & -1.03 & 0.3 \\
+\textbf{Forest\_typeMF} & -7.8 & 7.35 & -1.06 & 0.29 \\
+\textbf{Hydrological\_regimeSD} & 1.5 & 9.1 & 0.17 & 0.87 \\
 \bottomrule()
@@
 \endhead
-\textbf{s(log10(Area\_km2))} & 0.79 & 4 & 0.99 & 0.02 \\
-\textbf{s(Dryness)} & 4.64 & 9 & 2.26 & 0 \\
-\textbf{s(Length)} & 4.45 & 34 & 0.22 & 0.12 \\
+\textbf{s(log10(Area\_km2))} & 0.81 & 4 & 1.09 & 0.02 \\
+\textbf{s(Dryness)} & 4.59 & 9 & 2.25 & 0 \\
+\textbf{s(Length)} & 4.39 & 34 & 0.21 & 0.13 \\
 \bottomrule()
@@
 
-The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalised for the number of parameters).
+The overall explaining power of the model can be interpreted from the adjusted \emph{r\textsuperscript{2}} (which is penalized for the number of parameters).
 This indicates an adjusted \emph{r\textsuperscript{2}} of 0.45 and deviance explained is 0.49, suggesting the model only explains about 50\% of the variance in the data.
@@
 Inspecting the significance of the variables (Table \ref{tab:m-all-linear} and Table \ref{tab:m-all-smooth}) indicates some interesting features.
-The overall partial slope of the change in forest cover is -0.6, if all other variables are kept constant.
-This suggest quite strong change in streamflow, moving from fully forested to fully cleared.
-Over the whole forest cover range, this is a change of -120 mm, with other variables held constant.
+The overall partial slope of the change in forest cover is -0.61, if all other variables are kept constant.
+This suggest quite strong change in stream flow, moving from fully forested to fully cleared.
+Over the whole forest cover range, this is a change of -122 mm, with other variables held constant.
 This change is highly significant, as indicated by the low p-value.
 
-In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{Length} (p = 0.12)) explain variation in the data.
-For \emph{Length}, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in streamflow.
-
-Furthermore Table \ref{tab:m-all-linear} indicates that several of the assessment methods explain variation in the change in streamflow, which was also indicated in the correlation analysis.
+In addition, all the smoothed variables \emph{log10(Area (km\textsuperscript{2}))} (p = 0.02)), \emph{Dryness} (p = 0)) and \emph{Length} (p = 0.13)) explain variation in the data.
+For \emph{Length}, the p-value is not strictly smaller than 0.05, but still indicates some reasonable evidence that the variable explains some of the variation in the change in stream flow.
+
+Furthermore Table \ref{tab:m-all-linear} indicates that several of the assessment methods explain variation in the change in stream flow, which was also indicated in the correlation analysis.
 In particular, the assessment methods Paired Watersheds Experiments (PWE), Hydrological Modelling (HM) and Statistical modelling and hydrographs (SH) are important explaining variables (\(p < 0.05\)).
@@
 \endhead
-linear across range & 0.49 & 3182 \\
+linear across range & 0.49 & 3233 \\
 different for forestation and
-deforestation & 0.45 & 3227 \\
-non-linear across the range & 0.5 & 3182 \\
+deforestation & 0.45 & 3281 \\
+non-linear across the range & 0.51 & 3233 \\
 \bottomrule()
@@
 Figure \ref{fig:smoothsmodelall} highlights that the relationship between \emph{log10(Area km\textsuperscript{2})} and the change in flow is essentially linear.
-It indicates the negative slope that was also clear from \citet{zhang2017}, indicating that in larger catchments changes in forest cover have less impact on streamflow than for smaller catchments.
+It indicates the negative slope that was also clear from \citet{zhang2017}, indicating that in larger catchments changes in forest cover have less impact on stream flow than for smaller catchments.
 
@@
 Visually, the residuals appear approximately normal, although there is a noticable skew in a limited number of the data in the upper part of the distribution (Figure \ref{fig:gamcheckmodelall}, top left).
-This is related to a limited number catchments that have very high changes in streamflow in the data set.
+This is related to a limited number catchments that have very high changes in stream flow in the data set.
 In other words, the distribution of the residuals is somewhat fat-tailed.
@@
 
-The points related to catchments with very long studies (\textgreater{} 60 years) might be questionable, as changes other than forest cover change could affect streamflow.
+The points related to catchments with very long studies (\textgreater{} 60 years) might be questionable, as changes other than forest cover change could affect stream flow.
 In addition, a few of the catchments have Dryness values that are very large (\textgreater{} 5) and these values have high leverage in the data, affecting the residual distribution.
@@
 \endhead
-\textbf{(Intercept)} & -10.17 & 17.98 & -0.57 & 0.57 \\
-\textbf{DeltaF\_perc} & -0.59 & 0.08 & -7.45 & 0 \\
-\textbf{Forest\_SignIncrease} & 0.41 & 9.79 & 0.04 & 0.97 \\
-\textbf{Precip\_data\_typeOB} & -15.97 & 12.5 & -1.28 & 0.2 \\
-\textbf{Precip\_data\_typeSG} & 15.71 & 14.85 & 1.06 & 0.29 \\
-\textbf{Assessment\_techniqueEA, HM} & 20.38 & 41.03 & 0.5 & 0.62 \\
-\textbf{Assessment\_techniqueHM} & 26.42 & 11.4 & 2.32 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 28.51 & 12.15 & 2.35 & 0.02 \\
+\textbf{(Intercept)} & -10.69 & 17.85 & -0.6 & 0.55 \\
+\textbf{DeltaF\_perc} & -0.61 & 0.08 & -7.69 & 0 \\
+\textbf{Forest\_SignIncrease} & 1.12 & 9.72 & 0.12 & 0.91 \\
+\textbf{Precip\_data\_typeOB} & -16.23 & 12.46 & -1.3 & 0.19 \\
+\textbf{Precip\_data\_typeSG} & 15.96 & 14.86 & 1.07 & 0.28 \\
+\textbf{Assessment\_techniqueEA, HM} & 19.97 & 41.02 & 0.49 & 0.63 \\
+\textbf{Assessment\_techniqueHM} & 27.07 & 11.39 & 2.38 & 0.02 \\
+\textbf{Assessment\_techniquePWE} & 28.54 & 12.11 & 2.36 & 0.02 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 17.4 & 42.05 & 0.41 & 0.68 \\
-\textbf{Assessment\_techniqueQPW} & 41.49 & 19.53 & 2.12 & 0.03 \\
+HM} & 15.91 & 42.05 & 0.38 & 0.71 \\
+\textbf{Assessment\_techniqueQPW} & 43.3 & 19.52 & 2.22 & 0.03 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 24.81 & 23.32 & 1.06 & 0.29 \\
-\textbf{Assessment\_techniqueSH} & 47.26 & 11.49 & 4.11 & 0 \\
-\textbf{Forest\_typeCF} & -9.47 & 7.3 & -1.3 & 0.2 \\
-\textbf{Forest\_typeMF} & -6.01 & 7.35 & -0.82 & 0.41 \\
-\textbf{Hydrological\_regimeSD} & 2.5 & 8.89 & 0.28 & 0.78 \\
+EA} & 25.33 & 23.33 & 1.09 & 0.28 \\
+\textbf{Assessment\_techniqueSH} & 47.84 & 11.3 & 4.23 & 0 \\
+\textbf{Forest\_typeCF} & -7.89 & 7.22 & -1.09 & 0.27 \\
+\textbf{Forest\_typeMF} & -6.26 & 7.17 & -0.87 & 0.38 \\
+\textbf{Hydrological\_regimeSD} & 0.54 & 8.83 & 0.06 & 0.95 \\
 \bottomrule()
@@
 \endhead
-\textbf{s(Dryness)} & 4.02 & 9 & 2.16 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.87 & 4 & 1.53 & 0.01 \\
-\textbf{s(Length)} & 0 & 9 & 0 & 0.98 \\
+\textbf{s(Dryness)} & 3.82 & 9 & 2.15 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.89 & 4 & 1.66 & 0.01 \\
+\textbf{s(Length)} & 0 & 9 & 0 & 0.97 \\
 \bottomrule()
@@
 Therefore it is worth investigating what effect removing these few data points has on the overall model and the significance of the variables.
-Data that have \emph{Dryness} \textless= 5 and \emph{Length} \textless= 60 years were removed from the dataset and the model based on a reduction of the data set from 329 to 310 catchments is run again.
-
-This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.44 and a deviance explained of 0.48.
-
-Investigating the non-linear responses suggest that \emph{Dryness} has a clear non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on streamflow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
+Data that have \emph{Dryness} \textless= 5 and \emph{Length} \textless= 60 years were removed from the data set and the model based on a reduction of the data set from 334 to 315 catchments is run again.
+
+This model, which excludes data with long studies and very dry catchments explains only slightly less of the variation with an adjusted \emph{r\textsuperscript{2}} of 0.45 and a deviance explained of 0.48.
+
+Investigating the non-linear responses suggest that \emph{Dryness} has a clear non-linear response, which is significant, where changes in forest cover in drier catchments having a greater impact on stream flow (Figure \ref{fig:smoothsmodelredLength} and Table \ref{tab:restrictlength}).
 Catchment area (\emph{log10(Area (km\textsuperscript{2}))}) still has an impact on flow with p = 0.01, and the relationship looks almost linear.
@@
 \endhead
-PWE & 185 \\
+PWE & 187 \\
 HM & 57 \\
-SH & 42 \\
+SH & 45 \\
 EA & 32 \\
@@
 Therefore, the model was rerun excluding the combined assessment techniques (EA, HM), (PWE, HM) and (QPW, EA) and the assessment technique QPW, which were all non-significant (Table \ref{tab:tableassess}).
-This resulted in a data set of 323 catchment studies.
+This resulted in a data set of 328 catchment studies.
 
@@
 
-The generalised additive models appear to reach the same conclusions as the single variable regression in earlier papers \citep{zhang2017, filoso2017}. It appears that:
+The generalized additive models appear to reach the same conclusions as the single variable regression in earlier papers \citep{zhang2017, filoso2017}. It appears that:
 
@@
 \item
-  Larger catchments show lower impact of forest cover change on streamflow;
-\item
-  Drier catchments show a greater impact of forest cover change on streamflow; and
-\item
-  There is a general linear relationship between the change in forest cover and the change in streamflow.
+  Larger catchments show lower impact of forest cover change on stream flow;
+\item
+  Drier catchments show a greater impact of forest cover change on stream flow; and
+\item
+  There is a general linear relationship between the change in forest cover and the change in stream flow.
 \end{enumerate}
@@
 \item
-  There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest \% forest cover change and the largest variability in the flow response.
+  There is a clear relationship between size of catchments, area cleared and type of experiments, with particular Paired Watershed Experiments containing the smallest catchments, the largest percent forest cover change and the largest variability in the flow response.
 \end{enumerate}
@@
 
-It is possible that one of the reasons why \citet{zhang2017} separated their analysis in large (\textgreater{} 1000 km\textsuperscript{2}) and small (\textless{} 1000 km\textsuperscript{2}) catchments, is that they realised this difference in assessment methods and wanted to account for this.
+It is possible that one of the reasons why \citet{zhang2017} separated their analysis in large (\textgreater{} 1000 km\textsuperscript{2}) and small (\textless{} 1000 km\textsuperscript{2}) catchments, is that they realized this difference in assessment methods and wanted to account for this.
 However, this is not explicitly identified, and there is no real physical explanation of the 1000 km\textsuperscript{2} threshold.
@@
 Similarly, the data for the larger catchments containing smaller changes in forest cover are dominated by hydrological modelling studies, resulting in a further complication.
-If the reponse of the streamflow in the modelling studies is the result of the conceptualised relationship between streamflow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in streamflow is real, or simply a result of a pre-conceived model relationship.
-Is the smaller variation in the data for smaller changed in forest cover (Figure \ref{fig:overview}) a result of similar conceptualised model relationships, or actual variation between catchments and climate types?
+If the reponse of the stream flow in the modelling studies is the result of the conceptualized relationship between stream flow and forest cover (possibly from a subset of the paired catchment studies), then it is impossible to say if the change in stream flow is real, or simply a result of a pre-conceived model relationship.
+Is the smaller variation in the data for smaller changed in forest cover (Figure \ref{fig:overview}) a result of similar conceptualized model relationships, or actual variation between catchments and climate types?
 Currently this question cannot be answered.
@@
 This becomes problematic when extrapolated to larger scales.
-A clear example of this is the paper by \citet{hoekvandijke2022} where the conceptualised relationship between forest cover and streamflow pre-determines the outcomes of the global modelling.
-
-The only way to analyze changes in streamflow as a function of forest cover in larger catchments is to actually derive this from observed data of long term streamflow and forest cover (as was done in \citet{levy2018}).
+A clear example of this is the paper by \citet{hoekvandijke2022} where the conceptualized relationship between forest cover and stream flow pre-determines the outcomes of the global modelling.
+
+The only way to analyze changes in stream flow as a function of forest cover in larger catchments is to actually derive this from observed data of long term stream flow and forest cover (as was done in \citet{levy2018}).
 
@@
 In this case, the total \% area of forest was not included in the data.
-As a test, the total \% area of forest for the larger catchments (\textgreater{} 1000 km\textsuperscript{2} in \citet{zhang2017}) were added to the dataset and the model for just the large catchments was tested.
-This showed that the \% area of forest was not significant explaining the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
-While this might be an area of further research on the full dataset, it is complicated for two reasons:
+As a test, the total \% area of forest for the larger catchments (\textgreater{} 1000 km\textsuperscript{2} in \citet{zhang2017}) were added to the data set and the model for just the large catchments was tested.
+This showed that the \% area of forest was not significant to explain the change in flow for the larger catchments (retaining all other variables in the model, results not shown).
+While this might be an area of further research on the full data set, it is complicated for two reasons:
 
@@
   Another example is the paper by \citet{waterloo2007}.
-This modelling study in Fiji of the clearing of a catchment reports the changes in streamflow over parts of the year.
-For a period of 324 days the streamflow increased from 252 mm to 580 mm (a 230\% increase if calculated as \(580/252*100\)) and for a second period of 309 days the streamflow increased from 90 mm to 194 mm (a 215 \% increase).
+This modelling study in Fiji of the clearing of a catchment reports the changes in stream flow over parts of the year.
+For a period of 324 days the stream flow increased from 252 mm to 580 mm (a 230\% increase if calculated as \(580/252*100\)) and for a second period of 309 days the stream flow increased from 90 mm to 194 mm (a 215 \% increase).
 However, how we convert this to a change in annual flow (which most of the other data relate to) is difficult.
@@
 We suspect that results from \(252/580*100 \approx 50\) and \(90/194 \approx 50\).
+\item
+  A final example is around the choice of control or treatment.
+In the data base the assumption is that the change in flow is relative to the original situation.
+This can be either a ``before and after'' analysis, which can be problematic by itself due to climate variation, or a comparison of a treatment with a control.
+But even in this case comparing across catchments can be tricky.
+For example, at one extreme, some controls in the database are a bamboo catchment and a tea plantation \citep{blackie1979kimakia, blackie1979kericho}.
+A clearer example is the Brigalow catchment study in Queensland, Australia \citep{thornton2007brigalow}, catchment \#336 in the data set.
+This is a paired watershed experiment of conversion of native Brigalow (an Acacia species) into cropland and pasture.
+We chose to use the cropped catchment (C2 \citet{thornton2007brigalow}) as the deforestation treatment resulting in a change of flow of 140\%, based on Table 4 in \citet{thornton2007brigalow}.
+However, had we used the pasture catchment as the treatment then the change in flow would have been up to 165\%.
 \end{itemize}
@@
 As a result, it becomes difficult to assess how we interpret the change in flow as a result of a change in cover.
-In many cases we would expect the flow to change over time as a function of the recovery \citep{jones2017} and therefore the timeseries of the flow needs to be assessed over a longer time.
+In many cases we would expect the flow to change over time as a function of the recovery \citep{jones2017} and therefore the time series of the flow needs to be assessed over a longer time.
 
@@
 An additional complication related to combining studies related to wild fires or bush fires and logging studies is the differences in vegetation recovery.
-For example, \citet{heath2014} found that catchments with resprouting species around Sydney, Australia, indicated little change in the streamflow in comparison to species regrowing from seed further south on the continent \citep{zhou2015bushfire}.
+For example, \citet{heath2014} found that catchments with resprouting species around Sydney, Australia, indicated little change in the stream flow in comparison to species regrowing from seed further south on the continent \citep{zhou2015bushfire}.
 
 As a result, it can be difficult to exactly pinpoint the change in flow as a result of the change in cover, as well as being difficult to assess what the exact change in cover actually was.
-
-As indicated before, if the overall variation due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
+In addition, using only the overall change in stream flow can discard a lot of information from observations in individual years.
+Many papers give stream flow values for multiple years, often showing significant variation.
+Summarizing this into one average value discards all the information on the variance.
+
+As indicated before, if the overall variation in the data due to this issues is small, then this would not be an issue for upscaling the results, but the large variation for the smaller catchments suggest that effects could be considerable.
 As \citet{jones2017} indicate, this really needs time series analysis of the different experiments.
@@
 In the end, careful review of the data and the original papers can circumvent most of this issue.
-And, making the data available (as \citet{zhang2017}, \citet{zhou2015} and \citet{filoso2017} have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
+And, making the data available (as \citet{zhang2017}, \citet{zhou2015}, \citet{filoso2017} and this paper have done) provides an opportunity for review by other researchers, and over time most of the transcription errors can be resolved.
 
@@
 
-In this paper, a few studies have been singled out in the analysis.
-The choice of focus was mainly driven by the data that was made available by the authors of these papers \citep{zhang2017, filoso2017}, which provide a rich case study for the current paper.
+In this paper, a few studies have been singled out to demonstrate the main point that extrapolating individual studies to global scales is not that simple in natural systems.
+Not only is there significant natural variation and latent variables, interpretation and aggregation can cause further unforeseen problems.
+The choice of papers to focus on was mainly driven by the data that was made available by the authors of these papers \citep{zhang2017, filoso2017}, which provide a rich case study for the current paper.
 
@@
 This once again highlights that there is unexplained variation at the extremes of the distribution, once again related to the paired watershed experiments (Figure \ref{fig:assessment}).
-Generally,in statistical models, the approach would be to further normalise the residuals through transformations.
+Generally,in statistical models, the approach would be to further normalize the residuals through transformations.
 However, in this case this might be difficult and might not resolve all the issues due to the large variation in the data.
@@
 For example, the current database is mainly related to forest harvest, bush fire and reforestation/plantation management.
-It is tempting to use the result of a large scale analysis of this data to make inferences about overall landuse change \citep{li2017, wang2020}, but this would not be valid, as the deforestation studies are generally not a transition to an agricultural landuse or pasture, but regrowing into forest.
+It is tempting to use the result of a large scale analysis of this data to make inferences about overall land use change \citep{li2017, wang2020}, but this would not be valid, as not all deforestation studies are a transition to an agricultural land use or pasture, as in \citet{levy2018}.
+Many are logging or bushfire studies regrowing into forest after the initial treatment.
 Similarly, using the plantation studies to extrapolate to ``reforestation'' (as in \citet{filoso2017} and \citet{hoekvandijke2022}) is also tenuous.
@@
 Given the link between ecophysiology and water and carbon budgets \citep{jackson2005}, care should be taken in extrapolation, introducing a further error.
+
+As highlighted summarizing highly variable observed time series into single mean values introduces further potential problems, as well as discarding information about the variance.
 
@@
 Given the large number of studies that look at a time evolution of forest cover (i.e.~either clearing and regrowth, or burning and regrowth), this data can offer further insights into the dynamic response of catchments to changes in land cover.
-As highlighted, some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
+In addition, this might allow analysis of the variance of the response in addition to the mean response.
+While some of the older data is not fully recoverable, but there is often a series of papers related to one experiment, which at least would provide individual time points.
 
 More generally there is a clear need for a more in depth analysis of the data base of studies used here.
-In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, streamflow responses and responses of components of streamflow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
+In particular, more detailed data can potentially be extracted from many of the studies in terms of vegetation species, stream flow responses and responses of components of stream flow (slow flow, quick flow etc.), as well as a more in depth description of the management and actual experimental design.
 
@@
 Extrapolating paired watershed experiment results into models can possibly overlook landscape interactions that are visible at larger scales, but do not occur on smaller scales.
-For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on landuse.
-A carefully designed simulation study that specifically investigates the change in stream flow repsonse with scale using local field data for verification can help solve this problem.
-
-At the moment, providing answers to the impact of streamflow at larger scales should generally not be approached by simulation modelling.
-A better approach is analyzing streamflow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include landuse changes.
-The highlighted paper by \citet{levy2018} is currently the best example of a solid statistical approach to analyzing streamflow responses.
+For example, this could be the effects of lateral flow and groundwater connectivity and impacts of elevation on land use.
+A carefully designed simulation study that specifically investigates the change in stream flow response with scale using local field data for verification can help solve this problem.
+
+At the moment, providing answers to the impact of stream flow at larger scales should generally not be approached by simulation modelling.
+A better approach is analyzing stream flow data at multiple spatial and temporal scales for responses (rather than running simulations) and using satellite data to dynamically include land use changes.
+The highlighted paper by \citet{levy2018} is currently the best example of a solid statistical approach to analyzing stream flow responses.
 Simulation modelling can be an approach to analyze different scenarios, if there is clear recognition of the potential impact of the model structure (the algorithms and parameters that describe for example plantation tree growth) on the simulation outcomes.
@@
 
-This study demonstrates that analyzes of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
-
-While the analysis reveals similar conclusions in relation to the response of streamflow to forest cover, there are four major interlinked reasons why these results should be considered carefully.
-This subsequently has implications for meta-analyses in Environmental Science and Hydrology in general.
-The reasons higlighted in this paper are:
+This study demonstrates that the analysis of large databases of essentially ``aggregated data'' should be considered carefully and simple single variable regressions often present simplistic relationships that can be misleading.
+
+On first glance, stream flow will decrease with increases in forest cover, and increase with removal of forest cover.
+However, the conclusions on the exact relationship between stream flow change and forest cover change are highly uncertain.
+There are four major interlinked reasons why earlier conclusions should be considered carefully:
 
@@
 \item
-  The existence of latent variables in the data that create the appearance of a relationship that really does not exist;\\
-\item
-  The difficulty in fully interpreting the specifics of different studies;\\
-\item
-  The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and\\
+  The existence of latent variables in the data that create the appearance of a relationship that really does not exist.
+In this case the assessment technique used to assess the change in stream flow is highly significant;\\
+\item
+  The difficulty in fully interpreting the specifics of different studies. In this case the definition of vegetation, stream flow volume, control or time period was difficult to assess;\\
+\item
+  The difficulty of integrating data from seemingly similar studies, but with quite different objectives.
+This study highlighted the many logging studies followed by regrowth, or bushfire effects; and\\
 \item
@@
 
+While some of these issues might be overcome with careful analysis and transcription, not all can be directly or easily resolved.
 Any statistical analysis, including the one in this paper, needs to be considered ``conditional on the data'', and given the issues indicated, extrapolation of the results of summary studies to larger scales and into global hydrological models has to be done with great care.
 Better would be to analyze observed data and explicitly include uncertainty in the extrapolation of the results.
-
 This therefore has implications for the recent growth in meta-analysis review papers, which has been boosted by increased computational capacity and much better on-line accessible data bases with research data.
-Clearly, this requires careful definition of the search terms, and follow-up review of the harvested papers, as well as an understanding that the statistical relationships can be hiding other unknown factors.
-As the old adagium says: Correlation is not the same as causation.
+This is particularly true for natural systems involving climate variability and therefore extrapolations of experimental work in Environmental Science and Hydrology to global scales in general.
 

### 2022-10-07 paper/Forest_and_Water/SupplementaryMaterialPart1.Rmd

@@
 
-This supplementary material is related to 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
+This supplementary material is related to 'Generalizing the impact of forest cover on streamflow from experimental data: it is not that simple. Vervoort et al.'
 
@@
 The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively.
-So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%.
-corrected to 16\%.
+So the value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%.
+Corrected to 16\%.
 - 3, Baker Creek, @zhangwei2012. The original recorded 201.1\% change by @zhang2017 also seems wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected  

### 2022-10-07 paper/Forest_and_Water/SupplementaryMaterialPart3.tex

--- 2022-08-18 paper/Forest_and_Water/SupplementaryMaterialPart3.tex
@@
 \endhead
-\textbf{(Intercept)} & -7.47 & 15.87 & -0.47 & 0.64 \\
-\textbf{DeltaF\_perc} & -0.59 & 0.05 & -10.89 & 0 \\
-\textbf{Precip\_data\_typeOB} & -17.87 & 12.91 & -1.38 & 0.17 \\
-\textbf{Precip\_data\_typeSG} & 0.2 & 14.82 & 0.01 & 0.99 \\
-\textbf{Assessment\_techniqueEA, HM} & 18.67 & 41.7 & 0.45 & 0.65 \\
-\textbf{Assessment\_techniqueHM} & 26.61 & 11.43 & 2.33 & 0.02 \\
-\textbf{Assessment\_techniquePWE} & 30.77 & 11.68 & 2.63 & 0.01 \\
+\textbf{(Intercept)} & -5.37 & 16.19 & -0.33 & 0.74 \\
+\textbf{DeltaF\_perc} & -0.61 & 0.06 & -11.03 & 0 \\
+\textbf{Precip\_data\_typeOB} & -21.34 & 13.16 & -1.62 & 0.11 \\
+\textbf{Precip\_data\_typeSG} & 9.57 & 15.16 & 0.63 & 0.53 \\
+\textbf{Assessment\_techniqueEA, HM} & 20.32 & 42.72 & 0.48 & 0.63 \\
+\textbf{Assessment\_techniqueHM} & 23.51 & 11.69 & 2.01 & 0.05 \\
+\textbf{Assessment\_techniquePWE} & 30.71 & 11.92 & 2.58 & 0.01 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 15.8 & 42.22 & 0.37 & 0.71 \\
-\textbf{Assessment\_techniqueQPW} & 41.35 & 19.66 & 2.1 & 0.04 \\
+HM} & 15.79 & 43.24 & 0.37 & 0.72 \\
+\textbf{Assessment\_techniqueQPW} & 41.29 & 20.14 & 2.05 & 0.04 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 26.05 & 23.84 & 1.09 & 0.28 \\
-\textbf{Assessment\_techniqueSH} & 39.31 & 11.54 & 3.41 & 0 \\
-\textbf{Forest\_typeCF} & -9.28 & 7.41 & -1.25 & 0.21 \\
-\textbf{Forest\_typeMF} & -6.34 & 7.38 & -0.86 & 0.39 \\
-\textbf{Hydrological\_regimeSD} & 0.13 & 8.94 & 0.01 & 0.99 \\
+EA} & 25.16 & 24.41 & 1.03 & 0.3 \\
+\textbf{Assessment\_techniqueSH} & 46.03 & 11.65 & 3.95 & 0 \\
+\textbf{Forest\_typeCF} & -7.76 & 7.52 & -1.03 & 0.3 \\
+\textbf{Forest\_typeMF} & -7.8 & 7.35 & -1.06 & 0.29 \\
+\textbf{Hydrological\_regimeSD} & 1.5 & 9.1 & 0.17 & 0.87 \\
 \bottomrule()
@@
 The variable \texttt{DeltaF\_perc} clearly explains the variation in the change in flow with a p-value of 0.
-The overall model explains 50 \% of the variation in the data.
+The overall model explains 49 \% of the variation in the data.
 For comparison across models with different structures and different variables, the Aikaike Information Criteria (AIC) can be used.
-The AIC of the model is 3166.6801428.
+The AIC of the model is 3233.1130843.
 
@@
 \endhead
-\textbf{(Intercept)} & 10.91 & 19.47 & 0.56 & 0.58 \\
-\textbf{DeltaF\_perc\_pos} & 0.29 & 0.1 & 2.97 & 0 \\
-\textbf{Forest\_SignIncrease} & -52.26 & 7.62 & -6.86 & 0 \\
-\textbf{Precip\_data\_typeOB} & -15.75 & 14.22 & -1.11 & 0.27 \\
-\textbf{Precip\_data\_typeSG} & -5.2 & 16.81 & -0.31 & 0.76 \\
-\textbf{Assessment\_techniqueEA, HM} & 14.21 & 45.49 & 0.31 & 0.76 \\
-\textbf{Assessment\_techniqueHM} & 27 & 13.31 & 2.03 & 0.04 \\
-\textbf{Assessment\_techniquePWE} & 28.19 & 13.57 & 2.08 & 0.04 \\
+\textbf{(Intercept)} & 13.99 & 19.79 & 0.71 & 0.48 \\
+\textbf{DeltaF\_perc\_pos} & 0.3 & 0.1 & 2.98 & 0 \\
+\textbf{Forest\_SignIncrease} & -53.09 & 7.71 & -6.88 & 0 \\
+\textbf{Precip\_data\_typeOB} & -19.15 & 14.46 & -1.32 & 0.19 \\
+\textbf{Precip\_data\_typeSG} & 4.93 & 17.19 & 0.29 & 0.77 \\
+\textbf{Assessment\_techniqueEA, HM} & 18.4 & 46.6 & 0.39 & 0.69 \\
+\textbf{Assessment\_techniqueHM} & 22.56 & 13.55 & 1.67 & 0.1 \\
+\textbf{Assessment\_techniquePWE} & 27.64 & 13.79 & 2 & 0.05 \\
 \textbf{Assessment\_techniquePWE,
-HM} & 16.36 & 45.51 & 0.36 & 0.72 \\
-\textbf{Assessment\_techniqueQPW} & 28.63 & 21.65 & 1.32 & 0.19 \\
+HM} & 16.49 & 46.72 & 0.35 & 0.72 \\
+\textbf{Assessment\_techniqueQPW} & 27.18 & 22.18 & 1.23 & 0.22 \\
 \textbf{Assessment\_techniqueQPW,
-EA} & 35.76 & 25.74 & 1.39 & 0.17 \\
-\textbf{Assessment\_techniqueSH} & 33.52 & 12.94 & 2.59 & 0.01 \\
-\textbf{Forest\_typeCF} & -12.47 & 8.06 & -1.55 & 0.12 \\
-\textbf{Forest\_typeMF} & -5.1 & 8.18 & -0.62 & 0.53 \\
-\textbf{Hydrological\_regimeSD} & 3.45 & 9.76 & 0.35 & 0.72 \\
+EA} & 33.53 & 26.38 & 1.27 & 0.2 \\
+\textbf{Assessment\_techniqueSH} & 39.26 & 13.06 & 3.01 & 0 \\
+\textbf{Forest\_typeCF} & -10.87 & 8.18 & -1.33 & 0.19 \\
+\textbf{Forest\_typeMF} & -7.36 & 8.15 & -0.9 & 0.37 \\
+\textbf{Hydrological\_regimeSD} & 4.61 & 9.93 & 0.46 & 0.64 \\
 \bottomrule()
@@
 However, the results also show that in this case, the deforestation (\texttt{Forest\_Sign} \textless{} 0) has a different effect than reforestation (\texttt{Forest\_Sign} \textgreater{} 0).
-In fact, in absolute terms deforestration creates a -5226 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).
+In fact, in absolute terms deforestration creates a -5309 \% greater effect on streamflow than reforestation (and of course deforestation creates and increase in flow and reforestation a decrease).
 
-However, overall this model explains less of the overall variation in the data with a deviance explained of 0.46 and an AIC of 3212.700798.
+However, overall this model explains less of the overall variation in the data with a deviance explained of 0.45 and an AIC of 3281.055352.
 
@@
 \endhead
-\textbf{s(DeltaF\_perc)} & 1.11 & 9 & 12.72 & 0 \\
-\textbf{s(log10(Area\_km2))} & 0.76 & 4 & 0.8 & 0.03 \\
-\textbf{s(Dryness)} & 4.7 & 9 & 2.2 & 0 \\
-\textbf{s(length)} & 4.45 & 34 & 0.24 & 0.09 \\
+\textbf{s(DeltaF\_perc)} & 4.48 & 9 & 14.11 & 0 \\
+\textbf{s(log10(Area\_km2))} & 0.77 & 4 & 0.81 & 0.03 \\
+\textbf{s(Dryness)} & 4.7 & 9 & 2.42 & 0 \\
+\textbf{s(length)} & 4.65 & 34 & 0.25 & 0.08 \\
 \bottomrule()
@@
 In this case, the non-linear terms of the model need to be checked, which shows that once again, changes in forest cover impact changes in flow as indicated by the p-value of 0.
-The total deviance explained for the model is 0.5 and the model has an AIC of 3166.9317897.
+The total deviance explained for the model is 0.51 and the model has an AIC of 3232.9048803.
 These performance statistics are similar to the first (baseline) model.
@@
 ## Method: GCV   Optimizer: magic
-## Smoothing parameter selection converged after 15 iterations.
-## The RMS GCV score gradient at convergence was 0.004151441 .
+## Smoothing parameter selection converged after 10 iterations.
+## The RMS GCV score gradient at convergence was 0.005535096 .
 ## The Hessian was positive definite.
@@
 ##                        k'    edf k-index p-value  
-## s(DeltaF_perc)      9.000  1.113    0.89    0.03 *
-## s(log10(Area_km2))  4.000  0.756    0.96    0.20  
-## s(Dryness)          9.000  4.703    0.94    0.13  
-## s(length)          34.000  4.448    0.92    0.10 .
+## s(DeltaF_perc)      9.000  4.476    0.91   0.080 .
+## s(log10(Area_km2))  4.000  0.766    0.94   0.095 .
+## s(Dryness)          9.000  4.698    0.94   0.100 .
+## s(length)          34.000  4.645    0.94   0.090 .
 ## ---
